### 2024-09-20 paper/paper.Rmd
start repo

---
title: "A multi-stock, multi-region state-space age-structured assessment model with an application to black sea bass"
author: 
  - Timothy J. Miller^1^
  - Kierstin Curti
  - Alex Hansell
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
    includes:
      in_header: options.sty
  html_document:
    df_print: paged
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
bibliography: bsb-wham.bib
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=FALSE, tidy.opts=list(width.cutoff=60), warning = FALSE, message = FALSE, fig.align = "center")
#knitr::opts_knit$set(root.dir = '../')
library(knitr)
library(tidyverse)
library(pander)
library(kableExtra)
library(wham)
#library(ggplotFL)
library(viridis)
type <- "latex"
if(knitr::is_latex_output()) type <- "latex"
if(knitr::is_html_output()) type <- "html"
```

$^1$timothy.j.miller@noaa.gov, Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\

\pagebreak

## Main Message {-}

A multi-region, multi-stock generalization of WHAM with an application evaluating evidence of alternative hypotheses about temperature effects on black sea bass

## Abstract {-}

The Woods Hole Assessment Model (WHAM) is a state-space age-structured assessment model that is used to assess and manage many stocks in the Northeast US. We first describe a multi-stock, multi-region extension of WHAM that treats the population and fleet dynamics seasonally and allows movement by season and region to be functions of time- and age-varying autocorrelated random effects and environmental covariates. We then illustrate the model with northern and southern components of the NEUS black sea bass stock and evaluate alternative hypotheses of bottom temperature and random effects on recruitment and natural mortality. We show strong evidence for temperature effects on recruitment and no random effects or temperature effects on natural mortality, primarily for the northern stock component.

\pagebreak

# Introduction {-}

Why is this model needed.  Models ignoring movement can provide biased catch advice.  Using random effects for time-varying processes is considered  best practice.

The Woods Hole Assessment Model (WHAM) is an R package developed at the Northeast Fisheries Science Center [https://timjmiller.github.io/wham, @millerstock20; @stockmiller21]. Primary inputs and observation types are similar to ASAP and there is functionality built into WHAM to move information from ASAP input files to R. WHAM can be configured to fit SCAA models often identically to ASAP, so that existing assessments in the U.S. Northeast can be replicated and tested against state-space models with process errors and environmental effects in a single framework. WHAM models are built using the Template Model Builder package [TMB, @kristensenetal16] which provides efficient fitting of models that include random effects similar to the State-space Assessment Model [SAM, @nielsenberg14], which is used to manage many stocks within ICES. WHAM is currently used to manage 9 stocks in the NEUS.

The Woods Hole Assessment Model (WHAM) software package is maintained and developed at the Northeast Fisheries Science Center to enable state-space stock assessments, i.e. where processes such as the annual transitions in numbers-at-age (survival), natural mortality, selectivity, and catchability are treated as time- and(or) age-varying random effects. WHAM can also be configured without random effects as a traditional statistical catch-at-age model in order to bridge from current assessments which use Age-Structured Assessment Program (ASAP). Here, we present an extension of WHAM (Multi-WHAM) to model multiple stocks and(or) regions and allow movement of stocks among regions.

It has been reviewed in the literature and simulation tested [@stocketal21; @stockmiller21]. It has also been used as the operating model in the index-based assessment methods research track assessment [@legaultetal23] and is also the main tool for simulation studies in the research track on state-space assessment models.

A review of the essential features for next-generation stock assessments concluded that only WHAM and a similar State-Space Assessment Method (Nielsen and Berg 2014) model random effects correctly (Punt et al. 2020). WHAM is now being used as the management model for 9 stocks in the northeast U.S., but the standard version of WHAM can only be applied to a single stock and area (Miller and Stock 2020).

Here, we describe a multi-stock and multi-region extension of the WHAM package (version 2.0 https://github.com/timjmiller/wham/tree/lab). Numerous new configuration options are also included in this extension that can be useful even for single-stock or single region models. Any number of stocks or regions can be configured in this version of WHAM. The multi-stock extension was developed to accommodate movement of stocks or stock components in anticipation of black sea bass and Atlantic cod stocks where new modeling frameworks where recently examined through Northeast research track assessment process (cite RT docs). We demonstrate capabilities of the new version of WHAM with black sea bass and evaluate evidence for alternative hypotheses of temporal variation in natural mortality and effects of bottom temperature on recruitment and natural mortality of age 1 individuals.

# Methods {-}

## Model description {-}

Many of the options and equations of version 2.0 are are the same as those described by @stockmiller21, so we will only describe extensions and differences that have occurred since the first description of WHAM. The new version of WHAM can model multiple stocks, but survival, movement, harvest and natural mortality is are tracked for each stock. Therefore, much of the description below assumes a specific stock $s$, but this subscript is only explicitly used when necessary.

### The probability transition matrix {-}

Because individuals may be alive in one of several regions or harvested in one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model (reference). Multi-state models use a probability transition matrix (PTM) that describes the probability of individuals living and moving among regions or dying due to fishing or natural mortality over a time interval $i$ with duration $\delta_i$ in year $y$ for individuals at age $a$ on January 1. Each row and column of the PTM correspond to one the states: alive in region $r$, dead in fleet $f$, or dead from natural causes. The probabilities in each row sum to unity and assume an individual is in the corresponding state at the beginning of the interval. Given $n_R$ regions and $n_F$ fleets, the square PTM ($n_R + n_F + 1$ rows and columns) as a function of sub-matrices is
\begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix}
\end{equation}
where 
\begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*}
is the $n_R \times n_R$ matrix defining survival and occurring in each region at the end of the interval,
\begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*}
is the $n_R \times n_F$ matrix defining probabilities of being captured in each fleet during the interval, and $\mathbf{D}_{y,a,i}$ is the $n_R x 1$ matrix of probabilities of dying due to natural mortality during the interval. We have the identity matrix $\mathbf{I}_{H}$ for the states for  capture by each fleet and a 1 for the state for natural mortality because the probabilities of being in one of the mortality states given starting the interval in that state is unity.

WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would be multinomial distributed. The PTMs determine the expected numbers 1) in each state on January 1 of year $t+1$ at age $a+1$ given the abundances at age $a$ on January 1 of year $t$, 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.

### Single region PTMs {-}

When there is only one region,
\begin{equation}\label{eq:ptm_1_region}
\mathbf{P}_{y,a,i} = 
  \begin{bmatrix}
     S_{y,a,i} & \mathbf{H}_{y,a,i} & D_{y,a,i} \\
     0 & \mathbf{I}_{H} & 0\\
     0 & 0 & 1
  \end{bmatrix}
\end{equation}
where $S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}$, $\mathbf{H}_{y,a,i}$ is a 1 x $n_F$ matrix with elements for each fleet $f$: $\frac{F_{y,a,i,f}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)$, $D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)$, and $Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{y,a,i,f}$ is the total mortality rate.

### Multi-region PTMs {-}

When there is more than 1 region, WHAM can model survival and movement as processes occurring sequentially or simultaneously. The sequential assumption is used widely in spatially explicit model (e.g., SS3). Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval. WHAM is configured to have movement occur after survival and mortality:
\begin{equation*}
  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i}
\end{equation*}
where $\mathbf{S}_{y,a,i}$ is a $n_R \times n_R$ diagonal matrix of proportions surviving in each region (given they start in that region) 
\begin{equation*}
\mathbf{S}_{y,a,i} = 
  \begin{bmatrix}
    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
  \end{bmatrix}
\end{equation*}
and $\boldsymbol{\mu}_{y,a,i}$ is a $n_R \times n_R$ matrix of probabilities of moving from one region to another or staying in the region they occurred at the beginning of the interval: 
\begin{equation*}
\boldsymbol{\mu}_{y,a,i} = 
  \begin{bmatrix}
    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
    \vdots & \vdots & \ddots & \vdots \\
    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
  \end{bmatrix}
\end{equation*}

WHAM assumes each fleet $f$ can harvest in only 1 region ($r_f$) during specified seasons. So, for each fleet $f$, row $r_f$ and column $f$ of $\mathbf{H}_{y,a,i}$ will be $F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$ when fleet $f$ is harvesting during the interval $\delta_i$ and all other elements will be zero. Row $r$ of the single-column matrix $\mathbf{D}_{y,a,i}$ is  $M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$

When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates. We obtain the probability transition matrix over an interval $\delta_i$ by exponentiating the instantaneous rate matrix [@millerandersen08]
\begin{equation*}
\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
\end{equation*}
The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region. Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero. For two regions and one fleet operating in each region:
 \begin{equation*}
 \mathbf{A}_{y,a,i} = \begin{bmatrix}
 a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
 \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
 0 & 0 & 0 & 0 & 0 \\
 0 & 0 & 0 & 0 & 0 \\
 0 & 0 & 0 & 0 & 0
 \end{bmatrix}.
\end{equation*}
where $a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})$. When there is one region, $n_f$ fleets, and $\delta_i = 1$, exponentiating the instantaneous rate matrix results in the PTM defined in Eq. \ref{eq:ptm_1_region}.

### Seasonality {-}

Seasonality can be configured to accommodate characteristics of spawning, movement, and fleet-specific behavior. The annual time step can be divided into $K$ (any number) seasons and the interval size $\delta_i$ for each season $i$ need not be equal to any other seasonal interval. Under the Markov assumption, the PTM of surviving and moving and dying over $K$ intervals $\delta_1,\ldots, \delta_K$ (i.e., the entire year) is just the product of the PTMs for each interval:
$$ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).$$

For a stock spawning at some fraction of the year  $0<t_s<1$ in interval $\delta_j$, the fraction of time in season $j$ is 
$$\delta_{s,j} = t_s-\sum^{j-1}_{i=0}\delta_i
$$
and the PTM defining the proportions in each state at time $t_s$ for age $a$ is
\begin{equation}\label{eq:ptm_spawn}
\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{s,j}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{s,j}).
\end{equation}
Similarly, for an index $m$ occurring  at fraction of the year $t_m$ in interval $\delta_j$ the proportions in each state at the time of the observation is
\begin{equation} \label{eq:ptm_index} 
\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{m,j}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{m,j}).
\end{equation}

<!--\textbf{Still needs work because it is slow.}-->

### Numbers at age {-}

When there are $n_R$ regions and $n_F$ fleets, The vector of abundance in each state at age $a>1$ on January 1 is $\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}', \mathbf{0}')'$ where $\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})'$ is the number in the states corresponding to being alive in each region and $\mathbf{0}$ is a vector ($n_F+1$) for the numbers captured in each fleet and dead from natural mortality because no age $a$ fish have died yet on January 1.

Each stock $s$ is assumed to spawn and recruit in one region $r_s$. So for age $a=1$, $\mathbf{N}_{O,y,1}$ is 0 except for row $r = r_s$. Options for configuring recruitment ($N_{y,1,r_s}$) for each stock are the same as previous versions of WHAM. If recruitment is assumed to be a function of spawning stock biomass (SSB), it is only the spawning population in region $r_s$ at the time of spawning constitutes the SSB in the stock-recruit function. However, models can configure spawning individuals to occur in other regions at the time of spawning. Aside from treating recruitment as a random walk, the general model for annual recruitment as random effects is
\begin{equation*}
\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1,r_s} =  f\left(\text{SSB}_{y-1,r_s}\right) + \varepsilon_{y,1,r_s}
\end{equation*}
where  
$$\text{SSB}_{s,y} =  \sum^A_{a=1}  w_{s,y,a} \text{mat}_{s,y,a} \mathbf{O}_{s,y,a,r_s}(t_s)' \mathbf{N}_{O,s,y,a}$$
where $w_{s,y,a}$ is the mean weight at age of spawning individuals, $\text{mat}_{s,y,a}$ is the maturity at age, and $\mathbf{O}_{s,y,a,r_s}(t_s)$, the $r_s$ column of the upper-left submatrix of Eq. \ref{eq:ptm_spawn}, are the probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year.

As in previous versions of WHAM, the transitions in numbers at age from one year to another after recruitment can be treated deterministically or as functions of random effects. The predicted numbers at age in year $y$ at age $a$ for a given stock are vector analogs ($\mathbf{N}_{O,y,a}$) of the equations for numbers at age in the standard WHAM model [@stockmiller21]. For ages $a = 2,\ldots, A-1$, where $A$ is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first $n_R$ elements of the vector
$$\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.$$
The numbers alive in each region can also be modeled more simply using the sub-matrix $\mathbf{O}_{y,a}$. The general model for the transitions in abundance at age is 
\begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*}
for ages $a = 2,\ldots, A-1$, and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}.
\end{equation*}
When the transitions in abundance at age are treated deterministically, $\boldsymbol{\varepsilon}_{y,a} = 0$. The stock- and region-specific errors $\boldsymbol{\varepsilon}_{y,a}$ are independent, but the same correlation structures as previous versions are possible across ages and years for a given stock and region. When there is autocorrelation with age, WHAM now assumes this applies only to ages $a>1$ by default so that recruitment random effects are independent of those for for the annual transitions of older age classes. So the general covariance structure for a given stock at ages $a>1$ in region $r$ is
\begin{equation*}
  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma^2_{N,r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}.
\end{equation*}

### Initial numbers at age {-}

Initial numbers at age for each stock and region can be treated as age-specific fixed effects or with an equilibrium assumption as in previous versions of WHAM. For the equilibrium option there are two parameters for each stock: the stock-specific fully-selected fishing mortality rate $\log \widetilde{F}$ and the recruitment in year 1 $\log N_{1,1,r_s}$. A stock-specific equilibrium fishing mortality at age by fleet ${\tilde F}_{a,f}$ is the product of $\widetilde{F}$ and the selectivity across fleets in the first year
\begin{equation*}
  \text{sel}_{1,a,f} = \frac{F_{1,a,f}}{\max_a \sum_{f=1}^{n_F} F_{1,a,f}}.
\end{equation*}
We use $\widetilde{F}_{a,f}$ to define an equilibrium abundance per recruit by region at age $a$ conditional on recruiting to each region
\begin{equation}\label{eq:npraa}
 \widetilde{\mathbf{O}}_{a} = \left\{
 \begin{array}{ll}
\prod^{a-1}_{j=0}\mathbf{O}_{j}  & 1\leq a<A\\
\left[\prod^{a-1}_{j=0}\mathbf{O}_{j}\right] \left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1} & a = A
 \end{array}
\right.
\end{equation}
where $\mathbf{O}_{j}$ is the equilibrium probability of surviving age $a$ and occurring in each region and $\mathbf{O}_{0} = \mathbf{I}$. Natural mortality and movement rates in the first year of the model are also used in Eq. \ref{eq:npraa}. For the plus group $a=A$, $\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}$ is a ``fundamental matrix'' derived using the matrix version of the  geometric series [@kemenysnell60]. Recall that recruitment for stock $s$ only occur in region $r_s$ so, the equilibrium initial numbers at age $a$ by region are 
$$\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}' \mathbf{N}_{O,1,1}.$$

The initial abundances at age can also be treated as independent or autocorrelated (with age) random effects. Defining the vector of initial abundance at age in region $r$ as $\mathbf{N}_{O,1,r}$, the general model is 
$$\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}$$
where 
$$  Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.$$

### Parametizing movement {-}

For each season, there are at most $n_R-1$ parameters determining movement among regions given starting an the season in region $r$ in either the sequential or simultaneous configurations. Movement parameters are estimated on a transformed scale via a link function $g(\cdot)$. If survival and movement are occur simultaneously, the parameters are estimated a log link function is used and if they are separable, an additive logit link function (like a multinomial regression) is used. On the transformed scale, the general model for the movement parameter from region $r$ to $r'$ in season $i$ and year $y$ for individuals of age $a$ is a linear function of both random and environmental effects:
\begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
\end{equation*}
The random effects $\epsilon_{r\rightarrow r',y,a,i}$ are season-, and(or) region-to-region-specific and modeled most generally as a two-dimensional first-order autoregressive random effects with age and(or) year where the covariance is
\begin{equation*}
  Cov\left(\epsilon_{r\rightarrow r',y,a,i},\epsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
\end{equation*}
similar to how WHAM models variation in survival, natural mortality, and selectivity. Effects of covariate $E_k$ can be age-, season-, and(or) region-to-region-specific $\beta_{r\rightarrow r',a,i,k}$ and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.

There is currently no likelihood component for tagging data. Therefore, movement parameters would generally either need to be fixed or assumed to have some prior distribution, possibly based on external parameter estimates. We include prior distributions for the season and region-to-region specific (mean) movement parameters which are treated as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
  \begin{equation*}
  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
  \end{equation*}
When priors are used, the movement is defined instead as
  \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
  \end{equation*}

### Natural mortality {-}

Natural mortality options have been expanded in WHAM. When not estimated, (mean) mortality rates may be stock-, region-, and age-specific. When random effects are used, the same autocorrelation structures with age and year as described by @stockmiller21 can be configured for a given stock and region. Any environmental covariate effects can be stock-, region-, and age-specific. So the general model for natural mortality is
\begin{equation*}
  \log M_{y,a,r} = \theta_{M,r} + \epsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
\end{equation*}
The general covariance structure for random effects are modeled most generally as a two-dimensional first-order autoregressive random effects with age and(or) year where the covariance is
\begin{equation*}
  Cov\left(\epsilon_{M,y,a,r},\epsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
\end{equation*}

### Catch observations {-}

The log-normal distributional assumption for aggregate catch observations is the same as @stockmiller21, but the predicted catch is now a function of catch from each stock starting the year in each region. For a given stock and age, the numbers captured in each fleet over the year are
$$\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}' \mathbf{N}_{O,s,y,a}$$
The predicted numbers caught be each fleet across stocks is 
$$\widehat{\mathbf{N}}_{H,y,a} = \sum^{n_S}_{s=1} \widehat{\mathbf{N}}_{H,s,y,a}$$
and the predicted aggregate catch at age $a$ is
$$\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \widehat{\mathbf{N}}_{H,y,a}$$
where $\mathbf{c}_{y,a}$ is the vector of mean individual weight at age $a$ for each fleet and the aggregate catch by fleet is 
$$\widehat{\mathbf{C}}_y = \sum^{A}_{a=1} \widehat{\mathbf{C}}_{y,a}$$
The log-aggregate catch observations for fleet $f$ are normally distributed
$$ \log C_{y,f} \sim \text{N}\left(\log \widehat {C}_{y,f}, \sigma^2_{y,f}\right).$$

The predicted numbers caught for each fleet $f$ (row $f$ of $\widehat{\mathbf{N}}_{H,y,a}$) are used to make predicted age composition observations as described by @stockmiller21. Since then, three additional likelihood options for age composition observations have been added: a logistic-normal with AR(1) correlation structure [@francis14], the alternative Dirichlet-multinomial parameterization described by @thorsonetal17, and the multivariate Tweedie [@thorsonetal23].

### Index observations {-}

For index $m$ occurring in region $r_m$ at fraction of the year $t_m$, the predicted abundance at $t_m$ in region $r_m$ is 
$$\widehat{N}_{s,y,a,m} = \mathbf{O}_{s,y,a,r_m}(t_m)' \mathbf{N}_{O,s,y,a}$$
where $\mathbf{O}_{s,y,a,r_m}(t_m)$, the $r_m$ column of the upper-left submatrix of Eq. \ref{eq:ptm_index}, are the probabilities of surviving and occurring in region $r_m$ at time $t_m$ given being alive in each region at the start of the year. The predicted index at age is
$$\widehat{I}_{m,y,a} = q_{m,y} \text{sel}_{m,y,a}w_{m,y,a}\sum^{n_S}_{s = 1}\widehat{N}_{s,y,a,m}$$
where $q_{m,y}$ is the catchability of the index in year $y$, $\text{sel}_{m,y,a}$ is the selectivity and $w_{m,y,a}$ is the average weight of individuals at age $a$ if the index is quantified in biomass and $w_{m,y,a} = 1$ if the index is quantified in numbers. Predicted age composition observations are functions of $\widehat{I}_{m,y,a}$ as described by @stockmiller21 and the likelihood options are the same as those for catch explained above.

Catchability of the index can also be treated as functions of normal random effects and(or) environmental covariate effects 
$$log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}$$ 
where $u_{m}$ and $l_{m}$ are the upper and lower bounds of catchability for index $m$ (defaults are 0 and 1000) and the general covariance structure for the annual random effects is first-order autoregressive
$$Cov\left(\epsilon_{q,m,y},\epsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.$$

### Weight and Maturity at age {-}

Weight and maturity at age are treated similarly to @stockmiller21. Annual weight at age matrices for each fleet are used to calculate total catch and the weight at age is applied to catch numbers at age for any stocks caught by the fleet. Similarly, when indices and(or) associated age composition observations are measured in biomass, the weight at age matrices for the index are applied to predicted numbers at age of all stocks observed by the survey. Unique weight at age and maturity at age matrices are allowed for each stock to calculate spawning stock biomass.

### Reference points {-}

Currently a single F reference point $\widetilde F$ is estimated across stocks and regions and F by fleet and age is $\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}$. Selectivity is determined as before when there are multiple fleets where $\text{sel}_{f,a}$ is determined by averaging F at age over a user-defined set of years
\begin{equation*}
  \text{sel}_{a,f} = \frac{\overline F_{a,f}}{\max_a \sum^{n_F}_{f=1}{\overline F}_{a,f}}
\end{equation*}

The equilibrium spawning stock biomass per recruit for stock $s$ in region $r_s$ is defined as
\begin{equation}\label{eq:ssbpr}
 \upphi_s(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{s,a,r_s,\cdot} \mathbf{O}_{s,a,\cdot,r_s}(t_s) w_{s,a} m_{s,a}
\end{equation}
where $w_{s,a}$ and $m_{s,a}$ are the mean individual weight and probability of maturity at age $a$, $\widetilde{\mathbf{O}}_{s,a}$ are as described in Eq. \ref{eq:npraa}, and $\mathbf{O}_{s,a}(t_s)$ is the $n_R \times n_R$ upper-left sub-matrix of eq. \ref{eq:ptm_spawn} with the probabilities of surviving and occurring in each region $r'$ at age $a+t_s$ given starting in region $r$ at age $a$. The further subscripts $r_s,\cdot$ and $\cdot,r_s$ indictate row or column $r_s$, respectively. Using these rows and columns is required because of the assumption that spawning and recruitment only occur in region $r_s$.

The equilibrium spawning biomass per recruit (eq. \ref{eq:ssbpr}) is conditional on the region of recruitment $r_s$. The equilibrium recruitment in each region $\widetilde {\mathbf{N}}_{s,1}$, depends on the stock dynamics. This version of WHAM currently only allows complete spawning region fidelity so that a stock only spawns and recruits in a single region. In this case, $\widetilde {\mathbf{N}}_{s,1}$ will be positive in the spawning region ($r_s$) and zero elsewhere. Similarly, the row $r_s$ of $\mathbf{O}_s$ will be zero off of the diagonal. The matrices of probabilities of surviving and occurring in each region, $\widetilde{\mathbf{O}}_{s,a}$ and $\mathbf{O}_{s,a}(\delta_s)$, are functions of the fishing mortality rates for fleets in each region $\widetilde F_{f,a}$.

The matrix equilibrium yield per recruit as a function of $\widetilde F$ is calculated as
\begin{equation}\label{eq:ypr}
 \widetilde{Y}_s({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{s,a,r_s,\cdot} \mathbf{H}_{s,a} \mathbf{c}_{s,a}
\end{equation}
where $\mathbf{c}_a$ is the vector of mean individual weight at age for each fleet, and $\mathbf{H}_{s,a}$ is the submatrix of the probabilities of being captured in each fleet over the interval from $a$ to $a+1$, defined in eq. \ref{eq:ptm}.

As in previous versions of WHAM package, ``static'' reference points, typically meant to be defined for prevailing conditions, average all of the inputs to the spawning biomass and yield per recruit calculations over the user-specified years (e.g., last 5 years of the model). This same averaging is also applied to possibly time-varying movement parameters.

For $X\%$ SPR-based reference points, we use a Newton method and iterate
\begin{equation}\label{eq:newton}
  \log\widetilde{F}^{(i)} = \log\widetilde{F}^{(i-1)} - \frac{g\left(\log\widetilde{F}^{(i-1)}\right)}{g'\left(\log\widetilde{F}^{(i-1)}\right)}
\end{equation}
where $g(\log F)$ is the difference between the weighted sums of spawning biomass per recruit at $F$ and $X$\% of unfished spawning biomass per recruit across stocks:
\begin{equation*}
  g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
\end{equation*}
where $\upphi_s\left(F=0\right)$ is the equilibrium unfished spawning biomass per recruit. $g'(\log F)$ is the derivative of $g$ with respect to $\log F$, and the weights to use for each stock $\lambda_s$ can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate ``static'' equilibrium spawning biomass and yield.

When a Beverton-Holt or Ricker stock recruit relationship is assumed, an analogous Newton method is used to find $\log F$ that maximizes yield for MSY-based reference points. which are also a functions of the equilibrium yield per recruit (\ref{eq:ypr}) and equilibrium recruitment. The function $g(\log F)$ in Eq. \ref{eq:newton} is the first derivative of the yield curve with respect to $\log F$.

### Projections {-}

The projection options are generally the same as those for previous versions of WHAM. When there is movement of any stocks, the user has the option to project and use any random effects for time-varying movement or use the average over user specified years, analogous to how natural mortality can be treated in the projection period. The projections of any environmental covariates has been revised to better include error in the estimated latent covariate in any effects on the population in projection years.

<!--
## Model configurations  {-}

We compared the base models for each stock in the single-stock and multi-stock versions of wham.

```{r, echo = FALSE, out.width="75%", fig.align="center"}
#knitr::include_graphics(here::here("docs","plots","compare_one_stock_models.png"))
```
-->

# Application to black sea bass {-}

Prior to the most recent peer-reviewed assessment of black sea bass in the NEUS, the stock was assessed using the Age-Structured Assessment Program (ASAP) model [@legaultrestrepo99], a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects. Northern and southern components of the NEUS black sea bass stock divided by the Hudson Canyon were separately modeled in ASAP (reference map figure). Results from the separate ASAP models have been aggregated for a unit-stock assessment. Furthermore, the ASAP-based assessments have exhibited strong retrospective patterns [@mohn99], and exploring alternative modeling approaches for the northern and southern stock components has been a high priority for management.

In 2023, a working group composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above [@nefsc23]. This assessment included the spatial features and hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.

Below we describe the assumptions and configuration of the assessment model as determined by the working group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of environmental effects.

## Basic structure {-}

The first year being modeled for the population is 1989 and the fishery and index data used in the model span from 1989 to 2021. There are two stock components (north and south) modeled as separate populations that spawn and recruit in respective regions separated by the Hudson Canyon. We have observations for each of four total fishing fleets, where two fishing fleets (Recreational and Commercial) operate in each region.

There are 11 seasonal intervals within each calendar year: five monthly time intervals from Jan 1 to May 31, a spawning season from June 1 to July 31, and five monthly intervals from August 1 to December 31. The southern stock component is assumed to never move to the northern region (Figure \ref{fig:migration-diagram}). For the northern component, a proportion $\mu_1$ can move to the southern region each month during the last five months of the year, but no movement is allows from the south to the north during this period. During the first four intervals of the year a proportion $\mu_2$ the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period. In the fifth interval (May), all northern component individuals remaining in the south are assumed to move back to the north for the subsequent spawning period. Survival and movement occur sequentially in each interval and each of the two movement proportions are assumed constant across intervals, ages, and years.

The monthly movement matrices are
\begin{equation*}
\boldsymbol{\mu}_{1} = 
  \begin{bmatrix}
     1-\mu_{1\rightarrow 2} & \mu_{1\rightarrow 2} \\
     0 & 1 \\
  \end{bmatrix}
\end{equation*}
for the portion of the year after spawning and
\begin{equation*}
\boldsymbol{\mu}_{2} = 
  \begin{bmatrix}
     1 &  0 \\
     \mu_{2\rightarrow 1} & 1-\mu_{2\rightarrow 1} \\
  \end{bmatrix}
\end{equation*}
for the portion of the year before spawning.

The constant or mean log-natural mortality rate is assumed to be $\log(0.4)$

As noted in the description of the general WHAM model, tagging data are not yet allowed. However, the working group also fit a Stock Synthesis model [@methotwetzel13] which provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).

## Initial abundance at age {-}

With the movement configuration, the northern origin fish (ages 2+) can occur in the southern region on January 1. Estimating initial numbers at age as separate parameters can be challenging even in single-stock models. To avoid challenges with estimating initial numbers at age in each region for the northern component in the two stock model, we used the equilibrium assumption described above. Using this assumption two parameters are estimated for each stock component: an initial recruitment and an equilibrium fully-selected $F$.

## Recruitment and Survival {-}

For the northern population abundance at age 1 on January 1 (recruitment) is only allowed in the northern region, but given the monthly movement described above, older individuals that previously recruited in the northern region may occur in the southern region on January 1. Therefore, a model with survival random effects will model the transitions (survival/movement) of abundances at age of northern origin fish in each region. All of the initial runs assumed variance parameters for these random effects to be the same for northern origin fish occurring in both regions on January 1. The base model assumes very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model. We also allow 2DAR1 correlation for recruitment and survival for both the northern and southern components. Unique variance and correlation parameters for the recruitment and survival random effects are estimated for the northern and southern components.

## Uncertainty recreational CPA index observations {-}

The CVs provided by the analyses for the recreational catch per angler (RecCPA) ranged between 0.02 and 0.06 which the working group felt did not capture the true uncertainty in the index with regard to its relationship to stock abundance. In many of the initial runs as well as the base model we allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions. Models that successfully estimated these scalars indicated standard deviations for these surveys to be  approximately 5 times the input value and this value was fixed in many preliminary runs to avoid dealing with convergence problems. However, the base model successfully estimated these scalars. The model estimates are negligibly affected by estimating these scalars, but we felt estimating these parameters allowed uncertainty in model output to be more properly conveyed.

## Index and Catch age composition observations {-}

Table \ref{tab:age-comp-sel-table}

## Bottom Temperature effects {-}

All model fits also include bottom temperature observations for the northern and southern regions from 1963 to 2021 and estimated standard errors ranging between 0.03 and 0.09 degrees Celsius [@nefsc23]. We retained the assumption from the peer-reviewed assessment that treated the latent bottom temperature covariates in each region as first-order autoregressive processes.

We fit 14 models with alternative assumptions about the effects of bottom temperature covariates, ranging from no effects to effects on both regions for either recruitment or natural mortality at age 1 (Table \ref{tab:model-desc-table}). These analyses derive from the hypothesis that bottom temperature affects overwinter survival of fish where the fish turn from age 0 to age 1 on January 1 [@milleretal16_yoy_survival]. This temperature may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of of the year after becoming age 1. Furthermore, we have no direct observations of age 1 individuals from surveys until the spring season each year. Therefore, we fit models with effects of temperature on recruitment or natural mortality at age 1.

It is standard practice to treat annual recruitment as time-varying deviations from a mean model and all models here treat recruitment deviations as first-order autoregressive random effects. However, @milleretal18 showed how inferences of temperature effects on growth or maturity parameters can be very different whether the null model without the effect and models with effects also include random effects representing residual temporal variation in parameters. Therefore, we also explored whether including temporal random effects on age 1 natural mortality M affected inferences on corresponding temperature effects. Initially, we included random effects on age 1 natural mortality for both the northern and southern stock components, but estimates of these random effects and corresponding variance for the southern component converged to 0 so these were not included in models presented here.

We assume the covariate in year $y$ affects recruitment in the same years because the covariate observations are from months January to March. The fish are technically already 1 year old, but there are no observations of these individuals until later in the year except possibly in fishery catches which are accumulated over the whole year. Expected log-recruitment is a linear function of log-recruitment at the previous time step and bottom temperature

\begin{equation}\label{eq:expected-recruitment}
\log N_{s,r_s,y,1} = \mu_{R,s} + \beta_{R,s} x_y + \epsilon_{N,s,y,1}
\end{equation}

\begin{equation}\label{eq:expected-M1}
\log M_{s,y,1} = \mu_{M,s,1} + \beta_{M,s} x_y + \epsilon_{M,s,y}
\end{equation}

## Model fitting, diagnostics, and comparisons {-}

As in @milleretal16, we also assessed the stability of the AIC-based model selection over retrospective peels to guard against previously noted changes in perception of covariate effects on recruitment with increased length of the time series of observations [@myers98]. This retrospective examination was also recommended by @brooks24.

We examined retrospective patterns by fitting models where the terminal year is reduced sequentially by one year (peel) for seven years. Therefore, there are 7 fits of the proposed base model with the time series reduced by one to seven years. We  calculated Mohn's $\rho$ for recruitment, SSB, and fully-selected $F$. Absolute values of Mohn's $\rho$ near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended.

Specifically, we use version XX and commit XXX of the wham package (https://github.com/timjmiller/wham/tree/lab) for all results.

We performed a simulation self-test where new observations were simulated conditional on all random effects estimated in the proposed base model and the same model configuration was fit to each of the simulated data sets.

One-step-ahead residuals can now be calculated for all index, catch, environmental covariate observations using methods described by @thygesenetal17 and @trijouletetal23.

### One-step-ahead residuals {-}

<!--

We used one-step-ahead (OSA) residuals to evaluate mis-specification of the proposed base model. Pearson residuals do not have the appropriate properties (independent standard normal for a correctly specified model) for composition observations [@trijouletetal23] or for covariate and aggregate index and catch observations when temporal random effects are employed  because of the lack of independence of observations [@thygesenetal17].

There is no evidence of mis-specificaiton from the OSA residuals for aggregate catch of the northern commercial fleet (Figure \ref{fig:osa-North-comm-catch-summ}). There was some indication of trends in residuals of some of the age composition observations early in the time series and for the the first age class (Figures \ref{fig:osa-North-comm-paa-summ} and \ref{fig:osa-North-comm-paa}).

For the northern recreational fleet, the OSA residuals for the aggregate catch appeared satisfactory (Figure \ref{fig:osa-North-rec-catch-summ}). The OSA residuals for the age composition observations for the northern recreational fleet showed some tendency of underdispersion but were otherwise unnoteworthy (Figures \ref{fig:osa-North-rec-paa-summ} and \ref{fig:osa-North-rec-paa}).

Neither the OSA residuals for the aggregate catch or the age composition of the Northern recreational CPA index exhibited signs of mis-specification (Figures \ref{fig:osa-North-reccpa-catch-summ} to \ref{fig:osa-North-reccpa-paa}).

There was a little evidence of tendency toward negative OSA residuals for the Northern VAST index whereas there was some indication of positive residuals for the age composition observations, particularly at age 1 (Figures \ref{fig:osa-North-vast-catch-summ} to \ref{fig:osa-North-vast-paa})

THe OSA residuals for the aggregate catch of the southern commercial fleet tended to be negative, but the residuals for the age composition observations did not show any evidence of mis-specification (Figures \ref{fig:osa-South-comm-catch-summ} to \ref{fig:osa-South-comm-paa}).

The OSA residuals for the aggregate catch of the southern recreational fleet were under-dispersed and those for the age composition were somewhat underdispersed, but no trends in either types of observations were apparent (Figures \ref{fig:osa-South-rec-catch-summ} to \ref{fig:osa-South-rec-paa})

The OSA residuals for the aggregate catch of the southern recreational CPA index were also somewhat under-dispersed (Figures \ref{fig:osa-South-reccpa-catch-summ}). The residuals for the age composition observations exhibited a few large negative residuals at older ages, but otherwise exhibited no apparent trends (Figures \ref{fig:osa-South-reccpa-paa-summ} to \ref{fig:osa-South-reccpa-paa})

The OSA residuals for the aggregate southern VAST index were somewhat over-dispersed and the residuals for the age composition appeared to show some trend with observed proportion and age (Figures \ref{fig:osa-South-vast-catch-summ} to \ref{fig:osa-South-vast-paa}).
-->

### Retrospective patterns {-}

<!-- 
We also examined retrospective patterns by fitting models where the terminal year is reduced sequentially by one year (peel) for seven years. Therefore, there are 7 fits of the proposed base model with the time series reduced by one to seven years. Multi-WHAM performs these fits when specified and will calculate Mohn's $\rho$ for recruitment, SSB, and fully-selected $F$. Absolute values of Mohn's $\rho$ near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended. Although there was strong retrospective patterns in the most recent management track assessment for the northern component of the stock, there was no evidence of patterns for northern SSB (Mohn's $\rho$ = -0.058) and $F$ (Mohn's $\rho$ = 0.06) for the proposed base model (Figures \ref{fig:North-retro-ssb} and \ref{fig:North-retro-F}). Similarly, no patterns were exhibited for the southern component of SSB (Mohn's $\rho$ = -0.019) and regional $F$  (Mohn's $\rho$ = -0.05) (Figures \ref{fig:South-retro-ssb} and \ref{fig:South-retro-F}).
-->

# Results {-}

Although there was strong retrospective patterns in the most recent management track assessment for the northern component of the stock, there was no evidence of patterns for northern SSB (Mohn's $\rho$ = -0.058) and $F$ (Mohn's $\rho$ = 0.06) for the proposed base model (Figures \ref{fig:North-retro-ssb} and \ref{fig:North-retro-F}). Similarly, no patterns were exhibited for the southern component of SSB (Mohn's $\rho$ = -0.019) and regional $F$  (Mohn's $\rho$ = -0.05) (Figures \ref{fig:South-retro-ssb} and \ref{fig:South-retro-F}).

For 7 of the the simulated data sets the model failed to optimize. The maximum absolute gradient was $<10^{-6}$ for only 9 and $<10^{-4}$ for 52 of the 93 successfully fitted models. The poor convergence appeared to be attributable to the estimation of the scalar for the standard errors of the log-transformed northern Recreational CPA index for which estimates tended to 0 for nearly all of the fits ($<0.01$ for 83 fits). However, even across all fits including those with poor convergence, the SSB estimates appeared to be reliable (Figure \ref{fig:self-test-res-1}).

## Bottom temperature effect on recruitment {-}

We found the best model of bottom temperature covariate effects included the effect only on the recruitment of the northern stock, which is the configuration in the proposed base model although the difference in AIC for the model including effects on both regional components was small suggesting some evidence for this hypothesis as well (Table \ref{tab:compare-table}) . We also found AIC to strongly favor models that included at least an effect on the northern component across all retrospective peels (Table \ref{tab:peel-compare-table}). The posterior estimate of the bottom temperature covariate match the observations well because of the high precision of the observations (Figure \ref{fig:bottom-temp}). The residual variation in the standard deviation of recruitment random effects is reduced because the expected recruitment (Eq. \ref{eq:expected-recruitment}) is a function of the covariate  (Figure \ref{fig:relative-recruitment-bt}). This effect is included in the proposed base model (Figure \ref{fig:rec-bottom-temp}). Because the covariate is technically for temperature after the beginning of the calendar year when the fish are considered age 1 in the model, a comparison of the proposed base model to one with the bottom temperature effect on natural mortality at age 1 instead might be of interest in the future. An investigation of higher order orthogonal polynomial effects might also be of interest particularly for the southern component which might be experiencing more frequently higher temperatures less favorable to overwinter survival.

# Discussion {-}

The new features of Multi-WHAM include

* seasonal intervals within years
* variation in movement rates by stock, region-to-region, season, age, and year
* effects of environmental covariates on movement rates by stock, region-to-region, season, and age
* effects of environmental covariates on mortality rates by stock, region, and age
* effects of environmental covariates on recruitment by stock
* stock-specific stock-recruitment models
* priors for movement rates
* seasonal operation of fleets
* mortality and movement modeled sequentially or simultaneously
* options for weighting of stock-specific SSB/R for global SPR-based reference points

# References {-}

<div id="refs"></div>

\pagebreak

# Appendix A {-}

\setcounter{table}{0}
\renewcommand\thetable{A\arabic{table}}
\input{symboltable.tex}

\pagebreak

```{r migration-diagram, echo = FALSE, out.width="80%", fig.align="center", fig.cap = "Diagram of intervals within the year and configuation of the dynamics of each component of the BSB population."}
knitr::include_graphics(here::here("paper","bsb_movement_diagram.pdf"))
```

\setcounter{table}{0}
\renewcommand\thetable{\arabic{table}}

```{r, echo = F, eval = T}
age_comp_names <- c("North Commercial", "North Recreational", "South Commercial", "South Recreational",
  "North Recreational CPA", "North VAST", "South Recreational CPA", "South VAST")
age_comp_mod <- c("Dirichlet-Multinomial", "Logistic-normal (0s as missing)", "Logistic-normal (AR1, 0s as missing)","Logistic-normal (AR1, 0s as missing)", "Logistic-normal (0s as missing)","Dirichlet-Multinomial","Logistic-normal (AR1, 0s as missing)","Logistic-normal (AR1, 0s as missing)")
mean_sel_mod <- c("age-specific (flat-topped at ages > 3)",
  "age-specific (flat-topped at ages > 6)",
  "logistic", "logistic",
  "age-specific (flat-topped at ages > 1)",
  "age-specific (flat-topped at ages > 4)",
  "age-specific (flat-topped at ages > 2)",
  "age-specific (flat-topped at ages > 1)")
sel_re_mod <- c("2D-AR1 (age and year)", 
  "2D-AR1 (age and year)",
  "None",
  "None",
  "AR1 (year)",
  "2D-AR1 (age and year)",
  "None",
  "None")
out <- cbind.data.frame("Data component" = age_comp_names,  "Age Composition Likelihood" = age_comp_mod,  
  "Mean Selectivity model" = mean_sel_mod, "Random effects configuration" = sel_re_mod)

out %>% kable(format = type, booktabs = T, escape=F, row.names = F, label = "age-comp-sel-table",
    caption="Configuration of Age composition likelihood, mean selectivity model, and selectivity random effects for each age composition data component.", align = c(rep("l",4))) %>% landscape()
#knitr::knit_print(kableExtra::landscape(out))

```

```{r, echo = F, eval = T}
mod_names <- paste0("$M_{", 0:13,"}$")
col_names <- c("Model", "North", "South", "$M$ at age 1 random effects")

mod_table <- matrix(nrow = length(mod_names), ncol = length(col_names), dimnames = list(mod_names,col_names))

mod_table[,1] <- mod_names
mod_table[,2:4] <- "--"
mod_table[,4] <- rep(c("none", "time-varying"), each = 7)
mod_table[c(2,4),2] <- "Recruitment"
mod_table[c(3,4),3] <- "Recruitment"
mod_table[c(5,7),2] <- "$M$ at age 1"
mod_table[c(6,7),3] <- "$M$ at age 1"
mod_table[8:14,2:3] <- mod_table[1:7,2:3]

mod_table %>% kable(format = type, booktabs = T, escape=F, row.names = F, label = "model-desc-table",
    caption="Assumptions for temperature effects and random effects for age 1 natural mortality for each model.", align = c(rep("l",4))) %>%
  add_header_above(c(" " = 1, "Temperature Effect" = 2, " " = 1 ))
```

```{r, echo = F, eval = T}
diff_aic <- readRDS(here::here("results","diff_aic.RDS"))
diff_aic <- data.frame(round(diff_aic, 2))
diff_aic <- cbind(Model = row.names(diff_aic), diff_aic)
row.names(diff_aic) <- NULL
colnames(diff_aic) <- c("Model", as.character(0:7))
diff_aic %>%
  kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "diff-aic-table",
  caption = "Difference between AIC and the lowest AIC for each model by retrospective peel.") %>%
  add_header_above(c(" " = 1, "Peel" = 8))
```

```{r, echo = F, eval = T}
aic_wts <- readRDS("../results/aic_wts.RDS")
aic_wts <- data.frame(round(aic_wts, 2))
aic_wts <- cbind(Model = row.names(aic_wts), aic_wts)
row.names(aic_wts) <- NULL
colnames(aic_wts) <- c("Model", as.character(0:7))
aic_wts %>%
  kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "aic-wts-table",
  caption = "Model AIC weights for each retrospective peel.") %>%
  add_header_above(c(" " = 1, "Peel" = 8))
```

```{r, echo = F, eval = T}
rho_table <- readRDS("../results/rho_table.RDS")
row.names(rho_table) <- NULL
colnames(rho_table) <- c("Model", rep(c("North","South"), 3))
rho_table %>%
  kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "rho-table",
  caption = "Mohn's $\\rho$ for SSB, R, average F at ages 6 and 7 in northern and southern regions.") %>%
  add_header_above(c(" " = 1, "SSB" = 2, "Average F" = 2, "Recruitment" = 2))
```
\clearpage

# Supplemental Materials {-}

## Deriving the prior distribution for movement parameters {-}

The working group fit a Stock Synthesis model [@methotwetzel13] that included tagging data with 2 seasons (6 months each) and 2 regions where a proportion $\mu^*_1$ of the northern component moves to the south in one season and some proportion  $\mu^*_{2\rightarrow 1}$ move back to the south in the second season [@nefsc23]. The seasonal movement matrices for each season are
\begin{equation*}
\boldsymbol{\mu}^*_{1} = 
  \begin{bmatrix}
     1-\mu^*_{1\rightarrow 2} & \mu^*_{1\rightarrow 2} \\
     0 & 1 \\
  \end{bmatrix}
\end{equation*}
and
\begin{equation*}
\boldsymbol{\mu}_{2} = 
  \begin{bmatrix}
     1 &  0 \\
     \mu^*_{2\rightarrow 1} & 1-\mu^*_{2\rightarrow 1} \\
  \end{bmatrix}.
\end{equation*}

To obtain estimates of movement proportions for the monthly intervals in the WHAM model, the half-year movement matrices were converted to monthly movement matrices by taking the root $z_k$ of $\boldsymbol{\mu}^*_{k}$ which are defined by the number of months of movement for each season (5 and 4, respectively). The roots of the matrices are calculated using an eigen decomposition of the matrices
$$ \boldsymbol{\mu}_k =  \left(\boldsymbol{\mu}_k^*\right)^{z_k} = \mathbf{V}_k \mathbf{D}_k^{z_k} \mathbf{V}_k^{-1}$$
where $z_1 = 1/5$ for and $z_2 = 1/4$, and $\mathbf{V}_{k}$ and $\mathbf{D}_{k}$ are the matrix of eigenvectors (columnwise) and the diagonal matrix of corresponding eigenvalues of $\boldsymbol{\mu}^*_k$. The working group used a parametric bootstrap approach to determine an appropriate standard deviation for the prior distribution for the movement parameters. Stock Synthesis also estimates parameters on a transformed scale, but different from WHAM: 
$$\mu^*_{r\rightarrow r'} = \frac{1}{1 + 2e^{-x_{r\rightarrow r'}}}$$
The estimated parameters and standard errors from the Stock Synthesis model were $x_{1\rightarrow 2}=-1.44$ and $x_{2\rightarrow 1}=1.94$ and $SE(x_{1\rightarrow 2})) = 0.21$ and $SE(x_{2\rightarrow 1})) = 0.37$. The resulting in the estimated proportions were $\mu^*_{1\rightarrow 2}=0.11$ and $\mu^*_{2\rightarrow 1}=0.78$.

In WHAM, an additive logit transformation is used which is simply a logit transformation when there are only two regions: 
$$
\mu_{r\rightarrow r'} = \frac{1}{1+e^{-y_{r\rightarrow r'}}}.
$$
We simulated 1000 values from a normal distribution with mean and standard deviation defined by the parameter estimate and standard error $\tilde x_{{r\rightarrow r'},b} \sim N(x_{r\rightarrow r'}, SE(x_{r\rightarrow r'}))$ from the Stock Synthesis model. For each simulated value we constructed $\tilde {\boldsymbol{\mu}}^*_{{r\rightarrow r'},b}$, took the appropriate root and calculated inverse logit for $\tilde y_{{r\rightarrow r'},b}$. We calculated the mean and standard deviation of the values $y_{i,b}$. The mean values did not differ meaningfully from the transformation of the original estimates ($y_{1\rightarrow 2} = -3.79$ and $y_{2\rightarrow 1} = -0.79$) and the standard deviation was approximately 0.2 for both parameters.

### 2024-09-20 paper/paper.tex
start repo

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={A multi-stock, multi-region state-space age-structured assessment model with an application to black sea bass},
  pdfauthor={Timothy J. Miller1; Kierstin Curti; Alex Hansell},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{url}
\usepackage{setspace}
%\singlespacing
%\onehalfspacing
\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[belowskip=0pt,aboveskip=0pt]{caption}
\usepackage{relsize}
\usepackage{float}
% \usepackage[section]{placeins}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{amsmath,rotating}
\usepackage[scanall]{psfrag}
\usepackage{bm}
\usepackage{caption,graphics}
\usepackage{graphicx}
%\usepackage{natbib}
%\usepackage[nottoc]{tocbibind}
%\usepackage{indentfirst}
\usepackage{sectsty}
\usepackage{color}
\usepackage{fancyhdr}
\usepackage{xspace}
\usepackage{textcomp}
\usepackage{upgreek}

\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{A multi-stock, multi-region state-space age-structured assessment
model with an application to black sea bass}
\author{Timothy J. Miller\textsuperscript{1} \and Kierstin
Curti \and Alex Hansell}
\date{16 September, 2024}

\begin{document}
\maketitle

\(^1\)\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}},
Northeast Fisheries Science Center, National Marine Fisheries Service,
166 Water Street, Woods Hole, MA 02543, USA\\

\pagebreak

\hypertarget{main-message}{%
\subsection*{Main Message}\label{main-message}}
\addcontentsline{toc}{subsection}{Main Message}

A multi-region, multi-stock generalization of WHAM with an application
evaluating evidence of alternative hypotheses about temperature effects
on black sea bass

\hypertarget{abstract}{%
\subsection*{Abstract}\label{abstract}}
\addcontentsline{toc}{subsection}{Abstract}

The Woods Hole Assessment Model (WHAM) is a state-space age-structured
assessment model that is used to assess and manage many stocks in the
Northeast US. We first describe a multi-stock, multi-region extension of
WHAM that treats the population and fleet dynamics seasonally and allows
movement by season and region to be functions of time- and age-varying
autocorrelated random effects and environmental covariates. We then
illustrate the model with northern and southern components of the NEUS
black sea bass stock and evaluate alternative hypotheses of bottom
temperature and random effects on recruitment and natural mortality. We
show strong evidence for temperature effects on recruitment and no
random effects or temperature effects on natural mortality, primarily
for the northern stock component.

\pagebreak

\hypertarget{introduction}{%
\section*{Introduction}\label{introduction}}
\addcontentsline{toc}{section}{Introduction}

Why is this model needed. Models ignoring movement can provide biased
catch advice. Using random effects for time-varying processes is
considered best practice.

The Woods Hole Assessment Model (WHAM) is an R package developed at the
Northeast Fisheries Science Center
(\url{https://timjmiller.github.io/wham}, Miller and Stock 2020; Stock
and Miller 2021). Primary inputs and observation types are similar to
ASAP and there is functionality built into WHAM to move information from
ASAP input files to R. WHAM can be configured to fit SCAA models often
identically to ASAP, so that existing assessments in the U.S. Northeast
can be replicated and tested against state-space models with process
errors and environmental effects in a single framework. WHAM models are
built using the Template Model Builder package (TMB, Kristensen et al.
2016) which provides efficient fitting of models that include random
effects similar to the State-space Assessment Model (SAM, Nielsen and
Berg 2014), which is used to manage many stocks within ICES. WHAM is
currently used to manage 9 stocks in the NEUS.

The Woods Hole Assessment Model (WHAM) software package is maintained
and developed at the Northeast Fisheries Science Center to enable
state-space stock assessments, i.e.~where processes such as the annual
transitions in numbers-at-age (survival), natural mortality,
selectivity, and catchability are treated as time- and(or) age-varying
random effects. WHAM can also be configured without random effects as a
traditional statistical catch-at-age model in order to bridge from
current assessments which use Age-Structured Assessment Program (ASAP).
Here, we present an extension of WHAM (Multi-WHAM) to model multiple
stocks and(or) regions and allow movement of stocks among regions.

It has been reviewed in the literature and simulation tested (Stock et
al. 2021; Stock and Miller 2021). It has also been used as the operating
model in the index-based assessment methods research track assessment
(Legault et al. 2023) and is also the main tool for simulation studies
in the research track on state-space assessment models.

A review of the essential features for next-generation stock assessments
concluded that only WHAM and a similar State-Space Assessment Method
(Nielsen and Berg 2014) model random effects correctly (Punt et
al.~2020). WHAM is now being used as the management model for 9 stocks
in the northeast U.S., but the standard version of WHAM can only be
applied to a single stock and area (Miller and Stock 2020).

Here, we describe a multi-stock and multi-region extension of the WHAM
package (version 2.0 \url{https://github.com/timjmiller/wham/tree/lab}).
Numerous new configuration options are also included in this extension
that can be useful even for single-stock or single region models. Any
number of stocks or regions can be configured in this version of WHAM.
The multi-stock extension was developed to accommodate movement of
stocks or stock components in anticipation of black sea bass and
Atlantic cod stocks where new modeling frameworks where recently
examined through Northeast research track assessment process (cite RT
docs). We demonstrate capabilities of the new version of WHAM with black
sea bass and evaluate evidence for alternative hypotheses of temporal
variation in natural mortality and effects of bottom temperature on
recruitment and natural mortality of age 1 individuals.

\hypertarget{methods}{%
\section*{Methods}\label{methods}}
\addcontentsline{toc}{section}{Methods}

\hypertarget{model-description}{%
\subsection*{Model description}\label{model-description}}
\addcontentsline{toc}{subsection}{Model description}

Many of the options and equations of version 2.0 are are the same as
those described by Stock and Miller (2021), so we will only describe
extensions and differences that have occurred since the first
description of WHAM. The new version of WHAM can model multiple stocks,
but survival, movement, harvest and natural mortality is are tracked for
each stock. Therefore, much of the description below assumes a specific
stock \(s\), but this subscript is only explicitly used when necessary.

\hypertarget{the-probability-transition-matrix}{%
\subsubsection*{The probability transition
matrix}\label{the-probability-transition-matrix}}
\addcontentsline{toc}{subsubsection}{The probability transition matrix}

Because individuals may be alive in one of several regions or harvested
in one of several fleets, it is helpful to consider these as distinct
categories or states and treat the number of individuals occurring in
each category over time as a multi-state model (reference). Multi-state
models use a probability transition matrix (PTM) that describes the
probability of individuals living and moving among regions or dying due
to fishing or natural mortality over a time interval \(i\) with duration
\(\delta_i\) in year \(y\) for individuals at age \(a\) on January 1.
Each row and column of the PTM correspond to one the states: alive in
region \(r\), dead in fleet \(f\), or dead from natural causes. The
probabilities in each row sum to unity and assume an individual is in
the corresponding state at the beginning of the interval. Given \(n_R\)
regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and
columns) as a function of sub-matrices is \begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix}
\end{equation} where \begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*} is the \(n_R \times n_R\) matrix defining survival and
occurring in each region at the end of the interval, \begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*} is the \(n_R \times n_F\) matrix defining probabilities
of being captured in each fleet during the interval, and
\(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of
dying due to natural mortality during the interval. We have the identity
matrix \(\mathbf{I}_{H}\) for the states for capture by each fleet and a
1 for the state for natural mortality because the probabilities of being
in one of the mortality states given starting the interval in that state
is unity.

WHAM uses these PTMs to model abundance proportions in each state rather
than true probabilities where numbers in each state would be multinomial
distributed. The PTMs determine the expected numbers 1) in each state on
January 1 of year \(t+1\) at age \(a+1\) given the abundances at age
\(a\) on January 1 of year \(t\), 2) captured over the year in each
fleet, 3) available to each index, and 4) alive at the time and in the
region where spawning occurs.

\hypertarget{single-region-ptms}{%
\subsubsection*{Single region PTMs}\label{single-region-ptms}}
\addcontentsline{toc}{subsubsection}{Single region PTMs}

When there is only one region, \begin{equation}\label{eq:ptm_1_region}
\mathbf{P}_{y,a,i} = 
  \begin{bmatrix}
     S_{y,a,i} & \mathbf{H}_{y,a,i} & D_{y,a,i} \\
     0 & \mathbf{I}_{H} & 0\\
     0 & 0 & 1
  \end{bmatrix}
\end{equation} where \(S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}\),
\(\mathbf{H}_{y,a,i}\) is a 1 x \(n_F\) matrix with elements for each
fleet \(f\):
\(\frac{F_{y,a,i,f}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\),
\(D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\),
and \(Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{y,a,i,f}\) is the total
mortality rate.

\hypertarget{multi-region-ptms}{%
\subsubsection*{Multi-region PTMs}\label{multi-region-ptms}}
\addcontentsline{toc}{subsubsection}{Multi-region PTMs}

When there is more than 1 region, WHAM can model survival and movement
as processes occurring sequentially or simultaneously. The sequential
assumption is used widely in spatially explicit model (e.g., SS3). Under
the sequential assumption, survival and death occur over the interval
and movement among regions occurs instantly at either the beginning or
the end of the interval. WHAM is configured to have movement occur after
survival and mortality: \begin{equation*}
  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i}
\end{equation*} where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\)
diagonal matrix of proportions surviving in each region (given they
start in that region) \begin{equation*}
\mathbf{S}_{y,a,i} = 
  \begin{bmatrix}
    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
  \end{bmatrix}
\end{equation*} and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\)
matrix of probabilities of moving from one region to another or staying
in the region they occurred at the beginning of the interval:
\begin{equation*}
\boldsymbol{\mu}_{y,a,i} = 
  \begin{bmatrix}
    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
    \vdots & \vdots & \ddots & \vdots \\
    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
  \end{bmatrix}
\end{equation*}

WHAM assumes each fleet \(f\) can harvest in only 1 region (\(r_f\))
during specified seasons. So, for each fleet \(f\), row \(r_f\) and
column \(f\) of \(\mathbf{H}_{y,a,i}\) will be
\(F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\)
when fleet \(f\) is harvesting during the interval \(\delta_i\) and all
other elements will be zero. Row \(r\) of the single-column matrix
\(\mathbf{D}_{y,a,i}\) is
\(M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\)

When survival and movement are assumed to occur simultaneously, all
movement and mortality parameters are instantaneous rates. We obtain the
probability transition matrix over an interval \(\delta_i\) by
exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
\begin{equation*}
\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
\end{equation*} The instantaneous rate matrix takes rates of movement
between regions and the mortality rates for each fleet and region. Along
the diagonal is the negative of the sum of the other rates (the hazard)
so each row sums to zero. For two regions and one fleet operating in
each region: \begin{equation*}
 \mathbf{A}_{y,a,i} = \begin{bmatrix}
 a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
 \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
 0 & 0 & 0 & 0 & 0 \\
 0 & 0 & 0 & 0 & 0 \\
 0 & 0 & 0 & 0 & 0
 \end{bmatrix}.
\end{equation*} where
\(a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})\).
When there is one region, \(n_f\) fleets, and \(\delta_i = 1\),
exponentiating the instantaneous rate matrix results in the PTM defined
in Eq. \ref{eq:ptm_1_region}.

\hypertarget{seasonality}{%
\subsubsection*{Seasonality}\label{seasonality}}
\addcontentsline{toc}{subsubsection}{Seasonality}

Seasonality can be configured to accommodate characteristics of
spawning, movement, and fleet-specific behavior. The annual time step
can be divided into \(K\) (any number) seasons and the interval size
\(\delta_i\) for each season \(i\) need not be equal to any other
seasonal interval. Under the Markov assumption, the PTM of surviving and
moving and dying over \(K\) intervals \(\delta_1,\ldots, \delta_K\)
(i.e., the entire year) is just the product of the PTMs for each
interval:
\[ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).\]

For a stock spawning at some fraction of the year \(0<t_s<1\) in
interval \(\delta_j\), the fraction of time in season \(j\) is
\[\delta_{s,j} = t_s-\sum^{j-1}_{i=0}\delta_i
\] and the PTM defining the proportions in each state at time \(t_s\)
for age \(a\) is \begin{equation}\label{eq:ptm_spawn}
\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{s,j}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{s,j}).
\end{equation} Similarly, for an index \(m\) occurring at fraction of
the year \(t_m\) in interval \(\delta_j\) the proportions in each state
at the time of the observation is \begin{equation} \label{eq:ptm_index} 
\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{m,j}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{m,j}).
\end{equation}

\hypertarget{numbers-at-age}{%
\subsubsection*{Numbers at age}\label{numbers-at-age}}
\addcontentsline{toc}{subsubsection}{Numbers at age}

When there are \(n_R\) regions and \(n_F\) fleets, The vector of
abundance in each state at age \(a>1\) on January 1 is
\(\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}', \mathbf{0}')'\) where
\(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})'\) is the number
in the states corresponding to being alive in each region and
\(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each
fleet and dead from natural mortality because no age \(a\) fish have
died yet on January 1.

Each stock \(s\) is assumed to spawn and recruit in one region \(r_s\).
So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row
\(r = r_s\). Options for configuring recruitment (\(N_{y,1,r_s}\)) for
each stock are the same as previous versions of WHAM. If recruitment is
assumed to be a function of spawning stock biomass (SSB), it is only the
spawning population in region \(r_s\) at the time of spawning
constitutes the SSB in the stock-recruit function. However, models can
configure spawning individuals to occur in other regions at the time of
spawning. Aside from treating recruitment as a random walk, the general
model for annual recruitment as random effects is \begin{equation*}
\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1,r_s} =  f\left(\text{SSB}_{y-1,r_s}\right) + \varepsilon_{y,1,r_s}
\end{equation*} where\\
\[\text{SSB}_{s,y} =  \sum^A_{a=1}  w_{s,y,a} \text{mat}_{s,y,a} \mathbf{O}_{s,y,a,r_s}(t_s)' \mathbf{N}_{O,s,y,a}\]
where \(w_{s,y,a}\) is the mean weight at age of spawning individuals,
\(\text{mat}_{s,y,a}\) is the maturity at age, and
\(\mathbf{O}_{s,y,a,r_s}(t_s)\), the \(r_s\) column of the upper-left
submatrix of Eq. \ref{eq:ptm_spawn}, are the probabilities of surviving
and occurring in region \(r_s\) at time \(t_s\) given being alive in
each region at the start of the year.

As in previous versions of WHAM, the transitions in numbers at age from
one year to another after recruitment can be treated deterministically
or as functions of random effects. The predicted numbers at age in year
\(y\) at age \(a\) for a given stock are vector analogs
(\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the
standard WHAM model (Stock and Miller 2021). For ages
\(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected
number alive in each region at the beginning of the following year and
next age class age can be obtained from the first \(n_R\) elements of
the vector \[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\] The numbers
alive in each region can also be modeled more simply using the
sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions
in abundance at age is \begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*} for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}.
\end{equation*} When the transitions in abundance at age are treated
deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). The stock-
and region-specific errors \(\boldsymbol{\varepsilon}_{y,a}\) are
independent, but the same correlation structures as previous versions
are possible across ages and years for a given stock and region. When
there is autocorrelation with age, WHAM now assumes this applies only to
ages \(a>1\) by default so that recruitment random effects are
independent of those for for the annual transitions of older age
classes. So the general covariance structure for a given stock at ages
\(a>1\) in region \(r\) is \begin{equation*}
  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma^2_{N,r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}.
\end{equation*}

\hypertarget{initial-numbers-at-age}{%
\subsubsection*{Initial numbers at age}\label{initial-numbers-at-age}}
\addcontentsline{toc}{subsubsection}{Initial numbers at age}

Initial numbers at age for each stock and region can be treated as
age-specific fixed effects or with an equilibrium assumption as in
previous versions of WHAM. For the equilibrium option there are two
parameters for each stock: the stock-specific fully-selected fishing
mortality rate \(\log \widetilde{F}\) and the recruitment in year 1
\(\log N_{1,1,r_s}\). A stock-specific equilibrium fishing mortality at
age by fleet \({\tilde F}_{a,f}\) is the product of \(\widetilde{F}\)
and the selectivity across fleets in the first year \begin{equation*}
  \text{sel}_{1,a,f} = \frac{F_{1,a,f}}{\max_a \sum_{f=1}^{n_F} F_{1,a,f}}.
\end{equation*} We use \(\widetilde{F}_{a,f}\) to define an equilibrium
abundance per recruit by region at age \(a\) conditional on recruiting
to each region \begin{equation}\label{eq:npraa}
 \widetilde{\mathbf{O}}_{a} = \left\{
 \begin{array}{ll}
\prod^{a-1}_{j=0}\mathbf{O}_{j}  & 1\leq a<A\\
\left[\prod^{a-1}_{j=0}\mathbf{O}_{j}\right] \left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1} & a = A
 \end{array}
\right.
\end{equation} where \(\mathbf{O}_{j}\) is the equilibrium probability
of surviving age \(a\) and occurring in each region and
\(\mathbf{O}_{0} = \mathbf{I}\). Natural mortality and movement rates in
the first year of the model are also used in Eq. \ref{eq:npraa}. For the
plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is
a ``fundamental matrix'\,' derived using the matrix version of the
geometric series (Kemeny and Snell 1960). Recall that recruitment for
stock \(s\) only occur in region \(r_s\) so, the equilibrium initial
numbers at age \(a\) by region are
\[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}' \mathbf{N}_{O,1,1}.\]

The initial abundances at age can also be treated as independent or
autocorrelated (with age) random effects. Defining the vector of initial
abundance at age in region \(r\) as \(\mathbf{N}_{O,1,r}\), the general
model is
\[\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}\]
where
\[  Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.\]

\hypertarget{parametizing-movement}{%
\subsubsection*{Parametizing movement}\label{parametizing-movement}}
\addcontentsline{toc}{subsubsection}{Parametizing movement}

For each season, there are at most \(n_R-1\) parameters determining
movement among regions given starting an the season in region \(r\) in
either the sequential or simultaneous configurations. Movement
parameters are estimated on a transformed scale via a link function
\(g(\cdot)\). If survival and movement are occur simultaneously, the
parameters are estimated a log link function is used and if they are
separable, an additive logit link function (like a multinomial
regression) is used. On the transformed scale, the general model for the
movement parameter from region \(r\) to \(r'\) in season \(i\) and year
\(y\) for individuals of age \(a\) is a linear function of both random
and environmental effects: \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
\end{equation*} The random effects \(\epsilon_{r\rightarrow r',y,a,i}\)
are season-, and(or) region-to-region-specific and modeled most
generally as a two-dimensional first-order autoregressive random effects
with age and(or) year where the covariance is \begin{equation*}
  Cov\left(\epsilon_{r\rightarrow r',y,a,i},\epsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
\end{equation*} similar to how WHAM models variation in survival,
natural mortality, and selectivity. Effects of covariate \(E_k\) can be
age-, season-, and(or) region-to-region-specific
\(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial
options in the previous versions of WHAM for effects on recruitment and
natural mortality are available.

There is currently no likelihood component for tagging data. Therefore,
movement parameters would generally either need to be fixed or assumed
to have some prior distribution, possibly based on external parameter
estimates. We include prior distributions for the season and
region-to-region specific (mean) movement parameters which are treated
as random effects with the mean defined by the initial value of the
fixed effect counterpart and standard deviation \begin{equation*}
  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
  \end{equation*} When priors are used, the movement is defined instead
as \begin{equation*}
  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
  \end{equation*}

\hypertarget{natural-mortality}{%
\subsubsection*{Natural mortality}\label{natural-mortality}}
\addcontentsline{toc}{subsubsection}{Natural mortality}

Natural mortality options have been expanded in WHAM. When not
estimated, (mean) mortality rates may be stock-, region-, and
age-specific. When random effects are used, the same autocorrelation
structures with age and year as described by Stock and Miller (2021) can
be configured for a given stock and region. Any environmental covariate
effects can be stock-, region-, and age-specific. So the general model
for natural mortality is \begin{equation*}
  \log M_{y,a,r} = \theta_{M,r} + \epsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
\end{equation*} The general covariance structure for random effects are
modeled most generally as a two-dimensional first-order autoregressive
random effects with age and(or) year where the covariance is
\begin{equation*}
  Cov\left(\epsilon_{M,y,a,r},\epsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
\end{equation*}

\hypertarget{catch-observations}{%
\subsubsection*{Catch observations}\label{catch-observations}}
\addcontentsline{toc}{subsubsection}{Catch observations}

The log-normal distributional assumption for aggregate catch
observations is the same as Stock and Miller (2021), but the predicted
catch is now a function of catch from each stock starting the year in
each region. For a given stock and age, the numbers captured in each
fleet over the year are
\[\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}' \mathbf{N}_{O,s,y,a}\]
The predicted numbers caught be each fleet across stocks is
\[\widehat{\mathbf{N}}_{H,y,a} = \sum^{n_S}_{s=1} \widehat{\mathbf{N}}_{H,s,y,a}\]
and the predicted aggregate catch at age \(a\) is
\[\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \widehat{\mathbf{N}}_{H,y,a}\]
where \(\mathbf{c}_{y,a}\) is the vector of mean individual weight at
age \(a\) for each fleet and the aggregate catch by fleet is
\[\widehat{\mathbf{C}}_y = \sum^{A}_{a=1} \widehat{\mathbf{C}}_{y,a}\]
The log-aggregate catch observations for fleet \(f\) are normally
distributed
\[ \log C_{y,f} \sim \text{N}\left(\log \widehat {C}_{y,f}, \sigma^2_{y,f}\right).\]

The predicted numbers caught for each fleet \(f\) (row \(f\) of
\(\widehat{\mathbf{N}}_{H,y,a}\)) are used to make predicted age
composition observations as described by Stock and Miller (2021). Since
then, three additional likelihood options for age composition
observations have been added: a logistic-normal with AR(1) correlation
structure (Francis 2014), the alternative Dirichlet-multinomial
parameterization described by Thorson et al. (2017), and the
multivariate Tweedie (Thorson et al. 2023).

\hypertarget{index-observations}{%
\subsubsection*{Index observations}\label{index-observations}}
\addcontentsline{toc}{subsubsection}{Index observations}

For index \(m\) occurring in region \(r_m\) at fraction of the year
\(t_m\), the predicted abundance at \(t_m\) in region \(r_m\) is
\[\widehat{N}_{s,y,a,m} = \mathbf{O}_{s,y,a,r_m}(t_m)' \mathbf{N}_{O,s,y,a}\]
where \(\mathbf{O}_{s,y,a,r_m}(t_m)\), the \(r_m\) column of the
upper-left submatrix of Eq. \ref{eq:ptm_index}, are the probabilities of
surviving and occurring in region \(r_m\) at time \(t_m\) given being
alive in each region at the start of the year. The predicted index at
age is
\[\widehat{I}_{m,y,a} = q_{m,y} \text{sel}_{m,y,a}w_{m,y,a}\sum^{n_S}_{s = 1}\widehat{N}_{s,y,a,m}\]
where \(q_{m,y}\) is the catchability of the index in year \(y\),
\(\text{sel}_{m,y,a}\) is the selectivity and \(w_{m,y,a}\) is the
average weight of individuals at age \(a\) if the index is quantified in
biomass and \(w_{m,y,a} = 1\) if the index is quantified in numbers.
Predicted age composition observations are functions of
\(\widehat{I}_{m,y,a}\) as described by Stock and Miller (2021) and the
likelihood options are the same as those for catch explained above.

Catchability of the index can also be treated as functions of normal
random effects and(or) environmental covariate effects
\[log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}\]
where \(u_{m}\) and \(l_{m}\) are the upper and lower bounds of
catchability for index \(m\) (defaults are 0 and 1000) and the general
covariance structure for the annual random effects is first-order
autoregressive
\[Cov\left(\epsilon_{q,m,y},\epsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.\]

\hypertarget{weight-and-maturity-at-age}{%
\subsubsection*{Weight and Maturity at
age}\label{weight-and-maturity-at-age}}
\addcontentsline{toc}{subsubsection}{Weight and Maturity at age}

Weight and maturity at age are treated similarly to Stock and Miller
(2021). Annual weight at age matrices for each fleet are used to
calculate total catch and the weight at age is applied to catch numbers
at age for any stocks caught by the fleet. Similarly, when indices
and(or) associated age composition observations are measured in biomass,
the weight at age matrices for the index are applied to predicted
numbers at age of all stocks observed by the survey. Unique weight at
age and maturity at age matrices are allowed for each stock to calculate
spawning stock biomass.

\hypertarget{reference-points}{%
\subsubsection*{Reference points}\label{reference-points}}
\addcontentsline{toc}{subsubsection}{Reference points}

Currently a single F reference point \(\widetilde F\) is estimated
across stocks and regions and F by fleet and age is
\(\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}\). Selectivity is
determined as before when there are multiple fleets where
\(\text{sel}_{f,a}\) is determined by averaging F at age over a
user-defined set of years \begin{equation*}
  \text{sel}_{a,f} = \frac{\overline F_{a,f}}{\max_a \sum^{n_F}_{f=1}{\overline F}_{a,f}}
\end{equation*}

The equilibrium spawning stock biomass per recruit for stock \(s\) in
region \(r_s\) is defined as \begin{equation}\label{eq:ssbpr}
 \upphi_s(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{s,a,r_s,\cdot} \mathbf{O}_{s,a,\cdot,r_s}(t_s) w_{s,a} m_{s,a}
\end{equation} where \(w_{s,a}\) and \(m_{s,a}\) are the mean individual
weight and probability of maturity at age \(a\),
\(\widetilde{\mathbf{O}}_{s,a}\) are as described in Eq. \ref{eq:npraa},
and \(\mathbf{O}_{s,a}(t_s)\) is the \(n_R \times n_R\) upper-left
sub-matrix of eq. \ref{eq:ptm_spawn} with the probabilities of surviving
and occurring in each region \(r'\) at age \(a+t_s\) given starting in
region \(r\) at age \(a\). The further subscripts \(r_s,\cdot\) and
\(\cdot,r_s\) indictate row or column \(r_s\), respectively. Using these
rows and columns is required because of the assumption that spawning and
recruitment only occur in region \(r_s\).

The equilibrium spawning biomass per recruit (eq. \ref{eq:ssbpr}) is
conditional on the region of recruitment \(r_s\). The equilibrium
recruitment in each region \(\widetilde {\mathbf{N}}_{s,1}\), depends on
the stock dynamics. This version of WHAM currently only allows complete
spawning region fidelity so that a stock only spawns and recruits in a
single region. In this case, \(\widetilde {\mathbf{N}}_{s,1}\) will be
positive in the spawning region (\(r_s\)) and zero elsewhere. Similarly,
the row \(r_s\) of \(\mathbf{O}_s\) will be zero off of the diagonal.
The matrices of probabilities of surviving and occurring in each region,
\(\widetilde{\mathbf{O}}_{s,a}\) and \(\mathbf{O}_{s,a}(\delta_s)\), are
functions of the fishing mortality rates for fleets in each region
\(\widetilde F_{f,a}\).

The matrix equilibrium yield per recruit as a function of
\(\widetilde F\) is calculated as \begin{equation}\label{eq:ypr}
 \widetilde{Y}_s({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{s,a,r_s,\cdot} \mathbf{H}_{s,a} \mathbf{c}_{s,a}
\end{equation} where \(\mathbf{c}_a\) is the vector of mean individual
weight at age for each fleet, and \(\mathbf{H}_{s,a}\) is the submatrix
of the probabilities of being captured in each fleet over the interval
from \(a\) to \(a+1\), defined in eq. \ref{eq:ptm}.

As in previous versions of WHAM package, ``static'\,' reference points,
typically meant to be defined for prevailing conditions, average all of
the inputs to the spawning biomass and yield per recruit calculations
over the user-specified years (e.g., last 5 years of the model). This
same averaging is also applied to possibly time-varying movement
parameters.

For \(X\%\) SPR-based reference points, we use a Newton method and
iterate \begin{equation}\label{eq:newton}
  \log\widetilde{F}^{(i)} = \log\widetilde{F}^{(i-1)} - \frac{g\left(\log\widetilde{F}^{(i-1)}\right)}{g'\left(\log\widetilde{F}^{(i-1)}\right)}
\end{equation} where \(g(\log F)\) is the difference between the
weighted sums of spawning biomass per recruit at \(F\) and \(X\)\% of
unfished spawning biomass per recruit across stocks: \begin{equation*}
  g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
\end{equation*} where \(\upphi_s\left(F=0\right)\) is the equilibrium
unfished spawning biomass per recruit. \(g'(\log F)\) is the derivative
of \(g\) with respect to \(\log F\), and the weights to use for each
stock \(\lambda_s\) can be specified by the user or relative to the
average of recruitment for each stock over the same years the user
defines to calculate ``static'\,' equilibrium spawning biomass and
yield.

When a Beverton-Holt or Ricker stock recruit relationship is assumed, an
analogous Newton method is used to find \(\log F\) that maximizes yield
for MSY-based reference points. which are also a functions of the
equilibrium yield per recruit (\ref{eq:ypr}) and equilibrium
recruitment. The function \(g(\log F)\) in Eq. \ref{eq:newton} is the
first derivative of the yield curve with respect to \(\log F\).

\hypertarget{projections}{%
\subsubsection*{Projections}\label{projections}}
\addcontentsline{toc}{subsubsection}{Projections}

The projection options are generally the same as those for previous
versions of WHAM. When there is movement of any stocks, the user has the
option to project and use any random effects for time-varying movement
or use the average over user specified years, analogous to how natural
mortality can be treated in the projection period. The projections of
any environmental covariates has been revised to better include error in
the estimated latent covariate in any effects on the population in
projection years.

\hypertarget{application-to-black-sea-bass}{%
\section*{Application to black sea
bass}\label{application-to-black-sea-bass}}
\addcontentsline{toc}{section}{Application to black sea bass}

Prior to the most recent peer-reviewed assessment of black sea bass in
the NEUS, the stock was assessed using the Age-Structured Assessment
Program (ASAP) model (Legault and Restrepo 1999), a single-stock and
-region statistical catch-at-age model that estimates all model
parameters as fixed effects. Northern and southern components of the
NEUS black sea bass stock divided by the Hudson Canyon were separately
modeled in ASAP (reference map figure). Results from the separate ASAP
models have been aggregated for a unit-stock assessment. Furthermore,
the ASAP-based assessments have exhibited strong retrospective patterns
(Mohn 1999), and exploring alternative modeling approaches for the
northern and southern stock components has been a high priority for
management.

In 2023, a working group composed of scientists from federal, state, and
academic institutions determined an optimal data and model configuration
for the black sea bass stock using the multi-stock and multi-region
extension of WHAM described above (NEFSC 2023). This assessment included
the spatial features and hypothesized environmental drivers that were
prioritized research recommendations from previous black sea bass
assessments.

Below we describe the assumptions and configuration of the assessment
model as determined by the working group as well as the alternative
assumptions for recruitment and natural mortality in the models we fit
to evaluate alternative hypotheses of environmental effects.

\hypertarget{basic-structure}{%
\subsection*{Basic structure}\label{basic-structure}}
\addcontentsline{toc}{subsection}{Basic structure}

The first year being modeled for the population is 1989 and the fishery
and index data used in the model span from 1989 to 2021. There are two
stock components (north and south) modeled as separate populations that
spawn and recruit in respective regions separated by the Hudson Canyon.
We have observations for each of four total fishing fleets, where two
fishing fleets (Recreational and Commercial) operate in each region.

There are 11 seasonal intervals within each calendar year: five monthly
time intervals from Jan 1 to May 31, a spawning season from June 1 to
July 31, and five monthly intervals from August 1 to December 31. The
southern stock component is assumed to never move to the northern region
(Figure \ref{fig:migration-diagram}). For the northern component, a
proportion \(\mu_1\) can move to the southern region each month during
the last five months of the year, but no movement is allows from the
south to the north during this period. During the first four intervals
of the year a proportion \(\mu_2\) the northern component individuals in
the south can move back to the north, but no movement from the north to
south is allowed during this period. In the fifth interval (May), all
northern component individuals remaining in the south are assumed to
move back to the north for the subsequent spawning period. Survival and
movement occur sequentially in each interval and each of the two
movement proportions are assumed constant across intervals, ages, and
years.

The monthly movement matrices are \begin{equation*}
\boldsymbol{\mu}_{1} = 
  \begin{bmatrix}
     1-\mu_{1\rightarrow 2} & \mu_{1\rightarrow 2} \\
     0 & 1 \\
  \end{bmatrix}
\end{equation*} for the portion of the year after spawning and
\begin{equation*}
\boldsymbol{\mu}_{2} = 
  \begin{bmatrix}
     1 &  0 \\
     \mu_{2\rightarrow 1} & 1-\mu_{2\rightarrow 1} \\
  \end{bmatrix}
\end{equation*} for the portion of the year before spawning.

The constant or mean log-natural mortality rate is assumed to be
\(\log(0.4)\)

As noted in the description of the general WHAM model, tagging data are
not yet allowed. However, the working group also fit a Stock Synthesis
model (Methot and Wetzel 2013) which provided estimated movement rate
parameters and standard errors that were used to configure the priors
for WHAM (see Supplementary Materials).

\hypertarget{initial-abundance-at-age}{%
\subsection*{Initial abundance at age}\label{initial-abundance-at-age}}
\addcontentsline{toc}{subsection}{Initial abundance at age}

With the movement configuration, the northern origin fish (ages 2+) can
occur in the southern region on January 1. Estimating initial numbers at
age as separate parameters can be challenging even in single-stock
models. To avoid challenges with estimating initial numbers at age in
each region for the northern component in the two stock model, we used
the equilibrium assumption described above. Using this assumption two
parameters are estimated for each stock component: an initial
recruitment and an equilibrium fully-selected \(F\).

\hypertarget{recruitment-and-survival}{%
\subsection*{Recruitment and Survival}\label{recruitment-and-survival}}
\addcontentsline{toc}{subsection}{Recruitment and Survival}

For the northern population abundance at age 1 on January 1
(recruitment) is only allowed in the northern region, but given the
monthly movement described above, older individuals that previously
recruited in the northern region may occur in the southern region on
January 1. Therefore, a model with survival random effects will model
the transitions (survival/movement) of abundances at age of northern
origin fish in each region. All of the initial runs assumed variance
parameters for these random effects to be the same for northern origin
fish occurring in both regions on January 1. The base model assumes very
small variance for the transitions of northern fish in the southern
region, which is essentially the same as the deterministic transition
assumptions of a statistical catch at age model. We also allow 2DAR1
correlation for recruitment and survival for both the northern and
southern components. Unique variance and correlation parameters for the
recruitment and survival random effects are estimated for the northern
and southern components.

\hypertarget{uncertainty-recreational-cpa-index-observations}{%
\subsection*{Uncertainty recreational CPA index
observations}\label{uncertainty-recreational-cpa-index-observations}}
\addcontentsline{toc}{subsection}{Uncertainty recreational CPA index
observations}

The CVs provided by the analyses for the recreational catch per angler
(RecCPA) ranged between 0.02 and 0.06 which the working group felt did
not capture the true uncertainty in the index with regard to its
relationship to stock abundance. In many of the initial runs as well as
the base model we allowed a scalar multiple of the standard deviation of
the log aggregate index to be estimated for these indices in the
northern and southern regions. Models that successfully estimated these
scalars indicated standard deviations for these surveys to be
approximately 5 times the input value and this value was fixed in many
preliminary runs to avoid dealing with convergence problems. However,
the base model successfully estimated these scalars. The model estimates
are negligibly affected by estimating these scalars, but we felt
estimating these parameters allowed uncertainty in model output to be
more properly conveyed.

\hypertarget{index-and-catch-age-composition-observations}{%
\subsection*{Index and Catch age composition
observations}\label{index-and-catch-age-composition-observations}}
\addcontentsline{toc}{subsection}{Index and Catch age composition
observations}

Table \ref{tab:age-comp-sel-table}

\hypertarget{bottom-temperature-effects}{%
\subsection*{Bottom Temperature
effects}\label{bottom-temperature-effects}}
\addcontentsline{toc}{subsection}{Bottom Temperature effects}

All model fits also include bottom temperature observations for the
northern and southern regions from 1963 to 2021 and estimated standard
errors ranging between 0.03 and 0.09 degrees Celsius (NEFSC 2023). We
retained the assumption from the peer-reviewed assessment that treated
the latent bottom temperature covariates in each region as first-order
autoregressive processes.

We fit 14 models with alternative assumptions about the effects of
bottom temperature covariates, ranging from no effects to effects on
both regions for either recruitment or natural mortality at age 1 (Table
\ref{tab:model-desc-table}). These analyses derive from the hypothesis
that bottom temperature affects overwinter survival of fish where the
fish turn from age 0 to age 1 on January 1 (Miller et al. 2016a). This
temperature may be a proxy for temperature prior to January 1 and affect
survival during the end of the pre-recruit phase or natural mortality in
the early part of of the year after becoming age 1. Furthermore, we have
no direct observations of age 1 individuals from surveys until the
spring season each year. Therefore, we fit models with effects of
temperature on recruitment or natural mortality at age 1.

It is standard practice to treat annual recruitment as time-varying
deviations from a mean model and all models here treat recruitment
deviations as first-order autoregressive random effects. However, Miller
et al. (2018) showed how inferences of temperature effects on growth or
maturity parameters can be very different whether the null model without
the effect and models with effects also include random effects
representing residual temporal variation in parameters. Therefore, we
also explored whether including temporal random effects on age 1 natural
mortality M affected inferences on corresponding temperature effects.
Initially, we included random effects on age 1 natural mortality for
both the northern and southern stock components, but estimates of these
random effects and corresponding variance for the southern component
converged to 0 so these were not included in models presented here.

We assume the covariate in year \(y\) affects recruitment in the same
years because the covariate observations are from months January to
March. The fish are technically already 1 year old, but there are no
observations of these individuals until later in the year except
possibly in fishery catches which are accumulated over the whole year.
Expected log-recruitment is a linear function of log-recruitment at the
previous time step and bottom temperature

\begin{equation}\label{eq:expected-recruitment}
\log N_{s,r_s,y,1} = \mu_{R,s} + \beta_{R,s} x_y + \epsilon_{N,s,y,1}
\end{equation}

\begin{equation}\label{eq:expected-M1}
\log M_{s,y,1} = \mu_{M,s,1} + \beta_{M,s} x_y + \epsilon_{M,s,y}
\end{equation}

\hypertarget{model-fitting-diagnostics-and-comparisons}{%
\subsection*{Model fitting, diagnostics, and
comparisons}\label{model-fitting-diagnostics-and-comparisons}}
\addcontentsline{toc}{subsection}{Model fitting, diagnostics, and
comparisons}

As in Miller et al. (2016b), we also assessed the stability of the
AIC-based model selection over retrospective peels to guard against
previously noted changes in perception of covariate effects on
recruitment with increased length of the time series of observations
(Myers 1998). This retrospective examination was also recommended by
Brooks (2024).

We examined retrospective patterns by fitting models where the terminal
year is reduced sequentially by one year (peel) for seven years.
Therefore, there are 7 fits of the proposed base model with the time
series reduced by one to seven years. We calculated Mohn's \(\rho\) for
recruitment, SSB, and fully-selected \(F\). Absolute values of Mohn's
\(\rho\) near 0 imply no pattern in estimation of these quantities as
the time-series is sequentially extended.

Specifically, we use version XX and commit XXX of the wham package
(\url{https://github.com/timjmiller/wham/tree/lab}) for all results.

We performed a simulation self-test where new observations were
simulated conditional on all random effects estimated in the proposed
base model and the same model configuration was fit to each of the
simulated data sets.

One-step-ahead residuals can now be calculated for all index, catch,
environmental covariate observations using methods described by Thygesen
et al. (2017) and Trijoulet et al. (2023).

\hypertarget{one-step-ahead-residuals}{%
\subsubsection*{One-step-ahead
residuals}\label{one-step-ahead-residuals}}
\addcontentsline{toc}{subsubsection}{One-step-ahead residuals}

\hypertarget{retrospective-patterns}{%
\subsubsection*{Retrospective patterns}\label{retrospective-patterns}}
\addcontentsline{toc}{subsubsection}{Retrospective patterns}

\hypertarget{results}{%
\section*{Results}\label{results}}
\addcontentsline{toc}{section}{Results}

Although there was strong retrospective patterns in the most recent
management track assessment for the northern component of the stock,
there was no evidence of patterns for northern SSB (Mohn's \(\rho\) =
-0.058) and \(F\) (Mohn's \(\rho\) = 0.06) for the proposed base model
(Figures \ref{fig:North-retro-ssb} and \ref{fig:North-retro-F}).
Similarly, no patterns were exhibited for the southern component of SSB
(Mohn's \(\rho\) = -0.019) and regional \(F\) (Mohn's \(\rho\) = -0.05)
(Figures \ref{fig:South-retro-ssb} and \ref{fig:South-retro-F}).

For 7 of the the simulated data sets the model failed to optimize. The
maximum absolute gradient was \(<10^{-6}\) for only 9 and \(<10^{-4}\)
for 52 of the 93 successfully fitted models. The poor convergence
appeared to be attributable to the estimation of the scalar for the
standard errors of the log-transformed northern Recreational CPA index
for which estimates tended to 0 for nearly all of the fits (\(<0.01\)
for 83 fits). However, even across all fits including those with poor
convergence, the SSB estimates appeared to be reliable (Figure
\ref{fig:self-test-res-1}).

\hypertarget{bottom-temperature-effect-on-recruitment}{%
\subsection*{Bottom temperature effect on
recruitment}\label{bottom-temperature-effect-on-recruitment}}
\addcontentsline{toc}{subsection}{Bottom temperature effect on
recruitment}

We found the best model of bottom temperature covariate effects included
the effect only on the recruitment of the northern stock, which is the
configuration in the proposed base model although the difference in AIC
for the model including effects on both regional components was small
suggesting some evidence for this hypothesis as well (Table
\ref{tab:compare-table}) . We also found AIC to strongly favor models
that included at least an effect on the northern component across all
retrospective peels (Table \ref{tab:peel-compare-table}). The posterior
estimate of the bottom temperature covariate match the observations well
because of the high precision of the observations (Figure
\ref{fig:bottom-temp}). The residual variation in the standard deviation
of recruitment random effects is reduced because the expected
recruitment (Eq. \ref{eq:expected-recruitment}) is a function of the
covariate (Figure \ref{fig:relative-recruitment-bt}). This effect is
included in the proposed base model (Figure \ref{fig:rec-bottom-temp}).
Because the covariate is technically for temperature after the beginning
of the calendar year when the fish are considered age 1 in the model, a
comparison of the proposed base model to one with the bottom temperature
effect on natural mortality at age 1 instead might be of interest in the
future. An investigation of higher order orthogonal polynomial effects
might also be of interest particularly for the southern component which
might be experiencing more frequently higher temperatures less favorable
to overwinter survival.

\hypertarget{discussion}{%
\section*{Discussion}\label{discussion}}
\addcontentsline{toc}{section}{Discussion}

The new features of Multi-WHAM include

\begin{itemize}
\tightlist
\item
  seasonal intervals within years
\item
  variation in movement rates by stock, region-to-region, season, age,
  and year
\item
  effects of environmental covariates on movement rates by stock,
  region-to-region, season, and age
\item
  effects of environmental covariates on mortality rates by stock,
  region, and age
\item
  effects of environmental covariates on recruitment by stock
\item
  stock-specific stock-recruitment models
\item
  priors for movement rates
\item
  seasonal operation of fleets
\item
  mortality and movement modeled sequentially or simultaneously
\item
  options for weighting of stock-specific SSB/R for global SPR-based
  reference points
\end{itemize}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-brooks24}{}}%
Brooks, E.N. 2024. Pragmatic approaches to modeling recruitment in
fisheries stock assessment: A perspective. Fisheries Research
\textbf{270}: 106896.
doi:\href{https://doi.org/10.1016/j.fishres.2023.106896}{10.1016/j.fishres.2023.106896}.

\leavevmode\vadjust pre{\hypertarget{ref-francis14}{}}%
Francis, R.I.C.C. 2014. Replacing the multinomial in stock assessment
models: A first step. Fisheries Research \textbf{151}: 70--84.
doi:\href{https://doi.org/10.1016/j.fishres.2013.12.015}{10.1016/j.fishres.2013.12.015}.

\leavevmode\vadjust pre{\hypertarget{ref-kemenysnell60}{}}%
Kemeny, J.G., and Snell, J.L. 1960. Finite markov chains. D. Van
Nostrand Company, Princeton, New Jersey.

\leavevmode\vadjust pre{\hypertarget{ref-kristensenetal16}{}}%
Kristensen, K., Nielsen, A., Berg, C., Skaug, H., and Bell, B.M. 2016.
{TMB}: Automatic differentiation and {Laplace} approximation. Journal of
Statistical Software \textbf{70}: 1--21.
doi:\href{https://doi.org/10.18637/jss.v070.i05}{10.18637/jss.v070.i05}.

\leavevmode\vadjust pre{\hypertarget{ref-legaultrestrepo99}{}}%
Legault, C.M., and Restrepo, V.R. 1999. A flexible forward
age-structured assessment program. Col. Vol. Sci. Pap. ICCAT
\textbf{49}(2): 246--253.

\leavevmode\vadjust pre{\hypertarget{ref-legaultetal23}{}}%
Legault, C.M., Wiedenmann, J., Deroba, J.J., Fay, G., Miller, T.J.,
Brooks, E.N., Bell, R.J., Langan, J., Cournane, J., Jones, A.W., and
Muffley, B. 2023. Data rich but model resistant: An evaluation of
data-limited methods to manage fisheries with failed age-based stock
assessments. Canadian Journal of Fisheries and Aquatic Sciences
\textbf{80}(1): 27--42.
doi:\href{https://doi.org/10.1139/cjfas-2022-0045}{10.1139/cjfas-2022-0045}.

\leavevmode\vadjust pre{\hypertarget{ref-methotwetzel13}{}}%
Methot, R.D., and Wetzel, C.R. 2013. Stock {S}ynthesis: A biological and
statistical framework for fish stock assessment and fishery management.
Fisheries Research \textbf{142}(1): 86--99.

\leavevmode\vadjust pre{\hypertarget{ref-milleretal16_yoy_survival}{}}%
Miller, A.S., Shepherd, G.R., and Fratantoni, P.S. 2016a. {PLOS} {ONE}
\textbf{11}(1): e0147627.
doi:\href{https://doi.org/10.1371/journal.pone.0147627}{10.1371/journal.pone.0147627}.

\leavevmode\vadjust pre{\hypertarget{ref-millerandersen08}{}}%
Miller, T.J., and Andersen, P.K. 2008. A finite-state continuous-time
approach for inferring regional migration and mortality rates from
archival tagging and conventional tag-recovery experiments. Biometrics
\textbf{64}(4): 1196--1206.
doi:\href{https://doi.org/10.1111/j.1541-0420.2008.00996.x}{10.1111/j.1541-0420.2008.00996.x}.

\leavevmode\vadjust pre{\hypertarget{ref-milleretal16}{}}%
Miller, T.J., Hare, J.A., and Alade, L.A. 2016b. A state-space approach
to incorporating environmental effects on recruitment in an
age-structured assessment model with an application to southern {New
England} yellowtail flounder. Can. J. Fish. Aquat. Sci. \textbf{73}(8):
1261--1270.
doi:\href{https://doi.org/10.1139/cjfas-2015-0339}{10.1139/cjfas-2015-0339}.

\leavevmode\vadjust pre{\hypertarget{ref-milleretal18}{}}%
Miller, T.J., O'Brien, L., and Fratantoni, P.S. 2018. Temporal and
environmental variation in growth and maturity and effects on management
reference points of {Georges Bank Atlantic} cod. Can. J. Fish. Aquat.
Sci. \textbf{75}(12): 2159--2171.
doi:\href{https://doi.org/10.1139/cjfas-2017-0124}{10.1139/cjfas-2017-0124}.

\leavevmode\vadjust pre{\hypertarget{ref-millerstock20}{}}%
Miller, T.J., and Stock, B.C. 2020. The {Woods Hole Assessment Model}
({WHAM}). https://timjmiller.github.io/wham.

\leavevmode\vadjust pre{\hypertarget{ref-mohn99}{}}%
Mohn, R. 1999. The retrospective problem in sequential population
analysis: An investigation using cod fishery and simulated data. ICES
Journal of Marine Science \textbf{56}(4): 473--488.

\leavevmode\vadjust pre{\hypertarget{ref-myers98}{}}%
Myers, R.A. 1998. When do environment--recruitment correlations work?
Reviews in Fish Biology and Fisheries \textbf{8}(3): 229--249.
doi:\href{https://doi.org/10.1023/A:1008828730759}{10.1023/A:1008828730759}.

\leavevmode\vadjust pre{\hypertarget{ref-nefsc23}{}}%
NEFSC. 2023. Report of the black sea bass (\emph{{C}entropristis}
\emph{striata}) research track stock assessment working group.
{Available} at
https://www.mafmc.org/s/a\_2023\_BSB\_UNIT\_RTWG\_Report\_V2\_12\_2\_2023-1.pdf.

\leavevmode\vadjust pre{\hypertarget{ref-nielsenberg14}{}}%
Nielsen, A., and Berg, C.W. 2014. Estimation of time-varying selectivity
in stock assessments using state-space models. Fisheries Research
\textbf{158}: 96--101.
doi:\href{https://doi.org/10.1016/j.fishres.2014.01.014}{10.1016/j.fishres.2014.01.014}.

\leavevmode\vadjust pre{\hypertarget{ref-stockmiller21}{}}%
Stock, B.C., and Miller, T.J. 2021. The {Woods Hole Assessment Model}
({WHAM}): A general state-space assessment framework that incorporates
time- and age-varying processes via random effects and links to
environmental covariates. Fisheries Research \textbf{240}: 105967.
doi:\href{https://doi.org/10.1016/j.fishres.2021.105967}{10.1016/j.fishres.2021.105967}.

\leavevmode\vadjust pre{\hypertarget{ref-stocketal21}{}}%
Stock, B.C., Xu, H., Miller, T.J., Thorson, J.T., and Nye, J.A. 2021.
Implementing two-dimensional autocorrelation in either survival or
natural mortality improves a state-space assessment model for {Southern
New England}-{Mid Atlantic} yellowtail flounder. Fisheries Research
\textbf{237}: 105873.
doi:\href{https://doi.org/10.1016/j.fishres.2021.105873}{10.1016/j.fishres.2021.105873}.

\leavevmode\vadjust pre{\hypertarget{ref-thorsonetal17}{}}%
Thorson, J.T., Johnson, K.F., Methot, R.D., and Taylor, I.G. 2017.
Model-based estimates of effective sample size in stock assessment
models using the {D}irichlet-multinomial distribution. Fisheries
Research \textbf{192}: 84--93.
doi:\href{https://doi.org/10.1016/j.fishres.2016.06.005}{10.1016/j.fishres.2016.06.005}.

\leavevmode\vadjust pre{\hypertarget{ref-thorsonetal23}{}}%
Thorson, J.T., Miller, T.J., and Stock, B.C. 2023. The
multivariate-tweedie: A self-weighting likelihood for age and length
composition data arising from hierarchical sampling designs. ICES
Journal of Marine Science \textbf{80}(10): 2630--2641.
doi:\href{https://doi.org/10.1093/icesjms/fsac159}{10.1093/icesjms/fsac159}.

\leavevmode\vadjust pre{\hypertarget{ref-thygesenetal17}{}}%
Thygesen, U.H., Albertsen, C.M., Berg, C.W., Kristensen, K., and
Nielsen, A. 2017. {Validation of ecological state space models using the
Laplace approximation}. Environmental and Ecological Statistics
\textbf{24}(2): 317--339.
doi:\href{https://doi.org/10.1007/s10651-017-0372-4}{10.1007/s10651-017-0372-4}.

\leavevmode\vadjust pre{\hypertarget{ref-trijouletetal23}{}}%
Trijoulet, V., Albertsen, C.M., Kristensen, K., Legault, C.M., Miller,
T.J., and Nielsen, A. 2023. Model validation for compositional data in
stock assessment models: Calculating residuals with correct properties.
Fisheries Research \textbf{257}: 106487.
doi:\url{https://doi.org/10.1016/j.fishres.2022.106487}.

\end{CSLReferences}

\pagebreak

\hypertarget{appendix-a}{%
\section*{Appendix A}\label{appendix-a}}
\addcontentsline{toc}{section}{Appendix A}

\setcounter{table}{0}
\renewcommand\thetable{A\arabic{table}}
\input{symboltable.tex}

\pagebreak

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{bsb_movement_diagram}

}

\caption{Diagram of intervals within the year and configuation of the dynamics of each component of the BSB population.}\label{fig:migration-diagram}
\end{figure}

\setcounter{table}{0}
\renewcommand\thetable{\arabic{table}}

\begin{landscape}\begin{table}

\caption{\label{tab:age-comp-sel-table}Configuration of Age composition likelihood, mean selectivity model, and selectivity random effects for each age composition data component.}
\centering
\begin{tabular}[t]{llll}
\toprule
Data component & Age Composition Likelihood & Mean Selectivity model & Random effects configuration\\
\midrule
North Commercial & Dirichlet-Multinomial & age-specific (flat-topped at ages > 3) & 2D-AR1 (age and year)\\
North Recreational & Logistic-normal (0s as missing) & age-specific (flat-topped at ages > 6) & 2D-AR1 (age and year)\\
South Commercial & Logistic-normal (AR1, 0s as missing) & logistic & None\\
South Recreational & Logistic-normal (AR1, 0s as missing) & logistic & None\\
North Recreational CPA & Logistic-normal (0s as missing) & age-specific (flat-topped at ages > 1) & AR1 (year)\\
\addlinespace
North VAST & Dirichlet-Multinomial & age-specific (flat-topped at ages > 4) & 2D-AR1 (age and year)\\
South Recreational CPA & Logistic-normal (AR1, 0s as missing) & age-specific (flat-topped at ages > 2) & None\\
South VAST & Logistic-normal (AR1, 0s as missing) & age-specific (flat-topped at ages > 1) & None\\
\bottomrule
\end{tabular}
\end{table}
\end{landscape}

\begin{table}

\caption{\label{tab:model-desc-table}Assumptions for temperature effects and random effects for age 1 natural mortality for each model.}
\centering
\begin{tabular}[t]{llll}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Temperature Effect} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-3}
Model & North & South & $M$ at age 1 random effects\\
\midrule
$M_{0}$ & -- & -- & none\\
$M_{1}$ & Recruitment & -- & none\\
$M_{2}$ & -- & Recruitment & none\\
$M_{3}$ & Recruitment & Recruitment & none\\
$M_{4}$ & $M$ at age 1 & -- & none\\
\addlinespace
$M_{5}$ & -- & $M$ at age 1 & none\\
$M_{6}$ & $M$ at age 1 & $M$ at age 1 & none\\
$M_{7}$ & -- & -- & time-varying\\
$M_{8}$ & Recruitment & -- & time-varying\\
$M_{9}$ & -- & Recruitment & time-varying\\
\addlinespace
$M_{10}$ & Recruitment & Recruitment & time-varying\\
$M_{11}$ & $M$ at age 1 & -- & time-varying\\
$M_{12}$ & -- & $M$ at age 1 & time-varying\\
$M_{13}$ & $M$ at age 1 & $M$ at age 1 & time-varying\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:diff-aic-table}Difference between AIC and the lowest AIC for each model by retrospective peel.}
\centering
\begin{tabular}[t]{lrrrrrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Peel} \\
\cmidrule(l{3pt}r{3pt}){2-9}
Model & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
\midrule
$M_{0}$ & 14.53 & 16.09 & 13.78 & 13.69 & 13.79 & 17.02 & 8.47 & 8.22\\
$M_{1}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
$M_{2}$ & 15.29 & 16.77 & 14.80 & 14.71 & 14.85 & 18.11 & 9.20 & 9.22\\
$M_{3}$ & 3.49 & 5.42 & 4.42 & 4.69 & 5.02 & 8.75 & 0.73 & 1.00\\
$M_{4}$ & 16.51 & 18.05 & 15.63 & 15.45 & 15.63 & 18.79 & 10.08 & 9.85\\
\addlinespace
$M_{5}$ & 16.09 & 17.51 & 15.30 & 15.16 & 15.29 & 18.41 & 9.86 & 9.71\\
$M_{6}$ & 18.07 & 19.46 & 17.14 & 16.92 & 17.12 & 20.18 & 11.46 & 11.33\\
$M_{7}$ & 17.05 & 18.48 & 15.57 & 14.96 & 15.39 & 18.19 & 9.71 & 9.42\\
$M_{8}$ & 5.02 & 6.87 & 4.90 & 4.66 & 5.28 & 8.59 & 0.98 & 0.99\\
$M_{9}$ & 17.82 & 19.17 & 16.59 & 15.98 & 16.45 & 19.28 & 10.44 & 10.43\\
\addlinespace
$M_{10}$ & 5.78 & 7.56 & 5.92 & 5.68 & 6.34 & 9.68 & 1.71 & 1.99\\
$M_{11}$ & 19.02 & 20.48 & 17.57 & 16.96 & 17.39 & 20.19 & 11.68 & 11.39\\
$M_{12}$ & 18.61 & 19.90 & 17.09 & 16.43 & 16.88 & 19.58 & 11.10 & 10.92\\
$M_{13}$ & 20.58 & 21.89 & 19.09 & 18.43 & 18.88 & 21.57 & 13.07 & 12.88\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:aic-wts-table}Model AIC weights for each retrospective peel.}
\centering
\begin{tabular}[t]{lrrrrrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Peel} \\
\cmidrule(l{3pt}r{3pt}){2-9}
Model & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
\midrule
$M_{0}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.01 & 0.01\\
$M_{1}$ & 0.76 & 0.89 & 0.80 & 0.80 & 0.83 & 0.97 & 0.36 & 0.38\\
$M_{2}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
$M_{3}$ & 0.13 & 0.06 & 0.09 & 0.08 & 0.07 & 0.01 & 0.25 & 0.23\\
$M_{4}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
\addlinespace
$M_{5}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
$M_{6}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
$M_{7}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
$M_{8}$ & 0.06 & 0.03 & 0.07 & 0.08 & 0.06 & 0.01 & 0.22 & 0.23\\
$M_{9}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
\addlinespace
$M_{10}$ & 0.04 & 0.02 & 0.04 & 0.05 & 0.04 & 0.01 & 0.15 & 0.14\\
$M_{11}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
$M_{12}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
$M_{13}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:rho-table}Mohn's $\rho$ for SSB, R, average F at ages 6 and 7 in northern and southern regions.}
\centering
\begin{tabular}[t]{lrrrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{SSB} & \multicolumn{2}{c}{Average F} & \multicolumn{2}{c}{Recruitment} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
Model & North & South & North & South & North & South\\
\midrule
$M_{0}$ & -0.031 & -0.026 & 0.032 & -0.043 & 0.211 & -0.007\\
$M_{1}$ & -0.040 & -0.023 & 0.041 & -0.048 & 0.181 & -0.024\\
$M_{2}$ & -0.031 & -0.027 & 0.032 & -0.043 & 0.211 & -0.007\\
$M_{3}$ & -0.031 & -0.027 & 0.031 & -0.043 & 0.193 & -0.007\\
$M_{4}$ & -0.032 & -0.026 & 0.032 & -0.044 & 0.268 & -0.006\\
\addlinespace
$M_{5}$ & -0.031 & -0.026 & 0.032 & -0.045 & 0.210 & -0.008\\
$M_{6}$ & -0.032 & -0.025 & 0.032 & -0.045 & 0.267 & -0.007\\
$M_{7}$ & -0.033 & -0.026 & 0.033 & -0.043 & 0.225 & -0.007\\
$M_{8}$ & -0.035 & -0.027 & 0.035 & -0.043 & 0.186 & -0.007\\
$M_{9}$ & -0.033 & -0.027 & 0.033 & -0.043 & 0.224 & -0.008\\
\addlinespace
$M_{10}$ & -0.035 & -0.027 & 0.035 & -0.043 & 0.186 & -0.008\\
$M_{11}$ & -0.032 & -0.026 & 0.032 & -0.044 & 0.253 & -0.007\\
$M_{12}$ & -0.033 & -0.026 & 0.033 & -0.045 & 0.224 & -0.008\\
$M_{13}$ & -0.032 & -0.025 & 0.032 & -0.045 & 0.253 & -0.008\\
\bottomrule
\end{tabular}
\end{table}
\clearpage

\hypertarget{supplemental-materials}{%
\section*{Supplemental Materials}\label{supplemental-materials}}
\addcontentsline{toc}{section}{Supplemental Materials}

\hypertarget{deriving-the-prior-distribution-for-movement-parameters}{%
\subsection*{Deriving the prior distribution for movement
parameters}\label{deriving-the-prior-distribution-for-movement-parameters}}
\addcontentsline{toc}{subsection}{Deriving the prior distribution for
movement parameters}

The working group fit a Stock Synthesis model (Methot and Wetzel 2013)
that included tagging data with 2 seasons (6 months each) and 2 regions
where a proportion \(\mu^*_1\) of the northern component moves to the
south in one season and some proportion \(\mu^*_{2\rightarrow 1}\) move
back to the south in the second season (NEFSC 2023). The seasonal
movement matrices for each season are \begin{equation*}
\boldsymbol{\mu}^*_{1} = 
  \begin{bmatrix}
     1-\mu^*_{1\rightarrow 2} & \mu^*_{1\rightarrow 2} \\
     0 & 1 \\
  \end{bmatrix}
\end{equation*} and \begin{equation*}
\boldsymbol{\mu}_{2} = 
  \begin{bmatrix}
     1 &  0 \\
     \mu^*_{2\rightarrow 1} & 1-\mu^*_{2\rightarrow 1} \\
  \end{bmatrix}.
\end{equation*}

To obtain estimates of movement proportions for the monthly intervals in
the WHAM model, the half-year movement matrices were converted to
monthly movement matrices by taking the root \(z_k\) of
\(\boldsymbol{\mu}^*_{k}\) which are defined by the number of months of
movement for each season (5 and 4, respectively). The roots of the
matrices are calculated using an eigen decomposition of the matrices
\[ \boldsymbol{\mu}_k =  \left(\boldsymbol{\mu}_k^*\right)^{z_k} = \mathbf{V}_k \mathbf{D}_k^{z_k} \mathbf{V}_k^{-1}\]
where \(z_1 = 1/5\) for and \(z_2 = 1/4\), and \(\mathbf{V}_{k}\) and
\(\mathbf{D}_{k}\) are the matrix of eigenvectors (columnwise) and the
diagonal matrix of corresponding eigenvalues of
\(\boldsymbol{\mu}^*_k\). The working group used a parametric bootstrap
approach to determine an appropriate standard deviation for the prior
distribution for the movement parameters. Stock Synthesis also estimates
parameters on a transformed scale, but different from WHAM:
\[\mu^*_{r\rightarrow r'} = \frac{1}{1 + 2e^{-x_{r\rightarrow r'}}}\]
The estimated parameters and standard errors from the Stock Synthesis
model were \(x_{1\rightarrow 2}=-1.44\) and \(x_{2\rightarrow 1}=1.94\)
and \(SE(x_{1\rightarrow 2})) = 0.21\) and
\(SE(x_{2\rightarrow 1})) = 0.37\). The resulting in the estimated
proportions were \(\mu^*_{1\rightarrow 2}=0.11\) and
\(\mu^*_{2\rightarrow 1}=0.78\).

In WHAM, an additive logit transformation is used which is simply a
logit transformation when there are only two regions: \[
\mu_{r\rightarrow r'} = \frac{1}{1+e^{-y_{r\rightarrow r'}}}.
\] We simulated 1000 values from a normal distribution with mean and
standard deviation defined by the parameter estimate and standard error
\(\tilde x_{{r\rightarrow r'},b} \sim N(x_{r\rightarrow r'}, SE(x_{r\rightarrow r'}))\)
from the Stock Synthesis model. For each simulated value we constructed
\(\tilde {\boldsymbol{\mu}}^*_{{r\rightarrow r'},b}\), took the
appropriate root and calculated inverse logit for
\(\tilde y_{{r\rightarrow r'},b}\). We calculated the mean and standard
deviation of the values \(y_{i,b}\). The mean values did not differ
meaningfully from the transformation of the original estimates
(\(y_{1\rightarrow 2} = -3.79\) and \(y_{2\rightarrow 1} = -0.79\)) and
the standard deviation was approximately 0.2 for both parameters.

\end{document}

### 2024-09-20 paper/symboltable.tex
start repo

\begin{longtable}[c]{r p{0.85\textwidth}}
\caption{Definition of terms.\label{symbols}}%
\\ \hline \hline \endfirsthead 
\caption[]{(Continued)} %
\\ \hline \hline \endhead
 \hline \hline \endfoot
$i$ & Seasonal time interval\\
$\delta_i$ & Length of seasonal time interval $i$\\
$a$ & Age class\\
$y$ & Year\\
$A$ & Last age class (``plus group'')\\
$r$ & Region\\
$f$ & Fishing fleet\\
$n_F$ & Number of fishing fleets\\
$s$ & Stock\\
$n_R$ & Number of regions\\
$\mathbf{P}_{y,a,i}$ & Probability transition matrix for year $y$, age $a$, and season $i$\\
$\mathbf{O}_{y,a,i}$ & submatrix of $\mathbf{P}_{y,a,i}$ of probabilities of surviving and occuring in each region for year $y$, age $a$, and season $i$\\
$\mathbf{H}_{y,a,i}$ & submatrix of $\mathbf{P}_{y,a,i}$ of probabilities of being captured in each fishing fleet for year $y$, age $a$, and season $i$\\
$\mathbf{I}_{H}$ & $n_f$ x $n_f$ identity matrix \\
$m$ & Index observation\\
$O_{y,a,i}(r,r')$ & For year $y$, age $a$ and season $i$, the probability of surviving and occurring in region $r'$ given beginning the interval alive in region $r$\\
$H_{y,a,i}(r,f)$ & For year $y$, age $a$ and season $i$, the probability of being captured in fleet $f$ given beginning the interval alive in region $r$\\
$S_{y,a,i}$ & For year $y$, age $a$ and season $i$, the probability of surviving the interval (1 region model)\\
$F_{y,a,i,f}$ & Fishing mortality rate for fleet $f$ in year $y$ at age $a$ in seasonal interval $i$\\
$M_{y,a}$ &  Natural mortality rate in year $y$ at age $a$ (single region)\\
$M_{y,a,r}$ & Natural mortality rate in region $r$ and year $y$ at age $a$\\
$Z_{y,a,i}$ & Total mortality rate in year $y$ at age $a$ in seasonal interval $i$ (single region)\\
$Z_{y,a,i,r}$ & Total mortality rate in region $r$ and year $y$ at age $a$ in seasonal interval $i$\\
$\mathbf{S}_{y,a,i}$ & matrix of of probabilities of surviving in each region over the interval for season $i$, year $y$, age $a$ \\
$\boldsymbol{\mu}_{y,a,i}$ & matrix of of probabilities of moving or staying in each region at the end of season $i$ in year $y$ and age $a$ \\
$\mu_{r\rightarrow r',y,a,i}$ & For year $y$, age $a$ and seasonal interval $i$, either the probability of moving at the end of the interval or instantanteous rate of movement from region $r$ to region $r'$\\
$r_f$ & region where fleet $f$ operates\\
$\mathbf{N}_{y,a}$ & Column vector of abundances by region at age $a$ in year $y$\\
$\mathbf{A}_{y,a,i}$ & instantaneous rate matrix for seasonal interval $i$, year $y$, and age $a$\\
$a_{y,a,i,r}$ & For year $y$, age $a$ and seasonal interval $i$, the hazard or negative sum of the instantaneous rates of mortality and movement from the state corresonding to being alive in region $r$ \\
$\mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K)$ & Probability transition matrix for year $y$ and age $a$ over seasonal intervals $\delta_1,\ldots, \delta_K$\\
$K$ & Number of seasons in the annual time step\\
$t_s$ & fraction of the annual time step when spawning occurs for stock $s$\\
$\delta_{s,j}$ & fraction of the annual time step between $t_s$ and the end of season $j-1$\\
$t_m$ & fraction of the annual time step when index $m$ observers the population\\
$\delta_{m,j}$ & fraction of the annual time step between $t_m$ and the end of season $j-1$\\
$\mathbf{N}_{y,a}$ & Abundance at age $a$ in year $y$ in each of the living and mortality states on January 1\\
$\mathbf{N}_{O,y,a}$ & Abundance at age $a$ in year $y$ alive in each region on January 1\\
$N_{y,a,r}$ & Abundance at age $a$ in year $y$ alive in region $r$ on January 1\\
$r_s$ & region where stock $s$ spawns and recruits\\
$\text{SSB}_{s,y}$ & Spawning stock biomass for stock $s$ in year $y$\\
$\varepsilon_{y,a,r}$ & Random error for abundance at age $a$ in year $y$ in region $r$\\
$w_{s,y,a}$ & mean individual weight for stock $s$ at age $a$ in year $y$\\
$\text{mat}_{s,y,a}$ & proportion mature at age $a$ in year $y$ for stock $s$\\
$\mathbf{O}_{s,y,a,r_s}(t_s)$ & Probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year\\
$\boldsymbol{\varepsilon}_{y,a}$ & vector of random errors for abundance alive in each region on January 1 of year $y$ at age $a$\\
$\sigma_{N,r}$ & standard deviation parameter for abundance at age random effects in region $r$\\
$\rho_{N,\text{age},r}$ & first order auto-regressive correlation parameter across age for abundance at age random effects in region $r$\\
$\rho_{N,\text{year},r}$ & first order auto-regressive correlation parameter across year for abundance at age random effects in region $r$\\
$\text{sel}_{1,a,f}$ & selectivity at age $a$ for fleet $f$ in the first year\\
$F_{1,a,f}$ & fishing mortality rate at age $a$ for fleet $f$ in the first year\\
$\widetilde{F}_{a,f}$ & equilibrium fishing mortality rate at age $a$ for fleet $f$\\
$\mathbf{O}_{a}$ & proportion surviving the year at age $j$ and occuring in each region (columns) given alive on January 1 in each region (rows)\\
$\widetilde{\mathbf{O}}_{a}$ & equilibrium proportions alive in each region at age $a$ (columns) given recruitment in each region (rows)\\
$\mathbf{N}_{O,1,r}$ & vector of abundance by age in region $r$ in the first year\\
$\theta_{N_1,r}$ & mean parameter for initial numbers at age random effects in the first year for region $r$\\
$\boldsymbol{\varepsilon}_{N_1,r}$ & vector of random effects by age for initial numbers at age in the first year for region $r$\\ 
$\sigma_{N_1,r}$ & standard deviation parameter for initial numbers at age random effects in the first year for region $r$\\
$\rho_{N_1,r}$ & first order auto-regressive correlation parameter for initial numbers at age random effects in the first year for region $r$\\
$\mu_{r\rightarrow r',y,a,i}$ & movement from region $r$ to region $r'$ in seasonal interval $i$ and year $y$ at age $a$\\
$g(\mu_{r\rightarrow r',y,a,i})$ & link function for movement $\mu_{r\rightarrow r',y,a,i}$\\
$\theta_{r\rightarrow r',i}$ & mean parameter across age and year for movement from region $r$ to region $r'$ in seasonal interval $i$\\
$\epsilon_{r\rightarrow r',y,a,i}$ &  random effect parameter for movement from region $r$ to region $r'$ in seasonal interval $i$ and year $y$ at age $a$\\
$n_E$ & number of environmental covariates\\
$\beta_{r \rightarrow r',a,i,k}$ & effect of environmental covariate $k$ on movement from region $r$ to $r'$ at age $a$ in seasonal interval $i$\\
$E_{k,y}$ & latent environmetal covariate $k$ affecting the population in year $y$\\
$\sigma_{r \rightarrow r',i}$ & standard deviation parameter for movement random effects from region $r$ to $r'$ in seasonal interval $i$\\
$\rho_{r \rightarrow r',\text{age},i}$ & first order auto-regressive correlation parameter across age for movement random effects from region $r$ to $r'$ in seasonal interval $i$\\
$\rho_{r \rightarrow r',\text{year},i}$ & first order auto-regressive correlation parameter across year for movement random effects from region $r$ to $r'$ in seasonal interval $i$\\
$\gamma_{r\rightarrow r',i}$ & random effect for link-transformed mean movement from region $r$ to $r'$ in seasonal interval $i$ when a prior distribution is assumed\\
$\sigma_{r\rightarrow r',i}$ & standard deviation parameter for prior distribution of $\gamma_{r\rightarrow r',i}$\\
$M_{y,a,r}$ & natural mortality rate for age $a$ in year $y$ in region $r$\\
$\theta_{M,r}$ & mean parameter across age and year for nutural mortality in region $r$\\
$\epsilon_{M,r,y,a}$ & random effect parameter for natural mortality in region $r$ and year $y$ at age $a$\\
$\beta_{M,r,a,k}$ & effect of environmental covariate $k$ on natural mortality in region $r$ at age $a$\\
$\sigma_{M,r}$ & standard deviation parameter for natural mortality random effects in region $r$\\
$\rho_{M,\text{age},r}$ & first order auto-regressive correlation parameter across age for natural mortality random effects in region $r$\\
$\rho_{M,\text{year},r}$ & first order auto-regressive correlation parameter across year for natural mortality random effects in region $r$\\
$\widehat{\mathbf{N}}_{H,s,y,a}$ & vector of predicted numbers of stock $s$ at age $a$ in year $y$ captured by each fleet\\
$\widehat{\mathbf{N}}_{H,y,a}$ & vector of predicted numbers at age $a$ in year $y$ captured by each fleet across all stocks\\
$\widehat{\mathbf{C}}_{y,a}$ & vector of predicted biomass captured at age $a$ in year $y$ by each fleet across all stocks\\
$\mathbf{c}_{y,a}$ & vector of mean individual weight at age $a$ in year $y$ for each fleet\\
$\widehat{\mathbf{C}}_y$ & vector of predicted aggregate catch for each fleet in year $y$\\
$C_{y,f}$ & observed aggregate catch for fleet $f$ in year $y$\\
$\widehat C_{y,f}$ & predicted aggregate catch for fleet $f$ in year $y$\\
$\sigma_{y,f}$ & standard deviation of observed log-aggregate catch for fleet $f$ in year $y$\\
$\widehat{N}_{s,y,a,m}$ & predicted abundance at $t_m$ in region $r_m$\\
$\mathbf{O}_{s,y,a,r_m}(t_m)$ & the probabilities of surviving and occurring in region $r_m$ at time $t_m$ given being alive in each region at the start of the year which is the $r_m$ column of the upper-left submatrix of Eq. \ref{eq:ptm_index}\\
$\widehat I_{m,y,a}$ & Predicted relative abundance index for survey $d$ in year $y$ at age $a$ \\
$q_{m,y}$ & catchability of index $m$ in year $y$\\
$\text{sel}_{m,y,a}$ &  selectivity of index $m$ at age $a$ in year $y$ \\
$w_{m,y,a}$ & average weight of individuals at age $a$ for index $m$ if the index is quantified in biomass, otherwise it is unity\\
$u_{m}$ & upper bound for index $m$ catchability\\
$l_{m}$ & lower bound for index $m$ catchability\\
$\theta_{q,m}$ & mean index $m$ catchability parameter\\
$\varepsilon_{q,m,y}$ & index $m$ catchability random effect in year $y$\\
$\beta_{q,m,k}$ & effect of environmental covariate $k$ on index $m$ catchability\\
$\sigma_{q,r}$ & standard deviation parameter for index $m$ catchability random effects\\
$\rho_{q,m}$ & first order auto-regressive correlation parameter across year for index $m$ catchability random effects\\
\end{longtable}

### 2024-09-20 R/confirm_RT_Run34.R

@@ -1,0 +1,239 @@
+#like run 30, but fit models with and without temperature effects on recruitment
+
+#multi-wham (12/12/2023)
+#library("wham", lib.loc = "c:/work/wham/old_packages/4a44134")
+#pkgload::load_all("c:/work/wham/wham")
+
+Run34 <- readRDS(file.path("c:/work/BSB.2023.RT.Modeling", "2023.RT.Runs","Run34", "fit.RDS"))
+
+#devel branch is now multi-wham
+library(wham)
+
+library(here)
+asap <- read_asap3_dat(here("data",c("north.dat","south.dat")))
+temp <- prepare_wham_input(asap)
+
+#basic_info
+basic_info <- list(region_names = c("North", "South"), stock_names = paste0("BSB_", c("North", "South"))) #, NAA_where = array(1, dim = c(2,2,6)))
+#11 seasons each 1 month long except a 2 month interval in the model (June,July)
+seasons = c(rep(1,5),2,rep(1,5))/12
+basic_info$fracyr_seasons <- seasons
+#each age other than 1 (recruitment) for north stock can be in either region on Jan 1 
+basic_info$NAA_where <- array(1, dim = c(2,2,8))
+basic_info$NAA_where[1,2,1] = 0 #stock 1, age 1 can't be in region 2 
+basic_info$NAA_where[2,1,] = 0 #stock 2, any age can't be in region 1 (stock 2 doesn't move) 
+#average recruitment over years 2000+ for SSB40 BRPs
+basic_info$XSPR_R_avg_yrs <- which(temp$years>1999)
+basic_info$XSPR_R_opt <- 2 #use average of recruitments (random effects), not expected/predicted given last time step
+#############################################
+
+#############################################
+#ecov
+north_bt <- read.csv(here("data","bsb_bt_temp_nmab_1959-2022.csv"))
+south_bt <- read.csv(here("data","bsb_bt_temp_smab_1959-2022.csv"))
+ecov <- list(label = c("North_BT","South_BT"))
+ecov$mean <- cbind(north_bt[,'mean'], south_bt[,'mean'])
+ecov$logsigma <- log(cbind(north_bt[,'se'], south_bt[,'se']))
+ecov$year <- north_bt[,'year']
+ecov$use_obs <- matrix(1, NROW(ecov$mean),NCOL(ecov$mean))
+#ecov$lag <- 1
[…]

### 2024-09-20 R/fit_run.R

@@ -1,0 +1,435 @@
+#like run 30, but fit models with and without temperature effects on recruitment
+
+#multi-wham (12/12/2023)
+#library("wham", lib.loc = "c:/work/wham/old_packages/4a44134")
+
+Run34 <- readRDS(file.path("c:/work/BSB.2023.RT.Modeling", "2023.RT.Runs","Run34", "fit.RDS"))
+
+#devel branch is now multi-wham
+library(wham)
+
+library(here)
+asap <- read_asap3_dat(c("north.dat","south.dat"))
+
+#adjust input Neff for D-M
+change_max_Neff_fn <- function(asap, max_Neff = 1000){
+	for(i in 1:length(asap)) {
+		asap[[i]]$dat$catch_Neff[] <- max_Neff
+		asap[[i]]$dat$IAA_mats <- lapply(asap[[i]]$dat$IAA_mats, function(x) {
+			out <- x
+			out[which(out[,NCOL(out)]>0),NCOL(out)] <- max_Neff
+			return(out)
+		})
+	}
+	return(asap)
+}
+asap_dm <- change_max_Neff_fn(asap, 1000)
+
+# north_bt <- read.csv(here("2023.RT.Runs","Run33","bsb_bt_temp-nmab.csv"))
+# south_bt <- read.csv(here("2023.RT.Runs","Run33","bsb_bt_temp-smab.csv"))
+north_bt <- read.csv("bsb_bt_temp_nmab_1959-2022.csv")
+south_bt <- read.csv("bsb_bt_temp_smab_1959-2022.csv")
+
+ecov <- list(label = c("North_BT","South_BT"))
+ecov$mean <- cbind(north_bt[,'mean'], south_bt[,'mean'])
+ecov$logsigma <- log(cbind(north_bt[,'se'], south_bt[,'se']))
+ecov$year <- north_bt[,'year']
+ecov$use_obs <- matrix(1, NROW(ecov$mean),NCOL(ecov$mean))
+#ecov$lag <- 1
+ecov$process_model <- "ar1"
+ecov$process_mean_vals <- apply(ecov$mean, 2, mean)
+ecov$recruitment_how <- matrix(c("controlling-lag-0-linear","none","none","none"), 2,2)
+
+NAA_re = list(sigma = list("rec+1","rec+1"), cor = list("2dar1","2dar1"), N1_model = rep("equilibrium",2))
+#Wasn't decoupled in RT
+NAA_re$decouple_recruitment <- FALSE
+
[…]

### 2024-09-20 R/fits.R

@@ -1,0 +1,253 @@
+#like run 30, but fit models with and without temperature effects on recruitment
+
+#multi-wham (12/12/2023)
+#library("wham", lib.loc = "c:/work/wham/old_packages/4a44134")
+#pkgbuild::compile_dll("c:/work/wham/wham", debug = FALSE)
+#pkgload::load_all("c:/work/wham/wham")
+
+Run34 <- readRDS(file.path("c:/work/BSB.2023.RT.Modeling", "2023.RT.Runs","Run34", "fit.RDS"))
+
+#devel branch is now multi-wham
+library(wham)
+
+library(here)
+
+source(here("R","make_wham_inputs.R")
+
+# pkgload::load_all("c:/work/wham/wham")
+input_1 <- prepare_wham_input(asap, selectivity = sel, NAA_re = NAA_re, basic_info = basic_info, move = move, ecov = ecov_1, catch_info = catch_info,
+	index_info = index_info, age_comp = age_comp)
+input_1$fleet_names <- paste0(rep(c("North_", "South_"),each = 2), temp$fleet_names)
+input_1$index_names <- paste0(rep(c("North_", "South_"),c(2,2)), temp$index_names)
+
+nms <- sort(unique(names(Run34$opt$par)))
+temp <- input_1
+temp$par[nms] <- Run34$parList[nms]
+temp$par$logit_selpars <- input_1$par$logit_selpars
+temp$par$sel_repars <- input_1$par$sel_repars
+temp$par$logit_selpars <- Run34$parList$logit_selpars[-(5:8),]
+temp$par$sel_repars <- Run34$parList$sel_repars[-(5:8),]
+temp$par$sel_repars[,2:3] <- temp$par$sel_repars[,2:3]*2 #previous version had yet to change transformation for selectivity correlation parameters
+x <- temp$par$Ecov_process_pars
+x[2,] <- x[2,] + 0.5*log(1 - (-1 + 2/(1+exp(-x[3,])))^2) # previous version estimated marginal sig, instead of conditional
+temp$par$Ecov_process_pars <- x 
+
+
+input_1 <- temp
+# fit <- readRDS(here("results", "Run34_update_fit.RDS"))
+
[…]

### 2024-09-20 R/make_tables.R

@@ -1,0 +1,87 @@
+
+mod_names <- paste0("$M_{", 0:13,"}$")
+col_names <- c("Model", "Temperature effect (north)", "Temperature effect (south)", "$M$ at age 1 random effects")
+
+mod_table <- matrix(nrow = length(mod_names), ncol = length(col_names), dimnames = list(mod_names,col_names))
+
+mod_table[,1] <- mod_names
+mod_table[,2:4] <- "--"
+mod_table[,4] <- rep(c("none", "time-varying"), each = 7)
+mod_table[c(2,4),2] <- "Recruitment"
+mod_table[c(3,4),3] <- "Recruitment"
+mod_table[c(5,7),2] <- "$M$ at age 1"
+mod_table[c(6,7),3] <- "$M$ at age 1"
+mod_table[8:14,2:3] <- mod_table[1:7,2:3]
+
+
+fits <- readRDS(file.path("results","fits_no_M_re.RDS"))
+fits_M_re <- readRDS(file.path("results","fits_M_re_better.RDS"))
+
+# AIC
+
+aic1 <- sapply(fits, function(x) {
+	2*c(x$opt$obj + length(x$opt$par), sapply(x$peels, function(y) y$opt$obj + length(y$opt$par)))
+})
+aic2 <- sapply(fits_M_re, function(x) {
+	2*c(x$opt$obj + length(x$opt$par), sapply(x$peels, function(y) y$opt$obj + length(y$opt$par)))
+})
+
+diff_aic <- apply(cbind(aic1,aic2),1, function(x) x - min(x))
+aic_wts <- apply(aic,2, function(x) exp(-x/2)/sum(exp(-x/2)))
+rownames(diff_aic) <- rownames(aic_wts) <- mod_table[,1]
+colnames(diff_aic) <- colnames(aic_wts) <- paste0("Peel ", 0:7)
+saveRDS(diff_aic, file.path("results","diff_aic.RDS"))
+saveRDS(aic_wts, file.path("results","aic_wts.RDS"))
+
+
+rho_ssb <- rbind(t(sapply(fits, function(x) mohns_rho(x)$SSB)),
+ t(sapply(fits_M_re, function(x) mohns_rho(x)$SSB)))
+rho_Fbar <- rbind(t(sapply(fits, function(x) mohns_rho(x)$Fbar)),
+ t(sapply(fits_M_re, function(x) mohns_rho(x)$Fbar)))
[…]

### 2024-09-20 R/make_wham_inputs.R

@@ -1,0 +1,190 @@
+asap <- read_asap3_dat(here("data",c("north.dat","south.dat")))
+temp <- prepare_wham_input(asap)
+
+#basic_info
+basic_info <- list(region_names = c("North", "South"), stock_names = paste0("BSB_", c("North", "South"))) #, NAA_where = array(1, dim = c(2,2,6)))
+#11 seasons each 1 month long except a 2 month interval in the model (June,July)
+seasons = c(rep(1,5),2,rep(1,5))/12
+basic_info$fracyr_seasons <- seasons
+#each age other than 1 (recruitment) for north stock can be in either region on Jan 1 
+basic_info$NAA_where <- array(1, dim = c(2,2,8))
+basic_info$NAA_where[1,2,1] = 0 #stock 1, age 1 can't be in region 2 
+basic_info$NAA_where[2,1,] = 0 #stock 2, any age can't be in region 1 (stock 2 doesn't move) 
+#average recruitment over years 2000+ for SSB40 BRPs
+basic_info$XSPR_R_avg_yrs <- which(temp$years>1999)
+basic_info$XSPR_R_opt <- 2 #use average of recruitments (random effects), not expected/predicted given last time step
+#############################################
+
+#############################################
+#ecov
+north_bt <- read.csv(here("data","bsb_bt_temp_nmab_1959-2022.csv"))
+south_bt <- read.csv(here("data","bsb_bt_temp_smab_1959-2022.csv"))
+ecov <- list(label = c("North_BT","South_BT"))
+ecov$mean <- cbind(north_bt[,'mean'], south_bt[,'mean'])
+ecov$mean <- t(t(ecov$mean) - apply(ecov$mean,2,mean))
+ecov$logsigma <- log(cbind(north_bt[,'se'], south_bt[,'se']))
+ecov$year <- north_bt[,'year']
+ecov$use_obs <- matrix(1, NROW(ecov$mean),NCOL(ecov$mean))
+ecov$process_model <- "ar1"
+ecov$process_mean_vals <- apply(ecov$mean, 2, mean)
+
+ecov_0 <- ecov 
+ecov_0$recruitment_how <- matrix(c("none","none","none","none"), 2,2)
+ecov_1 <- ecov_2 <- ecov_0
+ecov_1$recruitment_how[1,1] <- "controlling-lag-0-linear" #north
+ecov_2$recruitment_how[2,2] <- "controlling-lag-0-linear" #south
[…]

### 2024-09-20 R/try_BH_SR.R

@@ -1,0 +1,56 @@
+fits_M_re <- readRDS(here("results", paste0("fits_M_re_better.RDS")))
+
+ny <- fits_M_re[[1]]$input$data$n_years_model
+plot(x[,1],x[,2])
+plot(x[,1],x[,2]/x[,1])
+
+x <- cbind(SSB = fits_M_re[[1]]$rep$SSB[-ny,2],R = fits_M_re[[1]]$rep$NAA[2,2,-1,1])
+plot(x[,1],x[,2])
+plot(x[,1],x[,2]/x[,1])
+plot(x[,1],log(x[,2]/x[,1]))
+
+x <- cbind(SSB = fits_M_re[[1]]$rep$SSB[-ny,1],R = fits_M_re[[1]]$rep$NAA[1,1,-1,1])
+fn <- function(par){
+	mu <- par[1] + log(x[,1]) - log(1 + exp(par[2])*x[,1])
+	return(-sum(dnorm(log(x[,2]), mu, sd = exp(par[3]), log = TRUE)))
+}
+phi0 <- mean(exp(fits_M_re[[1]]$rep$log_SPR0[,1]))
+opt <- nlminb(c(0,-5,0), fn)
+
+h <- exp(opt$par[1])*phi0/(4 + exp(opt$par[1])*phi0)
+
+x <- cbind(SSB = fits_M_re[[1]]$rep$SSB[-ny,2],R = fits_M_re[[1]]$rep$NAA[2,2,-1,1])
+phi0 <- c(phi0, mean(exp(fits_M_re[[1]]$rep$log_SPR0[,2])))
+
+fn <- function(par){
+	mu <- par[1] + log(x[,1]) - log(1 + exp(-6)*x[,1])
+	return(-sum(dnorm(log(x[,2]), mu, sd = exp(par[2]), log = TRUE)))
+}
+opt <- nlminb(c(0,0), fn)
+
+fn <- function(par){
+	h <- 0.2 + 0.8/(1 + exp(-par[1]))
+	R0 <- exp(par[2])
+	mu <- log(4 * R0 * h * x[,1]/((1-h)*R0*phi0[2] + (5*h - 1)* x[,1]))
+	return(-sum(dnorm(log(x[,2]), mu, sd = exp(par[3]), log = TRUE)))
+}
+opt <- nlminb(c(0,0,0), fn)
+
+h <- c(h,exp(opt$par[1])*phi0/(4 + exp(opt$par[1])*phi0))
+
+
[…]

### 2024-09-22 R/fits.R
add mles for no M re models

@@ -90,6 +90,8 @@ fits <- lapply(fits, do_reference_points)
 fits <- lapply(fits, do_sdreport)
 fits <- lapply(fits, make_osa_residuals)
 saveRDS(fits, here("results", paste0("fits_no_M_re.RDS")))
+parLists_fits <- sapply(fits, function(x) x$parList)
+saveRDS(parLists_fits, here::here("results", paste0("parLists_no_M_re.RDS")))
 
 input_M_re_0 <- prepare_wham_input(asap, selectivity = sel, NAA_re = NAA_re, basic_info = basic_info, move = move, ecov = ecov, catch_info = catch_info,
 	index_info = index_info, age_comp = age_comp, M = M_list)

### 2024-09-22 R/fits.R
change to script

@@ -90,7 +90,7 @@ fits <- lapply(fits, do_reference_points)
 fits <- lapply(fits, do_sdreport)
 fits <- lapply(fits, make_osa_residuals)
 saveRDS(fits, here("results", paste0("fits_no_M_re.RDS")))
-parLists_fits <- sapply(fits, function(x) x$parList)
+parLists_fits <- lapply(fits, function(x) x$parList)
 saveRDS(parLists_fits, here::here("results", paste0("parLists_no_M_re.RDS")))
 
 input_M_re_0 <- prepare_wham_input(asap, selectivity = sel, NAA_re = NAA_re, basic_info = basic_info, move = move, ecov = ecov, catch_info = catch_info,

### 2024-11-19 paper/paper.Rmd
saving work

@@
 ---
-title: "A multi-stock, multi-region state-space age-structured assessment model with an application to black sea bass"
+title: "Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass"
 author: 
@@
 
-A multi-region, multi-stock generalization of WHAM with an application evaluating evidence of alternative hypotheses about temperature effects on black sea bass
+Describes the multi-region, multi-stock generalization of WHAM and its usage to evaluate evidence of alternative hypotheses about temperature effects on recruitment of black sea bass stock components.
 
@@
 We first describe a multi-stock, multi-region extension of WHAM that treats the population and fleet dynamics seasonally and allows movement by season and region to be functions of time- and age-varying autocorrelated random effects and environmental covariates.
-We then illustrate the model with northern and southern components of the NEUS black sea bass stock and evaluate alternative hypotheses of bottom temperature and random effects on recruitment and natural mortality.
-We show strong evidence for temperature effects on recruitment and no random effects or temperature effects on natural mortality, primarily for the northern stock component.
+We then illustrate the model by applying it to data for the northern and southern components of the NEUS black sea bass stock and evaluate alternative hypotheses of bottom temperature and random effects on recruitment and natural mortality.
+We show strong evidence for temperature effects on recruitment, primarily for the northern stock component, and no evidence for including random effects or temperature effects on age 1 natural mortality.
 
@@
 
-Why is this model needed.  Models ignoring movement can provide biased catch advice.  Using random effects for time-varying processes is considered  best practice.
+A state-space statistical approach and maximum marginal likelihood or Bayesian fitting of stock assessment models allows estimation of time- and age-varying population attributes as random effects [@nielsenberg14,@cadigan16,@milleretal16].
+This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks [@puntetal20].
+The State-space Assessment Model [SAM, @nielsenberg14] continues to be developed and remains widely used within ICES to assess European fish stocks.
+Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada [@varkeyetal22;@perreaultetal20] and to the south, the Woods Hole Assessment Model [WHAM, @stockmiller21] is now used to assess many fish stocks in the Northwest Atlantic Ocean [@nefsc22;@nefsc22a;@nefsc24].
 
-The Woods Hole Assessment Model (WHAM) is an R package developed at the Northeast Fisheries Science Center [https://timjmiller.github.io/wham, @millerstock20; @stockmiller21].
-Primary inputs and observation types are similar to ASAP and there is functionality built into WHAM to move information from ASAP input files to R.
-WHAM can be configured to fit SCAA models often identically to ASAP, so that existing assessments in the U.S.
-Northeast can be replicated and tested against state-space models with process errors and environmental effects in a single framework.
-WHAM models are built using the Template Model Builder package [TMB, @kristensenetal16] which provides efficient fitting of models that include random effects similar to the State-space Assessment Model [SAM, @nielsenberg14], which is used to manage many stocks within ICES.
-WHAM is currently used to manage 9 stocks in the NEUS.
+WHAM is an R package developed and maintained at NOAA's Northeast Fisheries Science Center [https://timjmiller.github.io/wham, @millerstock20; @stockmiller21].
+WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and possibly effects of environmental covariates on various demographic parameters.
+Like SAM, WHAM models are built using the Template Model Builder package [TMB, @kristensenetal16] which provides a computationally efficient means of fitting an extremely wide class of models with random effects.
+WHAM has undergone active development since its creation and includes random effects options for in recruitment, inter-annual transitions numbers at age (hereafter referred to as "survival"), fishery and index selectivity, natural mortality, and catchability.
 
-The Woods Hole Assessment Model (WHAM) software package is maintained and developed at the Northeast Fisheries Science Center to enable state-space stock assessments, i.e.
-where processes such as the annual transitions in numbers-at-age (survival), natural mortality, selectivity, and catchability are treated as time- and(or) age-varying random effects.
-WHAM can also be configured without random effects as a traditional statistical catch-at-age model in order to bridge from current assessments which use Age-Structured Assessment Program (ASAP).
-Here, we present an extension of WHAM (Multi-WHAM) to model multiple stocks and(or) regions and allow movement of stocks among regions.
+However, WHAM has up to now only allowed models with one stock (component) and a single region and without seasonal changes in stock dynamics.
+Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice [@bosleyetal22;@yingetal11;@caoetal14].
+Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks [@puntetal20].
+We describe here the implementation of these features and other extensions since @stockmiller21 in WHAM version 2.0.
+Many of these new configuration options can be useful whether modeling 1 or more stocks and regions.
 
-It has been reviewed in the literature and simulation tested [@stocketal21; @stockmiller21].
-It has also been used as the operating model in the index-based assessment methods research track assessment [@legaultetal23] and is also the main tool for simulation studies in the research track on state-space assessment models.
-
-A review of the essential features for next-generation stock assessments concluded that only WHAM and a similar State-Space Assessment Method (Nielsen and Berg 2014) model random effects correctly (Punt et al.
-2020).
-WHAM is now being used as the management model for 9 stocks in the northeast U.S., but the standard version of WHAM can only be applied to a single stock and area (Miller and Stock 2020).
-
-Here, we describe a multi-stock and multi-region extension of the WHAM package (version 2.0 https://github.com/timjmiller/wham/tree/lab).
-Numerous new configuration options are also included in this extension that can be useful even for single-stock or single region models.
-Any number of stocks or regions can be configured in this version of WHAM.
-The multi-stock extension was developed to accommodate movement of stocks or stock components in anticipation of black sea bass and Atlantic cod stocks where new modeling frameworks where recently examined through Northeast research track assessment process (cite RT docs).
-We demonstrate capabilities of the new version of WHAM with black sea bass and evaluate evidence for alternative hypotheses of temporal variation in natural mortality and effects of bottom temperature on recruitment and natural mortality of age 1 individuals.
+This extension of WHAM was developed in concert with a stock assessment for black sea bass through the NEFSC research track assessment process where new modeling frameworks were recently examined and modeling multiple stocks or stock components simultaneously was of interest [@nefsc23].
+\textcolor{red}{[Text on black sea bass here]} We apply WHAM 2.0 to two stock components of black sea bass off the coast of the NEUS and evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
 
@@
 
-## Model description {-}
+## WHAM description {-}
 
-Many of the options and equations of version 2.0 are are the same as those described by @stockmiller21, so we will only describe extensions and differences that have occurred since the first description of WHAM.
-The new version of WHAM can model multiple stocks, but survival, movement, harvest and natural mortality is are tracked for each stock.
-Therefore, much of the description below assumes a specific stock $s$, but this subscript is only explicitly used when necessary.
+Many of the options and equations of WHAM version 2.0 are the same as those in @stockmiller21, so we will only describe extensions and differences that have occurred since their first description of WHAM.
+The new version of WHAM can model multiple stocks and survival, movement, harvest and natural mortality are tracked for each stock.
+Much of the description below is for a specific stock $s$, but, for simplicity, this subscript is implicit except when necessary.
 
@@
 
-Because individuals may be alive in one of several regions or harvested in one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model (reference).
-Multi-state models use a probability transition matrix (PTM) that describes the probability of individuals living and moving among regions or dying due to fishing or natural mortality over a time interval $i$ with duration $\delta_i$ in year $y$ for individuals at age $a$ on January 1.
+Because individuals may be alive in one of several regions or harvested in one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model.
+Approaches to modeling transitions among these tates may treat time discretely [e.g.,@arnason72;@schwarzetal93] or continuously [e.g., @hearnetal87;@commenges99; @andersenkeiding02].
+Multi-state models can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval $\delta$, conditional on being in each of those state at beginning of the interval.
+For fish populations these states would be defined as being alive in a particular region or being dead due to fishing from a particular fleet or natural mortality.
+The time interval $i$ with duration $\delta_i$ would be a season and the PTMs would be uniquely defined for each stock by season $i$, year $y$, and age $a$ on January 1.
 Each row and column of the PTM correspond to one the states: alive in region $r$, dead in fleet $f$, or dead from natural causes.
@@
 is the $n_R \times n_F$ matrix defining probabilities of being captured in each fleet during the interval, and $\mathbf{D}_{y,a,i}$ is the $n_R x 1$ matrix of probabilities of dying due to natural mortality during the interval.
-We have the identity matrix $\mathbf{I}_{H}$ for the states for  capture by each fleet and a 1 for the state for natural mortality because the probabilities of being in one of the mortality states given starting the interval in that state is unity.
+We have the identity matrix $\mathbf{I}_{H}$ for the states for  capture by each fleet and a 1 for the state for natural mortality because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).
 
-WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would be multinomial distributed.
+WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would be multinomial distributed as in a model for tagging data where fates of individual fish are assumed independent.
 The PTMs determine the expected numbers 1) in each state on January 1 of year $t+1$ at age $a+1$ given the abundances at age $a$ on January 1 of year $t$, 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.
@@
 When there is autocorrelation with age, WHAM now assumes this applies only to ages $a>1$ by default so that recruitment random effects are independent of those for for the annual transitions of older age classes.
-So the general covariance structure for a given stock at ages $a>1$ in region $r$ is
+So the general covariance structure for a given stock at ages $a>1$ in region $r$ is that of a two-dimensional first-order autoregressive (2DAR1) process
 \begin{equation*}
-  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma^2_{N,r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}.
+  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}
 \end{equation*}
+and that for age 1 is just AR1 across years 
+\begin{equation*}
+  Cov\left(\epsilon_{y,1},\epsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
+\end{equation*}
+Since recruitment for a given stock currently only occurs in one region $r_s$ there is only a single time-varying recruitment random effect for each stock.
 
@@
 
-The initial abundances at age can also be treated as independent or autocorrelated (with age) random effects.
-Defining the vector of initial abundance at age in region $r$ as $\mathbf{N}_{O,1,r}$, the general model is
+The initial abundances at age can also be treated as independent or as AR1 random effects. Defining the vector of initial abundance at age in region $r$ as $\mathbf{N}_{O,1,r}$, the general model is 
 $$\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}$$
@@
 \end{equation*}
-The random effects $\epsilon_{r\rightarrow r',y,a,i}$ are season-, and(or) region-to-region-specific and modeled most generally as a two-dimensional first-order autoregressive random effects with age and(or) year where the covariance is
+The random effects $\epsilon_{r\rightarrow r',y,a,i}$ are season-, and(or) region-to-region-specific and modeled most generally as 2DAR1 random effects with age and(or) year where the covariance is
 \begin{equation*}
@@
 When not estimated, (mean) mortality rates may be stock-, region-, and age-specific.
-When random effects are used, the same autocorrelation structures with age and year as described by @stockmiller21 can be configured for a given stock and region.
+When random effects are used, the same 2DAR1 structure with age and year as described by @stockmiller21 can be configured for a given stock and region.
 Any environmental covariate effects can be stock-, region-, and age-specific.
@@
 \end{equation*}
-The general covariance structure for random effects are modeled most generally as a two-dimensional first-order autoregressive random effects with age and(or) year where the covariance is
+The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
 \begin{equation*}
@@
 $$log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}$$ 
-where $u_{m}$ and $l_{m}$ are the upper and lower bounds of catchability for index $m$ (defaults are 0 and 1000) and the general covariance structure for the annual random effects is first-order autoregressive
+where $u_{m}$ and $l_{m}$ are the upper and lower bounds of catchability for index $m$ (defaults are 0 and 1000) and the general covariance structure for the annual random effects is AR1
 $$Cov\left(\epsilon_{q,m,y},\epsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.$$
@@
 When there is movement of any stocks, the user has the option to project and use any random effects for time-varying movement or use the average over user specified years, analogous to how natural mortality can be treated in the projection period.
-The projections of any environmental covariates has been revised to better include error in the estimated latent covariate in any effects on the population in projection years.
+The projection of any environmental covariates has been revised to better include error in the estimated latent covariate in any effects on the population in projection years.
 
@@
 
-Prior to the most recent peer-reviewed assessment of black sea bass in the NEUS, the stock was assessed using the Age-Structured Assessment Program (ASAP) model [@legaultrestrepo99], a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
-Northern and southern components of the NEUS black sea bass stock divided by the Hudson Canyon were separately modeled in ASAP (reference map figure).
-Results from the separate ASAP models have been aggregated for a unit-stock assessment.
-Furthermore, the ASAP-based assessments have exhibited strong retrospective patterns [@mohn99], and exploring alternative modeling approaches for the northern and southern stock components has been a high priority for management.
+Prior to its 2023 peer-reviewed assessment, the NEUS black sea bass stock was assessed using the Age-Structured Assessment Program (ASAP) model [@legaultrestrepo99], a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
+Northern and southern components of the NEUS black sea bass stock ascribed to regions divided by the Hudson Canyon were separately modeled in ASAP (reference map figure).
+Results from the separate ASAP models were combined for a unit-stock assessment.
+The ASAP-based assessments exhibited strong retrospective patterns [@mohn99], and exploring alternative modeling approaches for the northern and southern stock components has been a high priority for management.
 
-In 2023, a working group composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above [@nefsc23].
-This assessment included the spatial features and hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
+Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as "working group") composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above [@nefsc23].
+This assessment included the spatial features and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
 
-Below we describe the assumptions and configuration of the assessment model as determined by the working group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of environmental effects.
+Below we describe the assumptions and configuration of the assessment model as determined by the working group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.
 
@@
 The first year being modeled for the population is 1989 and the fishery and index data used in the model span from 1989 to 2021.
-There are two stock components (north and south) modeled as separate populations that spawn and recruit in respective regions separated by the Hudson Canyon.
+The north and south stock components are modeled as separate populations that spawn and recruit in respective regions.
 We have observations for each of four total fishing fleets, where two fishing fleets (Recreational and Commercial) operate in each region.
@@
 There are 11 seasonal intervals within each calendar year: five monthly time intervals from Jan 1 to May 31, a spawning season from June 1 to July 31, and five monthly intervals from August 1 to December 31.
-The southern stock component is assumed to never move to the northern region (Figure \ref{fig:migration-diagram}).
-For the northern component, a proportion $\mu_1$ can move to the southern region each month during the last five months of the year, but no movement is allows from the south to the north during this period.
-During the first four intervals of the year a proportion $\mu_2$ the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
+The southern stock component is assumed to never move to the northern region.
+For the northern component, a proportion $\mu_{\text{N}\rightarrow \text{S}}$ can move to the southern region each month during the last five months of the year, but no movement is allows from the south to the north during this period (Figure \ref{fig:migration-diagram}).
+During the first four intervals of the year a proportion $\mu_{\text{S}\rightarrow \text{N}}$ the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
 In the fifth interval (May), all northern component individuals remaining in the south are assumed to move back to the north for the subsequent spawning period.
@@
 
-The monthly movement matrices are
+The two monthly movement matrices are
 \begin{equation*}
@@
   \begin{bmatrix}
-     1-\mu_{1\rightarrow 2} & \mu_{1\rightarrow 2} \\
+     1-\mu_{\text{N}\rightarrow \text{S}} & \mu_{\text{N}\rightarrow \text{S}} \\
      0 & 1 \\
@@
      1 &  0 \\
-     \mu_{2\rightarrow 1} & 1-\mu_{2\rightarrow 1} \\
+     \mu_{\text{S}\rightarrow \text{N}} & 1-\mu_{\text{S}\rightarrow \text{N}} \\
   \end{bmatrix}
@@
 for the portion of the year before spawning.
-
-The constant or mean log-natural mortality rate is assumed to be $\log(0.4)$
-
 As noted in the description of the general WHAM model, tagging data are not yet allowed.
@@
 
-## Initial abundance at age {-}
-
 With the movement configuration, the northern origin fish (ages 2+) can occur in the southern region on January 1.
-Estimating initial numbers at age as separate parameters can be challenging even in single-stock models.
-To avoid challenges with estimating initial numbers at age in each region for the northern component in the two stock model, we used the equilibrium assumption described above.
-Using this assumption two parameters are estimated for each stock component: an initial recruitment and an equilibrium fully-selected $F$.
-
-## Recruitment and Survival {-}
+Estimating initial numbers at age as separate parameters can be challenging even in single-stock models, but for black sea bass the available data cannot distinguish the proportion of northern and southern component fish at each age in the southern region in the initial year of the model.
+Therefore, we used the simplifying equilibrium assumption described above where there are two parameters estimated for each stock component: an initial recruitment and an equilibrium fully-selected $F$ that determines the abundance at age in each region for each stock component.
 
@@
 All of the initial runs assumed variance parameters for these random effects to be the same for northern origin fish occurring in both regions on January 1.
-The base model assumes very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model.
+The base model assumes very small variance for the transitions of northern fish in the southern region, which is approximately the same as the deterministic transition assumptions of a statistical catch at age model.
 We also allow 2DAR1 correlation for recruitment and survival for both the northern and southern components.
@@
 
-The CVs provided by the analyses for the recreational catch per angler (RecCPA) ranged between 0.02 and 0.06 which the working group felt did not capture the true uncertainty in the index with regard to its relationship to stock abundance.
+The estimated coefficients of variation (CVs) provided by the analyses to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the working group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
 In many of the initial runs as well as the base model we allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
@@
 
-Table \ref{tab:age-comp-sel-table}
+The working group investigated many alternative assumptions for the probability models and selectivity models for the 8 different sets of age composition observations to reduce residual patterns and retrospective patterns.
+These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for 6 sets of of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
 
@@
 All model fits also include bottom temperature observations for the northern and southern regions from 1963 to 2021 and estimated standard errors ranging between 0.03 and 0.09 degrees Celsius [@nefsc23].
-We retained the assumption from the peer-reviewed assessment that treated the latent bottom temperature covariates in each region as first-order autoregressive processes.
+We retained the assumption from the peer-reviewed assessment that treated the latent bottom temperature covariates in each region as AR1 processes.
 
@@
 
-It is standard practice to treat annual recruitment as time-varying deviations from a mean model and all models here treat recruitment deviations as first-order autoregressive random effects.
-However, @milleretal18 showed how inferences of temperature effects on growth or maturity parameters can be very different whether the null model without the effect and models with effects also include random effects representing residual temporal variation in parameters.
-Therefore, we also explored whether including temporal random effects on age 1 natural mortality M affected inferences on corresponding temperature effects.
+It is standard practice to treat annual recruitment as time-varying deviations from a mean model and all models here treat recruitment deviations as AR1 random effects.
+However, @milleretal18 showed how inferences of temperature effects on growth or maturity parameters can be very different whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
+Therefore, we also explored whether including temporal random effects on age 1 natural mortality affected inferences on corresponding temperature effects.
 Initially, we included random effects on age 1 natural mortality for both the northern and southern stock components, but estimates of these random effects and corresponding variance for the southern component converged to 0 so these were not included in models presented here.
@@
 The fish are technically already 1 year old, but there are no observations of these individuals until later in the year except possibly in fishery catches which are accumulated over the whole year.
-Expected log-recruitment is a linear function of log-recruitment at the previous time step and bottom temperature
+Expected log-recruitment for a given stock is a linear function of bottom temperature
 
 \begin{equation}\label{eq:expected-recruitment}
-\log N_{s,r_s,y,1} = \mu_{R,s} + \beta_{R,s} x_y + \epsilon_{N,s,y,1}
+E\left(\log N_{y,1}|x_y\right) = \mu_{R} + \beta_{R} x_y.
 \end{equation}
+Similarly, expected log-natural morality as a function of bottom temperature is
+\begin{equation}\label{eq:expected-M1}
+E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y
+\end{equation}
+Because age 1 fish for the northern component can exist in both regions after January 1, natural mortality is acting in both regions for this stock.
+For models with covariate and/or annual random effects for age 1 fish we assume them only in the northern regions for the northern component.
+The corresponding random effects are
+\begin{equation}\label{eq:Rec-re}
+\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \epsilon_{y,1}
+\end{equation}
+and
+\begin{equation}\label{eq:M-re}
+\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \epsilon_{M,y}.
+\end{equation}
+The constant or mean log-natural mortality rate is assumed to be $\mu_{M,a} = \log(0.4)$ for all ages as recommended by the working group.
+Because the bottom temperature anomalies and random effects are centered at 0, mean log-natural mortality at age 1 over the time series should be approximately equal to $\mu_{M,1}$ for models that include those effects.
 
-\begin{equation}\label{eq:expected-M1}
-\log M_{s,y,1} = \mu_{M,s,1} + \beta_{M,s} x_y + \epsilon_{M,s,y}
-\end{equation}
+## Model fitting, diagnostics, and projections {-}
 
-## Model fitting, diagnostics, and comparisons {-}
+We used a development version ([commit fb8b089](https://github.com/timjmiller/wham/tree/fb8b089)) of the WHAM package prior to the release of version 2.0 for all results.
+All code for fitting models and generating results can be found at [github.com/timjmiller/wham_bsb_paper](https://github.com/timjmiller/wham_bsb_paper).
 
-As in @milleretal16, we also assessed the stability of the AIC-based model selection over retrospective peels to guard against previously noted changes in perception of covariate effects on recruitment with increased length of the time series of observations [@myers98].
+We examined retrospective patterns for all models by fitting corresponding models where the terminal year is reduced sequentially by one year (peel) for seven years.
+Therefore, there are eight fits of each model with the time series reduced by zero to seven years.
+We  calculated Mohn's $\rho$ for SSB, and average $F$ at ages 6 and 7.
+Absolute values of Mohn's $\rho$ near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended.
+
+As in @milleretal16, we also assessed the consistency of the AIC-based model selection over retrospective peels to guard against previously noted changes in perception of covariate effects on recruitment with increased length of the time series of observations [@myers98].
 This retrospective examination was also recommended by @brooks24.
 
-We examined retrospective patterns by fitting models where the terminal year is reduced sequentially by one year (peel) for seven years.
-Therefore, there are 7 fits of the proposed base model with the time series reduced by one to seven years.
-We  calculated Mohn's $\rho$ for recruitment, SSB, and fully-selected $F$.
-Absolute values of Mohn's $\rho$ near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended.
+We performed a jitter analysis of the base model $M_0$ and the best fitting model to investigate whether a local mimimum of the negative log-likelihood surface was obtained by the optimization.
+We used the jitter_wham function in the WHAM package which by default simulates starting values from a multivariate normal distribution with mean and covariance defined by the MLEs and estimated covariance matrix from the fitted model.
+See the Supplementary Materials for more details.
 
-Specifically, we use version XX and commit XXX of the wham package (https://github.com/timjmiller/wham/tree/lab) for all results.
+For the best performing model, we also performed so-called simulation self-tests where new observations were simulated conditional on all estimated fixed and random effects and the same model configuration was fit to each of the simulated data sets.
+We estimated median relative bias of SSB across these simulations.
+See the Supplementary Materials for more details.
 
-We performed a simulation self-test where new observations were simulated conditional on all random effects estimated in the proposed base model and the same model configuration was fit to each of the simulated data sets.
-
-One-step-ahead residuals can now be calculated for all index, catch, environmental covariate observations using methods described by @thygesenetal17 and @trijouletetal23.
-
-### One-step-ahead residuals {-}
-
-<!--
-
-We used one-step-ahead (OSA) residuals to evaluate mis-specification of the proposed base model.
-Pearson residuals do not have the appropriate properties (independent standard normal for a correctly specified model) for composition observations [@trijouletetal23] or for covariate and aggregate index and catch observations when temporal random effects are employed  because of the lack of independence of observations [@thygesenetal17].
-
-There is no evidence of mis-specificaiton from the OSA residuals for aggregate catch of the northern commercial fleet (Figure \ref{fig:osa-North-comm-catch-summ}).
-There was some indication of trends in residuals of some of the age composition observations early in the time series and for the the first age class (Figures \ref{fig:osa-North-comm-paa-summ} and \ref{fig:osa-North-comm-paa}).
-
-For the northern recreational fleet, the OSA residuals for the aggregate catch appeared satisfactory (Figure \ref{fig:osa-North-rec-catch-summ}).
-The OSA residuals for the age composition observations for the northern recreational fleet showed some tendency of underdispersion but were otherwise unnoteworthy (Figures \ref{fig:osa-North-rec-paa-summ} and \ref{fig:osa-North-rec-paa}).
-
-Neither the OSA residuals for the aggregate catch or the age composition of the Northern recreational CPA index exhibited signs of mis-specification (Figures \ref{fig:osa-North-reccpa-catch-summ} to \ref{fig:osa-North-reccpa-paa}).
-
-There was a little evidence of tendency toward negative OSA residuals for the Northern VAST index whereas there was some indication of positive residuals for the age composition observations, particularly at age 1 (Figures \ref{fig:osa-North-vast-catch-summ} to \ref{fig:osa-North-vast-paa})
-
-THe OSA residuals for the aggregate catch of the southern commercial fleet tended to be negative, but the residuals for the age composition observations did not show any evidence of mis-specification (Figures \ref{fig:osa-South-comm-catch-summ} to \ref{fig:osa-South-comm-paa}).
-
-The OSA residuals for the aggregate catch of the southern recreational fleet were under-dispersed and those for the age composition were somewhat underdispersed, but no trends in either types of observations were apparent (Figures \ref{fig:osa-South-rec-catch-summ} to \ref{fig:osa-South-rec-paa})
-
-The OSA residuals for the aggregate catch of the southern recreational CPA index were also somewhat under-dispersed (Figures \ref{fig:osa-South-reccpa-catch-summ}).
-The residuals for the age composition observations exhibited a few large negative residuals at older ages, but otherwise exhibited no apparent trends (Figures \ref{fig:osa-South-reccpa-paa-summ} to \ref{fig:osa-South-reccpa-paa})
-
-The OSA residuals for the aggregate southern VAST index were somewhat over-dispersed and the residuals for the age composition appeared to show some trend with observed proportion and age (Figures \ref{fig:osa-South-vast-catch-summ} to \ref{fig:osa-South-vast-paa}).
--->
-
-### Retrospective patterns {-}
-
-<!-- 
-We also examined retrospective patterns by fitting models where the terminal year is reduced sequentially by one year (peel) for seven years.
-Therefore, there are 7 fits of the proposed base model with the time series reduced by one to seven years.
-Multi-WHAM performs these fits when specified and will calculate Mohn's $\rho$ for recruitment, SSB, and fully-selected $F$.
-Absolute values of Mohn's $\rho$ near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended.
-Although there was strong retrospective patterns in the most recent management track assessment for the northern component of the stock, there was no evidence of patterns for northern SSB (Mohn's $\rho$ = -0.058) and $F$ (Mohn's $\rho$ = 0.06) for the proposed base model (Figures \ref{fig:North-retro-ssb} and \ref{fig:North-retro-F}).
-Similarly, no patterns were exhibited for the southern component of SSB (Mohn's $\rho$ = -0.019) and regional $F$  (Mohn's $\rho$ = -0.05) (Figures \ref{fig:South-retro-ssb} and \ref{fig:South-retro-F}).
--->
+To illustrate projections with environmental effects we projected the best black sea bass model under three alternative scenarios where the AR1 time-series model for bottom temperature continues into the projection years, where we project with bottom temperature being the average of the most recent 5 years, and where bottom temperature increases following a predictions from a simple linear regression of the estimated bottom temperature anomalies over time.
 
@@
 
-Although there was strong retrospective patterns in the most recent management track assessment for the northern component of the stock, there was no evidence of patterns for northern SSB (Mohn's $\rho$ = -0.058) and $F$ (Mohn's $\rho$ = 0.06) for the proposed base model (Figures \ref{fig:North-retro-ssb} and \ref{fig:North-retro-F}).
-Similarly, no patterns were exhibited for the southern component of SSB (Mohn's $\rho$ = -0.019) and regional $F$  (Mohn's $\rho$ = -0.05) (Figures \ref{fig:South-retro-ssb} and \ref{fig:South-retro-F}).
+We found the best model of bottom temperature covariate effects included the effect only on the recruitment of the northern stock (Table \ref{tab:diff-aic-table}).
+However, the difference in AIC for the model that also included effects on recruitment for the southern component suggested some evidence for this hypothesis as well (Table \ref{tab:aic-wts-table}).
 
-For 7 of the the simulated data sets the model failed to optimize.
-The maximum absolute gradient was $<10^{-6}$ for only 9 and $<10^{-4}$ for 52 of the 93 successfully fitted models.
-The poor convergence appeared to be attributable to the estimation of the scalar for the standard errors of the log-transformed northern Recreational CPA index for which estimates tended to 0 for nearly all of the fits ($<0.01$ for 83 fits).
-However, even across all fits including those with poor convergence, the SSB estimates appeared to be reliable (Figure \ref{fig:self-test-res-1}).
+Across retrospective peels where the terminal year of data was sequentially reduced, the ranking for the best model was consistent.
+However, when the terminal year was reduced 6 or more years, there was more evidence for further effects of temperature on recruitment for the southern component and for temporal variation in natural mortality at age 1 (Table \ref{tab:aic-wts-table}).
 
-## Bottom temperature effect on recruitment {-}
+The retrospective model fits did not indicate any evidence of patterns for northern or southern component SSB (all Mohn's $\rho \approx -0.03$) or fishing mortality (all Mohn's $\rho \approx 0.03$ and $-0.04$ for average $F$ at ages 6 and 7 in north and south regions, respectively) for any of the models (Table \ref{tab:rho-table}).
 
-We found the best model of bottom temperature covariate effects included the effect only on the recruitment of the northern stock, which is the configuration in the proposed base model although the difference in AIC for the model including effects on both regional components was small suggesting some evidence for this hypothesis as well (Table \ref{tab:compare-table}) .
-We also found AIC to strongly favor models that included at least an effect on the northern component across all retrospective peels (Table \ref{tab:peel-compare-table}).
-The posterior estimate of the bottom temperature covariate match the observations well because of the high precision of the observations (Figure \ref{fig:bottom-temp}).
+## Bottom temperature effects {-}
+
+The posterior estimate of the bottom temperature covariate match the observations well because of the high precision of the observations and anomalies in the north and south region appear highly correlated (Figure \ref{fig:bottom-temperature}).
+Because the temperature anomalies are treated as latent variables, when effects on recruitment are included, other data components in the model can affect the estimated anomalies [e.g., @milleretal18], but in this case the estimates are altered negligibly during the years where recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
+
+Estimated effects of bottom temperature on recruitment for either the northern or southern component were stable over the retrospective peels and differed negligibly whether effects for each component were estimated in isolation or together (Table \ref{tab:beta-sig-peel-table}).
+Estimates of residual variability in northern component recruitment, as measured by the conditional or marginal standard deviation of the recruitment random effects, increased slightly with the number of peeled years.
+However, the ratio of standard deviations of models with and without temperature effects was stable with about a 20% reduction in residual standard deviation when temperature effects were included.
+
+Although evidence for temperature effects on age 1 natural mortality was weak, it is notable that the sign of the estimated effect for the northern stock component differed depending on whether random effects on age 1 natural mortality were also included ($\widehat \beta_R = -1$ for $M_{6}$ or $M_{13}$) with the negative effect estimated with these random effects ($M_{13}$).
+
+## Stock size, fishing {-}
+
+For the best performing model $M_1$, estimates  of SSB
+Stock size is also increasing for the southern component, just not as quickly as the northern component.
+(plot of SSB time series for each component).
+
+Figure \ref{fig:M1-SSB-F-R}
+
+Figure \ref{fig:SSB-F-R-rel-M1} Annual estimates of SSB and F varied little among the 14 fitted models.
+
+plot of prior/posterior for movement rate.
+
+plot of average F by fleet,region.
+
+Sectivity random effects (Figure \ref{fig:selectivity-re})
+
+Estimates of effect size, sd, correlation for each peel (Table \ref{}
+
  The residual variation in the standard deviation of recruitment random effects is reduced because the expected recruitment (Eq.
-\ref{eq:expected-recruitment}) is a function of the covariate  (Figure \ref{fig:relative-recruitment-bt}).
-This effect is included in the proposed base model (Figure \ref{fig:rec-bottom-temp}).
-Because the covariate is technically for temperature after the beginning of the calendar year when the fish are considered age 1 in the model, a comparison of the proposed base model to one with the bottom temperature effect on natural mortality at age 1 instead might be of interest in the future.
-An investigation of higher order orthogonal polynomial effects might also be of interest particularly for the southern component which might be experiencing more frequently higher temperatures less favorable to overwinter survival.
+\ref{eq:expected-recruitment}) is a function of the covariate.
+This effect is included in the proposed base model (Figure \ref{fig:BT-Ecov-R}).
+
+plot of northern component predicted recruitment vs. time with color by bottom temperature
+
+## Reference points {-}
+
+## Projections {-}
+
+Projections of bottom temperature and recruitment and CVs (Figure \ref{fig:R-BT-proj})
+
+Uncertainty in both is much lower during the data years than the projection years.
+
+Uncertainty in BT RE propagates into both expected recruitment and estimated random effect for recruitment, but the latter also includes uncertainty due to the residual variability in recruitment.
 
@@
 
-The new features of Multi-WHAM include
+## General Model aspects {-}
 
-* seasonal intervals within years
-* variation in movement rates by stock, region-to-region, season, age, and year
-* effects of environmental covariates on movement rates by stock, region-to-region, season, and age
-* effects of environmental covariates on mortality rates by stock, region, and age
-* effects of environmental covariates on recruitment by stock
-* stock-specific stock-recruitment models
-* priors for movement rates
-* seasonal operation of fleets
-* mortality and movement modeled sequentially or simultaneously
-* options for weighting of stock-specific SSB/R for global SPR-based reference points
+WHAM provides a comprehensive treatment of environmental covariates and their effects on populations using state-space methods  [@maunder24].
+This framework accounts for the magnitude of differing uncertainties in observations and stochastic population dynamics processes.
+This version 2.0 extends these inferences to movement rates for multi-region models.
+
+## Future developments {-}
+
+An obvious limitation to WHAM is the inability to include tagging observations of any type.
+Such observations are critical to estimation of movement parameters [e.g., @goetheletal19], but also can inform mortality rate parameters [@hampton91].
+It is well known that natural mortality is seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality from other assessment model parameters [@leeetal11;@clark22].
+Estimation of natural mortality may be even more challenging within state-space assessment models with their greater flexibility from inclusion of time-varying random effects.
+For example, we fixed the mean natural mortality for age 1 fish and @cadigan16 and @stocketal21 also estimated natural mortality deviations.
+Therefore allowing tagging observations in WHAM should be a high priority even for models with a single stock and region.
+
+The work by @correaetal23 on incorporating length information and modeling growth within WHAM has not yet been merged into the multi-stock version 2.0. Growth and movement would be challenging.
+
+## Black sea bass {-}
+
+In our investigation of bottom temperature effects on recruitment and natural mortality and time-varying random effects on natural mortality at age 1, we found only including effects on recruitment for the northern stock component to be the best model with respect to AIC.
+This is the same configuration configuration accepted during the assessment peer-review process and it is the model currently used for management [@nefsc23].
+
+Although not relevant to the best performing model, the effect of bottom temperature on age 1 M would be expected to be opposite of that for recruitment because lower M with higher temperature would produce higher recruitment.
+This expected effect was estimated only when random effects on age 1 M were also included in the model.
+This seems to provide further support to including random effects when examining covariate effects on population parameters as demonstrated by @milleretal18 for effects on growth parameters.
+
+AIC suggested some weight for models with temperature effects on recruitment for the southern component and random effects on M at age 1.
+This might suggest making inferences on the population by using an ensemble of these and weighting by AIC [@burnhamanderson02].
+However, the estimates of assessment outputs relevant for management (SSB, F) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
+[but what about reference points?]
+
+### Temperature effects {-}
+
+BSB in the NEUS is at the northern extent of its range.
+hence the range extension to the north with increased temperature.
+The evidence for temperature effects on recruitment was strong only for the northern component which is the component of the stock at the limit of the species' range.
+Opposite effects of temperature on recruitment or population size would be expected for species in the same general area that are at the southern extent of their range [@gabriel92].
+For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer fall on the Northeast US shelf [@sullivanetal05,@milleretal16].
+
+### Self Tests {-}
+
+Simulation self tests performed by @stockmiller21 simulated process errors and observation errors and their simulations showed little bias in estimated assessment output such as SSB.
+@lietal24 used the same type of simulations and found what?.
+
+Here we conditioned on the estimated process errors for black sea bass.
+This conditional approach has been suggested as appropriate for tactical management where we are interested in inferences assuming our estimation of the population and fishing history from the original data is the truth [@zhengcadigan23,@cadiganetal24].
+We found some bias in SSB estimation for the black sea bass application (negatively for the northern stock), using the default model.
+We also found instability in the estimation of the observation variance parameters for the aggregate indices such that they often were estimated at the lower bound of 0.
+However, we found that the magnitude of bias was largely influenced by the large uncertainty estimated in some of the age composition observations assumed to have a logistic-normal distribution.
+When using the logistic-normal likelihood for management, we recommend inspecting the size of the estimated (marginal) standard deviation of the logistic normal model when self-test bias is non-negligible.
+We also recommend investigating the multiplicative transformation for the logistic normal rather than the additive transformation which is currently the only option in WHAM [@cadigan16].
+
+?Selectivity RE are only on index age comp in the north which may be why bias is larger there?
+
+?
+We found REML provided less biased estimation of SSB and recruitment which are functions of random effects.
+Whether REML should be the default approach for estimating these quantities from state-space stock assessment models requires further investigation.
+For example, improvement in bias may be negligible when data contain low observation error and/or there are few fixed effects parameters.
+? larger variance estimates allows greater spread of random effects estimates.
+Important for conditional self-test where estimated RE are of primary importance (Recruitment and SSB).
+When simulating RE and refitting, the sign of differences between estimates and true values can be random across simulations.
+Probably more important with more fixed effects and less informative data
+
+Including tagging data in WHAM would also be helpful for BSB because we are using the same data to estimate movement outside of the model as we are using to estimate the other parameters in WHAM.
+
+## Conclusions {-}
+
+WHAM version 2.0 extends the existing R package to allow multiple stocks and discrete spatial regions.
+It allows autocorrelated random effects and environmental covariate effects on movement rates.
+Estimates of movement rates from auxiliary studies can be used to construct prior distributions for movement rates in lieu of integrated likelihoods for tagging data.
+Version 2.0 also allows seasonal treatment of different fishing fleets and movement of individual stocks, and internal estimation of SPR- and MSY-based reference points accounting for seasonal spatial dynamics including movement and mortality and any environmental covariate effects.
+
+We applied this extended version to investigate alternative hypotheses about effects of bottom temperature on recruitment and age 1 natural mortality for black sea bass.
+Our analyses indicate evidence for effects of bottom temperature on recruitment for the northern component represents was stronger than other models that included effects on age 1 mortality and/or corresponding effects on the southern stock component.
+
+Future WHAM development should prioritize including tagging data to inform stock assessment parameter estimation and merging in the version created by @correaetal23 to allow modeling of growth and inclusion of length and length-at-age composition observations.
 
@@
 
+# Figures {-}
+
 ```{r migration-diagram, echo = FALSE, out.width="80%", fig.align="center", fig.cap = "Diagram of intervals within the year and configuation of the dynamics of each component of the BSB population."}
@@
 ```
+\pagebreak
+
+```{r move-prior-posterior, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Prior (black) and posterior (red) distributions of movement of northern component stock from north to south (top) and south to north (bottom)."}
+knitr::include_graphics(here::here("paper", "move_prior_post.pdf"))
+```
+\pagebreak
+
+```{r bottom-temperature, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Observations and 95\\% confidence intervals (points with vertical lines) and posterior estimates (lines) with 95\\% confidence intervals (polygons) of bottom temperature anomalies in the north and south regions from model $M_0$."}
+knitr::include_graphics(here::here("paper", "Ecov.pdf"))
+```
+\pagebreak
+
+```{r BT-Ecov-R, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Expected and Posterior Emperical Bayes estimates of annual recruitment for the northern stock component.
+Color defined by the corresponding annual bottom temperature anomaly."}
+knitr::include_graphics(here::here("paper", "best_R_Ecov.pdf"))
+```
+\pagebreak
+
+```{r M1-SSB-F-R, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of SSB, average F (ages 6-7), and recruitment from model $M_1$.
+Polygons represent 95\\% confidence intervals."}
+knitr::include_graphics(here::here("paper", "M_1_SSB_F_R.pdf"))
+```
+\pagebreak
+
+```{r R-BT-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Estimates of bottom temperature, recruitment, and expected recruitment, and coefficients of variation for the latter two from model $M_1$ assuming alternative bottom temperature scenarios described in the text.
+Polygons represent 95\\% confidence intervals."}
+knitr::include_graphics(here::here("paper", "R_proj_results.pdf"))
+```
+
+\pagebreak
+
+```{r selectivity-re, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Time and age-varying selectivty for fleets and indices in the northern region with autoregressive random effects."}
+knitr::include_graphics(here::here("paper", "selectivity_re_plot.pdf"))
+```
+\pagebreak
+
+\clearpage
+
+# Tables {-}
 
@@
 \renewcommand\thetable{\arabic{table}}
-
-```{r, echo = F, eval = T}
-age_comp_names <- c("North Commercial", "North Recreational", "South Commercial", "South Recreational",
-  "North Recreational CPA", "North VAST", "South Recreational CPA", "South VAST")
-age_comp_mod <- c("Dirichlet-Multinomial", "Logistic-normal (0s as missing)", "Logistic-normal (AR1, 0s as missing)","Logistic-normal (AR1, 0s as missing)", "Logistic-normal (0s as missing)","Dirichlet-Multinomial","Logistic-normal (AR1, 0s as missing)","Logistic-normal (AR1, 0s as missing)")
-mean_sel_mod <- c("age-specific (flat-topped at ages > 3)",
-  "age-specific (flat-topped at ages > 6)",
-  "logistic", "logistic",
-  "age-specific (flat-topped at ages > 1)",
-  "age-specific (flat-topped at ages > 4)",
-  "age-specific (flat-topped at ages > 2)",
-  "age-specific (flat-topped at ages > 1)")
-sel_re_mod <- c("2D-AR1 (age and year)", 
-  "2D-AR1 (age and year)",
-  "None",
-  "None",
-  "AR1 (year)",
-  "2D-AR1 (age and year)",
-  "None",
-  "None")
-out <- cbind.data.frame("Data component" = age_comp_names,  "Age Composition Likelihood" = age_comp_mod,  
-  "Mean Selectivity model" = mean_sel_mod, "Random effects configuration" = sel_re_mod)
-
-out %>% kable(format = type, booktabs = T, escape=F, row.names = F, label = "age-comp-sel-table",
-    caption="Configuration of Age composition likelihood, mean selectivity model, and selectivity random effects for each age composition data component.", align = c(rep("l",4))) %>% landscape()
-#knitr::knit_print(kableExtra::landscape(out))
-
-```
 
@@
 row.names(rho_table) <- NULL
-colnames(rho_table) <- c("Model", rep(c("North","South"), 3))
+colnames(rho_table) <- c("Model", rep(c("North","South"), 2))
 rho_table %>%
   kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "rho-table",
-  caption = "Mohn's $\\rho$ for SSB, R, average F at ages 6 and 7 in northern and southern regions.") %>%
-  add_header_above(c(" " = 1, "SSB" = 2, "Average F" = 2, "Recruitment" = 2))
+  caption = "Mohn's $\\rho$ for SSB, and average F at ages 6 and 7 in northern and southern regions.") %>%
+  add_header_above(c(" " = 1, "SSB" = 2, "Average F" = 2))
 ```
+```{r, echo = F, eval = T}
+par_names <- paste0("$M_{", c(1,2,3,3),"}$ ", "$\\widehat{\\beta}_{R,", c("N","S"), "}$")
+par_names <- c(par_names, paste0("$M_{", rep(0:1, 3),"}$ ", rep(c("Conditional ", "", "Marginal "),each = 2), "$\\widehat{\\", rep(c("sigma","rho","sigma"), each = 2),"}_{R,N}$"))
+peel_par_table <- t(readRDS(here::here("results/beta_sigma_par_peel_table.RDS"))[,1:10])
+# peel_par_table <- readRDS("../results/beta_sigma_par_peel_table.RDS")[,1:10]
+#colnames(peel_par_table) <- par_names
+peel_par_table <- cbind.data.frame(par_names, round(peel_par_table,3))
+colnames(peel_par_table) <- c("Parameter", as.character(0:7))
+row.names(peel_par_table) <- NULL
+peel_par_table %>%
+  kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "beta-sig-peel-table",
+  caption = "Estimates of temperature effects on recruitment and variance and autocorrelation parameters for recruitment for northern ($N$) and southern ($S$) components for models with no effects ($M_0$), effects on specific components ($M_1$ and $M_2$) or both components simultaneously ($M_3$) from the full model (Peel 0) and each retrospective peel.")%>%
+  add_header_above(c(" " = 1, "Peel" = 8))
+
+```
+
 \clearpage
@@
 The mean values did not differ meaningfully from the transformation of the original estimates ($y_{1\rightarrow 2} = -3.79$ and $y_{2\rightarrow 1} = -0.79$) and the standard deviation was approximately 0.2 for both parameters.
+
+## Diagnostics {-}
+
+### Jitter fits for model $M_0$
+
+WHAM by default completes three newton steps after the stats::nlminb minimization function completes to reduce the gradient at the minimized NLL.
+However, this generally has negligible effects on model estimates and the NLL.
+To reduce computation time, we did not complete these newton steps when performing jitter fits of the model.
+Without the Newton steps, the maximum (absolute) gradient sizes are generally less than 0.01 for models that converge satisfactorily.
+
+The 50 jitter fits demonstrated that a local minimum was obtained for the original fit of model $M_0$ (Figure \ref{fig:jitter-M0}).
+Some lower NLLs were obtained with unacceptable gradients, but a slightly lower NLL was found with a satisfactory gradient and with and invertible hessian.
+We therefore refit model $M_0$ and all remaining models using the better parameter estimates as initial values.
+
+### Jitter fits for model $M_1$
+
+The 50 jitter fits gave no evidence of a better minimization of the NLL.
+Three lower NLLs were obtained, but with unacceptably large gradients (Figure \ref{fig:jitter-M1}).
+The largest differences in parameter estimates for these three jitters were for numbers at age and selectivity random effects variance and correlation parameters.
+
+### Self test for model $M_1$
+
+Initial fits to simulated data from model $M_1$ showed estimation of the observation error standard deviation multiplier for the recreational catch-per angler indices in the north and south regions was unstable.
+Many of the fits to the simulated data produced implausible estimates at the 0 boundary for these parameters (very negative values on log-scale).
+Therefore, we completed self-tests with these parameters fixed at the true values.
+
+For 7 of the the simulated data sets the model failed to optimize.
+The maximum absolute gradient was $<10^{-6}$ for only 9 and $<10^{-4}$ for 52 of the 93 successfully fitted models.
+The poor convergence appeared to be attributable to the estimation of the scalar for the standard errors of the log-transformed northern Recreational CPA index for which estimates tended to 0 for nearly all of the fits ($<0.01$ for 83 fits).
+However, even across all fits including those with poor convergence, the SSB estimates appeared to be reliable (Figure \ref{fig:self-test-fig}).
+
+### One-step-ahead residuals for model $M_1$
+
+One-step-ahead residuals can now be calculated for all index, catch, and environmental covariate observations using methods described by @thygesenetal17 and @trijouletetal23.
+
+<!--
+
+We used one-step-ahead (OSA) residuals to evaluate mis-specification of the proposed base model.
+Pearson residuals do not have the appropriate properties (independent standard normal for a correctly specified model) for composition observations [@trijouletetal23] or for covariate and aggregate index and catch observations when temporal random effects are employed  because of the lack of independence of observations [@thygesenetal17].
+
+There is no evidence of mis-specificaiton from the OSA residuals for aggregate catch of the northern commercial fleet (Figure \ref{fig:osa-North-comm-catch-summ}).
+There was some indication of trends in residuals of some of the age composition observations early in the time series and for the the first age class (Figures \ref{fig:osa-North-comm-paa-summ} and \ref{fig:osa-North-comm-paa}).
+
+For the northern recreational fleet, the OSA residuals for the aggregate catch appeared satisfactory (Figure \ref{fig:osa-North-rec-catch-summ}).
+The OSA residuals for the age composition observations for the northern recreational fleet showed some tendency of underdispersion but were otherwise unnoteworthy (Figures \ref{fig:osa-North-rec-paa-summ} and \ref{fig:osa-North-rec-paa}).
+
+Neither the OSA residuals for the aggregate catch or the age composition of the Northern recreational CPA index exhibited signs of mis-specification (Figures \ref{fig:osa-North-reccpa-catch-summ} to \ref{fig:osa-North-reccpa-paa}).
+
+There was a little evidence of tendency toward negative OSA residuals for the Northern VAST index whereas there was some indication of positive residuals for the age composition observations, particularly at age 1 (Figures \ref{fig:osa-North-vast-catch-summ} to \ref{fig:osa-North-vast-paa})
+
+THe OSA residuals for the aggregate catch of the southern commercial fleet tended to be negative, but the residuals for the age composition observations did not show any evidence of mis-specification (Figures \ref{fig:osa-South-comm-catch-summ} to \ref{fig:osa-South-comm-paa}).
+
+The OSA residuals for the aggregate catch of the southern recreational fleet were under-dispersed and those for the age composition were somewhat underdispersed, but no trends in either types of observations were apparent (Figures \ref{fig:osa-South-rec-catch-summ} to \ref{fig:osa-South-rec-paa})
+
+The OSA residuals for the aggregate catch of the southern recreational CPA index were also somewhat under-dispersed (Figures \ref{fig:osa-South-reccpa-catch-summ}).
+The residuals for the age composition observations exhibited a few large negative residuals at older ages, but otherwise exhibited no apparent trends (Figures \ref{fig:osa-South-reccpa-paa-summ} to \ref{fig:osa-South-reccpa-paa})
+
+The OSA residuals for the aggregate southern VAST index were somewhat over-dispersed and the residuals for the age composition appeared to show some trend with observed proportion and age (Figures \ref{fig:osa-South-vast-catch-summ} to \ref{fig:osa-South-vast-paa}).
+-->
+
+\setcounter{table}{0}
+\renewcommand\thetable{S\arabic{table}}
+
+```{r, echo = F, eval = T}
+age_comp_names <- c("North commercial fleet", "North recreational fleet", "South commercial fleet", "South recreational fleet",
+  "North recreational CPA index", "North VAST index", "South recreational CPA index", "South VAST index")
+age_comp_mod <- c("Dirichlet-Multinomial", "Logistic-normal (Independent)", "Logistic-normal (AR1 correlation)","Logistic-normal (AR1 correlation)", "Logistic-normal (Independent)","Dirichlet-Multinomial","Logistic-normal (AR1 correlation)","Logistic-normal (AR1 correlation)")
+mean_sel_mod <- c("age-specific (ages > 3 fully selected)",
+  "age-specific (ages > 6 fully selected)",
+  "logistic", "logistic",
+  "age-specific (ages > 1 fully selected)",
+  "age-specific (ages > 4 fully selected)",
+  "age-specific (ages > 2 fully selected)",
+  "age-specific (ages > 1 fully selected)")
+sel_re_mod <- c("AR1 correlation by age and year", 
+  "AR1 correlation by age and year",
+  "None",
+  "None",
+  "AR1 correlation by year",
+  "AR1 correlation by age and year",
+  "None",
+  "None")
+out <- cbind.data.frame("Data component" = age_comp_names,  "Age Composition Likelihood" = age_comp_mod,  
+  "Mean Selectivity model" = mean_sel_mod, "Random effects Model" = sel_re_mod)
+
+out %>% kable(format = type, booktabs = T, escape=F, row.names = F, label = "age-comp-sel-table",
+    caption="Configuration of age composition likelihoods, mean selectivity models, and selectivity random effects models for each age composition data component.
+For all logistic-normal likelihoods, any ages observed as zeros are treated as missing.", align = c(rep("l",4))) %>% landscape()
+#knitr::knit_print(kableExtra::landscape(out))
+
+```
+
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\pagebreak
+
+```{r jitter-M0, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_0$.
+Black jitters had maximum absolute gradient values < $10^{-10}$, grey jitters had values > $10^{-10}$ and < 1, and red jitters had values > 1."}
+knitr::include_graphics(here::here("paper", "fit_0_jitter_plt.pdf"))
+```
+\pagebreak
+
+```{r jitter-M1, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_1$.
+Fits with black dots had maximum absolute gradient value < 0.01 and fits with red dots had values > 10."}
+knitr::include_graphics(here::here("paper", "fit_1_jitter_plt.pdf"))
+```
+\pagebreak
+
+\begin{landscape}
+
+```{r self-test-fig, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Median relative bias of SSB for the North and South stock components for estimation models where the observation variance of log-indices are fixed and estimation is by maximum marginal likelihood (ML) or Restricted Maximum Likelihood (REML) or where the standard deviation of logistic-normal age composition observations are assumed to be 0.1 (over sqrt(1000)) (Low Error).
+Black and Red dashed lines represent the median of the yeaerly medians and the median across all yearly relative errors, respectively."}
+knitr::include_graphics(here::here("paper", "self_test_results.pdf"))
+```
+\end{landscape}
+
+```{r SSB-F-R-rel-M1, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Estimates of SSB, F, and recruitment relative to those of the best performing model, $M_1$."}
+knitr::include_graphics(here::here("paper", "SSB_F_R_rel_M1.pdf"))
+```
+
+```{r Ecov-M1-rel-M0, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Relative differences in posterior estimates of northern region bottom temperature anomalies ($\\widehat X$) from the null model without effects on recruitment ($M_0$) and with effects on the northern stock component ($M_1$)."}
+knitr::include_graphics(here::here("paper", "Ecov_M1_rel_M0.pdf"))
+```
+
+\pagebreak

### 2024-11-19 paper/paper.tex
saving work

@@
 \hypersetup{
-  pdftitle={A multi-stock, multi-region state-space age-structured assessment model with an application to black sea bass},
+  pdftitle={Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass},
   pdfauthor={Timothy J. Miller1; Kierstin Curti; Alex Hansell},
@@
 
-\title{A multi-stock, multi-region state-space age-structured assessment
-model with an application to black sea bass}
+\title{Space for WHAM: a multi-region, multi-stock generalization of the
+Woods Hole Assessment Model with an application to black sea bass}
 \author{Timothy J. Miller\textsuperscript{1} \and Kierstin
 Curti \and Alex Hansell}
-\date{16 September, 2024}
+\date{14 November, 2024}
 
@@
 
-A multi-region, multi-stock generalization of WHAM with an application
-evaluating evidence of alternative hypotheses about temperature effects
-on black sea bass
+Describes the multi-region, multi-stock generalization of WHAM and its
+usage to evaluate evidence of alternative hypotheses about temperature
+effects on recruitment of black sea bass stock components.
 
@@
 autocorrelated random effects and environmental covariates. We then
-illustrate the model with northern and southern components of the NEUS
-black sea bass stock and evaluate alternative hypotheses of bottom
-temperature and random effects on recruitment and natural mortality. We
-show strong evidence for temperature effects on recruitment and no
-random effects or temperature effects on natural mortality, primarily
-for the northern stock component.
+illustrate the model by applying it to data for the northern and
+southern components of the NEUS black sea bass stock and evaluate
+alternative hypotheses of bottom temperature and random effects on
+recruitment and natural mortality. We show strong evidence for
+temperature effects on recruitment, primarily for the northern stock
+component, and no evidence for including random effects or temperature
+effects on age 1 natural mortality.
 
@@
 
-Why is this model needed. Models ignoring movement can provide biased
-catch advice. Using random effects for time-varying processes is
-considered best practice.
-
-The Woods Hole Assessment Model (WHAM) is an R package developed at the
-Northeast Fisheries Science Center
-(\url{https://timjmiller.github.io/wham}, Miller and Stock 2020; Stock
-and Miller 2021). Primary inputs and observation types are similar to
-ASAP and there is functionality built into WHAM to move information from
-ASAP input files to R. WHAM can be configured to fit SCAA models often
-identically to ASAP, so that existing assessments in the U.S. Northeast
-can be replicated and tested against state-space models with process
-errors and environmental effects in a single framework. WHAM models are
-built using the Template Model Builder package (TMB, Kristensen et al.
-2016) which provides efficient fitting of models that include random
-effects similar to the State-space Assessment Model (SAM, Nielsen and
-Berg 2014), which is used to manage many stocks within ICES. WHAM is
-currently used to manage 9 stocks in the NEUS.
-
-The Woods Hole Assessment Model (WHAM) software package is maintained
-and developed at the Northeast Fisheries Science Center to enable
-state-space stock assessments, i.e.~where processes such as the annual
-transitions in numbers-at-age (survival), natural mortality,
-selectivity, and catchability are treated as time- and(or) age-varying
-random effects. WHAM can also be configured without random effects as a
-traditional statistical catch-at-age model in order to bridge from
-current assessments which use Age-Structured Assessment Program (ASAP).
-Here, we present an extension of WHAM (Multi-WHAM) to model multiple
-stocks and(or) regions and allow movement of stocks among regions.
-
-It has been reviewed in the literature and simulation tested (Stock et
-al. 2021; Stock and Miller 2021). It has also been used as the operating
-model in the index-based assessment methods research track assessment
-(Legault et al. 2023) and is also the main tool for simulation studies
-in the research track on state-space assessment models.
-
-A review of the essential features for next-generation stock assessments
-concluded that only WHAM and a similar State-Space Assessment Method
-(Nielsen and Berg 2014) model random effects correctly (Punt et
-al.~2020). WHAM is now being used as the management model for 9 stocks
-in the northeast U.S., but the standard version of WHAM can only be
-applied to a single stock and area (Miller and Stock 2020).
-
-Here, we describe a multi-stock and multi-region extension of the WHAM
-package (version 2.0 \url{https://github.com/timjmiller/wham/tree/lab}).
-Numerous new configuration options are also included in this extension
-that can be useful even for single-stock or single region models. Any
-number of stocks or regions can be configured in this version of WHAM.
-The multi-stock extension was developed to accommodate movement of
-stocks or stock components in anticipation of black sea bass and
-Atlantic cod stocks where new modeling frameworks where recently
-examined through Northeast research track assessment process (cite RT
-docs). We demonstrate capabilities of the new version of WHAM with black
-sea bass and evaluate evidence for alternative hypotheses of temporal
-variation in natural mortality and effects of bottom temperature on
-recruitment and natural mortality of age 1 individuals.
+A state-space statistical approach and maximum marginal likelihood or
+Bayesian fitting of stock assessment models allows estimation of time-
+and age-varying population attributes as random effects Miller et al.
+(2016b). This estimation approach is considered an essential feature of
+gold-standard assessment models that we use in tactical management of
+commercially important fish stocks (Punt et al. 2020). The State-space
+Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed
+and remains widely used within ICES to assess European fish stocks.
+Various state-space models are being used to manage cod and plaice
+stocks in the waters of Eastern Canada (Perreault et al. 2020; Varkey et
+al. 2022) and to the south, the Woods Hole Assessment Model (WHAM, Stock
+and Miller 2021) is now used to assess many fish stocks in the Northwest
+Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).
+
+WHAM is an R package developed and maintained at NOAA's Northeast
+Fisheries Science Center (\url{https://timjmiller.github.io/wham},
+Miller and Stock 2020; Stock and Miller 2021). WHAM can be configured to
+fit a wide range of age-structured models from traditional statistical
+catch-at-age models without any random effects to models with several
+time and age varying process errors and possibly effects of
+environmental covariates on various demographic parameters. Like SAM,
+WHAM models are built using the Template Model Builder package (TMB,
+Kristensen et al. 2016) which provides a computationally efficient means
+of fitting an extremely wide class of models with random effects. WHAM
+has undergone active development since its creation and includes random
+effects options for in recruitment, inter-annual transitions numbers at
+age (hereafter referred to as ``survival''), fishery and index
+selectivity, natural mortality, and catchability.
+
+However, WHAM has up to now only allowed models with one stock
+(component) and a single region and without seasonal changes in stock
+dynamics. Using such models for stocks that have subcomponents with
+varying seasonal movement can provide incorrect inferences and poor
+management advice (Ying et al. 2011; Cao et al. 2014; Bosley et al.
+2022). Furthermore, the ability to account for spatial structure and
+model multiple stocks are also important features of leading-edge
+assessment modeling frameworks (Punt et al. 2020). We describe here the
+implementation of these features and other extensions since Stock and
+Miller (2021) in WHAM version 2.0. Many of these new configuration
+options can be useful whether modeling 1 or more stocks and regions.
+
+This extension of WHAM was developed in concert with a stock assessment
+for black sea bass through the NEFSC research track assessment process
+where new modeling frameworks were recently examined and modeling
+multiple stocks or stock components simultaneously was of interest
+(NEFSC 2023). \textcolor{red}{[Text on black sea bass here]} We apply
+WHAM 2.0 to two stock components of black sea bass off the coast of the
+NEUS and evaluate evidence for alternative hypotheses of temporal
+variation and effects specifically of bottom temperature on recruitment
+and natural mortality of age 1 individuals.
 
@@
 
-\hypertarget{model-description}{%
-\subsection*{Model description}\label{model-description}}
-\addcontentsline{toc}{subsection}{Model description}
-
-Many of the options and equations of version 2.0 are are the same as
-those described by Stock and Miller (2021), so we will only describe
-extensions and differences that have occurred since the first
-description of WHAM. The new version of WHAM can model multiple stocks,
-but survival, movement, harvest and natural mortality is are tracked for
-each stock. Therefore, much of the description below assumes a specific
-stock \(s\), but this subscript is only explicitly used when necessary.
+\hypertarget{wham-description}{%
+\subsection*{WHAM description}\label{wham-description}}
+\addcontentsline{toc}{subsection}{WHAM description}
+
+Many of the options and equations of WHAM version 2.0 are the same as
+those in Stock and Miller (2021), so we will only describe extensions
+and differences that have occurred since their first description of
+WHAM. The new version of WHAM can model multiple stocks and survival,
+movement, harvest and natural mortality are tracked for each stock. Much
+of the description below is for a specific stock \(s\), but, for
+simplicity, this subscript is implicit except when necessary.
 
@@
 categories or states and treat the number of individuals occurring in
-each category over time as a multi-state model (reference). Multi-state
-models use a probability transition matrix (PTM) that describes the
-probability of individuals living and moving among regions or dying due
-to fishing or natural mortality over a time interval \(i\) with duration
-\(\delta_i\) in year \(y\) for individuals at age \(a\) on January 1.
-Each row and column of the PTM correspond to one the states: alive in
-region \(r\), dead in fleet \(f\), or dead from natural causes. The
-probabilities in each row sum to unity and assume an individual is in
-the corresponding state at the beginning of the interval. Given \(n_R\)
-regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and
-columns) as a function of sub-matrices is \begin{equation}\label{eq:ptm}
+each category over time as a multi-state model. Approaches to modeling
+transitions among these tates may treat time discretely (e.g., Arnason
+1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987;
+Commenges 1999; Andersen and Keiding 2002). Multi-state models can
+define a probability transition matrix (PTM) that describes the
+probability of individuals occurring in different states at the end of a
+time interval \(\delta\), conditional on being in each of those state at
+beginning of the interval. For fish populations these states would be
+defined as being alive in a particular region or being dead due to
+fishing from a particular fleet or natural mortality. The time interval
+\(i\) with duration \(\delta_i\) would be a season and the PTMs would be
+uniquely defined for each stock by season \(i\), year \(y\), and age
+\(a\) on January 1. Each row and column of the PTM correspond to one the
+states: alive in region \(r\), dead in fleet \(f\), or dead from natural
+causes. The probabilities in each row sum to unity and assume an
+individual is in the corresponding state at the beginning of the
+interval. Given \(n_R\) regions and \(n_F\) fleets, the square PTM
+(\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
+\begin{equation}\label{eq:ptm}
   \mathbf{P}_{y,a,i} = \begin{bmatrix}
@@
 in one of the mortality states given starting the interval in that state
-is unity.
+is unity (no zombies allowed).
 
@@
 than true probabilities where numbers in each state would be multinomial
-distributed. The PTMs determine the expected numbers 1) in each state on
-January 1 of year \(t+1\) at age \(a+1\) given the abundances at age
-\(a\) on January 1 of year \(t\), 2) captured over the year in each
-fleet, 3) available to each index, and 4) alive at the time and in the
-region where spawning occurs.
+distributed as in a model for tagging data where fates of individual
+fish are assumed independent. The PTMs determine the expected numbers 1)
+in each state on January 1 of year \(t+1\) at age \(a+1\) given the
+abundances at age \(a\) on January 1 of year \(t\), 2) captured over the
+year in each fleet, 3) available to each index, and 4) alive at the time
+and in the region where spawning occurs.
 
@@
 classes. So the general covariance structure for a given stock at ages
-\(a>1\) in region \(r\) is \begin{equation*}
-  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma^2_{N,r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}.
-\end{equation*}
+\(a>1\) in region \(r\) is that of a two-dimensional first-order
+autoregressive (2DAR1) process \begin{equation*}
+  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}
+\end{equation*} and that for age 1 is just AR1 across years
+\begin{equation*}
+  Cov\left(\epsilon_{y,1},\epsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
+\end{equation*} Since recruitment for a given stock currently only
+occurs in one region \(r_s\) there is only a single time-varying
+recruitment random effect for each stock.
 
@@
 
-The initial abundances at age can also be treated as independent or
-autocorrelated (with age) random effects. Defining the vector of initial
-abundance at age in region \(r\) as \(\mathbf{N}_{O,1,r}\), the general
-model is
+The initial abundances at age can also be treated as independent or as
+AR1 random effects. Defining the vector of initial abundance at age in
+region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
 \[\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}\]
@@
 are season-, and(or) region-to-region-specific and modeled most
-generally as a two-dimensional first-order autoregressive random effects
-with age and(or) year where the covariance is \begin{equation*}
+generally as 2DAR1 random effects with age and(or) year where the
+covariance is \begin{equation*}
   Cov\left(\epsilon_{r\rightarrow r',y,a,i},\epsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
@@
 estimated, (mean) mortality rates may be stock-, region-, and
-age-specific. When random effects are used, the same autocorrelation
-structures with age and year as described by Stock and Miller (2021) can
-be configured for a given stock and region. Any environmental covariate
+age-specific. When random effects are used, the same 2DAR1 structure
+with age and year as described by Stock and Miller (2021) can be
+configured for a given stock and region. Any environmental covariate
 effects can be stock-, region-, and age-specific. So the general model
@@
 \end{equation*} The general covariance structure for random effects are
-modeled most generally as a two-dimensional first-order autoregressive
-random effects with age and(or) year where the covariance is
-\begin{equation*}
+modeled most generally as 2DAR1 random effect with age and(or) year
+where the covariance is \begin{equation*}
   Cov\left(\epsilon_{M,y,a,r},\epsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
@@
 catchability for index \(m\) (defaults are 0 and 1000) and the general
-covariance structure for the annual random effects is first-order
-autoregressive
+covariance structure for the annual random effects is AR1
 \[Cov\left(\epsilon_{q,m,y},\epsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.\]
@@
 or use the average over user specified years, analogous to how natural
-mortality can be treated in the projection period. The projections of
-any environmental covariates has been revised to better include error in
-the estimated latent covariate in any effects on the population in
+mortality can be treated in the projection period. The projection of any
+environmental covariates has been revised to better include error in the
+estimated latent covariate in any effects on the population in
 projection years.
@@
 
-Prior to the most recent peer-reviewed assessment of black sea bass in
-the NEUS, the stock was assessed using the Age-Structured Assessment
-Program (ASAP) model (Legault and Restrepo 1999), a single-stock and
--region statistical catch-at-age model that estimates all model
-parameters as fixed effects. Northern and southern components of the
-NEUS black sea bass stock divided by the Hudson Canyon were separately
-modeled in ASAP (reference map figure). Results from the separate ASAP
-models have been aggregated for a unit-stock assessment. Furthermore,
-the ASAP-based assessments have exhibited strong retrospective patterns
-(Mohn 1999), and exploring alternative modeling approaches for the
-northern and southern stock components has been a high priority for
-management.
-
-In 2023, a working group composed of scientists from federal, state, and
-academic institutions determined an optimal data and model configuration
-for the black sea bass stock using the multi-stock and multi-region
-extension of WHAM described above (NEFSC 2023). This assessment included
-the spatial features and hypothesized environmental drivers that were
-prioritized research recommendations from previous black sea bass
-assessments.
+Prior to its 2023 peer-reviewed assessment, the NEUS black sea bass
+stock was assessed using the Age-Structured Assessment Program (ASAP)
+model (Legault and Restrepo 1999), a single-stock and -region
+statistical catch-at-age model that estimates all model parameters as
+fixed effects. Northern and southern components of the NEUS black sea
+bass stock ascribed to regions divided by the Hudson Canyon were
+separately modeled in ASAP (reference map figure). Results from the
+separate ASAP models were combined for a unit-stock assessment. The
+ASAP-based assessments exhibited strong retrospective patterns (Mohn
+1999), and exploring alternative modeling approaches for the northern
+and southern stock components has been a high priority for management.
+
+Leading up to the 2023 peer-reviewed assessment, a working group
+(hereafter referred to as ``working group'') composed of scientists from
+federal, state, and academic institutions determined an optimal data and
+model configuration for the black sea bass stock using the multi-stock
+and multi-region extension of WHAM described above (NEFSC 2023). This
+assessment included the spatial features and investigated inclusion of
+hypothesized environmental drivers that were prioritized research
+recommendations from previous black sea bass assessments.
 
@@
 assumptions for recruitment and natural mortality in the models we fit
-to evaluate alternative hypotheses of environmental effects.
+to evaluate alternative hypotheses of bottom temperature effects on
+black sea bass.
 
@@
 The first year being modeled for the population is 1989 and the fishery
-and index data used in the model span from 1989 to 2021. There are two
-stock components (north and south) modeled as separate populations that
-spawn and recruit in respective regions separated by the Hudson Canyon.
-We have observations for each of four total fishing fleets, where two
-fishing fleets (Recreational and Commercial) operate in each region.
+and index data used in the model span from 1989 to 2021. The north and
+south stock components are modeled as separate populations that spawn
+and recruit in respective regions. We have observations for each of four
+total fishing fleets, where two fishing fleets (Recreational and
+Commercial) operate in each region.
 
@@
 July 31, and five monthly intervals from August 1 to December 31. The
-southern stock component is assumed to never move to the northern region
-(Figure \ref{fig:migration-diagram}). For the northern component, a
-proportion \(\mu_1\) can move to the southern region each month during
-the last five months of the year, but no movement is allows from the
-south to the north during this period. During the first four intervals
-of the year a proportion \(\mu_2\) the northern component individuals in
-the south can move back to the north, but no movement from the north to
-south is allowed during this period. In the fifth interval (May), all
-northern component individuals remaining in the south are assumed to
-move back to the north for the subsequent spawning period. Survival and
-movement occur sequentially in each interval and each of the two
-movement proportions are assumed constant across intervals, ages, and
-years.
-
-The monthly movement matrices are \begin{equation*}
+southern stock component is assumed to never move to the northern
+region. For the northern component, a proportion
+\(\mu_{\text{N}\rightarrow \text{S}}\) can move to the southern region
+each month during the last five months of the year, but no movement is
+allows from the south to the north during this period (Figure
+\ref{fig:migration-diagram}). During the first four intervals of the
+year a proportion \(\mu_{\text{S}\rightarrow \text{N}}\) the northern
+component individuals in the south can move back to the north, but no
+movement from the north to south is allowed during this period. In the
+fifth interval (May), all northern component individuals remaining in
+the south are assumed to move back to the north for the subsequent
+spawning period. Survival and movement occur sequentially in each
+interval and each of the two movement proportions are assumed constant
+across intervals, ages, and years.
+
+The two monthly movement matrices are \begin{equation*}
 \boldsymbol{\mu}_{1} = 
   \begin{bmatrix}
-     1-\mu_{1\rightarrow 2} & \mu_{1\rightarrow 2} \\
+     1-\mu_{\text{N}\rightarrow \text{S}} & \mu_{\text{N}\rightarrow \text{S}} \\
      0 & 1 \\
@@
      1 &  0 \\
-     \mu_{2\rightarrow 1} & 1-\mu_{2\rightarrow 1} \\
+     \mu_{\text{S}\rightarrow \text{N}} & 1-\mu_{\text{S}\rightarrow \text{N}} \\
   \end{bmatrix}
-\end{equation*} for the portion of the year before spawning.
-
-The constant or mean log-natural mortality rate is assumed to be
-\(\log(0.4)\)
-
-As noted in the description of the general WHAM model, tagging data are
-not yet allowed. However, the working group also fit a Stock Synthesis
-model (Methot and Wetzel 2013) which provided estimated movement rate
+\end{equation*} for the portion of the year before spawning. As noted in
+the description of the general WHAM model, tagging data are not yet
+allowed. However, the working group also fit a Stock Synthesis model
+(Methot and Wetzel 2013) which provided estimated movement rate
 parameters and standard errors that were used to configure the priors
 for WHAM (see Supplementary Materials).
-
-\hypertarget{initial-abundance-at-age}{%
-\subsection*{Initial abundance at age}\label{initial-abundance-at-age}}
-\addcontentsline{toc}{subsection}{Initial abundance at age}
 
@@
 age as separate parameters can be challenging even in single-stock
-models. To avoid challenges with estimating initial numbers at age in
-each region for the northern component in the two stock model, we used
-the equilibrium assumption described above. Using this assumption two
-parameters are estimated for each stock component: an initial
-recruitment and an equilibrium fully-selected \(F\).
-
-\hypertarget{recruitment-and-survival}{%
-\subsection*{Recruitment and Survival}\label{recruitment-and-survival}}
-\addcontentsline{toc}{subsection}{Recruitment and Survival}
+models, but for black sea bass the available data cannot distinguish the
+proportion of northern and southern component fish at each age in the
+southern region in the initial year of the model. Therefore, we used the
+simplifying equilibrium assumption described above where there are two
+parameters estimated for each stock component: an initial recruitment
+and an equilibrium fully-selected \(F\) that determines the abundance at
+age in each region for each stock component.
 
@@
 small variance for the transitions of northern fish in the southern
-region, which is essentially the same as the deterministic transition
+region, which is approximately the same as the deterministic transition
 assumptions of a statistical catch at age model. We also allow 2DAR1
@@
 
-The CVs provided by the analyses for the recreational catch per angler
-(RecCPA) ranged between 0.02 and 0.06 which the working group felt did
-not capture the true uncertainty in the index with regard to its
+The estimated coefficients of variation (CVs) provided by the analyses
+to generate the recreational catch per angler (recreational CPA) index
+ranged between 0.02 and 0.06 which the working group felt did not
+capture the true observation uncertainty in the index with regard to its
 relationship to stock abundance. In many of the initial runs as well as
@@
 
-Table \ref{tab:age-comp-sel-table}
+The working group investigated many alternative assumptions for the
+probability models and selectivity models for the 8 different sets of
+age composition observations to reduce residual patterns and
+retrospective patterns. These analyses resulted in use of selectivity
+random effects for the northern fleet and indices and logistic-normal
+likelihoods for 6 sets of of age composition observations and
+Dirichlet-multinomial likelihoods for one index and one fleet in the
+northern region (Table \ref{tab:age-comp-sel-table}).
 
@@
 retained the assumption from the peer-reviewed assessment that treated
-the latent bottom temperature covariates in each region as first-order
-autoregressive processes.
+the latent bottom temperature covariates in each region as AR1
+processes.
 
@@
 deviations from a mean model and all models here treat recruitment
-deviations as first-order autoregressive random effects. However, Miller
-et al. (2018) showed how inferences of temperature effects on growth or
-maturity parameters can be very different whether the null model without
-the effect and models with effects also include random effects
-representing residual temporal variation in parameters. Therefore, we
-also explored whether including temporal random effects on age 1 natural
-mortality M affected inferences on corresponding temperature effects.
-Initially, we included random effects on age 1 natural mortality for
-both the northern and southern stock components, but estimates of these
-random effects and corresponding variance for the southern component
-converged to 0 so these were not included in models presented here.
+deviations as AR1 random effects. However, Miller et al. (2018) showed
+how inferences of temperature effects on growth or maturity parameters
+can be very different whether the compared models with and without the
+effect also include random effects representing residual temporal
+variation in parameters. Therefore, we also explored whether including
+temporal random effects on age 1 natural mortality affected inferences
+on corresponding temperature effects. Initially, we included random
+effects on age 1 natural mortality for both the northern and southern
+stock components, but estimates of these random effects and
+corresponding variance for the southern component converged to 0 so
+these were not included in models presented here.
 
@@
 possibly in fishery catches which are accumulated over the whole year.
-Expected log-recruitment is a linear function of log-recruitment at the
-previous time step and bottom temperature
+Expected log-recruitment for a given stock is a linear function of
+bottom temperature
 
 \begin{equation}\label{eq:expected-recruitment}
-\log N_{s,r_s,y,1} = \mu_{R,s} + \beta_{R,s} x_y + \epsilon_{N,s,y,1}
-\end{equation}
-
-\begin{equation}\label{eq:expected-M1}
-\log M_{s,y,1} = \mu_{M,s,1} + \beta_{M,s} x_y + \epsilon_{M,s,y}
-\end{equation}
-
-\hypertarget{model-fitting-diagnostics-and-comparisons}{%
+E\left(\log N_{y,1}|x_y\right) = \mu_{R} + \beta_{R} x_y.
+\end{equation} Similarly, expected log-natural morality as a function of
+bottom temperature is \begin{equation}\label{eq:expected-M1}
+E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y
+\end{equation} Because age 1 fish for the northern component can exist
+in both regions after January 1, natural mortality is acting in both
+regions for this stock. For models with covariate and/or annual random
+effects for age 1 fish we assume them only in the northern regions for
+the northern component. The corresponding random effects are
+\begin{equation}\label{eq:Rec-re}
+\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \epsilon_{y,1}
+\end{equation} and \begin{equation}\label{eq:M-re}
+\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \epsilon_{M,y}.
+\end{equation} The constant or mean log-natural mortality rate is
+assumed to be \(\mu_{M,a} = \log(0.4)\) for all ages as recommended by
+the working group. Because the bottom temperature anomalies and random
+effects are centered at 0, mean log-natural mortality at age 1 over the
+time series should be approximately equal to \(\mu_{M,1}\) for models
+that include those effects.
+
+\hypertarget{model-fitting-diagnostics-and-projections}{%
 \subsection*{Model fitting, diagnostics, and
-comparisons}\label{model-fitting-diagnostics-and-comparisons}}
+projections}\label{model-fitting-diagnostics-and-projections}}
 \addcontentsline{toc}{subsection}{Model fitting, diagnostics, and
-comparisons}
-
-As in Miller et al. (2016b), we also assessed the stability of the
+projections}
+
+We used a development version
+(\href{https://github.com/timjmiller/wham/tree/fb8b089}{commit fb8b089})
+of the WHAM package prior to the release of version 2.0 for all results.
+All code for fitting models and generating results can be found at
+\href{https://github.com/timjmiller/wham_bsb_paper}{github.com/timjmiller/wham\_bsb\_paper}.
+
+We examined retrospective patterns for all models by fitting
+corresponding models where the terminal year is reduced sequentially by
+one year (peel) for seven years. Therefore, there are eight fits of each
+model with the time series reduced by zero to seven years. We calculated
+Mohn's \(\rho\) for SSB, and average \(F\) at ages 6 and 7. Absolute
+values of Mohn's \(\rho\) near 0 imply no pattern in estimation of these
+quantities as the time-series is sequentially extended.
+
+As in Miller et al. (2016b), we also assessed the consistency of the
 AIC-based model selection over retrospective peels to guard against
@@
 
-We examined retrospective patterns by fitting models where the terminal
-year is reduced sequentially by one year (peel) for seven years.
-Therefore, there are 7 fits of the proposed base model with the time
-series reduced by one to seven years. We calculated Mohn's \(\rho\) for
-recruitment, SSB, and fully-selected \(F\). Absolute values of Mohn's
-\(\rho\) near 0 imply no pattern in estimation of these quantities as
-the time-series is sequentially extended.
-
-Specifically, we use version XX and commit XXX of the wham package
-(\url{https://github.com/timjmiller/wham/tree/lab}) for all results.
-
-We performed a simulation self-test where new observations were
-simulated conditional on all random effects estimated in the proposed
-base model and the same model configuration was fit to each of the
-simulated data sets.
-
-One-step-ahead residuals can now be calculated for all index, catch,
-environmental covariate observations using methods described by Thygesen
-et al. (2017) and Trijoulet et al. (2023).
-
-\hypertarget{one-step-ahead-residuals}{%
-\subsubsection*{One-step-ahead
-residuals}\label{one-step-ahead-residuals}}
-\addcontentsline{toc}{subsubsection}{One-step-ahead residuals}
-
-\hypertarget{retrospective-patterns}{%
-\subsubsection*{Retrospective patterns}\label{retrospective-patterns}}
-\addcontentsline{toc}{subsubsection}{Retrospective patterns}
+We performed a jitter analysis of the base model \(M_0\) and the best
+fitting model to investigate whether a local mimimum of the negative
+log-likelihood surface was obtained by the optimization. We used the
+jitter\_wham function in the WHAM package which by default simulates
+starting values from a multivariate normal distribution with mean and
+covariance defined by the MLEs and estimated covariance matrix from the
+fitted model. See the Supplementary Materials for more details.
+
+For the best performing model, we also performed so-called simulation
+self-tests where new observations were simulated conditional on all
+estimated fixed and random effects and the same model configuration was
+fit to each of the simulated data sets. We estimated median relative
+bias of SSB across these simulations. See the Supplementary Materials
+for more details.
+
+To illustrate projections with environmental effects we projected the
+best black sea bass model under three alternative scenarios where the
+AR1 time-series model for bottom temperature continues into the
+projection years, where we project with bottom temperature being the
+average of the most recent 5 years, and where bottom temperature
+increases following a predictions from a simple linear regression of the
+estimated bottom temperature anomalies over time.
 
@@
 
-Although there was strong retrospective patterns in the most recent
-management track assessment for the northern component of the stock,
-there was no evidence of patterns for northern SSB (Mohn's \(\rho\) =
--0.058) and \(F\) (Mohn's \(\rho\) = 0.06) for the proposed base model
-(Figures \ref{fig:North-retro-ssb} and \ref{fig:North-retro-F}).
-Similarly, no patterns were exhibited for the southern component of SSB
-(Mohn's \(\rho\) = -0.019) and regional \(F\) (Mohn's \(\rho\) = -0.05)
-(Figures \ref{fig:South-retro-ssb} and \ref{fig:South-retro-F}).
-
-For 7 of the the simulated data sets the model failed to optimize. The
-maximum absolute gradient was \(<10^{-6}\) for only 9 and \(<10^{-4}\)
-for 52 of the 93 successfully fitted models. The poor convergence
-appeared to be attributable to the estimation of the scalar for the
-standard errors of the log-transformed northern Recreational CPA index
-for which estimates tended to 0 for nearly all of the fits (\(<0.01\)
-for 83 fits). However, even across all fits including those with poor
-convergence, the SSB estimates appeared to be reliable (Figure
-\ref{fig:self-test-res-1}).
-
-\hypertarget{bottom-temperature-effect-on-recruitment}{%
-\subsection*{Bottom temperature effect on
-recruitment}\label{bottom-temperature-effect-on-recruitment}}
-\addcontentsline{toc}{subsection}{Bottom temperature effect on
-recruitment}
-
 We found the best model of bottom temperature covariate effects included
-the effect only on the recruitment of the northern stock, which is the
-configuration in the proposed base model although the difference in AIC
-for the model including effects on both regional components was small
-suggesting some evidence for this hypothesis as well (Table
-\ref{tab:compare-table}) . We also found AIC to strongly favor models
-that included at least an effect on the northern component across all
-retrospective peels (Table \ref{tab:peel-compare-table}). The posterior
-estimate of the bottom temperature covariate match the observations well
-because of the high precision of the observations (Figure
-\ref{fig:bottom-temp}). The residual variation in the standard deviation
-of recruitment random effects is reduced because the expected
-recruitment (Eq. \ref{eq:expected-recruitment}) is a function of the
-covariate (Figure \ref{fig:relative-recruitment-bt}). This effect is
-included in the proposed base model (Figure \ref{fig:rec-bottom-temp}).
-Because the covariate is technically for temperature after the beginning
-of the calendar year when the fish are considered age 1 in the model, a
-comparison of the proposed base model to one with the bottom temperature
-effect on natural mortality at age 1 instead might be of interest in the
-future. An investigation of higher order orthogonal polynomial effects
-might also be of interest particularly for the southern component which
-might be experiencing more frequently higher temperatures less favorable
-to overwinter survival.
+the effect only on the recruitment of the northern stock (Table
+\ref{tab:diff-aic-table}). However, the difference in AIC for the model
+that also included effects on recruitment for the southern component
+suggested some evidence for this hypothesis as well (Table
+\ref{tab:aic-wts-table}).
+
+Across retrospective peels where the terminal year of data was
+sequentially reduced, the ranking for the best model was consistent.
+However, when the terminal year was reduced 6 or more years, there was
+more evidence for further effects of temperature on recruitment for the
+southern component and for temporal variation in natural mortality at
+age 1 (Table \ref{tab:aic-wts-table}).
+
+The retrospective model fits did not indicate any evidence of patterns
+for northern or southern component SSB (all Mohn's
+\(\rho \approx -0.03\)) or fishing mortality (all Mohn's
+\(\rho \approx 0.03\) and \(-0.04\) for average \(F\) at ages 6 and 7 in
+north and south regions, respectively) for any of the models (Table
+\ref{tab:rho-table}).
+
+\hypertarget{bottom-temperature-effects-1}{%
+\subsection*{Bottom temperature
+effects}\label{bottom-temperature-effects-1}}
+\addcontentsline{toc}{subsection}{Bottom temperature effects}
+
+The posterior estimate of the bottom temperature covariate match the
+observations well because of the high precision of the observations and
+anomalies in the north and south region appear highly correlated (Figure
+\ref{fig:bottom-temperature}). Because the temperature anomalies are
+treated as latent variables, when effects on recruitment are included,
+other data components in the model can affect the estimated anomalies
+(e.g., Miller et al. 2018), but in this case the estimates are altered
+negligibly during the years where recruitment is estimated (Figure
+\ref{fig:Ecov-M1-rel-M0}).
+
+Estimated effects of bottom temperature on recruitment for either the
+northern or southern component were stable over the retrospective peels
+and differed negligibly whether effects for each component were
+estimated in isolation or together (Table
+\ref{tab:beta-sig-peel-table}). Estimates of residual variability in
+northern component recruitment, as measured by the conditional or
+marginal standard deviation of the recruitment random effects, increased
+slightly with the number of peeled years. However, the ratio of standard
+deviations of models with and without temperature effects was stable
+with about a 20\% reduction in residual standard deviation when
+temperature effects were included.
+
+Although evidence for temperature effects on age 1 natural mortality was
+weak, it is notable that the sign of the estimated effect for the
+northern stock component differed depending on whether random effects on
+age 1 natural mortality were also included (\(\widehat \beta_R = -1\)
+for \(M_{6}\) or \(M_{13}\)) with the negative effect estimated with
+these random effects (\(M_{13}\)).
+
+\hypertarget{stock-size-fishing}{%
+\subsection*{Stock size, fishing}\label{stock-size-fishing}}
+\addcontentsline{toc}{subsection}{Stock size, fishing}
+
+For the best performing model \(M_1\), estimates of SSB Stock size is
+also increasing for the southern component, just not as quickly as the
+northern component. (plot of SSB time series for each component).
+
+Figure \ref{fig:M1-SSB-F-R}
+
+Figure \ref{fig:SSB-F-R-rel-M1} Annual estimates of SSB and F varied
+little among the 14 fitted models.
+
+plot of prior/posterior for movement rate.
+
+plot of average F by fleet,region.
+
+Sectivity random effects (Figure \ref{fig:selectivity-re})
+
+Estimates of effect size, sd, correlation for each peel (Table \ref{}
+
+The residual variation in the standard deviation of recruitment random
+effects is reduced because the expected recruitment (Eq.
+\ref{eq:expected-recruitment}) is a function of the covariate. This
+effect is included in the proposed base model (Figure
+\ref{fig:BT-Ecov-R}).
+
+plot of northern component predicted recruitment vs.~time with color by
+bottom temperature
+
+\hypertarget{reference-points-1}{%
+\subsection*{Reference points}\label{reference-points-1}}
+\addcontentsline{toc}{subsection}{Reference points}
+
+\hypertarget{projections-1}{%
+\subsection*{Projections}\label{projections-1}}
+\addcontentsline{toc}{subsection}{Projections}
+
+Projections of bottom temperature and recruitment and CVs (Figure
+\ref{fig:R-BT-proj})
+
+Uncertainty in both is much lower during the data years than the
+projection years.
+
+Uncertainty in BT RE propagates into both expected recruitment and
+estimated random effect for recruitment, but the latter also includes
+uncertainty due to the residual variability in recruitment.
 
@@
 
-The new features of Multi-WHAM include
-
-\begin{itemize}
-\tightlist
-\item
-  seasonal intervals within years
-\item
-  variation in movement rates by stock, region-to-region, season, age,
-  and year
-\item
-  effects of environmental covariates on movement rates by stock,
-  region-to-region, season, and age
-\item
-  effects of environmental covariates on mortality rates by stock,
-  region, and age
-\item
-  effects of environmental covariates on recruitment by stock
-\item
-  stock-specific stock-recruitment models
-\item
-  priors for movement rates
-\item
-  seasonal operation of fleets
-\item
-  mortality and movement modeled sequentially or simultaneously
-\item
-  options for weighting of stock-specific SSB/R for global SPR-based
-  reference points
-\end{itemize}
+\hypertarget{general-model-aspects}{%
+\subsection*{General Model aspects}\label{general-model-aspects}}
+\addcontentsline{toc}{subsection}{General Model aspects}
+
+WHAM provides a comprehensive treatment of environmental covariates and
+their effects on populations using state-space methods (Maunder 2024).
+This framework accounts for the magnitude of differing uncertainties in
+observations and stochastic population dynamics processes. This version
+2.0 extends these inferences to movement rates for multi-region models.
+
+\hypertarget{future-developments}{%
+\subsection*{Future developments}\label{future-developments}}
+\addcontentsline{toc}{subsection}{Future developments}
+
+An obvious limitation to WHAM is the inability to include tagging
+observations of any type. Such observations are critical to estimation
+of movement parameters (e.g., Goethel et al. 2019), but also can inform
+mortality rate parameters (Hampton 1991). It is well known that natural
+mortality is seldom estimated in assessment models because the observed
+data often provide little information to distinguish natural mortality
+from other assessment model parameters (Lee et al. 2011; Clark 2022).
+Estimation of natural mortality may be even more challenging within
+state-space assessment models with their greater flexibility from
+inclusion of time-varying random effects. For example, we fixed the mean
+natural mortality for age 1 fish and Cadigan (2016) and Stock et al.
+(2021) also estimated natural mortality deviations. Therefore allowing
+tagging observations in WHAM should be a high priority even for models
+with a single stock and region.
+
+The work by Correa et al. (2023) on incorporating length information and
+modeling growth within WHAM has not yet been merged into the multi-stock
+version 2.0. Growth and movement would be challenging.
+
+\hypertarget{black-sea-bass}{%
+\subsection*{Black sea bass}\label{black-sea-bass}}
+\addcontentsline{toc}{subsection}{Black sea bass}
+
+In our investigation of bottom temperature effects on recruitment and
+natural mortality and time-varying random effects on natural mortality
+at age 1, we found only including effects on recruitment for the
+northern stock component to be the best model with respect to AIC. This
+is the same configuration configuration accepted during the assessment
+peer-review process and it is the model currently used for management
+(NEFSC 2023).
+
+Although not relevant to the best performing model, the effect of bottom
+temperature on age 1 M would be expected to be opposite of that for
+recruitment because lower M with higher temperature would produce higher
+recruitment. This expected effect was estimated only when random effects
+on age 1 M were also included in the model. This seems to provide
+further support to including random effects when examining covariate
+effects on population parameters as demonstrated by Miller et al. (2018)
+for effects on growth parameters.
+
+AIC suggested some weight for models with temperature effects on
+recruitment for the southern component and random effects on M at age 1.
+This might suggest making inferences on the population by using an
+ensemble of these and weighting by AIC (Burnham and Anderson 2002).
+However, the estimates of assessment outputs relevant for management
+(SSB, F) were similar among the models and weighted estimates would
+differ very little from the estimates with the lowest AIC. {[}but what
+about reference points?{]}
+
+\hypertarget{temperature-effects}{%
+\subsubsection*{Temperature effects}\label{temperature-effects}}
+\addcontentsline{toc}{subsubsection}{Temperature effects}
+
+BSB in the NEUS is at the northern extent of its range. hence the range
+extension to the north with increased temperature. The evidence for
+temperature effects on recruitment was strong only for the northern
+component which is the component of the stock at the limit of the
+species' range. Opposite effects of temperature on recruitment or
+population size would be expected for species in the same general area
+that are at the southern extent of their range (Gabriel 1992). For
+example, higher recruitment of the Southern New England-Mid-Atlantic
+stock of yellowtail flounder is correlated with more cold water
+persistence into summer fall on the Northeast US shelf Miller et al.
+(2016b).
+
+\hypertarget{self-tests}{%
+\subsubsection*{Self Tests}\label{self-tests}}
+\addcontentsline{toc}{subsubsection}{Self Tests}
+
+Simulation self tests performed by Stock and Miller (2021) simulated
+process errors and observation errors and their simulations showed
+little bias in estimated assessment output such as SSB. Li et al. (2024)
+used the same type of simulations and found what?.
+
+Here we conditioned on the estimated process errors for black sea bass.
+This conditional approach has been suggested as appropriate for tactical
+management where we are interested in inferences assuming our estimation
+of the population and fishing history from the original data is the
+truth Cadigan et al. (2024). We found some bias in SSB estimation for
+the black sea bass application (negatively for the northern stock),
+using the default model. We also found instability in the estimation of
+the observation variance parameters for the aggregate indices such that
+they often were estimated at the lower bound of 0. However, we found
+that the magnitude of bias was largely influenced by the large
+uncertainty estimated in some of the age composition observations
+assumed to have a logistic-normal distribution. When using the
+logistic-normal likelihood for management, we recommend inspecting the
+size of the estimated (marginal) standard deviation of the logistic
+normal model when self-test bias is non-negligible. We also recommend
+investigating the multiplicative transformation for the logistic normal
+rather than the additive transformation which is currently the only
+option in WHAM (Cadigan 2016).
+
+?Selectivity RE are only on index age comp in the north which may be why
+bias is larger there?
+
+? We found REML provided less biased estimation of SSB and recruitment
+which are functions of random effects. Whether REML should be the
+default approach for estimating these quantities from state-space stock
+assessment models requires further investigation. For example,
+improvement in bias may be negligible when data contain low observation
+error and/or there are few fixed effects parameters. ? larger variance
+estimates allows greater spread of random effects estimates. Important
+for conditional self-test where estimated RE are of primary importance
+(Recruitment and SSB). When simulating RE and refitting, the sign of
+differences between estimates and true values can be random across
+simulations. Probably more important with more fixed effects and less
+informative data
+
+Including tagging data in WHAM would also be helpful for BSB because we
+are using the same data to estimate movement outside of the model as we
+are using to estimate the other parameters in WHAM.
+
+\hypertarget{conclusions}{%
+\subsection*{Conclusions}\label{conclusions}}
+\addcontentsline{toc}{subsection}{Conclusions}
+
+WHAM version 2.0 extends the existing R package to allow multiple stocks
+and discrete spatial regions. It allows autocorrelated random effects
+and environmental covariate effects on movement rates. Estimates of
+movement rates from auxiliary studies can be used to construct prior
+distributions for movement rates in lieu of integrated likelihoods for
+tagging data. Version 2.0 also allows seasonal treatment of different
+fishing fleets and movement of individual stocks, and internal
+estimation of SPR- and MSY-based reference points accounting for
+seasonal spatial dynamics including movement and mortality and any
+environmental covariate effects.
+
+We applied this extended version to investigate alternative hypotheses
+about effects of bottom temperature on recruitment and age 1 natural
+mortality for black sea bass. Our analyses indicate evidence for effects
+of bottom temperature on recruitment for the northern component
+represents was stronger than other models that included effects on age 1
+mortality and/or corresponding effects on the southern stock component.
+
+Future WHAM development should prioritize including tagging data to
+inform stock assessment parameter estimation and merging in the version
+created by Correa et al. (2023) to allow modeling of growth and
+inclusion of length and length-at-age composition observations.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-andersenkeiding02}{}}%
+Andersen, P.K., and Keiding, N. 2002. Multi-state models for event
+history analysis. Statistical Methods in Medical Research \textbf{11}:
+91--115.
+
+\leavevmode\vadjust pre{\hypertarget{ref-arnason72}{}}%
+Arnason, A.N. 1972. Parameter estimates from mark-recapture experiments
+on two populations subject to migration and death. Researches on
+Population Ecology \textbf{13}: 97--113.
+doi:\href{https://doi.org/10.1007/BF02521971}{10.1007/BF02521971}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bosleyetal22}{}}%
+Bosley, K.M., Schueller, A.M., Goethel, D.R., Hanselman, D.H., Fenske,
+K.H., Berger, A.M., Deroba, J.J., and Langseth, B.J. 2022. Finding the
+perfect mismatch: Evaluating misspecification of population structure
+within spatially explicit integrated population models. Fish and
+Fisheries \textbf{23}(2): 294--315.
+doi:\href{https://doi.org/10.1111/faf.12616}{10.1111/faf.12616}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-brooks24}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-burnhamanderson02}{}}%
+Burnham, K.P., and Anderson, D.R. 2002. Model selection and multimodel
+inference: A practical information-theoretic approach. Springer-Verlag,
+New York.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cadigan16}{}}%
+Cadigan, N.G. 2016. A state-space stock assessment model for northern
+cod, including under-reported catches and variable natural mortality
+rates. Canadian Journal of Fisheries and Aquatic Sciences
+\textbf{73}(2): 296--308.
+doi:\href{https://doi.org/10.1139/cjfas-2015-0047}{10.1139/cjfas-2015-0047}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cadiganetal24}{}}%
+Cadigan, N.G., Albertsen, C.M., Zheng, N., and Nielsen, A. 2024. Are
+state-space stock assessment model confidence intervals accurate? Case
+studies with SAM and barents sea stocks. Fisheries Research
+\textbf{272}: 106950.
+doi:\href{https://doi.org/10.1016/j.fishres.2024.106950}{10.1016/j.fishres.2024.106950}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-caoetal14}{}}%
+Cao, J., Truesdell, S.B., and Chen, Y. 2014. Impacts of seasonal stock
+mixing on the assessment of {A}tlantic cod in the {G}ulf of {M}aine.
+ICES Journal of Marine Science \textbf{71}(6): 1443--1457.
+doi:\href{https://doi.org/10.1093/icesjms/fsu066}{10.1093/icesjms/fsu066}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-clark22}{}}%
+Clark, W.G. 2022. Why natural mortality is estimable, in theory if not
+in practice, in a data-rich stock assessment. Fisheries Research
+\textbf{248}: 106203.
+doi:\href{https://doi.org/10.1016/j.fishres.2021.106203}{10.1016/j.fishres.2021.106203}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-commenges99}{}}%
+Commenges, D. 1999. Multi-state models in epidemiology. Lifetime Data
+Analysis \textbf{5}: 315--327.
+
+\leavevmode\vadjust pre{\hypertarget{ref-correaetal23}{}}%
+Correa, G.M., Monnahan, C.C., Sullivan, J.Y., Thorson, J.T., and Punt,
+A.E. 2023. Modelling time-varying growth in state-space stock
+assessments. ICES Journal of Marine Science \textbf{80}(7): 2036--2049.
+doi:\href{https://doi.org/10.1093/icesjms/fsad133}{10.1093/icesjms/fsad133}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-francis14}{}}%
@@
 doi:\href{https://doi.org/10.1016/j.fishres.2013.12.015}{10.1016/j.fishres.2013.12.015}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gabriel92}{}}%
+Gabriel, W.L. 1992. Persistence of demersal fish assemblages between
+{C}ape {H}atteras and {N}ova {S}cotia, {N}orthwest {A}tlantic. Journal
+of the Northwest Atlantic Fisheries Science \textbf{14}(1): 29--46.
+
+\leavevmode\vadjust pre{\hypertarget{ref-goetheletal19}{}}%
+Goethel, D.R., Bosley, K.M., Hanselman, D.H., Berger, A.M., Deroba,
+J.J., Langseth, B.J., and Schueller, A.M. 2019. Exploring the utility of
+different tag-recovery experimental designs for use in spatially
+explicit, tag-integrated stock assessment models. Fisheries Research
+\textbf{219}: 105320.
+doi:\href{https://doi.org/10.1016/j.fishres.2019.105320}{10.1016/j.fishres.2019.105320}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hampton91}{}}%
+Hampton, J. 1991. Estimation of southern bluefin tuna \emph{{T}hunnus
+maccoyii} natural mortality and movement rates from tagging experiments.
+Fishery Bulletin \textbf{89}(4): 591--610.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hearnetal87}{}}%
+Hearn, W.S., Sundland, R.L., and Hampton, J. 1987. Robust estimation of
+the natural mortality rate in a completed tagging experiment with
+variable fishing intensity. Journal Du Conseil International Pour
+L'exploration De La Mer \textbf{43}: 107--117.
+doi:\href{https://doi.org/10.1093/icesjms/43.2.107}{10.1093/icesjms/43.2.107}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-leeetal11}{}}%
+Lee, H.-H., Maunder, M.N., Piner, K.R., and Methot, R.D. 2011.
+Estimating natural mortality within a fisheries stock assessment model:
+An evaluation using simulation analysis based on twelve stock
+assessments. Fisheries Research \textbf{109}(1): 89--94.
+doi:\href{https://doi.org/10.1016/j.fishres.2011.01.021}{10.1016/j.fishres.2011.01.021}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-legaultrestrepo99}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-legaultetal23}{}}%
-Legault, C.M., Wiedenmann, J., Deroba, J.J., Fay, G., Miller, T.J.,
-Brooks, E.N., Bell, R.J., Langan, J., Cournane, J., Jones, A.W., and
-Muffley, B. 2023. Data rich but model resistant: An evaluation of
-data-limited methods to manage fisheries with failed age-based stock
-assessments. Canadian Journal of Fisheries and Aquatic Sciences
-\textbf{80}(1): 27--42.
-doi:\href{https://doi.org/10.1139/cjfas-2022-0045}{10.1139/cjfas-2022-0045}.
+\leavevmode\vadjust pre{\hypertarget{ref-lietal24}{}}%
+Li, C., Deroba, J.J., Miller, T.J., Legault, C.M., and Perretti, C.T.
+2024. An evaluation of common stock assessment diagnostic tools for
+choosing among state-space models with multiple random effects
+processes. Fisheries Research \textbf{273}: 106968.
+doi:\href{https://doi.org/10.1016/j.fishres.2024.106968}{10.1016/j.fishres.2024.106968}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-maunder24}{}}%
+Maunder, M.N. 2024. Towards a comprehensive framework for providing
+management advice from statistical inference using population dynamics
+models. Ecological Modelling \textbf{498}: 110836.
+doi:\href{https://doi.org/10.1016/j.ecolmodel.2024.110836}{10.1016/j.ecolmodel.2024.110836}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-milleretal16_yoy_survival}{}}%
-Miller, A.S., Shepherd, G.R., and Fratantoni, P.S. 2016a. {PLOS} {ONE}
-\textbf{11}(1): e0147627.
+Miller, A.S., Shepherd, G.R., and Fratantoni, P.S. 2016a. Offshore
+habitat preference of overwintering juvenile and adult black sea bass,
+\emph{{C}entropristis} \emph{striata}, and the relationship to
+year-class success. {PLOS} {ONE} \textbf{11}(1): e0147627.
 doi:\href{https://doi.org/10.1371/journal.pone.0147627}{10.1371/journal.pone.0147627}.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-nefsc22}{}}%
+NEFSC. 2022a. Final report of the haddock research track assessment
+working group. {Available} at
+https://s3.us-east-1.amazonaws.com/nefmc.org/14b\_EGB\_Research\_Track\_Haddock\_WG\_Report\_DRAFT.pdf.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nefsc22a}{}}%
+NEFSC. 2022b. Report of the american plaice research track working
+group. {Available} at
+https://s3.us-east-1.amazonaws.com/nefmc.org/2\_American-Plaice-WG-Report.pdf.
+
 \leavevmode\vadjust pre{\hypertarget{ref-nefsc23}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-nefsc24}{}}%
+NEFSC. 2024. Butterfish research track assessment report. US Dept Commer
+Northeast Fish Sci Cent Ref Doc. 24-03; 191 p.
+
 \leavevmode\vadjust pre{\hypertarget{ref-nielsenberg14}{}}%
@@
 doi:\href{https://doi.org/10.1016/j.fishres.2014.01.014}{10.1016/j.fishres.2014.01.014}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-perreaultetal20}{}}%
+Perreault, A.M.J., Wheeland, L.J., Morgan, M.J., and Cadigan, N.G. 2020.
+A state-space stock assessment model for {American} plaice on the
+{Grand} {Bank} of {Newfoundland}. Journal of Northwest Atlantic Fishery
+Science \textbf{51}: 45--104.
+doi:\href{https://doi.org/10.2960/j.v51.m727}{10.2960/j.v51.m727}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-puntetal20}{}}%
+Punt, A.E., Dunn, A., Elvarsson, B., Hampton, J., Hoyle, S.D., Maunder,
+M.N., Methot, R.D., and Nielsen, A. 2020. Essential features of the
+next-generation integrated fisheries stock assessment package: A
+perspective. Fisheries Research \textbf{229}: 105617.
+doi:\url{https://doi.org/10.1016/j.fishres.2020.105617}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-schwarzetal93}{}}%
+Schwarz, C.J., Schweigert, J.F., and Arnason, A.N. 1993. Estimating
+migration rates using tag-recovery data. Biometrics \textbf{49}:
+177--193. doi:\href{https://doi.org/10.2307/2532612}{10.2307/2532612}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-sullivanetal05}{}}%
+Sullivan, M.C., Cowen, R.K., and Steves, B.P. 2005. Evidence for
+atmosphere{--}ocean forcing of yellowtail flounder (\emph{{L}imanda}
+\emph{ferruginea}) recruitment in the {M}iddle {A}tlantic {B}ight.
+Fisheries Oceanography \textbf{14}(5): 386--399.
+
 \leavevmode\vadjust pre{\hypertarget{ref-thorsonetal17}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-varkeyetal22}{}}%
+Varkey, D.A., Babyn, J., Regular, P., Ings, D.W., Kumar, R., Rogers, B.,
+Champagnat, J., and Morgan, M.J. 2022. A state-space model for stock
+assessment of cod (gadus morhua) stock in NAFO subdivision 3Ps. {DFO}
+{Can.} {Sci.} {Advis.} {Sec.} {Res.} {Doc.} 2022/022. v + 78 p.
+
+\leavevmode\vadjust pre{\hypertarget{ref-yingetal11}{}}%
+Ying, Y., Chen, Y., Lin, L., and Gao, T. 2011. Risks of ignoring fish
+population spatial structure in fisheries management. Canadian Journal
+of Fisheries and Aquatic Sciences \textbf{68}(12): 2101--2120.
+doi:\href{https://doi.org/10.1139/f2011-116}{10.1139/f2011-116}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-zhengcadigan23}{}}%
+Zheng, N., and Cadigan, N. 2023. Frequentist conditional variance for
+nonlinear mixed-effects models. Journal of Statistical Theory and
+Practice \textbf{17}(3).
+doi:\href{https://doi.org/10.1007/s42519-022-00304-5}{10.1007/s42519-022-00304-5}.
+
 \end{CSLReferences}
@@
 
+\hypertarget{figures}{%
+\section*{Figures}\label{figures}}
+\addcontentsline{toc}{section}{Figures}
+
 \begin{figure}
@@
 \end{figure}
+\pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{move_prior_post}
+
+}
+
+\caption{Prior (black) and posterior (red) distributions of movement of northern component stock from north to south (top) and south to north (bottom).}\label{fig:move-prior-posterior}
+\end{figure}
+\pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{Ecov}
+
+}
+
+\caption{Observations and 95\% confidence intervals (points with vertical lines) and posterior estimates (lines) with 95\% confidence intervals (polygons) of bottom temperature anomalies in the north and south regions from model $M_0$.}\label{fig:bottom-temperature}
+\end{figure}
+\pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{best_R_Ecov}
+
+}
+
+\caption{Expected and Posterior Emperical Bayes estimates of annual recruitment for the northern stock component.
+Color defined by the corresponding annual bottom temperature anomaly.}\label{fig:BT-Ecov-R}
+\end{figure}
+\pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{M_1_SSB_F_R}
+
+}
+
+\caption{Annual estimates of SSB, average F (ages 6-7), and recruitment from model $M_1$. Polygons represent 95\% confidence intervals.}\label{fig:M1-SSB-F-R}
+\end{figure}
+\pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{R_proj_results}
+
+}
+
+\caption{Estimates of bottom temperature, recruitment, and expected recruitment, and coefficients of variation for the latter two from model $M_1$ assuming alternative bottom temperature scenarios described in the text.
+Polygons represent 95\% confidence intervals.}\label{fig:R-BT-proj}
+\end{figure}
+
+\pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{selectivity_re_plot}
+
+}
+
+\caption{Time and age-varying selectivty for fleets and indices in the northern region with autoregressive random effects.}\label{fig:selectivity-re}
+\end{figure}
+\pagebreak
+
+\clearpage
+
+\hypertarget{tables}{%
+\section*{Tables}\label{tables}}
+\addcontentsline{toc}{section}{Tables}
 
@@
 \renewcommand\thetable{\arabic{table}}
-
-\begin{landscape}\begin{table}
-
-\caption{\label{tab:age-comp-sel-table}Configuration of Age composition likelihood, mean selectivity model, and selectivity random effects for each age composition data component.}
-\centering
-\begin{tabular}[t]{llll}
-\toprule
-Data component & Age Composition Likelihood & Mean Selectivity model & Random effects configuration\\
-\midrule
-North Commercial & Dirichlet-Multinomial & age-specific (flat-topped at ages > 3) & 2D-AR1 (age and year)\\
-North Recreational & Logistic-normal (0s as missing) & age-specific (flat-topped at ages > 6) & 2D-AR1 (age and year)\\
-South Commercial & Logistic-normal (AR1, 0s as missing) & logistic & None\\
-South Recreational & Logistic-normal (AR1, 0s as missing) & logistic & None\\
-North Recreational CPA & Logistic-normal (0s as missing) & age-specific (flat-topped at ages > 1) & AR1 (year)\\
-\addlinespace
-North VAST & Dirichlet-Multinomial & age-specific (flat-topped at ages > 4) & 2D-AR1 (age and year)\\
-South Recreational CPA & Logistic-normal (AR1, 0s as missing) & age-specific (flat-topped at ages > 2) & None\\
-South VAST & Logistic-normal (AR1, 0s as missing) & age-specific (flat-topped at ages > 1) & None\\
-\bottomrule
-\end{tabular}
-\end{table}
-\end{landscape}
 
@@
 \midrule
-$M_{0}$ & 14.53 & 16.09 & 13.78 & 13.69 & 13.79 & 17.02 & 8.47 & 8.22\\
+$M_{0}$ & 11.83 & 11.39 & 10.41 & 10.05 & 9.86 & 9.41 & 8.47 & 8.22\\
 $M_{1}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{2}$ & 15.29 & 16.77 & 14.80 & 14.71 & 14.85 & 18.11 & 9.20 & 9.22\\
-$M_{3}$ & 3.49 & 5.42 & 4.42 & 4.69 & 5.02 & 8.75 & 0.73 & 1.00\\
-$M_{4}$ & 16.51 & 18.05 & 15.63 & 15.45 & 15.63 & 18.79 & 10.08 & 9.85\\
+$M_{2}$ & 12.63 & 12.03 & 11.41 & 11.06 & 10.91 & 10.34 & 9.20 & 9.22\\
+$M_{3}$ & 0.80 & 0.64 & 1.00 & 1.01 & 1.05 & 0.93 & 0.73 & 1.00\\
+$M_{4}$ & 13.81 & 13.35 & 12.26 & 11.82 & 11.71 & 11.18 & 10.08 & 9.85\\
 \addlinespace
-$M_{5}$ & 16.09 & 17.51 & 15.30 & 15.16 & 15.29 & 18.41 & 9.86 & 9.71\\
-$M_{6}$ & 18.07 & 19.46 & 17.14 & 16.92 & 17.12 & 20.18 & 11.46 & 11.33\\
-$M_{7}$ & 17.05 & 18.48 & 15.57 & 14.96 & 15.39 & 18.19 & 9.71 & 9.42\\
-$M_{8}$ & 5.02 & 6.87 & 4.90 & 4.66 & 5.28 & 8.59 & 0.98 & 0.99\\
-$M_{9}$ & 17.82 & 19.17 & 16.59 & 15.98 & 16.45 & 19.28 & 10.44 & 10.43\\
+$M_{5}$ & 13.25 & 12.55 & 11.68 & 11.21 & 11.07 & 10.41 & 9.86 & 9.71\\
+$M_{6}$ & 15.22 & 14.51 & 13.52 & 12.97 & 12.91 & 12.17 & 11.46 & 11.33\\
+$M_{7}$ & 14.32 & 13.75 & 12.17 & 11.30 & 11.43 & 10.55 & 9.71 & 9.42\\
+$M_{8}$ & 2.25 & 2.10 & 1.47 & 0.96 & 1.29 & 0.91 & 0.98 & 0.99\\
+$M_{9}$ & 15.12 & 14.39 & 13.17 & 12.31 & 12.48 & 11.49 & 10.44 & 10.43\\
 \addlinespace
-$M_{10}$ & 5.78 & 7.56 & 5.92 & 5.68 & 6.34 & 9.68 & 1.71 & 1.99\\
-$M_{11}$ & 19.02 & 20.48 & 17.57 & 16.96 & 17.39 & 20.19 & 11.68 & 11.39\\
-$M_{12}$ & 18.61 & 19.90 & 17.09 & 16.43 & 16.88 & 19.58 & 11.10 & 10.92\\
-$M_{13}$ & 20.58 & 21.89 & 19.09 & 18.43 & 18.88 & 21.57 & 13.07 & 12.88\\
+$M_{10}$ & 3.05 & 2.74 & 2.47 & 1.97 & 2.34 & 1.84 & 1.71 & 1.99\\
+$M_{11}$ & 16.29 & 15.74 & 14.17 & 13.30 & 13.43 & 12.55 & 11.68 & 11.39\\
+$M_{12}$ & 15.73 & 14.91 & 13.43 & 12.46 & 12.64 & 11.55 & 11.10 & 10.92\\
+$M_{13}$ & 17.70 & 16.91 & 15.43 & 14.46 & 14.64 & 13.55 & 13.07 & 12.88\\
 \bottomrule
@@
 $M_{0}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.01 & 0.01\\
-$M_{1}$ & 0.76 & 0.89 & 0.80 & 0.80 & 0.83 & 0.97 & 0.36 & 0.38\\
+$M_{1}$ & 0.45 & 0.43 & 0.42 & 0.38 & 0.41 & 0.37 & 0.36 & 0.38\\
 $M_{2}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{3}$ & 0.13 & 0.06 & 0.09 & 0.08 & 0.07 & 0.01 & 0.25 & 0.23\\
+$M_{3}$ & 0.30 & 0.31 & 0.25 & 0.23 & 0.24 & 0.23 & 0.25 & 0.23\\
 $M_{4}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
@@
 $M_{7}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{8}$ & 0.06 & 0.03 & 0.07 & 0.08 & 0.06 & 0.01 & 0.22 & 0.23\\
+$M_{8}$ & 0.15 & 0.15 & 0.20 & 0.24 & 0.21 & 0.24 & 0.22 & 0.23\\
 $M_{9}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
 \addlinespace
-$M_{10}$ & 0.04 & 0.02 & 0.04 & 0.05 & 0.04 & 0.01 & 0.15 & 0.14\\
+$M_{10}$ & 0.10 & 0.11 & 0.12 & 0.14 & 0.13 & 0.15 & 0.15 & 0.14\\
 $M_{11}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
@@
 
-\caption{\label{tab:rho-table}Mohn's $\rho$ for SSB, R, average F at ages 6 and 7 in northern and southern regions.}
+\caption{\label{tab:rho-table}Mohn's $\rho$ for SSB, and average F at ages 6 and 7 in northern and southern regions.}
 \centering
-\begin{tabular}[t]{lrrrrrr}
+\begin{tabular}[t]{lrrrr}
 \toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{SSB} & \multicolumn{2}{c}{Average F} & \multicolumn{2}{c}{Recruitment} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
-Model & North & South & North & South & North & South\\
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{SSB} & \multicolumn{2}{c}{Average F} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+Model & North & South & North & South\\
 \midrule
-$M_{0}$ & -0.031 & -0.026 & 0.032 & -0.043 & 0.211 & -0.007\\
-$M_{1}$ & -0.040 & -0.023 & 0.041 & -0.048 & 0.181 & -0.024\\
-$M_{2}$ & -0.031 & -0.027 & 0.032 & -0.043 & 0.211 & -0.007\\
-$M_{3}$ & -0.031 & -0.027 & 0.031 & -0.043 & 0.193 & -0.007\\
-$M_{4}$ & -0.032 & -0.026 & 0.032 & -0.044 & 0.268 & -0.006\\
+$M_{0}$ & -0.040 & -0.023 & 0.041 & -0.048\\
+$M_{1}$ & -0.040 & -0.023 & 0.041 & -0.048\\
+$M_{2}$ & -0.040 & -0.024 & 0.041 & -0.047\\
+$M_{3}$ & -0.040 & -0.024 & 0.041 & -0.047\\
+$M_{4}$ & -0.041 & -0.022 & 0.042 & -0.048\\
 \addlinespace
-$M_{5}$ & -0.031 & -0.026 & 0.032 & -0.045 & 0.210 & -0.008\\
-$M_{6}$ & -0.032 & -0.025 & 0.032 & -0.045 & 0.267 & -0.007\\
-$M_{7}$ & -0.033 & -0.026 & 0.033 & -0.043 & 0.225 & -0.007\\
-$M_{8}$ & -0.035 & -0.027 & 0.035 & -0.043 & 0.186 & -0.007\\
-$M_{9}$ & -0.033 & -0.027 & 0.033 & -0.043 & 0.224 & -0.008\\
+$M_{5}$ & -0.041 & -0.023 & 0.042 & -0.048\\
+$M_{6}$ & -0.042 & -0.023 & 0.042 & -0.048\\
+$M_{7}$ & -0.042 & -0.022 & 0.043 & -0.048\\
+$M_{8}$ & -0.044 & -0.023 & 0.044 & -0.048\\
+$M_{9}$ & -0.042 & -0.024 & 0.043 & -0.048\\
 \addlinespace
-$M_{10}$ & -0.035 & -0.027 & 0.035 & -0.043 & 0.186 & -0.008\\
-$M_{11}$ & -0.032 & -0.026 & 0.032 & -0.044 & 0.253 & -0.007\\
-$M_{12}$ & -0.033 & -0.026 & 0.033 & -0.045 & 0.224 & -0.008\\
-$M_{13}$ & -0.032 & -0.025 & 0.032 & -0.045 & 0.253 & -0.008\\
+$M_{10}$ & -0.044 & -0.024 & 0.044 & -0.047\\
+$M_{11}$ & -0.041 & -0.022 & 0.042 & -0.048\\
+$M_{12}$ & -0.042 & -0.023 & 0.043 & -0.048\\
+$M_{13}$ & -0.041 & -0.023 & 0.042 & -0.048\\
 \bottomrule
@@
 \end{table}
+\begin{table}
+
+\caption{\label{tab:beta-sig-peel-table}Estimates of temperature effects on recruitment and variance and autocorrelation parameters for recruitment for northern ($N$) and southern ($S$) components for models with no effects ($M_0$), effects on specific components ($M_1$ and $M_2$) or both components simultaneously ($M_3$) from the full model (Peel 0) and each retrospective peel.}
+\centering
+\begin{tabular}[t]{lrrrrrrrr}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Peel} \\
+\cmidrule(l{3pt}r{3pt}){2-9}
+Parameter & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
+\midrule
+$M_{1}$ $\widehat{\beta}_{R,N}$ & 0.474 & 0.480 & 0.485 & 0.476 & 0.464 & 0.468 & 0.439 & 0.445\\
+$M_{2}$ $\widehat{\beta}_{R,S}$ & 0.099 & 0.105 & 0.094 & 0.095 & 0.092 & 0.096 & 0.099 & 0.091\\
+$M_{3}$ $\widehat{\beta}_{R,N}$ & 0.474 & 0.480 & 0.485 & 0.476 & 0.464 & 0.468 & 0.439 & 0.445\\
+$M_{3}$ $\widehat{\beta}_{R,S}$ & 0.099 & 0.105 & 0.094 & 0.095 & 0.092 & 0.096 & 0.099 & 0.091\\
+$M_{0}$ Conditional $\widehat{\sigma}_{R,N}$ & 0.925 & 0.953 & 0.978 & 0.988 & 0.978 & 1.010 & 0.981 & 1.007\\
+\addlinespace
+$M_{1}$ Conditional $\widehat{\sigma}_{R,N}$ & 0.730 & 0.752 & 0.780 & 0.786 & 0.775 & 0.801 & 0.785 & 0.800\\
+$M_{0}$ $\widehat{\rho}_{R,N}$ & 0.362 & 0.375 & 0.385 & 0.405 & 0.424 & 0.432 & 0.427 & 0.431\\
+$M_{1}$ $\widehat{\rho}_{R,N}$ & 0.296 & 0.307 & 0.334 & 0.373 & 0.394 & 0.406 & 0.428 & 0.429\\
+$M_{0}$ Marginal $\widehat{\sigma}_{R,N}$ & 0.992 & 1.028 & 1.060 & 1.080 & 1.081 & 1.120 & 1.084 & 1.117\\
+$M_{1}$ Marginal $\widehat{\sigma}_{R,N}$ & 0.764 & 0.791 & 0.827 & 0.848 & 0.843 & 0.877 & 0.869 & 0.885\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \clearpage
@@
 
+\hypertarget{diagnostics}{%
+\subsection*{Diagnostics}\label{diagnostics}}
+\addcontentsline{toc}{subsection}{Diagnostics}
+
+\hypertarget{jitter-fits-for-model-m_0}{%
+\subsubsection{\texorpdfstring{Jitter fits for model
+\(M_0\)}{Jitter fits for model M\_0}}\label{jitter-fits-for-model-m_0}}
+
+WHAM by default completes three newton steps after the stats::nlminb
+minimization function completes to reduce the gradient at the minimized
+NLL. However, this generally has negligible effects on model estimates
+and the NLL. To reduce computation time, we did not complete these
+newton steps when performing jitter fits of the model. Without the
+Newton steps, the maximum (absolute) gradient sizes are generally less
+than 0.01 for models that converge satisfactorily.
+
+The 50 jitter fits demonstrated that a local minimum was obtained for
+the original fit of model \(M_0\) (Figure \ref{fig:jitter-M0}). Some
+lower NLLs were obtained with unacceptable gradients, but a slightly
+lower NLL was found with a satisfactory gradient and with and invertible
+hessian. We therefore refit model \(M_0\) and all remaining models using
+the better parameter estimates as initial values.
+
+\hypertarget{jitter-fits-for-model-m_1}{%
+\subsubsection{\texorpdfstring{Jitter fits for model
+\(M_1\)}{Jitter fits for model M\_1}}\label{jitter-fits-for-model-m_1}}
+
+The 50 jitter fits gave no evidence of a better minimization of the NLL.
+Three lower NLLs were obtained, but with unacceptably large gradients
+(Figure \ref{fig:jitter-M1}). The largest differences in parameter
+estimates for these three jitters were for numbers at age and
+selectivity random effects variance and correlation parameters.
+
+\hypertarget{self-test-for-model-m_1}{%
+\subsubsection{\texorpdfstring{Self test for model
+\(M_1\)}{Self test for model M\_1}}\label{self-test-for-model-m_1}}
+
+Initial fits to simulated data from model \(M_1\) showed estimation of
+the observation error standard deviation multiplier for the recreational
+catch-per angler indices in the north and south regions was unstable.
+Many of the fits to the simulated data produced implausible estimates at
+the 0 boundary for these parameters (very negative values on log-scale).
+Therefore, we completed self-tests with these parameters fixed at the
+true values.
+
+For 7 of the the simulated data sets the model failed to optimize. The
+maximum absolute gradient was \(<10^{-6}\) for only 9 and \(<10^{-4}\)
+for 52 of the 93 successfully fitted models. The poor convergence
+appeared to be attributable to the estimation of the scalar for the
+standard errors of the log-transformed northern Recreational CPA index
+for which estimates tended to 0 for nearly all of the fits (\(<0.01\)
+for 83 fits). However, even across all fits including those with poor
+convergence, the SSB estimates appeared to be reliable (Figure
+\ref{fig:self-test-fig}).
+
+\hypertarget{one-step-ahead-residuals-for-model-m_1}{%
+\subsubsection{\texorpdfstring{One-step-ahead residuals for model
+\(M_1\)}{One-step-ahead residuals for model M\_1}}\label{one-step-ahead-residuals-for-model-m_1}}
+
+One-step-ahead residuals can now be calculated for all index, catch, and
+environmental covariate observations using methods described by Thygesen
+et al. (2017) and Trijoulet et al. (2023).
+
+\setcounter{table}{0}
+\renewcommand\thetable{S\arabic{table}}
+
+\begin{landscape}\begin{table}
+
+\caption{\label{tab:age-comp-sel-table}Configuration of age composition likelihoods, mean selectivity models, and selectivity random effects models for each age composition data component.
+For all logistic-normal likelihoods, any ages observed as zeros are treated as missing.}
+\centering
+\begin{tabular}[t]{llll}
+\toprule
+Data component & Age Composition Likelihood & Mean Selectivity model & Random effects Model\\
+\midrule
+North commercial fleet & Dirichlet-Multinomial & age-specific (ages > 3 fully selected) & AR1 correlation by age and year\\
+North recreational fleet & Logistic-normal (Independent) & age-specific (ages > 6 fully selected) & AR1 correlation by age and year\\
+South commercial fleet & Logistic-normal (AR1 correlation) & logistic & None\\
+South recreational fleet & Logistic-normal (AR1 correlation) & logistic & None\\
+North recreational CPA index & Logistic-normal (Independent) & age-specific (ages > 1 fully selected) & AR1 correlation by year\\
+\addlinespace
+North VAST index & Dirichlet-Multinomial & age-specific (ages > 4 fully selected) & AR1 correlation by age and year\\
+South recreational CPA index & Logistic-normal (AR1 correlation) & age-specific (ages > 2 fully selected) & None\\
+South VAST index & Logistic-normal (AR1 correlation) & age-specific (ages > 1 fully selected) & None\\
+\bottomrule
+\end{tabular}
+\end{table}
+\end{landscape}
+
+\setcounter{figure}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{fit_0_jitter_plt}
+
+}
+
+\caption{Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_0$.
+Black jitters had maximum absolute gradient values < $10^{-10}$, grey jitters had values > $10^{-10}$ and < 1, and red jitters had values > 1.}\label{fig:jitter-M0}
+\end{figure}
+\pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{fit_1_jitter_plt}
+
+}
+
+\caption{Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_1$.
+Fits with black dots had maximum absolute gradient value < 0.01 and fits with red dots had values > 10.}\label{fig:jitter-M1}
+\end{figure}
+\pagebreak
+
+\begin{landscape}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{self_test_results}
+
+}
+
+\caption{Median relative bias of SSB for the North and South stock components for estimation models where the observation variance of log-indices are fixed and estimation is by maximum marginal likelihood (ML) or Restricted Maximum Likelihood (REML) or where the standard deviation of logistic-normal age composition observations are assumed to be 0.1 (over sqrt(1000)) (Low Error).
+Black and Red dashed lines represent the median of the yeaerly medians and the median across all yearly relative errors, respectively.}\label{fig:self-test-fig}
+\end{figure}
+\end{landscape}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{SSB_F_R_rel_M1}
+
+}
+
+\caption{Estimates of SSB, F, and recruitment relative to those of the best performing model, $M_1$.}\label{fig:SSB-F-R-rel-M1}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{Ecov_M1_rel_M0}
+
+}
+
+\caption{Relative differences in posterior estimates of northern region bottom temperature anomalies ($\widehat X$) from the null model without effects on recruitment ($M_0$) and with effects on the northern stock component ($M_1$).}\label{fig:Ecov-M1-rel-M0}
+\end{figure}
+
+\pagebreak
+
 \end{document}

### 2024-11-19 R/fit_0_using_RT_Run33.R

@@ -1,0 +1,27 @@
+Run34 <- readRDS(file.path("c:/work/BSB.2023.RT.Modeling", "2023.RT.Runs","Run34", "fit.RDS"))
+Run33_no_effect <- readRDS(file.path("c:/work/BSB.2023.RT.Modeling", "2023.RT.Runs","Run33", "fit_no_effect.RDS"))
+
+library(wham)
+
+library(here)
+
+source(here("R","make_wham_inputs.R"))
+
+input_0 <- prepare_wham_input(asap, selectivity = sel, NAA_re = NAA_re, basic_info = basic_info, move = move, ecov = ecov_0, catch_info = catch_info,
+	index_info = index_info, age_comp = age_comp)
+
+nms <- sort(unique(names(Run33_no_effect$opt$par)))
+temp <- input_0
+temp$par[nms] <- Run33_no_effect$parList[nms]
+temp$par$logit_selpars <- input_0$par$logit_selpars
+temp$par$sel_repars <- input_0$par$sel_repars
+temp$par$logit_selpars <- Run33_no_effect$parList$logit_selpars[-(5:8),]
+temp$par$sel_repars <- Run33_no_effect$parList$sel_repars[-(5:8),]
+temp$par$sel_repars[,2:3] <- temp$par$sel_repars[,2:3]*2 #previous version had yet to change transformation for selectivity correlation parameters
+x <- temp$par$Ecov_process_pars
+x[2,] <- x[2,] + 0.5*log(1 - (-1 + 2/(1+exp(-x[3,])))^2) # previous version estimated marginal sig, instead of conditional
+temp$par$Ecov_process_pars <- x 
+# temp$par$Ecov_beta_R[] <- 0
+input_0_Run33 <- temp
+fit_0_Run33 <- fit_wham(input_0_Run33, do.brps = FALSE, do.sdrep = FALSE, do.osa = FALSE, do.retro = FALSE)
+saveRDS(fit_0_Run33, here::here("results", "fit_0_Run33.RDS"))

### 2024-11-19 R/fits.R

@@ -11,10 +11,185 @@ Run34 <- readRDS(file.path("c:/work/BSB.2023.RT.Modeling", "2023.RT.Runs","Run34
 library(wham)
 
 library(here)
+asap <- read_asap3_dat(here("data",c("north.dat","south.dat")))
+temp <- prepare_wham_input(asap)
 
-source(here("R","make_wham_inputs.R")
+#basic_info
+basic_info <- list(region_names = c("North", "South"), stock_names = paste0("BSB_", c("North", "South"))) #, NAA_where = array(1, dim = c(2,2,6)))
+#11 seasons each 1 month long except a 2 month interval in the model (June,July)
+seasons = c(rep(1,5),2,rep(1,5))/12
+basic_info$fracyr_seasons <- seasons
+#each age other than 1 (recruitment) for north stock can be in either region on Jan 1 
+basic_info$NAA_where <- array(1, dim = c(2,2,8))
+basic_info$NAA_where[1,2,1] = 0 #stock 1, age 1 can't be in region 2 
+basic_info$NAA_where[2,1,] = 0 #stock 2, any age can't be in region 1 (stock 2 doesn't move) 
+#average recruitment over years 2000+ for SSB40 BRPs
+basic_info$XSPR_R_avg_yrs <- which(temp$years>1999)
+basic_info$XSPR_R_opt <- 2 #use average of recruitments (random effects), not expected/predicted given last time step
+#############################################
 
-# pkgload::load_all("c:/work/wham/wham")
+#############################################
+#ecov
+north_bt <- read.csv(here("data","bsb_bt_temp_nmab_1959-2022.csv"))
+south_bt <- read.csv(here("data","bsb_bt_temp_smab_1959-2022.csv"))
+ecov <- list(label = c("North_BT","South_BT"))
+ecov$mean <- cbind(north_bt[,'mean'], south_bt[,'mean'])
+ecov$mean <- t(t(ecov$mean) - apply(ecov$mean,2,mean))
+ecov$logsigma <- log(cbind(north_bt[,'se'], south_bt[,'se']))
+ecov$year <- north_bt[,'year']
+ecov$use_obs <- matrix(1, NROW(ecov$mean),NCOL(ecov$mean))
+ecov$process_model <- "ar1"
[…]

### 2024-11-19 R/fits_rev.R

@@ -1,0 +1,128 @@
+#multi-wham (12/12/2023)
+#library("wham", lib.loc = "c:/work/wham/old_packages/4a44134")
+
+#pkgbuild::compile_dll("c:/work/wham/wham", debug = FALSE)
+#pkgload::load_all("c:/work/wham/wham")
+
+#install.packages("systemfonts", repos = "https://cloud.r-project.org", type = "binary")
+#pak::pkg_install("timjmiller/wham@devel", upgrade = FALSE, ask = FALSE)
+
+#devel branch is now multi-wham
+library(wham)
+
+library(here)
+
+source(here("R","make_wham_inputs.R"))
+
+input_0 <- prepare_wham_input(asap, selectivity = sel, NAA_re = NAA_re, basic_info = basic_info, move = move, ecov = ecov_0, catch_info = catch_info,
+	index_info = index_info, age_comp = age_comp)
+# input_0$par <- fit_1$parList
+# input_0$par$Ecov_beta_R[] <- 0
+
+fit_0 <- fit_wham(input_0, do.brps = FALSE, do.sdrep = FALSE, do.osa = FALSE, do.retro = FALSE)
+saveRDS(fit_0, here::here("results", "fit_0_rev.RDS"))
+# fit_0 <- readRDS(here::here("results", "fit_0_rev.RDS"))
+res_dir <- here::here("results","jitter", "fit_0")
+dir.create(res_dir)
+fit_0_jit_res <- jitter_wham(fit_RDS = here::here("results","fit_0_rev.RDS"), n_jitter = 50, res_dir = res_dir)
+saveRDS(fit_0_jit_res, here::here(res_dir, "fit_0_jitter_results.RDS"))
+
+####################
+#this will give the same initial values as used by jitter_wham
+# cov <- fit_0$sdrep$cov.fixed
+# chol.L <- t(chol(cov))
+# set.seed(8675309)
+# initial_vals <- t(sapply(1:50, function(x) fit_0$opt$par + chol.L %*% cbind(rnorm(n= NCOL(cov)))))
+####################
+
+jitters <- fit_0_jit_res[[1]]
+sort(sapply(jitters, \(x) x$obj))
+y <- order(sapply(jitters, \(x) x$obj))
+
+input_0_better <- fit_0$input
+input_0_better$par <- fit_0$env$parList(jitters[[13]]$par)
[…]

### 2024-11-19 R/make_figures.R

@@ -1,0 +1,484 @@
+library(wham)
+library(ggplot2)
+library(dplyr)
+source(here::here("R","plot_functions.R"))
+
+fits <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))
+fits_M_re <- readRDS(file.path("results","fits_M_re_better.RDS"))
+
+
+
+res_dir <- here::here("results","jitter", "fit_0")
+fit_0 <- readRDS(here::here("results", "fit_0_rev.RDS"))
+fit_0_jit_res <- readRDS(here::here(res_dir, "fit_0_jitter_results.RDS"))
+# jitter_files <- dir(here::here("results","jitter","fit_0"), full.names = TRUE, pattern = glob2rx("jitter_sim*.RDS"))
+# jitters <- lapply(fit_0_jit_res, \(x) readRDS(x))
+jitters <- fit_0_jit_res[[1]]
+sort(sapply(jitters, \(x) x$obj))
+y <- order(sapply(jitters, \(x) x$obj))
+# sapply(jitters, \(x) max(abs(x$grad)))[y]
+cbind(round(jitters[[y[3]]]$par,2), round(jitters[[y[1]]]$par - jitters[[y[3]]]$par,2))
+
+temp <- jitter_wham(fit_RDS = here::here("results","fit_0_rev.RDS"), n_jitter = 1, initial_vals = rbind(fit_0_jit_res[[2]][y[1],]))
+jit_mod <- readRDS(here::here("results", "fit_0_rev.RDS"))
+jit_mod$env$last.par.best <- temp[[1]][[1]]$last.par.best
+jit_mod$fn(temp[[1]][[1]]$par) #works!
+
+jit_mod <- readRDS(here::here("results", "fit_0_rev.RDS"))
+jit_mod$par <- fit_0_jit_res[[2]][y[1],]
+temp1 <- fit_tmb(jit_mod, n.newton = 3, do.sdrep = FALSE)
+
+jit_mod <- readRDS(here::here("results", "fit_0_rev.RDS"))
+jit_mod$par <- fit_0_jit_res[[2]][y[1],]
+temp11 <- fit_tmb(jit_mod, n.newton = 0, do.sdrep = FALSE)
+jit_mod <- readRDS(here::here("results", "fit_0_rev.RDS"))
+jit_mod$fn(temp11$opt$par)
+
[…]

### 2024-11-19 R/make_tables.R

@@ -14,7 +14,7 @@ mod_table[c(6,7),3] <- "$M$ at age 1"
 mod_table[8:14,2:3] <- mod_table[1:7,2:3]
 
 
-fits <- readRDS(file.path("results","fits_no_M_re.RDS"))
+fits <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))
 fits_M_re <- readRDS(file.path("results","fits_M_re_better.RDS"))
 
 # AIC
@@ -27,61 +27,70 @@ aic2 <- sapply(fits_M_re, function(x) {
 })
 
 diff_aic <- apply(cbind(aic1,aic2),1, function(x) x - min(x))
-aic_wts <- apply(aic,2, function(x) exp(-x/2)/sum(exp(-x/2)))
+aic_wts <- apply(diff_aic,2, function(x) exp(-x/2)/sum(exp(-x/2)))
 rownames(diff_aic) <- rownames(aic_wts) <- mod_table[,1]
 colnames(diff_aic) <- colnames(aic_wts) <- paste0("Peel ", 0:7)
 saveRDS(diff_aic, file.path("results","diff_aic.RDS"))
 saveRDS(aic_wts, file.path("results","aic_wts.RDS"))
 
+#evidence for models with/without temporal variation in age 1 M.
+apply(aic_wts, 2, \(x) c(sum(x[1:7]), sum(x[8:14])))
+apply(diff_aic, 2, \(x) exp(-x[1:7]/2)/sapply(1:7, \(y) sum(exp(-x[c(y,y+7)]/2))))
+
+
+nlminb_conv <- sapply(c(fits,fits_M_re), function(x) {
+  x$opt$conv
+})
+
+nasd_conv <- sapply(c(fits,fits_M_re), function(x) {
+  x$na_sdrep
+})
+
+maxgr_conv <- sapply(c(fits,fits_M_re), function(x) {
+  max(abs(x$final_gradient))
+})
 
 rho_ssb <- rbind(t(sapply(fits, function(x) mohns_rho(x)$SSB)),
  t(sapply(fits_M_re, function(x) mohns_rho(x)$SSB)))
 rho_Fbar <- rbind(t(sapply(fits, function(x) mohns_rho(x)$Fbar)),
  t(sapply(fits_M_re, function(x) mohns_rho(x)$Fbar)))
-rho_R <- rbind(t(sapply(fits, function(x) c(mohns_rho(x)$naa[1,1,1],mohns_rho(x)$naa[2,2,1]))),
[…]

### 2024-11-19 R/plot_functions.R

@@ -1,0 +1,376 @@
+mypalette = function(n){
+  palette.fn <- colorRampPalette(c("dodgerblue","green","red"), space = "Lab")
+  palette.fn(n)
+}
+
+plot.ecov <- function(mod, plot.pad = FALSE, do.tex=FALSE, do.png=FALSE, fontfam="", res=72, od){
+  origpar <- par(no.readonly = TRUE)
+  dat = mod$env$data
+  ecov.pred = mod$rep$Ecov_x
+  ecov.pred.low <- ecov.pred.high <- ecov.pred.se <- matrix(NA, NROW(ecov.pred), NCOL(ecov.pred))
+  ecov.obs = dat$Ecov_obs[1:dat$n_years_Ecov,,drop=F]
+  years <- seq(from=mod$input$years_Ecov[1], by=1, length.out=NROW(ecov.obs))
+  years_full <- seq(from=mod$input$years_Ecov[1], by=1, length.out=NROW(ecov.pred))#dat$n_years_Ecov+dat$n_years_proj_Ecov)
+
+  if(class(mod$sdrep)[1] == "sdreport"){
+    temp <- list(TMB:::as.list.sdreport(mod$sdrep, what = "Est", report = T),
+      TMB:::as.list.sdreport(mod$sdrep, what = "Std", report = T))
+    ecov.pred.se[] <- temp[[2]]$Ecov_x #TMB:::as.list.sdreport(mod$sdrep, what = "Std.", report=TRUE)$Ecov_x
+    ecov.pred.low[] <- ecov.pred - 1.96 * ecov.pred.se
+    ecov.pred.high[] <- ecov.pred + 1.96 * ecov.pred.se
+  }
+
+  ecov.obs.sig = mod$rep$Ecov_obs_sigma # Ecov_obs_sigma is filled with fixed, or estimated values (fe or re) for each covariate depending on the respective options
+  ecov.use = dat$Ecov_use_obs[1:dat$n_years_Ecov,,drop=F]
+  ecov.obs.sig = ecov.obs.sig[1:dat$n_years_Ecov,,drop=F]
+  ecov.obs.sig[ecov.use == 0] <- NA
+
+  # default: don't plot the padded entries that weren't used in ecov likelihood
+  if(!plot.pad) ecov.obs[ecov.use == 0] <- NA
+
[…]

### 2024-11-19 R/projections.R

@@ -1,0 +1,128 @@
+library(wham)
+fit1 <-readRDS(file.path("results","fits_no_M_re_rev.RDS"))[[2]]
+
+tdat <- cbind.data.frame(time = fit1$input$years_Ecov, bt_N = fit1$rep$Ecov_x[,1],bt_S = fit1$rep$Ecov_x[,2])
+lmtdat <- lm(bt_N ~time, data = tdat)
+newdat <- cbind.data.frame(time = c(tdat$time,tail(tdat$time,1)+1:20), bt_N = c(tdat$bt_N, rep(NA,20)), bt_S = c(tdat$bt_S, rep(NA,20)))
+predlm <- predict(lmtdat, newdata = newdat, se.fit = TRUE)
+newdat$pred_N <- predlm$fit
+newdat$pred_lo_N <- newdat$pred_N + qnorm(0.025)*predlm$se.fit
+newdat$pred_hi_N <- newdat$pred_N+ qnorm(0.975)*predlm$se.fit
+
+plot(newdat$time, newdat$bt_N)
+lines(newdat$time, newdat$pred_N)
+polygon(c(newdat$time,rev(newdat$time)), c(newdat$pred_lo_N, rev(newdat$pred_hi_N)), border = "transparent", col = gray(0.7,alpha=0.3))
+
+lmtdat <- lm(bt_S ~time, data = tdat)
+predlm <- predict(lmtdat, newdata = newdat, se.fit = TRUE)
+newdat$pred_S <- predlm$fit
+newdat$pred_lo_S <- newdat$pred_S + qnorm(0.025)*predlm$se.fit
+newdat$pred_hi_S <- newdat$pred_S + qnorm(0.975)*predlm$se.fit
+
+plot(newdat$time, newdat$bt_S)
+lines(newdat$time, newdat$pred_S)
+polygon(c(newdat$time,rev(newdat$time)), c(newdat$pred_lo_S, rev(newdat$pred_hi_S)), border = "transparent", col = gray(0.7,alpha=0.3))
+
+proj.opts.list <- list()
+# proj.opts.list[[1]] <- list(n.yrs = 11, proj_F_opt = c(5,rep(3,10)), proj_Fcatch = rep(10000,11)) #continue Ecov process by default
[…]

### 2024-11-19 R/self_tests.R

@@ -1,0 +1,217 @@
+
+library(wham)
+#seeds <- 1:10
+seeds <- sample(0:1e9, 100, replace = TRUE)
+res_dir <- here::here("results","self_test", "fit_1")
+dir.create(res_dir)
+fit_1 <- readRDS(here::here("results","fit_1.RDS"))
+self_test_1 <- self_test(fit_RDS = here::here("results","fit_1.RDS"), n = 50, seeds = seeds, res_dir = res_dir)
+saveRDS(self_test_1, file = file.path(res_dir, "self_test_results.RDS"))
+
+test_files <- dir(here::here("results","self_test","fit_1"), full.names = TRUE, pattern = glob2rx("cond_sim*.RDS"))
+sims <- lapply(test_files, \(x) readRDS(x))
+# sims <- self_test_1[[1]]
+
+sd_scale_pars <- t(sapply(sims, \(x) x$par[names(x$par) =="log_index_sig_scale"]))
+
+use <- which(sapply(sims, \(x) !(is.na(max(abs(x$grad))) | max(abs(x$grad))>0.01)))
+
+resid_SSB_N <- sapply(sims[use], \(x) x$SSB[,1]/fits[[2]]$rep$SSB[,1] - 1)
+med_res_SSB_N <- apply(resid_SSB_N, 1, median, na.rm = TRUE)
+plot(fits[[2]]$years, med_res_SSB_N)
+abline(h = 0)
+abline(h = median(med_res_SSB_N, na.rm = TRUE), lty =2)
+abline(h = median(resid_SSB_N, na.rm = TRUE), lty =2, col = "red")
+
+resid_SSB_S <- sapply(sims[use], \(x) x$SSB[,2]/fits[[2]]$rep$SSB[,2] - 1)
+med_res_SSB_S <- apply(resid_SSB_S, 1, median, na.rm = TRUE)
+plot(fits[[2]]$years, med_res_SSB_S)
+abline(h = 0)
+abline(h = median(resid_SSB_S, na.rm = TRUE), lty =2)
+
+sapply(sims, \(x) max(abs(x$grad)))
+
+resid_par <- sapply(sims, \(x) x$par - fits[[2]]$opt$par)
+med_res_par <- apply(resid_par, 1, median)
+plot(med_res_par, xlab = "par")
+abline(h = 0)
+
+#seeds <- sample(0:1e9, 100, replace = TRUE)
+res_dir <- here::here("results","self_test", "fit_1_fix")
+dir.create(res_dir)
[…]

### 2025-01-02 paper/paper.Rmd
saving work

@@
 The State-space Assessment Model [SAM, @nielsenberg14] continues to be developed and remains widely used within ICES to assess European fish stocks.
-Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada [@varkeyetal22;@perreaultetal20] and to the south, the Woods Hole Assessment Model [WHAM, @stockmiller21] is now used to assess many fish stocks in the Northwest Atlantic Ocean [@nefsc22;@nefsc22a;@nefsc24].
-
-WHAM is an R package developed and maintained at NOAA's Northeast Fisheries Science Center [https://timjmiller.github.io/wham, @millerstock20; @stockmiller21].
-WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and possibly effects of environmental covariates on various demographic parameters.
+Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada [@varkeyetal22;@perreaultetal20] and to the south, the Woods Hole Assessment Model [WHAM, @millerstock20; @stockmiller21] is now used to assess many fish stocks in the Northwest Atlantic Ocean [@nefsc22;@nefsc22a;@nefsc24].
+
+WHAM is an R package developed and maintained at NOAA's Northeast Fisheries Science Center [https://timjmiller.github.io/wham].
+WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and effects of environmental covariates on various population parameters.
 Like SAM, WHAM models are built using the Template Model Builder package [TMB, @kristensenetal16] which provides a computationally efficient means of fitting an extremely wide class of models with random effects.
-WHAM has undergone active development since its creation and includes random effects options for in recruitment, inter-annual transitions numbers at age (hereafter referred to as "survival"), fishery and index selectivity, natural mortality, and catchability.
-
-However, WHAM has up to now only allowed models with one stock (component) and a single region and without seasonal changes in stock dynamics.
+WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as "survival"), fishery and index selectivity, natural mortality, and catchability.
+
+However, WHAM has only allowed one stock (component) and a single region and without seasonal changes in stock dynamics.
 Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice [@bosleyetal22;@yingetal11;@caoetal14].
@@
 We describe here the implementation of these features and other extensions since @stockmiller21 in WHAM version 2.0.
-Many of these new configuration options can be useful whether modeling 1 or more stocks and regions.
-
-This extension of WHAM was developed in concert with a stock assessment for black sea bass through the NEFSC research track assessment process where new modeling frameworks were recently examined and modeling multiple stocks or stock components simultaneously was of interest [@nefsc23].
-\textcolor{red}{[Text on black sea bass here]} We apply WHAM 2.0 to two stock components of black sea bass off the coast of the NEUS and evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
+Many of these new configuration options can also be useful when modeling a single stock and region.
+
+We developed this extension of WHAM in concert with a research track assessment process for black sea bass where new modeling frameworks were recently examined and modeling two stock components simultaneously and temporal changes in distribution was of interest [@nefsc23].
+To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the NEUS using most of the model assumptions that were accepted during the peer-review process for the black sea bass assessment.
+However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
+
+When modeling multiple stock components occurring in different regions further dimensionality is added to the class of possible equilbrium reference points [e.g., @kapuretal21].
+Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and SSB without external simulation exercises.
+Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points.
+We demonstrate some of options of WHAM version 2.0 provide biological reference points and short-term projections.
 
@@
 Many of the options and equations of WHAM version 2.0 are the same as those in @stockmiller21, so we will only describe extensions and differences that have occurred since their first description of WHAM.
-The new version of WHAM can model multiple stocks and survival, movement, harvest and natural mortality are tracked for each stock.
+The new version of WHAM can model multiple stocks each with their own movement, harvest, and natural mortality.
 Much of the description below is for a specific stock $s$, but, for simplicity, this subscript is implicit except when necessary.
@@
 \begin{equation*}
-\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}.
-\end{equation*}
-When the transitions in abundance at age are treated deterministically, $\boldsymbol{\varepsilon}_{y,a} = 0$.
-The stock- and region-specific errors $\boldsymbol{\varepsilon}_{y,a}$ are independent, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
+\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
+\end{equation*}
+where $\boldsymbol{\varepsilon}_{y,a}$ is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
 When there is autocorrelation with age, WHAM now assumes this applies only to ages $a>1$ by default so that recruitment random effects are independent of those for for the annual transitions of older age classes.
@@
 \begin{equation*}
-  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}
+  Cov\left(\varepsilon_{y,a,r},\varepsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}
 \end{equation*}
@@
 \begin{equation*}
-  Cov\left(\epsilon_{y,1},\epsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
-\end{equation*}
+  Cov\left(\varepsilon_{y,1},\varepsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
+\end{equation*}
+When the transitions in abundance at age are treated deterministically, $\boldsymbol{\varepsilon}_{y,a} = 0$.
 Since recruitment for a given stock currently only occurs in one region $r_s$ there is only a single time-varying recruitment random effect for each stock.
@@
 \begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
-\end{equation*}
-The random effects $\epsilon_{r\rightarrow r',y,a,i}$ are season-, and(or) region-to-region-specific and modeled most generally as 2DAR1 random effects with age and(or) year where the covariance is
-\begin{equation*}
-  Cov\left(\epsilon_{r\rightarrow r',y,a,i},\epsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
+  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
+\end{equation*}
+The random effects $\varepsilon_{r\rightarrow r',y,a,i}$ are season-, and(or) region-to-region-specific and modeled most generally as 2DAR1 random effects with age and(or) year where the covariance is
+\begin{equation*}
+  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
 \end{equation*}
@@
   \begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
+  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
   \end{equation*}
@@
 \begin{equation*}
-  \log M_{y,a,r} = \theta_{M,r} + \epsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
+  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
 \end{equation*}
@@
 \begin{equation*}
-  Cov\left(\epsilon_{M,y,a,r},\epsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
+  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
 \end{equation*}
@@
 where $u_{m}$ and $l_{m}$ are the upper and lower bounds of catchability for index $m$ (defaults are 0 and 1000) and the general covariance structure for the annual random effects is AR1
-$$Cov\left(\epsilon_{q,m,y},\epsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.$$
+$$Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.$$
 
@@
 where $g(\log F)$ is the difference between the weighted sums of spawning biomass per recruit at $F$ and $X$\% of unfished spawning biomass per recruit across stocks:
-\begin{equation*}
+\begin{equation}\label{eq:newton-obj}
   g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
-\end{equation*}
+\end{equation}
 where $\upphi_s\left(F=0\right)$ is the equilibrium unfished spawning biomass per recruit.
@@
 The projection of any environmental covariates has been revised to better include error in the estimated latent covariate in any effects on the population in projection years.
+Users can also specify a catch or fully-selected fishing mortality in each projection year and they can be fleet-specific.
 
@@
 # Application to black sea bass {-}
+
+Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging from the Gulf of Mexico to the Gulf of Maine.
+Fish north of Cape Hatteras, NC, are considered to comprise a single stock unit, but individual populations exhibit spawning site fidelity
+[@ablehales97;@fabrizioetal13].
+Fish in this northern stock unit perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring [@musickmercer77].
+Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude [@mosershepherd09].
 
@@
 
-Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as "working group") composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above [@nefsc23].
+Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as "Working Group") composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above [@nefsc23].
 This assessment included the spatial features and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
 
-Below we describe the assumptions and configuration of the assessment model as determined by the working group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.
+Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.
 
@@
 As noted in the description of the general WHAM model, tagging data are not yet allowed.
-However, the working group also fit a Stock Synthesis model [@methotwetzel13] which provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).
+However, the Working Group also fit a Stock Synthesis model [@methotwetzel13] which provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).
 
@@
 
-## Uncertainty recreational CPA index observations {-}
-
-The estimated coefficients of variation (CVs) provided by the analyses to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the working group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
+## Uncertainty in recreational CPA index observations {-}
+
+The estimated coefficients of variation (CVs) provided by the analyses to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the Working Group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
 In many of the initial runs as well as the base model we allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
@@
 
-The working group investigated many alternative assumptions for the probability models and selectivity models for the 8 different sets of age composition observations to reduce residual patterns and retrospective patterns.
+The Working Group investigated many alternative assumptions for the probability models and selectivity models for the 8 different sets of age composition observations to reduce residual patterns and retrospective patterns.
 These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for 6 sets of of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
@@
 \begin{equation}\label{eq:Rec-re}
-\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \epsilon_{y,1}
+\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \varepsilon_{y,1}
 \end{equation}
@@
 \begin{equation}\label{eq:M-re}
-\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \epsilon_{M,y}.
-\end{equation}
-The constant or mean log-natural mortality rate is assumed to be $\mu_{M,a} = \log(0.4)$ for all ages as recommended by the working group.
+\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \varepsilon_{M,y}.
+\end{equation}
+The constant or mean log-natural mortality rate is assumed to be $\mu_{M,a} = \log(0.4)$ for all ages as recommended by the Working Group.
 Because the bottom temperature anomalies and random effects are centered at 0, mean log-natural mortality at age 1 over the time series should be approximately equal to $\mu_{M,1}$ for models that include those effects.
 
-## Model fitting, diagnostics, and projections {-}
+## Model fitting, diagnostics, BRPs, and projections {-}
 
@@
 
-To illustrate projections with environmental effects we projected the best black sea bass model under three alternative scenarios where the AR1 time-series model for bottom temperature continues into the projection years, where we project with bottom temperature being the average of the most recent 5 years, and where bottom temperature increases following a predictions from a simple linear regression of the estimated bottom temperature anomalies over time.
+We produced SPR-based BRP estimates ($F_{40\%}$ and SSB$_{40\%}$) from the best performing model using a variety of options provided in WHAM.
+We generated annual BRP estimates based on the annual weight at age, maturity at age, M, fleet selectivity, and either the annual recruitment random effect (Eq.
+\ref{eq:Rec-re}) or the annual expected recruitment (Eq.
+\ref{eq:expected-recruitment}).
+We also calculated status as the ratios of annual average F and SSB to the corresponding reference points.
+Finally, we calculated "prevailing" BRPs using the average weight at age, maturity at age, M, fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status (Kobe plots) for the terminal year of the model (2021).
+All BRPs and status estimates are generated for each region as well as for the total stock.
+
+We projected the best black sea bass model under three alternative scenarios for the bottom temperature covariate where 1) the AR1 time-series model continues into the projection years, 2) the average of the most recent 5 years is projected, and 3) bottom temperature increases in projection years following a prediction from a simple linear regression of the estimated bottom temperature anomalies over time.
+We specified fishing mortality in projection years to be constant at the value in the last model year (2021).
+Average weight, natural mortality, movement, and maturity at age are used in projection years whereas the AR1 processes for all numbers at age random effects are continued.
 
@@
 
-We found the best model of bottom temperature covariate effects included the effect only on the recruitment of the northern stock (Table \ref{tab:diff-aic-table}).
-However, the difference in AIC for the model that also included effects on recruitment for the southern component suggested some evidence for this hypothesis as well (Table \ref{tab:aic-wts-table}).
-
-Across retrospective peels where the terminal year of data was sequentially reduced, the ranking for the best model was consistent.
-However, when the terminal year was reduced 6 or more years, there was more evidence for further effects of temperature on recruitment for the southern component and for temporal variation in natural mortality at age 1 (Table \ref{tab:aic-wts-table}).
-
-The retrospective model fits did not indicate any evidence of patterns for northern or southern component SSB (all Mohn's $\rho \approx -0.03$) or fishing mortality (all Mohn's $\rho \approx 0.03$ and $-0.04$ for average $F$ at ages 6 and 7 in north and south regions, respectively) for any of the models (Table \ref{tab:rho-table}).
-
-## Bottom temperature effects {-}
-
-The posterior estimate of the bottom temperature covariate match the observations well because of the high precision of the observations and anomalies in the north and south region appear highly correlated (Figure \ref{fig:bottom-temperature}).
-Because the temperature anomalies are treated as latent variables, when effects on recruitment are included, other data components in the model can affect the estimated anomalies [e.g., @milleretal18], but in this case the estimates are altered negligibly during the years where recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
+We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock and no random effects on age 1 natural mortality ($M_1$, Tables \ref{tab:diff-aic-table} and  \ref{tab:aic-wts-table}).
+Model $M_1$ consistently had the lowest AIC across retrospective peels where the terminal years of data are removed from the model.
+The difference in AIC for the model that also included bottom temperature effects on recruitment for the southern component ($M_3$) suggested some evidence for this hypothesis as well.
+There was little evidence for also including random effects on age 1 natural mortality for the northern stock component.
+However, when more than three terminal years of data were removed, the evidence for adding age 1 natural mortality random effects for the northern component ($M_8$) increased is similar to adding bottom temperature effects on southern component recruitment ($M_3$).
+
+Comparisons of estimates from retrospective peels did not show any indication of differences in model performance.
+There was very little variation in measures of retrospective patterns for SSB or fishing mortality across models.
+Mohn's $\rho \approx -0.03$ for SSB of both stock components and Mohn's $\rho \approx 0.03$ and $\approx -0.04$ for average $F$ at ages 6 and 7 in north and south regions, respectively (Table \ref{tab:rho-table}).
 
 Estimated effects of bottom temperature on recruitment for either the northern or southern component were stable over the retrospective peels and differed negligibly whether effects for each component were estimated in isolation or together (Table \ref{tab:beta-sig-peel-table}).
+The estimates for the effects on each stock component were essentially equivalent whether the effect was on one component or both.
 Estimates of residual variability in northern component recruitment, as measured by the conditional or marginal standard deviation of the recruitment random effects, increased slightly with the number of peeled years.
-However, the ratio of standard deviations of models with and without temperature effects was stable with about a 20% reduction in residual standard deviation when temperature effects were included.
-
-Although evidence for temperature effects on age 1 natural mortality was weak, it is notable that the sign of the estimated effect for the northern stock component differed depending on whether random effects on age 1 natural mortality were also included ($\widehat \beta_R = -1$ for $M_{6}$ or $M_{13}$) with the negative effect estimated with these random effects ($M_{13}$).
-
-## Stock size, fishing {-}
-
-For the best performing model $M_1$, estimates  of SSB
-Stock size is also increasing for the southern component, just not as quickly as the northern component.
-(plot of SSB time series for each component).
-
-Figure \ref{fig:M1-SSB-F-R}
-
-Figure \ref{fig:SSB-F-R-rel-M1} Annual estimates of SSB and F varied little among the 14 fitted models.
-
-plot of prior/posterior for movement rate.
-
-plot of average F by fleet,region.
-
-Sectivity random effects (Figure \ref{fig:selectivity-re})
-
-Estimates of effect size, sd, correlation for each peel (Table \ref{}
-
- The residual variation in the standard deviation of recruitment random effects is reduced because the expected recruitment (Eq.
-\ref{eq:expected-recruitment}) is a function of the covariate.
-This effect is included in the proposed base model (Figure \ref{fig:BT-Ecov-R}).
-
-plot of northern component predicted recruitment vs. time with color by bottom temperature
+However, the ratio of standard deviations of models with and without temperature effects was stable with approximately 20% reduction in residual standard deviation when temperature effects were included.
+The residual variation is reduced because the expected recruitment (Eq.
+\ref{eq:expected-recruitment}) is a function of the covariate (Figure \ref{fig:BT-Ecov-R}).
+
+The posterior estimates of the bottom temperature covariate match the observations well because of the high precision of the observed anomalies (Figure \ref{fig:bottom-temperature}).
+The anomalies for the north and south regions appear highly correlated presumably because of the proximity of the two regions.
+Because the temperature anomalies are treated as latent variables, other data components in the model can affect the estimated anomalies when effects on recruitment are included [e.g., @milleretal18], but in this case the estimates of anomalies are altered negligibly during the population model years when the effect on recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
+
+The effect of the anomalies on expected recruitment can be better observed by comparing the time series of the northern and southern components (Figure \ref{fig:E-R-SSB-F}).
+The expected recruitment for the southern component without any covariate effects is constant over time whereas the effect on expected recruitment for the northern component induces correlation of the bottom temperature anomalies and expected recruitment.
+For the northern stock component, most of the lowest estimated recruitment random effects occurred prior to 2000 and several large recruitment estimates occurred after 2010.
+For the southern stock, recruitment random effect estimates have been less variable and similar to the estimated median particularly since 2005.
+The estimated stock size of the northern component, as measured by SSB, increased substantially since 2005, whereas that for the southern component peaked in 2002.
+The increase in the northern component is also reflected in the total size of the combined components.
+Estimated average total fishing mortality rates  across ages 6 and 7 (combining the commercial and recreational fleets) were highest in the southern region prior to 1998 and have declined substantially since then.
+Estimated fishing mortality rates in the northern region have been less variable, but had largest peaks in 1996, 1999, and 2005.
+The decline in estimates in the southern component is also reflected in the decline in total fishing mortality across regions.
+
+The estimated movement from the north to the south for the northern stock component is much less than anticipated from the prior distribution that was parameterized based on the companion Stock Synthesis model fit completed by the Working Group (Figure \ref{fig:move-prior-posterior}).
+Conversely, the estimated movement rate from south to north was more consistent and slightly greater than that from the companion model.
+The estimated movement rates varied little across all fitted models implying that the estimation of movement was not sensitive to the alternative assumptions we considered.
+Furthermore, the very low estimated movement rate from north to south is also consistent with preliminary fits of the WHAM model without the prior distribution where the movement rate was estimated at the lower bound of 0 resulting in a lack of convergence.
+The estimated movement rates imply that a very small fraction of northern fish are ever in the southern region.
+
+The 2dAR1 random effects for the selectivity parameters of the northern recreational fleet indicate a decrease in selectivity at ages 2 to 6 over the time series which corresponds to increases in the size limits in regulations by states where these catches occurred (Figure \ref{fig:selectivity-re}).
+The model also estimated a more modest decrease in selectivity at ages 2 to 3 for the northern commercial fleet.
+The selectivity at the youngest ages for the two fishery independent indices varied over time but without notable trend.
 
@@
 
+Using the expected recruitment rather than the random effects as the multiplier with equilibrium SSB/R results in less variable estimates of annual SSB$_{40\%}$ (Figure \ref{fig:annual-BRPs}: top row).
+Estimates were less variable for the southern stock component than the northern stock component mainly because the expected recruitment is constant for the former whereas that for the northern component varies via the effect of the bottom temperature anomalies.
+Temporal variation for the southern stock component is mainly caused by annual variation in the inputs to the SSB/R calculation and $F_{40\%}$, but there is also some effect of the variation in the proportion of total recruitment for each stock component as used in the weighting terms to define the 40\% reduction in unfished SSB/R ($\lambda_s$ in Eq.
+\ref{eq:newton-obj}).
+Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB$_{40\%}$ for the southern stock component whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
+
+Ratios of SSB to SSB$_{40\%}$ should be viewed with caution because the SSB reference point is being defined using annual recruitments rather than some form of average recruitment and therefore, ratios larger than 2.5 which would correspond to unfished population sizes are possible (Figure \ref{fig:annual-BRPs}: second row).
+However, for the southern component, using the constant expected recruitment is comparable to using median recruitment and the annual variation in the inputs to SSB/R indicates no years where SSB was greater than unfished levels throughout the time series.
+The patterns in uncertainty of SSB ratios are the same as those for the SSB reference points (Figure \ref{fig:annual-SSB-status-cvs}).
+
+Differences between annual $F_{40\%}$ estimates using the alternative types of recruitment estimates are much less than those for SSB$_{40\%}$ because the recruitment estimates are only used in the weighting of the stock-specific SSB/R to determine $F_{40\%}$ ($\lambda_s$ in Eq.
+\ref{eq:newton-obj}, Figure \ref{fig:annual-BRPs}: third row).
+For similar reasons, estimates of uncertainty of $F_{40\%}$ are also similar between the type of recruitment estimates (Figure \ref{fig:annual-F40-cvs}).
+Annual estimates for the northern stock component were lowest in the earliest years of the time series whereas those for the soutern component are relatively stable over time.
+
+For a given type of recruitment used to calculate $F_{40\%}$, the annual ratios of average $F$ to the corresponding reference point are identical for each stock component and in total because of the selectivity used to define a single F reference point across both regions is exactly the same as that for the annual estimate of fishing mortality at age by fleet (Figure \ref{fig:annual-BRPs}: fourth row).
+As for annual $F_{40\%}$, estimates of uncertainty in the ratios are similar whether expected recruitment or the random effects are used to calculate the reference points (Figure \ref{fig:annual-F-status-cvs}).
+
+Unlike the annual ratios of $F$ to $F_{40\%}$, the $F_{40\%}$ under prevailing conditions uses a selectivity defined by the average $F$ at age and fleet over the last 5 years, which is different from that of the selectivity in the terminal year and therefore those ratios for each region and in total are different (Figure \ref{fig:annual-BRPs}: bottom row).
+Although the alternative types of recruitment estimates result in different points on the kobe plots for joint status, the uncertainty in the estimates suggest similar probabilities of alternative quadrants.
+However, the usage of the confidence regions of the ratios to defined probabilities is an approximation because these ratios are functions of both empirical Bayes estimates of random effects as well as maximum likelihood estimates of fixed effects.
+
 ## Projections {-}
 
-Projections of bottom temperature and recruitment and CVs (Figure \ref{fig:R-BT-proj})
-
-Uncertainty in both is much lower during the data years than the projection years.
-
-Uncertainty in BT RE propagates into both expected recruitment and estimated random effect for recruitment, but the latter also includes uncertainty due to the residual variability in recruitment.
+When continuing the AR1 model for the bottom temperature anomalies in the projection years we can see that the uncertainty grows and asymptotes to a level determined by the marginal variance of the process and that the predicted value approaches the estimated marginal mean value of approximately 0 (Figure \ref{fig:R-BT-proj}).
+There is no uncertainty in the projected linear trend because the values were specified as known whereas there is a small level of uncertainty in the projections of the recent average based on the uncertainty in the estimated mean.
+Comparing projected recruitment for the northern and southern stock components, the projected expected recruitment has much lower uncertainty than the projected random effects regardless of whether bottom temperature affects recruitment.
+For the northern component, CVs of projected expected recruitment are smaller than those of recruitment random effect, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for the future bottom temperature anomalies are made (Figure \ref{fig:R-F-SSB-CVs}).
+
+Because projected fishing mortality is specified to be the same as that in the terminal year the same uncertainty is also propagated in the projections (Figure \ref{fig:F-SSB-proj}).
+As we would expect, there is no effect of alternative assumptions about bottom temperature in the projections for the southern stock component because the effect of the bottom temperature anomaly is only on the northern component.
+However, there alternative bottom temperature assumptions do affect projections of northern component SSB and the projections do not stabilize when the anomalies are projected to increase linearly.
+When the same AR1 assumption for the bottom temperature anomalies is used in projections, CVs of northern and total SSB projections are larger than when the recent average or linear trend are assumed (Figure \ref{fig:R-F-SSB-CVs})
+
+Unlike the annual reference points during the model years (prior to 2022), components of the SSB/R calculations are constant in the projection years where the only changing annual values are the recruitments used in the weighting for $F_{40\%}$ and the recruitments used to calculate the resulting SSB$_{40\%}$ (Figure \ref{fig:brps-proj}).
+Estimated reference points in the first few projection years can differ depending on whether the expected recruitment or the random effects are used in their calculation, but longer term projected values are identical because the random effect estimates approach the expected values as the influence of observations diminishes.
+However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures \ref{fig:annual-F40-cvs} and \ref{fig:annual-SSB40-cvs}).
+The reduction in uncertainty using expected recruitment is less for $F_{40\%}$ than SSB$_{40\%}$ and for the northern component, that includes bottom temperature effects, than the southern component.
 
@@
 
-## General Model aspects {-}
-
-WHAM provides a comprehensive treatment of environmental covariates and their effects on populations using state-space methods  [@maunder24].
-This framework accounts for the magnitude of differing uncertainties in observations and stochastic population dynamics processes.
-This version 2.0 extends these inferences to movement rates for multi-region models.
-
-## Future developments {-}
-
-An obvious limitation to WHAM is the inability to include tagging observations of any type.
-Such observations are critical to estimation of movement parameters [e.g., @goetheletal19], but also can inform mortality rate parameters [@hampton91].
-It is well known that natural mortality is seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality from other assessment model parameters [@leeetal11;@clark22].
-Estimation of natural mortality may be even more challenging within state-space assessment models with their greater flexibility from inclusion of time-varying random effects.
-For example, we fixed the mean natural mortality for age 1 fish and @cadigan16 and @stocketal21 also estimated natural mortality deviations.
-Therefore allowing tagging observations in WHAM should be a high priority even for models with a single stock and region.
-
-The work by @correaetal23 on incorporating length information and modeling growth within WHAM has not yet been merged into the multi-stock version 2.0. Growth and movement would be challenging.
-
 ## Black sea bass {-}
@@
 In our investigation of bottom temperature effects on recruitment and natural mortality and time-varying random effects on natural mortality at age 1, we found only including effects on recruitment for the northern stock component to be the best model with respect to AIC.
-This is the same configuration configuration accepted during the assessment peer-review process and it is the model currently used for management [@nefsc23].
-
-Although not relevant to the best performing model, the effect of bottom temperature on age 1 M would be expected to be opposite of that for recruitment because lower M with higher temperature would produce higher recruitment.
-This expected effect was estimated only when random effects on age 1 M were also included in the model.
-This seems to provide further support to including random effects when examining covariate effects on population parameters as demonstrated by @milleretal18 for effects on growth parameters.
-
-AIC suggested some weight for models with temperature effects on recruitment for the southern component and random effects on M at age 1.
-This might suggest making inferences on the population by using an ensemble of these and weighting by AIC [@burnhamanderson02].
-However, the estimates of assessment outputs relevant for management (SSB, F) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
-[but what about reference points?]
-
-### Temperature effects {-}
-
-BSB in the NEUS is at the northern extent of its range.
-hence the range extension to the north with increased temperature.
-The evidence for temperature effects on recruitment was strong only for the northern component which is the component of the stock at the limit of the species' range.
+Black sea bass in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north.
 Opposite effects of temperature on recruitment or population size would be expected for species in the same general area that are at the southern extent of their range [@gabriel92].
 For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer fall on the Northeast US shelf [@sullivanetal05,@milleretal16].
+
+Our finding regarding bottom temperature effects is consistent with the bottom temperature effects on recruitment in the assessment model accepted during the peer-review process and which is currently used to assess the stock [@nefsc23].
+However, AIC suggested some weight for models with temperature effects on recruitment for the southern component and random effects on M at age 1.
+Although this might suggest making inferences on the population by using an ensemble of these and weighting by AIC [@burnhamanderson02], the estimates of assessment outputs relevant for management (e.g., SSB and $F$) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
+
+We showed that using expected recruitment rather than recruitment random effects to define BRPs can produce the same estimates but with lower uncertainty.
+Using expected bottom temperature anomalies rather than the random effects would also result in the same BRPs, but presumably with even less uncertainty than those we showed using expected recruitment.
+These results can be important for management when uncertainty in assessment output is used in making catch advice.
+The typical approach for defining stock and harvest status is to compare BRPs defined under prevailing conditions with current stock and harvest levels, but we could also make the comparisons with projected BRPs like those we demonstrated are possible.
+The uncertainty in these ratios could also be used in defining catch advice.
+
+As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
+Elaborate!
+
+<!--
 
@@
 
-Including tagging data in WHAM would also be helpful for BSB because we are using the same data to estimate movement outside of the model as we are using to estimate the other parameters in WHAM.
+~~Although not relevant to the best performing model, the effect of bottom temperature on age 1 M would be expected to be opposite of that for recruitment because lower M with higher temperature would produce higher recruitment.
+This expected effect was estimated only when random effects on age 1 M were also included in the model.
+This seems to provide further support to including random effects when examining covariate effects on population parameters as demonstrated by @milleretal18 for effects on growth parameters.~~
+
+-->
+
+## General Model aspects {-}
+
+WHAM provides a large class of models for time- and age-varying random effects for population attributes and treatment of environmental covariates and their effects on populations using state-space methods [@maunder24].
+This framework accounts for the magnitude of differing uncertainties in observations and stochastic population dynamics processes.
+Version 2.0 extends these inferences to movement rates for multi-region models and stock (component) specific configuration of random and environmental covariate effects.
+
+## Future developments {-}
+
+An obvious limitation to WHAM is the inability to include tagging observations of any type.
+Such observations are critical to estimation of movement parameters [e.g., @goetheletal19].
+We used estimates of movement parameters from a companion assessment model to define prior distributions for corresponding movement parameters in the state-space black sea bass model, but other than the tagging data, both of the assessment models used much the same data which can lead to inappropriate inferences.
+A better approach would be to estimate the movement parameters externally from just from the tagging data.
+
+Tagging data can also inform mortality rate parameters [@hampton91].
+It is well known that natural mortality is seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality from other assessment model parameters [@leeetal11;@clark22].
+Estimation of natural mortality may be even more challenging within state-space assessment models with their greater flexibility from inclusion of time-varying random effects.
+For example, we fixed the mean natural mortality for age 1 fish and @cadigan16 and @stocketal21 also estimated natural mortality deviations.
+Therefore allowing tagging observations in WHAM should be a high priority even for models with a single stock and region.
+This extension would also be beneficial for black sea bass because the current assessment uses much the same data to estimate movement outside of the model as we are using to estimate the other parameters in the WHAM model.
+
+WHAM version 2.0 only currently allows models where each stock spawns and recruits in a single region however, many stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region (references).
+Therefore, extending WHAM to allow such models would be beneficial.
+Similarly age composition data are scarce for many stocks incorporating the work by @correaetal23 on allowing length composition and modeling growth within WHAM would provide an assessment tool to include process errors in assessment with such data and possibly with multiple stocks and regions.
 
@@
 
-# Appendix A {-}
-
-\setcounter{table}{0}
-\renewcommand\thetable{A\arabic{table}}
-\input{symboltable.tex}
-
 \pagebreak
@@
 
-```{r bottom-temperature, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Observations and 95\\% confidence intervals (points with vertical lines) and posterior estimates (lines) with 95\\% confidence intervals (polygons) of bottom temperature anomalies in the north and south regions from model $M_0$."}
-knitr::include_graphics(here::here("paper", "Ecov.pdf"))
-```
-\pagebreak
-
-```{r BT-Ecov-R, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Expected and Posterior Emperical Bayes estimates of annual recruitment for the northern stock component.
-Color defined by the corresponding annual bottom temperature anomaly."}
-knitr::include_graphics(here::here("paper", "best_R_Ecov.pdf"))
-```
-\pagebreak
-
-```{r M1-SSB-F-R, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of SSB, average F (ages 6-7), and recruitment from model $M_1$.
-Polygons represent 95\\% confidence intervals."}
-knitr::include_graphics(here::here("paper", "M_1_SSB_F_R.pdf"))
-```
-\pagebreak
-
-```{r R-BT-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Estimates of bottom temperature, recruitment, and expected recruitment, and coefficients of variation for the latter two from model $M_1$ assuming alternative bottom temperature scenarios described in the text.
-Polygons represent 95\\% confidence intervals."}
-knitr::include_graphics(here::here("paper", "R_proj_results.pdf"))
-```
-
+```{r E-R-SSB-F, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual observations and posterior estimates of bottom temperature anomalies (top row), annual expected and random effect recruitment estimates (second row) and SSB and average fishing mortality at ages 6 and 7 from model $M_1$.
+Columns define estimates regionally (spawning region for SSB) and toal SSB and fishing mortality across regions.
+Polygons and vertical lines represent 95\\% confidence intervals."}
+knitr::include_graphics(here::here("paper", "E_R_SSB_F_fig.pdf"))
+```
 \pagebreak
@@
 
+```{r annual-BRPs, echo = FALSE, eval= TRUE, out.height = "85%", fit.align = "center", fig.cap = "Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio, equilibrium SSB at $F_{40\\%}$ based on annual inputs to SSB/R calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to SSB/R averaged over the last 5 years and recruitment is averaged over the time series.
+Red and blue in first 4 rows and black and gray in final row indicate alternative annual estimates of recruitment (random effect or expected).
+All results are based on model $M_1$.
+Polygons represent 95\\% confidence intervals or regions."}
+knitr::include_graphics(here::here("paper", "brp_status_results.pdf"))
+```
+
+```{r R-BT-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of bottom temperature in the northern region and alternative recruitment estimates (random effect or expected) by region.
+Estimates from years after 2021 are from projections of model $M1$ under three alternative assumptions for the bottom temperature anomalies in the northern region.
+Vertical dotted line is the last year of data and polygons represent 95\\% confidence intervals."}
+knitr::include_graphics(here::here("paper", "proj_ecov_Recruit_results.pdf"))
+```
+
+```{r F-SSB-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of average fishing mortality and SSB by region and in total.
+Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
+Vertical dotted line is the last year of data and polygons represent 95\\% confidence intervals."}
+knitr::include_graphics(here::here("paper", "proj_F_SSB.pdf"))
+```
+
+\begin{landscape}
+
+```{r brps-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates by region and in total of average equilibrium $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio and SSB at $F_{40\\%}$.
+Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region and average SSB/R inputs from the last 5 years of the unprojected model.
+Vertical dotted line is the last year of data and polygons represent 95\\% confidence intervals."}
+knitr::include_graphics(here::here("paper", "proj_brps.pdf"))
+```
+
+\end{landscape}
+
 \clearpage
-
-# Tables {-}
 
@@
 \renewcommand\thetable{\arabic{table}}
+
+# Tables {-}
 
@@
     caption="Assumptions for temperature effects and random effects for age 1 natural mortality for each model.", align = c(rep("l",4))) %>%
-  add_header_above(c(" " = 1, "Temperature Effect" = 2, " " = 1 ))
+  add_header_above(c(" " = 1, "Temperature Effect" = 2, " " = 1 )) %>% kable_styling(latex_options = "hold_position")
 ```
@@
   caption = "Difference between AIC and the lowest AIC for each model by retrospective peel.") %>%
-  add_header_above(c(" " = 1, "Peel" = 8))
-```
-
-```{r, echo = F, eval = T}
-aic_wts <- readRDS("../results/aic_wts.RDS")
-aic_wts <- data.frame(round(aic_wts, 2))
-aic_wts <- cbind(Model = row.names(aic_wts), aic_wts)
-row.names(aic_wts) <- NULL
-colnames(aic_wts) <- c("Model", as.character(0:7))
-aic_wts %>%
-  kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "aic-wts-table",
-  caption = "Model AIC weights for each retrospective peel.") %>%
   add_header_above(c(" " = 1, "Peel" = 8))
@@
 
+# Appendix A {-}
+
+\setcounter{table}{0}
+\renewcommand\thetable{A\arabic{table}}
+\input{symboltable.tex}
+
+\pagebreak
+
 # Supplemental Materials {-}
@@
 
-The working group fit a Stock Synthesis model [@methotwetzel13] that included tagging data with 2 seasons (6 months each) and 2 regions where a proportion $\mu^*_1$ of the northern component moves to the south in one season and some proportion  $\mu^*_{2\rightarrow 1}$ move back to the south in the second season [@nefsc23].
+The Working Group fit a Stock Synthesis model [@methotwetzel13] that included tagging data with 2 seasons (6 months each) and 2 regions where a proportion $\mu^*_1$ of the northern component moves to the south in one season and some proportion  $\mu^*_{2\rightarrow 1}$ move back to the south in the second season [@nefsc23].
 The seasonal movement matrices for each season are
@@
 where $z_1 = 1/5$ for and $z_2 = 1/4$, and $\mathbf{V}_{k}$ and $\mathbf{D}_{k}$ are the matrix of eigenvectors (columnwise) and the diagonal matrix of corresponding eigenvalues of $\boldsymbol{\mu}^*_k$.
-The working group used a parametric bootstrap approach to determine an appropriate standard deviation for the prior distribution for the movement parameters.
+The Working Group used a parametric bootstrap approach to determine an appropriate standard deviation for the prior distribution for the movement parameters.
 Stock Synthesis also estimates parameters on a transformed scale, but different from WHAM:
@@
 
-### Jitter fits for model $M_0$
+### Jitter fits for model $M_0$ {-}
 
@@
 
-### Jitter fits for model $M_1$
+### Jitter fits for model $M_1$ {-}
 
@@
 
-### Self test for model $M_1$
+### Self test for model $M_1$ {-}
 
@@
 However, even across all fits including those with poor convergence, the SSB estimates appeared to be reliable (Figure \ref{fig:self-test-fig}).
-
-### One-step-ahead residuals for model $M_1$
-
-One-step-ahead residuals can now be calculated for all index, catch, and environmental covariate observations using methods described by @thygesenetal17 and @trijouletetal23.
-
-<!--
-
-We used one-step-ahead (OSA) residuals to evaluate mis-specification of the proposed base model.
-Pearson residuals do not have the appropriate properties (independent standard normal for a correctly specified model) for composition observations [@trijouletetal23] or for covariate and aggregate index and catch observations when temporal random effects are employed  because of the lack of independence of observations [@thygesenetal17].
-
-There is no evidence of mis-specificaiton from the OSA residuals for aggregate catch of the northern commercial fleet (Figure \ref{fig:osa-North-comm-catch-summ}).
-There was some indication of trends in residuals of some of the age composition observations early in the time series and for the the first age class (Figures \ref{fig:osa-North-comm-paa-summ} and \ref{fig:osa-North-comm-paa}).
-
-For the northern recreational fleet, the OSA residuals for the aggregate catch appeared satisfactory (Figure \ref{fig:osa-North-rec-catch-summ}).
-The OSA residuals for the age composition observations for the northern recreational fleet showed some tendency of underdispersion but were otherwise unnoteworthy (Figures \ref{fig:osa-North-rec-paa-summ} and \ref{fig:osa-North-rec-paa}).
-
-Neither the OSA residuals for the aggregate catch or the age composition of the Northern recreational CPA index exhibited signs of mis-specification (Figures \ref{fig:osa-North-reccpa-catch-summ} to \ref{fig:osa-North-reccpa-paa}).
-
-There was a little evidence of tendency toward negative OSA residuals for the Northern VAST index whereas there was some indication of positive residuals for the age composition observations, particularly at age 1 (Figures \ref{fig:osa-North-vast-catch-summ} to \ref{fig:osa-North-vast-paa})
-
-THe OSA residuals for the aggregate catch of the southern commercial fleet tended to be negative, but the residuals for the age composition observations did not show any evidence of mis-specification (Figures \ref{fig:osa-South-comm-catch-summ} to \ref{fig:osa-South-comm-paa}).
-
-The OSA residuals for the aggregate catch of the southern recreational fleet were under-dispersed and those for the age composition were somewhat underdispersed, but no trends in either types of observations were apparent (Figures \ref{fig:osa-South-rec-catch-summ} to \ref{fig:osa-South-rec-paa})
-
-The OSA residuals for the aggregate catch of the southern recreational CPA index were also somewhat under-dispersed (Figures \ref{fig:osa-South-reccpa-catch-summ}).
-The residuals for the age composition observations exhibited a few large negative residuals at older ages, but otherwise exhibited no apparent trends (Figures \ref{fig:osa-South-reccpa-paa-summ} to \ref{fig:osa-South-reccpa-paa})
-
-The OSA residuals for the aggregate southern VAST index were somewhat over-dispersed and the residuals for the age composition appeared to show some trend with observed proportion and age (Figures \ref{fig:osa-South-vast-catch-summ} to \ref{fig:osa-South-vast-paa}).
--->
 
@@
 
+```{r, echo = F, eval = T}
+aic_wts <- readRDS("../results/aic_wts.RDS")
+aic_wts <- data.frame(round(aic_wts, 2))
+aic_wts <- cbind(Model = row.names(aic_wts), aic_wts)
+row.names(aic_wts) <- NULL
+colnames(aic_wts) <- c("Model", as.character(0:7))
+aic_wts %>%
+  kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "aic-wts-table",
+  caption = "Model AIC weights for each retrospective peel.") %>%
+  add_header_above(c(" " = 1, "Peel" = 8))
+```
+
 \setcounter{figure}{0}
@@
 ```
+
+```{r bottom-temperature, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Observations with 95\\% confidence intervals (points with vertical lines) and posterior estimates with 95\\% confidence intervals (lines with polygons) of bottom temperature anomalies in the north and south regions from model $M_1$.
+Gray vertical line defines the first year that the black sea bass stock is modeled."}
+knitr::include_graphics(here::here("paper", "BTA_full_fig.pdf"))
+```
+
 \end{landscape}
+\pagebreak
+
+```{r BT-Ecov-R, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Expected and random effect recruitment estimates for the northern stock component.
+Color of points defined by the corresponding annual bottom temperature anomaly."}
+knitr::include_graphics(here::here("paper", "best_R_Ecov.pdf"))
+```
+\pagebreak
 
@@
 
-\pagebreak
+\begin{landscape}
+
+```{r R-F-SSB-CVs, echo = FALSE, eval= TRUE, out.height = "90%", fit.align = "center", fig.cap = "Coefficients of variation for estimates of alternative recruitment estimates (random effects or expected), average fishing mortality at age 6 and 7, and SSB by region and in total from model $M_1$.
+Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
+knitr::include_graphics(here::here("paper", "R_SSB_F_cv_results.pdf"))
+```
+\end{landscape}
+
+```{r annual-F40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
+knitr::include_graphics(here::here("paper", "proj_F40_CV.pdf"))
+```
+
+```{r annual-SSB40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium SSB$_{40\\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations and alternative annual recruitment types.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
+knitr::include_graphics(here::here("paper", "proj_SSB40_CV.pdf"))
+```
+
+```{r F-status-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of ratios of fishing mortality to $F_{40\\%}$ by region and in total.
+Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
+Vertical dotted line is the last year of data and polygons represent 95\\% confidence intervals."}
+knitr::include_graphics(here::here("paper", "proj_F_status_results.pdf"))
+```
+```{r SSB-status-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of ratios of SSB to SSB$_{40\\%}$ by region and in total.
+Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
+Vertical dotted line is the last year of data and polygons represent 95\\% confidence intervals."}
+knitr::include_graphics(here::here("paper", "proj_SSB_status_results.pdf"))
+```
+
+```{r annual-SSB-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
+knitr::include_graphics(here::here("paper", "proj_SSB_status_CV.pdf"))
+```
+
+```{r annual-F-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\\bar{F}_{40\\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
+knitr::include_graphics(here::here("paper", "proj_F_status_CV.pdf"))
+```

### 2025-01-02 paper/paper.tex
saving work

@@
 Curti \and Alex Hansell}
-\date{14 November, 2024}
+\date{02 January, 2025}
 
@@
 stocks in the waters of Eastern Canada (Perreault et al. 2020; Varkey et
-al. 2022) and to the south, the Woods Hole Assessment Model (WHAM, Stock
-and Miller 2021) is now used to assess many fish stocks in the Northwest
-Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).
+al. 2022) and to the south, the Woods Hole Assessment Model (WHAM,
+Miller and Stock 2020; Stock and Miller 2021) is now used to assess many
+fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC
+2024).
 
 WHAM is an R package developed and maintained at NOAA's Northeast
-Fisheries Science Center (\url{https://timjmiller.github.io/wham},
-Miller and Stock 2020; Stock and Miller 2021). WHAM can be configured to
-fit a wide range of age-structured models from traditional statistical
-catch-at-age models without any random effects to models with several
-time and age varying process errors and possibly effects of
-environmental covariates on various demographic parameters. Like SAM,
+Fisheries Science Center {[}\url{https://timjmiller.github.io/wham}{]}.
+WHAM can be configured to fit a wide range of age-structured models from
+traditional statistical catch-at-age models without any random effects
+to models with several time and age varying process errors and effects
+of environmental covariates on various population parameters. Like SAM,
 WHAM models are built using the Template Model Builder package (TMB,
@@
 has undergone active development since its creation and includes random
-effects options for in recruitment, inter-annual transitions numbers at
+effects options for recruitment, inter-annual transitions of numbers at
 age (hereafter referred to as ``survival''), fishery and index
@@
 
-However, WHAM has up to now only allowed models with one stock
-(component) and a single region and without seasonal changes in stock
-dynamics. Using such models for stocks that have subcomponents with
-varying seasonal movement can provide incorrect inferences and poor
-management advice (Ying et al. 2011; Cao et al. 2014; Bosley et al.
-2022). Furthermore, the ability to account for spatial structure and
-model multiple stocks are also important features of leading-edge
-assessment modeling frameworks (Punt et al. 2020). We describe here the
-implementation of these features and other extensions since Stock and
-Miller (2021) in WHAM version 2.0. Many of these new configuration
-options can be useful whether modeling 1 or more stocks and regions.
-
-This extension of WHAM was developed in concert with a stock assessment
-for black sea bass through the NEFSC research track assessment process
-where new modeling frameworks were recently examined and modeling
-multiple stocks or stock components simultaneously was of interest
-(NEFSC 2023). \textcolor{red}{[Text on black sea bass here]} We apply
-WHAM 2.0 to two stock components of black sea bass off the coast of the
-NEUS and evaluate evidence for alternative hypotheses of temporal
-variation and effects specifically of bottom temperature on recruitment
-and natural mortality of age 1 individuals.
+However, WHAM has only allowed one stock (component) and a single region
+and without seasonal changes in stock dynamics. Using such models for
+stocks that have subcomponents with varying seasonal movement can
+provide incorrect inferences and poor management advice (Ying et al.
+2011; Cao et al. 2014; Bosley et al. 2022). Furthermore, the ability to
+account for spatial structure and model multiple stocks are also
+important features of leading-edge assessment modeling frameworks (Punt
+et al. 2020). We describe here the implementation of these features and
+other extensions since Stock and Miller (2021) in WHAM version 2.0. Many
+of these new configuration options can also be useful when modeling a
+single stock and region.
+
+We developed this extension of WHAM in concert with a research track
+assessment process for black sea bass where new modeling frameworks were
+recently examined and modeling two stock components simultaneously and
+temporal changes in distribution was of interest (NEFSC 2023). To
+illustrate the usage of WHAM 2.0 we apply it to two stock components of
+black sea bass off the coast of the NEUS using most of the model
+assumptions that were accepted during the peer-review process for the
+black sea bass assessment. However, here we evaluate evidence for
+alternative hypotheses of temporal variation and effects specifically of
+bottom temperature on recruitment and natural mortality of age 1
+individuals.
+
+When modeling multiple stock components occurring in different regions
+further dimensionality is added to the class of possible equilbrium
+reference points (e.g., Kapur et al. 2021). Short term projections are
+also an important part of the management process and the separation of
+observation and process errors in state-space models allows uncertainty
+in projected attributes such as recruitment and SSB without external
+simulation exercises. Finally, the inclusion of explicit effects of
+environmental covariates in the model also requires some assumptions on
+how they are treated in projections and reference points. We demonstrate
+some of options of WHAM version 2.0 provide biological reference points
+and short-term projections.
 
@@
 and differences that have occurred since their first description of
-WHAM. The new version of WHAM can model multiple stocks and survival,
-movement, harvest and natural mortality are tracked for each stock. Much
-of the description below is for a specific stock \(s\), but, for
-simplicity, this subscript is implicit except when necessary.
+WHAM. The new version of WHAM can model multiple stocks each with their
+own movement, harvest, and natural mortality. Much of the description
+below is for a specific stock \(s\), but, for simplicity, this subscript
+is implicit except when necessary.
 
@@
 \begin{equation*}
-\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}.
-\end{equation*} When the transitions in abundance at age are treated
-deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). The stock-
-and region-specific errors \(\boldsymbol{\varepsilon}_{y,a}\) are
-independent, but the same correlation structures as previous versions
-are possible across ages and years for a given stock and region. When
-there is autocorrelation with age, WHAM now assumes this applies only to
-ages \(a>1\) by default so that recruitment random effects are
-independent of those for for the annual transitions of older age
-classes. So the general covariance structure for a given stock at ages
-\(a>1\) in region \(r\) is that of a two-dimensional first-order
+\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
+\end{equation*} where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector
+of region-specific errors for a given stock, which are independent
+across stocks and regions, but the same correlation structures as
+previous versions are possible across ages and years for a given stock
+and region. When there is autocorrelation with age, WHAM now assumes
+this applies only to ages \(a>1\) by default so that recruitment random
+effects are independent of those for for the annual transitions of older
+age classes. So the general covariance structure for a given stock at
+ages \(a>1\) in region \(r\) is that of a two-dimensional first-order
 autoregressive (2DAR1) process \begin{equation*}
-  Cov\left(\epsilon_{y,a,r},\epsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}
+  Cov\left(\varepsilon_{y,a,r},\varepsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}
 \end{equation*} and that for age 1 is just AR1 across years
 \begin{equation*}
-  Cov\left(\epsilon_{y,1},\epsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
-\end{equation*} Since recruitment for a given stock currently only
-occurs in one region \(r_s\) there is only a single time-varying
-recruitment random effect for each stock.
+  Cov\left(\varepsilon_{y,1},\varepsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
+\end{equation*} When the transitions in abundance at age are treated
+deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). Since
+recruitment for a given stock currently only occurs in one region
+\(r_s\) there is only a single time-varying recruitment random effect
+for each stock.
 
@@
 and environmental effects: \begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
-\end{equation*} The random effects \(\epsilon_{r\rightarrow r',y,a,i}\)
-are season-, and(or) region-to-region-specific and modeled most
-generally as 2DAR1 random effects with age and(or) year where the
-covariance is \begin{equation*}
-  Cov\left(\epsilon_{r\rightarrow r',y,a,i},\epsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
+  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
+\end{equation*} The random effects
+\(\varepsilon_{r\rightarrow r',y,a,i}\) are season-, and(or)
+region-to-region-specific and modeled most generally as 2DAR1 random
+effects with age and(or) year where the covariance is \begin{equation*}
+  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
 \end{equation*} similar to how WHAM models variation in survival,
@@
 as \begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \epsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
+  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
   \end{equation*}
@@
 for natural mortality is \begin{equation*}
-  \log M_{y,a,r} = \theta_{M,r} + \epsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
+  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
 \end{equation*} The general covariance structure for random effects are
@@
 where the covariance is \begin{equation*}
-  Cov\left(\epsilon_{M,y,a,r},\epsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
+  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
 \end{equation*}
@@
 covariance structure for the annual random effects is AR1
-\[Cov\left(\epsilon_{q,m,y},\epsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.\]
+\[Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.\]
 
@@
 weighted sums of spawning biomass per recruit at \(F\) and \(X\)\% of
-unfished spawning biomass per recruit across stocks: \begin{equation*}
+unfished spawning biomass per recruit across stocks:
+\begin{equation}\label{eq:newton-obj}
   g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
-\end{equation*} where \(\upphi_s\left(F=0\right)\) is the equilibrium
+\end{equation} where \(\upphi_s\left(F=0\right)\) is the equilibrium
 unfished spawning biomass per recruit. \(g'(\log F)\) is the derivative
@@
 estimated latent covariate in any effects on the population in
-projection years.
+projection years. Users can also specify a catch or fully-selected
+fishing mortality in each projection year and they can be
+fleet-specific.
 
@@
 \addcontentsline{toc}{section}{Application to black sea bass}
+
+Black sea bass are a temperate reef fish in the western Atlantic Ocean
+ranging from the Gulf of Mexico to the Gulf of Maine. Fish north of Cape
+Hatteras, NC, are considered to comprise a single stock unit, but
+individual populations exhibit spawning site fidelity (Able and Stanton
+Hales, Jr. 1997; Fabrizio et al. 2013). Fish in this northern stock unit
+perform seasonal migrations out on the continental shelf in the fall and
+back to their inshore spawning areas during the spring (Musick and
+Mercer 1977). Analyses of tagging studies have demonstrated that the
+extent of seasonal migration varies along the coast such that fish from
+populations off of southern New England and further north move offshore
+and as far south as the coasts of Virgina and North Carolina, whereas
+fish in the southern portion of the stock range generally move shorter
+distances between inshore and offshore areas of similar latitude (Moser
+and Shepherd 2009).
 
@@
 Leading up to the 2023 peer-reviewed assessment, a working group
-(hereafter referred to as ``working group'') composed of scientists from
+(hereafter referred to as ``Working Group'') composed of scientists from
 federal, state, and academic institutions determined an optimal data and
@@
 Below we describe the assumptions and configuration of the assessment
-model as determined by the working group as well as the alternative
+model as determined by the Working Group as well as the alternative
 assumptions for recruitment and natural mortality in the models we fit
@@
 the description of the general WHAM model, tagging data are not yet
-allowed. However, the working group also fit a Stock Synthesis model
+allowed. However, the Working Group also fit a Stock Synthesis model
 (Methot and Wetzel 2013) which provided estimated movement rate
@@
 
-\hypertarget{uncertainty-recreational-cpa-index-observations}{%
-\subsection*{Uncertainty recreational CPA index
-observations}\label{uncertainty-recreational-cpa-index-observations}}
-\addcontentsline{toc}{subsection}{Uncertainty recreational CPA index
+\hypertarget{uncertainty-in-recreational-cpa-index-observations}{%
+\subsection*{Uncertainty in recreational CPA index
+observations}\label{uncertainty-in-recreational-cpa-index-observations}}
+\addcontentsline{toc}{subsection}{Uncertainty in recreational CPA index
 observations}
@@
 to generate the recreational catch per angler (recreational CPA) index
-ranged between 0.02 and 0.06 which the working group felt did not
+ranged between 0.02 and 0.06 which the Working Group felt did not
 capture the true observation uncertainty in the index with regard to its
@@
 
-The working group investigated many alternative assumptions for the
+The Working Group investigated many alternative assumptions for the
 probability models and selectivity models for the 8 different sets of
@@
 \begin{equation}\label{eq:Rec-re}
-\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \epsilon_{y,1}
+\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \varepsilon_{y,1}
 \end{equation} and \begin{equation}\label{eq:M-re}
-\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \epsilon_{M,y}.
+\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \varepsilon_{M,y}.
 \end{equation} The constant or mean log-natural mortality rate is
 assumed to be \(\mu_{M,a} = \log(0.4)\) for all ages as recommended by
-the working group. Because the bottom temperature anomalies and random
+the Working Group. Because the bottom temperature anomalies and random
 effects are centered at 0, mean log-natural mortality at age 1 over the
@@
 
-\hypertarget{model-fitting-diagnostics-and-projections}{%
-\subsection*{Model fitting, diagnostics, and
-projections}\label{model-fitting-diagnostics-and-projections}}
-\addcontentsline{toc}{subsection}{Model fitting, diagnostics, and
+\hypertarget{model-fitting-diagnostics-brps-and-projections}{%
+\subsection*{Model fitting, diagnostics, BRPs, and
+projections}\label{model-fitting-diagnostics-brps-and-projections}}
+\addcontentsline{toc}{subsection}{Model fitting, diagnostics, BRPs, and
 projections}
@@
 
-To illustrate projections with environmental effects we projected the
-best black sea bass model under three alternative scenarios where the
-AR1 time-series model for bottom temperature continues into the
-projection years, where we project with bottom temperature being the
-average of the most recent 5 years, and where bottom temperature
-increases following a predictions from a simple linear regression of the
-estimated bottom temperature anomalies over time.
+We produced SPR-based BRP estimates (\(F_{40\%}\) and SSB\(_{40\%}\))
+from the best performing model using a variety of options provided in
+WHAM. We generated annual BRP estimates based on the annual weight at
+age, maturity at age, M, fleet selectivity, and either the annual
+recruitment random effect (Eq. \ref{eq:Rec-re}) or the annual expected
+recruitment (Eq. \ref{eq:expected-recruitment}). We also calculated
+status as the ratios of annual average F and SSB to the corresponding
+reference points. Finally, we calculated ``prevailing'' BRPs using the
+average weight at age, maturity at age, M, fleet selectivity from the
+last 5 years (2017 to 2021) and average recruitment over years 2000 to
+2021, and made Kobe plots of joint status (Kobe plots) for the terminal
+year of the model (2021). All BRPs and status estimates are generated
+for each region as well as for the total stock.
+
+We projected the best black sea bass model under three alternative
+scenarios for the bottom temperature covariate where 1) the AR1
+time-series model continues into the projection years, 2) the average of
+the most recent 5 years is projected, and 3) bottom temperature
+increases in projection years following a prediction from a simple
+linear regression of the estimated bottom temperature anomalies over
+time. We specified fishing mortality in projection years to be constant
+at the value in the last model year (2021). Average weight, natural
+mortality, movement, and maturity at age are used in projection years
+whereas the AR1 processes for all numbers at age random effects are
+continued.
 
@@
 We found the best model of bottom temperature covariate effects included
-the effect only on the recruitment of the northern stock (Table
-\ref{tab:diff-aic-table}). However, the difference in AIC for the model
-that also included effects on recruitment for the southern component
-suggested some evidence for this hypothesis as well (Table
-\ref{tab:aic-wts-table}).
-
-Across retrospective peels where the terminal year of data was
-sequentially reduced, the ranking for the best model was consistent.
-However, when the terminal year was reduced 6 or more years, there was
-more evidence for further effects of temperature on recruitment for the
-southern component and for temporal variation in natural mortality at
-age 1 (Table \ref{tab:aic-wts-table}).
-
-The retrospective model fits did not indicate any evidence of patterns
-for northern or southern component SSB (all Mohn's
-\(\rho \approx -0.03\)) or fishing mortality (all Mohn's
-\(\rho \approx 0.03\) and \(-0.04\) for average \(F\) at ages 6 and 7 in
-north and south regions, respectively) for any of the models (Table
-\ref{tab:rho-table}).
-
-\hypertarget{bottom-temperature-effects-1}{%
-\subsection*{Bottom temperature
-effects}\label{bottom-temperature-effects-1}}
-\addcontentsline{toc}{subsection}{Bottom temperature effects}
-
-The posterior estimate of the bottom temperature covariate match the
-observations well because of the high precision of the observations and
-anomalies in the north and south region appear highly correlated (Figure
-\ref{fig:bottom-temperature}). Because the temperature anomalies are
-treated as latent variables, when effects on recruitment are included,
-other data components in the model can affect the estimated anomalies
-(e.g., Miller et al. 2018), but in this case the estimates are altered
-negligibly during the years where recruitment is estimated (Figure
-\ref{fig:Ecov-M1-rel-M0}).
+the effect only on recruitment of the northern stock and no random
+effects on age 1 natural mortality (\(M_1\), Tables
+\ref{tab:diff-aic-table} and \ref{tab:aic-wts-table}). Model \(M_1\)
+consistently had the lowest AIC across retrospective peels where the
+terminal years of data are removed from the model. The difference in AIC
+for the model that also included bottom temperature effects on
+recruitment for the southern component (\(M_3\)) suggested some evidence
+for this hypothesis as well. There was little evidence for also
+including random effects on age 1 natural mortality for the northern
+stock component. However, when more than three terminal years of data
+were removed, the evidence for adding age 1 natural mortality random
+effects for the northern component (\(M_8\)) increased is similar to
+adding bottom temperature effects on southern component recruitment
+(\(M_3\)).
+
+Comparisons of estimates from retrospective peels did not show any
+indication of differences in model performance. There was very little
+variation in measures of retrospective patterns for SSB or fishing
+mortality across models. Mohn's \(\rho \approx -0.03\) for SSB of both
+stock components and Mohn's \(\rho \approx 0.03\) and \(\approx -0.04\)
+for average \(F\) at ages 6 and 7 in north and south regions,
+respectively (Table \ref{tab:rho-table}).
 
@@
 estimated in isolation or together (Table
-\ref{tab:beta-sig-peel-table}). Estimates of residual variability in
-northern component recruitment, as measured by the conditional or
-marginal standard deviation of the recruitment random effects, increased
-slightly with the number of peeled years. However, the ratio of standard
+\ref{tab:beta-sig-peel-table}). The estimates for the effects on each
+stock component were essentially equivalent whether the effect was on
+one component or both. Estimates of residual variability in northern
+component recruitment, as measured by the conditional or marginal
+standard deviation of the recruitment random effects, increased slightly
+with the number of peeled years. However, the ratio of standard
 deviations of models with and without temperature effects was stable
-with about a 20\% reduction in residual standard deviation when
-temperature effects were included.
-
-Although evidence for temperature effects on age 1 natural mortality was
-weak, it is notable that the sign of the estimated effect for the
-northern stock component differed depending on whether random effects on
-age 1 natural mortality were also included (\(\widehat \beta_R = -1\)
-for \(M_{6}\) or \(M_{13}\)) with the negative effect estimated with
-these random effects (\(M_{13}\)).
-
-\hypertarget{stock-size-fishing}{%
-\subsection*{Stock size, fishing}\label{stock-size-fishing}}
-\addcontentsline{toc}{subsection}{Stock size, fishing}
-
-For the best performing model \(M_1\), estimates of SSB Stock size is
-also increasing for the southern component, just not as quickly as the
-northern component. (plot of SSB time series for each component).
-
-Figure \ref{fig:M1-SSB-F-R}
-
-Figure \ref{fig:SSB-F-R-rel-M1} Annual estimates of SSB and F varied
-little among the 14 fitted models.
-
-plot of prior/posterior for movement rate.
-
-plot of average F by fleet,region.
-
-Sectivity random effects (Figure \ref{fig:selectivity-re})
-
-Estimates of effect size, sd, correlation for each peel (Table \ref{}
-
-The residual variation in the standard deviation of recruitment random
-effects is reduced because the expected recruitment (Eq.
-\ref{eq:expected-recruitment}) is a function of the covariate. This
-effect is included in the proposed base model (Figure
-\ref{fig:BT-Ecov-R}).
-
-plot of northern component predicted recruitment vs.~time with color by
-bottom temperature
+with approximately 20\% reduction in residual standard deviation when
+temperature effects were included. The residual variation is reduced
+because the expected recruitment (Eq. \ref{eq:expected-recruitment}) is
+a function of the covariate (Figure \ref{fig:BT-Ecov-R}).
+
+The posterior estimates of the bottom temperature covariate match the
+observations well because of the high precision of the observed
+anomalies (Figure \ref{fig:bottom-temperature}). The anomalies for the
+north and south regions appear highly correlated presumably because of
+the proximity of the two regions. Because the temperature anomalies are
+treated as latent variables, other data components in the model can
+affect the estimated anomalies when effects on recruitment are included
+(e.g., Miller et al. 2018), but in this case the estimates of anomalies
+are altered negligibly during the population model years when the effect
+on recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
+
+The effect of the anomalies on expected recruitment can be better
+observed by comparing the time series of the northern and southern
+components (Figure \ref{fig:E-R-SSB-F}). The expected recruitment for
+the southern component without any covariate effects is constant over
+time whereas the effect on expected recruitment for the northern
+component induces correlation of the bottom temperature anomalies and
+expected recruitment. For the northern stock component, most of the
+lowest estimated recruitment random effects occurred prior to 2000 and
+several large recruitment estimates occurred after 2010. For the
+southern stock, recruitment random effect estimates have been less
+variable and similar to the estimated median particularly since 2005.
+The estimated stock size of the northern component, as measured by SSB,
+increased substantially since 2005, whereas that for the southern
+component peaked in 2002. The increase in the northern component is also
+reflected in the total size of the combined components. Estimated
+average total fishing mortality rates across ages 6 and 7 (combining the
+commercial and recreational fleets) were highest in the southern region
+prior to 1998 and have declined substantially since then. Estimated
+fishing mortality rates in the northern region have been less variable,
+but had largest peaks in 1996, 1999, and 2005. The decline in estimates
+in the southern component is also reflected in the decline in total
+fishing mortality across regions.
+
+The estimated movement from the north to the south for the northern
+stock component is much less than anticipated from the prior
+distribution that was parameterized based on the companion Stock
+Synthesis model fit completed by the Working Group (Figure
+\ref{fig:move-prior-posterior}). Conversely, the estimated movement rate
+from south to north was more consistent and slightly greater than that
+from the companion model. The estimated movement rates varied little
+across all fitted models implying that the estimation of movement was
+not sensitive to the alternative assumptions we considered. Furthermore,
+the very low estimated movement rate from north to south is also
+consistent with preliminary fits of the WHAM model without the prior
+distribution where the movement rate was estimated at the lower bound of
+0 resulting in a lack of convergence. The estimated movement rates imply
+that a very small fraction of northern fish are ever in the southern
+region.
+
+The 2dAR1 random effects for the selectivity parameters of the northern
+recreational fleet indicate a decrease in selectivity at ages 2 to 6
+over the time series which corresponds to increases in the size limits
+in regulations by states where these catches occurred (Figure
+\ref{fig:selectivity-re}). The model also estimated a more modest
+decrease in selectivity at ages 2 to 3 for the northern commercial
+fleet. The selectivity at the youngest ages for the two fishery
+independent indices varied over time but without notable trend.
 
@@
 
+Using the expected recruitment rather than the random effects as the
+multiplier with equilibrium SSB/R results in less variable estimates of
+annual SSB\(_{40\%}\) (Figure \ref{fig:annual-BRPs}: top row). Estimates
+were less variable for the southern stock component than the northern
+stock component mainly because the expected recruitment is constant for
+the former whereas that for the northern component varies via the effect
+of the bottom temperature anomalies. Temporal variation for the southern
+stock component is mainly caused by annual variation in the inputs to
+the SSB/R calculation and \(F_{40\%}\), but there is also some effect of
+the variation in the proportion of total recruitment for each stock
+component as used in the weighting terms to define the 40\% reduction in
+unfished SSB/R (\(\lambda_s\) in Eq. \ref{eq:newton-obj}). Using the
+recruitment random effects instead of expected recruitment creates more
+uncertain estimates of SSB\(_{40\%}\) for the southern stock component
+whereas there is little difference for the northern component because
+there is less difference in the types of recruitment estimates resulting
+from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
+
+Ratios of SSB to SSB\(_{40\%}\) should be viewed with caution because
+the SSB reference point is being defined using annual recruitments
+rather than some form of average recruitment and therefore, ratios
+larger than 2.5 which would correspond to unfished population sizes are
+possible (Figure \ref{fig:annual-BRPs}: second row). However, for the
+southern component, using the constant expected recruitment is
+comparable to using median recruitment and the annual variation in the
+inputs to SSB/R indicates no years where SSB was greater than unfished
+levels throughout the time series. The patterns in uncertainty of SSB
+ratios are the same as those for the SSB reference points (Figure
+\ref{fig:annual-SSB-status-cvs}).
+
+Differences between annual \(F_{40\%}\) estimates using the alternative
+types of recruitment estimates are much less than those for
+SSB\(_{40\%}\) because the recruitment estimates are only used in the
+weighting of the stock-specific SSB/R to determine \(F_{40\%}\)
+(\(\lambda_s\) in Eq. \ref{eq:newton-obj}, Figure \ref{fig:annual-BRPs}:
+third row). For similar reasons, estimates of uncertainty of
+\(F_{40\%}\) are also similar between the type of recruitment estimates
+(Figure \ref{fig:annual-F40-cvs}). Annual estimates for the northern
+stock component were lowest in the earliest years of the time series
+whereas those for the soutern component are relatively stable over time.
+
+For a given type of recruitment used to calculate \(F_{40\%}\), the
+annual ratios of average \(F\) to the corresponding reference point are
+identical for each stock component and in total because of the
+selectivity used to define a single F reference point across both
+regions is exactly the same as that for the annual estimate of fishing
+mortality at age by fleet (Figure \ref{fig:annual-BRPs}: fourth row). As
+for annual \(F_{40\%}\), estimates of uncertainty in the ratios are
+similar whether expected recruitment or the random effects are used to
+calculate the reference points (Figure \ref{fig:annual-F-status-cvs}).
+
+Unlike the annual ratios of \(F\) to \(F_{40\%}\), the \(F_{40\%}\)
+under prevailing conditions uses a selectivity defined by the average
+\(F\) at age and fleet over the last 5 years, which is different from
+that of the selectivity in the terminal year and therefore those ratios
+for each region and in total are different (Figure
+\ref{fig:annual-BRPs}: bottom row). Although the alternative types of
+recruitment estimates result in different points on the kobe plots for
+joint status, the uncertainty in the estimates suggest similar
+probabilities of alternative quadrants. However, the usage of the
+confidence regions of the ratios to defined probabilities is an
+approximation because these ratios are functions of both empirical Bayes
+estimates of random effects as well as maximum likelihood estimates of
+fixed effects.
+
 \hypertarget{projections-1}{%
@@
 
-Projections of bottom temperature and recruitment and CVs (Figure
-\ref{fig:R-BT-proj})
-
-Uncertainty in both is much lower during the data years than the
-projection years.
-
-Uncertainty in BT RE propagates into both expected recruitment and
-estimated random effect for recruitment, but the latter also includes
-uncertainty due to the residual variability in recruitment.
+When continuing the AR1 model for the bottom temperature anomalies in
+the projection years we can see that the uncertainty grows and
+asymptotes to a level determined by the marginal variance of the process
+and that the predicted value approaches the estimated marginal mean
+value of approximately 0 (Figure \ref{fig:R-BT-proj}). There is no
+uncertainty in the projected linear trend because the values were
+specified as known whereas there is a small level of uncertainty in the
+projections of the recent average based on the uncertainty in the
+estimated mean. Comparing projected recruitment for the northern and
+southern stock components, the projected expected recruitment has much
+lower uncertainty than the projected random effects regardless of
+whether bottom temperature affects recruitment. For the northern
+component, CVs of projected expected recruitment are smaller than those
+of recruitment random effect, but there are larger differences between
+CVs of projected expected recruitment assuming AR1 and the other
+scenarios for the future bottom temperature anomalies are made (Figure
+\ref{fig:R-F-SSB-CVs}).
+
+Because projected fishing mortality is specified to be the same as that
+in the terminal year the same uncertainty is also propagated in the
+projections (Figure \ref{fig:F-SSB-proj}). As we would expect, there is
+no effect of alternative assumptions about bottom temperature in the
+projections for the southern stock component because the effect of the
+bottom temperature anomaly is only on the northern component. However,
+there alternative bottom temperature assumptions do affect projections
+of northern component SSB and the projections do not stabilize when the
+anomalies are projected to increase linearly. When the same AR1
+assumption for the bottom temperature anomalies is used in projections,
+CVs of northern and total SSB projections are larger than when the
+recent average or linear trend are assumed (Figure
+\ref{fig:R-F-SSB-CVs})
+
+Unlike the annual reference points during the model years (prior to
+2022), components of the SSB/R calculations are constant in the
+projection years where the only changing annual values are the
+recruitments used in the weighting for \(F_{40\%}\) and the recruitments
+used to calculate the resulting SSB\(_{40\%}\) (Figure
+\ref{fig:brps-proj}). Estimated reference points in the first few
+projection years can differ depending on whether the expected
+recruitment or the random effects are used in their calculation, but
+longer term projected values are identical because the random effect
+estimates approach the expected values as the influence of observations
+diminishes. However, the estimates of uncertainty in the longer term
+reference points is lower using the expected recruitment rather than the
+random effects (Figures \ref{fig:annual-F40-cvs} and
+\ref{fig:annual-SSB40-cvs}). The reduction in uncertainty using expected
+recruitment is less for \(F_{40\%}\) than SSB\(_{40\%}\) and for the
+northern component, that includes bottom temperature effects, than the
+southern component.
 
@@
 
+\hypertarget{black-sea-bass}{%
+\subsection*{Black sea bass}\label{black-sea-bass}}
+\addcontentsline{toc}{subsection}{Black sea bass}
+
+In our investigation of bottom temperature effects on recruitment and
+natural mortality and time-varying random effects on natural mortality
+at age 1, we found only including effects on recruitment for the
+northern stock component to be the best model with respect to AIC. Black
+sea bass in the NEUS is at the northern extent of its range and the
+estimated increase in expected recruitment with temperature for the
+northern stock component is consistent with its range extension to the
+north. Opposite effects of temperature on recruitment or population size
+would be expected for species in the same general area that are at the
+southern extent of their range (Gabriel 1992). For example, higher
+recruitment of the Southern New England-Mid-Atlantic stock of yellowtail
+flounder is correlated with more cold water persistence into summer fall
+on the Northeast US shelf Miller et al. (2016b).
+
+Our finding regarding bottom temperature effects is consistent with the
+bottom temperature effects on recruitment in the assessment model
+accepted during the peer-review process and which is currently used to
+assess the stock (NEFSC 2023). However, AIC suggested some weight for
+models with temperature effects on recruitment for the southern
+component and random effects on M at age 1. Although this might suggest
+making inferences on the population by using an ensemble of these and
+weighting by AIC (Burnham and Anderson 2002), the estimates of
+assessment outputs relevant for management (e.g., SSB and \(F\)) were
+similar among the models and weighted estimates would differ very little
+from the estimates with the lowest AIC.
+
+We showed that using expected recruitment rather than recruitment random
+effects to define BRPs can produce the same estimates but with lower
+uncertainty. Using expected bottom temperature anomalies rather than the
+random effects would also result in the same BRPs, but presumably with
+even less uncertainty than those we showed using expected recruitment.
+These results can be important for management when uncertainty in
+assessment output is used in making catch advice. The typical approach
+for defining stock and harvest status is to compare BRPs defined under
+prevailing conditions with current stock and harvest levels, but we
+could also make the comparisons with projected BRPs like those we
+demonstrated are possible. The uncertainty in these ratios could also be
+used in defining catch advice.
+
+As more assessments begin including effects of environmental covariates
+explicitly in assessment models, assumptions about these covariates in
+calculated BRPs and population projections becomes an important decision
+in the management process. Elaborate!
+
 \hypertarget{general-model-aspects}{%
@@
 
-WHAM provides a comprehensive treatment of environmental covariates and
-their effects on populations using state-space methods (Maunder 2024).
-This framework accounts for the magnitude of differing uncertainties in
-observations and stochastic population dynamics processes. This version
-2.0 extends these inferences to movement rates for multi-region models.
+WHAM provides a large class of models for time- and age-varying random
+effects for population attributes and treatment of environmental
+covariates and their effects on populations using state-space methods
+(Maunder 2024). This framework accounts for the magnitude of differing
+uncertainties in observations and stochastic population dynamics
+processes. Version 2.0 extends these inferences to movement rates for
+multi-region models and stock (component) specific configuration of
+random and environmental covariate effects.
 
@@
 observations of any type. Such observations are critical to estimation
-of movement parameters (e.g., Goethel et al. 2019), but also can inform
-mortality rate parameters (Hampton 1991). It is well known that natural
-mortality is seldom estimated in assessment models because the observed
-data often provide little information to distinguish natural mortality
-from other assessment model parameters (Lee et al. 2011; Clark 2022).
-Estimation of natural mortality may be even more challenging within
-state-space assessment models with their greater flexibility from
-inclusion of time-varying random effects. For example, we fixed the mean
-natural mortality for age 1 fish and Cadigan (2016) and Stock et al.
-(2021) also estimated natural mortality deviations. Therefore allowing
-tagging observations in WHAM should be a high priority even for models
-with a single stock and region.
-
-The work by Correa et al. (2023) on incorporating length information and
-modeling growth within WHAM has not yet been merged into the multi-stock
-version 2.0. Growth and movement would be challenging.
-
-\hypertarget{black-sea-bass}{%
-\subsection*{Black sea bass}\label{black-sea-bass}}
-\addcontentsline{toc}{subsection}{Black sea bass}
-
-In our investigation of bottom temperature effects on recruitment and
-natural mortality and time-varying random effects on natural mortality
-at age 1, we found only including effects on recruitment for the
-northern stock component to be the best model with respect to AIC. This
-is the same configuration configuration accepted during the assessment
-peer-review process and it is the model currently used for management
-(NEFSC 2023).
-
-Although not relevant to the best performing model, the effect of bottom
-temperature on age 1 M would be expected to be opposite of that for
-recruitment because lower M with higher temperature would produce higher
-recruitment. This expected effect was estimated only when random effects
-on age 1 M were also included in the model. This seems to provide
-further support to including random effects when examining covariate
-effects on population parameters as demonstrated by Miller et al. (2018)
-for effects on growth parameters.
-
-AIC suggested some weight for models with temperature effects on
-recruitment for the southern component and random effects on M at age 1.
-This might suggest making inferences on the population by using an
-ensemble of these and weighting by AIC (Burnham and Anderson 2002).
-However, the estimates of assessment outputs relevant for management
-(SSB, F) were similar among the models and weighted estimates would
-differ very little from the estimates with the lowest AIC. {[}but what
-about reference points?{]}
-
-\hypertarget{temperature-effects}{%
-\subsubsection*{Temperature effects}\label{temperature-effects}}
-\addcontentsline{toc}{subsubsection}{Temperature effects}
-
-BSB in the NEUS is at the northern extent of its range. hence the range
-extension to the north with increased temperature. The evidence for
-temperature effects on recruitment was strong only for the northern
-component which is the component of the stock at the limit of the
-species' range. Opposite effects of temperature on recruitment or
-population size would be expected for species in the same general area
-that are at the southern extent of their range (Gabriel 1992). For
-example, higher recruitment of the Southern New England-Mid-Atlantic
-stock of yellowtail flounder is correlated with more cold water
-persistence into summer fall on the Northeast US shelf Miller et al.
-(2016b).
-
-\hypertarget{self-tests}{%
-\subsubsection*{Self Tests}\label{self-tests}}
-\addcontentsline{toc}{subsubsection}{Self Tests}
-
-Simulation self tests performed by Stock and Miller (2021) simulated
-process errors and observation errors and their simulations showed
-little bias in estimated assessment output such as SSB. Li et al. (2024)
-used the same type of simulations and found what?.
-
-Here we conditioned on the estimated process errors for black sea bass.
-This conditional approach has been suggested as appropriate for tactical
-management where we are interested in inferences assuming our estimation
-of the population and fishing history from the original data is the
-truth Cadigan et al. (2024). We found some bias in SSB estimation for
-the black sea bass application (negatively for the northern stock),
-using the default model. We also found instability in the estimation of
-the observation variance parameters for the aggregate indices such that
-they often were estimated at the lower bound of 0. However, we found
-that the magnitude of bias was largely influenced by the large
-uncertainty estimated in some of the age composition observations
-assumed to have a logistic-normal distribution. When using the
-logistic-normal likelihood for management, we recommend inspecting the
-size of the estimated (marginal) standard deviation of the logistic
-normal model when self-test bias is non-negligible. We also recommend
-investigating the multiplicative transformation for the logistic normal
-rather than the additive transformation which is currently the only
-option in WHAM (Cadigan 2016).
-
-?Selectivity RE are only on index age comp in the north which may be why
-bias is larger there?
-
-? We found REML provided less biased estimation of SSB and recruitment
-which are functions of random effects. Whether REML should be the
-default approach for estimating these quantities from state-space stock
-assessment models requires further investigation. For example,
-improvement in bias may be negligible when data contain low observation
-error and/or there are few fixed effects parameters. ? larger variance
-estimates allows greater spread of random effects estimates. Important
-for conditional self-test where estimated RE are of primary importance
-(Recruitment and SSB). When simulating RE and refitting, the sign of
-differences between estimates and true values can be random across
-simulations. Probably more important with more fixed effects and less
-informative data
-
-Including tagging data in WHAM would also be helpful for BSB because we
-are using the same data to estimate movement outside of the model as we
-are using to estimate the other parameters in WHAM.
+of movement parameters (e.g., Goethel et al. 2019). We used estimates of
+movement parameters from a companion assessment model to define prior
+distributions for corresponding movement parameters in the state-space
+black sea bass model, but other than the tagging data, both of the
+assessment models used much the same data which can lead to
+inappropriate inferences. A better approach would be to estimate the
+movement parameters externally from just from the tagging data.
+
+Tagging data can also inform mortality rate parameters (Hampton 1991).
+It is well known that natural mortality is seldom estimated in
+assessment models because the observed data often provide little
+information to distinguish natural mortality from other assessment model
+parameters (Lee et al. 2011; Clark 2022). Estimation of natural
+mortality may be even more challenging within state-space assessment
+models with their greater flexibility from inclusion of time-varying
+random effects. For example, we fixed the mean natural mortality for age
+1 fish and Cadigan (2016) and Stock et al. (2021) also estimated natural
+mortality deviations. Therefore allowing tagging observations in WHAM
+should be a high priority even for models with a single stock and
+region. This extension would also be beneficial for black sea bass
+because the current assessment uses much the same data to estimate
+movement outside of the model as we are using to estimate the other
+parameters in the WHAM model.
+
+WHAM version 2.0 only currently allows models where each stock spawns
+and recruits in a single region however, many stocks are assessed with
+spatial units that are assumed to have a global stock and recruitment
+model that distributes recruitment to each region (references).
+Therefore, extending WHAM to allow such models would be beneficial.
+Similarly age composition data are scarce for many stocks incorporating
+the work by Correa et al. (2023) on allowing length composition and
+modeling growth within WHAM would provide an assessment tool to include
+process errors in assessment with such data and possibly with multiple
+stocks and regions.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-ablehales97}{}}%
+Able, K.W., and Stanton Hales, Jr., L. 1997. Movements of juvenile black
+sea bass \emph{{C}entropristis striata} (linnaeus) in a southern {N}ew
+{J}ersey estuary. Journal of Experimental Marine Biology and Ecology
+\textbf{213}(2): 153--167.
+doi:\href{https://doi.org/10.1016/S0022-0981(96)02743-8}{10.1016/S0022-0981(96)02743-8}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-andersenkeiding02}{}}%
@@
 91--115.
+doi:\href{https://doi.org/10.1191/0962280202SM276ra}{10.1191/0962280202SM276ra}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-cadiganetal24}{}}%
-Cadigan, N.G., Albertsen, C.M., Zheng, N., and Nielsen, A. 2024. Are
-state-space stock assessment model confidence intervals accurate? Case
-studies with SAM and barents sea stocks. Fisheries Research
-\textbf{272}: 106950.
-doi:\href{https://doi.org/10.1016/j.fishres.2024.106950}{10.1016/j.fishres.2024.106950}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-caoetal14}{}}%
@@
 Analysis \textbf{5}: 315--327.
+doi:\href{https://doi.org/10.1023/A:1009636125294}{10.1023/A:1009636125294}.
 
@@
 doi:\href{https://doi.org/10.1093/icesjms/fsad133}{10.1093/icesjms/fsad133}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-fabrizioetal13}{}}%
+Fabrizio, M.C., Manderson, J.P., and Pessutti, J.P. 2013. Habitat
+associations and dispersal of black sea bass from a {M}id-{A}tlantic
+{B}ight reef. Marine Ecology Progress Series \textbf{482}: 241--253.
+doi:\href{https://doi.org/10.3354/meps10302}{10.3354/meps10302}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-kapuretal21}{}}%
+Kapur, M.S., Siple, M.C., Olmos, M., Privitera-Johnson, K.M., Adams, G.,
+Best, J., Castillo-Jordán, C., Cronin-Fine, L., Havron, A.M., Lee, Q.,
+Methot, R.D., and Punt, A.E. 2021. Equilibrium reference point
+calculations for the next generation of spatial assessments. Fisheries
+Research \textbf{244}: 106132.
+doi:\href{https://doi.org/10.1016/j.fishres.2021.106132}{10.1016/j.fishres.2021.106132}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-kemenysnell60}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lietal24}{}}%
-Li, C., Deroba, J.J., Miller, T.J., Legault, C.M., and Perretti, C.T.
-2024. An evaluation of common stock assessment diagnostic tools for
-choosing among state-space models with multiple random effects
-processes. Fisheries Research \textbf{273}: 106968.
-doi:\href{https://doi.org/10.1016/j.fishres.2024.106968}{10.1016/j.fishres.2024.106968}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-maunder24}{}}%
@@
 Fisheries Research \textbf{142}(1): 86--99.
+doi:\href{https://doi.org/10.1016/j.fishres.2012.10.012}{10.1016/j.fishres.2012.10.012}.
 
@@
 Journal of Marine Science \textbf{56}(4): 473--488.
+doi:\href{https://doi.org/10.1006/jmsc.1999.0481}{10.1006/jmsc.1999.0481}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mosershepherd09}{}}%
+Moser, J., and Shepherd, G.R. 2009. Seasonal distribution and movement
+of black sea bass (\emph{{C}entropristis striata}) in the {N}orthwest
+{A}tlantic as determined from a mark-recapture experiment. Journal of
+Northwest Atlantic Fishery Science \textbf{40}: 17--28.
+
+\leavevmode\vadjust pre{\hypertarget{ref-musickmercer77}{}}%
+Musick, J.A., and Mercer, L.P. 1977. Seasonal distribution of black sea
+bass, \emph{{C}entropristis striata}, in the {M}id-{A}tlantic {B}ight
+with comments on the ecology and fisheries of the species. Transactions
+of the American Fisheries Society \textbf{106}(1): 12--25.
+doi:\href{https://doi.org/10.1577/1548-8659(1977)106\%3C12:SDOBSB\%3E2.0.CO;2}{10.1577/1548-8659(1977)106\textless12:SDOBSB\textgreater2.0.CO;2}.
 
@@
 perspective. Fisheries Research \textbf{229}: 105617.
-doi:\url{https://doi.org/10.1016/j.fishres.2020.105617}.
+doi:\href{https://doi.org/10.1016/j.fishres.2020.105617}{10.1016/j.fishres.2020.105617}.
 
@@
 Fisheries Oceanography \textbf{14}(5): 386--399.
+doi:\href{https://doi.org/10.1111/j.1365-2419.2005.00343.x}{10.1111/j.1365-2419.2005.00343.x}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-thygesenetal17}{}}%
-Thygesen, U.H., Albertsen, C.M., Berg, C.W., Kristensen, K., and
-Nielsen, A. 2017. {Validation of ecological state space models using the
-Laplace approximation}. Environmental and Ecological Statistics
-\textbf{24}(2): 317--339.
-doi:\href{https://doi.org/10.1007/s10651-017-0372-4}{10.1007/s10651-017-0372-4}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-trijouletetal23}{}}%
-Trijoulet, V., Albertsen, C.M., Kristensen, K., Legault, C.M., Miller,
-T.J., and Nielsen, A. 2023. Model validation for compositional data in
-stock assessment models: Calculating residuals with correct properties.
-Fisheries Research \textbf{257}: 106487.
-doi:\url{https://doi.org/10.1016/j.fishres.2022.106487}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-varkeyetal22}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-zhengcadigan23}{}}%
-Zheng, N., and Cadigan, N. 2023. Frequentist conditional variance for
-nonlinear mixed-effects models. Journal of Statistical Theory and
-Practice \textbf{17}(3).
-doi:\href{https://doi.org/10.1007/s42519-022-00304-5}{10.1007/s42519-022-00304-5}.
-
 \end{CSLReferences}
@@
 \pagebreak
-
-\hypertarget{appendix-a}{%
-\section*{Appendix A}\label{appendix-a}}
-\addcontentsline{toc}{section}{Appendix A}
-
-\setcounter{table}{0}
-\renewcommand\thetable{A\arabic{table}}
-\input{symboltable.tex}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{Ecov}
+{\centering \includegraphics[height=0.95\textheight]{E_R_SSB_F_fig}
 
@@
 
-\caption{Observations and 95\% confidence intervals (points with vertical lines) and posterior estimates (lines) with 95\% confidence intervals (polygons) of bottom temperature anomalies in the north and south regions from model $M_0$.}\label{fig:bottom-temperature}
+\caption{Annual observations and posterior estimates of bottom temperature anomalies (top row), annual expected and random effect recruitment estimates (second row) and SSB and average fishing mortality at ages 6 and 7 from model $M_1$.
+Columns define estimates regionally (spawning region for SSB) and toal SSB and fishing mortality across regions.
+Polygons and vertical lines represent 95\% confidence intervals.}\label{fig:E-R-SSB-F}
 \end{figure}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{best_R_Ecov}
-
-}
-
-\caption{Expected and Posterior Emperical Bayes estimates of annual recruitment for the northern stock component.
-Color defined by the corresponding annual bottom temperature anomaly.}\label{fig:BT-Ecov-R}
-\end{figure}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{M_1_SSB_F_R}
-
-}
-
-\caption{Annual estimates of SSB, average F (ages 6-7), and recruitment from model $M_1$. Polygons represent 95\% confidence intervals.}\label{fig:M1-SSB-F-R}
-\end{figure}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{R_proj_results}
-
-}
-
-\caption{Estimates of bottom temperature, recruitment, and expected recruitment, and coefficients of variation for the latter two from model $M_1$ assuming alternative bottom temperature scenarios described in the text.
-Polygons represent 95\% confidence intervals.}\label{fig:R-BT-proj}
-\end{figure}
-
 \pagebreak
@@
 
+\begin{figure}
+
+{\centering \includegraphics[height=0.85\textheight]{brp_status_results}
+
+}
+
+\caption{Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio, equilibrium SSB at $F_{40\%}$ based on annual inputs to SSB/R calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to SSB/R averaged over the last 5 years and recruitment is averaged over the time series.
+Red and blue in first 4 rows and black and gray in final row indicate alternative annual estimates of recruitment (random effect or expected).
+All results are based on model $M_1$.
+Polygons represent 95\% confidence intervals or regions.}\label{fig:annual-BRPs}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_ecov_Recruit_results}
+
+}
+
+\caption{Annual estimates of bottom temperature in the northern region and alternative recruitment estimates (random effect or expected) by region.
+Estimates from years after 2021 are from projections of model $M1$ under three alternative assumptions for the bottom temperature anomalies in the northern region.
+Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:R-BT-proj}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_F_SSB}
+
+}
+
+\caption{Annual estimates of average fishing mortality and SSB by region and in total.
+Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
+Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:F-SSB-proj}
+\end{figure}
+
+\begin{landscape}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_brps}
+
+}
+
+\caption{Annual estimates by region and in total of average equilibrium $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio and SSB at $F_{40\%}$.
+Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region and average SSB/R inputs from the last 5 years of the unprojected model.
+Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:brps-proj}
+\end{figure}
+
+\end{landscape}
+
 \clearpage
+
+\setcounter{table}{0}
+\renewcommand\thetable{\arabic{table}}
 
@@
 
-\setcounter{table}{0}
-\renewcommand\thetable{\arabic{table}}
-
-\begin{table}
-
+\begin{table}[!h]
+\centering
 \caption{\label{tab:model-desc-table}Assumptions for temperature effects and random effects for age 1 natural mortality for each model.}
@@
 $M_{13}$ & 17.70 & 16.91 & 15.43 & 14.46 & 14.64 & 13.55 & 13.07 & 12.88\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:aic-wts-table}Model AIC weights for each retrospective peel.}
-\centering
-\begin{tabular}[t]{lrrrrrrrr}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Peel} \\
-\cmidrule(l{3pt}r{3pt}){2-9}
-Model & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
-\midrule
-$M_{0}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.01 & 0.01\\
-$M_{1}$ & 0.45 & 0.43 & 0.42 & 0.38 & 0.41 & 0.37 & 0.36 & 0.38\\
-$M_{2}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{3}$ & 0.30 & 0.31 & 0.25 & 0.23 & 0.24 & 0.23 & 0.25 & 0.23\\
-$M_{4}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-\addlinespace
-$M_{5}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{6}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{7}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{8}$ & 0.15 & 0.15 & 0.20 & 0.24 & 0.21 & 0.24 & 0.22 & 0.23\\
-$M_{9}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-\addlinespace
-$M_{10}$ & 0.10 & 0.11 & 0.12 & 0.14 & 0.13 & 0.15 & 0.15 & 0.14\\
-$M_{11}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{12}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{13}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
 \bottomrule
@@
 
+\hypertarget{appendix-a}{%
+\section*{Appendix A}\label{appendix-a}}
+\addcontentsline{toc}{section}{Appendix A}
+
+\setcounter{table}{0}
+\renewcommand\thetable{A\arabic{table}}
+\input{symboltable.tex}
+
+\pagebreak
+
 \hypertarget{supplemental-materials}{%
@@
 
-The working group fit a Stock Synthesis model (Methot and Wetzel 2013)
+The Working Group fit a Stock Synthesis model (Methot and Wetzel 2013)
 that included tagging data with 2 seasons (6 months each) and 2 regions
@@
 diagonal matrix of corresponding eigenvalues of
-\(\boldsymbol{\mu}^*_k\). The working group used a parametric bootstrap
+\(\boldsymbol{\mu}^*_k\). The Working Group used a parametric bootstrap
 approach to determine an appropriate standard deviation for the prior
@@
 \hypertarget{jitter-fits-for-model-m_0}{%
-\subsubsection{\texorpdfstring{Jitter fits for model
+\subsubsection*{\texorpdfstring{Jitter fits for model
 \(M_0\)}{Jitter fits for model M\_0}}\label{jitter-fits-for-model-m_0}}
+\addcontentsline{toc}{subsubsection}{Jitter fits for model \(M_0\)}
 
@@
 \hypertarget{jitter-fits-for-model-m_1}{%
-\subsubsection{\texorpdfstring{Jitter fits for model
+\subsubsection*{\texorpdfstring{Jitter fits for model
 \(M_1\)}{Jitter fits for model M\_1}}\label{jitter-fits-for-model-m_1}}
+\addcontentsline{toc}{subsubsection}{Jitter fits for model \(M_1\)}
 
@@
 \hypertarget{self-test-for-model-m_1}{%
-\subsubsection{\texorpdfstring{Self test for model
+\subsubsection*{\texorpdfstring{Self test for model
 \(M_1\)}{Self test for model M\_1}}\label{self-test-for-model-m_1}}
+\addcontentsline{toc}{subsubsection}{Self test for model \(M_1\)}
 
@@
 \ref{fig:self-test-fig}).
-
-\hypertarget{one-step-ahead-residuals-for-model-m_1}{%
-\subsubsection{\texorpdfstring{One-step-ahead residuals for model
-\(M_1\)}{One-step-ahead residuals for model M\_1}}\label{one-step-ahead-residuals-for-model-m_1}}
-
-One-step-ahead residuals can now be calculated for all index, catch, and
-environmental covariate observations using methods described by Thygesen
-et al. (2017) and Trijoulet et al. (2023).
 
@@
 
+\begin{table}
+
+\caption{\label{tab:aic-wts-table}Model AIC weights for each retrospective peel.}
+\centering
+\begin{tabular}[t]{lrrrrrrrr}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Peel} \\
+\cmidrule(l{3pt}r{3pt}){2-9}
+Model & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
+\midrule
+$M_{0}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.01 & 0.01\\
+$M_{1}$ & 0.45 & 0.43 & 0.42 & 0.38 & 0.41 & 0.37 & 0.36 & 0.38\\
+$M_{2}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
+$M_{3}$ & 0.30 & 0.31 & 0.25 & 0.23 & 0.24 & 0.23 & 0.25 & 0.23\\
+$M_{4}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
+\addlinespace
+$M_{5}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
+$M_{6}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
+$M_{7}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
+$M_{8}$ & 0.15 & 0.15 & 0.20 & 0.24 & 0.21 & 0.24 & 0.22 & 0.23\\
+$M_{9}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
+\addlinespace
+$M_{10}$ & 0.10 & 0.11 & 0.12 & 0.14 & 0.13 & 0.15 & 0.15 & 0.14\\
+$M_{11}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
+$M_{12}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
+$M_{13}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \setcounter{figure}{0}
@@
 \end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{BTA_full_fig}
+
+}
+
+\caption{Observations with 95\% confidence intervals (points with vertical lines) and posterior estimates with 95\% confidence intervals (lines with polygons) of bottom temperature anomalies in the north and south regions from model $M_1$.
+Gray vertical line defines the first year that the black sea bass stock is modeled.}\label{fig:bottom-temperature}
+\end{figure}
+
 \end{landscape}
+\pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{best_R_Ecov}
+
+}
+
+\caption{Expected and random effect recruitment estimates for the northern stock component. Color of points defined by the corresponding annual bottom temperature anomaly.}\label{fig:BT-Ecov-R}
+\end{figure}
+\pagebreak
 
@@
 
-\pagebreak
+\begin{landscape}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.9\textheight]{R_SSB_F_cv_results}
+
+}
+
+\caption{Coefficients of variation for estimates of alternative recruitment estimates (random effects or expected), average fishing mortality at age 6 and 7, and SSB by region and in total from model $M_1$.
+Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:R-F-SSB-CVs}
+\end{figure}
+\end{landscape}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_F40_CV}
+
+}
+
+\caption{Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-F40-cvs}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_SSB40_CV}
+
+}
+
+\caption{Coefficients of variation for annual equilibrium SSB$_{40\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations and alternative annual recruitment types.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-SSB40-cvs}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_F_status_results}
+
+}
+
+\caption{Annual estimates of ratios of fishing mortality to $F_{40\%}$ by region and in total.
+Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
+Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:F-status-proj}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_SSB_status_results}
+
+}
+
+\caption{Annual estimates of ratios of SSB to SSB$_{40\%}$ by region and in total.
+Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
+Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:SSB-status-proj}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_SSB_status_CV}
+
+}
+
+\caption{Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-SSB-status-cvs}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_F_status_CV}
+
+}
+
+\caption{Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\bar{F}_{40\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-F-status-cvs}
+\end{figure}
 

### 2025-01-02 paper/symboltable.tex
saving work

@@
 $\theta_{r\rightarrow r',i}$ & mean parameter across age and year for movement from region $r$ to region $r'$ in seasonal interval $i$\\
-$\epsilon_{r\rightarrow r',y,a,i}$ &  random effect parameter for movement from region $r$ to region $r'$ in seasonal interval $i$ and year $y$ at age $a$\\
+$\varepsilon_{r\rightarrow r',y,a,i}$ &  random effect parameter for movement from region $r$ to region $r'$ in seasonal interval $i$ and year $y$ at age $a$\\
 $n_E$ & number of environmental covariates\\
@@
 $\theta_{M,r}$ & mean parameter across age and year for nutural mortality in region $r$\\
-$\epsilon_{M,r,y,a}$ & random effect parameter for natural mortality in region $r$ and year $y$ at age $a$\\
+$\varepsilon_{M,r,y,a}$ & random effect parameter for natural mortality in region $r$ and year $y$ at age $a$\\
 $\beta_{M,r,a,k}$ & effect of environmental covariate $k$ on natural mortality in region $r$ at age $a$\\

### 2025-01-02 R/make_figures.R

@@ -1,7 +1,13 @@
 library(wham)
 library(ggplot2)
 library(dplyr)
+library(patchwork)
 source(here::here("R","plot_functions.R"))
+
+theme_set(theme_bw())
+theme_update(strip.background = element_blank(), strip.placement = "outside", strip.text = element_text(size = rel(2)), 
+      axis.title = element_text(size = rel(2)), axis.text = element_text(size = rel(2)), legend.text = element_text(size = rel(2)), #text = element_text(size = rel(2)), 
+      legend.title = element_text(size = rel(2)))
 
 fits <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))
 fits_M_re <- readRDS(file.path("results","fits_M_re_better.RDS"))
@@ -64,7 +70,7 @@ dev.off()
 
 
 cairo_pdf(file.path("paper", paste0("Ecov.pdf")), height = 10, width = 10)
-plot.ecov(fits[[1]])
+plot.ecov(fits[[2]])
 legend("topleft", legend = c("North", "South"), col = mypalette(2), lty =  1, pch = 19)
 dev.off()
 
@@ -138,7 +144,7 @@ df <- rbind(df, make_df(c(fits,fits_M_re), rep_name = "log_Fbar", label = "Avg. 
 temp_df <- make_df(c(fits,fits_M_re), rep_name = "log_NAA_rep", label = "Recruitment (Millions)")
 temp_df[,c("est","lo","hi")] <-temp_df[,c("est","lo","hi")]/1000
 df <- rbind(df,temp_df)
-
+df_M1 <- subset(df, model == levels(df$model)[2])
 plt <- ggplot(subset(df, model == levels(df$model)[2]), aes(x = year, y = est)) + scale_colour_viridis_d() + 
     geom_point(size = 1) + geom_line() +
     facet_grid(type ~ stock, scales = "free_y", switch = "y") + theme_bw() +
@@ -152,18 +158,31 @@ cairo_pdf(file.path("paper", "M_1_SSB_F_R.pdf"), width = 12, height = 16)
 plt
 dev.off()
 
-df <- cbind(df, rel_M1_est = NA)
+df_rel <- cbind(df, rel_M1_est = NA)
 for(i in unique(df$model)) for(j in unique(df$type)) for(k in unique(df$stock)) {
[…]

### 2025-01-02 R/plot_Ecov_R_SSB_F.R

@@ -1,0 +1,142 @@
+library(wham)
+library(ggplot2)
+library(dplyr)
+library(patchwork)
+source(here::here("R","plot_functions.R"))
+
+theme_set(theme_bw())
+theme_update(strip.background = element_blank(), strip.placement = "outside", strip.text = element_text(size = rel(2)), 
+      axis.title = element_text(size = rel(2)), axis.text = element_text(size = rel(2)), legend.text = element_text(size = rel(2)), #text = element_text(size = rel(2)), 
+      legend.title = element_text(size = rel(2)))
+
+fit <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))[[2]] #M_1
+proj_1 <- readRDS(here::here("results",paste0("m1_proj_",1,".RDS")))
+proj_1_fjp <- readRDS(here::here("results",paste0("m1_proj_",1,"_sdrep_fjp.RDS")))
+proj <- readRDS(here::here("results","m1_proj_3_R_opt1.RDS"))
+pred_R_N <- make_pred_R_cv(proj_1, proj_1_fjp, type = 1)
+pred_R_S <- make_pred_R_cv(proj_1, proj_1_fjp, type = 1, region = 2)
+
+
+df_use <- cbind.data.frame(Year = proj$input$years_Ecov, name = "Ecov_RE", val = TMB:::as.list.sdreport(proj$sdrep, report = TRUE, what = "Est")$Ecov_x[,1], 
+	se = TMB:::as.list.sdreport(proj$sdrep, report = TRUE, what = "Std")$Ecov_x[,1], Region = "North")
+df_use <- rbind(df_use, cbind.data.frame(Year = proj$input$years_Ecov, name = "Ecov_RE", val = TMB:::as.list.sdreport(proj$sdrep, report = TRUE, what = "Est")$Ecov_x[,2], 
+	se = TMB:::as.list.sdreport(proj$sdrep, report = TRUE, what = "Std")$Ecov_x[,2], Region = "South"))
+df_use <- rbind(df_use, cbind.data.frame(Year = proj$input$years_Ecov, name = "Ecov_obs", val = proj$input$data$Ecov_obs[,1], se = exp(proj$parList$Ecov_obs_logsigma[,1]), Region = "North"))
[…]

### 2025-01-02 R/plot_brps_status.R

@@ -1,0 +1,207 @@
+library(wham)
+library(ggplot2)
+library(dplyr)
+library(patchwork)
+source(here::here("R","plot_functions.R"))
+theme_set(theme_bw())
+theme_update(strip.background = element_blank(), strip.placement = "outside", strip.text = element_text(size = rel(2)), 
+      axis.title = element_text(size = rel(2)), axis.text = element_text(size = rel(2)), legend.text = element_text(size = rel(2)), #text = element_text(size = rel(2)), 
+      legend.title = element_text(size = rel(2)))
+
+proj_R1 <- lapply(1:3, \(x) readRDS(here::here("results",paste0("m1_proj_",x,"_R_opt1.RDS"))))
+proj_R3 <- lapply(1:3, \(x) readRDS(here::here("results",paste0("m1_proj_",x,"_R_opt3.RDS"))))
+se_R1 <- lapply(proj_R1, \(x) TMB:::as.list.sdreport(x$sdrep, what = "Std", report = TRUE))
+se_R3 <- lapply(proj_R3, \(x) TMB:::as.list.sdreport(x$sdrep, what = "Std", report = TRUE))
+
+#have to do se(predR) by hand
+source(here::here("R","plot_functions.R"))
+temp <- make_pred_R_cv(proj_R3[[1]], type = 1)
+temp <- rbind(temp, make_pred_R_cv(proj_R3[[1]], type = 1, region = 2))
+temp <- rbind(temp,make_pred_R_cv(proj_R3[[2]], type = 2))
+temp <- rbind(temp,make_pred_R_cv(proj_R3[[2]], type = 2, region = 2))
+temp <- rbind(temp,make_pred_R_cv(proj_R3[[3]], type = 3))
+temp <- rbind(temp,make_pred_R_cv(proj_R3[[3]], type = 3, region = 2))
+
+source(here::here("R","plot_functions.R"))
+brp_results <- lapply(c(proj_R1,proj_R3), get.brp.status.results)
+
+ny_full <- length(proj_R3[[1]]$years_full)
+df_annual <- cbind.data.frame(Year = proj_R3[[1]]$years_full, 
+  val = exp(c(sapply(c(proj_R1,proj_R3), \(x) c(x$rep$log_SSB_FXSPR)))),
+  cv = c(sapply(c(se_R1,se_R3), \(x) c(x$log_SSB_FXSPR))),
[…]

### 2025-01-02 R/plot_functions.R

@@ -49,6 +49,9 @@ plot.ecov <- function(mod, plot.pad = FALSE, do.tex=FALSE, do.png=FALSE, fontfam
     lines(years_full, ecov.pred[,i], col=plot.colors[i], lwd=3)
     if(length(years_full) > length(years)) abline(v=tail(years,1), lty=2)
   }
+  return(cbind.data.frame(Year = years_full, est = c(ecov.obs, ecov.pred), lo = c(ecov.low, ecov.pred.low), hi = c(ecov.high, ecov.pred.high), 
+    region = rep(c("North", "South"), each = length(years_full)),
+    type = rep(c("obs","pred"), each = 2*length(years_full))))
 }
 
 self_test_i <- function(i, fit_RDS = NULL, seeds = NULL, conditional = TRUE, map_change = NULL, res_dir = NULL, wham_location = NULL, test_dir = NULL, save_inputs = FALSE, 
@@ -242,68 +245,122 @@ plot_SSB <- function(sims, rep_name = "SSB", indices = 1:2, index_names = c("Nor
   }
 }
 
-make_pred_R_cv <- function(proj, fjp, type = 1){
+whichList <- function(mod, par = mod$env$last.par.best) {
+  nms <- names(mod$env$parameters)
+  which_by_name <- function(x) {
+    which_par <- which(names(par) == x)
+    y <- attr(mod$env$parameters[[x]],"shape")
+    if(is.null(y)) { #all estimated
+      y <- mod$env$parameters[[x]] 
+      y[] <- NA
+      y[] <- which_par
+    } else { #mapping used, not all estimated and/or some commonly estimated
+      y[] <- NA
+      f <- attr(mod$env$parameters[[x]], "map")
+      ind_unique <- unique(f[which(f>=0)]) #location of each unique estimated par
+      if(length(which_par)) for(l in 1:length(ind_unique)) {
+        y[which(f == ind_unique[l])] <- which_par[ind_unique[l]+1] #+1 because starts at 0
+      }
+    }
+    y
+  }
+  out <- lapply(nms,which_by_name)
+  names(out) <- nms
+  return(out)
+}
 
-  x <- TMB:::as.list.sdreport(proj$sdrep, report = FALSE, what = "Est")
[…]

### 2025-01-02 R/plot_projections.R

@@ -1,0 +1,711 @@
+library(wham)
+library(ggplot2)
+library(dplyr)
+library(patchwork)
+source(here::here("R","plot_functions.R"))
+theme_set(theme_bw())
+theme_update(strip.background = element_blank(), strip.placement = "outside", strip.text = element_text(size = rel(2)), 
+      axis.title = element_text(size = rel(2)), axis.text = element_text(size = rel(2)), legend.text = element_text(size = rel(2)), #text = element_text(size = rel(2)), 
+      legend.title = element_text(size = rel(2)))
+
+fit1 <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))[[2]] #M_1
+
+proj_R1 <- lapply(1:3, \(x) readRDS(here::here("results",paste0("m1_proj_",x,"_R_opt1.RDS"))))
+proj_R3 <- lapply(1:3, \(x) readRDS(here::here("results",paste0("m1_proj_",x,"_R_opt3.RDS"))))
+proj <- proj_R1[[3]]
+se_R1 <- lapply(proj_R1, \(x) TMB:::as.list.sdreport(x$sdrep, what = "Std", report = TRUE))
+se_R3 <- lapply(proj_R3, \(x) TMB:::as.list.sdreport(x$sdrep, what = "Std", report = TRUE))
+
+# fjp <- list() 
+# for(i in 1:3) fjp[[i]] <- readRDS(here::here("results",paste0("m1_proj_",i,"_sdrep_fjp.RDS")))
+
+#have to do se(predR) by hand
+source(here::here("R","plot_functions.R"))
+temp <- make_pred_R_cv(proj_R3[[1]], type = 1)
+temp <- rbind(temp, make_pred_R_cv(proj_R3[[1]], type = 1, region = 2))
+temp <- rbind(temp,make_pred_R_cv(proj_R3[[2]], type = 2))
+temp <- rbind(temp,make_pred_R_cv(proj_R3[[2]], type = 2, region = 2))
+temp <- rbind(temp,make_pred_R_cv(proj_R3[[3]], type = 3))
+temp <- rbind(temp,make_pred_R_cv(proj_R3[[3]], type = 3, region = 2))
+
+source(here::here("R","plot_functions.R"))
+brp_results <- lapply(c(proj_R1,proj_R3), get.brp.status.results)
+
+
+ny_full <- length(proj_R3[[1]]$years_full)
[…]

### 2025-01-02 R/projections.R

@@ -1,4 +1,5 @@
-library(wham)
+# library(wham)
+devtools::load_all("c:/work/wham/wham")
 fit1 <-readRDS(file.path("results","fits_no_M_re_rev.RDS"))[[2]]
 
 tdat <- cbind.data.frame(time = fit1$input$years_Ecov, bt_N = fit1$rep$Ecov_x[,1],bt_S = fit1$rep$Ecov_x[,2])
@@ -23,93 +24,56 @@ plot(newdat$time, newdat$bt_S)
 lines(newdat$time, newdat$pred_S)
 polygon(c(newdat$time,rev(newdat$time)), c(newdat$pred_lo_S, rev(newdat$pred_hi_S)), border = "transparent", col = gray(0.7,alpha=0.3))
 
+#Use terminal F in projections
 proj.opts.list <- list()
-# proj.opts.list[[1]] <- list(n.yrs = 11, proj_F_opt = c(5,rep(3,10)), proj_Fcatch = rep(10000,11)) #continue Ecov process by default
-# proj.opts.list[[2]] <- list(n.yrs = 11, proj_F_opt = c(5,rep(3,10)), proj_Fcatch = rep(10000,11),  cont.ecov=FALSE, use.last.ecov=FALSE, avg.ecov.yrs=tail(fit1$years,5)) #use average of last 5 years
-# proj.opts.list[[3]] <- list(n.yrs = 11, proj_F_opt = c(5,rep(3,10)), proj_Fcatch = rep(10000,11), cont.ecov=FALSE, use.last.ecov=FALSE, 
-# 	proj.ecov = cbind(newdat$pred_N,newdat$pred_S)[65:74,]) #use prediction from linear trend
-proj.opts.list[[1]] <- list(n.yrs = 10, proj_F_opt = rep(3,10), proj_Fcatch = rep(10000,10)) #continue Ecov process by default
-proj.opts.list[[2]] <- list(n.yrs = 10, proj_F_opt = rep(3,10), proj_Fcatch = rep(10000,10),  cont.ecov=FALSE, use.last.ecov=FALSE, avg.ecov.yrs=tail(fit1$years,5)) #use average of last 5 years
-proj.opts.list[[3]] <- list(n.yrs = 10, proj_F_opt = rep(3,10), proj_Fcatch = rep(10000,10), cont.ecov=FALSE, use.last.ecov=FALSE, 
[…]

### 2025-01-10 paper/paper.Rmd
Update paper.Rmd

@@
   - Kierstin Curti
-  - Alex Hansell
+  - Alexander C Hansell
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 The Woods Hole Assessment Model (WHAM) is a state-space age-structured assessment model that is used to assess and manage many stocks in the Northeast US.
-We first describe a multi-stock, multi-region extension of WHAM that treats the population and fleet dynamics seasonally and allows movement by season and region to be functions of time- and age-varying autocorrelated random effects and environmental covariates.
-We then illustrate the model by applying it to data for the northern and southern components of the NEUS black sea bass stock and evaluate alternative hypotheses of bottom temperature and random effects on recruitment and natural mortality.
+We first describe a multi-stock, multi-region extension of WHAM that treats the population and fleet dynamics seasonally and allows movement by season and region to be functions of time and age-varying autocorrelated random effects and environmental covariates.
+We then illustrate the model by applying it to data for the northern and southern components of the Northeast United States (NEUS) black sea bass stock and evaluate alternative hypotheses of bottom temperature and random effects on recruitment and natural mortality.
 We show strong evidence for temperature effects on recruitment, primarily for the northern stock component, and no evidence for including random effects or temperature effects on age 1 natural mortality.
@@
 
-A state-space statistical approach and maximum marginal likelihood or Bayesian fitting of stock assessment models allows estimation of time- and age-varying population attributes as random effects [@nielsenberg14,@cadigan16,@milleretal16].
+A state-space statistical approach and maximum marginal likelihood or Bayesian fitting of stock assessment models allows estimation of time and age-varying population attributes as random effects [@nielsenberg14,@cadigan16,@milleretal16].
 This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks [@puntetal20].
-The State-space Assessment Model [SAM, @nielsenberg14] continues to be developed and remains widely used within ICES to assess European fish stocks.
+The State-space Assessment Model [SAM, @nielsenberg14] continues to be developed and remains widely used within  the International Council for the Exploration of the Sea (ICES) to assess European fish stocks.
 Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada [@varkeyetal22;@perreaultetal20] and to the south, the Woods Hole Assessment Model [WHAM, @millerstock20; @stockmiller21] is now used to assess many fish stocks in the Northwest Atlantic Ocean [@nefsc22;@nefsc22a;@nefsc24].
@@
 
-We developed this extension of WHAM in concert with a research track assessment process for black sea bass where new modeling frameworks were recently examined and modeling two stock components simultaneously and temporal changes in distribution was of interest [@nefsc23].
-To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the NEUS using most of the model assumptions that were accepted during the peer-review process for the black sea bass assessment.
+We developed this extension of WHAM in concert with a research track assessment process for black sea bass (Centropristis striata) [@nefsc23].
+Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age assessment.
+The stock was split into two regions to account for spatial dynamics and improve model diagnostics (ASMFC, 2016).
+Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall [@mosershepherd09].
+Black sea bass distribution has been linked to warming waters on the northwest shelf (Bell et al.
+2016) and it is hypothesized that black sea bass are especially susceptible to climate change with likely increases in distribution and productivity due to warming water temperatures (Hare et al., 2015).
+Thus, a focus of the research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics [@nefsc23].To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the NEUS using most of the model assumptions that were accepted during the peer-review process.
 However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
@@
 When modeling multiple stock components occurring in different regions further dimensionality is added to the class of possible equilbrium reference points [e.g., @kapuretal21].
-Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and SSB without external simulation exercises.
+Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises.
 Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points.
@@
 
-### The probability transition matrix {-}
+### The probability transition matrix (PTM) {-}
 
 Because individuals may be alive in one of several regions or harvested in one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model.
-Approaches to modeling transitions among these tates may treat time discretely [e.g.,@arnason72;@schwarzetal93] or continuously [e.g., @hearnetal87;@commenges99; @andersenkeiding02].
+Approaches to modeling transitions among these states may treat time discretely [e.g.,@arnason72;@schwarzetal93] or continuously [e.g., @hearnetal87;@commenges99; @andersenkeiding02].
 Multi-state models can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval $\delta$, conditional on being in each of those state at beginning of the interval.
-For fish populations these states would be defined as being alive in a particular region or being dead due to fishing from a particular fleet or natural mortality.
+For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
 The time interval $i$ with duration $\delta_i$ would be a season and the PTMs would be uniquely defined for each stock by season $i$, year $y$, and age $a$ on January 1.
@@
 
-When there is more than 1 region, WHAM can model survival and movement as processes occurring sequentially or simultaneously.
-The sequential assumption is used widely in spatially explicit model (e.g., SS3).
+When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously.
+The sequential assumption is used widely in spatially explicit model (e.g., Stock Synthesis(Methot and Wetzel, 2013)).
 Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval.
@@
 
-WHAM assumes each fleet $f$ can harvest in only 1 region ($r_f$) during specified seasons.
+WHAM assumes each fleet $f$ can harvest in only one region ($r_f$) during specified seasons.
 So, for each fleet $f$, row $r_f$ and column $f$ of $\mathbf{H}_{y,a,i}$ will be $F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$ when fleet $f$ is harvesting during the interval $\delta_i$ and all other elements will be zero.
@@
 Seasonality can be configured to accommodate characteristics of spawning, movement, and fleet-specific behavior.
-The annual time step can be divided into $K$ (any number) seasons and the interval size $\delta_i$ for each season $i$ need not be equal to any other seasonal interval.
+The annual time step can be divided into $K$ (any number) seasons and the interval size $\delta_i$ for each season $i$ does not need to be equal to any other seasonal interval.
 Under the Markov assumption, the PTM of surviving and moving and dying over $K$ intervals $\delta_1,\ldots, \delta_K$ (i.e., the entire year) is just the product of the PTMs for each interval:
@@
 
-When there are $n_R$ regions and $n_F$ fleets, The vector of abundance in each state at age $a>1$ on January 1 is $\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}', \mathbf{0}')'$ where $\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})'$ is the number in the states corresponding to being alive in each region and $\mathbf{0}$ is a vector ($n_F+1$) for the numbers captured in each fleet and dead from natural mortality because no age $a$ fish have died yet on January 1.
+When there are $n_R$ regions and $n_F$ fleets, the vector of abundance in each state at age $a>1$ on January 1 is $\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}', \mathbf{0}')'$ where $\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})'$ is the number in the states corresponding to being alive in each region and $\mathbf{0}$ is a vector ($n_F+1$) for the numbers captured in each fleet and dead from natural mortality because no age $a$ fish have died yet on January 1.
 
@@
 So for age $a=1$, $\mathbf{N}_{O,y,1}$ is 0 except for row $r = r_s$.
-Options for configuring recruitment ($N_{y,1,r_s}$) for each stock are the same as previous versions of WHAM.
-If recruitment is assumed to be a function of spawning stock biomass (SSB), it is only the spawning population in region $r_s$ at the time of spawning constitutes the SSB in the stock-recruit function.
+Options for configuring recruitment ($N_{y,1,r_s}$) for each stock are the same as previous versions of WHAM [@millerstock20].
+If recruitment is assumed to be a function of SSB, it is only the spawning population in region $r_s$ at the time of spawning that constitutes the SSB in the stock-recruit function.
 However, models can configure spawning individuals to occur in other regions at the time of spawning.
@@
 Movement parameters are estimated on a transformed scale via a link function $g(\cdot)$.
-If survival and movement are occur simultaneously, the parameters are estimated a log link function is used and if they are separable, an additive logit link function (like a multinomial regression) is used.
+If survival and movement occur simultaneously, the parameters are estimated with a log link function  and if they are separable, an additive logit link function (like a multinomial regression) is used.
 On the transformed scale, the general model for the movement parameter from region $r$ to $r'$ in season $i$ and year $y$ for individuals of age $a$ is a linear function of both random and environmental effects:
@@
 
-Natural mortality options have been expanded in WHAM.
+Natural mortality options have been expanded since initial WHAM development.
 When not estimated, (mean) mortality rates may be stock-, region-, and age-specific.
@@
 When a Beverton-Holt or Ricker stock recruit relationship is assumed, an analogous Newton method is used to find $\log F$ that maximizes yield for MSY-based reference points.
-which are also a functions of the equilibrium yield per recruit (\ref{eq:ypr}) and equilibrium recruitment.
+which are also a functions of the equilibrium yield per recruit (Eq.
+\ref{eq:ypr}) and equilibrium recruitment.
 The function $g(\log F)$ in Eq.
@@
 
-The projection options are generally the same as those for previous versions of WHAM.
+The projection options are generally the same as those for previous versions of WHAM [@millerstock20].
 When there is movement of any stocks, the user has the option to project and use any random effects for time-varying movement or use the average over user specified years, analogous to how natural mortality can be treated in the projection period.
@@
 The north and south stock components are modeled as separate populations that spawn and recruit in respective regions.
-We have observations for each of four total fishing fleets, where two fishing fleets (Recreational and Commercial) operate in each region.
+We have observations for each of four total fishing fleets, where two fishing fleets (recreational and commercial) operate in each region.
 
@@
 
-## Uncertainty in recreational CPA index observations {-}
+## Uncertainty in recreational index observations {-}
 
@@
 
-## Index and Catch age composition observations {-}
-
-The Working Group investigated many alternative assumptions for the probability models and selectivity models for the 8 different sets of age composition observations to reduce residual patterns and retrospective patterns.
-These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for 6 sets of of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
+## Index and catch age composition observations {-}
+
+The Working Group investigated many alternative assumptions for the probability models and selectivity models for the eight different sets of age composition observations to reduce residual patterns and retrospective patterns.
+These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for six sets of of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
 
@@
 Because age 1 fish for the northern component can exist in both regions after January 1, natural mortality is acting in both regions for this stock.
-For models with covariate and/or annual random effects for age 1 fish we assume them only in the northern regions for the northern component.
+For models with covariate and(or) annual random effects for age 1 fish we assume them only in the northern regions for the northern component.
 The corresponding random effects are
@@
 
-## Model fitting, diagnostics, BRPs, and projections {-}
+## Model fitting, diagnostics, Biological Reference Points (BRPs), and projections {-}
 
@@
 We produced SPR-based BRP estimates ($F_{40\%}$ and SSB$_{40\%}$) from the best performing model using a variety of options provided in WHAM.
-We generated annual BRP estimates based on the annual weight at age, maturity at age, M, fleet selectivity, and either the annual recruitment random effect (Eq.
+We generated annual BRP estimates based on the annual weight at age, maturity at age, natural mortality, fleet selectivity, and either the annual recruitment random effect (Eq.
 \ref{eq:Rec-re}) or the annual expected recruitment (Eq.
@@
 We also calculated status as the ratios of annual average F and SSB to the corresponding reference points.
-Finally, we calculated "prevailing" BRPs using the average weight at age, maturity at age, M, fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status (Kobe plots) for the terminal year of the model (2021).
+Finally, we calculated "prevailing" BRPs using the average weight at age, maturity at age, M, fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
 All BRPs and status estimates are generated for each region as well as for the total stock.
@@
 
-As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
+For black sea bass, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figure 6).
+In the projection period, the AR1 approach reverts to the long-term mean of the environmental covariate.
+This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypothesis of increasing water temperature and changing environmental conditions (Hare et al.
+2015, Pershing et al.
+2021).
+Because of these concerns, the most recent stock assessment for yellowtail flounder used a change point analysis on environmental covariates to determine current conditions.
+The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections (NEFSC, 2024).
+Unfortunately, this methodology still has limitations because it is unlikely that current environmental conditions are stable.
+Similarly, a linear increase of temperature is also not likely in the future (see past trends, Figure 6 top row).As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
 Elaborate!
@@
 We used estimates of movement parameters from a companion assessment model to define prior distributions for corresponding movement parameters in the state-space black sea bass model, but other than the tagging data, both of the assessment models used much the same data which can lead to inappropriate inferences.
-A better approach would be to estimate the movement parameters externally from just from the tagging data.
+A better approach would be to estimate the movement parameters externally from just the tagging data.
 

### 2025-01-21 paper/paper.Rmd
saving work

--- 2025-01-02 paper/paper.Rmd
@@
 
-A state-space statistical approach and maximum marginal likelihood or Bayesian fitting of stock assessment models allows estimation of time- and age-varying population attributes as random effects [@nielsenberg14,@cadigan16,@milleretal16].
+A state-space statistical approach and maximum marginal likelihood or Bayesian fitting of stock assessment models allows estimation of time- and age-varying population attributes as random effects [@nielsenberg14;@cadigan16;@milleretal16].
 This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks [@puntetal20].

### 2025-02-13 paper/paper.Rmd
added in Alex's edits

--- 2025-01-10 paper/paper.Rmd
@@
   - Kierstin Curti
-  - Alexander C Hansell
+  - Alexander C. Hansell
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 
-The Woods Hole Assessment Model (WHAM) is a state-space age-structured assessment model that is used to assess and manage many stocks in the Northeast US.
+The Woods Hole Assessment Model is a state-space age-structured assessment model that is used to assess and manage many stocks in the Northeast US.
 We first describe a multi-stock, multi-region extension of WHAM that treats the population and fleet dynamics seasonally and allows movement by season and region to be functions of time and age-varying autocorrelated random effects and environmental covariates.
-We then illustrate the model by applying it to data for the northern and southern components of the Northeast United States (NEUS) black sea bass stock and evaluate alternative hypotheses of bottom temperature and random effects on recruitment and natural mortality.
+We then illustrate the model by applying it to data for the northern and southern components of the Northeast United States black sea bass stock and evaluate alternative hypotheses of bottom temperature and random effects on recruitment and natural mortality.
 We show strong evidence for temperature effects on recruitment, primarily for the northern stock component, and no evidence for including random effects or temperature effects on age 1 natural mortality.
@@
 
-A state-space statistical approach and maximum marginal likelihood or Bayesian fitting of stock assessment models allows estimation of time and age-varying population attributes as random effects [@nielsenberg14,@cadigan16,@milleretal16].
+A state-space statistical approach and maximum marginal likelihood or Bayesian fitting of stock assessment models allows estimation of time and age-varying population attributes as random effects [@nielsenberg14;@cadigan16;@milleretal16].
 This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks [@puntetal20].
@@
 
-We developed this extension of WHAM in concert with a research track assessment process for black sea bass (Centropristis striata) [@nefsc23].
+We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) [@nefsc23].
 Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age assessment.
-The stock was split into two regions to account for spatial dynamics and improve model diagnostics (ASMFC, 2016).
+The stock was split into two regions to account for spatial dynamics and improve model diagnostics [@asmfc16].
 Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall [@mosershepherd09].
-Black sea bass distribution has been linked to warming waters on the northwest shelf (Bell et al.
-2016) and it is hypothesized that black sea bass are especially susceptible to climate change with likely increases in distribution and productivity due to warming water temperatures (Hare et al., 2015).
+Black sea bass distribution has been linked to warming waters on the northwest shelf [@belletal15] and it is hypothesized that black sea bass are especially susceptible to climate change with likely increases in distribution and productivity due to warming water temperatures [@hareetal16].
 Thus, a focus of the research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics [@nefsc23].To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the NEUS using most of the model assumptions that were accepted during the peer-review process.
@@
 
-### The probability transition matrix (PTM) {-}
+### The probability transition matrix {-}
 
@@
 When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously.
-The sequential assumption is used widely in spatially explicit model (e.g., Stock Synthesis(Methot and Wetzel, 2013)).
+The sequential assumption is used widely in spatially explicit model [e.g., Stock Synthesis, @methotwetzel13].
 Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval.
@@
 
-## Bottom Temperature effects {-}
+## Bottom temperature effects {-}
 
@@
 
-## Model fitting, diagnostics, Biological Reference Points (BRPs), and projections {-}
+## Model fitting, diagnostics, reference points, and projections {-}
 
@@
 In the projection period, the AR1 approach reverts to the long-term mean of the environmental covariate.
-This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypothesis of increasing water temperature and changing environmental conditions (Hare et al.
-2015, Pershing et al.
-2021).
+This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypothesis of increasing water temperature and changing environmental conditions [@hareetal16;@pershingetal21].
 Because of these concerns, the most recent stock assessment for yellowtail flounder used a change point analysis on environmental covariates to determine current conditions.
@@
 Similarly, a linear increase of temperature is also not likely in the future (see past trends, Figure 6 top row).As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
-Elaborate!
 

### 2025-02-13 paper/paper.tex
added in Alex's edits

@@
   pdftitle={Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass},
-  pdfauthor={Timothy J. Miller1; Kierstin Curti; Alex Hansell},
+  pdfauthor={Timothy J. Miller1; Kierstin Curti; Alexander C. Hansell},
   hidelinks,
@@
 \author{Timothy J. Miller\textsuperscript{1} \and Kierstin
-Curti \and Alex Hansell}
-\date{02 January, 2025}
+Curti \and Alexander C. Hansell}
+\date{21 January, 2025}
 
@@
 
-The Woods Hole Assessment Model (WHAM) is a state-space age-structured
+The Woods Hole Assessment Model is a state-space age-structured
 assessment model that is used to assess and manage many stocks in the
@@
 WHAM that treats the population and fleet dynamics seasonally and allows
-movement by season and region to be functions of time- and age-varying
+movement by season and region to be functions of time and age-varying
 autocorrelated random effects and environmental covariates. We then
 illustrate the model by applying it to data for the northern and
-southern components of the NEUS black sea bass stock and evaluate
-alternative hypotheses of bottom temperature and random effects on
-recruitment and natural mortality. We show strong evidence for
-temperature effects on recruitment, primarily for the northern stock
+southern components of the Northeast United States black sea bass stock
+and evaluate alternative hypotheses of bottom temperature and random
+effects on recruitment and natural mortality. We show strong evidence
+for temperature effects on recruitment, primarily for the northern stock
 component, and no evidence for including random effects or temperature
@@
 A state-space statistical approach and maximum marginal likelihood or
-Bayesian fitting of stock assessment models allows estimation of time-
-and age-varying population attributes as random effects Miller et al.
-(2016b). This estimation approach is considered an essential feature of
-gold-standard assessment models that we use in tactical management of
-commercially important fish stocks (Punt et al. 2020). The State-space
-Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed
-and remains widely used within ICES to assess European fish stocks.
-Various state-space models are being used to manage cod and plaice
-stocks in the waters of Eastern Canada (Perreault et al. 2020; Varkey et
-al. 2022) and to the south, the Woods Hole Assessment Model (WHAM,
-Miller and Stock 2020; Stock and Miller 2021) is now used to assess many
-fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC
-2024).
+Bayesian fitting of stock assessment models allows estimation of time
+and age-varying population attributes as random effects (Nielsen and
+Berg 2014; Cadigan 2016; Miller et al. 2016b). This estimation approach
+is considered an essential feature of gold-standard assessment models
+that we use in tactical management of commercially important fish stocks
+(Punt et al. 2020). The State-space Assessment Model (SAM, Nielsen and
+Berg 2014) continues to be developed and remains widely used within the
+International Council for the Exploration of the Sea (ICES) to assess
+European fish stocks. Various state-space models are being used to
+manage cod and plaice stocks in the waters of Eastern Canada (Perreault
+et al. 2020; Varkey et al. 2022) and to the south, the Woods Hole
+Assessment Model (WHAM, Miller and Stock 2020; Stock and Miller 2021) is
+now used to assess many fish stocks in the Northwest Atlantic Ocean
+(NEFSC 2022a, 2022b; NEFSC 2024).
 
@@
 We developed this extension of WHAM in concert with a research track
-assessment process for black sea bass where new modeling frameworks were
-recently examined and modeling two stock components simultaneously and
-temporal changes in distribution was of interest (NEFSC 2023). To
+assessment process for black sea bass (\textit{Centropristis striata})
+(NEFSC 2023). Previous black sea bass assessments assumed two stock
+components (North and South), each assessed with a separate statistical
+catch at age assessment. The stock was split into two regions to account
+for spatial dynamics and improve model diagnostics (ASMFC 2016).
+Additionally, black sea bass have complex spatial dynamics and make
+seasonal migrations, moving inshore in the spring and offshore in the
+fall (Moser and Shepherd 2009). Black sea bass distribution has been
+linked to warming waters on the northwest shelf (Bell et al. 2015) and
+it is hypothesized that black sea bass are especially susceptible to
+climate change with likely increases in distribution and productivity
+due to warming water temperatures (Hare 2016). Thus, a focus of the
+research track assessment was to develop an assessment platform that
+could capture multiple stock components, complex seasonal movement and
+explore environmental drivers of population dynamics (NEFSC 2023).To
 illustrate the usage of WHAM 2.0 we apply it to two stock components of
 black sea bass off the coast of the NEUS using most of the model
-assumptions that were accepted during the peer-review process for the
-black sea bass assessment. However, here we evaluate evidence for
-alternative hypotheses of temporal variation and effects specifically of
-bottom temperature on recruitment and natural mortality of age 1
-individuals.
+assumptions that were accepted during the peer-review process. However,
+here we evaluate evidence for alternative hypotheses of temporal
+variation and effects specifically of bottom temperature on recruitment
+and natural mortality of age 1 individuals.
 
@@
 observation and process errors in state-space models allows uncertainty
-in projected attributes such as recruitment and SSB without external
-simulation exercises. Finally, the inclusion of explicit effects of
-environmental covariates in the model also requires some assumptions on
-how they are treated in projections and reference points. We demonstrate
-some of options of WHAM version 2.0 provide biological reference points
-and short-term projections.
+in projected attributes such as recruitment and spawning stock biomass
+(SSB) without external simulation exercises. Finally, the inclusion of
+explicit effects of environmental covariates in the model also requires
+some assumptions on how they are treated in projections and reference
+points. We demonstrate some of options of WHAM version 2.0 provide
+biological reference points and short-term projections.
 
@@
 each category over time as a multi-state model. Approaches to modeling
-transitions among these tates may treat time discretely (e.g., Arnason
+transitions among these states may treat time discretely (e.g., Arnason
 1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987;
@@
 time interval \(\delta\), conditional on being in each of those state at
-beginning of the interval. For fish populations these states would be
-defined as being alive in a particular region or being dead due to
-fishing from a particular fleet or natural mortality. The time interval
-\(i\) with duration \(\delta_i\) would be a season and the PTMs would be
+beginning of the interval. For fish populations, these states would be
+defined as being alive in a particular region or deceased due to fishing
+from a particular fleet or natural mortality. The time interval \(i\)
+with duration \(\delta_i\) would be a season and the PTMs would be
 uniquely defined for each stock by season \(i\), year \(y\), and age
@@
 
-When there is more than 1 region, WHAM can model survival and movement
+When there is more than one region, WHAM can model survival and movement
 as processes occurring sequentially or simultaneously. The sequential
-assumption is used widely in spatially explicit model (e.g., SS3). Under
-the sequential assumption, survival and death occur over the interval
-and movement among regions occurs instantly at either the beginning or
-the end of the interval. WHAM is configured to have movement occur after
-survival and mortality: \begin{equation*}
+assumption is used widely in spatially explicit model (e.g., Stock
+Synthesis, Methot and Wetzel 2013). Under the sequential assumption,
+survival and death occur over the interval and movement among regions
+occurs instantly at either the beginning or the end of the interval.
+WHAM is configured to have movement occur after survival and mortality:
+\begin{equation*}
   \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i}
@@
 
-WHAM assumes each fleet \(f\) can harvest in only 1 region (\(r_f\))
+WHAM assumes each fleet \(f\) can harvest in only one region (\(r_f\))
 during specified seasons. So, for each fleet \(f\), row \(r_f\) and
@@
 can be divided into \(K\) (any number) seasons and the interval size
-\(\delta_i\) for each season \(i\) need not be equal to any other
-seasonal interval. Under the Markov assumption, the PTM of surviving and
-moving and dying over \(K\) intervals \(\delta_1,\ldots, \delta_K\)
-(i.e., the entire year) is just the product of the PTMs for each
-interval:
+\(\delta_i\) for each season \(i\) does not need to be equal to any
+other seasonal interval. Under the Markov assumption, the PTM of
+surviving and moving and dying over \(K\) intervals
+\(\delta_1,\ldots, \delta_K\) (i.e., the entire year) is just the
+product of the PTMs for each interval:
 \[ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).\]
@@
 
-When there are \(n_R\) regions and \(n_F\) fleets, The vector of
+When there are \(n_R\) regions and \(n_F\) fleets, the vector of
 abundance in each state at age \(a>1\) on January 1 is
@@
 \(r = r_s\). Options for configuring recruitment (\(N_{y,1,r_s}\)) for
-each stock are the same as previous versions of WHAM. If recruitment is
-assumed to be a function of spawning stock biomass (SSB), it is only the
-spawning population in region \(r_s\) at the time of spawning
+each stock are the same as previous versions of WHAM (Miller and Stock
+2020). If recruitment is assumed to be a function of SSB, it is only the
+spawning population in region \(r_s\) at the time of spawning that
 constitutes the SSB in the stock-recruit function. However, models can
@@
 parameters are estimated on a transformed scale via a link function
-\(g(\cdot)\). If survival and movement are occur simultaneously, the
-parameters are estimated a log link function is used and if they are
+\(g(\cdot)\). If survival and movement occur simultaneously, the
+parameters are estimated with a log link function and if they are
 separable, an additive logit link function (like a multinomial
@@
 
-Natural mortality options have been expanded in WHAM. When not
-estimated, (mean) mortality rates may be stock-, region-, and
-age-specific. When random effects are used, the same 2DAR1 structure
-with age and year as described by Stock and Miller (2021) can be
-configured for a given stock and region. Any environmental covariate
+Natural mortality options have been expanded since initial WHAM
+development. When not estimated, (mean) mortality rates may be stock-,
+region-, and age-specific. When random effects are used, the same 2DAR1
+structure with age and year as described by Stock and Miller (2021) can
+be configured for a given stock and region. Any environmental covariate
 effects can be stock-, region-, and age-specific. So the general model
@@
 for MSY-based reference points. which are also a functions of the
-equilibrium yield per recruit (\ref{eq:ypr}) and equilibrium
+equilibrium yield per recruit (Eq. \ref{eq:ypr}) and equilibrium
 recruitment. The function \(g(\log F)\) in Eq. \ref{eq:newton} is the
@@
 The projection options are generally the same as those for previous
-versions of WHAM. When there is movement of any stocks, the user has the
-option to project and use any random effects for time-varying movement
-or use the average over user specified years, analogous to how natural
-mortality can be treated in the projection period. The projection of any
-environmental covariates has been revised to better include error in the
-estimated latent covariate in any effects on the population in
-projection years. Users can also specify a catch or fully-selected
-fishing mortality in each projection year and they can be
+versions of WHAM (Miller and Stock 2020). When there is movement of any
+stocks, the user has the option to project and use any random effects
+for time-varying movement or use the average over user specified years,
+analogous to how natural mortality can be treated in the projection
+period. The projection of any environmental covariates has been revised
+to better include error in the estimated latent covariate in any effects
+on the population in projection years. Users can also specify a catch or
+fully-selected fishing mortality in each projection year and they can be
 fleet-specific.
@@
 and recruit in respective regions. We have observations for each of four
-total fishing fleets, where two fishing fleets (Recreational and
-Commercial) operate in each region.
+total fishing fleets, where two fishing fleets (recreational and
+commercial) operate in each region.
 
@@
 
-\hypertarget{uncertainty-in-recreational-cpa-index-observations}{%
-\subsection*{Uncertainty in recreational CPA index
-observations}\label{uncertainty-in-recreational-cpa-index-observations}}
-\addcontentsline{toc}{subsection}{Uncertainty in recreational CPA index
+\hypertarget{uncertainty-in-recreational-index-observations}{%
+\subsection*{Uncertainty in recreational index
+observations}\label{uncertainty-in-recreational-index-observations}}
+\addcontentsline{toc}{subsection}{Uncertainty in recreational index
 observations}
@@
 \hypertarget{index-and-catch-age-composition-observations}{%
-\subsection*{Index and Catch age composition
+\subsection*{Index and catch age composition
 observations}\label{index-and-catch-age-composition-observations}}
-\addcontentsline{toc}{subsection}{Index and Catch age composition
+\addcontentsline{toc}{subsection}{Index and catch age composition
 observations}
@@
 The Working Group investigated many alternative assumptions for the
-probability models and selectivity models for the 8 different sets of
-age composition observations to reduce residual patterns and
+probability models and selectivity models for the eight different sets
+of age composition observations to reduce residual patterns and
 retrospective patterns. These analyses resulted in use of selectivity
 random effects for the northern fleet and indices and logistic-normal
-likelihoods for 6 sets of of age composition observations and
+likelihoods for six sets of of age composition observations and
 Dirichlet-multinomial likelihoods for one index and one fleet in the
@@
 \hypertarget{bottom-temperature-effects}{%
-\subsection*{Bottom Temperature
+\subsection*{Bottom temperature
 effects}\label{bottom-temperature-effects}}
-\addcontentsline{toc}{subsection}{Bottom Temperature effects}
+\addcontentsline{toc}{subsection}{Bottom temperature effects}
 
@@
 in both regions after January 1, natural mortality is acting in both
-regions for this stock. For models with covariate and/or annual random
+regions for this stock. For models with covariate and(or) annual random
 effects for age 1 fish we assume them only in the northern regions for
@@
 
-\hypertarget{model-fitting-diagnostics-brps-and-projections}{%
-\subsection*{Model fitting, diagnostics, BRPs, and
-projections}\label{model-fitting-diagnostics-brps-and-projections}}
-\addcontentsline{toc}{subsection}{Model fitting, diagnostics, BRPs, and
-projections}
+\hypertarget{model-fitting-diagnostics-reference-points-and-projections}{%
+\subsection*{Model fitting, diagnostics, reference points, and
+projections}\label{model-fitting-diagnostics-reference-points-and-projections}}
+\addcontentsline{toc}{subsection}{Model fitting, diagnostics, reference
+points, and projections}
 
@@
 WHAM. We generated annual BRP estimates based on the annual weight at
-age, maturity at age, M, fleet selectivity, and either the annual
-recruitment random effect (Eq. \ref{eq:Rec-re}) or the annual expected
-recruitment (Eq. \ref{eq:expected-recruitment}). We also calculated
-status as the ratios of annual average F and SSB to the corresponding
-reference points. Finally, we calculated ``prevailing'' BRPs using the
-average weight at age, maturity at age, M, fleet selectivity from the
-last 5 years (2017 to 2021) and average recruitment over years 2000 to
-2021, and made Kobe plots of joint status (Kobe plots) for the terminal
-year of the model (2021). All BRPs and status estimates are generated
-for each region as well as for the total stock.
+age, maturity at age, natural mortality, fleet selectivity, and either
+the annual recruitment random effect (Eq. \ref{eq:Rec-re}) or the annual
+expected recruitment (Eq. \ref{eq:expected-recruitment}). We also
+calculated status as the ratios of annual average F and SSB to the
+corresponding reference points. Finally, we calculated ``prevailing''
+BRPs using the average weight at age, maturity at age, M, fleet
+selectivity from the last 5 years (2017 to 2021) and average recruitment
+over years 2000 to 2021, and made Kobe plots of joint status for the
+terminal year of the model (2021). All BRPs and status estimates are
+generated for each region as well as for the total stock.
 
@@
 
-As more assessments begin including effects of environmental covariates
-explicitly in assessment models, assumptions about these covariates in
-calculated BRPs and population projections becomes an important decision
-in the management process. Elaborate!
+For black sea bass, it is evident that the projection methodology of the
+environmental covariate has an impact on short-term projection estimates
+of recruitment and SSB (Figure 6). In the projection period, the AR1
+approach reverts to the long-term mean of the environmental covariate.
+This is most likely not an appropriate assumption for stocks in the NEUS
+given observations and hypothesis of increasing water temperature and
+changing environmental conditions (Hare 2016; Pershing et al. 2021).
+Because of these concerns, the most recent stock assessment for
+yellowtail flounder used a change point analysis on environmental
+covariates to determine current conditions. The mean environmental
+covariate from the current condition was then used as a data input in
+WHAM to inform short-term projections (NEFSC, 2024). Unfortunately, this
+methodology still has limitations because it is unlikely that current
+environmental conditions are stable. Similarly, a linear increase of
+temperature is also not likely in the future (see past trends, Figure 6
+top row).As more assessments begin including effects of environmental
+covariates explicitly in assessment models, assumptions about these
+covariates in calculated BRPs and population projections becomes an
+important decision in the management process.
 
@@
 inappropriate inferences. A better approach would be to estimate the
-movement parameters externally from just from the tagging data.
+movement parameters externally from just the tagging data.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-asmfc16}{}}%
+ASMFC. 2016. Black sea bass stock assessment. {Available} at
+http://www.asmfc.org/uploads/file/5953f11d2016BlackSeaBassStockAssmt.pdf.
+
+\leavevmode\vadjust pre{\hypertarget{ref-belletal15}{}}%
+Bell, R.J., Richardson, D.E., Hare, J.A., Lynch, P.D., and Fratantoni,
+P.S. 2015. Disentangling the effects of climate, abundance, and size on
+the distribution of marine fish: An example based on four stocks from
+the {N}ortheast {US} shelf. ICES Journal of Marine Science
+\textbf{72}(5): 1311--1322.
+doi:\href{https://doi.org/10.1093/icesjms/fsu217}{10.1093/icesjms/fsu217}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-bosleyetal22}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-hareetal16}{}}%
+Hare, W.E.A.N., Jonathan A. AND Morrison. 2016. A vulnerability
+assessment of fish and invertebrates to climate change on the northeast
+u.s. Continental shelf. PLOS ONE \textbf{11}: 1--30.
+doi:\href{https://doi.org/10.1371/journal.pone.0146756}{10.1371/journal.pone.0146756}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-hearnetal87}{}}%
@@
 doi:\href{https://doi.org/10.2960/j.v51.m727}{10.2960/j.v51.m727}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pershingetal21}{}}%
+Pershing, A.J., Alexander, M.A., Brady, D.C., Brickman, D., Curchitser,
+E.N., Diamond, A.W., McClenachan, L., Mills, K.E., Nichols, O.C.,
+Pendleton, D.E., Record, N.R., Scott, J.D., Staudinger, M.D., and Wang,
+Y. 2021. Climate impacts on the {G}ulf of {M}aine ecosystem: A review of
+observed and expected changes in 2050 from rising temperatures.
+Elementa: Science of the Anthropocene \textbf{9}(1): 00076.
+doi:\href{https://doi.org/10.1525/elementa.2020.00076}{10.1525/elementa.2020.00076}.
 

### 2025-02-28 paper/paper.Rmd
good draft

@@
 
-Describes the multi-region, multi-stock generalization of WHAM and its usage to evaluate evidence of alternative hypotheses about temperature effects on recruitment of black sea bass stock components.
+Describes the multi-region, multi-stock generalization of WHAM and its usage to evaluate evidence of alternative hypotheses about temperature effects on recruitment and age 1 natural mortality of black sea bass stock components.
 
@@
 
-The Woods Hole Assessment Model is a state-space age-structured assessment model that is used to assess and manage many stocks in the Northeast US.
-We first describe a multi-stock, multi-region extension of WHAM that treats the population and fleet dynamics seasonally and allows movement by season and region to be functions of time and age-varying autocorrelated random effects and environmental covariates.
-We then illustrate the model by applying it to data for the northern and southern components of the Northeast United States black sea bass stock and evaluate alternative hypotheses of bottom temperature and random effects on recruitment and natural mortality.
+The Woods Hole Assessment Model is a general state-space age-structured assessment model that is used to assess and manage many stocks in the Northeast US.
+We first describe an extension of the model allowing any number of stocks (or stock components) and and regions with movement among regions as well as seasonal variation in stock and fleet dynamics.
+Movement rates can be a be functions of time- and age-varying random effects and environmental covariates.
+We then illustrate the model by applying it to data for the northern and southern components of the Northeast US black sea bass stock and evaluate alternative hypotheses of bottom temperature and time-varying random effects on recruitment and natural mortality.
 We show strong evidence for temperature effects on recruitment, primarily for the northern stock component, and no evidence for including random effects or temperature effects on age 1 natural mortality.
@@
 
-A state-space statistical approach and maximum marginal likelihood or Bayesian fitting of stock assessment models allows estimation of time and age-varying population attributes as random effects [@nielsenberg14;@cadigan16;@milleretal16].
+A state-space age-structured stock assessment models can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures [@nielsenberg14;@cadigan16;@milleretal16].
 This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks [@puntetal20].
@@
 Like SAM, WHAM models are built using the Template Model Builder package [TMB, @kristensenetal16] which provides a computationally efficient means of fitting an extremely wide class of models with random effects.
-WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as "survival"), fishery and index selectivity, natural mortality, and catchability.
+WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as "apparent survival"), fishery and index selectivity, natural mortality, and catchability.
 
@@
 Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall [@mosershepherd09].
-Black sea bass distribution has been linked to warming waters on the northwest shelf [@belletal15] and it is hypothesized that black sea bass are especially susceptible to climate change with likely increases in distribution and productivity due to warming water temperatures [@hareetal16].
+Black sea bass distribution has been linked to warming waters on the northwest shelf [@belletal15] and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north [@hareetal16].
 Thus, a focus of the research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics [@nefsc23].To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the NEUS using most of the model assumptions that were accepted during the peer-review process.
@@
 Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points.
-We demonstrate some of options of WHAM version 2.0 provide biological reference points and short-term projections.
+We demonstrate some of options of WHAM version 2.0 to provide biological reference points and short-term projections.
 
@@
 Approaches to modeling transitions among these states may treat time discretely [e.g.,@arnason72;@schwarzetal93] or continuously [e.g., @hearnetal87;@commenges99; @andersenkeiding02].
-Multi-state models can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval $\delta$, conditional on being in each of those state at beginning of the interval.
+We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval $\delta$ (columns), conditional on being in each of those state at beginning of the interval (rows).
 For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
@@
 Each row and column of the PTM correspond to one the states: alive in region $r$, dead in fleet $f$, or dead from natural causes.
-The probabilities in each row sum to unity and assume an individual is in the corresponding state at the beginning of the interval.
+The probabilities in each row sum to unity (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval.
 Given $n_R$ regions and $n_F$ fleets, the square PTM ($n_R + n_F + 1$ rows and columns) as a function of sub-matrices is
@@
 is the $n_R \times n_F$ matrix defining probabilities of being captured in each fleet during the interval, and $\mathbf{D}_{y,a,i}$ is the $n_R x 1$ matrix of probabilities of dying due to natural mortality during the interval.
-We have the identity matrix $\mathbf{I}_{H}$ for the states for  capture by each fleet and a 1 for the state for natural mortality because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).
-
-WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would be multinomial distributed as in a model for tagging data where fates of individual fish are assumed independent.
+We have the identity matrix $\mathbf{I}_{H}$ for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).
+
+WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would be multinomial distributed as in a model for tagging data where fates of individual tagged fish are assumed independent.
 The PTMs determine the expected numbers 1) in each state on January 1 of year $t+1$ at age $a+1$ given the abundances at age $a$ on January 1 of year $t$, 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.
@@
 Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval.
-WHAM is configured to have movement occur after survival and mortality:
+WHAM is configured to have movement occur after survival:
 \begin{equation*}
@@
 Seasonality can be configured to accommodate characteristics of spawning, movement, and fleet-specific behavior.
-The annual time step can be divided into $K$ (any number) seasons and the interval size $\delta_i$ for each season $i$ does not need to be equal to any other seasonal interval.
+The annual time step can be divided into $K$ seasons and the interval size $\delta_i$ for each season $i$ does not need to be equal to any other seasonal interval.
 Under the Markov assumption, the PTM of surviving and moving and dying over $K$ intervals $\delta_1,\ldots, \delta_K$ (i.e., the entire year) is just the product of the PTMs for each interval:
@@
 where  
-$$\text{SSB}_{s,y} =  \sum^A_{a=1}  w_{s,y,a} \text{mat}_{s,y,a} \mathbf{O}_{s,y,a,r_s}(t_s)' \mathbf{N}_{O,s,y,a}$$
-where $w_{s,y,a}$ is the mean weight at age of spawning individuals, $\text{mat}_{s,y,a}$ is the maturity at age, and $\mathbf{O}_{s,y,a,r_s}(t_s)$, the $r_s$ column of the upper-left submatrix of Eq.
-\ref{eq:ptm_spawn}, are the probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year.
+$$\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}(t_s)' \mathbf{N}_{O,y,a}$$
+and $w_{y,a}$ is the mean weight at age of spawning individuals, $\text{mat}_{y,a}$ is the maturity at age, and $\mathbf{O}_{y,a,r_s}(t_s)$ are the probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year (the $r_s$ column of the upper-left submatrix of Eq.
+\ref{eq:ptm_spawn}).
 
@@
 When the transitions in abundance at age are treated deterministically, $\boldsymbol{\varepsilon}_{y,a} = 0$.
-Since recruitment for a given stock currently only occurs in one region $r_s$ there is only a single time-varying recruitment random effect for each stock.
+Since recruitment for a given stock currently only occurs in one region $r_s$ there is only a single vector of time-varying recruitment random effects for each stock.
 
@@
 
-The initial abundances at age can also be treated as independent or as AR1 random effects. Defining the vector of initial abundance at age in region $r$ as $\mathbf{N}_{O,1,r}$, the general model is 
+The initial abundances at age can also be treated as independent or AR1 random effects. Defining the vector of initial abundance at age in region $r$ as $\mathbf{N}_{O,1,r}$, the general model is 
 $$\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}$$
@@
 
-For each season, there are at most $n_R-1$ parameters determining movement among regions given starting an the season in region $r$ in either the sequential or simultaneous configurations.
+For each season, there are at most $n_R-1$ parameters determining movement among regions given starting the season in region $r$ in either the sequential or simultaneous configurations.
 Movement parameters are estimated on a transformed scale via a link function $g(\cdot)$.
 If survival and movement occur simultaneously, the parameters are estimated with a log link function  and if they are separable, an additive logit link function (like a multinomial regression) is used.
-On the transformed scale, the general model for the movement parameter from region $r$ to $r'$ in season $i$ and year $y$ for individuals of age $a$ is a linear function of both random and environmental effects:
+On the transformed scale, the general model for the movement parameter from region $r$ to $r'$ in season $i$ and year $y$ for individuals of age $a$ is a linear function of an intercept or mean parameter ($\theta_{r\rightarrow r',i}$) and both random and environmental effects:
 \begin{equation*}
@@
 \end{equation*}
-The random effects $\varepsilon_{r\rightarrow r',y,a,i}$ are season-, and(or) region-to-region-specific and modeled most generally as 2DAR1 random effects with age and(or) year where the covariance is
+The random effects $\varepsilon_{r\rightarrow r',y,a,i}$ are season-, and region-to-region-specific and modeled most generally as 2DAR1 random effects with age and(or) year where the covariance is
 \begin{equation*}
@@
 similar to how WHAM models variation in survival, natural mortality, and selectivity.
-Effects of covariate $E_k$ can be age-, season-, and(or) region-to-region-specific $\beta_{r\rightarrow r',a,i,k}$ and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.
+Effects of covariate $E_k$ can be age-, season-, and region-to-region-specific $\beta_{r\rightarrow r',a,i,k}$ and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.
 
 There is currently no likelihood component for tagging data.
-Therefore, movement parameters would generally either need to be fixed or assumed to have some prior distribution, possibly based on external parameter estimates.
-We include prior distributions for the season and region-to-region specific (mean) movement parameters which are treated as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
+Therefore, movement parameters would usually either need to be fixed or assumed to have some prior distribution, possibly based on external parameter estimates.
+WHAM includes prior distributions for the season and region-to-region specific (mean) movement parameters and treats them as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
   \begin{equation*}
@@
 
-Natural mortality options have been expanded since initial WHAM development.
-When not estimated, (mean) mortality rates may be stock-, region-, and age-specific.
-When random effects are used, the same 2DAR1 structure with age and year as described by @stockmiller21 can be configured for a given stock and region.
-Any environmental covariate effects can be stock-, region-, and age-specific.
+When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific.
+When random effects are used, the same 2DAR1 structure with age and year as described by @stockmiller21 can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
 So the general model for natural mortality is
@@
 
-Catchability of the index can also be treated as functions of normal random effects and(or) environmental covariate effects 
+Catchability of the index can also be treated as logit functions of normal random effects and(or) environmental covariate effects 
 $$log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}$$ 
-where $u_{m}$ and $l_{m}$ are the upper and lower bounds of catchability for index $m$ (defaults are 0 and 1000) and the general covariance structure for the annual random effects is AR1
+where $\theta_{q,m}$ is an intercept or mean parameter, and $u_{m}$ and $l_{m}$ are the upper and lower bounds of catchability for index $m$ (defaults are 0 and 1000).
+The general covariance structure for the annual random effects is AR1
 $$Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.$$
 
-### Weight and Maturity at age {-}
-
-Weight and maturity at age are treated similarly to @stockmiller21.
-Annual weight at age matrices for each fleet are used to calculate total catch and the weight at age is applied to catch numbers at age for any stocks caught by the fleet.
-Similarly, when indices and(or) associated age composition observations are measured in biomass, the weight at age matrices for the index are applied to predicted numbers at age of all stocks observed by the survey.
-Unique weight at age and maturity at age matrices are allowed for each stock to calculate spawning stock biomass.
-
 ### Reference points {-}
 
-Currently a single F reference point $\widetilde F$ is estimated across stocks and regions and F by fleet and age is $\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}$.
-Selectivity is determined as before when there are multiple fleets where $\text{sel}_{f,a}$ is determined by averaging F at age over a user-defined set of years
-\begin{equation*}
+Currently a single fishing mortality reference point $\widetilde F$ is estimated across stocks and regions and the fleet- and age-specific reference points is $\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}$.
+Selectivity is determined across fleets and ages as in the equilibrium assumption for initial abundance at age except here it is a function of average F at age over a user-defined set of years
+\begin{equation}\label{eq:spr_sel}
   \text{sel}_{a,f} = \frac{\overline F_{a,f}}{\max_a \sum^{n_F}_{f=1}{\overline F}_{a,f}}
-\end{equation*}
-
-The equilibrium spawning stock biomass per recruit for stock $s$ in region $r_s$ is defined as
+\end{equation}
+
+The equilibrium spawning stock biomass per recruit for a given stock in region $r_s$ is defined as
 \begin{equation}\label{eq:ssbpr}
- \upphi_s(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{s,a,r_s,\cdot} \mathbf{O}_{s,a,\cdot,r_s}(t_s) w_{s,a} m_{s,a}
-\end{equation}
-where $w_{s,a}$ and $m_{s,a}$ are the mean individual weight and probability of maturity at age $a$, $\widetilde{\mathbf{O}}_{s,a}$ are as described in Eq.
-\ref{eq:npraa}, and $\mathbf{O}_{s,a}(t_s)$ is the $n_R \times n_R$ upper-left sub-matrix of eq.
+ \upphi(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{O}_{a,\cdot,r_s}(t_s) w_{a} m_{a}
+\end{equation}
+where $w_{a}$ and $m_{a}$ are the mean individual weight and probability of maturity at age $a$, $\widetilde{\mathbf{O}}_{a}$ are as described in Eq.
+\ref{eq:npraa}, and $\mathbf{O}_{a}(t_s)$ is the $n_R \times n_R$ upper-left sub-matrix of eq.
 \ref{eq:ptm_spawn} with the probabilities of surviving and occurring in each region $r'$ at age $a+t_s$ given starting in region $r$ at age $a$.
-The further subscripts $r_s,\cdot$ and $\cdot,r_s$ indictate row or column $r_s$, respectively.
+The further subscripts $r_s,\cdot$ and $\cdot,r_s$ indicate row or column $r_s$, respectively.
 Using these rows and columns is required because of the assumption that spawning and recruitment only occur in region $r_s$.
@@
 \ref{eq:ssbpr}) is conditional on the region of recruitment $r_s$.
-The equilibrium recruitment in each region $\widetilde {\mathbf{N}}_{s,1}$, depends on the stock dynamics.
-This version of WHAM currently only allows complete spawning region fidelity so that a stock only spawns and recruits in a single region.
-In this case, $\widetilde {\mathbf{N}}_{s,1}$ will be positive in the spawning region ($r_s$) and zero elsewhere.
-Similarly, the row $r_s$ of $\mathbf{O}_s$ will be zero off of the diagonal.
-The matrices of probabilities of surviving and occurring in each region, $\widetilde{\mathbf{O}}_{s,a}$ and $\mathbf{O}_{s,a}(\delta_s)$, are functions of the fishing mortality rates for fleets in each region $\widetilde F_{f,a}$.
-
-The matrix equilibrium yield per recruit as a function of $\widetilde F$ is calculated as
+The equilibrium recruitment depends on the stock dynamics.
+WHAM currently only allows complete spawning region fidelity so that a stock only spawns and recruits in a single region.
+Therefore, the equilibrium vector of numbers at age 1 for each state, $\widetilde {\mathbf{N}}_{1}$, will only be nonzero in the row corresponding to the spawning region ($r_s$).
+The matrices of probabilities of surviving and occurring in each region, $\widetilde{\mathbf{O}}_{s,a}$ and $\mathbf{O}_{s,a}(t_s)$, are functions of the fishing mortality rates for fleets in each region $\widetilde F_{f,a}$.
+
+The matrix of equilibrium yield per recruit by region and fleet as a function of $\widetilde F$ is calculated as
 \begin{equation}\label{eq:ypr}
- \widetilde{Y}_s({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{s,a,r_s,\cdot} \mathbf{H}_{s,a} \mathbf{c}_{s,a}
+ \widetilde{\nu}({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{H}_{a} \mathbf{c}_{s,a}
 \end{equation}
@@
 
-As in previous versions of WHAM package, ``static'' reference points, typically meant to be defined for prevailing conditions, average all of the inputs to the spawning biomass and yield per recruit calculations over the user-specified years (e.g., last 5 years of the model).
-This same averaging is also applied to possibly time-varying movement parameters.
+As in previous versions of WHAM package, reference points for prevailing conditions, average all of the inputs to the spawning biomass and yield per recruit calculations (weight, maturity, natural mortality, and movement) over a set of user-specified years as shown in eq.
+\ref{eq:spr_sel} for selectivity (e.g., last 5 years of the model).
 
@@
 This assessment included the spatial features and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
-
 Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.
@@
 
-There are 11 seasonal intervals within each calendar year: five monthly time intervals from Jan 1 to May 31, a spawning season from June 1 to July 31, and five monthly intervals from August 1 to December 31.
+There are 11 "seasonal" intervals within each calendar year: five monthly time intervals from Jan 1 to May 31, a spawning season from June 1 to July 31, and five monthly intervals from August 1 to December 31.
 The southern stock component is assumed to never move to the northern region.
@@
 As noted in the description of the general WHAM model, tagging data are not yet allowed.
-However, the Working Group also fit a Stock Synthesis model [@methotwetzel13] which provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).
-
-With the movement configuration, the northern origin fish (ages 2+) can occur in the southern region on January 1.
+However, the Working Group also fit a Stock Synthesis model [@methotwetzel13] which included tagging data and provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).
+
+With movement as configured, the northern origin fish (ages 2+) can occur in the southern region on January 1.
 Estimating initial numbers at age as separate parameters can be challenging even in single-stock models, but for black sea bass the available data cannot distinguish the proportion of northern and southern component fish at each age in the southern region in the initial year of the model.
-Therefore, we used the simplifying equilibrium assumption described above where there are two parameters estimated for each stock component: an initial recruitment and an equilibrium fully-selected $F$ that determines the abundance at age in each region for each stock component.
-
-For the northern population abundance at age 1 on January 1 (recruitment) is only allowed in the northern region, but given the monthly movement described above, older individuals that previously recruited in the northern region may occur in the southern region on January 1.
-Therefore, a model with survival random effects will model the transitions (survival/movement) of abundances at age of northern origin fish in each region.
-All of the initial runs assumed variance parameters for these random effects to be the same for northern origin fish occurring in both regions on January 1.
-The base model assumes very small variance for the transitions of northern fish in the southern region, which is approximately the same as the deterministic transition assumptions of a statistical catch at age model.
-We also allow 2DAR1 correlation for recruitment and survival for both the northern and southern components.
-Unique variance and correlation parameters for the recruitment and survival random effects are estimated for the northern and southern components.
+Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component: an initial recruitment and an equilibrium fully-selected $F$ that determines the abundance at age in each region for each stock component.
+
+For subsequent years, all models allow AR1 correlation of recruitment and 2DAR1 correlation for survival for both the northern and southern stock components.
+Models with apparent survival random effects will model the transitions (conditional on both survival and movement) of abundances at age of northern origin fish in each region, but there appears to be little ability of the existing data from the southern region to distinguish apparent survival random effects for each stock component.
+All models we considered assume very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model.
+Unique variance and correlation parameters for the recruitment and apparent survival random effects are estimated for the northern and southern components.
 
@@
 
-The estimated coefficients of variation (CVs) provided by the analyses to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the Working Group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
-In many of the initial runs as well as the base model we allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
-Models that successfully estimated these scalars indicated standard deviations for these surveys to be  approximately 5 times the input value and this value was fixed in many preliminary runs to avoid dealing with convergence problems.
-However, the base model successfully estimated these scalars.
-The model estimates are negligibly affected by estimating these scalars, but we felt estimating these parameters allowed uncertainty in model output to be more properly conveyed.
+The estimated coefficients of variation (CVs) provided by the analyses by the Working Group to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the Working Group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
+All models we considered allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
+However, lack of model convergence was observed in many fits for self-test simulations.
+The model estimates are negligibly affected by estimating these scalars, but estimating rather than fixing these parameters allowed uncertainty in model output to be more properly conveyed.
 
@@
 
-The Working Group investigated many alternative assumptions for the probability models and selectivity models for the eight different sets of age composition observations to reduce residual patterns and retrospective patterns.
+The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual patterns and retrospective patterns.
 These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for six sets of of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
@@
 
-All model fits also include bottom temperature observations for the northern and southern regions from 1963 to 2021 and estimated standard errors ranging between 0.03 and 0.09 degrees Celsius [@nefsc23].
+All models include bottom temperature observations for the northern and southern regions from 1963 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius [@nefsc23].
 We retained the assumption from the peer-reviewed assessment that treated the latent bottom temperature covariates in each region as AR1 processes.
@@
 These analyses derive from the hypothesis that bottom temperature affects overwinter survival of fish where the fish turn from age 0 to age 1 on January 1 [@milleretal16_yoy_survival].
-This temperature may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of of the year after becoming age 1.
+This covariate may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of of the year after becoming age 1.
 Furthermore, we have no direct observations of age 1 individuals from surveys until the spring season each year.
@@
 
-It is standard practice to treat annual recruitment as time-varying deviations from a mean model and all models here treat recruitment deviations as AR1 random effects.
+All models treat annual recruitment as AR1 time-varying random effects with mean parameters for each stock component.
 However, @milleretal18 showed how inferences of temperature effects on growth or maturity parameters can be very different whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
@@
 
-## Model fitting, diagnostics, reference points, and projections {-}
+## Model fitting and diagnostics {-}
 
@@
 Absolute values of Mohn's $\rho$ near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended.
-
 As in @milleretal16, we also assessed the consistency of the AIC-based model selection over retrospective peels to guard against previously noted changes in perception of covariate effects on recruitment with increased length of the time series of observations [@myers98].
@@
 We performed a jitter analysis of the base model $M_0$ and the best fitting model to investigate whether a local mimimum of the negative log-likelihood surface was obtained by the optimization.
-We used the jitter_wham function in the WHAM package which by default simulates starting values from a multivariate normal distribution with mean and covariance defined by the MLEs and estimated covariance matrix from the fitted model.
+We used the jitter_wham function in the WHAM package which by default simulates starting values from a multivariate normal distribution with mean and covariance defined by the MLEs and hessian-based covariance matrix from the fitted model.
 See the Supplementary Materials for more details.
 
-For the best performing model, we also performed so-called simulation self-tests where new observations were simulated conditional on all estimated fixed and random effects and the same model configuration was fit to each of the simulated data sets.
+For the best performing model, we also performed simulation self-tests where new observations were simulated conditional on all estimated fixed and random effects and the same model configuration was fit to each of the simulated data sets.
 We estimated median relative bias of SSB across these simulations.
 See the Supplementary Materials for more details.
+
+## Reference points, and projections {-}
 
@@
 We specified fishing mortality in projection years to be constant at the value in the last model year (2021).
-Average weight, natural mortality, movement, and maturity at age are used in projection years whereas the AR1 processes for all numbers at age random effects are continued.
+Average weight, natural mortality, movement, and maturity at age over the last 5 years are used in projection years whereas the AR1 processes for all numbers at age random effects are continued.
 
@@
 
-The 2dAR1 random effects for the selectivity parameters of the northern recreational fleet indicate a decrease in selectivity at ages 2 to 6 over the time series which corresponds to increases in the size limits in regulations by states where these catches occurred (Figure \ref{fig:selectivity-re}).
+The 2DAR1 random effects for the selectivity parameters of the northern recreational fleet indicate a decrease in selectivity at ages 2 to 6 over the time series which corresponds to increases in the size limits in regulations by states where these catches occurred (Figure \ref{fig:selectivity-re}).
 The model also estimated a more modest decrease in selectivity at ages 2 to 3 for the northern commercial fleet.
@@
 
-Using the expected recruitment rather than the random effects as the multiplier with equilibrium SSB/R results in less variable estimates of annual SSB$_{40\%}$ (Figure \ref{fig:annual-BRPs}: top row).
+Using the expected recruitment rather than the random effects as the multiplier with equilibrium spawning biomass per recruit ($\upphi(\widetilde{F})$, eq.
+\ref{eq:ssbpr}) results in less variable estimates of annual SSB$_{40\%}$ (Figure \ref{fig:annual-BRPs}: top row).
 Estimates were less variable for the southern stock component than the northern stock component mainly because the expected recruitment is constant for the former whereas that for the northern component varies via the effect of the bottom temperature anomalies.
-Temporal variation for the southern stock component is mainly caused by annual variation in the inputs to the SSB/R calculation and $F_{40\%}$, but there is also some effect of the variation in the proportion of total recruitment for each stock component as used in the weighting terms to define the 40\% reduction in unfished SSB/R ($\lambda_s$ in Eq.
+Temporal variation for the southern stock component is mainly caused by annual variation in the inputs to the $\upphi(\widetilde{F})$ calculation and $F_{40\%}$, but there is also some effect of the variation in the proportion of total recruitment for each stock component as used in the weighting terms to define the 40\% reduction in unfished spawning biomass per recruit ($\lambda_s$ in Eq.
 \ref{eq:newton-obj}).
@@
 Ratios of SSB to SSB$_{40\%}$ should be viewed with caution because the SSB reference point is being defined using annual recruitments rather than some form of average recruitment and therefore, ratios larger than 2.5 which would correspond to unfished population sizes are possible (Figure \ref{fig:annual-BRPs}: second row).
-However, for the southern component, using the constant expected recruitment is comparable to using median recruitment and the annual variation in the inputs to SSB/R indicates no years where SSB was greater than unfished levels throughout the time series.
+However, for the southern component, using the constant expected recruitment is comparable to using median recruitment and the annual variation in the inputs to $\upphi(\widetilde{F})$ indicates no years where SSB was greater than unfished levels throughout the time series.
 The patterns in uncertainty of SSB ratios are the same as those for the SSB reference points (Figure \ref{fig:annual-SSB-status-cvs}).
 
-Differences between annual $F_{40\%}$ estimates using the alternative types of recruitment estimates are much less than those for SSB$_{40\%}$ because the recruitment estimates are only used in the weighting of the stock-specific SSB/R to determine $F_{40\%}$ ($\lambda_s$ in Eq.
+Differences between annual $F_{40\%}$ estimates using the alternative types of recruitment estimates are much less than those for SSB$_{40\%}$ because the recruitment estimates are only used in the weighting of the stock-specific $\upphi(\widetilde{F})$ to determine $F_{40\%}$ ($\lambda_s$ in Eq.
 \ref{eq:newton-obj}, Figure \ref{fig:annual-BRPs}: third row).
@@
 Unlike the annual ratios of $F$ to $F_{40\%}$, the $F_{40\%}$ under prevailing conditions uses a selectivity defined by the average $F$ at age and fleet over the last 5 years, which is different from that of the selectivity in the terminal year and therefore those ratios for each region and in total are different (Figure \ref{fig:annual-BRPs}: bottom row).
-Although the alternative types of recruitment estimates result in different points on the kobe plots for joint status, the uncertainty in the estimates suggest similar probabilities of alternative quadrants.
+Although the alternative types of recruitment estimates result in different points on the Kobe plots for joint status, the uncertainty in the estimates suggest similar probabilities of alternative quadrants.
 However, the usage of the confidence regions of the ratios to defined probabilities is an approximation because these ratios are functions of both empirical Bayes estimates of random effects as well as maximum likelihood estimates of fixed effects.
@@
 
-When continuing the AR1 model for the bottom temperature anomalies in the projection years we can see that the uncertainty grows and asymptotes to a level determined by the marginal variance of the process and that the predicted value approaches the estimated marginal mean value of approximately 0 (Figure \ref{fig:R-BT-proj}).
+When continuing the AR1 model for the bottom temperature anomalies in the projection years, the uncertainty grows and asymptotes to a level determined by the marginal variance of the process and the predicted value approaches the estimated marginal mean value of approximately 0 (Figure \ref{fig:R-BT-proj}).
 There is no uncertainty in the projected linear trend because the values were specified as known whereas there is a small level of uncertainty in the projections of the recent average based on the uncertainty in the estimated mean.
@@
 As we would expect, there is no effect of alternative assumptions about bottom temperature in the projections for the southern stock component because the effect of the bottom temperature anomaly is only on the northern component.
-However, there alternative bottom temperature assumptions do affect projections of northern component SSB and the projections do not stabilize when the anomalies are projected to increase linearly.
-When the same AR1 assumption for the bottom temperature anomalies is used in projections, CVs of northern and total SSB projections are larger than when the recent average or linear trend are assumed (Figure \ref{fig:R-F-SSB-CVs})
-
-Unlike the annual reference points during the model years (prior to 2022), components of the SSB/R calculations are constant in the projection years where the only changing annual values are the recruitments used in the weighting for $F_{40\%}$ and the recruitments used to calculate the resulting SSB$_{40\%}$ (Figure \ref{fig:brps-proj}).
+However, the alternative bottom temperature assumptions do affect projections of northern component SSB and the projections do not stabilize when the anomalies are projected to increase linearly.
+When the same AR1 assumption for the bottom temperature anomalies is used in projections, CVs of northern and total SSB projections are larger than when the recent average or linear trend are assumed (Figure \ref{fig:R-F-SSB-CVs}).
+
+Unlike the annual reference points during the model years (prior to 2022), components of the $\upphi(\widetilde{F})$ calculations are constant in the projection years where the only changing annual values are the recruitments used in the weighting for $F_{40\%}$ and the recruitments used to calculate the resulting SSB$_{40\%}$ (Figure \ref{fig:brps-proj}).
 Estimated reference points in the first few projection years can differ depending on whether the expected recruitment or the random effects are used in their calculation, but longer term projected values are identical because the random effect estimates approach the expected values as the influence of observations diminishes.
 However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures \ref{fig:annual-F40-cvs} and \ref{fig:annual-SSB40-cvs}).
-The reduction in uncertainty using expected recruitment is less for $F_{40\%}$ than SSB$_{40\%}$ and for the northern component, that includes bottom temperature effects, than the southern component.
+The reduction in uncertainty using expected recruitment is less for $F_{40\%}$ than SSB$_{40\%}$ and for the northern component which includes bottom temperature effects than the southern component.
 
@@
 In our investigation of bottom temperature effects on recruitment and natural mortality and time-varying random effects on natural mortality at age 1, we found only including effects on recruitment for the northern stock component to be the best model with respect to AIC.
-Black sea bass in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north.
+Black sea bass in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north [@mcmahanetal20].
 Opposite effects of temperature on recruitment or population size would be expected for species in the same general area that are at the southern extent of their range [@gabriel92].
-For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer fall on the Northeast US shelf [@sullivanetal05,@milleretal16].
+For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer fall on the Northeast US shelf [@sullivanetal05;@milleretal16].
 
@@
 Although this might suggest making inferences on the population by using an ensemble of these and weighting by AIC [@burnhamanderson02], the estimates of assessment outputs relevant for management (e.g., SSB and $F$) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
+Nevertheless, continued monitoring of effects on the southern component may be prudent.
 
 We showed that using expected recruitment rather than recruitment random effects to define BRPs can produce the same estimates but with lower uncertainty.
-Using expected bottom temperature anomalies rather than the random effects would also result in the same BRPs, but presumably with even less uncertainty than those we showed using expected recruitment.
+Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same BRPs and even less uncertainty than those we showed using expected recruitment.
 These results can be important for management when uncertainty in assessment output is used in making catch advice.
@@
 For black sea bass, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figure 6).
-In the projection period, the AR1 approach reverts to the long-term mean of the environmental covariate.
+In the projection period, the AR1 models for bottom temperature and recuitment revert to the long-term means of those processes.
 This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypothesis of increasing water temperature and changing environmental conditions [@hareetal16;@pershingetal21].
@@
 We used estimates of movement parameters from a companion assessment model to define prior distributions for corresponding movement parameters in the state-space black sea bass model, but other than the tagging data, both of the assessment models used much the same data which can lead to inappropriate inferences.
-A better approach would be to estimate the movement parameters externally from just the tagging data.
+In lieu of integrating an observation model for the tagging data, a better approach would be to estimate the movement parameters externally from just the tagging data.
 
@@
 For example, we fixed the mean natural mortality for age 1 fish and @cadigan16 and @stocketal21 also estimated natural mortality deviations.
-Therefore allowing tagging observations in WHAM should be a high priority even for models with a single stock and region.
+However, the simulation studies of @milleretalinreview1 suggest estimating natural mortality in WHAM is feasible in some situations with typical assessment data.
+Regardless, tagging observations can provide improved inferences and allowing tagging observations in WHAM should be a high priority even for models with a single stock and region.
 This extension would also be beneficial for black sea bass because the current assessment uses much the same data to estimate movement outside of the model as we are using to estimate the other parameters in the WHAM model.
 
-WHAM version 2.0 only currently allows models where each stock spawns and recruits in a single region however, many stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region (references).
+WHAM version 2.0 only allows models where each stock spawns and recruits in a single region however, some stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region [@kapuretal21].
 Therefore, extending WHAM to allow such models would be beneficial.
@@
 We applied this extended version to investigate alternative hypotheses about effects of bottom temperature on recruitment and age 1 natural mortality for black sea bass.
-Our analyses indicate evidence for effects of bottom temperature on recruitment for the northern component represents was stronger than other models that included effects on age 1 mortality and/or corresponding effects on the southern stock component.
+Our analyses indicate evidence for effects of bottom temperature on recruitment for the northern component was stronger than other models that included effects on age 1 mortality and/or corresponding effects on the southern stock component.
 
@@
 
-```{r annual-BRPs, echo = FALSE, eval= TRUE, out.height = "85%", fit.align = "center", fig.cap = "Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio, equilibrium SSB at $F_{40\\%}$ based on annual inputs to SSB/R calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to SSB/R averaged over the last 5 years and recruitment is averaged over the time series.
+```{r annual-BRPs, echo = FALSE, eval= TRUE, out.height = "85%", fit.align = "center", fig.cap = "Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio, equilibrium SSB at $F_{40\\%}$ based on annual inputs to $\\upphi(\\widetilde{F})$ calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to $\\upphi(\\widetilde{F})$ averaged over the last 5 years and recruitment is averaged over the time series.
 Red and blue in first 4 rows and black and gray in final row indicate alternative annual estimates of recruitment (random effect or expected).
@@
 ```{r brps-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates by region and in total of average equilibrium $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio and SSB at $F_{40\\%}$.
-Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region and average SSB/R inputs from the last 5 years of the unprojected model.
+Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region and average $\\upphi(\\widetilde{F})$ inputs from the last 5 years of the unprojected model.
 Vertical dotted line is the last year of data and polygons represent 95\\% confidence intervals."}
@@
 
-```{r annual-F40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+```{r annual-F40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
 Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
@@
 
-```{r annual-SSB40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium SSB$_{40\\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations and alternative annual recruitment types.
+```{r annual-SSB40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium SSB$_{40\\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations and alternative annual recruitment types.
 Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
@@
 
-```{r annual-SSB-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+```{r annual-SSB-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
 Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
@@
 
-```{r annual-F-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\\bar{F}_{40\\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+```{r annual-F-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\\bar{F}_{40\\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
 Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}

### 2025-02-28 paper/paper.tex
good draft

@@
 \newcommand{\elandscape}{\end{landscape}}
+\newcommand{\noop}[1]{}
 \usepackage{booktabs}
@@
 Curti \and Alexander C. Hansell}
-\date{21 January, 2025}
+\date{28 February, 2025}
 
@@
 usage to evaluate evidence of alternative hypotheses about temperature
-effects on recruitment of black sea bass stock components.
+effects on recruitment and age 1 natural mortality of black sea bass
+stock components.
 
@@
 
-The Woods Hole Assessment Model is a state-space age-structured
+The Woods Hole Assessment Model is a general state-space age-structured
 assessment model that is used to assess and manage many stocks in the
-Northeast US. We first describe a multi-stock, multi-region extension of
-WHAM that treats the population and fleet dynamics seasonally and allows
-movement by season and region to be functions of time and age-varying
-autocorrelated random effects and environmental covariates. We then
-illustrate the model by applying it to data for the northern and
-southern components of the Northeast United States black sea bass stock
-and evaluate alternative hypotheses of bottom temperature and random
-effects on recruitment and natural mortality. We show strong evidence
-for temperature effects on recruitment, primarily for the northern stock
-component, and no evidence for including random effects or temperature
-effects on age 1 natural mortality.
+Northeast US. We first describe an extension of the model allowing any
+number of stocks (or stock components) and and regions with movement
+among regions as well as seasonal variation in stock and fleet dynamics.
+Movement rates can be a be functions of time- and age-varying random
+effects and environmental covariates. We then illustrate the model by
+applying it to data for the northern and southern components of the
+Northeast US black sea bass stock and evaluate alternative hypotheses of
+bottom temperature and time-varying random effects on recruitment and
+natural mortality. We show strong evidence for temperature effects on
+recruitment, primarily for the northern stock component, and no evidence
+for including random effects or temperature effects on age 1 natural
+mortality.
 
@@
 
-A state-space statistical approach and maximum marginal likelihood or
-Bayesian fitting of stock assessment models allows estimation of time
-and age-varying population attributes as random effects (Nielsen and
-Berg 2014; Cadigan 2016; Miller et al. 2016b). This estimation approach
-is considered an essential feature of gold-standard assessment models
-that we use in tactical management of commercially important fish stocks
-(Punt et al. 2020). The State-space Assessment Model (SAM, Nielsen and
-Berg 2014) continues to be developed and remains widely used within the
-International Council for the Exploration of the Sea (ICES) to assess
-European fish stocks. Various state-space models are being used to
-manage cod and plaice stocks in the waters of Eastern Canada (Perreault
-et al. 2020; Varkey et al. 2022) and to the south, the Woods Hole
-Assessment Model (WHAM, Miller and Stock 2020; Stock and Miller 2021) is
-now used to assess many fish stocks in the Northwest Atlantic Ocean
-(NEFSC 2022a, 2022b; NEFSC 2024).
+A state-space age-structured stock assessment models can be used to
+estiamte time and age-varying population attributes as random effects
+using maximum marginal likelihood or Bayesian fitting procedures
+(Nielsen and Berg 2014; Cadigan 2016; Miller et al. 2016b). This
+estimation approach is considered an essential feature of gold-standard
+assessment models that we use in tactical management of commercially
+important fish stocks (Punt et al. 2020). The State-space Assessment
+Model (SAM, Nielsen and Berg 2014) continues to be developed and remains
+widely used within the International Council for the Exploration of the
+Sea (ICES) to assess European fish stocks. Various state-space models
+are being used to manage cod and plaice stocks in the waters of Eastern
+Canada (Perreault et al. 2020; Varkey et al. 2022) and to the south, the
+Woods Hole Assessment Model (WHAM, Miller and Stock 2020; Stock and
+Miller 2021) is now used to assess many fish stocks in the Northwest
+Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).
 
@@
 effects options for recruitment, inter-annual transitions of numbers at
-age (hereafter referred to as ``survival''), fishery and index
+age (hereafter referred to as ``apparent survival''), fishery and index
 selectivity, natural mortality, and catchability.
@@
 linked to warming waters on the northwest shelf (Bell et al. 2015) and
-it is hypothesized that black sea bass are especially susceptible to
-climate change with likely increases in distribution and productivity
-due to warming water temperatures (Hare 2016). Thus, a focus of the
-research track assessment was to develop an assessment platform that
-could capture multiple stock components, complex seasonal movement and
-explore environmental drivers of population dynamics (NEFSC 2023).To
-illustrate the usage of WHAM 2.0 we apply it to two stock components of
-black sea bass off the coast of the NEUS using most of the model
-assumptions that were accepted during the peer-review process. However,
-here we evaluate evidence for alternative hypotheses of temporal
-variation and effects specifically of bottom temperature on recruitment
-and natural mortality of age 1 individuals.
+it is hypothesized that the distribution and productivity of the black
+sea bass stock is especially susceptible to long-term changes in
+temperature causing increases in abundance further north (Hare 2016).
+Thus, a focus of the research track assessment was to develop an
+assessment platform that could capture multiple stock components,
+complex seasonal movement and explore environmental drivers of
+population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we
+apply it to two stock components of black sea bass off the coast of the
+NEUS using most of the model assumptions that were accepted during the
+peer-review process. However, here we evaluate evidence for alternative
+hypotheses of temporal variation and effects specifically of bottom
+temperature on recruitment and natural mortality of age 1 individuals.
 
@@
 some assumptions on how they are treated in projections and reference
-points. We demonstrate some of options of WHAM version 2.0 provide
+points. We demonstrate some of options of WHAM version 2.0 to provide
 biological reference points and short-term projections.
@@
 1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987;
-Commenges 1999; Andersen and Keiding 2002). Multi-state models can
-define a probability transition matrix (PTM) that describes the
-probability of individuals occurring in different states at the end of a
-time interval \(\delta\), conditional on being in each of those state at
-beginning of the interval. For fish populations, these states would be
-defined as being alive in a particular region or deceased due to fishing
-from a particular fleet or natural mortality. The time interval \(i\)
-with duration \(\delta_i\) would be a season and the PTMs would be
-uniquely defined for each stock by season \(i\), year \(y\), and age
-\(a\) on January 1. Each row and column of the PTM correspond to one the
-states: alive in region \(r\), dead in fleet \(f\), or dead from natural
-causes. The probabilities in each row sum to unity and assume an
-individual is in the corresponding state at the beginning of the
-interval. Given \(n_R\) regions and \(n_F\) fleets, the square PTM
-(\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
-\begin{equation}\label{eq:ptm}
+Commenges 1999; Andersen and Keiding 2002). We can define a probability
+transition matrix (PTM) that describes the probability of individuals
+occurring in different states at the end of a time interval \(\delta\)
+(columns), conditional on being in each of those state at beginning of
+the interval (rows). For fish populations, these states would be defined
+as being alive in a particular region or deceased due to fishing from a
+particular fleet or natural mortality. The time interval \(i\) with
+duration \(\delta_i\) would be a season and the PTMs would be uniquely
+defined for each stock by season \(i\), year \(y\), and age \(a\) on
+January 1. Each row and column of the PTM correspond to one the states:
+alive in region \(r\), dead in fleet \(f\), or dead from natural causes.
+The probabilities in each row sum to unity (must be in one of the
+states) and assume an individual is in the state that corresponds to
+that row at the beginning of the interval. Given \(n_R\) regions and
+\(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a
+function of sub-matrices is \begin{equation}\label{eq:ptm}
   \mathbf{P}_{y,a,i} = \begin{bmatrix}
@@
 dying due to natural mortality during the interval. We have the identity
-matrix \(\mathbf{I}_{H}\) for the states for capture by each fleet and a
-1 for the state for natural mortality because the probabilities of being
-in one of the mortality states given starting the interval in that state
-is unity (no zombies allowed).
+matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a
+1 for the natural mortality state because the probabilities of being in
+one of the mortality states given starting the interval in that state is
+unity (no zombies allowed).
 
@@
 distributed as in a model for tagging data where fates of individual
-fish are assumed independent. The PTMs determine the expected numbers 1)
-in each state on January 1 of year \(t+1\) at age \(a+1\) given the
-abundances at age \(a\) on January 1 of year \(t\), 2) captured over the
-year in each fleet, 3) available to each index, and 4) alive at the time
-and in the region where spawning occurs.
+tagged fish are assumed independent. The PTMs determine the expected
+numbers 1) in each state on January 1 of year \(t+1\) at age \(a+1\)
+given the abundances at age \(a\) on January 1 of year \(t\), 2)
+captured over the year in each fleet, 3) available to each index, and 4)
+alive at the time and in the region where spawning occurs.
 
@@
 occurs instantly at either the beginning or the end of the interval.
-WHAM is configured to have movement occur after survival and mortality:
+WHAM is configured to have movement occur after survival:
 \begin{equation*}
@@
 spawning, movement, and fleet-specific behavior. The annual time step
-can be divided into \(K\) (any number) seasons and the interval size
-\(\delta_i\) for each season \(i\) does not need to be equal to any
-other seasonal interval. Under the Markov assumption, the PTM of
-surviving and moving and dying over \(K\) intervals
-\(\delta_1,\ldots, \delta_K\) (i.e., the entire year) is just the
-product of the PTMs for each interval:
+can be divided into \(K\) seasons and the interval size \(\delta_i\) for
+each season \(i\) does not need to be equal to any other seasonal
+interval. Under the Markov assumption, the PTM of surviving and moving
+and dying over \(K\) intervals \(\delta_1,\ldots, \delta_K\) (i.e., the
+entire year) is just the product of the PTMs for each interval:
 \[ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).\]
@@
 \end{equation*} where\\
-\[\text{SSB}_{s,y} =  \sum^A_{a=1}  w_{s,y,a} \text{mat}_{s,y,a} \mathbf{O}_{s,y,a,r_s}(t_s)' \mathbf{N}_{O,s,y,a}\]
-where \(w_{s,y,a}\) is the mean weight at age of spawning individuals,
-\(\text{mat}_{s,y,a}\) is the maturity at age, and
-\(\mathbf{O}_{s,y,a,r_s}(t_s)\), the \(r_s\) column of the upper-left
-submatrix of Eq. \ref{eq:ptm_spawn}, are the probabilities of surviving
-and occurring in region \(r_s\) at time \(t_s\) given being alive in
-each region at the start of the year.
+\[\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}(t_s)' \mathbf{N}_{O,y,a}\]
+and \(w_{y,a}\) is the mean weight at age of spawning individuals,
+\(\text{mat}_{y,a}\) is the maturity at age, and
+\(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and
+occurring in region \(r_s\) at time \(t_s\) given being alive in each
+region at the start of the year (the \(r_s\) column of the upper-left
+submatrix of Eq. \ref{eq:ptm_spawn}).
 
@@
 recruitment for a given stock currently only occurs in one region
-\(r_s\) there is only a single time-varying recruitment random effect
-for each stock.
+\(r_s\) there is only a single vector of time-varying recruitment random
+effects for each stock.
 
@@
 
-The initial abundances at age can also be treated as independent or as
-AR1 random effects. Defining the vector of initial abundance at age in
+The initial abundances at age can also be treated as independent or AR1
+random effects. Defining the vector of initial abundance at age in
 region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
@@
 For each season, there are at most \(n_R-1\) parameters determining
-movement among regions given starting an the season in region \(r\) in
+movement among regions given starting the season in region \(r\) in
 either the sequential or simultaneous configurations. Movement
@@
 movement parameter from region \(r\) to \(r'\) in season \(i\) and year
-\(y\) for individuals of age \(a\) is a linear function of both random
-and environmental effects: \begin{equation*}
+\(y\) for individuals of age \(a\) is a linear function of an intercept
+or mean parameter (\(\theta_{r\rightarrow r',i}\)) and both random and
+environmental effects: \begin{equation*}
   g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
 \end{equation*} The random effects
-\(\varepsilon_{r\rightarrow r',y,a,i}\) are season-, and(or)
+\(\varepsilon_{r\rightarrow r',y,a,i}\) are season-, and
 region-to-region-specific and modeled most generally as 2DAR1 random
@@
 natural mortality, and selectivity. Effects of covariate \(E_k\) can be
-age-, season-, and(or) region-to-region-specific
+age-, season-, and region-to-region-specific
 \(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial
@@
 There is currently no likelihood component for tagging data. Therefore,
-movement parameters would generally either need to be fixed or assumed
-to have some prior distribution, possibly based on external parameter
-estimates. We include prior distributions for the season and
-region-to-region specific (mean) movement parameters which are treated
-as random effects with the mean defined by the initial value of the
-fixed effect counterpart and standard deviation \begin{equation*}
+movement parameters would usually either need to be fixed or assumed to
+have some prior distribution, possibly based on external parameter
+estimates. WHAM includes prior distributions for the season and
+region-to-region specific (mean) movement parameters and treats them as
+random effects with the mean defined by the initial value of the fixed
+effect counterpart and standard deviation \begin{equation*}
   \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
@@
 
-Natural mortality options have been expanded since initial WHAM
-development. When not estimated, (mean) mortality rates may be stock-,
-region-, and age-specific. When random effects are used, the same 2DAR1
-structure with age and year as described by Stock and Miller (2021) can
-be configured for a given stock and region. Any environmental covariate
-effects can be stock-, region-, and age-specific. So the general model
-for natural mortality is \begin{equation*}
+When treated as constant parameters, natural mortality rates may be
+stock-, region-, and age-specific. When random effects are used, the
+same 2DAR1 structure with age and year as described by Stock and Miller
+(2021) can be configured for a given stock and region and any
+environmental covariate effects can be stock-, region-, and
+age-specific. So the general model for natural mortality is
+\begin{equation*}
   \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
@@
 
-Catchability of the index can also be treated as functions of normal
-random effects and(or) environmental covariate effects
+Catchability of the index can also be treated as logit functions of
+normal random effects and(or) environmental covariate effects
 \[log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}\]
-where \(u_{m}\) and \(l_{m}\) are the upper and lower bounds of
-catchability for index \(m\) (defaults are 0 and 1000) and the general
-covariance structure for the annual random effects is AR1
+where \(\theta_{q,m}\) is an intercept or mean parameter, and \(u_{m}\)
+and \(l_{m}\) are the upper and lower bounds of catchability for index
+\(m\) (defaults are 0 and 1000). The general covariance structure for
+the annual random effects is AR1
 \[Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.\]
-
-\hypertarget{weight-and-maturity-at-age}{%
-\subsubsection*{Weight and Maturity at
-age}\label{weight-and-maturity-at-age}}
-\addcontentsline{toc}{subsubsection}{Weight and Maturity at age}
-
-Weight and maturity at age are treated similarly to Stock and Miller
-(2021). Annual weight at age matrices for each fleet are used to
-calculate total catch and the weight at age is applied to catch numbers
-at age for any stocks caught by the fleet. Similarly, when indices
-and(or) associated age composition observations are measured in biomass,
-the weight at age matrices for the index are applied to predicted
-numbers at age of all stocks observed by the survey. Unique weight at
-age and maturity at age matrices are allowed for each stock to calculate
-spawning stock biomass.
 
@@
 
-Currently a single F reference point \(\widetilde F\) is estimated
-across stocks and regions and F by fleet and age is
+Currently a single fishing mortality reference point \(\widetilde F\) is
+estimated across stocks and regions and the fleet- and age-specific
+reference points is
 \(\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}\). Selectivity is
-determined as before when there are multiple fleets where
-\(\text{sel}_{f,a}\) is determined by averaging F at age over a
-user-defined set of years \begin{equation*}
+determined across fleets and ages as in the equilibrium assumption for
+initial abundance at age except here it is a function of average F at
+age over a user-defined set of years \begin{equation}\label{eq:spr_sel}
   \text{sel}_{a,f} = \frac{\overline F_{a,f}}{\max_a \sum^{n_F}_{f=1}{\overline F}_{a,f}}
-\end{equation*}
-
-The equilibrium spawning stock biomass per recruit for stock \(s\) in
+\end{equation}
+
+The equilibrium spawning stock biomass per recruit for a given stock in
 region \(r_s\) is defined as \begin{equation}\label{eq:ssbpr}
- \upphi_s(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{s,a,r_s,\cdot} \mathbf{O}_{s,a,\cdot,r_s}(t_s) w_{s,a} m_{s,a}
-\end{equation} where \(w_{s,a}\) and \(m_{s,a}\) are the mean individual
+ \upphi(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{O}_{a,\cdot,r_s}(t_s) w_{a} m_{a}
+\end{equation} where \(w_{a}\) and \(m_{a}\) are the mean individual
 weight and probability of maturity at age \(a\),
-\(\widetilde{\mathbf{O}}_{s,a}\) are as described in Eq. \ref{eq:npraa},
-and \(\mathbf{O}_{s,a}(t_s)\) is the \(n_R \times n_R\) upper-left
+\(\widetilde{\mathbf{O}}_{a}\) are as described in Eq. \ref{eq:npraa},
+and \(\mathbf{O}_{a}(t_s)\) is the \(n_R \times n_R\) upper-left
 sub-matrix of eq. \ref{eq:ptm_spawn} with the probabilities of surviving
@@
 region \(r\) at age \(a\). The further subscripts \(r_s,\cdot\) and
-\(\cdot,r_s\) indictate row or column \(r_s\), respectively. Using these
+\(\cdot,r_s\) indicate row or column \(r_s\), respectively. Using these
 rows and columns is required because of the assumption that spawning and
@@
 conditional on the region of recruitment \(r_s\). The equilibrium
-recruitment in each region \(\widetilde {\mathbf{N}}_{s,1}\), depends on
-the stock dynamics. This version of WHAM currently only allows complete
-spawning region fidelity so that a stock only spawns and recruits in a
-single region. In this case, \(\widetilde {\mathbf{N}}_{s,1}\) will be
-positive in the spawning region (\(r_s\)) and zero elsewhere. Similarly,
-the row \(r_s\) of \(\mathbf{O}_s\) will be zero off of the diagonal.
-The matrices of probabilities of surviving and occurring in each region,
-\(\widetilde{\mathbf{O}}_{s,a}\) and \(\mathbf{O}_{s,a}(\delta_s)\), are
-functions of the fishing mortality rates for fleets in each region
-\(\widetilde F_{f,a}\).
-
-The matrix equilibrium yield per recruit as a function of
-\(\widetilde F\) is calculated as \begin{equation}\label{eq:ypr}
- \widetilde{Y}_s({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{s,a,r_s,\cdot} \mathbf{H}_{s,a} \mathbf{c}_{s,a}
+recruitment depends on the stock dynamics. WHAM currently only allows
+complete spawning region fidelity so that a stock only spawns and
+recruits in a single region. Therefore, the equilibrium vector of
+numbers at age 1 for each state, \(\widetilde {\mathbf{N}}_{1}\), will
+only be nonzero in the row corresponding to the spawning region
+(\(r_s\)). The matrices of probabilities of surviving and occurring in
+each region, \(\widetilde{\mathbf{O}}_{s,a}\) and
+\(\mathbf{O}_{s,a}(t_s)\), are functions of the fishing mortality rates
+for fleets in each region \(\widetilde F_{f,a}\).
+
+The matrix of equilibrium yield per recruit by region and fleet as a
+function of \(\widetilde F\) is calculated as
+\begin{equation}\label{eq:ypr}
+ \widetilde{\nu}({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{H}_{a} \mathbf{c}_{s,a}
 \end{equation} where \(\mathbf{c}_a\) is the vector of mean individual
@@
 
-As in previous versions of WHAM package, ``static'\,' reference points,
-typically meant to be defined for prevailing conditions, average all of
-the inputs to the spawning biomass and yield per recruit calculations
-over the user-specified years (e.g., last 5 years of the model). This
-same averaging is also applied to possibly time-varying movement
-parameters.
+As in previous versions of WHAM package, reference points for prevailing
+conditions, average all of the inputs to the spawning biomass and yield
+per recruit calculations (weight, maturity, natural mortality, and
+movement) over a set of user-specified years as shown in eq.
+\ref{eq:spr_sel} for selectivity (e.g., last 5 years of the model).
 
@@
 hypothesized environmental drivers that were prioritized research
-recommendations from previous black sea bass assessments.
-
-Below we describe the assumptions and configuration of the assessment
-model as determined by the Working Group as well as the alternative
-assumptions for recruitment and natural mortality in the models we fit
-to evaluate alternative hypotheses of bottom temperature effects on
-black sea bass.
+recommendations from previous black sea bass assessments. Below we
+describe the assumptions and configuration of the assessment model as
+determined by the Working Group as well as the alternative assumptions
+for recruitment and natural mortality in the models we fit to evaluate
+alternative hypotheses of bottom temperature effects on black sea bass.
 
@@
 
-There are 11 seasonal intervals within each calendar year: five monthly
-time intervals from Jan 1 to May 31, a spawning season from June 1 to
-July 31, and five monthly intervals from August 1 to December 31. The
-southern stock component is assumed to never move to the northern
+There are 11 ``seasonal'' intervals within each calendar year: five
+monthly time intervals from Jan 1 to May 31, a spawning season from June
+1 to July 31, and five monthly intervals from August 1 to December 31.
+The southern stock component is assumed to never move to the northern
 region. For the northern component, a proportion
@@
 allowed. However, the Working Group also fit a Stock Synthesis model
-(Methot and Wetzel 2013) which provided estimated movement rate
-parameters and standard errors that were used to configure the priors
-for WHAM (see Supplementary Materials).
-
-With the movement configuration, the northern origin fish (ages 2+) can
+(Methot and Wetzel 2013) which included tagging data and provided
+estimated movement rate parameters and standard errors that were used to
+configure the priors for WHAM (see Supplementary Materials).
+
+With movement as configured, the northern origin fish (ages 2+) can
 occur in the southern region on January 1. Estimating initial numbers at
@@
 southern region in the initial year of the model. Therefore, we used the
-simplifying equilibrium assumption described above where there are two
-parameters estimated for each stock component: an initial recruitment
-and an equilibrium fully-selected \(F\) that determines the abundance at
-age in each region for each stock component.
-
-For the northern population abundance at age 1 on January 1
-(recruitment) is only allowed in the northern region, but given the
-monthly movement described above, older individuals that previously
-recruited in the northern region may occur in the southern region on
-January 1. Therefore, a model with survival random effects will model
-the transitions (survival/movement) of abundances at age of northern
-origin fish in each region. All of the initial runs assumed variance
-parameters for these random effects to be the same for northern origin
-fish occurring in both regions on January 1. The base model assumes very
-small variance for the transitions of northern fish in the southern
-region, which is approximately the same as the deterministic transition
-assumptions of a statistical catch at age model. We also allow 2DAR1
-correlation for recruitment and survival for both the northern and
-southern components. Unique variance and correlation parameters for the
-recruitment and survival random effects are estimated for the northern
-and southern components.
+simplifying equilibrium assumption where there are two parameters
+estimated for each stock component: an initial recruitment and an
+equilibrium fully-selected \(F\) that determines the abundance at age in
+each region for each stock component.
+
+For subsequent years, all models allow AR1 correlation of recruitment
+and 2DAR1 correlation for survival for both the northern and southern
+stock components. Models with apparent survival random effects will
+model the transitions (conditional on both survival and movement) of
+abundances at age of northern origin fish in each region, but there
+appears to be little ability of the existing data from the southern
+region to distinguish apparent survival random effects for each stock
+component. All models we considered assume very small variance for the
+transitions of northern fish in the southern region, which is
+essentially the same as the deterministic transition assumptions of a
+statistical catch at age model. Unique variance and correlation
+parameters for the recruitment and apparent survival random effects are
+estimated for the northern and southern components.
 
@@
 The estimated coefficients of variation (CVs) provided by the analyses
-to generate the recreational catch per angler (recreational CPA) index
-ranged between 0.02 and 0.06 which the Working Group felt did not
-capture the true observation uncertainty in the index with regard to its
-relationship to stock abundance. In many of the initial runs as well as
-the base model we allowed a scalar multiple of the standard deviation of
-the log aggregate index to be estimated for these indices in the
-northern and southern regions. Models that successfully estimated these
-scalars indicated standard deviations for these surveys to be
-approximately 5 times the input value and this value was fixed in many
-preliminary runs to avoid dealing with convergence problems. However,
-the base model successfully estimated these scalars. The model estimates
-are negligibly affected by estimating these scalars, but we felt
-estimating these parameters allowed uncertainty in model output to be
-more properly conveyed.
+by the Working Group to generate the recreational catch per angler
+(recreational CPA) index ranged between 0.02 and 0.06 which the Working
+Group felt did not capture the true observation uncertainty in the index
+with regard to its relationship to stock abundance. All models we
+considered allowed a scalar multiple of the standard deviation of the
+log aggregate index to be estimated for these indices in the northern
+and southern regions. However, lack of model convergence was observed in
+many fits for self-test simulations. The model estimates are negligibly
+affected by estimating these scalars, but estimating rather than fixing
+these parameters allowed uncertainty in model output to be more properly
+conveyed.
 
@@
 The Working Group investigated many alternative assumptions for the
-probability models and selectivity models for the eight different sets
-of age composition observations to reduce residual patterns and
+probability models and selectivity forms for the eight different sets of
+age composition observations to reduce residual patterns and
 retrospective patterns. These analyses resulted in use of selectivity
@@
 
-All model fits also include bottom temperature observations for the
-northern and southern regions from 1963 to 2021 and estimated standard
-errors ranging between 0.03 and 0.09 degrees Celsius (NEFSC 2023). We
-retained the assumption from the peer-reviewed assessment that treated
-the latent bottom temperature covariates in each region as AR1
-processes.
+All models include bottom temperature observations for the northern and
+southern regions from 1963 to 2021 and standard errors of observations
+ranged between 0.03 and 0.09 degrees Celsius (NEFSC 2023). We retained
+the assumption from the peer-reviewed assessment that treated the latent
+bottom temperature covariates in each region as AR1 processes.
 
@@
 fish turn from age 0 to age 1 on January 1 (Miller et al. 2016a). This
-temperature may be a proxy for temperature prior to January 1 and affect
+covariate may be a proxy for temperature prior to January 1 and affect
 survival during the end of the pre-recruit phase or natural mortality in
@@
 
-It is standard practice to treat annual recruitment as time-varying
-deviations from a mean model and all models here treat recruitment
-deviations as AR1 random effects. However, Miller et al. (2018) showed
-how inferences of temperature effects on growth or maturity parameters
-can be very different whether the compared models with and without the
-effect also include random effects representing residual temporal
-variation in parameters. Therefore, we also explored whether including
-temporal random effects on age 1 natural mortality affected inferences
-on corresponding temperature effects. Initially, we included random
-effects on age 1 natural mortality for both the northern and southern
-stock components, but estimates of these random effects and
+All models treat annual recruitment as AR1 time-varying random effects
+with mean parameters for each stock component. However, Miller et al.
+(2018) showed how inferences of temperature effects on growth or
+maturity parameters can be very different whether the compared models
+with and without the effect also include random effects representing
+residual temporal variation in parameters. Therefore, we also explored
+whether including temporal random effects on age 1 natural mortality
+affected inferences on corresponding temperature effects. Initially, we
+included random effects on age 1 natural mortality for both the northern
+and southern stock components, but estimates of these random effects and
 corresponding variance for the southern component converged to 0 so
@@
 
-\hypertarget{model-fitting-diagnostics-reference-points-and-projections}{%
-\subsection*{Model fitting, diagnostics, reference points, and
-projections}\label{model-fitting-diagnostics-reference-points-and-projections}}
-\addcontentsline{toc}{subsection}{Model fitting, diagnostics, reference
-points, and projections}
+\hypertarget{model-fitting-and-diagnostics}{%
+\subsection*{Model fitting and
+diagnostics}\label{model-fitting-and-diagnostics}}
+\addcontentsline{toc}{subsection}{Model fitting and diagnostics}
 
@@
 values of Mohn's \(\rho\) near 0 imply no pattern in estimation of these
-quantities as the time-series is sequentially extended.
-
-As in Miller et al. (2016b), we also assessed the consistency of the
-AIC-based model selection over retrospective peels to guard against
-previously noted changes in perception of covariate effects on
-recruitment with increased length of the time series of observations
-(Myers 1998). This retrospective examination was also recommended by
-Brooks (2024).
+quantities as the time-series is sequentially extended. As in Miller et
+al. (2016b), we also assessed the consistency of the AIC-based model
+selection over retrospective peels to guard against previously noted
+changes in perception of covariate effects on recruitment with increased
+length of the time series of observations (Myers 1998). This
+retrospective examination was also recommended by Brooks (2024).
 
@@
 starting values from a multivariate normal distribution with mean and
-covariance defined by the MLEs and estimated covariance matrix from the
-fitted model. See the Supplementary Materials for more details.
-
-For the best performing model, we also performed so-called simulation
-self-tests where new observations were simulated conditional on all
-estimated fixed and random effects and the same model configuration was
-fit to each of the simulated data sets. We estimated median relative
-bias of SSB across these simulations. See the Supplementary Materials
-for more details.
+covariance defined by the MLEs and hessian-based covariance matrix from
+the fitted model. See the Supplementary Materials for more details.
+
+For the best performing model, we also performed simulation self-tests
+where new observations were simulated conditional on all estimated fixed
+and random effects and the same model configuration was fit to each of
+the simulated data sets. We estimated median relative bias of SSB across
+these simulations. See the Supplementary Materials for more details.
+
+\hypertarget{reference-points-and-projections}{%
+\subsection*{Reference points, and
+projections}\label{reference-points-and-projections}}
+\addcontentsline{toc}{subsection}{Reference points, and projections}
 
@@
 at the value in the last model year (2021). Average weight, natural
-mortality, movement, and maturity at age are used in projection years
-whereas the AR1 processes for all numbers at age random effects are
-continued.
+mortality, movement, and maturity at age over the last 5 years are used
+in projection years whereas the AR1 processes for all numbers at age
+random effects are continued.
 
@@
 
-The 2dAR1 random effects for the selectivity parameters of the northern
+The 2DAR1 random effects for the selectivity parameters of the northern
 recreational fleet indicate a decrease in selectivity at ages 2 to 6
@@
 Using the expected recruitment rather than the random effects as the
-multiplier with equilibrium SSB/R results in less variable estimates of
-annual SSB\(_{40\%}\) (Figure \ref{fig:annual-BRPs}: top row). Estimates
-were less variable for the southern stock component than the northern
-stock component mainly because the expected recruitment is constant for
-the former whereas that for the northern component varies via the effect
-of the bottom temperature anomalies. Temporal variation for the southern
-stock component is mainly caused by annual variation in the inputs to
-the SSB/R calculation and \(F_{40\%}\), but there is also some effect of
-the variation in the proportion of total recruitment for each stock
-component as used in the weighting terms to define the 40\% reduction in
-unfished SSB/R (\(\lambda_s\) in Eq. \ref{eq:newton-obj}). Using the
-recruitment random effects instead of expected recruitment creates more
-uncertain estimates of SSB\(_{40\%}\) for the southern stock component
-whereas there is little difference for the northern component because
-there is less difference in the types of recruitment estimates resulting
-from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
+multiplier with equilibrium spawning biomass per recruit
+(\(\upphi(\widetilde{F})\), eq. \ref{eq:ssbpr}) results in less variable
+estimates of annual SSB\(_{40\%}\) (Figure \ref{fig:annual-BRPs}: top
+row). Estimates were less variable for the southern stock component than
+the northern stock component mainly because the expected recruitment is
+constant for the former whereas that for the northern component varies
+via the effect of the bottom temperature anomalies. Temporal variation
+for the southern stock component is mainly caused by annual variation in
+the inputs to the \(\upphi(\widetilde{F})\) calculation and
+\(F_{40\%}\), but there is also some effect of the variation in the
+proportion of total recruitment for each stock component as used in the
+weighting terms to define the 40\% reduction in unfished spawning
+biomass per recruit (\(\lambda_s\) in Eq. \ref{eq:newton-obj}). Using
+the recruitment random effects instead of expected recruitment creates
+more uncertain estimates of SSB\(_{40\%}\) for the southern stock
+component whereas there is little difference for the northern component
+because there is less difference in the types of recruitment estimates
+resulting from the covariate effects (Figure
+\ref{fig:annual-SSB40-cvs}).
 
@@
 comparable to using median recruitment and the annual variation in the
-inputs to SSB/R indicates no years where SSB was greater than unfished
-levels throughout the time series. The patterns in uncertainty of SSB
-ratios are the same as those for the SSB reference points (Figure
-\ref{fig:annual-SSB-status-cvs}).
+inputs to \(\upphi(\widetilde{F})\) indicates no years where SSB was
+greater than unfished levels throughout the time series. The patterns in
+uncertainty of SSB ratios are the same as those for the SSB reference
+points (Figure \ref{fig:annual-SSB-status-cvs}).
 
@@
 SSB\(_{40\%}\) because the recruitment estimates are only used in the
-weighting of the stock-specific SSB/R to determine \(F_{40\%}\)
-(\(\lambda_s\) in Eq. \ref{eq:newton-obj}, Figure \ref{fig:annual-BRPs}:
-third row). For similar reasons, estimates of uncertainty of
-\(F_{40\%}\) are also similar between the type of recruitment estimates
-(Figure \ref{fig:annual-F40-cvs}). Annual estimates for the northern
-stock component were lowest in the earliest years of the time series
-whereas those for the soutern component are relatively stable over time.
+weighting of the stock-specific \(\upphi(\widetilde{F})\) to determine
+\(F_{40\%}\) (\(\lambda_s\) in Eq. \ref{eq:newton-obj}, Figure
+\ref{fig:annual-BRPs}: third row). For similar reasons, estimates of
+uncertainty of \(F_{40\%}\) are also similar between the type of
+recruitment estimates (Figure \ref{fig:annual-F40-cvs}). Annual
+estimates for the northern stock component were lowest in the earliest
+years of the time series whereas those for the soutern component are
+relatively stable over time.
 
@@
 \ref{fig:annual-BRPs}: bottom row). Although the alternative types of
-recruitment estimates result in different points on the kobe plots for
+recruitment estimates result in different points on the Kobe plots for
 joint status, the uncertainty in the estimates suggest similar
@@
 When continuing the AR1 model for the bottom temperature anomalies in
-the projection years we can see that the uncertainty grows and
-asymptotes to a level determined by the marginal variance of the process
-and that the predicted value approaches the estimated marginal mean
-value of approximately 0 (Figure \ref{fig:R-BT-proj}). There is no
-uncertainty in the projected linear trend because the values were
-specified as known whereas there is a small level of uncertainty in the
-projections of the recent average based on the uncertainty in the
-estimated mean. Comparing projected recruitment for the northern and
-southern stock components, the projected expected recruitment has much
-lower uncertainty than the projected random effects regardless of
-whether bottom temperature affects recruitment. For the northern
-component, CVs of projected expected recruitment are smaller than those
-of recruitment random effect, but there are larger differences between
-CVs of projected expected recruitment assuming AR1 and the other
-scenarios for the future bottom temperature anomalies are made (Figure
-\ref{fig:R-F-SSB-CVs}).
+the projection years, the uncertainty grows and asymptotes to a level
+determined by the marginal variance of the process and the predicted
+value approaches the estimated marginal mean value of approximately 0
+(Figure \ref{fig:R-BT-proj}). There is no uncertainty in the projected
+linear trend because the values were specified as known whereas there is
+a small level of uncertainty in the projections of the recent average
+based on the uncertainty in the estimated mean. Comparing projected
+recruitment for the northern and southern stock components, the
+projected expected recruitment has much lower uncertainty than the
+projected random effects regardless of whether bottom temperature
+affects recruitment. For the northern component, CVs of projected
+expected recruitment are smaller than those of recruitment random
+effect, but there are larger differences between CVs of projected
+expected recruitment assuming AR1 and the other scenarios for the future
+bottom temperature anomalies are made (Figure \ref{fig:R-F-SSB-CVs}).
 
@@
 bottom temperature anomaly is only on the northern component. However,
-there alternative bottom temperature assumptions do affect projections
-of northern component SSB and the projections do not stabilize when the
+the alternative bottom temperature assumptions do affect projections of
+northern component SSB and the projections do not stabilize when the
 anomalies are projected to increase linearly. When the same AR1
@@
 recent average or linear trend are assumed (Figure
-\ref{fig:R-F-SSB-CVs})
+\ref{fig:R-F-SSB-CVs}).
 
 Unlike the annual reference points during the model years (prior to
-2022), components of the SSB/R calculations are constant in the
-projection years where the only changing annual values are the
-recruitments used in the weighting for \(F_{40\%}\) and the recruitments
-used to calculate the resulting SSB\(_{40\%}\) (Figure
+2022), components of the \(\upphi(\widetilde{F})\) calculations are
+constant in the projection years where the only changing annual values
+are the recruitments used in the weighting for \(F_{40\%}\) and the
+recruitments used to calculate the resulting SSB\(_{40\%}\) (Figure
 \ref{fig:brps-proj}). Estimated reference points in the first few
@@
 recruitment is less for \(F_{40\%}\) than SSB\(_{40\%}\) and for the
-northern component, that includes bottom temperature effects, than the
+northern component which includes bottom temperature effects than the
 southern component.
@@
 northern stock component is consistent with its range extension to the
-north. Opposite effects of temperature on recruitment or population size
-would be expected for species in the same general area that are at the
-southern extent of their range (Gabriel 1992). For example, higher
-recruitment of the Southern New England-Mid-Atlantic stock of yellowtail
-flounder is correlated with more cold water persistence into summer fall
-on the Northeast US shelf Miller et al. (2016b).
+north (McMahan et al. 2020). Opposite effects of temperature on
+recruitment or population size would be expected for species in the same
+general area that are at the southern extent of their range (Gabriel
+1992). For example, higher recruitment of the Southern New
+England-Mid-Atlantic stock of yellowtail flounder is correlated with
+more cold water persistence into summer fall on the Northeast US shelf
+(Sullivan et al. 2005; Miller et al. 2016b).
 
@@
 similar among the models and weighted estimates would differ very little
-from the estimates with the lowest AIC.
+from the estimates with the lowest AIC. Nevertheless, continued
+monitoring of effects on the southern component may be prudent.
 
@@
 effects to define BRPs can produce the same estimates but with lower
-uncertainty. Using expected bottom temperature anomalies rather than the
-random effects would also result in the same BRPs, but presumably with
-even less uncertainty than those we showed using expected recruitment.
-These results can be important for management when uncertainty in
-assessment output is used in making catch advice. The typical approach
-for defining stock and harvest status is to compare BRPs defined under
-prevailing conditions with current stock and harvest levels, but we
-could also make the comparisons with projected BRPs like those we
-demonstrated are possible. The uncertainty in these ratios could also be
-used in defining catch advice.
+uncertainty. Although the option is unavailable currently in WHAM, we
+expect using expected bottom temperature anomalies rather than the
+random effects would also result in the same BRPs and even less
+uncertainty than those we showed using expected recruitment. These
+results can be important for management when uncertainty in assessment
+output is used in making catch advice. The typical approach for defining
+stock and harvest status is to compare BRPs defined under prevailing
+conditions with current stock and harvest levels, but we could also make
+the comparisons with projected BRPs like those we demonstrated are
+possible. The uncertainty in these ratios could also be used in defining
+catch advice.
 
@@
 of recruitment and SSB (Figure 6). In the projection period, the AR1
-approach reverts to the long-term mean of the environmental covariate.
-This is most likely not an appropriate assumption for stocks in the NEUS
-given observations and hypothesis of increasing water temperature and
-changing environmental conditions (Hare 2016; Pershing et al. 2021).
-Because of these concerns, the most recent stock assessment for
-yellowtail flounder used a change point analysis on environmental
-covariates to determine current conditions. The mean environmental
-covariate from the current condition was then used as a data input in
-WHAM to inform short-term projections (NEFSC, 2024). Unfortunately, this
-methodology still has limitations because it is unlikely that current
-environmental conditions are stable. Similarly, a linear increase of
-temperature is also not likely in the future (see past trends, Figure 6
-top row).As more assessments begin including effects of environmental
-covariates explicitly in assessment models, assumptions about these
-covariates in calculated BRPs and population projections becomes an
-important decision in the management process.
+models for bottom temperature and recuitment revert to the long-term
+means of those processes. This is most likely not an appropriate
+assumption for stocks in the NEUS given observations and hypothesis of
+increasing water temperature and changing environmental conditions (Hare
+2016; Pershing et al. 2021). Because of these concerns, the most recent
+stock assessment for yellowtail flounder used a change point analysis on
+environmental covariates to determine current conditions. The mean
+environmental covariate from the current condition was then used as a
+data input in WHAM to inform short-term projections (NEFSC, 2024).
+Unfortunately, this methodology still has limitations because it is
+unlikely that current environmental conditions are stable. Similarly, a
+linear increase of temperature is also not likely in the future (see
+past trends, Figure 6 top row).As more assessments begin including
+effects of environmental covariates explicitly in assessment models,
+assumptions about these covariates in calculated BRPs and population
+projections becomes an important decision in the management process.
 
@@
 assessment models used much the same data which can lead to
-inappropriate inferences. A better approach would be to estimate the
+inappropriate inferences. In lieu of integrating an observation model
+for the tagging data, a better approach would be to estimate the
 movement parameters externally from just the tagging data.
@@
 1 fish and Cadigan (2016) and Stock et al. (2021) also estimated natural
-mortality deviations. Therefore allowing tagging observations in WHAM
-should be a high priority even for models with a single stock and
-region. This extension would also be beneficial for black sea bass
-because the current assessment uses much the same data to estimate
-movement outside of the model as we are using to estimate the other
-parameters in the WHAM model.
-
-WHAM version 2.0 only currently allows models where each stock spawns
-and recruits in a single region however, many stocks are assessed with
-spatial units that are assumed to have a global stock and recruitment
-model that distributes recruitment to each region (references).
-Therefore, extending WHAM to allow such models would be beneficial.
-Similarly age composition data are scarce for many stocks incorporating
-the work by Correa et al. (2023) on allowing length composition and
-modeling growth within WHAM would provide an assessment tool to include
-process errors in assessment with such data and possibly with multiple
-stocks and regions.
+mortality deviations. However, the simulation studies of Miller et al.
+(In review) suggest estimating natural mortality in WHAM is feasible in
+some situations with typical assessment data. Regardless, tagging
+observations can provide improved inferences and allowing tagging
+observations in WHAM should be a high priority even for models with a
+single stock and region. This extension would also be beneficial for
+black sea bass because the current assessment uses much the same data to
+estimate movement outside of the model as we are using to estimate the
+other parameters in the WHAM model.
+
+WHAM version 2.0 only allows models where each stock spawns and recruits
+in a single region however, some stocks are assessed with spatial units
+that are assumed to have a global stock and recruitment model that
+distributes recruitment to each region (Kapur et al. 2021). Therefore,
+extending WHAM to allow such models would be beneficial. Similarly age
+composition data are scarce for many stocks incorporating the work by
+Correa et al. (2023) on allowing length composition and modeling growth
+within WHAM would provide an assessment tool to include process errors
+in assessment with such data and possibly with multiple stocks and
+regions.
 
@@
 mortality for black sea bass. Our analyses indicate evidence for effects
-of bottom temperature on recruitment for the northern component
-represents was stronger than other models that included effects on age 1
-mortality and/or corresponding effects on the southern stock component.
+of bottom temperature on recruitment for the northern component was
+stronger than other models that included effects on age 1 mortality
+and/or corresponding effects on the southern stock component.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-mcmahanetal20}{}}%
+McMahan, M.D., Sherwood, G.D., and Grabowski, J.H. 2020. Geographic
+variation in life-history traits of black sea bass
+(\emph{{C}entropristis striata}) during a rapid range expansion.
+Frontiers in Marine Science \textbf{7}: 567758.
+doi:\href{https://doi.org/10.3389/fmars.2020.567758}{10.3389/fmars.2020.567758}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-methotwetzel13}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-milleretalinreview1}{}}%
+Miller, T.J., Britten, G., Brooks, E.N., Fay, G., Hansell, A., Legault,
+C.M., Li, C., Muffley, B., Stock, B.C., and Wiedenmann, J. In review.
+Factors affecting reliablity of state-space age-structured assessment
+models. Canadian Journal of Fisheries Science.
+
 \leavevmode\vadjust pre{\hypertarget{ref-milleretal16}{}}%
@@
 age-structured assessment model with an application to southern {New
-England} yellowtail flounder. Can. J. Fish. Aquat. Sci. \textbf{73}(8):
-1261--1270.
+England} yellowtail flounder. Canadian Journal of Fisheries Science
+\textbf{73}(8): 1261--1270.
 doi:\href{https://doi.org/10.1139/cjfas-2015-0339}{10.1139/cjfas-2015-0339}.
@@
 environmental variation in growth and maturity and effects on management
-reference points of {Georges Bank Atlantic} cod. Can. J. Fish. Aquat.
-Sci. \textbf{75}(12): 2159--2171.
+reference points of {Georges Bank Atlantic} cod. Canadian Journal of
+Fisheries Science \textbf{75}(12): 2159--2171.
 doi:\href{https://doi.org/10.1139/cjfas-2017-0124}{10.1139/cjfas-2017-0124}.
@@
 
-\caption{Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio, equilibrium SSB at $F_{40\%}$ based on annual inputs to SSB/R calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to SSB/R averaged over the last 5 years and recruitment is averaged over the time series.
+\caption{Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio, equilibrium SSB at $F_{40\%}$ based on annual inputs to $\upphi(\widetilde{F})$ calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to $\upphi(\widetilde{F})$ averaged over the last 5 years and recruitment is averaged over the time series.
 Red and blue in first 4 rows and black and gray in final row indicate alternative annual estimates of recruitment (random effect or expected).
@@
 \caption{Annual estimates by region and in total of average equilibrium $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio and SSB at $F_{40\%}$.
-Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region and average SSB/R inputs from the last 5 years of the unprojected model.
+Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region and average $\upphi(\widetilde{F})$ inputs from the last 5 years of the unprojected model.
 Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:brps-proj}
@@
 
-\caption{Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+\caption{Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
 Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-F40-cvs}
@@
 
-\caption{Coefficients of variation for annual equilibrium SSB$_{40\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations and alternative annual recruitment types.
+\caption{Coefficients of variation for annual equilibrium SSB$_{40\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations and alternative annual recruitment types.
 Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-SSB40-cvs}
@@
 
-\caption{Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+\caption{Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
 Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-SSB-status-cvs}
@@
 
-\caption{Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\bar{F}_{40\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to SSB/R calculations.
+\caption{Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\bar{F}_{40\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
 Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-F-status-cvs}

### 2025-03-04 paper/paper.Rmd
add edits, map fig

@@
 output:
-  pdf_document:
+  bookdown::pdf_document2:
     keep_tex: yes
+    toc: no
+    number_sections: yes
     fig_caption: yes
-    number_sections: yes
     includes:
       in_header: options.sty
-  html_document:
-    df_print: paged
+#    citation_package: biblatex
+#biblio-style: authoryear
+#  pdf_document:
+#    keep_tex: yes
+#    fig_caption: yes
+#    number_sections: yes
+#    includes:
+#      in_header: options.sty
+#  html_document:
+#    df_print: paged
 csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
@@
 Prior to its 2023 peer-reviewed assessment, the NEUS black sea bass stock was assessed using the Age-Structured Assessment Program (ASAP) model [@legaultrestrepo99], a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
-Northern and southern components of the NEUS black sea bass stock ascribed to regions divided by the Hudson Canyon were separately modeled in ASAP (reference map figure).
+Northern and southern components of the NEUS black sea bass stock ascribed to regions divided by the Hudson Canyon were separately modeled in ASAP (Figure \ref{fig:map}).
 Results from the separate ASAP models were combined for a unit-stock assessment.
@@
 
-## General Model aspects {-}
+## General model {-}
 
@@
 Version 2.0 extends these inferences to movement rates for multi-region models and stock (component) specific configuration of random and environmental covariate effects.
-
-## Future developments {-}
 
@@
 
+<!--
+
 ## Conclusions {-}
@@
 
+-->
+
 # References {-}
@@
 # Figures {-}
+
+<!--
+(ref:caption)
+
+\begin{figure}
+\begin{center}
+\includegraphics{map}
+\end{center}
+\caption{The Northwest Atlantic coastal area where the NEUS black sea bass stock occurs.
+Red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations \cite{milleretal16_yoy_survival} and the bathymetric contour is the 400 m isocline.}
+\end{figure}
+\pagebreak
+-->
+
+(ref:caption) The Northwest Atlantic coastal area where the NEUS black sea bass stock occurs.
+Red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations [after @milleretal16_yoy_survival] and the bathymetric contour is the 400 m isocline.
+
+```{r map, echo = FALSE, out.width="100%", fig.align="center", fig.cap = "(ref:caption)"}
+knitr::include_graphics(here::here("paper","map.pdf"))
+```
+\pagebreak
 

### 2025-03-04 paper/paper.tex
add edits, map fig

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 
-\title{Space for WHAM: a multi-region, multi-stock generalization of the
-Woods Hole Assessment Model with an application to black sea bass}
-\author{Timothy J. Miller\textsuperscript{1} \and Kierstin
-Curti \and Alexander C. Hansell}
-\date{28 February, 2025}
+\title{Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass}
+\author{Timothy J. Miller\textsuperscript{1} \and Kierstin Curti \and Alexander C. Hansell}
+\date{04 March, 2025}
 
@@
 
-\(^1\)\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}},
-Northeast Fisheries Science Center, National Marine Fisheries Service,
-166 Water Street, Woods Hole, MA 02543, USA\\
+\(^1\)\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}, Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\\
 
@@
 
-Describes the multi-region, multi-stock generalization of WHAM and its
-usage to evaluate evidence of alternative hypotheses about temperature
-effects on recruitment and age 1 natural mortality of black sea bass
-stock components.
+Describes the multi-region, multi-stock generalization of WHAM and its usage to evaluate evidence of alternative hypotheses about temperature effects on recruitment and age 1 natural mortality of black sea bass stock components.
 
@@
 
-The Woods Hole Assessment Model is a general state-space age-structured
-assessment model that is used to assess and manage many stocks in the
-Northeast US. We first describe an extension of the model allowing any
-number of stocks (or stock components) and and regions with movement
-among regions as well as seasonal variation in stock and fleet dynamics.
-Movement rates can be a be functions of time- and age-varying random
-effects and environmental covariates. We then illustrate the model by
-applying it to data for the northern and southern components of the
-Northeast US black sea bass stock and evaluate alternative hypotheses of
-bottom temperature and time-varying random effects on recruitment and
-natural mortality. We show strong evidence for temperature effects on
-recruitment, primarily for the northern stock component, and no evidence
-for including random effects or temperature effects on age 1 natural
-mortality.
+The Woods Hole Assessment Model is a general state-space age-structured assessment model that is used to assess and manage many stocks in the Northeast US.
+We first describe an extension of the model allowing any number of stocks (or stock components) and and regions with movement among regions as well as seasonal variation in stock and fleet dynamics.
+Movement rates can be a be functions of time- and age-varying random effects and environmental covariates.
+We then illustrate the model by applying it to data for the northern and southern components of the Northeast US black sea bass stock and evaluate alternative hypotheses of bottom temperature and time-varying random effects on recruitment and natural mortality.
+We show strong evidence for temperature effects on recruitment, primarily for the northern stock component, and no evidence for including random effects or temperature effects on age 1 natural mortality.
 
@@
 
-A state-space age-structured stock assessment models can be used to
-estiamte time and age-varying population attributes as random effects
-using maximum marginal likelihood or Bayesian fitting procedures
-(Nielsen and Berg 2014; Cadigan 2016; Miller et al. 2016b). This
-estimation approach is considered an essential feature of gold-standard
-assessment models that we use in tactical management of commercially
-important fish stocks (Punt et al. 2020). The State-space Assessment
-Model (SAM, Nielsen and Berg 2014) continues to be developed and remains
-widely used within the International Council for the Exploration of the
-Sea (ICES) to assess European fish stocks. Various state-space models
-are being used to manage cod and plaice stocks in the waters of Eastern
-Canada (Perreault et al. 2020; Varkey et al. 2022) and to the south, the
-Woods Hole Assessment Model (WHAM, Miller and Stock 2020; Stock and
-Miller 2021) is now used to assess many fish stocks in the Northwest
-Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).
-
-WHAM is an R package developed and maintained at NOAA's Northeast
-Fisheries Science Center {[}\url{https://timjmiller.github.io/wham}{]}.
-WHAM can be configured to fit a wide range of age-structured models from
-traditional statistical catch-at-age models without any random effects
-to models with several time and age varying process errors and effects
-of environmental covariates on various population parameters. Like SAM,
-WHAM models are built using the Template Model Builder package (TMB,
-Kristensen et al. 2016) which provides a computationally efficient means
-of fitting an extremely wide class of models with random effects. WHAM
-has undergone active development since its creation and includes random
-effects options for recruitment, inter-annual transitions of numbers at
-age (hereafter referred to as ``apparent survival''), fishery and index
-selectivity, natural mortality, and catchability.
-
-However, WHAM has only allowed one stock (component) and a single region
-and without seasonal changes in stock dynamics. Using such models for
-stocks that have subcomponents with varying seasonal movement can
-provide incorrect inferences and poor management advice (Ying et al.
-2011; Cao et al. 2014; Bosley et al. 2022). Furthermore, the ability to
-account for spatial structure and model multiple stocks are also
-important features of leading-edge assessment modeling frameworks (Punt
-et al. 2020). We describe here the implementation of these features and
-other extensions since Stock and Miller (2021) in WHAM version 2.0. Many
-of these new configuration options can also be useful when modeling a
-single stock and region.
-
-We developed this extension of WHAM in concert with a research track
-assessment process for black sea bass (\textit{Centropristis striata})
-(NEFSC 2023). Previous black sea bass assessments assumed two stock
-components (North and South), each assessed with a separate statistical
-catch at age assessment. The stock was split into two regions to account
-for spatial dynamics and improve model diagnostics (ASMFC 2016).
-Additionally, black sea bass have complex spatial dynamics and make
-seasonal migrations, moving inshore in the spring and offshore in the
-fall (Moser and Shepherd 2009). Black sea bass distribution has been
-linked to warming waters on the northwest shelf (Bell et al. 2015) and
-it is hypothesized that the distribution and productivity of the black
-sea bass stock is especially susceptible to long-term changes in
-temperature causing increases in abundance further north (Hare 2016).
-Thus, a focus of the research track assessment was to develop an
-assessment platform that could capture multiple stock components,
-complex seasonal movement and explore environmental drivers of
-population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we
-apply it to two stock components of black sea bass off the coast of the
-NEUS using most of the model assumptions that were accepted during the
-peer-review process. However, here we evaluate evidence for alternative
-hypotheses of temporal variation and effects specifically of bottom
-temperature on recruitment and natural mortality of age 1 individuals.
-
-When modeling multiple stock components occurring in different regions
-further dimensionality is added to the class of possible equilbrium
-reference points (e.g., Kapur et al. 2021). Short term projections are
-also an important part of the management process and the separation of
-observation and process errors in state-space models allows uncertainty
-in projected attributes such as recruitment and spawning stock biomass
-(SSB) without external simulation exercises. Finally, the inclusion of
-explicit effects of environmental covariates in the model also requires
-some assumptions on how they are treated in projections and reference
-points. We demonstrate some of options of WHAM version 2.0 to provide
-biological reference points and short-term projections.
+A state-space age-structured stock assessment models can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al.
+2016b).
+This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks (Punt et al.
+2020).
+The State-space Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed and remains widely used within the International Council for the Exploration of the Sea (ICES) to assess European fish stocks.
+Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada (Perreault et al.
+2020; Varkey et al.
+2022) and to the south, the Woods Hole Assessment Model (WHAM, Miller and Stock 2020; Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).
+
+WHAM is an R package developed and maintained at NOAA's Northeast Fisheries Science Center {[}\url{https://timjmiller.github.io/wham}{]}.
+WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and effects of environmental covariates on various population parameters.
+Like SAM, WHAM models are built using the Template Model Builder package (TMB, Kristensen et al.
+2016) which provides a computationally efficient means of fitting an extremely wide class of models with random effects.
+WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as ``apparent survival''), fishery and index selectivity, natural mortality, and catchability.
+
+However, WHAM has only allowed one stock (component) and a single region and without seasonal changes in stock dynamics.
+Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice (Ying et al.
+2011; Cao et al.
+2014; Bosley et al.
+2022).
+Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks (Punt et al.
+2020).
+We describe here the implementation of these features and other extensions since Stock and Miller (2021) in WHAM version 2.0.
+Many of these new configuration options can also be useful when modeling a single stock and region.
+
+We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) (NEFSC 2023).
+Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age assessment.
+The stock was split into two regions to account for spatial dynamics and improve model diagnostics (ASMFC 2016).
+Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (Moser and Shepherd 2009).
+Black sea bass distribution has been linked to warming waters on the northwest shelf (Bell et al.
+2015) and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare 2016).
+Thus, a focus of the research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the NEUS using most of the model assumptions that were accepted during the peer-review process.
+However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
+
+When modeling multiple stock components occurring in different regions further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al.
+2021).
+Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises.
+Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points.
+We demonstrate some of options of WHAM version 2.0 to provide biological reference points and short-term projections.
 
@@
 
-Many of the options and equations of WHAM version 2.0 are the same as
-those in Stock and Miller (2021), so we will only describe extensions
-and differences that have occurred since their first description of
-WHAM. The new version of WHAM can model multiple stocks each with their
-own movement, harvest, and natural mortality. Much of the description
-below is for a specific stock \(s\), but, for simplicity, this subscript
-is implicit except when necessary.
+Many of the options and equations of WHAM version 2.0 are the same as those in Stock and Miller (2021), so we will only describe extensions and differences that have occurred since their first description of WHAM.
+The new version of WHAM can model multiple stocks each with their own movement, harvest, and natural mortality.
+Much of the description below is for a specific stock \(s\), but, for simplicity, this subscript is implicit except when necessary.
 
 \hypertarget{the-probability-transition-matrix}{%
-\subsubsection*{The probability transition
-matrix}\label{the-probability-transition-matrix}}
+\subsubsection*{The probability transition matrix}\label{the-probability-transition-matrix}}
 \addcontentsline{toc}{subsubsection}{The probability transition matrix}
 
-Because individuals may be alive in one of several regions or harvested
-in one of several fleets, it is helpful to consider these as distinct
-categories or states and treat the number of individuals occurring in
-each category over time as a multi-state model. Approaches to modeling
-transitions among these states may treat time discretely (e.g., Arnason
-1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987;
-Commenges 1999; Andersen and Keiding 2002). We can define a probability
-transition matrix (PTM) that describes the probability of individuals
-occurring in different states at the end of a time interval \(\delta\)
-(columns), conditional on being in each of those state at beginning of
-the interval (rows). For fish populations, these states would be defined
-as being alive in a particular region or deceased due to fishing from a
-particular fleet or natural mortality. The time interval \(i\) with
-duration \(\delta_i\) would be a season and the PTMs would be uniquely
-defined for each stock by season \(i\), year \(y\), and age \(a\) on
-January 1. Each row and column of the PTM correspond to one the states:
-alive in region \(r\), dead in fleet \(f\), or dead from natural causes.
-The probabilities in each row sum to unity (must be in one of the
-states) and assume an individual is in the state that corresponds to
-that row at the beginning of the interval. Given \(n_R\) regions and
-\(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a
-function of sub-matrices is \begin{equation}\label{eq:ptm}
+Because individuals may be alive in one of several regions or harvested in one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model.
+Approaches to modeling transitions among these states may treat time discretely (e.g., Arnason 1972; Schwarz et al.
+1993) or continuously (e.g., Hearn et al.
+1987; Commenges 1999; Andersen and Keiding 2002).
+We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval \(\delta\) (columns), conditional on being in each of those state at beginning of the interval (rows).
+For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
+The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by season \(i\), year \(y\), and age \(a\) on January 1.
+Each row and column of the PTM correspond to one the states: alive in region \(r\), dead in fleet \(f\), or dead from natural causes.
+The probabilities in each row sum to unity (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval.
+Given \(n_R\) regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
+\begin{equation}\label{eq:ptm}
   \mathbf{P}_{y,a,i} = \begin{bmatrix}
@@
   \end{bmatrix}
-\end{equation} where \begin{equation*}
+\end{equation}
+where
+\begin{equation*}
   \mathbf{O}_{y,a,i} = 
@@
   \end{bmatrix}
-\end{equation*} is the \(n_R \times n_R\) matrix defining survival and
-occurring in each region at the end of the interval, \begin{equation*} 
+\end{equation*}
+is the \(n_R \times n_R\) matrix defining survival and occurring in each region at the end of the interval,
+\begin{equation*} 
   \mathbf{H}_{y,a,i} = 
@@
   \end{bmatrix}
-\end{equation*} is the \(n_R \times n_F\) matrix defining probabilities
-of being captured in each fleet during the interval, and
-\(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of
-dying due to natural mortality during the interval. We have the identity
-matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a
-1 for the natural mortality state because the probabilities of being in
-one of the mortality states given starting the interval in that state is
-unity (no zombies allowed).
-
-WHAM uses these PTMs to model abundance proportions in each state rather
-than true probabilities where numbers in each state would be multinomial
-distributed as in a model for tagging data where fates of individual
-tagged fish are assumed independent. The PTMs determine the expected
-numbers 1) in each state on January 1 of year \(t+1\) at age \(a+1\)
-given the abundances at age \(a\) on January 1 of year \(t\), 2)
-captured over the year in each fleet, 3) available to each index, and 4)
-alive at the time and in the region where spawning occurs.
+\end{equation*}
+is the \(n_R \times n_F\) matrix defining probabilities of being captured in each fleet during the interval, and \(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of dying due to natural mortality during the interval.
+We have the identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).
+
+WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would be multinomial distributed as in a model for tagging data where fates of individual tagged fish are assumed independent.
+The PTMs determine the expected numbers 1) in each state on January 1 of year \(t+1\) at age \(a+1\) given the abundances at age \(a\) on January 1 of year \(t\), 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.
 
@@
 
-When there is only one region, \begin{equation}\label{eq:ptm_1_region}
+When there is only one region,
+\begin{equation}\label{eq:ptm_1_region}
 \mathbf{P}_{y,a,i} = 
@@
   \end{bmatrix}
-\end{equation} where \(S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}\),
-\(\mathbf{H}_{y,a,i}\) is a 1 x \(n_F\) matrix with elements for each
-fleet \(f\):
-\(\frac{F_{y,a,i,f}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\),
-\(D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\),
-and \(Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{y,a,i,f}\) is the total
-mortality rate.
+\end{equation}
+where \(S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}\), \(\mathbf{H}_{y,a,i}\) is a 1 x \(n_F\) matrix with elements for each fleet \(f\): \(\frac{F_{y,a,i,f}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), \(D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), and \(Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{y,a,i,f}\) is the total mortality rate.
 
@@
 
-When there is more than one region, WHAM can model survival and movement
-as processes occurring sequentially or simultaneously. The sequential
-assumption is used widely in spatially explicit model (e.g., Stock
-Synthesis, Methot and Wetzel 2013). Under the sequential assumption,
-survival and death occur over the interval and movement among regions
-occurs instantly at either the beginning or the end of the interval.
+When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously.
+The sequential assumption is used widely in spatially explicit model (e.g., Stock Synthesis, Methot and Wetzel 2013).
+Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval.
 WHAM is configured to have movement occur after survival:
@@
   \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i}
-\end{equation*} where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\)
-diagonal matrix of proportions surviving in each region (given they
-start in that region) \begin{equation*}
+\end{equation*}
+where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\) diagonal matrix of proportions surviving in each region (given they start in that region)
+\begin{equation*}
 \mathbf{S}_{y,a,i} = 
@@
   \end{bmatrix}
-\end{equation*} and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\)
-matrix of probabilities of moving from one region to another or staying
-in the region they occurred at the beginning of the interval:
+\end{equation*}
+and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\) matrix of probabilities of moving from one region to another or staying in the region they occurred at the beginning of the interval:
 \begin{equation*}
@@
 
-WHAM assumes each fleet \(f\) can harvest in only one region (\(r_f\))
-during specified seasons. So, for each fleet \(f\), row \(r_f\) and
-column \(f\) of \(\mathbf{H}_{y,a,i}\) will be
-\(F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\)
-when fleet \(f\) is harvesting during the interval \(\delta_i\) and all
-other elements will be zero. Row \(r\) of the single-column matrix
-\(\mathbf{D}_{y,a,i}\) is
-\(M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\)
-
-When survival and movement are assumed to occur simultaneously, all
-movement and mortality parameters are instantaneous rates. We obtain the
-probability transition matrix over an interval \(\delta_i\) by
-exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
+WHAM assumes each fleet \(f\) can harvest in only one region (\(r_f\)) during specified seasons.
+So, for each fleet \(f\), row \(r_f\) and column \(f\) of \(\mathbf{H}_{y,a,i}\) will be \(F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\) when fleet \(f\) is harvesting during the interval \(\delta_i\) and all other elements will be zero.
+Row \(r\) of the single-column matrix \(\mathbf{D}_{y,a,i}\) is \(M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\)
+
+When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
+We obtain the probability transition matrix over an interval \(\delta_i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
 \begin{equation*}
 \mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
-\end{equation*} The instantaneous rate matrix takes rates of movement
-between regions and the mortality rates for each fleet and region. Along
-the diagonal is the negative of the sum of the other rates (the hazard)
-so each row sums to zero. For two regions and one fleet operating in
-each region: \begin{equation*}
+\end{equation*}
+The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region.
+Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero.
+For two regions and one fleet operating in each region:
+\begin{equation*}
  \mathbf{A}_{y,a,i} = \begin{bmatrix}
@@
  \end{bmatrix}.
-\end{equation*} where
-\(a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})\).
-When there is one region, \(n_f\) fleets, and \(\delta_i = 1\),
-exponentiating the instantaneous rate matrix results in the PTM defined
-in Eq. \ref{eq:ptm_1_region}.
+\end{equation*}
+where \(a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})\).
+When there is one region, \(n_f\) fleets, and \(\delta_i = 1\), exponentiating the instantaneous rate matrix results in the PTM defined in Eq.
+\ref{eq:ptm_1_region}.
 
@@
 
-Seasonality can be configured to accommodate characteristics of
-spawning, movement, and fleet-specific behavior. The annual time step
-can be divided into \(K\) seasons and the interval size \(\delta_i\) for
-each season \(i\) does not need to be equal to any other seasonal
-interval. Under the Markov assumption, the PTM of surviving and moving
-and dying over \(K\) intervals \(\delta_1,\ldots, \delta_K\) (i.e., the
-entire year) is just the product of the PTMs for each interval:
+Seasonality can be configured to accommodate characteristics of spawning, movement, and fleet-specific behavior.
+The annual time step can be divided into \(K\) seasons and the interval size \(\delta_i\) for each season \(i\) does not need to be equal to any other seasonal interval.
+Under the Markov assumption, the PTM of surviving and moving and dying over \(K\) intervals \(\delta_1,\ldots, \delta_K\) (i.e., the entire year) is just the product of the PTMs for each interval:
 \[ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).\]
 
-For a stock spawning at some fraction of the year \(0<t_s<1\) in
-interval \(\delta_j\), the fraction of time in season \(j\) is
+For a stock spawning at some fraction of the year \(0<t_s<1\) in interval \(\delta_j\), the fraction of time in season \(j\) is
 \[\delta_{s,j} = t_s-\sum^{j-1}_{i=0}\delta_i
-\] and the PTM defining the proportions in each state at time \(t_s\)
-for age \(a\) is \begin{equation}\label{eq:ptm_spawn}
+\]
+and the PTM defining the proportions in each state at time \(t_s\) for age \(a\) is
+\begin{equation}\label{eq:ptm_spawn}
 \mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{s,j}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{s,j}).
-\end{equation} Similarly, for an index \(m\) occurring at fraction of
-the year \(t_m\) in interval \(\delta_j\) the proportions in each state
-at the time of the observation is \begin{equation} \label{eq:ptm_index} 
+\end{equation}
+Similarly, for an index \(m\) occurring at fraction of the year \(t_m\) in interval \(\delta_j\) the proportions in each state at the time of the observation is
+\begin{equation} \label{eq:ptm_index} 
 \mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{m,j}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{m,j}).
@@
 
-When there are \(n_R\) regions and \(n_F\) fleets, the vector of
-abundance in each state at age \(a>1\) on January 1 is
-\(\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}', \mathbf{0}')'\) where
-\(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})'\) is the number
-in the states corresponding to being alive in each region and
-\(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each
-fleet and dead from natural mortality because no age \(a\) fish have
-died yet on January 1.
+When there are \(n_R\) regions and \(n_F\) fleets, the vector of abundance in each state at age \(a>1\) on January 1 is \(\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}', \mathbf{0}')'\) where \(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})'\) is the number in the states corresponding to being alive in each region and \(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each fleet and dead from natural mortality because no age \(a\) fish have died yet on January 1.
 
 Each stock \(s\) is assumed to spawn and recruit in one region \(r_s\).
-So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row
-\(r = r_s\). Options for configuring recruitment (\(N_{y,1,r_s}\)) for
-each stock are the same as previous versions of WHAM (Miller and Stock
-2020). If recruitment is assumed to be a function of SSB, it is only the
-spawning population in region \(r_s\) at the time of spawning that
-constitutes the SSB in the stock-recruit function. However, models can
-configure spawning individuals to occur in other regions at the time of
-spawning. Aside from treating recruitment as a random walk, the general
-model for annual recruitment as random effects is \begin{equation*}
+So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row \(r = r_s\).
+Options for configuring recruitment (\(N_{y,1,r_s}\)) for each stock are the same as previous versions of WHAM (Miller and Stock 2020).
+If recruitment is assumed to be a function of SSB, it is only the spawning population in region \(r_s\) at the time of spawning that constitutes the SSB in the stock-recruit function.
+However, models can configure spawning individuals to occur in other regions at the time of spawning.
+Aside from treating recruitment as a random walk, the general model for annual recruitment as random effects is
+\begin{equation*}
 \log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1,r_s} =  f\left(\text{SSB}_{y-1,r_s}\right) + \varepsilon_{y,1,r_s}
-\end{equation*} where\\
+\end{equation*}
+where\\
 \[\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}(t_s)' \mathbf{N}_{O,y,a}\]
-and \(w_{y,a}\) is the mean weight at age of spawning individuals,
-\(\text{mat}_{y,a}\) is the maturity at age, and
-\(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and
-occurring in region \(r_s\) at time \(t_s\) given being alive in each
-region at the start of the year (the \(r_s\) column of the upper-left
-submatrix of Eq. \ref{eq:ptm_spawn}).
-
-As in previous versions of WHAM, the transitions in numbers at age from
-one year to another after recruitment can be treated deterministically
-or as functions of random effects. The predicted numbers at age in year
-\(y\) at age \(a\) for a given stock are vector analogs
-(\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the
-standard WHAM model (Stock and Miller 2021). For ages
-\(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected
-number alive in each region at the beginning of the following year and
-next age class age can be obtained from the first \(n_R\) elements of
-the vector \[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\] The numbers
-alive in each region can also be modeled more simply using the
-sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions
-in abundance at age is \begin{equation*}
+and \(w_{y,a}\) is the mean weight at age of spawning individuals, \(\text{mat}_{y,a}\) is the maturity at age, and \(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and occurring in region \(r_s\) at time \(t_s\) given being alive in each region at the start of the year (the \(r_s\) column of the upper-left submatrix of Eq.
+\ref{eq:ptm_spawn}).
+
+As in previous versions of WHAM, the transitions in numbers at age from one year to another after recruitment can be treated deterministically or as functions of random effects.
+The predicted numbers at age in year \(y\) at age \(a\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (Stock and Miller 2021).
+For ages \(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first \(n_R\) elements of the vector
+\[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\]
+The numbers alive in each region can also be modeled more simply using the sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions in abundance at age is
+\begin{equation*}
 \log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
-\end{equation*} for ages \(a = 2,\ldots, A-1\), and for the plus group
+\end{equation*}
+for ages \(a = 2,\ldots, A-1\), and for the plus group
 \begin{equation*}
 \log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
-\end{equation*} where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector
-of region-specific errors for a given stock, which are independent
-across stocks and regions, but the same correlation structures as
-previous versions are possible across ages and years for a given stock
-and region. When there is autocorrelation with age, WHAM now assumes
-this applies only to ages \(a>1\) by default so that recruitment random
-effects are independent of those for for the annual transitions of older
-age classes. So the general covariance structure for a given stock at
-ages \(a>1\) in region \(r\) is that of a two-dimensional first-order
-autoregressive (2DAR1) process \begin{equation*}
+\end{equation*}
+where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
+When there is autocorrelation with age, WHAM now assumes this applies only to ages \(a>1\) by default so that recruitment random effects are independent of those for for the annual transitions of older age classes.
+So the general covariance structure for a given stock at ages \(a>1\) in region \(r\) is that of a two-dimensional first-order autoregressive (2DAR1) process
+\begin{equation*}
   Cov\left(\varepsilon_{y,a,r},\varepsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}
-\end{equation*} and that for age 1 is just AR1 across years
+\end{equation*}
+and that for age 1 is just AR1 across years
 \begin{equation*}
   Cov\left(\varepsilon_{y,1},\varepsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
-\end{equation*} When the transitions in abundance at age are treated
-deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). Since
-recruitment for a given stock currently only occurs in one region
-\(r_s\) there is only a single vector of time-varying recruitment random
-effects for each stock.
+\end{equation*}
+When the transitions in abundance at age are treated deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\).
+Since recruitment for a given stock currently only occurs in one region \(r_s\) there is only a single vector of time-varying recruitment random effects for each stock.
 
@@
 
-Initial numbers at age for each stock and region can be treated as
-age-specific fixed effects or with an equilibrium assumption as in
-previous versions of WHAM. For the equilibrium option there are two
-parameters for each stock: the stock-specific fully-selected fishing
-mortality rate \(\log \widetilde{F}\) and the recruitment in year 1
-\(\log N_{1,1,r_s}\). A stock-specific equilibrium fishing mortality at
-age by fleet \({\tilde F}_{a,f}\) is the product of \(\widetilde{F}\)
-and the selectivity across fleets in the first year \begin{equation*}
+Initial numbers at age for each stock and region can be treated as age-specific fixed effects or with an equilibrium assumption as in previous versions of WHAM.
+For the equilibrium option there are two parameters for each stock: the stock-specific fully-selected fishing mortality rate \(\log \widetilde{F}\) and the recruitment in year 1 \(\log N_{1,1,r_s}\).
+A stock-specific equilibrium fishing mortality at age by fleet \({\tilde F}_{a,f}\) is the product of \(\widetilde{F}\) and the selectivity across fleets in the first year
+\begin{equation*}
   \text{sel}_{1,a,f} = \frac{F_{1,a,f}}{\max_a \sum_{f=1}^{n_F} F_{1,a,f}}.
-\end{equation*} We use \(\widetilde{F}_{a,f}\) to define an equilibrium
-abundance per recruit by region at age \(a\) conditional on recruiting
-to each region \begin{equation}\label{eq:npraa}
+\end{equation*}
+We use \(\widetilde{F}_{a,f}\) to define an equilibrium abundance per recruit by region at age \(a\) conditional on recruiting to each region
+\begin{equation}\label{eq:npraa}
  \widetilde{\mathbf{O}}_{a} = \left\{
@@
 \right.
-\end{equation} where \(\mathbf{O}_{j}\) is the equilibrium probability
-of surviving age \(a\) and occurring in each region and
-\(\mathbf{O}_{0} = \mathbf{I}\). Natural mortality and movement rates in
-the first year of the model are also used in Eq. \ref{eq:npraa}. For the
-plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is
-a ``fundamental matrix'\,' derived using the matrix version of the
-geometric series (Kemeny and Snell 1960). Recall that recruitment for
-stock \(s\) only occur in region \(r_s\) so, the equilibrium initial
-numbers at age \(a\) by region are
+\end{equation}
+where \(\mathbf{O}_{j}\) is the equilibrium probability of surviving age \(a\) and occurring in each region and \(\mathbf{O}_{0} = \mathbf{I}\).
+Natural mortality and movement rates in the first year of the model are also used in Eq.
+\ref{eq:npraa}.
+For the plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is a ``fundamental matrix'\,' derived using the matrix version of the geometric series (Kemeny and Snell 1960).
+Recall that recruitment for stock \(s\) only occur in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
 \[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}' \mathbf{N}_{O,1,1}.\]
 
-The initial abundances at age can also be treated as independent or AR1
-random effects. Defining the vector of initial abundance at age in
-region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
+The initial abundances at age can also be treated as independent or AR1 random effects. Defining the vector of initial abundance at age in region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
 \[\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}\]
@@
 
-For each season, there are at most \(n_R-1\) parameters determining
-movement among regions given starting the season in region \(r\) in
-either the sequential or simultaneous configurations. Movement
-parameters are estimated on a transformed scale via a link function
-\(g(\cdot)\). If survival and movement occur simultaneously, the
-parameters are estimated with a log link function and if they are
-separable, an additive logit link function (like a multinomial
-regression) is used. On the transformed scale, the general model for the
-movement parameter from region \(r\) to \(r'\) in season \(i\) and year
-\(y\) for individuals of age \(a\) is a linear function of an intercept
-or mean parameter (\(\theta_{r\rightarrow r',i}\)) and both random and
-environmental effects: \begin{equation*}
+For each season, there are at most \(n_R-1\) parameters determining movement among regions given starting the season in region \(r\) in either the sequential or simultaneous configurations.
+Movement parameters are estimated on a transformed scale via a link function \(g(\cdot)\).
+If survival and movement occur simultaneously, the parameters are estimated with a log link function and if they are separable, an additive logit link function (like a multinomial regression) is used.
+On the transformed scale, the general model for the movement parameter from region \(r\) to \(r'\) in season \(i\) and year \(y\) for individuals of age \(a\) is a linear function of an intercept or mean parameter (\(\theta_{r\rightarrow r',i}\)) and both random and environmental effects:
+\begin{equation*}
   g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
-\end{equation*} The random effects
-\(\varepsilon_{r\rightarrow r',y,a,i}\) are season-, and
-region-to-region-specific and modeled most generally as 2DAR1 random
-effects with age and(or) year where the covariance is \begin{equation*}
+\end{equation*}
+The random effects \(\varepsilon_{r\rightarrow r',y,a,i}\) are season-, and region-to-region-specific and modeled most generally as 2DAR1 random effects with age and(or) year where the covariance is
+\begin{equation*}
   Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
-\end{equation*} similar to how WHAM models variation in survival,
-natural mortality, and selectivity. Effects of covariate \(E_k\) can be
-age-, season-, and region-to-region-specific
-\(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial
-options in the previous versions of WHAM for effects on recruitment and
-natural mortality are available.
-
-There is currently no likelihood component for tagging data. Therefore,
-movement parameters would usually either need to be fixed or assumed to
-have some prior distribution, possibly based on external parameter
-estimates. WHAM includes prior distributions for the season and
-region-to-region specific (mean) movement parameters and treats them as
-random effects with the mean defined by the initial value of the fixed
-effect counterpart and standard deviation \begin{equation*}
+\end{equation*}
+similar to how WHAM models variation in survival, natural mortality, and selectivity.
+Effects of covariate \(E_k\) can be age-, season-, and region-to-region-specific \(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.
+
+There is currently no likelihood component for tagging data.
+Therefore, movement parameters would usually either need to be fixed or assumed to have some prior distribution, possibly based on external parameter estimates.
+WHAM includes prior distributions for the season and region-to-region specific (mean) movement parameters and treats them as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
+\begin{equation*}
   \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
-  \end{equation*} When priors are used, the movement is defined instead
-as \begin{equation*}
+  \end{equation*}
+When priors are used, the movement is defined instead as
+\begin{equation*}
   g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
@@
 
-When treated as constant parameters, natural mortality rates may be
-stock-, region-, and age-specific. When random effects are used, the
-same 2DAR1 structure with age and year as described by Stock and Miller
-(2021) can be configured for a given stock and region and any
-environmental covariate effects can be stock-, region-, and
-age-specific. So the general model for natural mortality is
+When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific.
+When random effects are used, the same 2DAR1 structure with age and year as described by Stock and Miller (2021) can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
+So the general model for natural mortality is
 \begin{equation*}
   \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
-\end{equation*} The general covariance structure for random effects are
-modeled most generally as 2DAR1 random effect with age and(or) year
-where the covariance is \begin{equation*}
+\end{equation*}
+The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
+\begin{equation*}
   Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
@@
 
-The log-normal distributional assumption for aggregate catch
-observations is the same as Stock and Miller (2021), but the predicted
-catch is now a function of catch from each stock starting the year in
-each region. For a given stock and age, the numbers captured in each
-fleet over the year are
+The log-normal distributional assumption for aggregate catch observations is the same as Stock and Miller (2021), but the predicted catch is now a function of catch from each stock starting the year in each region.
+For a given stock and age, the numbers captured in each fleet over the year are
 \[\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}' \mathbf{N}_{O,s,y,a}\]
@@
 \[\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \widehat{\mathbf{N}}_{H,y,a}\]
-where \(\mathbf{c}_{y,a}\) is the vector of mean individual weight at
-age \(a\) for each fleet and the aggregate catch by fleet is
+where \(\mathbf{c}_{y,a}\) is the vector of mean individual weight at age \(a\) for each fleet and the aggregate catch by fleet is
 \[\widehat{\mathbf{C}}_y = \sum^{A}_{a=1} \widehat{\mathbf{C}}_{y,a}\]
-The log-aggregate catch observations for fleet \(f\) are normally
-distributed
+The log-aggregate catch observations for fleet \(f\) are normally distributed
 \[ \log C_{y,f} \sim \text{N}\left(\log \widehat {C}_{y,f}, \sigma^2_{y,f}\right).\]
 
-The predicted numbers caught for each fleet \(f\) (row \(f\) of
-\(\widehat{\mathbf{N}}_{H,y,a}\)) are used to make predicted age
-composition observations as described by Stock and Miller (2021). Since
-then, three additional likelihood options for age composition
-observations have been added: a logistic-normal with AR(1) correlation
-structure (Francis 2014), the alternative Dirichlet-multinomial
-parameterization described by Thorson et al. (2017), and the
-multivariate Tweedie (Thorson et al. 2023).
+The predicted numbers caught for each fleet \(f\) (row \(f\) of \(\widehat{\mathbf{N}}_{H,y,a}\)) are used to make predicted age composition observations as described by Stock and Miller (2021).
+Since then, three additional likelihood options for age composition observations have been added: a logistic-normal with AR(1) correlation structure (Francis 2014), the alternative Dirichlet-multinomial parameterization described by Thorson et al.
+(2017), and the multivariate Tweedie (Thorson et al.
+2023).
 
@@
 
-For index \(m\) occurring in region \(r_m\) at fraction of the year
-\(t_m\), the predicted abundance at \(t_m\) in region \(r_m\) is
+For index \(m\) occurring in region \(r_m\) at fraction of the year \(t_m\), the predicted abundance at \(t_m\) in region \(r_m\) is
 \[\widehat{N}_{s,y,a,m} = \mathbf{O}_{s,y,a,r_m}(t_m)' \mathbf{N}_{O,s,y,a}\]
-where \(\mathbf{O}_{s,y,a,r_m}(t_m)\), the \(r_m\) column of the
-upper-left submatrix of Eq. \ref{eq:ptm_index}, are the probabilities of
-surviving and occurring in region \(r_m\) at time \(t_m\) given being
-alive in each region at the start of the year. The predicted index at
-age is
+where \(\mathbf{O}_{s,y,a,r_m}(t_m)\), the \(r_m\) column of the upper-left submatrix of Eq.
+\ref{eq:ptm_index}, are the probabilities of surviving and occurring in region \(r_m\) at time \(t_m\) given being alive in each region at the start of the year.
+The predicted index at age is
 \[\widehat{I}_{m,y,a} = q_{m,y} \text{sel}_{m,y,a}w_{m,y,a}\sum^{n_S}_{s = 1}\widehat{N}_{s,y,a,m}\]
-where \(q_{m,y}\) is the catchability of the index in year \(y\),
-\(\text{sel}_{m,y,a}\) is the selectivity and \(w_{m,y,a}\) is the
-average weight of individuals at age \(a\) if the index is quantified in
-biomass and \(w_{m,y,a} = 1\) if the index is quantified in numbers.
-Predicted age composition observations are functions of
-\(\widehat{I}_{m,y,a}\) as described by Stock and Miller (2021) and the
-likelihood options are the same as those for catch explained above.
-
-Catchability of the index can also be treated as logit functions of
-normal random effects and(or) environmental covariate effects
+where \(q_{m,y}\) is the catchability of the index in year \(y\), \(\text{sel}_{m,y,a}\) is the selectivity and \(w_{m,y,a}\) is the average weight of individuals at age \(a\) if the index is quantified in biomass and \(w_{m,y,a} = 1\) if the index is quantified in numbers.
+Predicted age composition observations are functions of \(\widehat{I}_{m,y,a}\) as described by Stock and Miller (2021) and the likelihood options are the same as those for catch explained above.
+
+Catchability of the index can also be treated as logit functions of normal random effects and(or) environmental covariate effects
 \[log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}\]
-where \(\theta_{q,m}\) is an intercept or mean parameter, and \(u_{m}\)
-and \(l_{m}\) are the upper and lower bounds of catchability for index
-\(m\) (defaults are 0 and 1000). The general covariance structure for
-the annual random effects is AR1
+where \(\theta_{q,m}\) is an intercept or mean parameter, and \(u_{m}\) and \(l_{m}\) are the upper and lower bounds of catchability for index \(m\) (defaults are 0 and 1000).
+The general covariance structure for the annual random effects is AR1
 \[Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.\]
@@
 
-Currently a single fishing mortality reference point \(\widetilde F\) is
-estimated across stocks and regions and the fleet- and age-specific
-reference points is
-\(\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}\). Selectivity is
-determined across fleets and ages as in the equilibrium assumption for
-initial abundance at age except here it is a function of average F at
-age over a user-defined set of years \begin{equation}\label{eq:spr_sel}
+Currently a single fishing mortality reference point \(\widetilde F\) is estimated across stocks and regions and the fleet- and age-specific reference points is \(\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}\).
+Selectivity is determined across fleets and ages as in the equilibrium assumption for initial abundance at age except here it is a function of average F at age over a user-defined set of years
+\begin{equation}\label{eq:spr_sel}
   \text{sel}_{a,f} = \frac{\overline F_{a,f}}{\max_a \sum^{n_F}_{f=1}{\overline F}_{a,f}}
@@
 
-The equilibrium spawning stock biomass per recruit for a given stock in
-region \(r_s\) is defined as \begin{equation}\label{eq:ssbpr}
+The equilibrium spawning stock biomass per recruit for a given stock in region \(r_s\) is defined as
+\begin{equation}\label{eq:ssbpr}
  \upphi(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{O}_{a,\cdot,r_s}(t_s) w_{a} m_{a}
-\end{equation} where \(w_{a}\) and \(m_{a}\) are the mean individual
-weight and probability of maturity at age \(a\),
-\(\widetilde{\mathbf{O}}_{a}\) are as described in Eq. \ref{eq:npraa},
-and \(\mathbf{O}_{a}(t_s)\) is the \(n_R \times n_R\) upper-left
-sub-matrix of eq. \ref{eq:ptm_spawn} with the probabilities of surviving
-and occurring in each region \(r'\) at age \(a+t_s\) given starting in
-region \(r\) at age \(a\). The further subscripts \(r_s,\cdot\) and
-\(\cdot,r_s\) indicate row or column \(r_s\), respectively. Using these
-rows and columns is required because of the assumption that spawning and
-recruitment only occur in region \(r_s\).
-
-The equilibrium spawning biomass per recruit (eq. \ref{eq:ssbpr}) is
-conditional on the region of recruitment \(r_s\). The equilibrium
-recruitment depends on the stock dynamics. WHAM currently only allows
-complete spawning region fidelity so that a stock only spawns and
-recruits in a single region. Therefore, the equilibrium vector of
-numbers at age 1 for each state, \(\widetilde {\mathbf{N}}_{1}\), will
-only be nonzero in the row corresponding to the spawning region
-(\(r_s\)). The matrices of probabilities of surviving and occurring in
-each region, \(\widetilde{\mathbf{O}}_{s,a}\) and
-\(\mathbf{O}_{s,a}(t_s)\), are functions of the fishing mortality rates
-for fleets in each region \(\widetilde F_{f,a}\).
-
-The matrix of equilibrium yield per recruit by region and fleet as a
-function of \(\widetilde F\) is calculated as
+\end{equation}
+where \(w_{a}\) and \(m_{a}\) are the mean individual weight and probability of maturity at age \(a\), \(\widetilde{\mathbf{O}}_{a}\) are as described in Eq.
+\ref{eq:npraa}, and \(\mathbf{O}_{a}(t_s)\) is the \(n_R \times n_R\) upper-left sub-matrix of eq.
+\ref{eq:ptm_spawn} with the probabilities of surviving and occurring in each region \(r'\) at age \(a+t_s\) given starting in region \(r\) at age \(a\).
+The further subscripts \(r_s,\cdot\) and \(\cdot,r_s\) indicate row or column \(r_s\), respectively.
+Using these rows and columns is required because of the assumption that spawning and recruitment only occur in region \(r_s\).
+
+The equilibrium spawning biomass per recruit (eq.
+\ref{eq:ssbpr}) is conditional on the region of recruitment \(r_s\).
+The equilibrium recruitment depends on the stock dynamics.
+WHAM currently only allows complete spawning region fidelity so that a stock only spawns and recruits in a single region.
+Therefore, the equilibrium vector of numbers at age 1 for each state, \(\widetilde {\mathbf{N}}_{1}\), will only be nonzero in the row corresponding to the spawning region (\(r_s\)).
+The matrices of probabilities of surviving and occurring in each region, \(\widetilde{\mathbf{O}}_{s,a}\) and \(\mathbf{O}_{s,a}(t_s)\), are functions of the fishing mortality rates for fleets in each region \(\widetilde F_{f,a}\).
+
+The matrix of equilibrium yield per recruit by region and fleet as a function of \(\widetilde F\) is calculated as
 \begin{equation}\label{eq:ypr}
  \widetilde{\nu}({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{H}_{a} \mathbf{c}_{s,a}
-\end{equation} where \(\mathbf{c}_a\) is the vector of mean individual
-weight at age for each fleet, and \(\mathbf{H}_{s,a}\) is the submatrix
-of the probabilities of being captured in each fleet over the interval
-from \(a\) to \(a+1\), defined in eq. \ref{eq:ptm}.
-
-As in previous versions of WHAM package, reference points for prevailing
-conditions, average all of the inputs to the spawning biomass and yield
-per recruit calculations (weight, maturity, natural mortality, and
-movement) over a set of user-specified years as shown in eq.
+\end{equation}
+where \(\mathbf{c}_a\) is the vector of mean individual weight at age for each fleet, and \(\mathbf{H}_{s,a}\) is the submatrix of the probabilities of being captured in each fleet over the interval from \(a\) to \(a+1\), defined in eq.
+\ref{eq:ptm}.
+
+As in previous versions of WHAM package, reference points for prevailing conditions, average all of the inputs to the spawning biomass and yield per recruit calculations (weight, maturity, natural mortality, and movement) over a set of user-specified years as shown in eq.
 \ref{eq:spr_sel} for selectivity (e.g., last 5 years of the model).
 
-For \(X\%\) SPR-based reference points, we use a Newton method and
-iterate \begin{equation}\label{eq:newton}
+For \(X\%\) SPR-based reference points, we use a Newton method and iterate
+\begin{equation}\label{eq:newton}
   \log\widetilde{F}^{(i)} = \log\widetilde{F}^{(i-1)} - \frac{g\left(\log\widetilde{F}^{(i-1)}\right)}{g'\left(\log\widetilde{F}^{(i-1)}\right)}
-\end{equation} where \(g(\log F)\) is the difference between the
-weighted sums of spawning biomass per recruit at \(F\) and \(X\)\% of
-unfished spawning biomass per recruit across stocks:
+\end{equation}
+where \(g(\log F)\) is the difference between the weighted sums of spawning biomass per recruit at \(F\) and \(X\)\% of unfished spawning biomass per recruit across stocks:
 \begin{equation}\label{eq:newton-obj}
   g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
-\end{equation} where \(\upphi_s\left(F=0\right)\) is the equilibrium
-unfished spawning biomass per recruit. \(g'(\log F)\) is the derivative
-of \(g\) with respect to \(\log F\), and the weights to use for each
-stock \(\lambda_s\) can be specified by the user or relative to the
-average of recruitment for each stock over the same years the user
-defines to calculate ``static'\,' equilibrium spawning biomass and
-yield.
-
-When a Beverton-Holt or Ricker stock recruit relationship is assumed, an
-analogous Newton method is used to find \(\log F\) that maximizes yield
-for MSY-based reference points. which are also a functions of the
-equilibrium yield per recruit (Eq. \ref{eq:ypr}) and equilibrium
-recruitment. The function \(g(\log F)\) in Eq. \ref{eq:newton} is the
-first derivative of the yield curve with respect to \(\log F\).
+\end{equation}
+where \(\upphi_s\left(F=0\right)\) is the equilibrium unfished spawning biomass per recruit.
+\(g'(\log F)\) is the derivative of \(g\) with respect to \(\log F\), and the weights to use for each stock \(\lambda_s\) can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate ``static'\,' equilibrium spawning biomass and yield.
+
+When a Beverton-Holt or Ricker stock recruit relationship is assumed, an analogous Newton method is used to find \(\log F\) that maximizes yield for MSY-based reference points.
+which are also a functions of the equilibrium yield per recruit (Eq.
+\ref{eq:ypr}) and equilibrium recruitment.
+The function \(g(\log F)\) in Eq.
+\ref{eq:newton} is the first derivative of the yield curve with respect to \(\log F\).
 
@@
 
-The projection options are generally the same as those for previous
-versions of WHAM (Miller and Stock 2020). When there is movement of any
-stocks, the user has the option to project and use any random effects
-for time-varying movement or use the average over user specified years,
-analogous to how natural mortality can be treated in the projection
-period. The projection of any environmental covariates has been revised
-to better include error in the estimated latent covariate in any effects
-on the population in projection years. Users can also specify a catch or
-fully-selected fishing mortality in each projection year and they can be
-fleet-specific.
+The projection options are generally the same as those for previous versions of WHAM (Miller and Stock 2020).
+When there is movement of any stocks, the user has the option to project and use any random effects for time-varying movement or use the average over user specified years, analogous to how natural mortality can be treated in the projection period.
+The projection of any environmental covariates has been revised to better include error in the estimated latent covariate in any effects on the population in projection years.
+Users can also specify a catch or fully-selected fishing mortality in each projection year and they can be fleet-specific.
 
 \hypertarget{application-to-black-sea-bass}{%
-\section*{Application to black sea
-bass}\label{application-to-black-sea-bass}}
+\section*{Application to black sea bass}\label{application-to-black-sea-bass}}
 \addcontentsline{toc}{section}{Application to black sea bass}
 
-Black sea bass are a temperate reef fish in the western Atlantic Ocean
-ranging from the Gulf of Mexico to the Gulf of Maine. Fish north of Cape
-Hatteras, NC, are considered to comprise a single stock unit, but
-individual populations exhibit spawning site fidelity (Able and Stanton
-Hales, Jr. 1997; Fabrizio et al. 2013). Fish in this northern stock unit
-perform seasonal migrations out on the continental shelf in the fall and
-back to their inshore spawning areas during the spring (Musick and
-Mercer 1977). Analyses of tagging studies have demonstrated that the
-extent of seasonal migration varies along the coast such that fish from
-populations off of southern New England and further north move offshore
-and as far south as the coasts of Virgina and North Carolina, whereas
-fish in the southern portion of the stock range generally move shorter
-distances between inshore and offshore areas of similar latitude (Moser
-and Shepherd 2009).
-
-Prior to its 2023 peer-reviewed assessment, the NEUS black sea bass
-stock was assessed using the Age-Structured Assessment Program (ASAP)
-model (Legault and Restrepo 1999), a single-stock and -region
-statistical catch-at-age model that estimates all model parameters as
-fixed effects. Northern and southern components of the NEUS black sea
-bass stock ascribed to regions divided by the Hudson Canyon were
-separately modeled in ASAP (reference map figure). Results from the
-separate ASAP models were combined for a unit-stock assessment. The
-ASAP-based assessments exhibited strong retrospective patterns (Mohn
-1999), and exploring alternative modeling approaches for the northern
-and southern stock components has been a high priority for management.
-
-Leading up to the 2023 peer-reviewed assessment, a working group
-(hereafter referred to as ``Working Group'') composed of scientists from
-federal, state, and academic institutions determined an optimal data and
-model configuration for the black sea bass stock using the multi-stock
-and multi-region extension of WHAM described above (NEFSC 2023). This
-assessment included the spatial features and investigated inclusion of
-hypothesized environmental drivers that were prioritized research
-recommendations from previous black sea bass assessments. Below we
-describe the assumptions and configuration of the assessment model as
-determined by the Working Group as well as the alternative assumptions
-for recruitment and natural mortality in the models we fit to evaluate
-alternative hypotheses of bottom temperature effects on black sea bass.
+Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging from the Gulf of Mexico to the Gulf of Maine.
+Fish north of Cape Hatteras, NC, are considered to comprise a single stock unit, but individual populations exhibit spawning site fidelity
+(Able and Stanton Hales, Jr.
+1997; Fabrizio et al.
+2013).
+Fish in this northern stock unit perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring (Musick and Mercer 1977).
+Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude (Moser and Shepherd 2009).
+
+Prior to its 2023 peer-reviewed assessment, the NEUS black sea bass stock was assessed using the Age-Structured Assessment Program (ASAP) model (Legault and Restrepo 1999), a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
+Northern and southern components of the NEUS black sea bass stock ascribed to regions divided by the Hudson Canyon were separately modeled in ASAP (Figure \ref{fig:map}).
+Results from the separate ASAP models were combined for a unit-stock assessment.
+The ASAP-based assessments exhibited strong retrospective patterns (Mohn 1999), and exploring alternative modeling approaches for the northern and southern stock components has been a high priority for management.
+
+Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as ``Working Group'') composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above (NEFSC 2023).
+This assessment included the spatial features and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
+Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.
 
@@
 
-The first year being modeled for the population is 1989 and the fishery
-and index data used in the model span from 1989 to 2021. The north and
-south stock components are modeled as separate populations that spawn
-and recruit in respective regions. We have observations for each of four
-total fishing fleets, where two fishing fleets (recreational and
-commercial) operate in each region.
-
-There are 11 ``seasonal'' intervals within each calendar year: five
-monthly time intervals from Jan 1 to May 31, a spawning season from June
-1 to July 31, and five monthly intervals from August 1 to December 31.
-The southern stock component is assumed to never move to the northern
-region. For the northern component, a proportion
-\(\mu_{\text{N}\rightarrow \text{S}}\) can move to the southern region
-each month during the last five months of the year, but no movement is
-allows from the south to the north during this period (Figure
-\ref{fig:migration-diagram}). During the first four intervals of the
-year a proportion \(\mu_{\text{S}\rightarrow \text{N}}\) the northern
-component individuals in the south can move back to the north, but no
-movement from the north to south is allowed during this period. In the
-fifth interval (May), all northern component individuals remaining in
-the south are assumed to move back to the north for the subsequent
-spawning period. Survival and movement occur sequentially in each
-interval and each of the two movement proportions are assumed constant
-across intervals, ages, and years.
-
-The two monthly movement matrices are \begin{equation*}
+The first year being modeled for the population is 1989 and the fishery and index data used in the model span from 1989 to 2021.
+The north and south stock components are modeled as separate populations that spawn and recruit in respective regions.
+We have observations for each of four total fishing fleets, where two fishing fleets (recreational and commercial) operate in each region.
+
+There are 11 ``seasonal'' intervals within each calendar year: five monthly time intervals from Jan 1 to May 31, a spawning season from June 1 to July 31, and five monthly intervals from August 1 to December 31.
+The southern stock component is assumed to never move to the northern region.
+For the northern component, a proportion \(\mu_{\text{N}\rightarrow \text{S}}\) can move to the southern region each month during the last five months of the year, but no movement is allows from the south to the north during this period (Figure \ref{fig:migration-diagram}).
+During the first four intervals of the year a proportion \(\mu_{\text{S}\rightarrow \text{N}}\) the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
+In the fifth interval (May), all northern component individuals remaining in the south are assumed to move back to the north for the subsequent spawning period.
+Survival and movement occur sequentially in each interval and each of the two movement proportions are assumed constant across intervals, ages, and years.
+
+The two monthly movement matrices are
+\begin{equation*}
 \boldsymbol{\mu}_{1} = 
@@
   \end{bmatrix}
-\end{equation*} for the portion of the year after spawning and
+\end{equation*}
+for the portion of the year after spawning and
 \begin{equation*}
@@
   \end{bmatrix}
-\end{equation*} for the portion of the year before spawning. As noted in
-the description of the general WHAM model, tagging data are not yet
-allowed. However, the Working Group also fit a Stock Synthesis model
-(Methot and Wetzel 2013) which included tagging data and provided
-estimated movement rate parameters and standard errors that were used to
-configure the priors for WHAM (see Supplementary Materials).
-
-With movement as configured, the northern origin fish (ages 2+) can
-occur in the southern region on January 1. Estimating initial numbers at
-age as separate parameters can be challenging even in single-stock
-models, but for black sea bass the available data cannot distinguish the
-proportion of northern and southern component fish at each age in the
-southern region in the initial year of the model. Therefore, we used the
-simplifying equilibrium assumption where there are two parameters
-estimated for each stock component: an initial recruitment and an
-equilibrium fully-selected \(F\) that determines the abundance at age in
-each region for each stock component.
-
-For subsequent years, all models allow AR1 correlation of recruitment
-and 2DAR1 correlation for survival for both the northern and southern
-stock components. Models with apparent survival random effects will
-model the transitions (conditional on both survival and movement) of
-abundances at age of northern origin fish in each region, but there
-appears to be little ability of the existing data from the southern
-region to distinguish apparent survival random effects for each stock
-component. All models we considered assume very small variance for the
-transitions of northern fish in the southern region, which is
-essentially the same as the deterministic transition assumptions of a
-statistical catch at age model. Unique variance and correlation
-parameters for the recruitment and apparent survival random effects are
-estimated for the northern and southern components.
+\end{equation*}
+for the portion of the year before spawning.
+As noted in the description of the general WHAM model, tagging data are not yet allowed.
+However, the Working Group also fit a Stock Synthesis model (Methot and Wetzel 2013) which included tagging data and provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).
+
+With movement as configured, the northern origin fish (ages 2+) can occur in the southern region on January 1.
+Estimating initial numbers at age as separate parameters can be challenging even in single-stock models, but for black sea bass the available data cannot distinguish the proportion of northern and southern component fish at each age in the southern region in the initial year of the model.
+Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component: an initial recruitment and an equilibrium fully-selected \(F\) that determines the abundance at age in each region for each stock component.
+
+For subsequent years, all models allow AR1 correlation of recruitment and 2DAR1 correlation for survival for both the northern and southern stock components.
+Models with apparent survival random effects will model the transitions (conditional on both survival and movement) of abundances at age of northern origin fish in each region, but there appears to be little ability of the existing data from the southern region to distinguish apparent survival random effects for each stock component.
+All models we considered assume very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model.
+Unique variance and correlation parameters for the recruitment and apparent survival random effects are estimated for the northern and southern components.
 
 \hypertarget{uncertainty-in-recreational-index-observations}{%
-\subsection*{Uncertainty in recreational index
-observations}\label{uncertainty-in-recreational-index-observations}}
-\addcontentsline{toc}{subsection}{Uncertainty in recreational index
-observations}
-
-The estimated coefficients of variation (CVs) provided by the analyses
-by the Working Group to generate the recreational catch per angler
-(recreational CPA) index ranged between 0.02 and 0.06 which the Working
-Group felt did not capture the true observation uncertainty in the index
-with regard to its relationship to stock abundance. All models we
-considered allowed a scalar multiple of the standard deviation of the
-log aggregate index to be estimated for these indices in the northern
-and southern regions. However, lack of model convergence was observed in
-many fits for self-test simulations. The model estimates are negligibly
-affected by estimating these scalars, but estimating rather than fixing
-these parameters allowed uncertainty in model output to be more properly
-conveyed.
+\subsection*{Uncertainty in recreational index observations}\label{uncertainty-in-recreational-index-observations}}
+\addcontentsline{toc}{subsection}{Uncertainty in recreational index observations}
+
+The estimated coefficients of variation (CVs) provided by the analyses by the Working Group to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the Working Group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
+All models we considered allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
+However, lack of model convergence was observed in many fits for self-test simulations.
+The model estimates are negligibly affected by estimating these scalars, but estimating rather than fixing these parameters allowed uncertainty in model output to be more properly conveyed.
 
 \hypertarget{index-and-catch-age-composition-observations}{%
-\subsection*{Index and catch age composition
-observations}\label{index-and-catch-age-composition-observations}}
-\addcontentsline{toc}{subsection}{Index and catch age composition
-observations}
-
-The Working Group investigated many alternative assumptions for the
-probability models and selectivity forms for the eight different sets of
-age composition observations to reduce residual patterns and
-retrospective patterns. These analyses resulted in use of selectivity
-random effects for the northern fleet and indices and logistic-normal
-likelihoods for six sets of of age composition observations and
-Dirichlet-multinomial likelihoods for one index and one fleet in the
-northern region (Table \ref{tab:age-comp-sel-table}).
+\subsection*{Index and catch age composition observations}\label{index-and-catch-age-composition-observations}}
+\addcontentsline{toc}{subsection}{Index and catch age composition observations}
+
+The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual patterns and retrospective patterns.
+These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for six sets of of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
 
 \hypertarget{bottom-temperature-effects}{%
-\subsection*{Bottom temperature
-effects}\label{bottom-temperature-effects}}
+\subsection*{Bottom temperature effects}\label{bottom-temperature-effects}}
 \addcontentsline{toc}{subsection}{Bottom temperature effects}
 
-All models include bottom temperature observations for the northern and
-southern regions from 1963 to 2021 and standard errors of observations
-ranged between 0.03 and 0.09 degrees Celsius (NEFSC 2023). We retained
-the assumption from the peer-reviewed assessment that treated the latent
-bottom temperature covariates in each region as AR1 processes.
-
-We fit 14 models with alternative assumptions about the effects of
-bottom temperature covariates, ranging from no effects to effects on
-both regions for either recruitment or natural mortality at age 1 (Table
-\ref{tab:model-desc-table}). These analyses derive from the hypothesis
-that bottom temperature affects overwinter survival of fish where the
-fish turn from age 0 to age 1 on January 1 (Miller et al. 2016a). This
-covariate may be a proxy for temperature prior to January 1 and affect
-survival during the end of the pre-recruit phase or natural mortality in
-the early part of of the year after becoming age 1. Furthermore, we have
-no direct observations of age 1 individuals from surveys until the
-spring season each year. Therefore, we fit models with effects of
-temperature on recruitment or natural mortality at age 1.
-
-All models treat annual recruitment as AR1 time-varying random effects
-with mean parameters for each stock component. However, Miller et al.
-(2018) showed how inferences of temperature effects on growth or
-maturity parameters can be very different whether the compared models
-with and without the effect also include random effects representing
-residual temporal variation in parameters. Therefore, we also explored
-whether including temporal random effects on age 1 natural mortality
-affected inferences on corresponding temperature effects. Initially, we
-included random effects on age 1 natural mortality for both the northern
-and southern stock components, but estimates of these random effects and
-corresponding variance for the southern component converged to 0 so
-these were not included in models presented here.
-
-We assume the covariate in year \(y\) affects recruitment in the same
-years because the covariate observations are from months January to
-March. The fish are technically already 1 year old, but there are no
-observations of these individuals until later in the year except
-possibly in fishery catches which are accumulated over the whole year.
-Expected log-recruitment for a given stock is a linear function of
-bottom temperature
+All models include bottom temperature observations for the northern and southern regions from 1963 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius (NEFSC 2023).
+We retained the assumption from the peer-reviewed assessment that treated the latent bottom temperature covariates in each region as AR1 processes.
+
+We fit 14 models with alternative assumptions about the effects of bottom temperature covariates, ranging from no effects to effects on both regions for either recruitment or natural mortality at age 1 (Table \ref{tab:model-desc-table}).
+These analyses derive from the hypothesis that bottom temperature affects overwinter survival of fish where the fish turn from age 0 to age 1 on January 1 (Miller et al.
+2016a).
+This covariate may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of of the year after becoming age 1.
+Furthermore, we have no direct observations of age 1 individuals from surveys until the spring season each year.
+Therefore, we fit models with effects of temperature on recruitment or natural mortality at age 1.
+
+All models treat annual recruitment as AR1 time-varying random effects with mean parameters for each stock component.
+However, Miller et al.
+(2018) showed how inferences of temperature effects on growth or maturity parameters can be very different whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
+Therefore, we also explored whether including temporal random effects on age 1 natural mortality affected inferences on corresponding temperature effects.
+Initially, we included random effects on age 1 natural mortality for both the northern and southern stock components, but estimates of these random effects and corresponding variance for the southern component converged to 0 so these were not included in models presented here.
+
+We assume the covariate in year \(y\) affects recruitment in the same years because the covariate observations are from months January to March.
+The fish are technically already 1 year old, but there are no observations of these individuals until later in the year except possibly in fishery catches which are accumulated over the whole year.
+Expected log-recruitment for a given stock is a linear function of bottom temperature
 
@@
 E\left(\log N_{y,1}|x_y\right) = \mu_{R} + \beta_{R} x_y.
-\end{equation} Similarly, expected log-natural morality as a function of
-bottom temperature is \begin{equation}\label{eq:expected-M1}
+\end{equation}
+Similarly, expected log-natural morality as a function of bottom temperature is
+\begin{equation}\label{eq:expected-M1}
 E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y
-\end{equation} Because age 1 fish for the northern component can exist
-in both regions after January 1, natural mortality is acting in both
-regions for this stock. For models with covariate and(or) annual random
-effects for age 1 fish we assume them only in the northern regions for
-the northern component. The corresponding random effects are
+\end{equation}
+Because age 1 fish for the northern component can exist in both regions after January 1, natural mortality is acting in both regions for this stock.
+For models with covariate and(or) annual random effects for age 1 fish we assume them only in the northern regions for the northern component.
+The corresponding random effects are
 \begin{equation}\label{eq:Rec-re}
 \log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \varepsilon_{y,1}
-\end{equation} and \begin{equation}\label{eq:M-re}
+\end{equation}
+and
+\begin{equation}\label{eq:M-re}
 \log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \varepsilon_{M,y}.
-\end{equation} The constant or mean log-natural mortality rate is
-assumed to be \(\mu_{M,a} = \log(0.4)\) for all ages as recommended by
-the Working Group. Because the bottom temperature anomalies and random
-effects are centered at 0, mean log-natural mortality at age 1 over the
-time series should be approximately equal to \(\mu_{M,1}\) for models
-that include those effects.
+\end{equation}
+The constant or mean log-natural mortality rate is assumed to be \(\mu_{M,a} = \log(0.4)\) for all ages as recommended by the Working Group.
+Because the bottom temperature anomalies and random effects are centered at 0, mean log-natural mortality at age 1 over the time series should be approximately equal to \(\mu_{M,1}\) for models that include those effects.
 
 \hypertarget{model-fitting-and-diagnostics}{%
-\subsection*{Model fitting and
-diagnostics}\label{model-fitting-and-diagnostics}}
+\subsection*{Model fitting and diagnostics}\label{model-fitting-and-diagnostics}}
 \addcontentsline{toc}{subsection}{Model fitting and diagnostics}
 
-We used a development version
-(\href{https://github.com/timjmiller/wham/tree/fb8b089}{commit fb8b089})
-of the WHAM package prior to the release of version 2.0 for all results.
-All code for fitting models and generating results can be found at
-\href{https://github.com/timjmiller/wham_bsb_paper}{github.com/timjmiller/wham\_bsb\_paper}.
-
-We examined retrospective patterns for all models by fitting
-corresponding models where the terminal year is reduced sequentially by
-one year (peel) for seven years. Therefore, there are eight fits of each
-model with the time series reduced by zero to seven years. We calculated
-Mohn's \(\rho\) for SSB, and average \(F\) at ages 6 and 7. Absolute
-values of Mohn's \(\rho\) near 0 imply no pattern in estimation of these
-quantities as the time-series is sequentially extended. As in Miller et
-al. (2016b), we also assessed the consistency of the AIC-based model
-selection over retrospective peels to guard against previously noted
-changes in perception of covariate effects on recruitment with increased
-length of the time series of observations (Myers 1998). This
-retrospective examination was also recommended by Brooks (2024).
-
-We performed a jitter analysis of the base model \(M_0\) and the best
-fitting model to investigate whether a local mimimum of the negative
-log-likelihood surface was obtained by the optimization. We used the
-jitter\_wham function in the WHAM package which by default simulates
-starting values from a multivariate normal distribution with mean and
-covariance defined by the MLEs and hessian-based covariance matrix from
-the fitted model. See the Supplementary Materials for more details.
-
-For the best performing model, we also performed simulation self-tests
-where new observations were simulated conditional on all estimated fixed
-and random effects and the same model configuration was fit to each of
-the simulated data sets. We estimated median relative bias of SSB across
-these simulations. See the Supplementary Materials for more details.
+We used a development version (\href{https://github.com/timjmiller/wham/tree/fb8b089}{commit fb8b089}) of the WHAM package prior to the release of version 2.0 for all results.
+All code for fitting models and generating results can be found at \href{https://github.com/timjmiller/wham_bsb_paper}{github.com/timjmiller/wham\_bsb\_paper}.
+
+We examined retrospective patterns for all models by fitting corresponding models where the terminal year is reduced sequentially by one year (peel) for seven years.
+Therefore, there are eight fits of each model with the time series reduced by zero to seven years.
+We calculated Mohn's \(\rho\) for SSB, and average \(F\) at ages 6 and 7.
+Absolute values of Mohn's \(\rho\) near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended.
+As in Miller et al.
+(2016b), we also assessed the consistency of the AIC-based model selection over retrospective peels to guard against previously noted changes in perception of covariate effects on recruitment with increased length of the time series of observations (Myers 1998).
+This retrospective examination was also recommended by Brooks (2024).
+
+We performed a jitter analysis of the base model \(M_0\) and the best fitting model to investigate whether a local mimimum of the negative log-likelihood surface was obtained by the optimization.
+We used the jitter\_wham function in the WHAM package which by default simulates starting values from a multivariate normal distribution with mean and covariance defined by the MLEs and hessian-based covariance matrix from the fitted model.
+See the Supplementary Materials for more details.
+
+For the best performing model, we also performed simulation self-tests where new observations were simulated conditional on all estimated fixed and random effects and the same model configuration was fit to each of the simulated data sets.
+We estimated median relative bias of SSB across these simulations.
+See the Supplementary Materials for more details.
 
 \hypertarget{reference-points-and-projections}{%
-\subsection*{Reference points, and
-projections}\label{reference-points-and-projections}}
+\subsection*{Reference points, and projections}\label{reference-points-and-projections}}
 \addcontentsline{toc}{subsection}{Reference points, and projections}
 
-We produced SPR-based BRP estimates (\(F_{40\%}\) and SSB\(_{40\%}\))
-from the best performing model using a variety of options provided in
-WHAM. We generated annual BRP estimates based on the annual weight at
-age, maturity at age, natural mortality, fleet selectivity, and either
-the annual recruitment random effect (Eq. \ref{eq:Rec-re}) or the annual
-expected recruitment (Eq. \ref{eq:expected-recruitment}). We also
-calculated status as the ratios of annual average F and SSB to the
-corresponding reference points. Finally, we calculated ``prevailing''
-BRPs using the average weight at age, maturity at age, M, fleet
-selectivity from the last 5 years (2017 to 2021) and average recruitment
-over years 2000 to 2021, and made Kobe plots of joint status for the
-terminal year of the model (2021). All BRPs and status estimates are
-generated for each region as well as for the total stock.
-
-We projected the best black sea bass model under three alternative
-scenarios for the bottom temperature covariate where 1) the AR1
-time-series model continues into the projection years, 2) the average of
-the most recent 5 years is projected, and 3) bottom temperature
-increases in projection years following a prediction from a simple
-linear regression of the estimated bottom temperature anomalies over
-time. We specified fishing mortality in projection years to be constant
-at the value in the last model year (2021). Average weight, natural
-mortality, movement, and maturity at age over the last 5 years are used
-in projection years whereas the AR1 processes for all numbers at age
-random effects are continued.
+We produced SPR-based BRP estimates (\(F_{40\%}\) and SSB\(_{40\%}\)) from the best performing model using a variety of options provided in WHAM.
+We generated annual BRP estimates based on the annual weight at age, maturity at age, natural mortality, fleet selectivity, and either the annual recruitment random effect (Eq.
+\ref{eq:Rec-re}) or the annual expected recruitment (Eq.
+\ref{eq:expected-recruitment}).
+We also calculated status as the ratios of annual average F and SSB to the corresponding reference points.
+Finally, we calculated ``prevailing'' BRPs using the average weight at age, maturity at age, M, fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
+All BRPs and status estimates are generated for each region as well as for the total stock.
+
+We projected the best black sea bass model under three alternative scenarios for the bottom temperature covariate where 1) the AR1 time-series model continues into the projection years, 2) the average of the most recent 5 years is projected, and 3) bottom temperature increases in projection years following a prediction from a simple linear regression of the estimated bottom temperature anomalies over time.
+We specified fishing mortality in projection years to be constant at the value in the last model year (2021).
+Average weight, natural mortality, movement, and maturity at age over the last 5 years are used in projection years whereas the AR1 processes for all numbers at age random effects are continued.
 
@@
 
-We found the best model of bottom temperature covariate effects included
-the effect only on recruitment of the northern stock and no random
-effects on age 1 natural mortality (\(M_1\), Tables
-\ref{tab:diff-aic-table} and \ref{tab:aic-wts-table}). Model \(M_1\)
-consistently had the lowest AIC across retrospective peels where the
-terminal years of data are removed from the model. The difference in AIC
-for the model that also included bottom temperature effects on
-recruitment for the southern component (\(M_3\)) suggested some evidence
-for this hypothesis as well. There was little evidence for also
-including random effects on age 1 natural mortality for the northern
-stock component. However, when more than three terminal years of data
-were removed, the evidence for adding age 1 natural mortality random
-effects for the northern component (\(M_8\)) increased is similar to
-adding bottom temperature effects on southern component recruitment
-(\(M_3\)).
-
-Comparisons of estimates from retrospective peels did not show any
-indication of differences in model performance. There was very little
-variation in measures of retrospective patterns for SSB or fishing
-mortality across models. Mohn's \(\rho \approx -0.03\) for SSB of both
-stock components and Mohn's \(\rho \approx 0.03\) and \(\approx -0.04\)
-for average \(F\) at ages 6 and 7 in north and south regions,
-respectively (Table \ref{tab:rho-table}).
-
-Estimated effects of bottom temperature on recruitment for either the
-northern or southern component were stable over the retrospective peels
-and differed negligibly whether effects for each component were
-estimated in isolation or together (Table
-\ref{tab:beta-sig-peel-table}). The estimates for the effects on each
-stock component were essentially equivalent whether the effect was on
-one component or both. Estimates of residual variability in northern
-component recruitment, as measured by the conditional or marginal
-standard deviation of the recruitment random effects, increased slightly
-with the number of peeled years. However, the ratio of standard
-deviations of models with and without temperature effects was stable
-with approximately 20\% reduction in residual standard deviation when
-temperature effects were included. The residual variation is reduced
-because the expected recruitment (Eq. \ref{eq:expected-recruitment}) is
-a function of the covariate (Figure \ref{fig:BT-Ecov-R}).
-
-The posterior estimates of the bottom temperature covariate match the
-observations well because of the high precision of the observed
-anomalies (Figure \ref{fig:bottom-temperature}). The anomalies for the
-north and south regions appear highly correlated presumably because of
-the proximity of the two regions. Because the temperature anomalies are
-treated as latent variables, other data components in the model can
-affect the estimated anomalies when effects on recruitment are included
-(e.g., Miller et al. 2018), but in this case the estimates of anomalies
-are altered negligibly during the population model years when the effect
-on recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
-
-The effect of the anomalies on expected recruitment can be better
-observed by comparing the time series of the northern and southern
-components (Figure \ref{fig:E-R-SSB-F}). The expected recruitment for
-the southern component without any covariate effects is constant over
-time whereas the effect on expected recruitment for the northern
-component induces correlation of the bottom temperature anomalies and
-expected recruitment. For the northern stock component, most of the
-lowest estimated recruitment random effects occurred prior to 2000 and
-several large recruitment estimates occurred after 2010. For the
-southern stock, recruitment random effect estimates have been less
-variable and similar to the estimated median particularly since 2005.
-The estimated stock size of the northern component, as measured by SSB,
-increased substantially since 2005, whereas that for the southern
-component peaked in 2002. The increase in the northern component is also
-reflected in the total size of the combined components. Estimated
-average total fishing mortality rates across ages 6 and 7 (combining the
-commercial and recreational fleets) were highest in the southern region
-prior to 1998 and have declined substantially since then. Estimated
-fishing mortality rates in the northern region have been less variable,
-but had largest peaks in 1996, 1999, and 2005. The decline in estimates
-in the southern component is also reflected in the decline in total
-fishing mortality across regions.
-
-The estimated movement from the north to the south for the northern
-stock component is much less than anticipated from the prior
-distribution that was parameterized based on the companion Stock
-Synthesis model fit completed by the Working Group (Figure
-\ref{fig:move-prior-posterior}). Conversely, the estimated movement rate
-from south to north was more consistent and slightly greater than that
-from the companion model. The estimated movement rates varied little
-across all fitted models implying that the estimation of movement was
-not sensitive to the alternative assumptions we considered. Furthermore,
-the very low estimated movement rate from north to south is also
-consistent with preliminary fits of the WHAM model without the prior
-distribution where the movement rate was estimated at the lower bound of
-0 resulting in a lack of convergence. The estimated movement rates imply
-that a very small fraction of northern fish are ever in the southern
-region.
-
-The 2DAR1 random effects for the selectivity parameters of the northern
-recreational fleet indicate a decrease in selectivity at ages 2 to 6
-over the time series which corresponds to increases in the size limits
-in regulations by states where these catches occurred (Figure
-\ref{fig:selectivity-re}). The model also estimated a more modest
-decrease in selectivity at ages 2 to 3 for the northern commercial
-fleet. The selectivity at the youngest ages for the two fishery
-independent indices varied over time but without notable trend.
+We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock and no random effects on age 1 natural mortality (\(M_1\), Tables \ref{tab:diff-aic-table} and \ref{tab:aic-wts-table}).
+Model \(M_1\) consistently had the lowest AIC across retrospective peels where the terminal years of data are removed from the model.
+The difference in AIC for the model that also included bottom temperature effects on recruitment for the southern component (\(M_3\)) suggested some evidence for this hypothesis as well.
+There was little evidence for also including random effects on age 1 natural mortality for the northern stock component.
+However, when more than three terminal years of data were removed, the evidence for adding age 1 natural mortality random effects for the northern component (\(M_8\)) increased is similar to adding bottom temperature effects on southern component recruitment (\(M_3\)).
+
+Comparisons of estimates from retrospective peels did not show any indication of differences in model performance.
+There was very little variation in measures of retrospective patterns for SSB or fishing mortality across models.
+Mohn's \(\rho \approx -0.03\) for SSB of both stock components and Mohn's \(\rho \approx 0.03\) and \(\approx -0.04\) for average \(F\) at ages 6 and 7 in north and south regions, respectively (Table \ref{tab:rho-table}).
+
+Estimated effects of bottom temperature on recruitment for either the northern or southern component were stable over the retrospective peels and differed negligibly whether effects for each component were estimated in isolation or together (Table \ref{tab:beta-sig-peel-table}).
+The estimates for the effects on each stock component were essentially equivalent whether the effect was on one component or both.
+Estimates of residual variability in northern component recruitment, as measured by the conditional or marginal standard deviation of the recruitment random effects, increased slightly with the number of peeled years.
+However, the ratio of standard deviations of models with and without temperature effects was stable with approximately 20\% reduction in residual standard deviation when temperature effects were included.
+The residual variation is reduced because the expected recruitment (Eq.
+\ref{eq:expected-recruitment}) is a function of the covariate (Figure \ref{fig:BT-Ecov-R}).
+
+The posterior estimates of the bottom temperature covariate match the observations well because of the high precision of the observed anomalies (Figure \ref{fig:bottom-temperature}).
+The anomalies for the north and south regions appear highly correlated presumably because of the proximity of the two regions.
+Because the temperature anomalies are treated as latent variables, other data components in the model can affect the estimated anomalies when effects on recruitment are included (e.g., Miller et al.
+2018), but in this case the estimates of anomalies are altered negligibly during the population model years when the effect on recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
+
+The effect of the anomalies on expected recruitment can be better observed by comparing the time series of the northern and southern components (Figure \ref{fig:E-R-SSB-F}).
+The expected recruitment for the southern component without any covariate effects is constant over time whereas the effect on expected recruitment for the northern component induces correlation of the bottom temperature anomalies and expected recruitment.
+For the northern stock component, most of the lowest estimated recruitment random effects occurred prior to 2000 and several large recruitment estimates occurred after 2010.
+For the southern stock, recruitment random effect estimates have been less variable and similar to the estimated median particularly since 2005.
+The estimated stock size of the northern component, as measured by SSB, increased substantially since 2005, whereas that for the southern component peaked in 2002.
+The increase in the northern component is also reflected in the total size of the combined components.
+Estimated average total fishing mortality rates across ages 6 and 7 (combining the commercial and recreational fleets) were highest in the southern region prior to 1998 and have declined substantially since then.
+Estimated fishing mortality rates in the northern region have been less variable, but had largest peaks in 1996, 1999, and 2005.
+The decline in estimates in the southern component is also reflected in the decline in total fishing mortality across regions.
+
+The estimated movement from the north to the south for the northern stock component is much less than anticipated from the prior distribution that was parameterized based on the companion Stock Synthesis model fit completed by the Working Group (Figure \ref{fig:move-prior-posterior}).
+Conversely, the estimated movement rate from south to north was more consistent and slightly greater than that from the companion model.
+The estimated movement rates varied little across all fitted models implying that the estimation of movement was not sensitive to the alternative assumptions we considered.
+Furthermore, the very low estimated movement rate from north to south is also consistent with preliminary fits of the WHAM model without the prior distribution where the movement rate was estimated at the lower bound of 0 resulting in a lack of convergence.
+The estimated movement rates imply that a very small fraction of northern fish are ever in the southern region.
+
+The 2DAR1 random effects for the selectivity parameters of the northern recreational fleet indicate a decrease in selectivity at ages 2 to 6 over the time series which corresponds to increases in the size limits in regulations by states where these catches occurred (Figure \ref{fig:selectivity-re}).
+The model also estimated a more modest decrease in selectivity at ages 2 to 3 for the northern commercial fleet.
+The selectivity at the youngest ages for the two fishery independent indices varied over time but without notable trend.
 
@@
 
-Using the expected recruitment rather than the random effects as the
-multiplier with equilibrium spawning biomass per recruit
-(\(\upphi(\widetilde{F})\), eq. \ref{eq:ssbpr}) results in less variable
-estimates of annual SSB\(_{40\%}\) (Figure \ref{fig:annual-BRPs}: top
-row). Estimates were less variable for the southern stock component than
-the northern stock component mainly because the expected recruitment is
-constant for the former whereas that for the northern component varies
-via the effect of the bottom temperature anomalies. Temporal variation
-for the southern stock component is mainly caused by annual variation in
-the inputs to the \(\upphi(\widetilde{F})\) calculation and
-\(F_{40\%}\), but there is also some effect of the variation in the
-proportion of total recruitment for each stock component as used in the
-weighting terms to define the 40\% reduction in unfished spawning
-biomass per recruit (\(\lambda_s\) in Eq. \ref{eq:newton-obj}). Using
-the recruitment random effects instead of expected recruitment creates
-more uncertain estimates of SSB\(_{40\%}\) for the southern stock
-component whereas there is little difference for the northern component
-because there is less difference in the types of recruitment estimates
-resulting from the covariate effects (Figure
-\ref{fig:annual-SSB40-cvs}).
-
-Ratios of SSB to SSB\(_{40\%}\) should be viewed with caution because
-the SSB reference point is being defined using annual recruitments
-rather than some form of average recruitment and therefore, ratios
-larger than 2.5 which would correspond to unfished population sizes are
-possible (Figure \ref{fig:annual-BRPs}: second row). However, for the
-southern component, using the constant expected recruitment is
-comparable to using median recruitment and the annual variation in the
-inputs to \(\upphi(\widetilde{F})\) indicates no years where SSB was
-greater than unfished levels throughout the time series. The patterns in
-uncertainty of SSB ratios are the same as those for the SSB reference
-points (Figure \ref{fig:annual-SSB-status-cvs}).
-
-Differences between annual \(F_{40\%}\) estimates using the alternative
-types of recruitment estimates are much less than those for
-SSB\(_{40\%}\) because the recruitment estimates are only used in the
-weighting of the stock-specific \(\upphi(\widetilde{F})\) to determine
-\(F_{40\%}\) (\(\lambda_s\) in Eq. \ref{eq:newton-obj}, Figure
-\ref{fig:annual-BRPs}: third row). For similar reasons, estimates of
-uncertainty of \(F_{40\%}\) are also similar between the type of
-recruitment estimates (Figure \ref{fig:annual-F40-cvs}). Annual
-estimates for the northern stock component were lowest in the earliest
-years of the time series whereas those for the soutern component are
-relatively stable over time.
-
-For a given type of recruitment used to calculate \(F_{40\%}\), the
-annual ratios of average \(F\) to the corresponding reference point are
-identical for each stock component and in total because of the
-selectivity used to define a single F reference point across both
-regions is exactly the same as that for the annual estimate of fishing
-mortality at age by fleet (Figure \ref{fig:annual-BRPs}: fourth row). As
-for annual \(F_{40\%}\), estimates of uncertainty in the ratios are
-similar whether expected recruitment or the random effects are used to
-calculate the reference points (Figure \ref{fig:annual-F-status-cvs}).
-
-Unlike the annual ratios of \(F\) to \(F_{40\%}\), the \(F_{40\%}\)
-under prevailing conditions uses a selectivity defined by the average
-\(F\) at age and fleet over the last 5 years, which is different from
-that of the selectivity in the terminal year and therefore those ratios
-for each region and in total are different (Figure
-\ref{fig:annual-BRPs}: bottom row). Although the alternative types of
-recruitment estimates result in different points on the Kobe plots for
-joint status, the uncertainty in the estimates suggest similar
-probabilities of alternative quadrants. However, the usage of the
-confidence regions of the ratios to defined probabilities is an
-approximation because these ratios are functions of both empirical Bayes
-estimates of random effects as well as maximum likelihood estimates of
-fixed effects.
+Using the expected recruitment rather than the random effects as the multiplier with equilibrium spawning biomass per recruit (\(\upphi(\widetilde{F})\), eq.
+\ref{eq:ssbpr}) results in less variable estimates of annual SSB\(_{40\%}\) (Figure \ref{fig:annual-BRPs}: top row).
+Estimates were less variable for the southern stock component than the northern stock component mainly because the expected recruitment is constant for the former whereas that for the northern component varies via the effect of the bottom temperature anomalies.
+Temporal variation for the southern stock component is mainly caused by annual variation in the inputs to the \(\upphi(\widetilde{F})\) calculation and \(F_{40\%}\), but there is also some effect of the variation in the proportion of total recruitment for each stock component as used in the weighting terms to define the 40\% reduction in unfished spawning biomass per recruit (\(\lambda_s\) in Eq.
+\ref{eq:newton-obj}).
+Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB\(_{40\%}\) for the southern stock component whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
+
+Ratios of SSB to SSB\(_{40\%}\) should be viewed with caution because the SSB reference point is being defined using annual recruitments rather than some form of average recruitment and therefore, ratios larger than 2.5 which would correspond to unfished population sizes are possible (Figure \ref{fig:annual-BRPs}: second row).
+However, for the southern component, using the constant expected recruitment is comparable to using median recruitment and the annual variation in the inputs to \(\upphi(\widetilde{F})\) indicates no years where SSB was greater than unfished levels throughout the time series.
+The patterns in uncertainty of SSB ratios are the same as those for the SSB reference points (Figure \ref{fig:annual-SSB-status-cvs}).
+
+Differences between annual \(F_{40\%}\) estimates using the alternative types of recruitment estimates are much less than those for SSB\(_{40\%}\) because the recruitment estimates are only used in the weighting of the stock-specific \(\upphi(\widetilde{F})\) to determine \(F_{40\%}\) (\(\lambda_s\) in Eq.
+\ref{eq:newton-obj}, Figure \ref{fig:annual-BRPs}: third row).
+For similar reasons, estimates of uncertainty of \(F_{40\%}\) are also similar between the type of recruitment estimates (Figure \ref{fig:annual-F40-cvs}).
+Annual estimates for the northern stock component were lowest in the earliest years of the time series whereas those for the soutern component are relatively stable over time.
+
+For a given type of recruitment used to calculate \(F_{40\%}\), the annual ratios of average \(F\) to the corresponding reference point are identical for each stock component and in total because of the selectivity used to define a single F reference point across both regions is exactly the same as that for the annual estimate of fishing mortality at age by fleet (Figure \ref{fig:annual-BRPs}: fourth row).
+As for annual \(F_{40\%}\), estimates of uncertainty in the ratios are similar whether expected recruitment or the random effects are used to calculate the reference points (Figure \ref{fig:annual-F-status-cvs}).
+
+Unlike the annual ratios of \(F\) to \(F_{40\%}\), the \(F_{40\%}\) under prevailing conditions uses a selectivity defined by the average \(F\) at age and fleet over the last 5 years, which is different from that of the selectivity in the terminal year and therefore those ratios for each region and in total are different (Figure \ref{fig:annual-BRPs}: bottom row).
+Although the alternative types of recruitment estimates result in different points on the Kobe plots for joint status, the uncertainty in the estimates suggest similar probabilities of alternative quadrants.
+However, the usage of the confidence regions of the ratios to defined probabilities is an approximation because these ratios are functions of both empirical Bayes estimates of random effects as well as maximum likelihood estimates of fixed effects.
 
@@
 
-When continuing the AR1 model for the bottom temperature anomalies in
-the projection years, the uncertainty grows and asymptotes to a level
-determined by the marginal variance of the process and the predicted
-value approaches the estimated marginal mean value of approximately 0
-(Figure \ref{fig:R-BT-proj}). There is no uncertainty in the projected
-linear trend because the values were specified as known whereas there is
-a small level of uncertainty in the projections of the recent average
-based on the uncertainty in the estimated mean. Comparing projected
-recruitment for the northern and southern stock components, the
-projected expected recruitment has much lower uncertainty than the
-projected random effects regardless of whether bottom temperature
-affects recruitment. For the northern component, CVs of projected
-expected recruitment are smaller than those of recruitment random
-effect, but there are larger differences between CVs of projected
-expected recruitment assuming AR1 and the other scenarios for the future
-bottom temperature anomalies are made (Figure \ref{fig:R-F-SSB-CVs}).
-
-Because projected fishing mortality is specified to be the same as that
-in the terminal year the same uncertainty is also propagated in the
-projections (Figure \ref{fig:F-SSB-proj}). As we would expect, there is
-no effect of alternative assumptions about bottom temperature in the
-projections for the southern stock component because the effect of the
-bottom temperature anomaly is only on the northern component. However,
-the alternative bottom temperature assumptions do affect projections of
-northern component SSB and the projections do not stabilize when the
-anomalies are projected to increase linearly. When the same AR1
-assumption for the bottom temperature anomalies is used in projections,
-CVs of northern and total SSB projections are larger than when the
-recent average or linear trend are assumed (Figure
-\ref{fig:R-F-SSB-CVs}).
-
-Unlike the annual reference points during the model years (prior to
-2022), components of the \(\upphi(\widetilde{F})\) calculations are
-constant in the projection years where the only changing annual values
-are the recruitments used in the weighting for \(F_{40\%}\) and the
-recruitments used to calculate the resulting SSB\(_{40\%}\) (Figure
-\ref{fig:brps-proj}). Estimated reference points in the first few
-projection years can differ depending on whether the expected
-recruitment or the random effects are used in their calculation, but
-longer term projected values are identical because the random effect
-estimates approach the expected values as the influence of observations
-diminishes. However, the estimates of uncertainty in the longer term
-reference points is lower using the expected recruitment rather than the
-random effects (Figures \ref{fig:annual-F40-cvs} and
-\ref{fig:annual-SSB40-cvs}). The reduction in uncertainty using expected
-recruitment is less for \(F_{40\%}\) than SSB\(_{40\%}\) and for the
-northern component which includes bottom temperature effects than the
-southern component.
+When continuing the AR1 model for the bottom temperature anomalies in the projection years, the uncertainty grows and asymptotes to a level determined by the marginal variance of the process and the predicted value approaches the estimated marginal mean value of approximately 0 (Figure \ref{fig:R-BT-proj}).
+There is no uncertainty in the projected linear trend because the values were specified as known whereas there is a small level of uncertainty in the projections of the recent average based on the uncertainty in the estimated mean.
+Comparing projected recruitment for the northern and southern stock components, the projected expected recruitment has much lower uncertainty than the projected random effects regardless of whether bottom temperature affects recruitment.
+For the northern component, CVs of projected expected recruitment are smaller than those of recruitment random effect, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for the future bottom temperature anomalies are made (Figure \ref{fig:R-F-SSB-CVs}).
+
+Because projected fishing mortality is specified to be the same as that in the terminal year the same uncertainty is also propagated in the projections (Figure \ref{fig:F-SSB-proj}).
+As we would expect, there is no effect of alternative assumptions about bottom temperature in the projections for the southern stock component because the effect of the bottom temperature anomaly is only on the northern component.
+However, the alternative bottom temperature assumptions do affect projections of northern component SSB and the projections do not stabilize when the anomalies are projected to increase linearly.
+When the same AR1 assumption for the bottom temperature anomalies is used in projections, CVs of northern and total SSB projections are larger than when the recent average or linear trend are assumed (Figure \ref{fig:R-F-SSB-CVs}).
+
+Unlike the annual reference points during the model years (prior to 2022), components of the \(\upphi(\widetilde{F})\) calculations are constant in the projection years where the only changing annual values are the recruitments used in the weighting for \(F_{40\%}\) and the recruitments used to calculate the resulting SSB\(_{40\%}\) (Figure \ref{fig:brps-proj}).
+Estimated reference points in the first few projection years can differ depending on whether the expected recruitment or the random effects are used in their calculation, but longer term projected values are identical because the random effect estimates approach the expected values as the influence of observations diminishes.
+However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures \ref{fig:annual-F40-cvs} and \ref{fig:annual-SSB40-cvs}).
+The reduction in uncertainty using expected recruitment is less for \(F_{40\%}\) than SSB\(_{40\%}\) and for the northern component which includes bottom temperature effects than the southern component.
 
@@
 
-In our investigation of bottom temperature effects on recruitment and
-natural mortality and time-varying random effects on natural mortality
-at age 1, we found only including effects on recruitment for the
-northern stock component to be the best model with respect to AIC. Black
-sea bass in the NEUS is at the northern extent of its range and the
-estimated increase in expected recruitment with temperature for the
-northern stock component is consistent with its range extension to the
-north (McMahan et al. 2020). Opposite effects of temperature on
-recruitment or population size would be expected for species in the same
-general area that are at the southern extent of their range (Gabriel
-1992). For example, higher recruitment of the Southern New
-England-Mid-Atlantic stock of yellowtail flounder is correlated with
-more cold water persistence into summer fall on the Northeast US shelf
-(Sullivan et al. 2005; Miller et al. 2016b).
-
-Our finding regarding bottom temperature effects is consistent with the
-bottom temperature effects on recruitment in the assessment model
-accepted during the peer-review process and which is currently used to
-assess the stock (NEFSC 2023). However, AIC suggested some weight for
-models with temperature effects on recruitment for the southern
-component and random effects on M at age 1. Although this might suggest
-making inferences on the population by using an ensemble of these and
-weighting by AIC (Burnham and Anderson 2002), the estimates of
-assessment outputs relevant for management (e.g., SSB and \(F\)) were
-similar among the models and weighted estimates would differ very little
-from the estimates with the lowest AIC. Nevertheless, continued
-monitoring of effects on the southern component may be prudent.
-
-We showed that using expected recruitment rather than recruitment random
-effects to define BRPs can produce the same estimates but with lower
-uncertainty. Although the option is unavailable currently in WHAM, we
-expect using expected bottom temperature anomalies rather than the
-random effects would also result in the same BRPs and even less
-uncertainty than those we showed using expected recruitment. These
-results can be important for management when uncertainty in assessment
-output is used in making catch advice. The typical approach for defining
-stock and harvest status is to compare BRPs defined under prevailing
-conditions with current stock and harvest levels, but we could also make
-the comparisons with projected BRPs like those we demonstrated are
-possible. The uncertainty in these ratios could also be used in defining
-catch advice.
-
-For black sea bass, it is evident that the projection methodology of the
-environmental covariate has an impact on short-term projection estimates
-of recruitment and SSB (Figure 6). In the projection period, the AR1
-models for bottom temperature and recuitment revert to the long-term
-means of those processes. This is most likely not an appropriate
-assumption for stocks in the NEUS given observations and hypothesis of
-increasing water temperature and changing environmental conditions (Hare
-2016; Pershing et al. 2021). Because of these concerns, the most recent
-stock assessment for yellowtail flounder used a change point analysis on
-environmental covariates to determine current conditions. The mean
-environmental covariate from the current condition was then used as a
-data input in WHAM to inform short-term projections (NEFSC, 2024).
-Unfortunately, this methodology still has limitations because it is
-unlikely that current environmental conditions are stable. Similarly, a
-linear increase of temperature is also not likely in the future (see
-past trends, Figure 6 top row).As more assessments begin including
-effects of environmental covariates explicitly in assessment models,
-assumptions about these covariates in calculated BRPs and population
-projections becomes an important decision in the management process.
-
-\hypertarget{general-model-aspects}{%
-\subsection*{General Model aspects}\label{general-model-aspects}}
-\addcontentsline{toc}{subsection}{General Model aspects}
-
-WHAM provides a large class of models for time- and age-varying random
-effects for population attributes and treatment of environmental
-covariates and their effects on populations using state-space methods
-(Maunder 2024). This framework accounts for the magnitude of differing
-uncertainties in observations and stochastic population dynamics
-processes. Version 2.0 extends these inferences to movement rates for
-multi-region models and stock (component) specific configuration of
-random and environmental covariate effects.
-
-\hypertarget{future-developments}{%
-\subsection*{Future developments}\label{future-developments}}
-\addcontentsline{toc}{subsection}{Future developments}
-
-An obvious limitation to WHAM is the inability to include tagging
-observations of any type. Such observations are critical to estimation
-of movement parameters (e.g., Goethel et al. 2019). We used estimates of
-movement parameters from a companion assessment model to define prior
-distributions for corresponding movement parameters in the state-space
-black sea bass model, but other than the tagging data, both of the
-assessment models used much the same data which can lead to
-inappropriate inferences. In lieu of integrating an observation model
-for the tagging data, a better approach would be to estimate the
-movement parameters externally from just the tagging data.
+In our investigation of bottom temperature effects on recruitment and natural mortality and time-varying random effects on natural mortality at age 1, we found only including effects on recruitment for the northern stock component to be the best model with respect to AIC.
+Black sea bass in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north (McMahan et al.
+2020).
+Opposite effects of temperature on recruitment or population size would be expected for species in the same general area that are at the southern extent of their range (Gabriel 1992).
+For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer fall on the Northeast US shelf (Sullivan et al.
+2005; Miller et al.
+2016b).
+
+Our finding regarding bottom temperature effects is consistent with the bottom temperature effects on recruitment in the assessment model accepted during the peer-review process and which is currently used to assess the stock (NEFSC 2023).
+However, AIC suggested some weight for models with temperature effects on recruitment for the southern component and random effects on M at age 1.
+Although this might suggest making inferences on the population by using an ensemble of these and weighting by AIC (Burnham and Anderson 2002), the estimates of assessment outputs relevant for management (e.g., SSB and \(F\)) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
+Nevertheless, continued monitoring of effects on the southern component may be prudent.
+
+We showed that using expected recruitment rather than recruitment random effects to define BRPs can produce the same estimates but with lower uncertainty.
+Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same BRPs and even less uncertainty than those we showed using expected recruitment.
+These results can be important for management when uncertainty in assessment output is used in making catch advice.
+The typical approach for defining stock and harvest status is to compare BRPs defined under prevailing conditions with current stock and harvest levels, but we could also make the comparisons with projected BRPs like those we demonstrated are possible.
+The uncertainty in these ratios could also be used in defining catch advice.
+
+For black sea bass, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figure 6).
+In the projection period, the AR1 models for bottom temperature and recuitment revert to the long-term means of those processes.
+This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypothesis of increasing water temperature and changing environmental conditions (Hare 2016; Pershing et al.
+2021).
+Because of these concerns, the most recent stock assessment for yellowtail flounder used a change point analysis on environmental covariates to determine current conditions.
+The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections (NEFSC, 2024).
+Unfortunately, this methodology still has limitations because it is unlikely that current environmental conditions are stable.
+Similarly, a linear increase of temperature is also not likely in the future (see past trends, Figure 6 top row).As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
+
+\hypertarget{general-model}{%
+\subsection*{General model}\label{general-model}}
+\addcontentsline{toc}{subsection}{General model}
+
+WHAM provides a large class of models for time- and age-varying random effects for population attributes and treatment of environmental covariates and their effects on populations using state-space methods (Maunder 2024).
+This framework accounts for the magnitude of differing uncertainties in observations and stochastic population dynamics processes.
+Version 2.0 extends these inferences to movement rates for multi-region models and stock (component) specific configuration of random and environmental covariate effects.
+
+An obvious limitation to WHAM is the inability to include tagging observations of any type.
+Such observations are critical to estimation of movement parameters (e.g., Goethel et al.
+2019).
+We used estimates of movement parameters from a companion assessment model to define prior distributions for corresponding movement parameters in the state-space black sea bass model, but other than the tagging data, both of the assessment models used much the same data which can lead to inappropriate inferences.
+In lieu of integrating an observation model for the tagging data, a better approach would be to estimate the movement parameters externally from just the tagging data.
 
 Tagging data can also inform mortality rate parameters (Hampton 1991).
-It is well known that natural mortality is seldom estimated in
-assessment models because the observed data often provide little
-information to distinguish natural mortality from other assessment model
-parameters (Lee et al. 2011; Clark 2022). Estimation of natural
-mortality may be even more challenging within state-space assessment
-models with their greater flexibility from inclusion of time-varying
-random effects. For example, we fixed the mean natural mortality for age
-1 fish and Cadigan (2016) and Stock et al. (2021) also estimated natural
-mortality deviations. However, the simulation studies of Miller et al.
-(In review) suggest estimating natural mortality in WHAM is feasible in
-some situations with typical assessment data. Regardless, tagging
-observations can provide improved inferences and allowing tagging
-observations in WHAM should be a high priority even for models with a
-single stock and region. This extension would also be beneficial for
-black sea bass because the current assessment uses much the same data to
-estimate movement outside of the model as we are using to estimate the
-other parameters in the WHAM model.
-
-WHAM version 2.0 only allows models where each stock spawns and recruits
-in a single region however, some stocks are assessed with spatial units
-that are assumed to have a global stock and recruitment model that
-distributes recruitment to each region (Kapur et al. 2021). Therefore,
-extending WHAM to allow such models would be beneficial. Similarly age
-composition data are scarce for many stocks incorporating the work by
-Correa et al. (2023) on allowing length composition and modeling growth
-within WHAM would provide an assessment tool to include process errors
-in assessment with such data and possibly with multiple stocks and
-regions.
-
-\hypertarget{conclusions}{%
-\subsection*{Conclusions}\label{conclusions}}
-\addcontentsline{toc}{subsection}{Conclusions}
-
-WHAM version 2.0 extends the existing R package to allow multiple stocks
-and discrete spatial regions. It allows autocorrelated random effects
-and environmental covariate effects on movement rates. Estimates of
-movement rates from auxiliary studies can be used to construct prior
-distributions for movement rates in lieu of integrated likelihoods for
-tagging data. Version 2.0 also allows seasonal treatment of different
-fishing fleets and movement of individual stocks, and internal
-estimation of SPR- and MSY-based reference points accounting for
-seasonal spatial dynamics including movement and mortality and any
-environmental covariate effects.
-
-We applied this extended version to investigate alternative hypotheses
-about effects of bottom temperature on recruitment and age 1 natural
-mortality for black sea bass. Our analyses indicate evidence for effects
-of bottom temperature on recruitment for the northern component was
-stronger than other models that included effects on age 1 mortality
-and/or corresponding effects on the southern stock component.
-
-Future WHAM development should prioritize including tagging data to
-inform stock assessment parameter estimation and merging in the version
-created by Correa et al. (2023) to allow modeling of growth and
-inclusion of length and length-at-age composition observations.
+It is well known that natural mortality is seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality from other assessment model parameters (Lee et al.
+2011; Clark 2022).
+Estimation of natural mortality may be even more challenging within state-space assessment models with their greater flexibility from inclusion of time-varying random effects.
+For example, we fixed the mean natural mortality for age 1 fish and Cadigan (2016) and Stock et al.
+(2021) also estimated natural mortality deviations.
+However, the simulation studies of Miller et al.
+(In review) suggest estimating natural mortality in WHAM is feasible in some situations with typical assessment data.
+Regardless, tagging observations can provide improved inferences and allowing tagging observations in WHAM should be a high priority even for models with a single stock and region.
+This extension would also be beneficial for black sea bass because the current assessment uses much the same data to estimate movement outside of the model as we are using to estimate the other parameters in the WHAM model.
+
+WHAM version 2.0 only allows models where each stock spawns and recruits in a single region however, some stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region (Kapur et al.
+2021).
+Therefore, extending WHAM to allow such models would be beneficial.
+Similarly age composition data are scarce for many stocks incorporating the work by Correa et al.
+(2023) on allowing length composition and modeling growth within WHAM would provide an assessment tool to include process errors in assessment with such data and possibly with multiple stocks and regions.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ablehales97}{}}%
-Able, K.W., and Stanton Hales, Jr., L. 1997. Movements of juvenile black
-sea bass \emph{{C}entropristis striata} (linnaeus) in a southern {N}ew
-{J}ersey estuary. Journal of Experimental Marine Biology and Ecology
-\textbf{213}(2): 153--167.
+Able, K.W., and Stanton Hales, Jr., L.
+1997.
+Movements of juvenile black sea bass \emph{{C}entropristis striata} (linnaeus) in a southern {N}ew {J}ersey estuary.
+Journal of Experimental Marine Biology and Ecology \textbf{213}(2): 153--167.
 doi:\href{https://doi.org/10.1016/S0022-0981(96)02743-8}{10.1016/S0022-0981(96)02743-8}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-andersenkeiding02}{}}%
-Andersen, P.K., and Keiding, N. 2002. Multi-state models for event
-history analysis. Statistical Methods in Medical Research \textbf{11}:
-91--115.
+Andersen, P.K., and Keiding, N.
+2002.
+Multi-state models for event history analysis.
+Statistical Methods in Medical Research \textbf{11}: 91--115.
 doi:\href{https://doi.org/10.1191/0962280202SM276ra}{10.1191/0962280202SM276ra}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-arnason72}{}}%
-Arnason, A.N. 1972. Parameter estimates from mark-recapture experiments
-on two populations subject to migration and death. Researches on
-Population Ecology \textbf{13}: 97--113.
+Arnason, A.N.
+1972.
+Parameter estimates from mark-recapture experiments on two populations subject to migration and death.
+Researches on Population Ecology \textbf{13}: 97--113.
 doi:\href{https://doi.org/10.1007/BF02521971}{10.1007/BF02521971}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-asmfc16}{}}%
-ASMFC. 2016. Black sea bass stock assessment. {Available} at
-http://www.asmfc.org/uploads/file/5953f11d2016BlackSeaBassStockAssmt.pdf.
+ASMFC. 2016. Black sea bass stock assessment. {Available} at http://www.asmfc.org/uploads/file/5953f11d2016BlackSeaBassStockAssmt.pdf.
 
 \leavevmode\vadjust pre{\hypertarget{ref-belletal15}{}}%
-Bell, R.J., Richardson, D.E., Hare, J.A., Lynch, P.D., and Fratantoni,
-P.S. 2015. Disentangling the effects of climate, abundance, and size on
-the distribution of marine fish: An example based on four stocks from
-the {N}ortheast {US} shelf. ICES Journal of Marine Science
-\textbf{72}(5): 1311--1322.
+Bell, R.J., Richardson, D.E., Hare, J.A., Lynch, P.D., and Fratantoni, P.S.
+2015.
+Disentangling the effects of climate, abundance, and size on the distribution of marine fish: An example based on four stocks from the {N}ortheast {US} shelf.
+ICES Journal of Marine Science \textbf{72}(5): 1311--1322.
 doi:\href{https://doi.org/10.1093/icesjms/fsu217}{10.1093/icesjms/fsu217}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bosleyetal22}{}}%
-Bosley, K.M., Schueller, A.M., Goethel, D.R., Hanselman, D.H., Fenske,
-K.H., Berger, A.M., Deroba, J.J., and Langseth, B.J. 2022. Finding the
-perfect mismatch: Evaluating misspecification of population structure
-within spatially explicit integrated population models. Fish and
-Fisheries \textbf{23}(2): 294--315.
+Bosley, K.M., Schueller, A.M., Goethel, D.R., Hanselman, D.H., Fenske, K.H., Berger, A.M., Deroba, J.J., and Langseth, B.J.
+2022.
+Finding the perfect mismatch: Evaluating misspecification of population structure within spatially explicit integrated population models.
+Fish and Fisheries \textbf{23}(2): 294--315.
 doi:\href{https://doi.org/10.1111/faf.12616}{10.1111/faf.12616}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-brooks24}{}}%
-Brooks, E.N. 2024. Pragmatic approaches to modeling recruitment in
-fisheries stock assessment: A perspective. Fisheries Research
-\textbf{270}: 106896.
+Brooks, E.N.
+2024.
+Pragmatic approaches to modeling recruitment in fisheries stock assessment: A perspective.
+Fisheries Research \textbf{270}: 106896.
 doi:\href{https://doi.org/10.1016/j.fishres.2023.106896}{10.1016/j.fishres.2023.106896}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-burnhamanderson02}{}}%
-Burnham, K.P., and Anderson, D.R. 2002. Model selection and multimodel
-inference: A practical information-theoretic approach. Springer-Verlag,
-New York.
+Burnham, K.P., and Anderson, D.R. 2002. Model selection and multimodel inference: A practical information-theoretic approach. Springer-Verlag, New York.
 
 \leavevmode\vadjust pre{\hypertarget{ref-cadigan16}{}}%
-Cadigan, N.G. 2016. A state-space stock assessment model for northern
-cod, including under-reported catches and variable natural mortality
-rates. Canadian Journal of Fisheries and Aquatic Sciences
-\textbf{73}(2): 296--308.
+Cadigan, N.G.
+2016.
+A state-space stock assessment model for northern cod, including under-reported catches and variable natural mortality rates.
+Canadian Journal of Fisheries and Aquatic Sciences \textbf{73}(2): 296--308.
 doi:\href{https://doi.org/10.1139/cjfas-2015-0047}{10.1139/cjfas-2015-0047}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-caoetal14}{}}%
-Cao, J., Truesdell, S.B., and Chen, Y. 2014. Impacts of seasonal stock
-mixing on the assessment of {A}tlantic cod in the {G}ulf of {M}aine.
+Cao, J., Truesdell, S.B., and Chen, Y.
+2014.
+Impacts of seasonal stock mixing on the assessment of {A}tlantic cod in the {G}ulf of {M}aine.
 ICES Journal of Marine Science \textbf{71}(6): 1443--1457.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-clark22}{}}%
-Clark, W.G. 2022. Why natural mortality is estimable, in theory if not
-in practice, in a data-rich stock assessment. Fisheries Research
-\textbf{248}: 106203.
+Clark, W.G.
+2022.
+Why natural mortality is estimable, in theory if not in practice, in a data-rich stock assessment.
+Fisheries Research \textbf{248}: 106203.
 doi:\href{https://doi.org/10.1016/j.fishres.2021.106203}{10.1016/j.fishres.2021.106203}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-commenges99}{}}%
-Commenges, D. 1999. Multi-state models in epidemiology. Lifetime Data
-Analysis \textbf{5}: 315--327.
-doi:\href{https://doi.org/10.1023/A:1009636125294}{10.1023/A:1009636125294}.
+Commenges, D. 1999. Multi-state models in epidemiology. Lifetime Data Analysis \textbf{5}: 315--327. doi:\href{https://doi.org/10.1023/A:1009636125294}{10.1023/A:1009636125294}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-correaetal23}{}}%
-Correa, G.M., Monnahan, C.C., Sullivan, J.Y., Thorson, J.T., and Punt,
-A.E. 2023. Modelling time-varying growth in state-space stock
-assessments. ICES Journal of Marine Science \textbf{80}(7): 2036--2049.
+Correa, G.M., Monnahan, C.C., Sullivan, J.Y., Thorson, J.T., and Punt, A.E.
+2023.
+Modelling time-varying growth in state-space stock assessments.
+ICES Journal of Marine Science \textbf{80}(7): 2036--2049.
 doi:\href{https://doi.org/10.1093/icesjms/fsad133}{10.1093/icesjms/fsad133}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-fabrizioetal13}{}}%
-Fabrizio, M.C., Manderson, J.P., and Pessutti, J.P. 2013. Habitat
-associations and dispersal of black sea bass from a {M}id-{A}tlantic
-{B}ight reef. Marine Ecology Progress Series \textbf{482}: 241--253.
+Fabrizio, M.C., Manderson, J.P., and Pessutti, J.P.
+2013.
+Habitat associations and dispersal of black sea bass from a {M}id-{A}tlantic {B}ight reef.
+Marine Ecology Progress Series \textbf{482}: 241--253.
 doi:\href{https://doi.org/10.3354/meps10302}{10.3354/meps10302}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-francis14}{}}%
-Francis, R.I.C.C. 2014. Replacing the multinomial in stock assessment
-models: A first step. Fisheries Research \textbf{151}: 70--84.
+Francis, R.I.C.C.
+2014.
+Replacing the multinomial in stock assessment models: A first step.
+Fisheries Research \textbf{151}: 70--84.
 doi:\href{https://doi.org/10.1016/j.fishres.2013.12.015}{10.1016/j.fishres.2013.12.015}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gabriel92}{}}%
-Gabriel, W.L. 1992. Persistence of demersal fish assemblages between
-{C}ape {H}atteras and {N}ova {S}cotia, {N}orthwest {A}tlantic. Journal
-of the Northwest Atlantic Fisheries Science \textbf{14}(1): 29--46.
+Gabriel, W.L.
+1992.
+Persistence of demersal fish assemblages between {C}ape {H}atteras and {N}ova {S}cotia, {N}orthwest {A}tlantic.
+Journal of the Northwest Atlantic Fisheries Science \textbf{14}(1): 29--46.
 
 \leavevmode\vadjust pre{\hypertarget{ref-goetheletal19}{}}%
-Goethel, D.R., Bosley, K.M., Hanselman, D.H., Berger, A.M., Deroba,
-J.J., Langseth, B.J., and Schueller, A.M. 2019. Exploring the utility of
-different tag-recovery experimental designs for use in spatially
-explicit, tag-integrated stock assessment models. Fisheries Research
-\textbf{219}: 105320.
+Goethel, D.R., Bosley, K.M., Hanselman, D.H., Berger, A.M., Deroba, J.J., Langseth, B.J., and Schueller, A.M.
+2019.
+Exploring the utility of different tag-recovery experimental designs for use in spatially explicit, tag-integrated stock assessment models.
+Fisheries Research \textbf{219}: 105320.
 doi:\href{https://doi.org/10.1016/j.fishres.2019.105320}{10.1016/j.fishres.2019.105320}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hampton91}{}}%
-Hampton, J. 1991. Estimation of southern bluefin tuna \emph{{T}hunnus
-maccoyii} natural mortality and movement rates from tagging experiments.
-Fishery Bulletin \textbf{89}(4): 591--610.
+Hampton, J. 1991. Estimation of southern bluefin tuna \emph{{T}hunnus maccoyii} natural mortality and movement rates from tagging experiments. Fishery Bulletin \textbf{89}(4): 591--610.
 
 \leavevmode\vadjust pre{\hypertarget{ref-hareetal16}{}}%
-Hare, W.E.A.N., Jonathan A. AND Morrison. 2016. A vulnerability
-assessment of fish and invertebrates to climate change on the northeast
-u.s. Continental shelf. PLOS ONE \textbf{11}: 1--30.
+Hare, W.E.A.N., Jonathan A.
+AND Morrison.
+2016.
+A vulnerability assessment of fish and invertebrates to climate change on the northeast u.s.
+Continental shelf.
+PLOS ONE \textbf{11}: 1--30.
 doi:\href{https://doi.org/10.1371/journal.pone.0146756}{10.1371/journal.pone.0146756}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hearnetal87}{}}%
-Hearn, W.S., Sundland, R.L., and Hampton, J. 1987. Robust estimation of
-the natural mortality rate in a completed tagging experiment with
-variable fishing intensity. Journal Du Conseil International Pour
-L'exploration De La Mer \textbf{43}: 107--117.
+Hearn, W.S., Sundland, R.L., and Hampton, J.
+1987.
+Robust estimation of the natural mortality rate in a completed tagging experiment with variable fishing intensity.
+Journal Du Conseil International Pour L'exploration De La Mer \textbf{43}: 107--117.
 doi:\href{https://doi.org/10.1093/icesjms/43.2.107}{10.1093/icesjms/43.2.107}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kapuretal21}{}}%
-Kapur, M.S., Siple, M.C., Olmos, M., Privitera-Johnson, K.M., Adams, G.,
-Best, J., Castillo-Jordán, C., Cronin-Fine, L., Havron, A.M., Lee, Q.,
-Methot, R.D., and Punt, A.E. 2021. Equilibrium reference point
-calculations for the next generation of spatial assessments. Fisheries
-Research \textbf{244}: 106132.
+Kapur, M.S., Siple, M.C., Olmos, M., Privitera-Johnson, K.M., Adams, G., Best, J., Castillo-Jordán, C., Cronin-Fine, L., Havron, A.M., Lee, Q., Methot, R.D., and Punt, A.E.
+2021.
+Equilibrium reference point calculations for the next generation of spatial assessments.
+Fisheries Research \textbf{244}: 106132.
 doi:\href{https://doi.org/10.1016/j.fishres.2021.106132}{10.1016/j.fishres.2021.106132}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kemenysnell60}{}}%
-Kemeny, J.G., and Snell, J.L. 1960. Finite markov chains. D. Van
-Nostrand Company, Princeton, New Jersey.
+Kemeny, J.G., and Snell, J.L. 1960. Finite markov chains. D. Van Nostrand Company, Princeton, New Jersey.
 
 \leavevmode\vadjust pre{\hypertarget{ref-kristensenetal16}{}}%
-Kristensen, K., Nielsen, A., Berg, C., Skaug, H., and Bell, B.M. 2016.
-{TMB}: Automatic differentiation and {Laplace} approximation. Journal of
-Statistical Software \textbf{70}: 1--21.
+Kristensen, K., Nielsen, A., Berg, C., Skaug, H., and Bell, B.M.
+2016.
+{TMB}: Automatic differentiation and {Laplace} approximation.
+Journal of Statistical Software \textbf{70}: 1--21.
 doi:\href{https://doi.org/10.18637/jss.v070.i05}{10.18637/jss.v070.i05}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-leeetal11}{}}%
-Lee, H.-H., Maunder, M.N., Piner, K.R., and Methot, R.D. 2011.
-Estimating natural mortality within a fisheries stock assessment model:
-An evaluation using simulation analysis based on twelve stock
-assessments. Fisheries Research \textbf{109}(1): 89--94.
+Lee, H.-H., Maunder, M.N., Piner, K.R., and Methot, R.D.
+2011.
+Estimating natural mortality within a fisheries stock assessment model: An evaluation using simulation analysis based on twelve stock assessments.
+Fisheries Research \textbf{109}(1): 89--94.
 doi:\href{https://doi.org/10.1016/j.fishres.2011.01.021}{10.1016/j.fishres.2011.01.021}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-legaultrestrepo99}{}}%
-Legault, C.M., and Restrepo, V.R. 1999. A flexible forward
-age-structured assessment program. Col. Vol. Sci. Pap. ICCAT
-\textbf{49}(2): 246--253.
+Legault, C.M., and Restrepo, V.R. 1999. A flexible forward age-structured assessment program. Col. Vol. Sci. Pap. ICCAT \textbf{49}(2): 246--253.
 
 \leavevmode\vadjust pre{\hypertarget{ref-maunder24}{}}%
-Maunder, M.N. 2024. Towards a comprehensive framework for providing
-management advice from statistical inference using population dynamics
-models. Ecological Modelling \textbf{498}: 110836.
+Maunder, M.N.
+2024.
+Towards a comprehensive framework for providing management advice from statistical inference using population dynamics models.
+Ecological Modelling \textbf{498}: 110836.
 doi:\href{https://doi.org/10.1016/j.ecolmodel.2024.110836}{10.1016/j.ecolmodel.2024.110836}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-mcmahanetal20}{}}%
-McMahan, M.D., Sherwood, G.D., and Grabowski, J.H. 2020. Geographic
-variation in life-history traits of black sea bass
-(\emph{{C}entropristis striata}) during a rapid range expansion.
+McMahan, M.D., Sherwood, G.D., and Grabowski, J.H.
+2020.
+Geographic variation in life-history traits of black sea bass (\emph{{C}entropristis striata}) during a rapid range expansion.
 Frontiers in Marine Science \textbf{7}: 567758.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-methotwetzel13}{}}%
-Methot, R.D., and Wetzel, C.R. 2013. Stock {S}ynthesis: A biological and
-statistical framework for fish stock assessment and fishery management.
+Methot, R.D., and Wetzel, C.R.
+2013.
+Stock {S}ynthesis: A biological and statistical framework for fish stock assessment and fishery management.
 Fisheries Research \textbf{142}(1): 86--99.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-milleretal16_yoy_survival}{}}%
-Miller, A.S., Shepherd, G.R., and Fratantoni, P.S. 2016a. Offshore
-habitat preference of overwintering juvenile and adult black sea bass,
-\emph{{C}entropristis} \emph{striata}, and the relationship to
-year-class success. {PLOS} {ONE} \textbf{11}(1): e0147627.
+Miller, A.S., Shepherd, G.R., and Fratantoni, P.S.
+2016a.
+Offshore habitat preference of overwintering juvenile and adult black sea bass, \emph{{C}entropristis} \emph{striata}, and the relationship to year-class success.
+{PLOS} {ONE} \textbf{11}(1): e0147627.
 doi:\href{https://doi.org/10.1371/journal.pone.0147627}{10.1371/journal.pone.0147627}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-millerandersen08}{}}%
-Miller, T.J., and Andersen, P.K. 2008. A finite-state continuous-time
-approach for inferring regional migration and mortality rates from
-archival tagging and conventional tag-recovery experiments. Biometrics
-\textbf{64}(4): 1196--1206.
+Miller, T.J., and Andersen, P.K.
+2008.
+A finite-state continuous-time approach for inferring regional migration and mortality rates from archival tagging and conventional tag-recovery experiments.
+Biometrics \textbf{64}(4): 1196--1206.
 doi:\href{https://doi.org/10.1111/j.1541-0420.2008.00996.x}{10.1111/j.1541-0420.2008.00996.x}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-milleretalinreview1}{}}%
-Miller, T.J., Britten, G., Brooks, E.N., Fay, G., Hansell, A., Legault,
-C.M., Li, C., Muffley, B., Stock, B.C., and Wiedenmann, J. In review.
-Factors affecting reliablity of state-space age-structured assessment
-models. Canadian Journal of Fisheries Science.
+Miller, T.J., Britten, G., Brooks, E.N., Fay, G., Hansell, A., Legault, C.M., Li, C., Muffley, B., Stock, B.C., and Wiedenmann, J.
+In review.
+Factors affecting reliablity of state-space age-structured assessment models.
+Canadian Journal of Fisheries Science.
 
 \leavevmode\vadjust pre{\hypertarget{ref-milleretal16}{}}%
-Miller, T.J., Hare, J.A., and Alade, L.A. 2016b. A state-space approach
-to incorporating environmental effects on recruitment in an
-age-structured assessment model with an application to southern {New
-England} yellowtail flounder. Canadian Journal of Fisheries Science
-\textbf{73}(8): 1261--1270.
+Miller, T.J., Hare, J.A., and Alade, L.A.
+2016b.
+A state-space approach to incorporating environmental effects on recruitment in an age-structured assessment model with an application to southern {New England} yellowtail flounder.
+Canadian Journal of Fisheries Science \textbf{73}(8): 1261--1270.
 doi:\href{https://doi.org/10.1139/cjfas-2015-0339}{10.1139/cjfas-2015-0339}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-milleretal18}{}}%
-Miller, T.J., O'Brien, L., and Fratantoni, P.S. 2018. Temporal and
-environmental variation in growth and maturity and effects on management
-reference points of {Georges Bank Atlantic} cod. Canadian Journal of
-Fisheries Science \textbf{75}(12): 2159--2171.
+Miller, T.J., O'Brien, L., and Fratantoni, P.S.
+2018.
+Temporal and environmental variation in growth and maturity and effects on management reference points of {Georges Bank Atlantic} cod.
+Canadian Journal of Fisheries Science \textbf{75}(12): 2159--2171.
 doi:\href{https://doi.org/10.1139/cjfas-2017-0124}{10.1139/cjfas-2017-0124}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-millerstock20}{}}%
-Miller, T.J., and Stock, B.C. 2020. The {Woods Hole Assessment Model}
-({WHAM}). https://timjmiller.github.io/wham.
+Miller, T.J., and Stock, B.C. 2020. The {Woods Hole Assessment Model} ({WHAM}). https://timjmiller.github.io/wham.
 
 \leavevmode\vadjust pre{\hypertarget{ref-mohn99}{}}%
-Mohn, R. 1999. The retrospective problem in sequential population
-analysis: An investigation using cod fishery and simulated data. ICES
-Journal of Marine Science \textbf{56}(4): 473--488.
+Mohn, R.
+1999.
+The retrospective problem in sequential population analysis: An investigation using cod fishery and simulated data.
+ICES Journal of Marine Science \textbf{56}(4): 473--488.
 doi:\href{https://doi.org/10.1006/jmsc.1999.0481}{10.1006/jmsc.1999.0481}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-mosershepherd09}{}}%
-Moser, J., and Shepherd, G.R. 2009. Seasonal distribution and movement
-of black sea bass (\emph{{C}entropristis striata}) in the {N}orthwest
-{A}tlantic as determined from a mark-recapture experiment. Journal of
-Northwest Atlantic Fishery Science \textbf{40}: 17--28.
+Moser, J., and Shepherd, G.R.
+2009.
+Seasonal distribution and movement of black sea bass (\emph{{C}entropristis striata}) in the {N}orthwest {A}tlantic as determined from a mark-recapture experiment.
+Journal of Northwest Atlantic Fishery Science \textbf{40}: 17--28.
 
 \leavevmode\vadjust pre{\hypertarget{ref-musickmercer77}{}}%
-Musick, J.A., and Mercer, L.P. 1977. Seasonal distribution of black sea
-bass, \emph{{C}entropristis striata}, in the {M}id-{A}tlantic {B}ight
-with comments on the ecology and fisheries of the species. Transactions
-of the American Fisheries Society \textbf{106}(1): 12--25.
+Musick, J.A., and Mercer, L.P.
+1977.
+Seasonal distribution of black sea bass, \emph{{C}entropristis striata}, in the {M}id-{A}tlantic {B}ight with comments on the ecology and fisheries of the species.
+Transactions of the American Fisheries Society \textbf{106}(1): 12--25.
 doi:\href{https://doi.org/10.1577/1548-8659(1977)106\%3C12:SDOBSB\%3E2.0.CO;2}{10.1577/1548-8659(1977)106\textless12:SDOBSB\textgreater2.0.CO;2}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-myers98}{}}%
-Myers, R.A. 1998. When do environment--recruitment correlations work?
+Myers, R.A.
+1998.
+When do environment--recruitment correlations work?
 Reviews in Fish Biology and Fisheries \textbf{8}(3): 229--249.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nefsc22}{}}%
-NEFSC. 2022a. Final report of the haddock research track assessment
-working group. {Available} at
-https://s3.us-east-1.amazonaws.com/nefmc.org/14b\_EGB\_Research\_Track\_Haddock\_WG\_Report\_DRAFT.pdf.
+NEFSC.
+2022a.
+Final report of the haddock research track assessment working group.
+{Available} at https://s3.us-east-1.amazonaws.com/nefmc.org/14b\_EGB\_Research\_Track\_Haddock\_WG\_Report\_DRAFT.pdf.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nefsc22a}{}}%
-NEFSC. 2022b. Report of the american plaice research track working
-group. {Available} at
-https://s3.us-east-1.amazonaws.com/nefmc.org/2\_American-Plaice-WG-Report.pdf.
+NEFSC. 2022b. Report of the american plaice research track working group. {Available} at https://s3.us-east-1.amazonaws.com/nefmc.org/2\_American-Plaice-WG-Report.pdf.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nefsc23}{}}%
-NEFSC. 2023. Report of the black sea bass (\emph{{C}entropristis}
-\emph{striata}) research track stock assessment working group.
-{Available} at
-https://www.mafmc.org/s/a\_2023\_BSB\_UNIT\_RTWG\_Report\_V2\_12\_2\_2023-1.pdf.
+NEFSC.
+2023.
+Report of the black sea bass (\emph{{C}entropristis} \emph{striata}) research track stock assessment working group.
+{Available} at https://www.mafmc.org/s/a\_2023\_BSB\_UNIT\_RTWG\_Report\_V2\_12\_2\_2023-1.pdf.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nefsc24}{}}%
-NEFSC. 2024. Butterfish research track assessment report. US Dept Commer
-Northeast Fish Sci Cent Ref Doc. 24-03; 191 p.
+NEFSC. 2024. Butterfish research track assessment report. US Dept Commer Northeast Fish Sci Cent Ref Doc. 24-03; 191 p.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nielsenberg14}{}}%
-Nielsen, A., and Berg, C.W. 2014. Estimation of time-varying selectivity
-in stock assessments using state-space models. Fisheries Research
-\textbf{158}: 96--101.
+Nielsen, A., and Berg, C.W.
+2014.
+Estimation of time-varying selectivity in stock assessments using state-space models.
+Fisheries Research \textbf{158}: 96--101.
 doi:\href{https://doi.org/10.1016/j.fishres.2014.01.014}{10.1016/j.fishres.2014.01.014}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-perreaultetal20}{}}%
-Perreault, A.M.J., Wheeland, L.J., Morgan, M.J., and Cadigan, N.G. 2020.
-A state-space stock assessment model for {American} plaice on the
-{Grand} {Bank} of {Newfoundland}. Journal of Northwest Atlantic Fishery
-Science \textbf{51}: 45--104.
+Perreault, A.M.J., Wheeland, L.J., Morgan, M.J., and Cadigan, N.G.
+2020.
+A state-space stock assessment model for {American} plaice on the {Grand} {Bank} of {Newfoundland}.
+Journal of Northwest Atlantic Fishery Science \textbf{51}: 45--104.
 doi:\href{https://doi.org/10.2960/j.v51.m727}{10.2960/j.v51.m727}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pershingetal21}{}}%
-Pershing, A.J., Alexander, M.A., Brady, D.C., Brickman, D., Curchitser,
-E.N., Diamond, A.W., McClenachan, L., Mills, K.E., Nichols, O.C.,
-Pendleton, D.E., Record, N.R., Scott, J.D., Staudinger, M.D., and Wang,
-Y. 2021. Climate impacts on the {G}ulf of {M}aine ecosystem: A review of
-observed and expected changes in 2050 from rising temperatures.
+Pershing, A.J., Alexander, M.A., Brady, D.C., Brickman, D., Curchitser, E.N., Diamond, A.W., McClenachan, L., Mills, K.E., Nichols, O.C., Pendleton, D.E., Record, N.R., Scott, J.D., Staudinger, M.D., and Wang, Y.
+2021.
+Climate impacts on the {G}ulf of {M}aine ecosystem: A review of observed and expected changes in 2050 from rising temperatures.
 Elementa: Science of the Anthropocene \textbf{9}(1): 00076.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-puntetal20}{}}%
-Punt, A.E., Dunn, A., Elvarsson, B., Hampton, J., Hoyle, S.D., Maunder,
-M.N., Methot, R.D., and Nielsen, A. 2020. Essential features of the
-next-generation integrated fisheries stock assessment package: A
-perspective. Fisheries Research \textbf{229}: 105617.
+Punt, A.E., Dunn, A., Elvarsson, B., Hampton, J., Hoyle, S.D., Maunder, M.N., Methot, R.D., and Nielsen, A.
+2020.
+Essential features of the next-generation integrated fisheries stock assessment package: A perspective.
+Fisheries Research \textbf{229}: 105617.
 doi:\href{https://doi.org/10.1016/j.fishres.2020.105617}{10.1016/j.fishres.2020.105617}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schwarzetal93}{}}%
-Schwarz, C.J., Schweigert, J.F., and Arnason, A.N. 1993. Estimating
-migration rates using tag-recovery data. Biometrics \textbf{49}:
-177--193. doi:\href{https://doi.org/10.2307/2532612}{10.2307/2532612}.
+Schwarz, C.J., Schweigert, J.F., and Arnason, A.N.
+1993.
+Estimating migration rates using tag-recovery data.
+Biometrics \textbf{49}: 177--193.
+doi:\href{https://doi.org/10.2307/2532612}{10.2307/2532612}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-stockmiller21}{}}%
-Stock, B.C., and Miller, T.J. 2021. The {Woods Hole Assessment Model}
-({WHAM}): A general state-space assessment framework that incorporates
-time- and age-varying processes via random effects and links to
-environmental covariates. Fisheries Research \textbf{240}: 105967.
+Stock, B.C., and Miller, T.J.
+2021.
+The {Woods Hole Assessment Model} ({WHAM}): A general state-space assessment framework that incorporates time- and age-varying processes via random effects and links to environmental covariates.
+Fisheries Research \textbf{240}: 105967.
 doi:\href{https://doi.org/10.1016/j.fishres.2021.105967}{10.1016/j.fishres.2021.105967}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stocketal21}{}}%
-Stock, B.C., Xu, H., Miller, T.J., Thorson, J.T., and Nye, J.A. 2021.
-Implementing two-dimensional autocorrelation in either survival or
-natural mortality improves a state-space assessment model for {Southern
-New England}-{Mid Atlantic} yellowtail flounder. Fisheries Research
-\textbf{237}: 105873.
+Stock, B.C., Xu, H., Miller, T.J., Thorson, J.T., and Nye, J.A.
+2021.
+Implementing two-dimensional autocorrelation in either survival or natural mortality improves a state-space assessment model for {Southern New England}-{Mid Atlantic} yellowtail flounder.
+Fisheries Research \textbf{237}: 105873.
 doi:\href{https://doi.org/10.1016/j.fishres.2021.105873}{10.1016/j.fishres.2021.105873}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-sullivanetal05}{}}%
-Sullivan, M.C., Cowen, R.K., and Steves, B.P. 2005. Evidence for
-atmosphere{--}ocean forcing of yellowtail flounder (\emph{{L}imanda}
-\emph{ferruginea}) recruitment in the {M}iddle {A}tlantic {B}ight.
+Sullivan, M.C., Cowen, R.K., and Steves, B.P.
+2005.
+Evidence for atmosphere{--}ocean forcing of yellowtail flounder (\emph{{L}imanda} \emph{ferruginea}) recruitment in the {M}iddle {A}tlantic {B}ight.
 Fisheries Oceanography \textbf{14}(5): 386--399.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorsonetal17}{}}%
-Thorson, J.T., Johnson, K.F., Methot, R.D., and Taylor, I.G. 2017.
-Model-based estimates of effective sample size in stock assessment
-models using the {D}irichlet-multinomial distribution. Fisheries
-Research \textbf{192}: 84--93.
+Thorson, J.T., Johnson, K.F., Methot, R.D., and Taylor, I.G.
+2017.
+Model-based estimates of effective sample size in stock assessment models using the {D}irichlet-multinomial distribution.
+Fisheries Research \textbf{192}: 84--93.
 doi:\href{https://doi.org/10.1016/j.fishres.2016.06.005}{10.1016/j.fishres.2016.06.005}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorsonetal23}{}}%
-Thorson, J.T., Miller, T.J., and Stock, B.C. 2023. The
-multivariate-tweedie: A self-weighting likelihood for age and length
-composition data arising from hierarchical sampling designs. ICES
-Journal of Marine Science \textbf{80}(10): 2630--2641.
+Thorson, J.T., Miller, T.J., and Stock, B.C.
+2023.
+The multivariate-tweedie: A self-weighting likelihood for age and length composition data arising from hierarchical sampling designs.
+ICES Journal of Marine Science \textbf{80}(10): 2630--2641.
 doi:\href{https://doi.org/10.1093/icesjms/fsac159}{10.1093/icesjms/fsac159}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-varkeyetal22}{}}%
-Varkey, D.A., Babyn, J., Regular, P., Ings, D.W., Kumar, R., Rogers, B.,
-Champagnat, J., and Morgan, M.J. 2022. A state-space model for stock
-assessment of cod (gadus morhua) stock in NAFO subdivision 3Ps. {DFO}
-{Can.} {Sci.} {Advis.} {Sec.} {Res.} {Doc.} 2022/022. v + 78 p.
+Varkey, D.A., Babyn, J., Regular, P., Ings, D.W., Kumar, R., Rogers, B., Champagnat, J., and Morgan, M.J.
+2022.
+A state-space model for stock assessment of cod (gadus morhua) stock in NAFO subdivision 3Ps.
+{DFO} {Can.} {Sci.} {Advis.} {Sec.} {Res.} {Doc.} 2022/022.
+v + 78 p.
 
 \leavevmode\vadjust pre{\hypertarget{ref-yingetal11}{}}%
-Ying, Y., Chen, Y., Lin, L., and Gao, T. 2011. Risks of ignoring fish
-population spatial structure in fisheries management. Canadian Journal
-of Fisheries and Aquatic Sciences \textbf{68}(12): 2101--2120.
+Ying, Y., Chen, Y., Lin, L., and Gao, T.
+2011.
+Risks of ignoring fish population spatial structure in fisheries management.
+Canadian Journal of Fisheries and Aquatic Sciences \textbf{68}(12): 2101--2120.
 doi:\href{https://doi.org/10.1139/f2011-116}{10.1139/f2011-116}.
@@
 \addcontentsline{toc}{section}{Figures}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{map}
+
+}
+
+\caption{The Northwest Atlantic coastal area where the NEUS black sea bass stock occurs.
+Red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations (after Miller et al.
+2016a) and the bathymetric contour is the 400 m isocline.}\label{fig:map}
+\end{figure}
+\pagebreak
 
@@
 \hypertarget{deriving-the-prior-distribution-for-movement-parameters}{%
-\subsection*{Deriving the prior distribution for movement
-parameters}\label{deriving-the-prior-distribution-for-movement-parameters}}
-\addcontentsline{toc}{subsection}{Deriving the prior distribution for
-movement parameters}
-
-The Working Group fit a Stock Synthesis model (Methot and Wetzel 2013)
-that included tagging data with 2 seasons (6 months each) and 2 regions
-where a proportion \(\mu^*_1\) of the northern component moves to the
-south in one season and some proportion \(\mu^*_{2\rightarrow 1}\) move
-back to the south in the second season (NEFSC 2023). The seasonal
-movement matrices for each season are \begin{equation*}
+\subsection*{Deriving the prior distribution for movement parameters}\label{deriving-the-prior-distribution-for-movement-parameters}}
+\addcontentsline{toc}{subsection}{Deriving the prior distribution for movement parameters}
+
+The Working Group fit a Stock Synthesis model (Methot and Wetzel 2013) that included tagging data with 2 seasons (6 months each) and 2 regions where a proportion \(\mu^*_1\) of the northern component moves to the south in one season and some proportion \(\mu^*_{2\rightarrow 1}\) move back to the south in the second season (NEFSC 2023).
+The seasonal movement matrices for each season are
+\begin{equation*}
 \boldsymbol{\mu}^*_{1} = 
@@
   \end{bmatrix}
-\end{equation*} and \begin{equation*}
+\end{equation*}
+and
+\begin{equation*}
 \boldsymbol{\mu}_{2} = 
@@
 
-To obtain estimates of movement proportions for the monthly intervals in
-the WHAM model, the half-year movement matrices were converted to
-monthly movement matrices by taking the root \(z_k\) of
-\(\boldsymbol{\mu}^*_{k}\) which are defined by the number of months of
-movement for each season (5 and 4, respectively). The roots of the
-matrices are calculated using an eigen decomposition of the matrices
+To obtain estimates of movement proportions for the monthly intervals in the WHAM model, the half-year movement matrices were converted to monthly movement matrices by taking the root \(z_k\) of \(\boldsymbol{\mu}^*_{k}\) which are defined by the number of months of movement for each season (5 and 4, respectively).
+The roots of the matrices are calculated using an eigen decomposition of the matrices
 \[ \boldsymbol{\mu}_k =  \left(\boldsymbol{\mu}_k^*\right)^{z_k} = \mathbf{V}_k \mathbf{D}_k^{z_k} \mathbf{V}_k^{-1}\]
-where \(z_1 = 1/5\) for and \(z_2 = 1/4\), and \(\mathbf{V}_{k}\) and
-\(\mathbf{D}_{k}\) are the matrix of eigenvectors (columnwise) and the
-diagonal matrix of corresponding eigenvalues of
-\(\boldsymbol{\mu}^*_k\). The Working Group used a parametric bootstrap
-approach to determine an appropriate standard deviation for the prior
-distribution for the movement parameters. Stock Synthesis also estimates
-parameters on a transformed scale, but different from WHAM:
+where \(z_1 = 1/5\) for and \(z_2 = 1/4\), and \(\mathbf{V}_{k}\) and \(\mathbf{D}_{k}\) are the matrix of eigenvectors (columnwise) and the diagonal matrix of corresponding eigenvalues of \(\boldsymbol{\mu}^*_k\).
+The Working Group used a parametric bootstrap approach to determine an appropriate standard deviation for the prior distribution for the movement parameters.
+Stock Synthesis also estimates parameters on a transformed scale, but different from WHAM:
 \[\mu^*_{r\rightarrow r'} = \frac{1}{1 + 2e^{-x_{r\rightarrow r'}}}\]
-The estimated parameters and standard errors from the Stock Synthesis
-model were \(x_{1\rightarrow 2}=-1.44\) and \(x_{2\rightarrow 1}=1.94\)
-and \(SE(x_{1\rightarrow 2})) = 0.21\) and
-\(SE(x_{2\rightarrow 1})) = 0.37\). The resulting in the estimated
-proportions were \(\mu^*_{1\rightarrow 2}=0.11\) and
-\(\mu^*_{2\rightarrow 1}=0.78\).
-
-In WHAM, an additive logit transformation is used which is simply a
-logit transformation when there are only two regions: \[
+The estimated parameters and standard errors from the Stock Synthesis model were \(x_{1\rightarrow 2}=-1.44\) and \(x_{2\rightarrow 1}=1.94\) and \(SE(x_{1\rightarrow 2})) = 0.21\) and \(SE(x_{2\rightarrow 1})) = 0.37\).
+The resulting in the estimated proportions were \(\mu^*_{1\rightarrow 2}=0.11\) and \(\mu^*_{2\rightarrow 1}=0.78\).
+
+In WHAM, an additive logit transformation is used which is simply a logit transformation when there are only two regions:
+\[
 \mu_{r\rightarrow r'} = \frac{1}{1+e^{-y_{r\rightarrow r'}}}.
-\] We simulated 1000 values from a normal distribution with mean and
-standard deviation defined by the parameter estimate and standard error
-\(\tilde x_{{r\rightarrow r'},b} \sim N(x_{r\rightarrow r'}, SE(x_{r\rightarrow r'}))\)
-from the Stock Synthesis model. For each simulated value we constructed
-\(\tilde {\boldsymbol{\mu}}^*_{{r\rightarrow r'},b}\), took the
-appropriate root and calculated inverse logit for
-\(\tilde y_{{r\rightarrow r'},b}\). We calculated the mean and standard
-deviation of the values \(y_{i,b}\). The mean values did not differ
-meaningfully from the transformation of the original estimates
-(\(y_{1\rightarrow 2} = -3.79\) and \(y_{2\rightarrow 1} = -0.79\)) and
-the standard deviation was approximately 0.2 for both parameters.
+\]
+We simulated 1000 values from a normal distribution with mean and standard deviation defined by the parameter estimate and standard error \(\tilde x_{{r\rightarrow r'},b} \sim N(x_{r\rightarrow r'}, SE(x_{r\rightarrow r'}))\) from the Stock Synthesis model.
+For each simulated value we constructed \(\tilde {\boldsymbol{\mu}}^*_{{r\rightarrow r'},b}\), took the appropriate root and calculated inverse logit for \(\tilde y_{{r\rightarrow r'},b}\).
+We calculated the mean and standard deviation of the values \(y_{i,b}\).
+The mean values did not differ meaningfully from the transformation of the original estimates (\(y_{1\rightarrow 2} = -3.79\) and \(y_{2\rightarrow 1} = -0.79\)) and the standard deviation was approximately 0.2 for both parameters.
 
@@
 \hypertarget{jitter-fits-for-model-m_0}{%
-\subsubsection*{\texorpdfstring{Jitter fits for model
-\(M_0\)}{Jitter fits for model M\_0}}\label{jitter-fits-for-model-m_0}}
+\subsubsection*{\texorpdfstring{Jitter fits for model \(M_0\)}{Jitter fits for model M\_0}}\label{jitter-fits-for-model-m_0}}
 \addcontentsline{toc}{subsubsection}{Jitter fits for model \(M_0\)}
 
-WHAM by default completes three newton steps after the stats::nlminb
-minimization function completes to reduce the gradient at the minimized
-NLL. However, this generally has negligible effects on model estimates
-and the NLL. To reduce computation time, we did not complete these
-newton steps when performing jitter fits of the model. Without the
-Newton steps, the maximum (absolute) gradient sizes are generally less
-than 0.01 for models that converge satisfactorily.
-
-The 50 jitter fits demonstrated that a local minimum was obtained for
-the original fit of model \(M_0\) (Figure \ref{fig:jitter-M0}). Some
-lower NLLs were obtained with unacceptable gradients, but a slightly
-lower NLL was found with a satisfactory gradient and with and invertible
-hessian. We therefore refit model \(M_0\) and all remaining models using
-the better parameter estimates as initial values.
+WHAM by default completes three newton steps after the stats::nlminb minimization function completes to reduce the gradient at the minimized NLL.
+However, this generally has negligible effects on model estimates and the NLL.
+To reduce computation time, we did not complete these newton steps when performing jitter fits of the model.
+Without the Newton steps, the maximum (absolute) gradient sizes are generally less than 0.01 for models that converge satisfactorily.
+
+The 50 jitter fits demonstrated that a local minimum was obtained for the original fit of model \(M_0\) (Figure \ref{fig:jitter-M0}).
+Some lower NLLs were obtained with unacceptable gradients, but a slightly lower NLL was found with a satisfactory gradient and with and invertible hessian.
+We therefore refit model \(M_0\) and all remaining models using the better parameter estimates as initial values.
 
 \hypertarget{jitter-fits-for-model-m_1}{%
-\subsubsection*{\texorpdfstring{Jitter fits for model
-\(M_1\)}{Jitter fits for model M\_1}}\label{jitter-fits-for-model-m_1}}
+\subsubsection*{\texorpdfstring{Jitter fits for model \(M_1\)}{Jitter fits for model M\_1}}\label{jitter-fits-for-model-m_1}}
 \addcontentsline{toc}{subsubsection}{Jitter fits for model \(M_1\)}
@@
 The 50 jitter fits gave no evidence of a better minimization of the NLL.
-Three lower NLLs were obtained, but with unacceptably large gradients
-(Figure \ref{fig:jitter-M1}). The largest differences in parameter
-estimates for these three jitters were for numbers at age and
-selectivity random effects variance and correlation parameters.
+Three lower NLLs were obtained, but with unacceptably large gradients (Figure \ref{fig:jitter-M1}).
+The largest differences in parameter estimates for these three jitters were for numbers at age and selectivity random effects variance and correlation parameters.
 
 \hypertarget{self-test-for-model-m_1}{%
-\subsubsection*{\texorpdfstring{Self test for model
-\(M_1\)}{Self test for model M\_1}}\label{self-test-for-model-m_1}}
+\subsubsection*{\texorpdfstring{Self test for model \(M_1\)}{Self test for model M\_1}}\label{self-test-for-model-m_1}}
 \addcontentsline{toc}{subsubsection}{Self test for model \(M_1\)}
 
-Initial fits to simulated data from model \(M_1\) showed estimation of
-the observation error standard deviation multiplier for the recreational
-catch-per angler indices in the north and south regions was unstable.
-Many of the fits to the simulated data produced implausible estimates at
-the 0 boundary for these parameters (very negative values on log-scale).
-Therefore, we completed self-tests with these parameters fixed at the
-true values.
-
-For 7 of the the simulated data sets the model failed to optimize. The
-maximum absolute gradient was \(<10^{-6}\) for only 9 and \(<10^{-4}\)
-for 52 of the 93 successfully fitted models. The poor convergence
-appeared to be attributable to the estimation of the scalar for the
-standard errors of the log-transformed northern Recreational CPA index
-for which estimates tended to 0 for nearly all of the fits (\(<0.01\)
-for 83 fits). However, even across all fits including those with poor
-convergence, the SSB estimates appeared to be reliable (Figure
-\ref{fig:self-test-fig}).
+Initial fits to simulated data from model \(M_1\) showed estimation of the observation error standard deviation multiplier for the recreational catch-per angler indices in the north and south regions was unstable.
+Many of the fits to the simulated data produced implausible estimates at the 0 boundary for these parameters (very negative values on log-scale).
+Therefore, we completed self-tests with these parameters fixed at the true values.
+
+For 7 of the the simulated data sets the model failed to optimize.
+The maximum absolute gradient was \(<10^{-6}\) for only 9 and \(<10^{-4}\) for 52 of the 93 successfully fitted models.
+The poor convergence appeared to be attributable to the estimation of the scalar for the standard errors of the log-transformed northern Recreational CPA index for which estimates tended to 0 for nearly all of the fits (\(<0.01\) for 83 fits).
+However, even across all fits including those with poor convergence, the SSB estimates appeared to be reliable (Figure \ref{fig:self-test-fig}).
 

### 2025-03-04 R/studyareamap.R

@@ -1,0 +1,50 @@
+
+# load the appropriate R package libraries
+library(rnaturalearth)
+library(maps)
+library(sp)
+library(sf)
+library(ggplot2)
+
+#hudson
+#hudson.n_UTM <- data.frame(x=c(357212.46,19314.01,658471.31,935303.78),y=c(4261346,4566676,5099985,4635884))
+spRef_UTM_19 <- "+proj=utm +zone=19 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
+spRef_DD <- "+proj=longlat +ellps=GRS80 +datum=NAD83 +no_defs"
+hudson.n_UTM <- data.frame(x=c(357212.46,19314.01),y=c(4261346,4566676))
+hudson.n_UTM.rl <- rbind(hudson.n_UTM, hudson.n_UTM[1,])
+hudson.n_UTM.l <- Line(hudson.n_UTM.rl)
+hudson.n_UTM.l <- Lines(list(hudson.n_UTM.l), ID = 'n')
+hudson.n_UTM.spL <- SpatialLines(list(hudson.n_UTM.l))
+proj4string(hudson.n_UTM.spL) <- CRS(spRef_UTM_19)
+hudson.n <- spTransform(hudson.n_UTM.spL, CRS(spRef_DD))
+hudson.n.sf <- st_as_sf(hudson.n,coords = c("x", "y"))
+world <- st_as_sf(maps::map("world", plot = FALSE, fill = TRUE))
+states <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))
+
+# contours <- sf::st_as_sf(sf::read_sf(("c:/work/shapefiles/Contours"), quiet=TRUE))
+# contour_400m <- subset(contours, elev_m == -400)
+# saveRDS(contour_400m, here::here("data","contour_400m.RDS"))
+contour_400m <- readRDS(here::here("data","contour_400m.RDS"))
+lat.range <- c(34.5,42)
+lon.range <- c(-78,-70)
+lat.range <- c(35,45)
+lon.range <- c(-76,-66)
+
+state_prov <- rnaturalearth::ne_states(c("united states of america", "canada"))
+
+# ggmap <- ggplot(states) +
+ggmap <- ggplot(state_prov) +
+  geom_sf(alpha = 0.5) +
+  # geom_sf(data = states, alpha = 0.5) +
+  geom_sf(data = contour_400m, linewidth = 0.1) +
[…]

### 2025-03-18 paper/paper.Rmd
add edits from Chris and a little tidying of the repo

@@
   - Timothy J. Miller^1^
-  - Kierstin Curti
+  - Kiersten Curti
   - Alexander C. Hansell
@@
 
-## Main Message {-}
-
-Describes the multi-region, multi-stock generalization of WHAM and its usage to evaluate evidence of alternative hypotheses about temperature effects on recruitment and age 1 natural mortality of black sea bass stock components.
+**keywords:** stock assessment, state-space, multi-stock, multi-region, age-structured, black sea bass
 
@@
 The Woods Hole Assessment Model is a general state-space age-structured assessment model that is used to assess and manage many stocks in the Northeast US.
-We first describe an extension of the model allowing any number of stocks (or stock components) and and regions with movement among regions as well as seasonal variation in stock and fleet dynamics.
+We first describe an extension of the model allowing any number of stocks (or stock components) and regions with movement among regions as well as seasonal variation in stock and fleet dynamics.
 Movement rates can be a be functions of time- and age-varying random effects and environmental covariates.
-We then illustrate the model by applying it to data for the northern and southern components of the Northeast US black sea bass stock and evaluate alternative hypotheses of bottom temperature and time-varying random effects on recruitment and natural mortality.
+We then illustrate the model by applying it to data for the northern and southern components of the Northeast United States black sea bass stock and evaluate alternative hypotheses of bottom temperature and time-varying random effects on recruitment and natural mortality.
 We show strong evidence for temperature effects on recruitment, primarily for the northern stock component, and no evidence for including random effects or temperature effects on age 1 natural mortality.
@@
 
-A state-space age-structured stock assessment models can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures [@nielsenberg14;@cadigan16;@milleretal16].
+A state-space age-structured stock assessment model can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures [@nielsenberg14;@cadigan16;@milleretal16].
 This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks [@puntetal20].
@@
 Black sea bass distribution has been linked to warming waters on the northwest shelf [@belletal15] and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north [@hareetal16].
-Thus, a focus of the research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics [@nefsc23].To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the NEUS using most of the model assumptions that were accepted during the peer-review process.
+Thus, a focus of the black sea bass research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics [@nefsc23].To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process.
 However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
@@
 Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points.
-We demonstrate some of options of WHAM version 2.0 to provide biological reference points and short-term projections.
+We demonstrate some of the options of WHAM version 2.0 to provide biological reference points and short-term projections.
 
@@
 For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
-The time interval $i$ with duration $\delta_i$ would be a season and the PTMs would be uniquely defined for each stock by season $i$, year $y$, and age $a$ on January 1.
-Each row and column of the PTM correspond to one the states: alive in region $r$, dead in fleet $f$, or dead from natural causes.
+The time interval $i$ with duration $\delta_i$ would be a season and the PTMs would be uniquely defined for each stock by year $y$, age $a$, and season $i$ on January 1.
+Each row and column of the PTM correspond to one of the states: alive in region $r$, dead in fleet $f$, or dead from natural causes.
 The probabilities in each row sum to unity (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval.
@@
 When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously.
-The sequential assumption is used widely in spatially explicit model [e.g., Stock Synthesis, @methotwetzel13].
+The sequential assumption is used widely in spatially explicit models [e.g., Stock Synthesis, @methotwetzel13].
 Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval.
@@
 When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
-We obtain the probability transition matrix over an interval $\delta_i$ by exponentiating the instantaneous rate matrix [@millerandersen08]
+We obtain the PTM over an interval $\delta_i$ by exponentiating the instantaneous rate matrix [@millerandersen08]
 \begin{equation*}
@@
  0 & 0 & 0 & 0 & 0
- \end{bmatrix}.
+ \end{bmatrix}
 \end{equation*}
@@
 
-<!--\textbf{Still needs work because it is slow.}-->
-
 ### Numbers at age {-}
 
-When there are $n_R$ regions and $n_F$ fleets, the vector of abundance in each state at age $a>1$ on January 1 is $\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}', \mathbf{0}')'$ where $\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})'$ is the number in the states corresponding to being alive in each region and $\mathbf{0}$ is a vector ($n_F+1$) for the numbers captured in each fleet and dead from natural mortality because no age $a$ fish have died yet on January 1.
+When there are $n_R$ regions and $n_F$ fleets, the vector of abundance in each state at age $a>1$ on January 1 is $\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}^T, \mathbf{0}^T)^T$ where $\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})^T$ is the number in the states corresponding to being alive in each region and $\mathbf{0}$ is a vector ($n_F+1$) for the numbers captured in each fleet and dead from natural mortality because no age $a$ fish have died yet on January 1.
 
@@
 where  
-$$\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}(t_s)' \mathbf{N}_{O,y,a}$$
+$$\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}^T(t_s) \mathbf{N}_{O,y,a}$$
 and $w_{y,a}$ is the mean weight at age of spawning individuals, $\text{mat}_{y,a}$ is the maturity at age, and $\mathbf{O}_{y,a,r_s}(t_s)$ are the probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year (the $r_s$ column of the upper-left submatrix of Eq.
@@
 
-As in previous versions of WHAM, the transitions in numbers at age from one year to another after recruitment can be treated deterministically or as functions of random effects.
+As in previous versions of WHAM, the apparent survival (and movement) from one year to another after recruitment can be treated deterministically or as functions of random effects.
 The predicted numbers at age in year $y$ at age $a$ for a given stock are vector analogs ($\mathbf{N}_{O,y,a}$) of the equations for numbers at age in the standard WHAM model [@stockmiller21].
 For ages $a = 2,\ldots, A-1$, where $A$ is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first $n_R$ elements of the vector
-$$\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.$$
-The numbers alive in each region can also be modeled more simply using the sub-matrix $\mathbf{O}_{y,a}$. The general model for the transitions in abundance at age is 
-\begin{equation*}
-\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
+$$\mathbf{P}_{y-1,a-1}^T \mathbf{N}_{y-1,a-1}.$$
+The numbers alive in each region can also be modeled more simply using the sub-matrix $\mathbf{O}_{y,a}$.
+The general model for the vector of abundance at age in year $y$ at age $a$ given the vector in year $y-1$ at age $a-1$ is
+\begin{equation*}
+\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}^T \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
 \end{equation*}
@@
 \begin{equation*}
-\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
+\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,A-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}^T \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}^T \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
 \end{equation*}
 where $\boldsymbol{\varepsilon}_{y,a}$ is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
-When there is autocorrelation with age, WHAM now assumes this applies only to ages $a>1$ by default so that recruitment random effects are independent of those for for the annual transitions of older age classes.
+When there is autocorrelation with age, WHAM now assumes this applies only to ages $a>1$ by default so that recruitment random effects are independent of those for the apparent survival of older age classes.
 So the general covariance structure for a given stock at ages $a>1$ in region $r$ is that of a two-dimensional first-order autoregressive (2DAR1) process
@@
 \end{equation*}
-When the transitions in abundance at age are treated deterministically, $\boldsymbol{\varepsilon}_{y,a} = 0$.
+When the annual changes in abundance at age are treated deterministically, $\boldsymbol{\varepsilon}_{y,a} = 0$.
 Since recruitment for a given stock currently only occurs in one region $r_s$ there is only a single vector of time-varying recruitment random effects for each stock.
@@
 Recall that recruitment for stock $s$ only occur in region $r_s$ so, the equilibrium initial numbers at age $a$ by region are
-$$\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}' \mathbf{N}_{O,1,1}.$$
+$$\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}^T \mathbf{N}_{O,1,1}.$$
 
@@
 
-For each season, there are at most $n_R-1$ parameters determining movement among regions given starting the season in region $r$ in either the sequential or simultaneous configurations.
+For each season, there are at most $n_R-1$ parameters determining movement for a given stock among regions given starting the season in region $r$ in either the sequential or simultaneous configurations.
 Movement parameters are estimated on a transformed scale via a link function $g(\cdot)$.
@@
   \begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
+  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y}.
   \end{equation*}
@@
 
-When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific.
+When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific, but currently are constant across seasons.
 When random effects are used, the same 2DAR1 structure with age and year as described by @stockmiller21 can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
@@
 
-The log-normal distributional assumption for aggregate catch observations is the same as @stockmiller21, but the predicted catch is now a function of catch from each stock starting the year in each region.
-For a given stock and age, the numbers captured in each fleet over the year are
-$$\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}' \mathbf{N}_{O,s,y,a}$$
-The predicted numbers caught be each fleet across stocks is 
+The log-normal distributional assumption for aggregate catch observations is the same as @stockmiller21, but the predicted catch is now a function of catch from each stock starting the year in each region and any seasonal movement.
+For a given stock and age, the vector of numbers captured by each fleet over the year are
+$$\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}^T \mathbf{N}_{O,s,y,a}$$
+(recall each fleet can harvest in only one region as described above). The vector of predicted numbers caught be each fleet across stocks is 
 $$\widehat{\mathbf{N}}_{H,y,a} = \sum^{n_S}_{s=1} \widehat{\mathbf{N}}_{H,s,y,a}$$
-and the predicted aggregate catch at age $a$ is
+and the vector of predicted aggregate catch by each fleet at age $a$ is
 $$\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \widehat{\mathbf{N}}_{H,y,a}$$
 where $\mathbf{c}_{y,a}$ is the vector of mean individual weight at age $a$ for each fleet and the aggregate catch by fleet is 
-$$\widehat{\mathbf{C}}_y = \sum^{A}_{a=1} \widehat{\mathbf{C}}_{y,a}$$
+$$\widehat{\mathbf{C}}_y = \sum^{A}_{a=1} \widehat{\mathbf{C}}_{y,a}.$$
 The log-aggregate catch observations for fleet $f$ are normally distributed
@@
 
-For index $m$ occurring in region $r_m$ at fraction of the year $t_m$, the predicted abundance at $t_m$ in region $r_m$ is 
-$$\widehat{N}_{s,y,a,m} = \mathbf{O}_{s,y,a,r_m}(t_m)' \mathbf{N}_{O,s,y,a}$$
+Like fishing fleets, an index $m$ can observe the population in a single region $r_m$ at fraction of the year $t_m$, the predicted abundance at $t_m$ in region $r_m$ is 
+$$\widehat{N}_{s,y,a,m} = \mathbf{O}_{s,y,a,r_m}^T(t_m) \mathbf{N}_{O,s,y,a}$$
 where $\mathbf{O}_{s,y,a,r_m}(t_m)$, the $r_m$ column of the upper-left submatrix of Eq.
@@
 
-As in previous versions of WHAM package, reference points for prevailing conditions, average all of the inputs to the spawning biomass and yield per recruit calculations (weight, maturity, natural mortality, and movement) over a set of user-specified years as shown in eq.
+As in previous versions of the WHAM package, reference points for prevailing conditions, average all of the inputs to the spawning biomass and yield per recruit calculations (weight, maturity, natural mortality, and movement) over a set of user-specified years as shown in eq.
 \ref{eq:spr_sel} for selectivity (e.g., last 5 years of the model).
 
-For $X\%$ SPR-based reference points, we use a Newton method and iterate
+For $X\%$ SPR-based reference points, we use a Newton method and iterate for steps $j = 1,\ldots,10$
 \begin{equation}\label{eq:newton}
-  \log\widetilde{F}^{(i)} = \log\widetilde{F}^{(i-1)} - \frac{g\left(\log\widetilde{F}^{(i-1)}\right)}{g'\left(\log\widetilde{F}^{(i-1)}\right)}
+  \log\widetilde{F}^{j} = \log\widetilde{F}^{j-1} - \frac{g\left(\log\widetilde{F}^{j-1}\right)}{g^{(1)}\left(\log\widetilde{F}^{j-1}\right)}
 \end{equation}
@@
 where $\upphi_s\left(F=0\right)$ is the equilibrium unfished spawning biomass per recruit.
-$g'(\log F)$ is the derivative of $g$ with respect to $\log F$, and the weights to use for each stock $\lambda_s$ can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate ``static'' equilibrium spawning biomass and yield.
+$g^{(1)}(\log F)$ is the derivative of $g$ with respect to $\log F$, and the weights to use for each stock $\lambda_s$ can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate "static" equilibrium spawning biomass and yield.
 
@@
 \ref{eq:ypr}) and equilibrium recruitment.
-The function $g(\log F)$ in Eq.
-\ref{eq:newton} is the first derivative of the yield curve with respect to $\log F$.
+The function $g(\log F)$ and $g^{(1)}(\log F)$ in Eq.
+\ref{eq:newton} are the first and second derivatives of the sum of the stock-specific equilibrium yields with respect to $\log F$ [analogous to that described by @milleretal16].
 
@@
 
-<!--
-## Model configurations  {-}
-
-We compared the base models for each stock in the single-stock and multi-stock versions of wham.
-
-```{r, echo = FALSE, out.width="75%", fig.align="center"}
-#knitr::include_graphics(here::here("docs","plots","compare_one_stock_models.png"))
-```
--->
-
 # Application to black sea bass {-}
@@
 Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging from the Gulf of Mexico to the Gulf of Maine.
-Fish north of Cape Hatteras, NC, are considered to comprise a single stock unit, but individual populations exhibit spawning site fidelity
-[@ablehales97;@fabrizioetal13].
+Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity [@ablehales97;@fabrizioetal13].
 Fish in this northern stock unit perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring [@musickmercer77].
@@
 Estimating initial numbers at age as separate parameters can be challenging even in single-stock models, but for black sea bass the available data cannot distinguish the proportion of northern and southern component fish at each age in the southern region in the initial year of the model.
-Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component: an initial recruitment and an equilibrium fully-selected $F$ that determines the abundance at age in each region for each stock component.
-
-For subsequent years, all models allow AR1 correlation of recruitment and 2DAR1 correlation for survival for both the northern and southern stock components.
+Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component: an initial recruitment and an equilibrium fully-selected $F$ (across all fleets) that determines the initial abundance at age in each region for each stock component.
+
+For subsequent years, all models allow AR1 correlation of recruitment and 2DAR1 correlation for apparent survival random effects for both the northern and southern stock components.
 Models with apparent survival random effects will model the transitions (conditional on both survival and movement) of abundances at age of northern origin fish in each region, but there appears to be little ability of the existing data from the southern region to distinguish apparent survival random effects for each stock component.
-All models we considered assume very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model.
+All models we considered assume a very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model.
 Unique variance and correlation parameters for the recruitment and apparent survival random effects are estimated for the northern and southern components.
@@
 All models we considered allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
-However, lack of model convergence was observed in many fits for self-test simulations.
-The model estimates are negligibly affected by estimating these scalars, but estimating rather than fixing these parameters allowed uncertainty in model output to be more properly conveyed.
+Although lack of model convergence was observed in many fits for self-test simulations, the estimation of critical model output (e.g., SSB, fishing mortality) was reliable and robust the poor estimation of these scalars (Figure \ref{fig:self-test-fig}), and estimating these parameters rather than fixing them in the fit to the actual data allows uncertainty in model output to be more properly conveyed.
 
@@
 The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual patterns and retrospective patterns.
-These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for six sets of of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
+These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for six sets of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
 
@@
 These analyses derive from the hypothesis that bottom temperature affects overwinter survival of fish where the fish turn from age 0 to age 1 on January 1 [@milleretal16_yoy_survival].
-This covariate may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of of the year after becoming age 1.
+This covariate may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of the year after becoming age 1.
 Furthermore, we have no direct observations of age 1 individuals from surveys until the spring season each year.
@@
 
-We assume the covariate in year $y$ affects recruitment in the same years because the covariate observations are from months January to March.
+We assume the covariate in year $y$ affects recruitment in the same year because the covariate observations are from months January to March.
 The fish are technically already 1 year old, but there are no observations of these individuals until later in the year except possibly in fishery catches which are accumulated over the whole year.
-Expected log-recruitment for a given stock is a linear function of bottom temperature
+Expected log-recruitment for a given stock component is a linear function of bottom temperature
 
@@
 \begin{equation}\label{eq:expected-M1}
-E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y
-\end{equation}
-Because age 1 fish for the northern component can exist in both regions after January 1, natural mortality is acting in both regions for this stock.
+E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y.
+\end{equation}
+Because age 1 fish for the northern component can exist in both regions after January 1 other the spawning period, natural mortality is acting in both regions for this stock.
 For models with covariate and(or) annual random effects for age 1 fish we assume them only in the northern regions for the northern component.
@@
 We used a development version ([commit fb8b089](https://github.com/timjmiller/wham/tree/fb8b089)) of the WHAM package prior to the release of version 2.0 for all results.
-All code for fitting models and generating results can be found at [github.com/timjmiller/wham_bsb_paper](https://github.com/timjmiller/wham_bsb_paper).
+All code for fitting models and generating results can be found at [github.com/timjmiller/multi_wham_bsb](https://github.com/timjmiller/multi_wham_bsb).
 
@@
 Finally, we calculated "prevailing" BRPs using the average weight at age, maturity at age, M, fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
-All BRPs and status estimates are generated for each region as well as for the total stock.
+All BRPs and status estimates are generated for each region as well as for the total stock area.
 
@@
 
-We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock and no random effects on age 1 natural mortality ($M_1$, Tables \ref{tab:diff-aic-table} and  \ref{tab:aic-wts-table}).
+We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock component and no random effects on age 1 natural mortality ($M_1$, Tables \ref{tab:diff-aic-table} and  \ref{tab:aic-wts-table}).
 Model $M_1$ consistently had the lowest AIC across retrospective peels where the terminal years of data are removed from the model.
@@
 Comparisons of estimates from retrospective peels did not show any indication of differences in model performance.
-There was very little variation in measures of retrospective patterns for SSB or fishing mortality across models.
+There was little variation in measures of retrospective patterns for SSB or fishing mortality across models.
 Mohn's $\rho \approx -0.03$ for SSB of both stock components and Mohn's $\rho \approx 0.03$ and $\approx -0.04$ for average $F$ at ages 6 and 7 in north and south regions, respectively (Table \ref{tab:rho-table}).
@@
 The effect of the anomalies on expected recruitment can be better observed by comparing the time series of the northern and southern components (Figure \ref{fig:E-R-SSB-F}).
-The expected recruitment for the southern component without any covariate effects is constant over time whereas the effect on expected recruitment for the northern component induces correlation of the bottom temperature anomalies and expected recruitment.
+The expected recruitment for the southern component without any covariate effects is constant over time whereas the effect of the anomalies on expected recruitment for the northern component induces annual variation in expected recruitment.
+The annual variation in expected recruitment (and the temperature anomalies) is correlated with the recruitment random effects of the northern component and therefore reduces the residual variation in the recruitment random effects.
 For the northern stock component, most of the lowest estimated recruitment random effects occurred prior to 2000 and several large recruitment estimates occurred after 2010.
-For the southern stock, recruitment random effect estimates have been less variable and similar to the estimated median particularly since 2005.
-The estimated stock size of the northern component, as measured by SSB, increased substantially since 2005, whereas that for the southern component peaked in 2002.
+For the southern stock component, recruitment random effect estimates have been less variable and similar to the estimated median particularly since 2005.
+The estimated size of the northern stock component, as measured by SSB, increased substantially since 2005, whereas that for the southern component peaked in 2002.
 The increase in the northern component is also reflected in the total size of the combined components.
@@
 The selectivity at the youngest ages for the two fishery independent indices varied over time but without notable trend.
+The selectivity of fleets and indices in the south increased with age, but were constant over time (Figure \ref{fig:selectivity-south}).
 
@@
 
-Differences between annual $F_{40\%}$ estimates using the alternative types of recruitment estimates are much less than those for SSB$_{40\%}$ because the recruitment estimates are only used in the weighting of the stock-specific $\upphi(\widetilde{F})$ to determine $F_{40\%}$ ($\lambda_s$ in Eq.
+Differences between annual $F_{40\%}$ estimates using the alternative types of recruitment estimates are much less than those for SSB$_{40\%}$ because the recruitment estimates are only used in the weighting of the stock component-specific $\upphi(\widetilde{F})$ to determine $F_{40\%}$ ($\lambda_s$ in Eq.
 \ref{eq:newton-obj}, Figure \ref{fig:annual-BRPs}: third row).
 For similar reasons, estimates of uncertainty of $F_{40\%}$ are also similar between the type of recruitment estimates (Figure \ref{fig:annual-F40-cvs}).
-Annual estimates for the northern stock component were lowest in the earliest years of the time series whereas those for the soutern component are relatively stable over time.
+Annual estimates for the northern stock component were lowest in the earliest years of the time series whereas those for the southern component are relatively stable over time.
 
@@
 Although the alternative types of recruitment estimates result in different points on the Kobe plots for joint status, the uncertainty in the estimates suggest similar probabilities of alternative quadrants.
-However, the usage of the confidence regions of the ratios to defined probabilities is an approximation because these ratios are functions of both empirical Bayes estimates of random effects as well as maximum likelihood estimates of fixed effects.
+However, the usage of the confidence regions of the ratios to define probabilities is an approximation because these ratios are functions of both empirical Bayes estimates of random effects as well as maximum likelihood estimates of fixed effects.
 
@@
 Opposite effects of temperature on recruitment or population size would be expected for species in the same general area that are at the southern extent of their range [@gabriel92].
-For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer fall on the Northeast US shelf [@sullivanetal05;@milleretal16].
+For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer and fall on the Northeast US shelf [@sullivanetal05;@milleretal16].
 
@@
 Unfortunately, this methodology still has limitations because it is unlikely that current environmental conditions are stable.
-Similarly, a linear increase of temperature is also not likely in the future (see past trends, Figure 6 top row).As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
-
-<!--
-
-### Self Tests {-}
-
-Simulation self tests performed by @stockmiller21 simulated process errors and observation errors and their simulations showed little bias in estimated assessment output such as SSB.
-@lietal24 used the same type of simulations and found what?.
-
-Here we conditioned on the estimated process errors for black sea bass.
-This conditional approach has been suggested as appropriate for tactical management where we are interested in inferences assuming our estimation of the population and fishing history from the original data is the truth [@zhengcadigan23,@cadiganetal24].
-We found some bias in SSB estimation for the black sea bass application (negatively for the northern stock), using the default model.
-We also found instability in the estimation of the observation variance parameters for the aggregate indices such that they often were estimated at the lower bound of 0.
-However, we found that the magnitude of bias was largely influenced by the large uncertainty estimated in some of the age composition observations assumed to have a logistic-normal distribution.
-When using the logistic-normal likelihood for management, we recommend inspecting the size of the estimated (marginal) standard deviation of the logistic normal model when self-test bias is non-negligible.
-We also recommend investigating the multiplicative transformation for the logistic normal rather than the additive transformation which is currently the only option in WHAM [@cadigan16].
-
-?Selectivity RE are only on index age comp in the north which may be why bias is larger there?
-
-?
-We found REML provided less biased estimation of SSB and recruitment which are functions of random effects.
-Whether REML should be the default approach for estimating these quantities from state-space stock assessment models requires further investigation.
-For example, improvement in bias may be negligible when data contain low observation error and/or there are few fixed effects parameters.
-? larger variance estimates allows greater spread of random effects estimates.
-Important for conditional self-test where estimated RE are of primary importance (Recruitment and SSB).
-When simulating RE and refitting, the sign of differences between estimates and true values can be random across simulations.
-Probably more important with more fixed effects and less informative data
-
-~~Although not relevant to the best performing model, the effect of bottom temperature on age 1 M would be expected to be opposite of that for recruitment because lower M with higher temperature would produce higher recruitment.
-This expected effect was estimated only when random effects on age 1 M were also included in the model.
-This seems to provide further support to including random effects when examining covariate effects on population parameters as demonstrated by @milleretal18 for effects on growth parameters.~~
-
--->
+Similarly, a linear increase of temperature is also not likely in the future (see past trends, top row of Figure 6).
+As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
 
@@
 Therefore, extending WHAM to allow such models would be beneficial.
-Similarly age composition data are scarce for many stocks incorporating the work by @correaetal23 on allowing length composition and modeling growth within WHAM would provide an assessment tool to include process errors in assessment with such data and possibly with multiple stocks and regions.
-
-<!--
-
-## Conclusions {-}
-
-WHAM version 2.0 extends the existing R package to allow multiple stocks and discrete spatial regions.
-It allows autocorrelated random effects and environmental covariate effects on movement rates.
-Estimates of movement rates from auxiliary studies can be used to construct prior distributions for movement rates in lieu of integrated likelihoods for tagging data.
-Version 2.0 also allows seasonal treatment of different fishing fleets and movement of individual stocks, and internal estimation of SPR- and MSY-based reference points accounting for seasonal spatial dynamics including movement and mortality and any environmental covariate effects.
-
-We applied this extended version to investigate alternative hypotheses about effects of bottom temperature on recruitment and age 1 natural mortality for black sea bass.
-Our analyses indicate evidence for effects of bottom temperature on recruitment for the northern component was stronger than other models that included effects on age 1 mortality and/or corresponding effects on the southern stock component.
-
-Future WHAM development should prioritize including tagging data to inform stock assessment parameter estimation and merging in the version created by @correaetal23 to allow modeling of growth and inclusion of length and length-at-age composition observations.
-
--->
+Finally, because age composition data are scarce for many stocks, incorporating the work by @correaetal23 on allowing length composition and modeling growth within WHAM would provide an assessment tool to include process errors in assessment with such data and possibly with multiple stocks and regions.
+
+# Acknowledgements
+
+We express our gratitude to the many scientists who have collected and processed the data that went into the assessment model, including survey biologists, port samplers, observers and agers.
+We thank the black sea bass research track assessment working group and peer-reviewers for the work compiling data and determining many of the configurations of the black sea bass assessment model that we used for our analyses.
+We also thank Chris Legault for comments and suggestions that improved the clarity of this paper.
 
@@
 # Figures {-}
-
-<!--
-(ref:caption)
-
-\begin{figure}
-\begin{center}
-\includegraphics{map}
-\end{center}
-\caption{The Northwest Atlantic coastal area where the NEUS black sea bass stock occurs.
-Red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations \cite{milleretal16_yoy_survival} and the bathymetric contour is the 400 m isocline.}
-\end{figure}
-\pagebreak
--->
 
@@
 
-```{r move-prior-posterior, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Prior (black) and posterior (red) distributions of movement of northern component stock from north to south (top) and south to north (bottom)."}
+```{r move-prior-posterior, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Prior (black) and posterior (red) distributions of movement of the northern stock component from north to south (top) and south to north (bottom)."}
 knitr::include_graphics(here::here("paper", "move_prior_post.pdf"))
@@
 Many of the fits to the simulated data produced implausible estimates at the 0 boundary for these parameters (very negative values on log-scale).
-Therefore, we completed self-tests with these parameters fixed at the true values.
-
-For 7 of the the simulated data sets the model failed to optimize.
-The maximum absolute gradient was $<10^{-6}$ for only 9 and $<10^{-4}$ for 52 of the 93 successfully fitted models.
-The poor convergence appeared to be attributable to the estimation of the scalar for the standard errors of the log-transformed northern Recreational CPA index for which estimates tended to 0 for nearly all of the fits ($<0.01$ for 83 fits).
-However, even across all fits including those with poor convergence, the SSB estimates appeared to be reliable (Figure \ref{fig:self-test-fig}).
+However, across all fits including those with poor convergence, estimation of SSB and fishing mortality was reliable (Figure \ref{fig:self-test-fig}).
+We also fit analogous models with the multiplier parameters fixed at the true values, which did improve convergence, but larger bias was estimated for fishing mortality and SSB for the northern component.
 
@@
 
+```{r self-test-fig, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Median relative error of SSB (Total and by stock component) and total fully-selected fishing mortality for estimation models fitted to simulated data from model $M_1$ where the observation variance of log-indices are fixed and estimated.
+Black and Red dashed lines represent the median of the annual medians and the median across all annual relative errors, respectively."}
+knitr::include_graphics(here::here("paper", "self_test_results.pdf"))
+```
+
 \begin{landscape}
-
-```{r self-test-fig, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Median relative bias of SSB for the North and South stock components for estimation models where the observation variance of log-indices are fixed and estimation is by maximum marginal likelihood (ML) or Restricted Maximum Likelihood (REML) or where the standard deviation of logistic-normal age composition observations are assumed to be 0.1 (over sqrt(1000)) (Low Error).
-Black and Red dashed lines represent the median of the yeaerly medians and the median across all yearly relative errors, respectively."}
-knitr::include_graphics(here::here("paper", "self_test_results.pdf"))
-```
 
@@
 ```
+
+```{r selectivity-south, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Selectivty for fleets and indices in the southern region."}
+knitr::include_graphics(here::here("paper", "selectivity_south_plot.pdf"))
+```
+\pagebreak
 
@@
 
+```{r annual-F-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\\bar{F}_{40\\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
+knitr::include_graphics(here::here("paper", "proj_F_status_CV.pdf"))
+```
+
 ```{r annual-SSB-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
@@
 ```
-
-```{r annual-F-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\\bar{F}_{40\\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
-knitr::include_graphics(here::here("paper", "proj_F_status_CV.pdf"))
-```

### 2025-03-18 paper/paper.tex
add edits from Chris and a little tidying of the repo

@@
   pdftitle={Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass},
-  pdfauthor={Timothy J. Miller1; Kierstin Curti; Alexander C. Hansell},
+  pdfauthor={Timothy J. Miller1; Kiersten Curti; Alexander C. Hansell},
   hidelinks,
@@
 \title{Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass}
-\author{Timothy J. Miller\textsuperscript{1} \and Kierstin Curti \and Alexander C. Hansell}
-\date{04 March, 2025}
+\author{Timothy J. Miller\textsuperscript{1} \and Kiersten Curti \and Alexander C. Hansell}
+\date{18 March, 2025}
 
@@
 
-\hypertarget{main-message}{%
-\subsection*{Main Message}\label{main-message}}
-\addcontentsline{toc}{subsection}{Main Message}
-
-Describes the multi-region, multi-stock generalization of WHAM and its usage to evaluate evidence of alternative hypotheses about temperature effects on recruitment and age 1 natural mortality of black sea bass stock components.
+\textbf{keywords:} stock assessment, state-space, multi-stock, multi-region, age-structured, black sea bass
 
@@
 
-A state-space age-structured stock assessment models can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al.
+A state-space age-structured stock assessment model can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al.
 2016b).
@@
 Black sea bass distribution has been linked to warming waters on the northwest shelf (Bell et al.
-2015) and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare 2016).
-Thus, a focus of the research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the NEUS using most of the model assumptions that were accepted during the peer-review process.
+2015) and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare et al.
+2016).
+Thus, a focus of the black sea bass research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process.
 However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
@@
 For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
-The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by season \(i\), year \(y\), and age \(a\) on January 1.
-Each row and column of the PTM correspond to one the states: alive in region \(r\), dead in fleet \(f\), or dead from natural causes.
+The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by year \(y\), age \(a\), and season \(i\) on January 1.
+Each row and column of the PTM correspond to one of the states: alive in region \(r\), dead in fleet \(f\), or dead from natural causes.
 The probabilities in each row sum to unity (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval.
@@
 When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously.
-The sequential assumption is used widely in spatially explicit model (e.g., Stock Synthesis, Methot and Wetzel 2013).
+The sequential assumption is used widely in spatially explicit models (e.g., Stock Synthesis, Methot and Wetzel 2013).
 Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval.
@@
 When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
-We obtain the probability transition matrix over an interval \(\delta_i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
+We obtain the PTM over an interval \(\delta_i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
 \begin{equation*}
@@
 
-When there are \(n_R\) regions and \(n_F\) fleets, the vector of abundance in each state at age \(a>1\) on January 1 is \(\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}', \mathbf{0}')'\) where \(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})'\) is the number in the states corresponding to being alive in each region and \(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each fleet and dead from natural mortality because no age \(a\) fish have died yet on January 1.
+When there are \(n_R\) regions and \(n_F\) fleets, the vector of abundance in each state at age \(a>1\) on January 1 is \(\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}^T, \mathbf{0}^T)^T\) where \(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})^T\) is the number in the states corresponding to being alive in each region and \(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each fleet and dead from natural mortality because no age \(a\) fish have died yet on January 1.
 
@@
 where\\
-\[\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}(t_s)' \mathbf{N}_{O,y,a}\]
+\[\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}^T(t_s) \mathbf{N}_{O,y,a}\]
 and \(w_{y,a}\) is the mean weight at age of spawning individuals, \(\text{mat}_{y,a}\) is the maturity at age, and \(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and occurring in region \(r_s\) at time \(t_s\) given being alive in each region at the start of the year (the \(r_s\) column of the upper-left submatrix of Eq.
@@
 
-As in previous versions of WHAM, the transitions in numbers at age from one year to another after recruitment can be treated deterministically or as functions of random effects.
+As in previous versions of WHAM, the apparent survival (and movement) from one year to another after recruitment can be treated deterministically or as functions of random effects.
 The predicted numbers at age in year \(y\) at age \(a\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (Stock and Miller 2021).
 For ages \(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first \(n_R\) elements of the vector
-\[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\]
-The numbers alive in each region can also be modeled more simply using the sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions in abundance at age is
-\begin{equation*}
-\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
+\[\mathbf{P}_{y-1,a-1}^T \mathbf{N}_{y-1,a-1}.\]
+The numbers alive in each region can also be modeled more simply using the sub-matrix \(\mathbf{O}_{y,a}\).
+The general model for the vector of abundance at age in year \(y\) at age \(a\) given the vector in year \(y-1\) at age \(a-1\) is
+\begin{equation*}
+\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}^T \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
 \end{equation*}
@@
 \begin{equation*}
-\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
+\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,A-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}^T \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}^T \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
 \end{equation*}
 where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
-When there is autocorrelation with age, WHAM now assumes this applies only to ages \(a>1\) by default so that recruitment random effects are independent of those for for the annual transitions of older age classes.
+When there is autocorrelation with age, WHAM now assumes this applies only to ages \(a>1\) by default so that recruitment random effects are independent of those for the apparent survival of older age classes.
 So the general covariance structure for a given stock at ages \(a>1\) in region \(r\) is that of a two-dimensional first-order autoregressive (2DAR1) process
@@
 \end{equation*}
-When the transitions in abundance at age are treated deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\).
+When the annual changes in abundance at age are treated deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\).
 Since recruitment for a given stock currently only occurs in one region \(r_s\) there is only a single vector of time-varying recruitment random effects for each stock.
@@
 Recall that recruitment for stock \(s\) only occur in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
-\[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}' \mathbf{N}_{O,1,1}.\]
+\[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}^T \mathbf{N}_{O,1,1}.\]
 
@@
 
-For each season, there are at most \(n_R-1\) parameters determining movement among regions given starting the season in region \(r\) in either the sequential or simultaneous configurations.
+For each season, there are at most \(n_R-1\) parameters determining movement for a given stock among regions given starting the season in region \(r\) in either the sequential or simultaneous configurations.
 Movement parameters are estimated on a transformed scale via a link function \(g(\cdot)\).
@@
 \begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
+  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y}.
   \end{equation*}
@@
 
-When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific.
+When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific, but currently are constant across seasons.
 When random effects are used, the same 2DAR1 structure with age and year as described by Stock and Miller (2021) can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
@@
 
-The log-normal distributional assumption for aggregate catch observations is the same as Stock and Miller (2021), but the predicted catch is now a function of catch from each stock starting the year in each region.
-For a given stock and age, the numbers captured in each fleet over the year are
-\[\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}' \mathbf{N}_{O,s,y,a}\]
-The predicted numbers caught be each fleet across stocks is
+The log-normal distributional assumption for aggregate catch observations is the same as Stock and Miller (2021), but the predicted catch is now a function of catch from each stock starting the year in each region and any seasonal movement.
+For a given stock and age, the vector of numbers captured by each fleet over the year are
+\[\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}^T \mathbf{N}_{O,s,y,a}\]
+(recall each fleet can harvest in only one region as described above). The vector of predicted numbers caught be each fleet across stocks is
 \[\widehat{\mathbf{N}}_{H,y,a} = \sum^{n_S}_{s=1} \widehat{\mathbf{N}}_{H,s,y,a}\]
-and the predicted aggregate catch at age \(a\) is
+and the vector of predicted aggregate catch by each fleet at age \(a\) is
 \[\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \widehat{\mathbf{N}}_{H,y,a}\]
 where \(\mathbf{c}_{y,a}\) is the vector of mean individual weight at age \(a\) for each fleet and the aggregate catch by fleet is
-\[\widehat{\mathbf{C}}_y = \sum^{A}_{a=1} \widehat{\mathbf{C}}_{y,a}\]
+\[\widehat{\mathbf{C}}_y = \sum^{A}_{a=1} \widehat{\mathbf{C}}_{y,a}.\]
 The log-aggregate catch observations for fleet \(f\) are normally distributed
@@
 
-For index \(m\) occurring in region \(r_m\) at fraction of the year \(t_m\), the predicted abundance at \(t_m\) in region \(r_m\) is
-\[\widehat{N}_{s,y,a,m} = \mathbf{O}_{s,y,a,r_m}(t_m)' \mathbf{N}_{O,s,y,a}\]
+Like fishing fleets, an index \(m\) can observe the population in a single region \(r_m\) at fraction of the year \(t_m\), the predicted abundance at \(t_m\) in region \(r_m\) is
+\[\widehat{N}_{s,y,a,m} = \mathbf{O}_{s,y,a,r_m}^T(t_m) \mathbf{N}_{O,s,y,a}\]
 where \(\mathbf{O}_{s,y,a,r_m}(t_m)\), the \(r_m\) column of the upper-left submatrix of Eq.
@@
 
-As in previous versions of WHAM package, reference points for prevailing conditions, average all of the inputs to the spawning biomass and yield per recruit calculations (weight, maturity, natural mortality, and movement) over a set of user-specified years as shown in eq.
+As in previous versions of the WHAM package, reference points for prevailing conditions, average all of the inputs to the spawning biomass and yield per recruit calculations (weight, maturity, natural mortality, and movement) over a set of user-specified years as shown in eq.
 \ref{eq:spr_sel} for selectivity (e.g., last 5 years of the model).
 
-For \(X\%\) SPR-based reference points, we use a Newton method and iterate
+For \(X\%\) SPR-based reference points, we use a Newton method and iterate for steps \(j = 1,\ldots,10\)
 \begin{equation}\label{eq:newton}
-  \log\widetilde{F}^{(i)} = \log\widetilde{F}^{(i-1)} - \frac{g\left(\log\widetilde{F}^{(i-1)}\right)}{g'\left(\log\widetilde{F}^{(i-1)}\right)}
+  \log\widetilde{F}^{j} = \log\widetilde{F}^{j-1} - \frac{g\left(\log\widetilde{F}^{j-1}\right)}{g^{(1)}\left(\log\widetilde{F}^{j-1}\right)}
 \end{equation}
@@
 where \(\upphi_s\left(F=0\right)\) is the equilibrium unfished spawning biomass per recruit.
-\(g'(\log F)\) is the derivative of \(g\) with respect to \(\log F\), and the weights to use for each stock \(\lambda_s\) can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate ``static'\,' equilibrium spawning biomass and yield.
+\(g^{(1)}(\log F)\) is the derivative of \(g\) with respect to \(\log F\), and the weights to use for each stock \(\lambda_s\) can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate ``static'' equilibrium spawning biomass and yield.
 
@@
 \ref{eq:ypr}) and equilibrium recruitment.
-The function \(g(\log F)\) in Eq.
-\ref{eq:newton} is the first derivative of the yield curve with respect to \(\log F\).
+The function \(g(\log F)\) and \(g^{(1)}(\log F)\) in Eq.
+\ref{eq:newton} are the first and second derivatives of the sum of the stock-specific equilibrium yields with respect to \(\log F\) (analogous to that described by Miller et al.
+2016b).
 
@@
 Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging from the Gulf of Mexico to the Gulf of Maine.
-Fish north of Cape Hatteras, NC, are considered to comprise a single stock unit, but individual populations exhibit spawning site fidelity
-(Able and Stanton Hales, Jr.
+Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity (Able and Stanton Hales, Jr.
 1997; Fabrizio et al.
@@
 Estimating initial numbers at age as separate parameters can be challenging even in single-stock models, but for black sea bass the available data cannot distinguish the proportion of northern and southern component fish at each age in the southern region in the initial year of the model.
-Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component: an initial recruitment and an equilibrium fully-selected \(F\) that determines the abundance at age in each region for each stock component.
-
-For subsequent years, all models allow AR1 correlation of recruitment and 2DAR1 correlation for survival for both the northern and southern stock components.
+Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component: an initial recruitment and an equilibrium fully-selected \(F\) (across all fleets) that determines the initial abundance at age in each region for each stock component.
+
+For subsequent years, all models allow AR1 correlation of recruitment and 2DAR1 correlation for apparent survival random effects for both the northern and southern stock components.
 Models with apparent survival random effects will model the transitions (conditional on both survival and movement) of abundances at age of northern origin fish in each region, but there appears to be little ability of the existing data from the southern region to distinguish apparent survival random effects for each stock component.
-All models we considered assume very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model.
+All models we considered assume a very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model.
 Unique variance and correlation parameters for the recruitment and apparent survival random effects are estimated for the northern and southern components.
@@
 2016a).
-This covariate may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of of the year after becoming age 1.
+This covariate may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of the year after becoming age 1.
 Furthermore, we have no direct observations of age 1 individuals from surveys until the spring season each year.
@@
 
-We assume the covariate in year \(y\) affects recruitment in the same years because the covariate observations are from months January to March.
+We assume the covariate in year \(y\) affects recruitment in the same year because the covariate observations are from months January to March.
 The fish are technically already 1 year old, but there are no observations of these individuals until later in the year except possibly in fishery catches which are accumulated over the whole year.
-Expected log-recruitment for a given stock is a linear function of bottom temperature
+Expected log-recruitment for a given stock component is a linear function of bottom temperature
 
@@
 \begin{equation}\label{eq:expected-M1}
-E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y
+E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y.
 \end{equation}
-Because age 1 fish for the northern component can exist in both regions after January 1, natural mortality is acting in both regions for this stock.
+Because age 1 fish for the northern component can exist in both regions after January 1 other the spawning period, natural mortality is acting in both regions for this stock.
 For models with covariate and(or) annual random effects for age 1 fish we assume them only in the northern regions for the northern component.
@@
 We used a development version (\href{https://github.com/timjmiller/wham/tree/fb8b089}{commit fb8b089}) of the WHAM package prior to the release of version 2.0 for all results.
-All code for fitting models and generating results can be found at \href{https://github.com/timjmiller/wham_bsb_paper}{github.com/timjmiller/wham\_bsb\_paper}.
+All code for fitting models and generating results can be found at \href{https://github.com/timjmiller/multi_wham_bsb}{github.com/timjmiller/multi\_wham\_bsb}.
 
@@
 Finally, we calculated ``prevailing'' BRPs using the average weight at age, maturity at age, M, fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
-All BRPs and status estimates are generated for each region as well as for the total stock.
+All BRPs and status estimates are generated for each region as well as for the total stock area.
 
@@
 
-We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock and no random effects on age 1 natural mortality (\(M_1\), Tables \ref{tab:diff-aic-table} and \ref{tab:aic-wts-table}).
+We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock component and no random effects on age 1 natural mortality (\(M_1\), Tables \ref{tab:diff-aic-table} and \ref{tab:aic-wts-table}).
 Model \(M_1\) consistently had the lowest AIC across retrospective peels where the terminal years of data are removed from the model.
@@
 Comparisons of estimates from retrospective peels did not show any indication of differences in model performance.
-There was very little variation in measures of retrospective patterns for SSB or fishing mortality across models.
+There was little variation in measures of retrospective patterns for SSB or fishing mortality across models.
 Mohn's \(\rho \approx -0.03\) for SSB of both stock components and Mohn's \(\rho \approx 0.03\) and \(\approx -0.04\) for average \(F\) at ages 6 and 7 in north and south regions, respectively (Table \ref{tab:rho-table}).
@@
 
-Differences between annual \(F_{40\%}\) estimates using the alternative types of recruitment estimates are much less than those for SSB\(_{40\%}\) because the recruitment estimates are only used in the weighting of the stock-specific \(\upphi(\widetilde{F})\) to determine \(F_{40\%}\) (\(\lambda_s\) in Eq.
+Differences between annual \(F_{40\%}\) estimates using the alternative types of recruitment estimates are much less than those for SSB\(_{40\%}\) because the recruitment estimates are only used in the weighting of the stock component-specific \(\upphi(\widetilde{F})\) to determine \(F_{40\%}\) (\(\lambda_s\) in Eq.
 \ref{eq:newton-obj}, Figure \ref{fig:annual-BRPs}: third row).
 For similar reasons, estimates of uncertainty of \(F_{40\%}\) are also similar between the type of recruitment estimates (Figure \ref{fig:annual-F40-cvs}).
-Annual estimates for the northern stock component were lowest in the earliest years of the time series whereas those for the soutern component are relatively stable over time.
+Annual estimates for the northern stock component were lowest in the earliest years of the time series whereas those for the southern component are relatively stable over time.
 
@@
 Opposite effects of temperature on recruitment or population size would be expected for species in the same general area that are at the southern extent of their range (Gabriel 1992).
-For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer fall on the Northeast US shelf (Sullivan et al.
+For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer and fall on the Northeast US shelf (Sullivan et al.
 2005; Miller et al.
@@
 In the projection period, the AR1 models for bottom temperature and recuitment revert to the long-term means of those processes.
-This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypothesis of increasing water temperature and changing environmental conditions (Hare 2016; Pershing et al.
+This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypothesis of increasing water temperature and changing environmental conditions (Hare et al.
+2016; Pershing et al.
 2021).
@@
 Unfortunately, this methodology still has limitations because it is unlikely that current environmental conditions are stable.
-Similarly, a linear increase of temperature is also not likely in the future (see past trends, Figure 6 top row).As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
+Similarly, a linear increase of temperature is also not likely in the future (see past trends, top row of Figure 6).
+As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
 
@@
 Therefore, extending WHAM to allow such models would be beneficial.
-Similarly age composition data are scarce for many stocks incorporating the work by Correa et al.
+Finally, because age composition data are scarce for many stocks, incorporating the work by Correa et al.
 (2023) on allowing length composition and modeling growth within WHAM would provide an assessment tool to include process errors in assessment with such data and possibly with multiple stocks and regions.
+
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+We express our gratitude to the many scientists who have collected and processed the data that went into the assessment model, including survey biologists, port samplers, observers and agers.
+We thank the black sea bass research track assessment working group and peer-reviewers for the work compiling data and determining many of the configurations of the black sea bass assessment model that we used for our analyses.
+We also thank Chris Legault for comments and suggestions that improved the clarity of this paper.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hareetal16}{}}%
-Hare, W.E.A.N., Jonathan A.
-AND Morrison.
+Hare, J.A., Morrison, W.E., Nelson, M.W., Stachura, M.M., Teeters, E.J., Griffis, R.B., Alexander, M.A., Scott, J.D., Alade, L., Bell, R.J., Chute, A.S., Curti, K.L., Curtis, T.H., Kircheis, D., Kocik, J.F., Lucey, S.M., McCandless, C.T., Milke, L.M., Richardson, D.E., Robillard, E., Walsh, H.J., McManus, M.C., Marancik, K.E., and Griswold, C.A.
 2016.
-A vulnerability assessment of fish and invertebrates to climate change on the northeast u.s.
+A vulnerability assessment of fish and invertebrates to climate change on the northeast {U}.{S}.
 Continental shelf.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kemenysnell60}{}}%
-Kemeny, J.G., and Snell, J.L. 1960. Finite markov chains. D. Van Nostrand Company, Princeton, New Jersey.
+Kemeny, J.G., and Snell, J.L. 1960. Finite {M}arkov chains. D. Van Nostrand Company, Princeton, New Jersey.
 
@@
 2023.
-The multivariate-tweedie: A self-weighting likelihood for age and length composition data arising from hierarchical sampling designs.
+The multivariate-{T}weedie: A self-weighting likelihood for age and length composition data arising from hierarchical sampling designs.
 ICES Journal of Marine Science \textbf{80}(10): 2630--2641.
@@
 2022.
-A state-space model for stock assessment of cod (gadus morhua) stock in NAFO subdivision 3Ps.
+A state-space model for stock assessment of cod (\emph{{G}adus morhua}) stock in NAFO subdivision 3Ps.
 {DFO} {Can.} {Sci.} {Advis.} {Sec.} {Res.} {Doc.} 2022/022.
@@
 
-\caption{Prior (black) and posterior (red) distributions of movement of northern component stock from north to south (top) and south to north (bottom).}\label{fig:move-prior-posterior}
+\caption{Prior (black) and posterior (red) distributions of movement of the northern stock component from north to south (top) and south to north (bottom).}\label{fig:move-prior-posterior}
 \end{figure}
@@
 Many of the fits to the simulated data produced implausible estimates at the 0 boundary for these parameters (very negative values on log-scale).
-Therefore, we completed self-tests with these parameters fixed at the true values.
-
-For 7 of the the simulated data sets the model failed to optimize.
-The maximum absolute gradient was \(<10^{-6}\) for only 9 and \(<10^{-4}\) for 52 of the 93 successfully fitted models.
-The poor convergence appeared to be attributable to the estimation of the scalar for the standard errors of the log-transformed northern Recreational CPA index for which estimates tended to 0 for nearly all of the fits (\(<0.01\) for 83 fits).
-However, even across all fits including those with poor convergence, the SSB estimates appeared to be reliable (Figure \ref{fig:self-test-fig}).
+However, across all fits including those with poor convergence, estimation of SSB and fishing mortality was reliable (Figure \ref{fig:self-test-fig}).
+We also fit analogous models with the multiplier parameters fixed at the true values, which did improve convergence, but larger bias was estimated for fishing mortality and SSB for the northern component.
 
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{self_test_results}
+
+}
+
+\caption{Median relative error of SSB (Total and by stock component) and total fully-selected fishing mortality for estimation models fitted to simulated data from model $M_1$ where the observation variance of log-indices are fixed and estimated.
+Black and Red dashed lines represent the median of the annual medians and the median across all annual relative errors, respectively.}\label{fig:self-test-fig}
+\end{figure}
+
 \begin{landscape}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{self_test_results}
-
-}
-
-\caption{Median relative bias of SSB for the North and South stock components for estimation models where the observation variance of log-indices are fixed and estimation is by maximum marginal likelihood (ML) or Restricted Maximum Likelihood (REML) or where the standard deviation of logistic-normal age composition observations are assumed to be 0.1 (over sqrt(1000)) (Low Error).
-Black and Red dashed lines represent the median of the yeaerly medians and the median across all yearly relative errors, respectively.}\label{fig:self-test-fig}
-\end{figure}
 
@@
 \end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{selectivity_south_plot}
+
+}
+
+\caption{Selectivty for fleets and indices in the southern region.}\label{fig:selectivity-south}
+\end{figure}
+\pagebreak
 
@@
 
+{\centering \includegraphics[height=0.95\textheight]{proj_F_status_CV}
+
+}
+
+\caption{Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\bar{F}_{40\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-F-status-cvs}
+\end{figure}
+
+\begin{figure}
+
 {\centering \includegraphics[height=0.95\textheight]{proj_SSB_status_CV}
@@
 
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_F_status_CV}
-
-}
-
-\caption{Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\bar{F}_{40\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-F-status-cvs}
-\end{figure}
-
 \end{document}

### 2025-03-18 R/fits_rev.R

@@ -21,7 +21,8 @@ input_0 <- prepare_wham_input(asap, selectivity = sel, NAA_re = NAA_re, basic_in
 
 fit_0 <- fit_wham(input_0, do.brps = FALSE, do.sdrep = FALSE, do.osa = FALSE, do.retro = FALSE)
 saveRDS(fit_0, here::here("results", "fit_0_rev.RDS"))
-# fit_0 <- readRDS(here::here("results", "fit_0_rev.RDS"))
+
+#Do jitter for null model
 res_dir <- here::here("results","jitter", "fit_0")
 dir.create(res_dir)
 fit_0_jit_res <- jitter_wham(fit_RDS = here::here("results","fit_0_rev.RDS"), n_jitter = 50, res_dir = res_dir)
@@ -35,9 +36,6 @@ saveRDS(fit_0_jit_res, here::here(res_dir, "fit_0_jitter_results.RDS"))
 # initial_vals <- t(sapply(1:50, function(x) fit_0$opt$par + chol.L %*% cbind(rnorm(n= NCOL(cov)))))
 ####################
 
-jitters <- fit_0_jit_res[[1]]
-sort(sapply(jitters, \(x) x$obj))
-y <- order(sapply(jitters, \(x) x$obj))
 
 input_0_better <- fit_0$input
 input_0_better$par <- fit_0$env$parList(jitters[[13]]$par)
@@ -63,6 +61,10 @@ input_1$par <- fit_0$parList
 fit_1 <- fit_wham(input_1, do.brps = FALSE, do.sdrep = FALSE, do.osa = FALSE, do.retro = FALSE)
 saveRDS(fit_1, here::here("results", "fit_1.RDS"))
 
+res_dir <- here::here("results","jitter", "fit_1")
+dir.create(res_dir)
+fit_1_jit_res <- jitter_wham(fit_RDS = here::here("results","fit_1.RDS"), n_jitter = 50, res_dir = res_dir)
+saveRDS(fit_1_jit_res, here::here(res_dir, "fit_1_jitter_results.RDS"))
 
 for(i in 2:6){
 	assign(paste0("input_",i), prepare_wham_input(asap, selectivity = sel, NAA_re = NAA_re, basic_info = basic_info, move = move, ecov = get(paste0("ecov_", i)), catch_info = catch_info,
@@ -83,18 +85,7 @@ fits <- lapply(fits, make_osa_residuals)
[…]

### 2025-03-18 R/make_figures.R

@@ -1,71 +1,118 @@
+# load the appropriate R package libraries
+library(rnaturalearth)
+library(maps)
+library(sp)
+library(sf)
 library(wham)
 library(ggplot2)
 library(dplyr)
 library(patchwork)
+# load some helper functions
 source(here::here("R","plot_functions.R"))
 
-theme_set(theme_bw())
-theme_update(strip.background = element_blank(), strip.placement = "outside", strip.text = element_text(size = rel(2)), 
-      axis.title = element_text(size = rel(2)), axis.text = element_text(size = rel(2)), legend.text = element_text(size = rel(2)), #text = element_text(size = rel(2)), 
-      legend.title = element_text(size = rel(2)))
-
+#load fits
 fits <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))
 fits_M_re <- readRDS(file.path("results","fits_M_re_better.RDS"))
 
+##############################################################################################
 
 
-res_dir <- here::here("results","jitter", "fit_0")
-fit_0 <- readRDS(here::here("results", "fit_0_rev.RDS"))
-fit_0_jit_res <- readRDS(here::here(res_dir, "fit_0_jitter_results.RDS"))
-# jitter_files <- dir(here::here("results","jitter","fit_0"), full.names = TRUE, pattern = glob2rx("jitter_sim*.RDS"))
-# jitters <- lapply(fit_0_jit_res, \(x) readRDS(x))
-jitters <- fit_0_jit_res[[1]]
-sort(sapply(jitters, \(x) x$obj))
-y <- order(sapply(jitters, \(x) x$obj))
-# sapply(jitters, \(x) max(abs(x$grad)))[y]
-cbind(round(jitters[[y[3]]]$par,2), round(jitters[[y[1]]]$par - jitters[[y[3]]]$par,2))
+#Fig 1.
+#hudson
+#hudson.n_UTM <- data.frame(x=c(357212.46,19314.01,658471.31,935303.78),y=c(4261346,4566676,5099985,4635884))
+spRef_UTM_19 <- "+proj=utm +zone=19 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
[…]

### 2025-03-18 R/plot_Ecov_R_SSB_F.R

@@ -64,9 +64,12 @@ p_E_full <- ggplot(this_df, aes(x = Year, y = val, colour = name, fill = name)) 
     geom_ribbon(data =  subset(this_df, name == "Random Effect"), aes(ymin=lo, ymax=hi, fill = name), alpha=0.4, linetype = 0) 
 p_E_full
 
+###########################################################################################
+# Fig. S4
 cairo_pdf(file.path("paper", "BTA_full_fig.pdf"), width = 20, height = 10)
 p_E_full
 dev.off()
+###########################################################################################
 
 this_df <- subset(df_use, Year > 1988 & Year < 2022 & name %in% c("Ecov_obs", "Ecov_RE"))
 this_df$name <- c("Random Effect", "Observed")[match(this_df$name, c("Ecov_RE","Ecov_obs"))]
@@ -132,6 +135,8 @@ plt_F <- ggplot(this_df, aes(x = Year, y = val)) +
 plt_F
 
 
+###########################################################################################
+# Fig. 4
 cairo_pdf(file.path("paper", "E_R_SSB_F_fig.pdf"), width = 20, height = 16)
 design <- c(area(1,1,1,778), area(2,1,2,778), area(3,1,3,1000), area(4,1,4,1000))
 (p_E +  xlab("")  + theme(axis.title.x = element_blank(), strip.text=element_text(margin=margin(b = 5)), axis.text.x=element_blank(), plot.margin = margin(b = 1, t = 0))) + 
@@ -140,3 +145,4 @@ design <- c(area(1,1,1,778), area(2,1,2,778), area(3,1,3,1000), area(4,1,4,1000)
   (plt_F + theme(legend.position="none", strip.clip = "off", strip.text = element_text(margin=margin(t=-15)), plot.margin = margin(b = 1, t = 0))) + 
   plot_layout(design = design)
 dev.off()
+###########################################################################################

### 2025-03-18 R/plot_brps_status.R

@@ -164,6 +164,7 @@ theme_update(strip.background = element_blank(), strip.placement = "outside", st
       axis.title = element_text(size = rel(2)), axis.text = element_text(size = rel(2)), legend.text = element_text(size = rel(2)), #text = element_text(size = rel(2)), 
       legend.title = element_text(size = rel(2)))
 
+#Fig 6
 cairo_pdf(file.path("paper", "brp_status_results.pdf"), width = 15, height = 20)
 design <- c(area(1,1,1,1), area(2,1,2,1), area(3,1,3,1), area(4,1,4,1), area(5,1,5,1))
 	(plt_SSB40 +  xlab("")  + theme(legend.position="none", axis.title.x = element_blank(), strip.text=element_text(margin=margin(b = 5)), axis.text.x=element_blank(), plot.margin = margin(b = 1, t = 0))) + 

### 2025-03-18 R/plot_projections.R

@@ -204,6 +204,7 @@ R_Exp_proj <- ggplot(this_df, aes(x = Year, y = val)) + xlab("Year") +
     geom_ribbon(data = north_proj_df, aes(ymin = lo, ymax = hi, fill = Proj_type), alpha=0.4, linetype = 0)
 R_Exp_proj
 
+#Fig. 7
 cairo_pdf(file.path("paper", "proj_ecov_Recruit_results.pdf"), width = 20, height = 20)
 design <- c(area(1,1,1,572), area(2,1,2,1000), area(3,1,3,1000))
 (BTA_proj +  xlab("")  + theme(axis.title.x = element_blank(), strip.text=element_text(margin=margin(b = 5)), axis.text.x=element_blank(), plot.margin = margin(b = 1, t = 0))) + 
@@ -256,6 +257,8 @@ SSB_proj <- ggplot(this_df, aes(x = Year, y = val)) + xlab("Year") +
     geom_line(proj_df, mapping = aes(x = Year, y = val, colour = Proj_type)) + 
     geom_ribbon(proj_df, mapping = aes(x = Year, y = val, ymin = lo, ymax = hi, fill = Proj_type), alpha=0.4, linetype = 0)
 SSB_proj
+
+#Fig. 8
 
 cairo_pdf(file.path("paper", "proj_F_SSB.pdf"), width = 20, height = 20)
 design <- c(area(1,1,1,1), area(2,1,2,1))
@@ -310,6 +313,7 @@ cairo_pdf(file.path("paper", "proj_F40_results.pdf"), width = 20, height = 20)
 F40_proj
 dev.off()
 
+#Fig. 9
 cairo_pdf(file.path("paper", "proj_brps.pdf"), width = 24, height = 16)
 design <- c(area(1,1,1,1), area(1,2,1,2))
 (F40_proj +  xlab("Year")  + theme(legend.position="none", axis.text.x=element_text(size = rel(0.7)))) + 
@@ -336,9 +340,12 @@ SSB_status_proj <- ggplot(this_df, aes(x = Year, y = val)) + xlab("Year") +
     geom_ribbon(proj_df, mapping = aes(x = Year, y = val, ymin = lo, ymax = hi, fill = Proj_type), alpha=0.4, linetype = 0)
 SSB_status_proj
 
+################################################################
+#Fig. S13
 cairo_pdf(file.path("paper", "proj_SSB_status_results.pdf"), width = 20, height = 20)
 SSB_status_proj
 dev.off()
+################################################################
 
 
 
[…]

### 2025-03-18 R/self_tests.R

@@ -1,37 +1,43 @@
 
 library(wham)
 #seeds <- 1:10
+fit_1 <- readRDS(here::here("results","fit_1.RDS"))
 seeds <- sample(0:1e9, 100, replace = TRUE)
 res_dir <- here::here("results","self_test", "fit_1")
 dir.create(res_dir)
-fit_1 <- readRDS(here::here("results","fit_1.RDS"))
 self_test_1 <- self_test(fit_RDS = here::here("results","fit_1.RDS"), n = 50, seeds = seeds, res_dir = res_dir)
 saveRDS(self_test_1, file = file.path(res_dir, "self_test_results.RDS"))
 
+self_test_1 <- readRDS(file = file.path(res_dir, "self_test_results.RDS"))
+
 test_files <- dir(here::here("results","self_test","fit_1"), full.names = TRUE, pattern = glob2rx("cond_sim*.RDS"))
 sims <- lapply(test_files, \(x) readRDS(x))
+
 # sims <- self_test_1[[1]]
 
 sd_scale_pars <- t(sapply(sims, \(x) x$par[names(x$par) =="log_index_sig_scale"]))
 
-use <- which(sapply(sims, \(x) !(is.na(max(abs(x$grad))) | max(abs(x$grad))>0.01)))
+# use <- which(sapply(sims, \(x) !(is.na(max(abs(x$grad))) | max(abs(x$grad))>0.01)))
+use <- 1:length(sims)
 
-resid_SSB_N <- sapply(sims[use], \(x) x$SSB[,1]/fits[[2]]$rep$SSB[,1] - 1)
+par(mfrow= c(2,2))
+resid_SSB_N <- sapply(sims[use], \(x) x$SSB[,1]/fit_1$rep$SSB[,1] - 1)
 med_res_SSB_N <- apply(resid_SSB_N, 1, median, na.rm = TRUE)
-plot(fits[[2]]$years, med_res_SSB_N)
+plot(fit_1$years, med_res_SSB_N)
 abline(h = 0)
 abline(h = median(med_res_SSB_N, na.rm = TRUE), lty =2)
 abline(h = median(resid_SSB_N, na.rm = TRUE), lty =2, col = "red")
 
-resid_SSB_S <- sapply(sims[use], \(x) x$SSB[,2]/fits[[2]]$rep$SSB[,2] - 1)
[…]

### 2025-03-19 paper/paper.Rmd
saving edits from co-authors

@@
   - Timothy J. Miller^1^
-  - Kiersten Curti
+  - Kiersten L. Curti
   - Alexander C. Hansell
@@
 We first describe an extension of the model allowing any number of stocks (or stock components) and regions with movement among regions as well as seasonal variation in stock and fleet dynamics.
-Movement rates can be a be functions of time- and age-varying random effects and environmental covariates.
+Movement rates can be functions of time- and age-varying random effects and environmental covariates.
 We then illustrate the model by applying it to data for the northern and southern components of the Northeast United States black sea bass stock and evaluate alternative hypotheses of bottom temperature and time-varying random effects on recruitment and natural mortality.
@@
 
-A state-space age-structured stock assessment model can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures [@nielsenberg14;@cadigan16;@milleretal16].
+State-space age-structured stock assessment models can be used to estimate time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures [@nielsenberg14;@cadigan16;@milleretal16].
 This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks [@puntetal20].
@@
 
-WHAM is an R package developed and maintained at NOAA's Northeast Fisheries Science Center [https://timjmiller.github.io/wham].
+WHAM is an R package developed and maintained by scientists at NOAA's Northeast Fisheries Science Center (https://timjmiller.github.io/wham).
 WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and effects of environmental covariates on various population parameters.
@@
 We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) [@nefsc23].
-Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age assessment.
+Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age model with results subsequently combined to inform management.
 The stock was split into two regions to account for spatial dynamics and improve model diagnostics [@asmfc16].
@@
 
-When modeling multiple stock components occurring in different regions further dimensionality is added to the class of possible equilbrium reference points [e.g., @kapuretal21].
+When modeling multiple stock components occurring in different regions, further dimensionality is added to the class of possible equilbrium reference points [e.g., @kapuretal21].
 Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises.
-Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points.
+Finally, the inclusion of explicit effects of environmental covariates in the model also requires assumptions on how they are treated in projections and reference points.
 We demonstrate some of the options of WHAM version 2.0 to provide biological reference points and short-term projections.
@@
 Approaches to modeling transitions among these states may treat time discretely [e.g.,@arnason72;@schwarzetal93] or continuously [e.g., @hearnetal87;@commenges99; @andersenkeiding02].
-We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval $\delta$ (columns), conditional on being in each of those state at beginning of the interval (rows).
+We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval $\delta$ (columns), conditional on being in each of those states at the beginning of the interval (rows).
 For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
@@
 
-WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would be multinomial distributed as in a model for tagging data where fates of individual tagged fish are assumed independent.
+WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would have a multinomial distribution as in a model for tagging data where fates of individual tagged fish are assumed independent.
 The PTMs determine the expected numbers 1) in each state on January 1 of year $t+1$ at age $a+1$ given the abundances at age $a$ on January 1 of year $t$, 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.
@@
 where $\boldsymbol{\varepsilon}_{y,a}$ is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
+
 When there is autocorrelation with age, WHAM now assumes this applies only to ages $a>1$ by default so that recruitment random effects are independent of those for the apparent survival of older age classes.
@@
 For the plus group $a=A$, $\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}$ is a ``fundamental matrix'' derived using the matrix version of the  geometric series [@kemenysnell60].
-Recall that recruitment for stock $s$ only occur in region $r_s$ so, the equilibrium initial numbers at age $a$ by region are
+Recall that recruitment for stock $s$ only occurs in region $r_s$ so, the equilibrium initial numbers at age $a$ by region are
 $$\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}^T \mathbf{N}_{O,1,1}.$$
@@
 $$\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}^T \mathbf{N}_{O,s,y,a}$$
-(recall each fleet can harvest in only one region as described above). The vector of predicted numbers caught be each fleet across stocks is 
+(recall each fleet can harvest in only one region as described above). The vector of predicted numbers caught by each fleet across stocks is 
 $$\widehat{\mathbf{N}}_{H,y,a} = \sum^{n_S}_{s=1} \widehat{\mathbf{N}}_{H,s,y,a}$$
@@
 
-Currently a single fishing mortality reference point $\widetilde F$ is estimated across stocks and regions and the fleet- and age-specific reference points is $\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}$.
+Currently a single fishing mortality reference point $\widetilde F$ is estimated across stocks and regions and the fleet- and age-specific reference point is $\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}$.
 Selectivity is determined across fleets and ages as in the equilibrium assumption for initial abundance at age except here it is a function of average F at age over a user-defined set of years
@@
 
-When a Beverton-Holt or Ricker stock recruit relationship is assumed, an analogous Newton method is used to find $\log F$ that maximizes yield for MSY-based reference points.
-which are also a functions of the equilibrium yield per recruit (Eq.
+When a Beverton-Holt or Ricker stock recruit relationship is assumed, an analogous Newton method is used to find $\log F$ that maximizes yield for MSY-based reference points, which are also a functions of the equilibrium yield per recruit (Eq.
 \ref{eq:ypr}) and equilibrium recruitment.
@@
 
-Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging from the Gulf of Mexico to the Gulf of Maine.
+Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging the entire east coast of the United States.
 Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity [@ablehales97;@fabrizioetal13].
@@
 Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as "Working Group") composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above [@nefsc23].
-This assessment included the spatial features and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
+This assessment included the two regions and two stock components and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
 Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.
@@
 For the northern component, a proportion $\mu_{\text{N}\rightarrow \text{S}}$ can move to the southern region each month during the last five months of the year, but no movement is allows from the south to the north during this period (Figure \ref{fig:migration-diagram}).
-During the first four intervals of the year a proportion $\mu_{\text{S}\rightarrow \text{N}}$ the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
+During the first four intervals of the year a proportion $\mu_{\text{S}\rightarrow \text{N}}$ of the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
 In the fifth interval (May), all northern component individuals remaining in the south are assumed to move back to the north for the subsequent spawning period.
@@
 
-The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual patterns and retrospective patterns.
-These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for six sets of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
+The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residualand retrospective patterns.
+These analyses resulted in use of selectivity random effects for the northern fleet and indices, logistic-normal likelihoods for six sets of age composition observations, and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
 
@@
 We fit 14 models with alternative assumptions about the effects of bottom temperature covariates, ranging from no effects to effects on both regions for either recruitment or natural mortality at age 1 (Table \ref{tab:model-desc-table}).
-These analyses derive from the hypothesis that bottom temperature affects overwinter survival of fish where the fish turn from age 0 to age 1 on January 1 [@milleretal16_yoy_survival].
+These analyses derive from the hypothesis that bottom temperature affects survival over the first winter where fish turn from age 0 to age 1 on January 1 [@milleretal16_yoy_survival].
 This covariate may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of the year after becoming age 1.
@@
 All models treat annual recruitment as AR1 time-varying random effects with mean parameters for each stock component.
-However, @milleretal18 showed how inferences of temperature effects on growth or maturity parameters can be very different whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
+However, @milleretal18 showed how inferences of temperature effects on growth or maturity parameters can be very different depending on whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
 Therefore, we also explored whether including temporal random effects on age 1 natural mortality affected inferences on corresponding temperature effects.
@@
 We assume the covariate in year $y$ affects recruitment in the same year because the covariate observations are from months January to March.
-The fish are technically already 1 year old, but there are no observations of these individuals until later in the year except possibly in fishery catches which are accumulated over the whole year.
+The fish are technically already 1 year old, but there are no observations of these individuals until later in the spring except possibly in fishery catches which are accumulated over the whole year.
 Expected log-recruitment for a given stock component is a linear function of bottom temperature
@@
 Because age 1 fish for the northern component can exist in both regions after January 1 other the spawning period, natural mortality is acting in both regions for this stock.
-For models with covariate and(or) annual random effects for age 1 fish we assume them only in the northern regions for the northern component.
+For models with covariate and(or) annual random effects for age 1 fish we assume them only in the northern region for the northern component.
 The corresponding random effects are
@@
 We also calculated status as the ratios of annual average F and SSB to the corresponding reference points.
-Finally, we calculated "prevailing" BRPs using the average weight at age, maturity at age, M, fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
+Finally, we calculated "prevailing" BRPs using the average weight at age, maturity at age, M, and fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
 All BRPs and status estimates are generated for each region as well as for the total stock area.
@@
 There was little evidence for also including random effects on age 1 natural mortality for the northern stock component.
-However, when more than three terminal years of data were removed, the evidence for adding age 1 natural mortality random effects for the northern component ($M_8$) increased is similar to adding bottom temperature effects on southern component recruitment ($M_3$).
+However, when more than three terminal years of data were removed, the evidence for adding age 1 natural mortality random effects for the northern component ($M_8$) increased similar to adding bottom temperature effects on southern component recruitment ($M_3$).
 
@@
 Comparing projected recruitment for the northern and southern stock components, the projected expected recruitment has much lower uncertainty than the projected random effects regardless of whether bottom temperature affects recruitment.
-For the northern component, CVs of projected expected recruitment are smaller than those of recruitment random effect, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for the future bottom temperature anomalies are made (Figure \ref{fig:R-F-SSB-CVs}).
+For the northern component, CVs of projected expected recruitments are smaller than those of recruitment random effects, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for the future bottom temperature anomalies are made (Figure \ref{fig:R-F-SSB-CVs}).
 
@@
 
-For black sea bass, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figure 6).
+For black sea bass, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figures \ref{fig:R-BT-proj} and \ref{fig:F-SSB-proj}).
 In the projection period, the AR1 models for bottom temperature and recuitment revert to the long-term means of those processes.
-This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypothesis of increasing water temperature and changing environmental conditions [@hareetal16;@pershingetal21].
+This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypotheses of increasing water temperature and changing environmental conditions [@hareetal16;@pershingetal21].
 Because of these concerns, the most recent stock assessment for yellowtail flounder used a change point analysis on environmental covariates to determine current conditions.
-The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections (NEFSC, 2024).
+The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections [@nefsc24a].
 Unfortunately, this methodology still has limitations because it is unlikely that current environmental conditions are stable.
-Similarly, a linear increase of temperature is also not likely in the future (see past trends, top row of Figure 6).
-As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
+Similarly, a linear increase of temperature is also not likely in the future (top row of Figure \ref{fig:E-R-SSB-F}).
+As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in BRPs and population projections becomes an important decision in the management process.
 
@@
 Estimation of natural mortality may be even more challenging within state-space assessment models with their greater flexibility from inclusion of time-varying random effects.
-For example, we fixed the mean natural mortality for age 1 fish and @cadigan16 and @stocketal21 also estimated natural mortality deviations.
+For example, we fixed the mean natural mortality for age 1 fish and both @cadigan16 and @stocketal21 also estimated natural mortality deviations.
 However, the simulation studies of @milleretalinreview1 suggest estimating natural mortality in WHAM is feasible in some situations with typical assessment data.

### 2025-05-19 paper/paper_R1.Rmd
revised paper

--- 2025-03-19 paper/paper.Rmd
@@
 
-However, WHAM has only allowed one stock (component) and a single region and without seasonal changes in stock dynamics.
+However, WHAM has only allowed one stock and a single region and without seasonal changes in stock dynamics.
 Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice [@bosleyetal22;@yingetal11;@caoetal14].
@@
 We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) [@nefsc23].
-Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age model with results subsequently combined to inform management.
-The stock was split into two regions to account for spatial dynamics and improve model diagnostics [@asmfc16].
+Black sea bass is a high profile stock in the Northeast U.S.
+and important to both commercial fishing fleets and recreational anglers.
+Black sea bass are targeted from Maine to North Carolina in both state and federal waters.
+Past stock assessments have estimated that the stock is healthy (not overfished and overfishing not occurring).
+Previous assessments assumed two stock components (north and south), each assessed with a separate statistical catch at age model with results subsequently combined to inform management, but these assessments have exhibited strong retrospective patterns.
+The stock was split into two regional components to account for spatial dynamics and improve model diagnostics [@asmfc16].
 Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall [@mosershepherd09].
 Black sea bass distribution has been linked to warming waters on the northwest shelf [@belletal15] and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north [@hareetal16].
-Thus, a focus of the black sea bass research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics [@nefsc23].To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process.
+Thus, a focus of the black sea bass research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement, and explore environmental drivers of population dynamics [@nefsc23].
+We denote the fish in the north as separate stock components rather than stocks because both components are managed as a single stock, but recognizing differences in migration between fish originating in the two areas.
+
+To illustrate the usage of WHAM 2.0 we apply it to the two stock components of black sea bass off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process.
 However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
@@
 Approaches to modeling transitions among these states may treat time discretely [e.g.,@arnason72;@schwarzetal93] or continuously [e.g., @hearnetal87;@commenges99; @andersenkeiding02].
-We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval $\delta$ (columns), conditional on being in each of those states at the beginning of the interval (rows).
+We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval with duration $\delta$ (columns), conditional on being in each of those states at the beginning of the interval (rows).
 For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
@@
 is the $n_R \times n_F$ matrix defining probabilities of being captured in each fleet during the interval, and $\mathbf{D}_{y,a,i}$ is the $n_R x 1$ matrix of probabilities of dying due to natural mortality during the interval.
-We have the identity matrix $\mathbf{I}_{H}$ for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).
+We have the $n_F\times n_F$ identity matrix $\mathbf{I}_{H}$ for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).
 
@@
 WHAM assumes each fleet $f$ can harvest in only one region ($r_f$) during specified seasons.
-So, for each fleet $f$, row $r_f$ and column $f$ of $\mathbf{H}_{y,a,i}$ will be $F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$ when fleet $f$ is harvesting during the interval $\delta_i$ and all other elements will be zero.
-Row $r$ of the single-column matrix $\mathbf{D}_{y,a,i}$ is  $M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$
-
-When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
-We obtain the PTM over an interval $\delta_i$ by exponentiating the instantaneous rate matrix [@millerandersen08]
-\begin{equation*}
-\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
-\end{equation*}
-The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region.
-Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero.
-For two regions and one fleet operating in each region:
- \begin{equation*}
- \mathbf{A}_{y,a,i} = \begin{bmatrix}
- a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
- \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
- 0 & 0 & 0 & 0 & 0 \\
- 0 & 0 & 0 & 0 & 0 \\
- 0 & 0 & 0 & 0 & 0
- \end{bmatrix}
-\end{equation*}
-where $a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})$.
-When there is one region, $n_f$ fleets, and $\delta_i = 1$, exponentiating the instantaneous rate matrix results in the PTM defined in Eq.
-\ref{eq:ptm_1_region}.
+So, for each fleet $f$, row $r_f$ and column $f$ of $\mathbf{H}_{y,a,i}$ will be $F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$ when fleet $f$ is harvesting during the interval $i$ and all other elements will be zero.
+Row $r$ of the single-column matrix $\mathbf{D}_{y,a,i}$ is  $M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$.
+For details defining PTMs with movement and mortality occuring simultaneously see Appendix B.
 
@@
 Seasonality can be configured to accommodate characteristics of spawning, movement, and fleet-specific behavior.
-The annual time step can be divided into $K$ seasons and the interval size $\delta_i$ for each season $i$ does not need to be equal to any other seasonal interval.
-Under the Markov assumption, the PTM of surviving and moving and dying over $K$ intervals $\delta_1,\ldots, \delta_K$ (i.e., the entire year) is just the product of the PTMs for each interval:
+The annual time step can be divided into $K$ seasons and the interval duration $\delta_i$ for each season $i$ does not need to be equal to any other seasonal interval.
+Under the Markov assumption the PTMs for each consecutive interval are independent and the PTM of surviving and moving and dying over $K$ intervals $1,\ldots, K$ (i.e., the entire year) is just the product of the PTMs for each interval:
 $$ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).$$
 
-For a stock spawning at some fraction of the year $0<t_s<1$ in interval $\delta_j$, the fraction of time in season $j$ is 
-$$\delta_{s,j} = t_s-\sum^{j-1}_{i=0}\delta_i
-$$
-and the PTM defining the proportions in each state at time $t_s$ for age $a$ is
+For a stock spawning at some fraction of the year $0<t_s<1$ in interval $i \in \{1,\ldots,K\}$, the fraction of time in the interval is 
+$$\delta_{s,i} = t_s-\sum^{i-1}_{j=0}\delta_j$$
+(where $\delta_0 = 0$) and the PTM defining the proportions in each state at time $t_s$ for age $a$ is
 \begin{equation}\label{eq:ptm_spawn}
-\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{s,j}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{s,j}).
-\end{equation}
-Similarly, for an index $m$ occurring  at fraction of the year $t_m$ in interval $\delta_j$ the proportions in each state at the time of the observation is
+\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{s,i}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,j}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{s,i})
+\end{equation}
+where $\mathbf{P}_{y,a,0}$ is an identity matrix. Similarly, for an index $m$ occurring  at fraction of the year $t_m$ in interval $i$ the proportions in each state at the time of the observation is
 \begin{equation} \label{eq:ptm_index} 
-\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{m,j}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{m,j}).
+\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{m,i}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,i}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{m,i}).
 \end{equation}
@@
 If recruitment is assumed to be a function of SSB, it is only the spawning population in region $r_s$ at the time of spawning that constitutes the SSB in the stock-recruit function.
-However, models can configure spawning individuals to occur in other regions at the time of spawning.
+However, models can configure individuals to occur in other regions at the time of spawning.
 Aside from treating recruitment as a random walk, the general model for annual recruitment as random effects is
@@
 $$\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}^T(t_s) \mathbf{N}_{O,y,a}$$
-and $w_{y,a}$ is the mean weight at age of spawning individuals, $\text{mat}_{y,a}$ is the maturity at age, and $\mathbf{O}_{y,a,r_s}(t_s)$ are the probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year (the $r_s$ column of the upper-left submatrix of Eq.
+and $w_{y,a}$ and $\text{mat}_{y,a}$ are the stock-specific mean weight at age of spawning individuals and maturity at age, respectively, and $\mathbf{O}_{y,a,r_s}(t_s)$ are the probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year (the $r_s$ column of the upper-left submatrix of Eq.
 \ref{eq:ptm_spawn}).
@@
 As in previous versions of WHAM, the apparent survival (and movement) from one year to another after recruitment can be treated deterministically or as functions of random effects.
-The predicted numbers at age in year $y$ at age $a$ for a given stock are vector analogs ($\mathbf{N}_{O,y,a}$) of the equations for numbers at age in the standard WHAM model [@stockmiller21].
+The predicted numbers at age $a$ in year $y$ for a given stock are vector analogs ($\mathbf{N}_{O,y,a}$) of the equations for numbers at age in the standard WHAM model [@stockmiller21].
 For ages $a = 2,\ldots, A-1$, where $A$ is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first $n_R$ elements of the vector
@@
 $$\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}^T \mathbf{N}_{O,1,1}.$$
-
-The initial abundances at age can also be treated as independent or AR1 random effects. Defining the vector of initial abundance at age in region $r$ as $\mathbf{N}_{O,1,r}$, the general model is 
-$$\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}$$
-where 
-$$  Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.$$
+Initial abundance at age can also be estimated as random effects as described in Appendix B.
 
@@
 When random effects are used, the same 2DAR1 structure with age and year as described by @stockmiller21 can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
-So the general model for natural mortality is
-\begin{equation*}
-  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
-\end{equation*}
-The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
-\begin{equation*}
-  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
-\end{equation*}
+See Appendix B for details.
 
@@
 Predicted age composition observations are functions of $\widehat{I}_{m,y,a}$ as described by @stockmiller21 and the likelihood options are the same as those for catch explained above.
-
-Catchability of the index can also be treated as logit functions of normal random effects and(or) environmental covariate effects 
-$$log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}$$ 
-where $\theta_{q,m}$ is an intercept or mean parameter, and $u_{m}$ and $l_{m}$ are the upper and lower bounds of catchability for index $m$ (defaults are 0 and 1000).
-The general covariance structure for the annual random effects is AR1
-$$Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.$$
+See Appendix B for options to model temporal variation in catchability.
 
@@
 \begin{equation}\label{eq:newton}
-  \log\widetilde{F}^{j} = \log\widetilde{F}^{j-1} - \frac{g\left(\log\widetilde{F}^{j-1}\right)}{g^{(1)}\left(\log\widetilde{F}^{j-1}\right)}
-\end{equation}
-where $g(\log F)$ is the difference between the weighted sums of spawning biomass per recruit at $F$ and $X$\% of unfished spawning biomass per recruit across stocks:
+  \log\widetilde{F}^{j} = \log\widetilde{F}^{j-1} - \frac{h\left(\log\widetilde{F}^{j-1}\right)}{h^{(1)}\left(\log\widetilde{F}^{j-1}\right)}
+\end{equation}
+where $h(\log F)$ is the difference between the weighted sums of spawning biomass per recruit at $F$ and $X$\% of unfished spawning biomass per recruit across stocks:
 \begin{equation}\label{eq:newton-obj}
-  g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
+  h(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
 \end{equation}
 where $\upphi_s\left(F=0\right)$ is the equilibrium unfished spawning biomass per recruit.
-$g^{(1)}(\log F)$ is the derivative of $g$ with respect to $\log F$, and the weights to use for each stock $\lambda_s$ can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate "static" equilibrium spawning biomass and yield.
+$h^{(1)}(\log F)$ is the derivative of $h$ with respect to $\log F$, and the weights to use for each stock $\lambda_s$ can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate "static" equilibrium spawning biomass and yield.
 
@@
 \ref{eq:ypr}) and equilibrium recruitment.
-The function $g(\log F)$ and $g^{(1)}(\log F)$ in Eq.
+The function $h(\log F)$ and $h^{(1)}(\log F)$ in Eq.
 \ref{eq:newton} are the first and second derivatives of the sum of the stock-specific equilibrium yields with respect to $\log F$ [analogous to that described by @milleretal16].
@@
 Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity [@ablehales97;@fabrizioetal13].
-Fish in this northern stock unit perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring [@musickmercer77].
+Fish in this northern stock component perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring [@musickmercer77].
 Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude [@mosershepherd09].
@@
 The first year being modeled for the population is 1989 and the fishery and index data used in the model span from 1989 to 2021.
-The north and south stock components are modeled as separate populations that spawn and recruit in respective regions.
+The north and south stock components are modeled as separate populations that spawn and recruit in respective regions and there are eight age classes with the last being a plus group ($a \in \{1,\ldots,8+\}$).
 We have observations for each of four total fishing fleets, where two fishing fleets (recreational and commercial) operate in each region.
@@
 The southern stock component is assumed to never move to the northern region.
-For the northern component, a proportion $\mu_{\text{N}\rightarrow \text{S}}$ can move to the southern region each month during the last five months of the year, but no movement is allows from the south to the north during this period (Figure \ref{fig:migration-diagram}).
+For the northern component, a proportion $\mu_{\text{N}\rightarrow \text{S}}$ can move to the southern region each month during the last five months of the year, but no movement is allowed from the south to the north during this period (Figure \ref{fig:migration-diagram}).
 During the first four intervals of the year a proportion $\mu_{\text{S}\rightarrow \text{N}}$ of the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
@@
 All models we considered allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
-Although lack of model convergence was observed in many fits for self-test simulations, the estimation of critical model output (e.g., SSB, fishing mortality) was reliable and robust the poor estimation of these scalars (Figure \ref{fig:self-test-fig}), and estimating these parameters rather than fixing them in the fit to the actual data allows uncertainty in model output to be more properly conveyed.
+Although lack of model convergence was observed in many fits for self-test simulations, the estimation of critical model output (e.g., SSB, fishing mortality) was reliable and robust to the poor estimation of these scalars (Figure \ref{fig:self-test-fig}), and estimating these parameters rather than fixing them in the fit to the actual data allows uncertainty in model output to be more properly conveyed.
 
@@
 
-The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residualand retrospective patterns.
+The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual and retrospective patterns.
 These analyses resulted in use of selectivity random effects for the northern fleet and indices, logistic-normal likelihoods for six sets of age composition observations, and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
@@
 
-All models include bottom temperature observations for the northern and southern regions from 1963 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius [@nefsc23].
+All models include observations of winter bottom temperature anomalies for the northern and southern regions from 1959 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius [@nefsc23].
+See the Supplementary Materials for details about the derivation of these observations.
 We retained the assumption from the peer-reviewed assessment that treated the latent bottom temperature covariates in each region as AR1 processes.
@@
 
-We assume the covariate in year $y$ affects recruitment in the same year because the covariate observations are from months January to March.
+We assume the covariate in year $y$ affects recruitment in the same year because the covariate observations are from winter months.
 The fish are technically already 1 year old, but there are no observations of these individuals until later in the spring except possibly in fishery catches which are accumulated over the whole year.
@@
 
-We used a development version ([commit fb8b089](https://github.com/timjmiller/wham/tree/fb8b089)) of the WHAM package prior to the release of version 2.0 for all results.
+We used a development version ([commit 53e236b](https://github.com/timjmiller/wham/tree/53e236b)) of the WHAM package after the release of version 2.0 for all results.
 All code for fitting models and generating results can be found at [github.com/timjmiller/multi_wham_bsb](https://github.com/timjmiller/multi_wham_bsb).
@@
 We estimated median relative bias of SSB across these simulations.
+Finally, we fitted three sensitivity models which differed from the best performing model, by 1) assuming no movement for either stock component, 2) movement parameters are fixed at the means for the prior distributions, and 3) a constant natural mortality rate is estimated across all regions and stocks.
 See the Supplementary Materials for more details.
 
-## Reference points, and projections {-}
+## Reference points and projections {-}
 
@@
 
-Estimated effects of bottom temperature on recruitment for either the northern or southern component were stable over the retrospective peels and differed negligibly whether effects for each component were estimated in isolation or together (Table \ref{tab:beta-sig-peel-table}).
+Estimated positive effects of bottom temperature on recruitment for the northern or southern components were stable over the retrospective peels and differed negligibly whether effects for each component were estimated in isolation or together (Table \ref{tab:beta-sig-peel-table}).
 The estimates for the effects on each stock component were essentially equivalent whether the effect was on one component or both.
@@
 
+The three sensitivity fits suggest the estimates of stock size and fishing mortality are fairly robust to whether the movement rates are fixed at the values from the companion Stock Synthesis fit or whether no movement is assumed (Figure \ref{fig:sensitivity-plots}).
+A constant natural mortality rate was estimable, but the estimated rate (0.72) is nearly twice that determined by the Working Group and resulted in generally higher stock sizes and lower fishing mortality rates.
+
 ## Reference points {-}
@@
 Although this might suggest making inferences on the population by using an ensemble of these and weighting by AIC [@burnhamanderson02], the estimates of assessment outputs relevant for management (e.g., SSB and $F$) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
-Nevertheless, continued monitoring of effects on the southern component may be prudent.
+Continued monitoring of temperature effects on the southern component and investigations of non-linear effects of temperature for both components may be prudent because of the possibility of an optimal temperature range for survival via effects on growth during the pre-recruit phase [@warelambert85;@anderson88].
+Furthermore, our inferences about effects on natural mortality at age 1 may be influences by our assumptions of constant median (or constant depending on the model) natural mortality rate and movement rates for all ages.
+Further investigations of variation in these parameters with age and effects of bottom temperature could improve our understanding of the demography for this stock.
 
 We showed that using expected recruitment rather than recruitment random effects to define BRPs can produce the same estimates but with lower uncertainty.
-Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same BRPs and even less uncertainty than those we showed using expected recruitment.
+Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same BRPs and even lower uncertainty.
 These results can be important for management when uncertainty in assessment output is used in making catch advice.
@@
 The uncertainty in these ratios could also be used in defining catch advice.
+
+BRPS under prevailing conditions use average values for recruitment and other inputs and MSY-based BRPs are based on the stock-recruit relationship which is also typically constant over time.
+If BRPs should be defined in such a way that they are less variable than the stock over time, expected inputs to the BRPs (instead of annual values) would be more appropriate.
+Lower variability in BRPs would presumably correlate with lower variability in catch advice which might be desirable [e.g., @cochraneetal98;@kelletal05], but the appropriate level of temporal variability in BRPs and catch advice will dependent on the characteristics of the fish stock, associated fisheries, and management structure [@holland10;@puntetal16].
 
@@
 
-# Acknowledgements
+Our application to black sea bass is a particular combination of the spatial modeling options described by @bergeretal24.
+Other combinations of the described options are possible in WHAM 2.0.
+For example, WHAM 2.0 can accommodate all of fleet structure options described in Table 2 by @bergeretal24 and we expect more of these options to become available in future versions of WHAM.
+
+# Acknowledgements {-}
 
@@
 We thank the black sea bass research track assessment working group and peer-reviewers for the work compiling data and determining many of the configurations of the black sea bass assessment model that we used for our analyses.
-We also thank Chris Legault for comments and suggestions that improved the clarity of this paper.
+We also thank Chris Legault, two anonymous reviewers, and the associate editor for comments and suggestions that improved the clarity of this paper.
 
@@
 
-```{r move-prior-posterior, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Prior (black) and posterior (red) distributions of movement of the northern stock component from north to south (top) and south to north (bottom)."}
+```{r move-prior-posterior, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Prior (black) and posterior (red) distributions of movement of the northern stock component from north to south (top) and south to north (bottom).
+Vertical lines are the prior and posterior estimates."}
 knitr::include_graphics(here::here("paper", "move_prior_post.pdf"))
@@
 \renewcommand\thetable{A\arabic{table}}
-\input{symboltable.tex}
-
-\pagebreak
-
-# Supplemental Materials {-}
+\input{symboltable_R1.tex}
+
+# Appendix B {-}
+
+New features of WHAM 2.0 that are not used in the application to black sea bass are described here..
+
+## Simultaneous movement and mortality {-}
+
+When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
+We obtain the PTM over an interval $i$ by exponentiating the instantaneous rate matrix [@millerandersen08]
+\begin{equation*}
+\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
+\end{equation*}
+The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region.
+Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero.
+For two regions and one fleet operating in each region:
+ \begin{equation*}
+ \mathbf{A}_{y,a,i} = \begin{bmatrix}
+ a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
+ \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
+ 0 & 0 & 0 & 0 & 0 \\
+ 0 & 0 & 0 & 0 & 0 \\
+ 0 & 0 & 0 & 0 & 0
+ \end{bmatrix}
+\end{equation*}
+where $a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})$.
+When there is one region, $n_f$ fleets, and $\delta_i = 1$, exponentiating the instantaneous rate matrix results in the PTM defined in Eq.
+\ref{eq:ptm_1_region}.
+
+## Initial abundance at age as random effects {-}
+
+The initial abundances at age can also be treated as independent or AR1 random effects. Defining the vector of initial abundance at age in region $r$ as $\mathbf{N}_{O,1,r}$, the general model is 
+\begin{equation*}
+\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}
+\end{equation*}
+where 
+\begin{equation*}
+Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.
+\end{equation*}
+
+## General natural mortality model {-}
+
+The general model for natural mortality at age $a$ for a given stock in region $r$ is a function of random and covariate effects
+\begin{equation*}
+  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
+\end{equation*}
+The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
+\begin{equation*}
+  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
+\end{equation*}
+
+## General catchability model {-}
+
+Catchability of each index can be treated as logit functions of normal random effects and(or) environmental covariate effects
+\begin{equation*}
+log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}
+\end{equation*}
+where $\theta_{q,m}$ is an intercept or mean parameter, and $u_{m}$ and $l_{m}$ are the upper and lower bounds of catchability for index $m$ (defaults are 0 and 1000).
+The general covariance structure for the annual random effects is AR1
+\begin{equation*}
+Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.
+\end{equation*}
+
+\pagebreak
+
+# Supplementary Materials {-}
 
@@
 
+## Bottom temperature anomalies {-}
+
+The Working Group created bottom temperature observations from a high resolution ocean bottom temperature product by @dupontaviceetal23.
+The annual observations for each region are defined by the average over all spatial bottom temperature values for February and March by region and year.
+Similarly, the Working Group calculated standard errors from the standard deviation of all values in the region and the total number of values for a given year.
+We created regional bottom temperature anomalies by subtracting means for each region across all years.
+
 ## Diagnostics {-}
@@
 The 50 jitter fits demonstrated that a local minimum was obtained for the original fit of model $M_0$ (Figure \ref{fig:jitter-M0}).
-Some lower NLLs were obtained with unacceptable gradients, but a slightly lower NLL was found with a satisfactory gradient and with and invertible hessian.
-We therefore refit model $M_0$ and all remaining models using the better parameter estimates as initial values.
+One lower NLL was obtained with unacceptable gradients (No.
+25), but a slightly lower NLL was found with a satisfactory gradient for 3 of the jitters (Nos.
+9, 13, 29).
+However, one of the jitter fits (No.
+9) did not provide a non-zero estimate of the variance parameter for one of the indices and the other two provided identical results and we refit model $M_0$ and all remaining models using the better parameter estimates as initial values.
 
@@
 ```{r jitter-M0, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_0$.
-Black jitters had maximum absolute gradient values < $10^{-10}$, grey jitters had values > $10^{-10}$ and < 1, and red jitters had values > 1."}
+Black jitters had maximum absolute gradient values < $10^{-2}$ and red jitters had values > 1."}
 knitr::include_graphics(here::here("paper", "fit_0_jitter_plt.pdf"))
@@
 ```{r self-test-fig, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Median relative error of SSB (Total and by stock component) and total fully-selected fishing mortality for estimation models fitted to simulated data from model $M_1$ where the observation variance of log-indices are fixed and estimated.
-Black and Red dashed lines represent the median of the annual medians and the median across all annual relative errors, respectively."}
+Black and Red dashed lines represent the median of the annual medians and the median across all annual relative errors, respectively.
+Vertical lines represent 95\\% confidence intervals."}
 knitr::include_graphics(here::here("paper", "self_test_results.pdf"))
@@
 ```
+
+\begin{landscape}
+
+```{r sensitivity-plots, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Estimates of annual SSB and fishing mortality rates for the best performing model $M_1$ and models that are otherwise the same except where  1) movement rates are fixed at the means for the prior distribution, 2) a constant natural mortality rate is estimated, or 3) there is no movement for either stock component."}
+knitr::include_graphics(here::here("paper", "SSB_F_sensitivity_plots.pdf"))
+```
+\end{landscape}

### 2025-05-19 paper/paper_R1_natbib.Rmd
revised paper

--- 2025-05-19 paper/paper_R1.Rmd
@@
     fig_caption: yes
+    citation_package: natbib
     includes:
@@
 #    df_print: paged
-csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
-bibliography: bsb-wham.bib
+# csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
+natbiboptions: round
+biblio-style: cjfas.bst
+# bibliography: bsb-wham.bib
 ---
@@
 
-# References {-}
+\pagebreak
+
+\bibliography{bsb-wham}
 
 <div id="refs"></div>
-
-\pagebreak
 

### 2025-05-19 paper/symboltable_R1.tex
revised paper

--- 2025-01-02 paper/symboltable.tex
@@
 $t_s$ & fraction of the annual time step when spawning occurs for stock $s$\\
-$\delta_{s,j}$ & fraction of the annual time step between $t_s$ and the end of season $j-1$\\
+$\delta_{s,i}$ & fraction of the annual time step between $t_s$ and the beginning of season $i$\\
 $t_m$ & fraction of the annual time step when index $m$ observers the population\\
-$\delta_{m,j}$ & fraction of the annual time step between $t_m$ and the end of season $j-1$\\
+$\delta_{m,i}$ & fraction of the annual time step between $t_m$ and the beginning of season $i$\\
 $\mathbf{N}_{y,a}$ & Abundance at age $a$ in year $y$ in each of the living and mortality states on January 1\\
@@
 $r_s$ & region where stock $s$ spawns and recruits\\
-$\text{SSB}_{s,y}$ & Spawning stock biomass for stock $s$ in year $y$\\
+$\text{SSB}_{y,r_s}$ & Spawning stock biomass for a given stock in region $r_s$ in year $y$\\
 $\varepsilon_{y,a,r}$ & Random error for abundance at age $a$ in year $y$ in region $r$\\
-$w_{s,y,a}$ & mean individual weight for stock $s$ at age $a$ in year $y$\\
-$\text{mat}_{s,y,a}$ & proportion mature at age $a$ in year $y$ for stock $s$\\
-$\mathbf{O}_{s,y,a,r_s}(t_s)$ & Probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year\\
+$w_{y,a}$ & mean individual weight for a given stock at age $a$ in year $y$\\
+$\text{mat}_{y,a}$ & proportion mature at age $a$ in year $y$ for a given stock\\
+$\mathbf{O}_{y,a,r_s}(t_s)$ & Probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year\\
 $\boldsymbol{\varepsilon}_{y,a}$ & vector of random errors for abundance alive in each region on January 1 of year $y$ at age $a$\\
@@
 $\rho_{N,\text{year},r}$ & first order auto-regressive correlation parameter across year for abundance at age random effects in region $r$\\
-$\text{sel}_{1,a,f}$ & selectivity at age $a$ for fleet $f$ in the first year\\
+$\text{sel}_{1,a,f}$ & selectivity scalar at age $a$ for fleet $f$ in the first year\\
 $F_{1,a,f}$ & fishing mortality rate at age $a$ for fleet $f$ in the first year\\
 $\widetilde{F}_{a,f}$ & equilibrium fishing mortality rate at age $a$ for fleet $f$\\
-$\mathbf{O}_{a}$ & proportion surviving the year at age $j$ and occuring in each region (columns) given alive on January 1 in each region (rows)\\
+$\mathbf{O}_{a}$ & proportion surviving the year at age $a$ and occuring in each region (columns) given alive on January 1 in each region (rows)\\
 $\widetilde{\mathbf{O}}_{a}$ & equilibrium proportions alive in each region at age $a$ (columns) given recruitment in each region (rows)\\
@@
 $\rho_{N_1,r}$ & first order auto-regressive correlation parameter for initial numbers at age random effects in the first year for region $r$\\
-$\mu_{r\rightarrow r',y,a,i}$ & movement from region $r$ to region $r'$ in seasonal interval $i$ and year $y$ at age $a$\\
-$g(\mu_{r\rightarrow r',y,a,i})$ & link function for movement $\mu_{r\rightarrow r',y,a,i}$\\
-$\theta_{r\rightarrow r',i}$ & mean parameter across age and year for movement from region $r$ to region $r'$ in seasonal interval $i$\\
-$\varepsilon_{r\rightarrow r',y,a,i}$ &  random effect parameter for movement from region $r$ to region $r'$ in seasonal interval $i$ and year $y$ at age $a$\\
+$\mu_{r\rightarrow r',y,a,i}$ & movement rate from region $r$ to region $r'$ in seasonal interval $i$ and year $y$ at age $a$\\
+$g(\mu_{r\rightarrow r',y,a,i})$ & link function for movement rate $\mu_{r\rightarrow r',y,a,i}$\\
+$\theta_{r\rightarrow r',i}$ & mean or intercept parameter across age and year for movement from region $r$ to region $r'$ in seasonal interval $i$\\
+$\varepsilon_{r\rightarrow r',y,a,i}$ & random effect for movement from region $r$ to region $r'$ in seasonal interval $i$ and year $y$ at age $a$\\
 $n_E$ & number of environmental covariates\\
@@
 $\rho_{q,m}$ & first order auto-regressive correlation parameter across year for index $m$ catchability random effects\\
+$\text{sel}_{f,a}$ & selectivity of fleet $f$ at age $a$ for SPR-based reference points\\
+$\overline{F}_{f,a}$ & Average fishing mortality for fleet $f$ at age $a$ over user-specified set of years\\
+$\upphi(\widetilde{F})$ & equilibrium spawning biomass per recruit for stock $s$ at fully-selected fishing mortality rate $\widetilde{F}$\\
+$\widetilde{\mathbf{O}}_{a,r_s,\cdot}$ & proportion surviving the year at age $a$ and occuring in each region (columns) given alive on January 1 in region $r_s$\\
+$\mathbf{O}_{a,\cdot,r_s}(t_s)$ & proportion surviving the year at age $a$ starting in each regin (rows) and occuring in region $r_s$ at spawning time $t_s$\\
+$\widetilde{\nu}({\widetilde{F}})$ & matrix of equilibrium yield per recruit by region and fleet at fully-selected fishing mortality rate $\widetilde{F}$\\
+$\lambda_s$ & weight for stock $s$ to use for spawning biomass per recruit for defining SPR-based fishing mortality reference point
 \end{longtable}

### 2025-05-19 R/fits_rev.R

@@ -1,5 +1,5 @@
 #multi-wham (12/12/2023)
-#library("wham", lib.loc = "c:/work/wham/old_packages/4a44134")
+#library("wham", lib.loc = "c:/work/wham/old_packages/24c8156")
 
 #pkgbuild::compile_dll("c:/work/wham/wham", debug = FALSE)
 #pkgload::load_all("c:/work/wham/wham")
@@ -8,7 +8,9 @@
 #pak::pkg_install("timjmiller/wham@devel", upgrade = FALSE, ask = FALSE)
 
 #devel branch is now multi-wham
-library(wham)
+#library(wham)
+library("wham", lib.loc = "c:/work/wham/old_packages/53e236b")
+# library("wham", lib.loc = "c:/work/wham/old_packages/fb8b089")
 
 library(here)
 
@@ -16,17 +18,9 @@ source(here("R","make_wham_inputs.R"))
 
 input_0 <- prepare_wham_input(asap, selectivity = sel, NAA_re = NAA_re, basic_info = basic_info, move = move, ecov = ecov_0, catch_info = catch_info,
 	index_info = index_info, age_comp = age_comp)
-# input_0$par <- fit_1$parList
-# input_0$par$Ecov_beta_R[] <- 0
-
 fit_0 <- fit_wham(input_0, do.brps = FALSE, do.sdrep = FALSE, do.osa = FALSE, do.retro = FALSE)
+fit_0 <- do_sdreport(fit_0)
 saveRDS(fit_0, here::here("results", "fit_0_rev.RDS"))
-
-#Do jitter for null model
-res_dir <- here::here("results","jitter", "fit_0")
-dir.create(res_dir)
-fit_0_jit_res <- jitter_wham(fit_RDS = here::here("results","fit_0_rev.RDS"), n_jitter = 50, res_dir = res_dir)
-saveRDS(fit_0_jit_res, here::here(res_dir, "fit_0_jitter_results.RDS"))
 
 ####################
 #this will give the same initial values as used by jitter_wham
@@ -36,32 +30,38 @@ saveRDS(fit_0_jit_res, here::here(res_dir, "fit_0_jitter_results.RDS"))
 # initial_vals <- t(sapply(1:50, function(x) fit_0$opt$par + chol.L %*% cbind(rnorm(n= NCOL(cov)))))
 ####################
 
+#Do jitter for null model
[…]

### 2025-05-19 R/make_figures.R

@@ -3,7 +3,8 @@ library(rnaturalearth)
 library(maps)
 library(sp)
 library(sf)
-library(wham)
+library("wham", lib.loc = "c:/work/wham/old_packages/53e236b")
+# library(wham)
 library(ggplot2)
 library(dplyr)
 library(patchwork)
@@ -11,8 +12,8 @@ library(patchwork)
 source(here::here("R","plot_functions.R"))
 
 #load fits
-fits <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))
-fits_M_re <- readRDS(file.path("results","fits_M_re_better.RDS"))
+fits <- readRDS(file.path("results","fits_no_M_re_rev_1.RDS"))
+fits_M_re <- readRDS(file.path("results","fits_M_re_best.RDS"))
 
 ##############################################################################################
 
@@ -76,6 +77,8 @@ dlogitnorm <- function(x, p, sigma){
   return(fx)
 }
 
+fit1 <- readRDS(here::here("results","fit_1.RDS")) #M_1
+
 cairo_pdf(file.path("paper", "move_prior_post.pdf"), width = 7, height = 11)
 par(mfrow = c(2,1), mar = c(4,4,1,1), oma = c(1,1,1,1))
 ps <- seq(0,0.1,0.001)
@@ -105,7 +108,7 @@ sapply(c(fits,fits_M_re), \(x) x$rep$mu[1,8,1,1,2,1])
 #plot selectivity at time and age for blocks with RE.
 source(here::here("R","plot_functions.R"))
 cairo_pdf(file.path("paper", "selectivity_re_plot.pdf"), width = 16, height = 12)
-plot.selectivity(fits[[2]], blocks = c(1:2,5:6), block.names = c("North Commercial Fleet", "North Recreational Fleet", "North Recreational Catch/Effort Index", "North Spring Index"))
+plot.selectivity(fit1, blocks = c(1:2,5:6), block.names = c("North Commercial Fleet", "North Recreational Fleet", "North Recreational Catch/Effort Index", "North Spring Index"))
 dev.off()
 
 
@@ -117,7 +120,7 @@ dev.off()
 
 
 cairo_pdf(file.path("paper", paste0("Ecov.pdf")), height = 10, width = 10)
[…]

### 2025-05-19 R/make_tables.R

@@ -1,3 +1,4 @@
+library("wham", lib.loc = "c:/work/wham/old_packages/53e236b")
 
 mod_names <- paste0("$M_{", 0:13,"}$")
 col_names <- c("Model", "Temperature effect (north)", "Temperature effect (south)", "$M$ at age 1 random effects")
@@ -14,8 +15,8 @@ mod_table[c(6,7),3] <- "$M$ at age 1"
 mod_table[8:14,2:3] <- mod_table[1:7,2:3]
 
 
-fits <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))
-fits_M_re <- readRDS(file.path("results","fits_M_re_better.RDS"))
+fits <- readRDS(file.path("results","fits_no_M_re_rev_1.RDS"))
+fits_M_re <- readRDS(file.path("results","fits_M_re_best.RDS"))
 
 # AIC
 
@@ -87,10 +88,10 @@ saveRDS(par_peels, file.path("results", "beta_sigma_par_peel_table.RDS"))
 ########################################################
 #static BRP table
 
-sapply(fits, \(x) x$rep$log_FXSPR_static)
-sapply(fits, \(x) x$rep$log_SSB_FXSPR_static)
-sapply(fits, \(x) x$rep$log_SPR0[,3])
-exp(sapply(fits, \(x) x$rep$log_SSB_FXSPR[,3]))
+# sapply(fits, \(x) x$rep$log_FXSPR_static)
+# sapply(fits, \(x) x$rep$log_SSB_FXSPR_static)
+# sapply(fits, \(x) x$rep$log_SPR0[,3])
+# exp(sapply(fits, \(x) x$rep$log_SSB_FXSPR[,3]))
 
 
-fits[[1]]$rep$log_SPR0[,3])
+# fits[[1]]$rep$log_SPR0[,3])

### 2025-05-19 R/plot_Ecov_R_SSB_F.R

@@ -1,4 +1,4 @@
-library(wham)
+library("wham", lib.loc = "c:/work/wham/old_packages/53e236b")
 library(ggplot2)
 library(dplyr)
 library(patchwork)
@@ -9,12 +9,13 @@ theme_update(strip.background = element_blank(), strip.placement = "outside", st
       axis.title = element_text(size = rel(2)), axis.text = element_text(size = rel(2)), legend.text = element_text(size = rel(2)), #text = element_text(size = rel(2)), 
       legend.title = element_text(size = rel(2)))
 
-fit <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))[[2]] #M_1
-proj_1 <- readRDS(here::here("results",paste0("m1_proj_",1,".RDS")))
-proj_1_fjp <- readRDS(here::here("results",paste0("m1_proj_",1,"_sdrep_fjp.RDS")))
+# fit <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))[[2]] #M_1
+fit <- readRDS(file.path("results","fit_1.RDS"))#M_1
+proj_1 <- readRDS(here::here("results",paste0("m1_proj_1_R_opt1.RDS")))
+# proj_1_fjp <- readRDS(here::here("results",paste0("m1_proj_",1,"_sdrep_fjp.RDS")))
 proj <- readRDS(here::here("results","m1_proj_3_R_opt1.RDS"))
-pred_R_N <- make_pred_R_cv(proj_1, proj_1_fjp, type = 1)
-pred_R_S <- make_pred_R_cv(proj_1, proj_1_fjp, type = 1, region = 2)
+pred_R_N <- make_pred_R_cv(proj_1, type = 1)
+pred_R_S <- make_pred_R_cv(proj_1, type = 1, region = 2)
 
 
 df_use <- cbind.data.frame(Year = proj$input$years_Ecov, name = "Ecov_RE", val = TMB:::as.list.sdreport(proj$sdrep, report = TRUE, what = "Est")$Ecov_x[,1], 

### 2025-05-19 R/plot_brps_status.R

@@ -1,4 +1,4 @@
-library(wham)
+library("wham", lib.loc = "c:/work/wham/old_packages/53e236b")
 library(ggplot2)
 library(dplyr)
 library(patchwork)

### 2025-05-19 R/plot_projections.R

@@ -1,4 +1,4 @@
-library(wham)
+library("wham", lib.loc = "c:/work/wham/old_packages/53e236b")
 library(ggplot2)
 library(dplyr)
 library(patchwork)
@@ -8,7 +8,8 @@ theme_update(strip.background = element_blank(), strip.placement = "outside", st
       axis.title = element_text(size = rel(2)), axis.text = element_text(size = rel(2)), legend.text = element_text(size = rel(2)), #text = element_text(size = rel(2)), 
       legend.title = element_text(size = rel(2)))
 
-fit1 <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))[[2]] #M_1
+# fit1 <- readRDS(file.path("results","fits_no_M_re_rev.RDS"))[[2]] #M_1
+fit1 <- readRDS(here::here("results","fit_1.RDS")) #M_1
 
 proj_R1 <- lapply(1:3, \(x) readRDS(here::here("results",paste0("m1_proj_",x,"_R_opt1.RDS"))))
 proj_R3 <- lapply(1:3, \(x) readRDS(here::here("results",paste0("m1_proj_",x,"_R_opt3.RDS"))))
@@ -536,149 +537,149 @@ dev.off()
 #######################################################################################
 #old
 
-proj <- list() 
-for(i in 1:3) proj[[i]] <- readRDS(here::here("results",paste0("m1_proj_",i,".RDS")))
-fjp <- list() 
-for(i in 1:3) fjp[[i]] <- readRDS(here::here("results",paste0("m1_proj_",i,"_sdrep_fjp.RDS")))
-for(i in 1:3) proj[[i]]$sdrep <- fjp[[i]]
+# proj <- list() 
+# for(i in 1:3) proj[[i]] <- readRDS(here::here("results",paste0("m1_proj_",i,".RDS")))
+# fjp <- list() 
+# for(i in 1:3) fjp[[i]] <- readRDS(here::here("results",paste0("m1_proj_",i,"_sdrep_fjp.RDS")))
+# for(i in 1:3) proj[[i]]$sdrep <- fjp[[i]]
 
[…]

### 2025-05-19 R/projections.R

@@ -1,7 +1,7 @@
 # library(wham)
-devtools::load_all("c:/work/wham/wham")
-fit1 <-readRDS(file.path("results","fits_no_M_re_rev.RDS"))[[2]]
-
+library("wham", lib.loc = "c:/work/wham/old_packages/53e236b")
+# fit1 <-readRDS(file.path("results","fits_no_M_re_rev_1.RDS"))[[2]]
+fit1 <- readRDS(here::here("results","fit_1.RDS"))
 tdat <- cbind.data.frame(time = fit1$input$years_Ecov, bt_N = fit1$rep$Ecov_x[,1],bt_S = fit1$rep$Ecov_x[,2])
 lmtdat <- lm(bt_N ~time, data = tdat)
 newdat <- cbind.data.frame(time = c(tdat$time,tail(tdat$time,1)+1:20), bt_N = c(tdat$bt_N, rep(NA,20)), bt_S = c(tdat$bt_S, rep(NA,20)))
@@ -33,11 +33,11 @@ proj.opts.list[[3]] <- list(n.yrs = 10, proj_F_opt = rep(1,10), proj_Fcatch = re
 saveRDS(proj.opts.list, here::here("results", "m1_proj_opts_list.RDS"))
 # temp <- project_wham(fit1, proj.opts = proj.opts.list[[1]], check.version = F, save.sdrep = FALSE)
 proj.list <- list()
-proj.list[[1]] <- project_wham(fit1, proj.opts = proj.opts.list[[1]], check.version = F, do.sdrep = TRUE, TMB.jointPrecision = TRUE)
+proj.list[[1]] <- project_wham(fit1, proj.opts = proj.opts.list[[1]], do.sdrep = TRUE, TMB.jointPrecision = TRUE)
 saveRDS(proj.list[[1]], here::here("results","m1_proj_1_R_opt2.RDS"))
-proj.list[[2]] <- project_wham(fit1, proj.opts = proj.opts.list[[2]], check.version = F, do.sdrep = TRUE, TMB.jointPrecision = TRUE)
+proj.list[[2]] <- project_wham(fit1, proj.opts = proj.opts.list[[2]], do.sdrep = TRUE, TMB.jointPrecision = TRUE)
 saveRDS(proj.list[[2]], here::here("results","m1_proj_2_R_opt2.RDS"))
[…]

### 2025-05-19 R/self_tests.R

@@ -1,11 +1,13 @@
 
-library(wham)
+# library(wham)
+library("wham", lib.loc = "c:/work/wham/old_packages/53e236b")
 #seeds <- 1:10
 fit_1 <- readRDS(here::here("results","fit_1.RDS"))
-seeds <- sample(0:1e9, 100, replace = TRUE)
-res_dir <- here::here("results","self_test", "fit_1")
+# seeds <- sample(0:1e9, 100, replace = TRUE)
+res_dir <- here::here("results","self_test", "fit_1_rev")
 dir.create(res_dir)
-self_test_1 <- self_test(fit_RDS = here::here("results","fit_1.RDS"), n = 50, seeds = seeds, res_dir = res_dir)
+# self_test_1 <- self_test(fit_RDS = here::here("results","fit_1.RDS"), n = 50, seeds = seeds, res_dir = res_dir)
+self_test_1 <- self_test(fit_RDS = here::here("results","fit_1.RDS"), n = 50, res_dir = res_dir)
 saveRDS(self_test_1, file = file.path(res_dir, "self_test_results.RDS"))
 
 self_test_1 <- readRDS(file = file.path(res_dir, "self_test_results.RDS"))
@@ -43,7 +45,8 @@ plot(med_res_par, xlab = "par")
 abline(h = 0)
 
 #seeds <- sample(0:1e9, 100, replace = TRUE)
-res_dir <- here::here("results","self_test", "fit_1_fix")
+seeds <- self_test_1$seeds
+res_dir <- here::here("results","self_test", "fit_1_rev_fix")
 dir.create(res_dir)
 fit_1 <- readRDS(here::here("results","fit_1.RDS"))
 source(here::here("R","plot_functions.R"))
@@ -99,135 +102,135 @@ self_test_fix[[1]][[1]]$sim_input$data$mu[1,8,1,1,,]
 self_test_fix[[1]][[1]]$rep$mu[1,8,1,1,,]
 fit_1$rep$mu[1,8,1,1,,]
 
-ini <- self_test_fix[[1]][[38]]$sim_input #bad negative median across all years
-ini$random <- c(ini$random, c("mean_rec_pars","log_N1", "logit_q","F_pars","logit_selpars","Ecov_beta_R"))
-reml_fit <- fit_wham(ini, do.sdrep = FALSE, do.retro = FALSE, do.osa = FALSE, do.brps = FALSE)
[…]

### 2025-05-19 R/sensitivity_fits.R

@@ -1,0 +1,103 @@
+library("wham", lib.loc = "c:/work/wham/old_packages/53e236b")
+# library("wham", lib.loc = "c:/work/wham/old_packages/24c8156")
+library(here)
+source(here::here("R","make_wham_inputs.R"))
+
+#don't use prior
+move$use_prior <- array(0, dim = c(2,length(seasons),2,1))
+
+fit1 <- readRDS(here::here("results", paste0("fits_no_M_re_rev_1.RDS")))[[2]]
+
+
+#don't use prior
+input_1 <- prepare_wham_input(asap, selectivity = sel, NAA_re = NAA_re, basic_info = basic_info, move = move, ecov = ecov_1, catch_info = catch_info,
+	index_info = index_info, age_comp = age_comp)
+input_1$fleet_names <- paste0(rep(c("North_", "South_"),each = 2), temp$fleet_names)
+input_1$index_names <- paste0(rep(c("North_", "South_"),c(2,2)), temp$index_names)
+input_1$par <- fit1$parList
+input_1$par$mu_prior_re[] <- 0
+input_1$map$trans_mu <- factor(rep(NA, length(input_1$par$trans_mu)))
+# input_1$par <- fit_0$parList
+# fit_1_1 <- fit_wham(input_1, do.fit = FALSE, do.brps = FALSE, do.sdrep = FALSE, do.osa = FALSE, do.retro = FALSE)
+fit_1_1 <- fit_wham(input_1, do.brps = FALSE, do.sdrep = FALSE, do.osa = FALSE, do.retro = FALSE)
+saveRDS(fit_1_1, here::here("results", "fit_1_fix_mu.RDS"))
+
+
+#estimate median M (constant across stocks, regions)
+source(here::here("R","make_wham_inputs.R"))
+M <- list(mean_model = "estimate-M")
+input_2 <- prepare_wham_input(asap, selectivity = sel, NAA_re = NAA_re, basic_info = basic_info, move = move, ecov = ecov_1, catch_info = catch_info, M = M,
+	index_info = index_info, age_comp = age_comp)
+input_2$fleet_names <- paste0(rep(c("North_", "South_"),each = 2), temp$fleet_names)
+input_2$index_names <- paste0(rep(c("North_", "South_"),c(2,2)), temp$index_names)
+input_2$par <- fits[[2]]$parList
+
+#input_2$par <- fit_1_1$parList
+
[…]

### 2025-06-17 paper/paper_R2.tex
revisions complete

--- 2025-03-18 paper/paper.tex
@@
   pdftitle={Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass},
-  pdfauthor={Timothy J. Miller1; Kiersten Curti; Alexander C. Hansell},
+  pdfauthor={Timothy J. Miller1; Kiersten L. Curti; Alexander C. Hansell},
   hidelinks,
@@
 \title{Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass}
-\author{Timothy J. Miller\textsuperscript{1} \and Kiersten Curti \and Alexander C. Hansell}
-\date{18 March, 2025}
+\author{Timothy J. Miller\textsuperscript{1} \and Kiersten L. Curti \and Alexander C. Hansell}
+\date{16 June, 2025}
 
@@
 
-A state-space age-structured stock assessment model can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al.
+State-space age-structured stock assessment models can be used to estimate time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al.
 2016b).
@@
 2020; Varkey et al.
-2022) and to the south, the Woods Hole Assessment Model (WHAM, Miller and Stock 2020; Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).
-
-WHAM is an R package developed and maintained at NOAA's Northeast Fisheries Science Center {[}\url{https://timjmiller.github.io/wham}{]}.
-WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and effects of environmental covariates on various population parameters.
+2022) and to the south, the Woods Hole Assessment Model (WHAM, Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024a).
+
+WHAM is an R package developed and maintained by scientists at NOAA's Northeast Fisheries Science Center (\url{https://timjmiller.github.io/wham}).
+WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects, to models with several time and age varying process errors and effects of environmental covariates on various population parameters.
 Like SAM, WHAM models are built using the Template Model Builder package (TMB, Kristensen et al.
@@
 
-However, WHAM has only allowed one stock (component) and a single region and without seasonal changes in stock dynamics.
+However, WHAM has only allowed consideration of a single stock and region, and no seasonal changes in stock dynamics.
 Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice (Ying et al.
@@
 
-We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) (NEFSC 2023).
-Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age assessment.
-The stock was split into two regions to account for spatial dynamics and improve model diagnostics (ASMFC 2016).
-Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (Moser and Shepherd 2009).
-Black sea bass distribution has been linked to warming waters on the northwest shelf (Bell et al.
-2015) and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare et al.
+We developed this extension of WHAM in concert with a research track assessment process for black sea bass (BSB, \textit{Centropristis striata}; NEFSC 2023), which is a high profile stock in the Northeast U.S.
+and important to both commercial fishing fleets and recreational anglers.
+BSB are targeted from Maine to North Carolina in both state and federal waters.
+Past stock assessments have estimated that the stock is healthy (not overfished and overfishing not occurring).
+Those assessments assumed two stock components (north and south), each assessed with a separate statistical catch at age model, with results subsequently combined to inform management.
+However, the assessments exhibited strong retrospective patterns.
+The stock was split into two regional components to account for spatial dynamics and improve model diagnostics (ASMFC 2016).
+Additionally, BSB have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (Moser and Shepherd 2009).
+BSB distribution has been linked to warming waters on the northwest shelf (Bell et al.
+2015) and it is hypothesized that the distribution and productivity of the BSB stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare et al.
 2016).
-Thus, a focus of the black sea bass research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process.
-However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
-
-When modeling multiple stock components occurring in different regions further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al.
+Thus, a focus of the BSB research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement, and explore environmental drivers of population dynamics (NEFSC 2023).
+We denote the fish in the north as separate stock components rather than stocks because both components are managed as a single stock, but recognizing differences in migration between fish originating in the two areas.
+
+To illustrate the usage of WHAM 2.0 we apply it to the two stock components of BSB off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process (NEFSC 2023).
+However, here we evaluate evidence for alternative hypotheses of temporal variation and effects of bottom temperature, specifically, on recruitment and natural mortality of age 1 individuals.
+
+When modeling multiple stock components and regions, further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al.
 2021).
 Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises.
-Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points.
+Finally, the inclusion of explicit effects of environmental covariates in the model also requires assumptions on how they are treated in projections and reference points.
 We demonstrate some of the options of WHAM version 2.0 to provide biological reference points and short-term projections.
@@
 Many of the options and equations of WHAM version 2.0 are the same as those in Stock and Miller (2021), so we will only describe extensions and differences that have occurred since their first description of WHAM.
-The new version of WHAM can model multiple stocks each with their own movement, harvest, and natural mortality.
+The new version of WHAM can model multiple stocks, each with their own movement, harvest, and natural mortality.
 Much of the description below is for a specific stock \(s\), but, for simplicity, this subscript is implicit except when necessary.
@@
 
-Because individuals may be alive in one of several regions or harvested in one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model.
+Because individuals may be alive in one of several regions or harvested by one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model.
 Approaches to modeling transitions among these states may treat time discretely (e.g., Arnason 1972; Schwarz et al.
@@
 1987; Commenges 1999; Andersen and Keiding 2002).
-We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval \(\delta\) (columns), conditional on being in each of those state at beginning of the interval (rows).
+We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval with duration \(\delta\) (columns), conditional on being in each of those states at the beginning of the interval (rows).
 For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
-The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by year \(y\), age \(a\), and season \(i\) on January 1.
+The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by year \(y\), age \(a\), and season \(i\).
 Each row and column of the PTM correspond to one of the states: alive in region \(r\), dead in fleet \(f\), or dead from natural causes.
-The probabilities in each row sum to unity (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval.
+The probabilities in each row sum to one (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval.
 Given \(n_R\) regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
@@
     0 & 0 & 1
-  \end{bmatrix}
+  \end{bmatrix},
 \end{equation}
@@
 \end{equation*}
-is the \(n_R \times n_F\) matrix defining probabilities of being captured in each fleet during the interval, and \(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of dying due to natural mortality during the interval.
-We have the identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).
-
-WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would be multinomial distributed as in a model for tagging data where fates of individual tagged fish are assumed independent.
-The PTMs determine the expected numbers 1) in each state on January 1 of year \(t+1\) at age \(a+1\) given the abundances at age \(a\) on January 1 of year \(t\), 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.
+is the \(n_R \times n_F\) matrix defining probabilities of being captured in each fleet during the interval, and \(\mathbf{D}_{y,a,i}\) is the \(n_R \times 1\) matrix of probabilities of dying due to natural mortality during the interval.
+We have the \(n_F\times n_F\) identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is one (no zombies allowed).
+
+WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would have a multinomial distribution as in a model for tagging data where fates of individual tagged fish are assumed independent.
+The PTMs determine the expected numbers 1) in each state on January 1 of year \(y+1\) at age \(a+1\) given the abundances at age \(a\) on January 1 of year \(y\), 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.
 
@@
      0 & 0 & 1
-  \end{bmatrix}
+  \end{bmatrix},
 \end{equation}
-where \(S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}\), \(\mathbf{H}_{y,a,i}\) is a 1 x \(n_F\) matrix with elements for each fleet \(f\): \(\frac{F_{y,a,i,f}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), \(D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), and \(Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{y,a,i,f}\) is the total mortality rate.
+where \(S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}\), \(\mathbf{H}_{y,a,i}\) is a 1 x \(n_F\) matrix with elements for each fleet \(f\): \(\frac{F_{f,y,a,i}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), \(D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), and \(Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{f,y,a,i}\) is the total mortality rate.
 
@@
 \begin{equation*}
-  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i}
+  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i},
 \end{equation*}
@@
 \end{equation*}
-and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\) matrix of probabilities of moving from one region to another or staying in the region they occurred at the beginning of the interval:
+and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\) matrix of proportions moving from one region to another or staying in the region they occurred at the beginning of the interval:
 \begin{equation*}
@@
 WHAM assumes each fleet \(f\) can harvest in only one region (\(r_f\)) during specified seasons.
-So, for each fleet \(f\), row \(r_f\) and column \(f\) of \(\mathbf{H}_{y,a,i}\) will be \(F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\) when fleet \(f\) is harvesting during the interval \(\delta_i\) and all other elements will be zero.
-Row \(r\) of the single-column matrix \(\mathbf{D}_{y,a,i}\) is \(M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\)
-
-When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
-We obtain the PTM over an interval \(\delta_i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
-\begin{equation*}
-\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
-\end{equation*}
-The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region.
-Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero.
-For two regions and one fleet operating in each region:
-\begin{equation*}
- \mathbf{A}_{y,a,i} = \begin{bmatrix}
- a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
- \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
- 0 & 0 & 0 & 0 & 0 \\
- 0 & 0 & 0 & 0 & 0 \\
- 0 & 0 & 0 & 0 & 0
- \end{bmatrix}
-\end{equation*}
-where \(a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})\).
-When there is one region, \(n_f\) fleets, and \(\delta_i = 1\), exponentiating the instantaneous rate matrix results in the PTM defined in Eq.
-\ref{eq:ptm_1_region}.
+So, for each fleet \(f\), row \(r_f\) and column \(f\) of \(\mathbf{H}_{y,a,i}\) will be \(F_{f,y,a,i}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\) when fleet \(f\) is harvesting during the interval \(i\) and all other elements will be zero.
+Row \(r\) of the single-column matrix \(\mathbf{D}_{y,a,i}\) is \(M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\).
+For details defining PTMs with movement and mortality occurring simultaneously see Appendix B.
 
@@
 Seasonality can be configured to accommodate characteristics of spawning, movement, and fleet-specific behavior.
-The annual time step can be divided into \(K\) seasons and the interval size \(\delta_i\) for each season \(i\) does not need to be equal to any other seasonal interval.
-Under the Markov assumption, the PTM of surviving and moving and dying over \(K\) intervals \(\delta_1,\ldots, \delta_K\) (i.e., the entire year) is just the product of the PTMs for each interval:
+The annual time step can be divided into \(K\) seasons and the interval duration \(\delta_i\) for each season \(i\) does not need to be equal to any other seasonal interval.
+Under the Markov assumption the PTMs for each consecutive interval are independent and the PTM of surviving, moving, and dying over \(K\) intervals \(1,\ldots, K\) (i.e., the entire year) is just the product of the PTMs for each interval:
 \[ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).\]
 
-For a stock spawning at some fraction of the year \(0<t_s<1\) in interval \(\delta_j\), the fraction of time in season \(j\) is
-\[\delta_{s,j} = t_s-\sum^{j-1}_{i=0}\delta_i
-\]
-and the PTM defining the proportions in each state at time \(t_s\) for age \(a\) is
+For a stock spawning at some fraction of the year \(0<t_s<1\) in interval \(i \in \{1,\ldots,K\}\), the fraction of time in the interval is
+\[\delta_{s,i} = t_s-\sum^{i-1}_{j=0}\delta_j\]
+(where \(\delta_0 = 0\)), and the PTM defining the proportions in each state at time \(t_s\) for age \(a\) is
 \begin{equation}\label{eq:ptm_spawn}
-\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{s,j}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{s,j}).
+\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{s,i}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,j}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{s,i}),
 \end{equation}
-Similarly, for an index \(m\) occurring at fraction of the year \(t_m\) in interval \(\delta_j\) the proportions in each state at the time of the observation is
+where \(\mathbf{P}_{y,a,0}\) is an identity matrix.
+Similarly, for an index \(m\) occurring at fraction of the year \(t_m\) in interval \(i\) the proportions in each state at the time of the observation is
 \begin{equation} \label{eq:ptm_index} 
-\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{m,j}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{m,j}).
+\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{m,i}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,i}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{m,i}).
 \end{equation}
@@
 
-When there are \(n_R\) regions and \(n_F\) fleets, the vector of abundance in each state at age \(a>1\) on January 1 is \(\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}^T, \mathbf{0}^T)^T\) where \(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})^T\) is the number in the states corresponding to being alive in each region and \(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each fleet and dead from natural mortality because no age \(a\) fish have died yet on January 1.
+When there are \(n_R\) regions and \(n_F\) fleets, the vector of abundance for a given stock (component) in each state at age \(a>1\) on January 1 is \(\mathbf{N}_{y,a}^T = (\mathbf{N}_{O,y,a}^T, \mathbf{0}^T)\), where \(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})^T\) is the number in the states corresponding to being alive in each region and \(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each fleet or dead from natural mortality because no age \(a\) fish have died yet on January 1.
 
 Each stock \(s\) is assumed to spawn and recruit in one region \(r_s\).
-So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row \(r = r_s\).
-Options for configuring recruitment (\(N_{y,1,r_s}\)) for each stock are the same as previous versions of WHAM (Miller and Stock 2020).
+So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row \(r_s\).
+Options for configuring recruitment (\(N_{y,1,r_s}\)) for each stock are the same as previous versions of WHAM.
 If recruitment is assumed to be a function of SSB, it is only the spawning population in region \(r_s\) at the time of spawning that constitutes the SSB in the stock-recruit function.
-However, models can configure spawning individuals to occur in other regions at the time of spawning.
+However, models can configure individuals to occur in other regions at the time of spawning.
 Aside from treating recruitment as a random walk, the general model for annual recruitment as random effects is
 \begin{equation*}
-\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1,r_s} =  f\left(\text{SSB}_{y-1,r_s}\right) + \varepsilon_{y,1,r_s}
+\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1} =  f\left(\text{SSB}_{y-1}\right) + \varepsilon_{y,1,r_s},
 \end{equation*}
 where\\
-\[\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}^T(t_s) \mathbf{N}_{O,y,a}\]
-and \(w_{y,a}\) is the mean weight at age of spawning individuals, \(\text{mat}_{y,a}\) is the maturity at age, and \(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and occurring in region \(r_s\) at time \(t_s\) given being alive in each region at the start of the year (the \(r_s\) column of the upper-left submatrix of Eq.
+\[\text{SSB}_{y} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,\cdot,r_s}^T(t_s) \mathbf{N}_{O,y,a}\]
+and \(w_{y,a}\) and \(\text{mat}_{y,a}\) are the stock-specific mean weight at age of spawning individuals and maturity at age, respectively, and \(\mathbf{O}_{y,a,\cdot,r_s}(t_s)\) are the probabilities of surviving and occurring in region \(r_s\) at time \(t_s\) given being alive in each region at the start of the year (column \(r_s\) of the upper-left submatrix of Eq.
 \ref{eq:ptm_spawn}).
@@
 As in previous versions of WHAM, the apparent survival (and movement) from one year to another after recruitment can be treated deterministically or as functions of random effects.
-The predicted numbers at age in year \(y\) at age \(a\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (Stock and Miller 2021).
+The predicted numbers at age \(a\) in year \(y\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (Stock and Miller 2021).
 For ages \(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first \(n_R\) elements of the vector
@@
 \begin{equation*}
-\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,A-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}^T \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}^T \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
+\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,A-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}^T \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}^T \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A},
 \end{equation*}
 where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
+
 When there is autocorrelation with age, WHAM now assumes this applies only to ages \(a>1\) by default so that recruitment random effects are independent of those for the apparent survival of older age classes.
-So the general covariance structure for a given stock at ages \(a>1\) in region \(r\) is that of a two-dimensional first-order autoregressive (2DAR1) process
+The general covariance structure for a given stock at ages \(a>1\) in region \(r\) is a two-dimensional first-order autoregressive (2DAR1) process
 \begin{equation*}
@@
 \end{equation*}
-When the annual changes in abundance at age are treated deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\).
+If the annual changes in abundance at age are treated deterministically, then \(\boldsymbol{\varepsilon}_{y,a} = \mathbf{0}\).
 Since recruitment for a given stock currently only occurs in one region \(r_s\) there is only a single vector of time-varying recruitment random effects for each stock.
@@
 Initial numbers at age for each stock and region can be treated as age-specific fixed effects or with an equilibrium assumption as in previous versions of WHAM.
-For the equilibrium option there are two parameters for each stock: the stock-specific fully-selected fishing mortality rate \(\log \widetilde{F}\) and the recruitment in year 1 \(\log N_{1,1,r_s}\).
-A stock-specific equilibrium fishing mortality at age by fleet \({\tilde F}_{a,f}\) is the product of \(\widetilde{F}\) and the selectivity across fleets in the first year
-\begin{equation*}
-  \text{sel}_{1,a,f} = \frac{F_{1,a,f}}{\max_a \sum_{f=1}^{n_F} F_{1,a,f}}.
-\end{equation*}
-We use \(\widetilde{F}_{a,f}\) to define an equilibrium abundance per recruit by region at age \(a\) conditional on recruiting to each region
+For the equilibrium option there are two parameters for each stock: the stock-specific equilibrium fully-selected fishing mortality rate (\(\widetilde{F}_1\)) and the recruitment in year 1 (\(N_{1,1,r_s}\)).
+A stock-specific equilibrium fishing mortality at age by fleet (\(\widetilde{F}_{f,1,a}\)) is the product of \(\widetilde{F}_1\) and the selectivity across fleets in the first year
+\begin{equation*}
+  \text{sel}_{f,1,a} = \frac{F_{f,1,a}}{\max_a \sum_{f=1}^{n_F} F_{f,1,a}}.
+\end{equation*}
+We use \(\widetilde{F}_{f,1,a}\) to define an equilibrium abundance per recruit by region at age \(a\) conditional on recruiting to each region
 \begin{equation}\label{eq:npraa}
@@
 \left[\prod^{a-1}_{j=0}\mathbf{O}_{j}\right] \left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1} & a = A
- \end{array}
+ \end{array},
 \right.
 \end{equation}
-where \(\mathbf{O}_{j}\) is the equilibrium probability of surviving age \(a\) and occurring in each region and \(\mathbf{O}_{0} = \mathbf{I}\).
+where \(\mathbf{O}_{j}\) is the equilibrium proportions surviving age \(a\) and occurring in each region and \(\mathbf{O}_{0} = \mathbf{I}\), the identity matrix.
 Natural mortality and movement rates in the first year of the model are also used in Eq.
@@
 For the plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is a ``fundamental matrix'\,' derived using the matrix version of the geometric series (Kemeny and Snell 1960).
-Recall that recruitment for stock \(s\) only occur in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
-\[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}^T \mathbf{N}_{O,1,1}.\]
-
-The initial abundances at age can also be treated as independent or AR1 random effects. Defining the vector of initial abundance at age in region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
-\[\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}\]
-where
-\[  Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.\]
+Recall that recruitment for a given stock \(s\) only occurs in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
+\[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a,r_s,\cdot} N_{1,1,r_s},\]
+where \(\widetilde{\mathbf{O}}_{a,r_s,\cdot}\) is the vector of values in row \(r_s\) of \(\widetilde{\mathbf{O}}_{a}\).
+Initial abundance at age can also be estimated as random effects as described in Appendix B.
 
@@
 If survival and movement occur simultaneously, the parameters are estimated with a log link function and if they are separable, an additive logit link function (like a multinomial regression) is used.
-On the transformed scale, the general model for the movement parameter from region \(r\) to \(r'\) in season \(i\) and year \(y\) for individuals of age \(a\) is a linear function of an intercept or mean parameter (\(\theta_{r\rightarrow r',i}\)) and both random and environmental effects:
-\begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
-\end{equation*}
-The random effects \(\varepsilon_{r\rightarrow r',y,a,i}\) are season-, and region-to-region-specific and modeled most generally as 2DAR1 random effects with age and(or) year where the covariance is
-\begin{equation*}
-  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
-\end{equation*}
-similar to how WHAM models variation in survival, natural mortality, and selectivity.
-Effects of covariate \(E_k\) can be age-, season-, and region-to-region-specific \(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.
+On the transformed scale, the general model for the movement parameter from region \(r\) to \(r'\) in season \(i\) and year \(y\) for individuals of age \(a\) is a linear function of an intercept or mean parameter (\(\theta_{r\rightarrow r',i}\)) and both random and covariate effects:
+\begin{equation*}
+  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} X_{k,y}.
+\end{equation*}
+The random effects \(\varepsilon_{r\rightarrow r',y,a,i}\) are region-to-region- and season-specific, and modeled most generally as 2DAR1 random effects with age and(or) year, where the covariance is
+\begin{equation*}
+  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}.
+\end{equation*}
+This is similar to how WHAM models variation in survival, natural mortality, and selectivity.
+Covariate \(X_k\) effects (\(\beta_{r\rightarrow r',a,i,k}\)) can be region-to-region-, age-, and season-specific, and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.
 
@@
 Therefore, movement parameters would usually either need to be fixed or assumed to have some prior distribution, possibly based on external parameter estimates.
-WHAM includes prior distributions for the season and region-to-region specific (mean) movement parameters and treats them as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
-\begin{equation*}
-  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
+WHAM includes prior distributions for the region-to-region- and season-specific (mean) movement parameters and treats them as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
+\begin{equation*}
+  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{\gamma,r\rightarrow r',i}\right).
   \end{equation*}
@@
 \begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y}.
+  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} X_{k,y}.
   \end{equation*}
@@
 When random effects are used, the same 2DAR1 structure with age and year as described by Stock and Miller (2021) can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
-So the general model for natural mortality is
-\begin{equation*}
-  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
-\end{equation*}
-The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
-\begin{equation*}
-  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
-\end{equation*}
+We provide details in Appendix B.
 
@@
 The log-normal distributional assumption for aggregate catch observations is the same as Stock and Miller (2021), but the predicted catch is now a function of catch from each stock starting the year in each region and any seasonal movement.
-For a given stock and age, the vector of numbers captured by each fleet over the year are
-\[\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}^T \mathbf{N}_{O,s,y,a}\]
-(recall each fleet can harvest in only one region as described above). The vector of predicted numbers caught be each fleet across stocks is
-\[\widehat{\mathbf{N}}_{H,y,a} = \sum^{n_S}_{s=1} \widehat{\mathbf{N}}_{H,s,y,a}\]
+For a given stock \(s\) at age \(a\), the vector of numbers captured by each fleet over the year \(y\) is
+\[\mathbf{N}_{H,s,y,a} = \mathbf{H}_{s,y,a}^T \mathbf{N}_{O,s,y,a}\]
+(recall each fleet can harvest in only one region as described above). The vector of numbers caught by each fleet across stocks is
+\[\mathbf{N}_{H,y,a} = \sum^{n_S}_{s=1} \mathbf{N}_{H,s,y,a}\]
 and the vector of predicted aggregate catch by each fleet at age \(a\) is
-\[\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \widehat{\mathbf{N}}_{H,y,a}\]
+\[\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \mathbf{N}_{H,y,a},\]
 where \(\mathbf{c}_{y,a}\) is the vector of mean individual weight at age \(a\) for each fleet and the aggregate catch by fleet is
@@
 The log-aggregate catch observations for fleet \(f\) are normally distributed
-\[ \log C_{y,f} \sim \text{N}\left(\log \widehat {C}_{y,f}, \sigma^2_{y,f}\right).\]
-
-The predicted numbers caught for each fleet \(f\) (row \(f\) of \(\widehat{\mathbf{N}}_{H,y,a}\)) are used to make predicted age composition observations as described by Stock and Miller (2021).
+\[ \log C_{f,y} \sim \text{N}\left(\log \widehat {C}_{f,y}, \sigma^2_{f,y}\right),\]
+
+where \(\widehat C_{f,y}\) is an element of \(\widehat{\mathbf{C}}_y\).
+The predicted numbers caught for each fleet \(f\) (row \(f\) of \(\mathbf{N}_{H,y,a}\)) are used to make predicted age composition observations as described by Stock and Miller (2021).
 Since then, three additional likelihood options for age composition observations have been added: a logistic-normal with AR(1) correlation structure (Francis 2014), the alternative Dirichlet-multinomial parameterization described by Thorson et al.
@@
 
-Like fishing fleets, an index \(m\) can observe the population in a single region \(r_m\) at fraction of the year \(t_m\), the predicted abundance at \(t_m\) in region \(r_m\) is
-\[\widehat{N}_{s,y,a,m} = \mathbf{O}_{s,y,a,r_m}^T(t_m) \mathbf{N}_{O,s,y,a}\]
-where \(\mathbf{O}_{s,y,a,r_m}(t_m)\), the \(r_m\) column of the upper-left submatrix of Eq.
-\ref{eq:ptm_index}, are the probabilities of surviving and occurring in region \(r_m\) at time \(t_m\) given being alive in each region at the start of the year.
+An index \(m\) can observe any stocks that occur in a single region \(r_m\) at fraction of the year \(t_m\). The abundance at \(t_m\) in region \(r_m\) is
+\[N_{m,s,y,a} = \mathbf{O}_{s,y,a,\cdot,r_m}^T(t_m) \mathbf{N}_{O,s,y,a},\]
+where \(\mathbf{O}_{s,y,a,\cdot,r_m}(t_m)\), column \(r_m\) of the upper-left submatrix of Eq.
+\ref{eq:ptm_index}, is the vector of proportions surviving and occurring in region \(r_m\) at time \(t_m\) given being alive in each region at the start of the year.
 The predicted index at age is
-\[\widehat{I}_{m,y,a} = q_{m,y} \text{sel}_{m,y,a}w_{m,y,a}\sum^{n_S}_{s = 1}\widehat{N}_{s,y,a,m}\]
-where \(q_{m,y}\) is the catchability of the index in year \(y\), \(\text{sel}_{m,y,a}\) is the selectivity and \(w_{m,y,a}\) is the average weight of individuals at age \(a\) if the index is quantified in biomass and \(w_{m,y,a} = 1\) if the index is quantified in numbers.
+\[\widehat{I}_{m,y,a} = q_{m,y} \text{sel}_{m,y,a}w_{m,y,a}\sum^{n_S}_{s = 1}N_{m,s,y,a},\]
+where \(q_{m,y}\) is the catchability of the index in year \(y\), \(\text{sel}_{m,y,a}\) is the selectivity, and \(w_{m,y,a}\) is the average weight of individuals at age \(a\) if the index is quantified in biomass and \(w_{m,y,a} = 1\) if the index is quantified in numbers.
 Predicted age composition observations are functions of \(\widehat{I}_{m,y,a}\) as described by Stock and Miller (2021) and the likelihood options are the same as those for catch explained above.
-
-Catchability of the index can also be treated as logit functions of normal random effects and(or) environmental covariate effects
-\[log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}\]
-where \(\theta_{q,m}\) is an intercept or mean parameter, and \(u_{m}\) and \(l_{m}\) are the upper and lower bounds of catchability for index \(m\) (defaults are 0 and 1000).
-The general covariance structure for the annual random effects is AR1
-\[Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.\]
+We describe options to model temporal variation in catchability in Appendix B.
 
@@
 
-Currently a single fishing mortality reference point \(\widetilde F\) is estimated across stocks and regions and the fleet- and age-specific reference points is \(\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}\).
-Selectivity is determined across fleets and ages as in the equilibrium assumption for initial abundance at age except here it is a function of average F at age over a user-defined set of years
+WHAM calculates annual reference points using biological inputs each year and it also provides the same reference points under prevailing conditions where biological inputs are averaged over a user-defined set of recent years.
+A single fishing mortality reference point \(\widetilde{F}\) is estimated across stocks and regions and the fleet- and age-specific reference point is \(\widetilde{F}_{f,a} = \widetilde{F} \text{sel}_{f,a}\).
+Selectivity is determined across fleets and ages as in the equilibrium assumption for initial abundance at age except here it is a function of fishing mortality at age, possibly averaged over years,
 \begin{equation}\label{eq:spr_sel}
-  \text{sel}_{a,f} = \frac{\overline F_{a,f}}{\max_a \sum^{n_F}_{f=1}{\overline F}_{a,f}}
+  \text{sel}_{f,a} = \frac{\overline{F}_{f,a}}{\max_a \sum^{n_F}_{f=1}{\overline{F}}_{f,a}}.
 \end{equation}
 
-The equilibrium spawning stock biomass per recruit for a given stock in region \(r_s\) is defined as
+The equilibrium spawning stock biomass-per-recruit for a given stock in region \(r_s\) is defined as
 \begin{equation}\label{eq:ssbpr}
- \upphi(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{O}_{a,\cdot,r_s}(t_s) w_{a} m_{a}
+ \upphi(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{O}_{a,\cdot,r_s}(t_s) w_{a} m_{a},
 \end{equation}
-where \(w_{a}\) and \(m_{a}\) are the mean individual weight and probability of maturity at age \(a\), \(\widetilde{\mathbf{O}}_{a}\) are as described in Eq.
-\ref{eq:npraa}, and \(\mathbf{O}_{a}(t_s)\) is the \(n_R \times n_R\) upper-left sub-matrix of eq.
-\ref{eq:ptm_spawn} with the probabilities of surviving and occurring in each region \(r'\) at age \(a+t_s\) given starting in region \(r\) at age \(a\).
-The further subscripts \(r_s,\cdot\) and \(\cdot,r_s\) indicate row or column \(r_s\), respectively.
+where \(w_{a}\) and \(m_{a}\) are the mean individual weight and proportion mature at age \(a\), \(\widetilde{\mathbf{O}}_{a,r_s,\cdot}\) is row \(r_s\) of \(\widetilde{\mathbf{O}}_a\) in Eq.
+\ref{eq:npraa}, and \(\mathbf{O}_{a,\cdot,r_s}(t_s)\) is column \(r_s\) of the matrix of proportions surviving and occurring in each region at age \(a+t_s\) (the upper-left sub-matrix of eq.
+\ref{eq:ptm_spawn}).
 Using these rows and columns is required because of the assumption that spawning and recruitment only occur in region \(r_s\).
-
-The equilibrium spawning biomass per recruit (eq.
-\ref{eq:ssbpr}) is conditional on the region of recruitment \(r_s\).
-The equilibrium recruitment depends on the stock dynamics.
-WHAM currently only allows complete spawning region fidelity so that a stock only spawns and recruits in a single region.
-Therefore, the equilibrium vector of numbers at age 1 for each state, \(\widetilde {\mathbf{N}}_{1}\), will only be nonzero in the row corresponding to the spawning region (\(r_s\)).
-The matrices of probabilities of surviving and occurring in each region, \(\widetilde{\mathbf{O}}_{s,a}\) and \(\mathbf{O}_{s,a}(t_s)\), are functions of the fishing mortality rates for fleets in each region \(\widetilde F_{f,a}\).
-
-The matrix of equilibrium yield per recruit by region and fleet as a function of \(\widetilde F\) is calculated as
+For prevailing conditions, the years to average for \(w_a\), \(m_a\), and natural mortality and movement rates to parameterize \(\mathbf{O}_{a}(t_s)\) and \(\widetilde{\mathbf{O}}_{a}\), are defined by the user.
+
+The equilibrium yield-per-recruit as a function of \(\widetilde{F}\) for a given stock \(s\) is calculated as
 \begin{equation}\label{eq:ypr}
- \widetilde{\nu}({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{H}_{a} \mathbf{c}_{s,a}
+ \upnu({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{H}_{a} \mathbf{c}_{a},
 \end{equation}
-where \(\mathbf{c}_a\) is the vector of mean individual weight at age for each fleet, and \(\mathbf{H}_{s,a}\) is the submatrix of the probabilities of being captured in each fleet over the interval from \(a\) to \(a+1\), defined in eq.
+where \(\mathbf{c}_{a}\) is the vector of mean individual weight at age for each fleet, and \(\mathbf{H}_{a}\) is the submatrix of the proportions captured in each fleet over the interval from \(a\) to \(a+1\), defined in eq.
 \ref{eq:ptm}.
-
-As in previous versions of the WHAM package, reference points for prevailing conditions, average all of the inputs to the spawning biomass and yield per recruit calculations (weight, maturity, natural mortality, and movement) over a set of user-specified years as shown in eq.
-\ref{eq:spr_sel} for selectivity (e.g., last 5 years of the model).
-
-For \(X\%\) SPR-based reference points, we use a Newton method and iterate for steps \(j = 1,\ldots,10\)
+Other than mean individual weight, \(\upnu({\widetilde{F}})\) uses all of the same (averaged) inputs as \(\upphi({\widetilde{F}})\) in Eq.
+\ref{eq:ssbpr}.
+
+For reference points based on reductions of \(\upphi({\widetilde{F}})\) to \(p\%\) of the unfished value, \(\upphi(0)\) (\(p/100\) is known as the spawning potential ratio; SPR), we use a Newton-Raphson method and iterate for steps \(j = 1,\ldots,10\)
 \begin{equation}\label{eq:newton}
-  \log\widetilde{F}^{j} = \log\widetilde{F}^{j-1} - \frac{g\left(\log\widetilde{F}^{j-1}\right)}{g^{(1)}\left(\log\widetilde{F}^{j-1}\right)}
+  \log\widetilde{F}^{j} = \log\widetilde{F}^{j-1} - \frac{h\left(\log\widetilde{F}^{j-1}\right)}{h^{(1)}\left(\log\widetilde{F}^{j-1}\right)},
 \end{equation}
-where \(g(\log F)\) is the difference between the weighted sums of spawning biomass per recruit at \(F\) and \(X\)\% of unfished spawning biomass per recruit across stocks:
+where \(h(\log \widetilde{F})\) is the sum of weighted differences between \(\upphi({\widetilde{F}})\) and \(p\)\% of \(\upphi_s\left(0\right)\) across stocks:
 \begin{equation}\label{eq:newton-obj}
-  g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
+  h(\log \widetilde{F}) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(e^{\log \widetilde{F}}\right) - \frac{p}{100}\upphi_s\left(0\right)\right],
 \end{equation}
-where \(\upphi_s\left(F=0\right)\) is the equilibrium unfished spawning biomass per recruit.
-\(g^{(1)}(\log F)\) is the derivative of \(g\) with respect to \(\log F\), and the weights to use for each stock \(\lambda_s\) can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate ``static'' equilibrium spawning biomass and yield.
-
-When a Beverton-Holt or Ricker stock recruit relationship is assumed, an analogous Newton method is used to find \(\log F\) that maximizes yield for MSY-based reference points.
-which are also a functions of the equilibrium yield per recruit (Eq.
-\ref{eq:ypr}) and equilibrium recruitment.
-The function \(g(\log F)\) and \(g^{(1)}(\log F)\) in Eq.
-\ref{eq:newton} are the first and second derivatives of the sum of the stock-specific equilibrium yields with respect to \(\log F\) (analogous to that described by Miller et al.
+and \(h^{(1)}(\log \widetilde{F})\) is the derivative of \(h(\log \widetilde{F})\) with respect to \(\log \widetilde{F}\) (analogous to the approach for SPR-based reference points in Miller and Legault 2017).
+The weights for each stock (\(\lambda_s\)) can be specific values supplied by the user or defined to be the proportions of average recruitment for each stock where the years used to average are those the user defines to calculate prevailing equilibrium spawning biomass and yield reference points, which are the product of the defined average recruitment and \(\upphi({\widetilde{F}})\) and \(\upnu({\widetilde{F}})\), respectively.
+The Newton-Raphson method is used by WHAM internally to allow uncertainty in all estimated parameters to be propagated into uncertainty estimates for the reference points.
+
+When a Beverton-Holt or Ricker stock-recruit relationship is assumed, we use the derivative (with respect to \(\widetilde{F}\)) of the weighted sum of stock-specific equilibrium yields (the product of \(\upnu({\widetilde{F}})\) and equilibrium recruitment from the stock-recruit relationship) as \(h(\log \widetilde{F})\) to find \(\widetilde{F}\) which maximizes the weighted total yield across stocks for MSY-based reference points (\(F_{\text{MSY}}\), analogous to the approach for MSY-based reference points in Miller et al.
 2016b).
@@
 
-The projection options are generally the same as those for previous versions of WHAM (Miller and Stock 2020).
+The projection options are generally the same as those for previous versions of WHAM.
 When there is movement of any stocks, the user has the option to project and use any random effects for time-varying movement or use the average over user specified years, analogous to how natural mortality can be treated in the projection period.
@@
 
-Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging from the Gulf of Mexico to the Gulf of Maine.
+BSB are a temperate reef fish in the western Atlantic Ocean ranging the entire east coast of the United States.
 Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity (Able and Stanton Hales, Jr.
@@
 2013).
-Fish in this northern stock unit perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring (Musick and Mercer 1977).
+Fish in this NEUS stock perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring (Musick and Mercer 1977).
 Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude (Moser and Shepherd 2009).
 
-Prior to its 2023 peer-reviewed assessment, the NEUS black sea bass stock was assessed using the Age-Structured Assessment Program (ASAP) model (Legault and Restrepo 1999), a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
-Northern and southern components of the NEUS black sea bass stock ascribed to regions divided by the Hudson Canyon were separately modeled in ASAP (Figure \ref{fig:map}).
+Prior to its 2023 peer-reviewed assessment, the NEUS BSB stock was assessed using the Age-Structured Assessment Program (ASAP) model (Legault and Restrepo 1999), a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
+Northern and southern components of the NEUS BSB stock ascribed to regions divided by the Hudson Canyon were separately modeled in ASAP (Figure \ref{fig:map}).
 Results from the separate ASAP models were combined for a unit-stock assessment.
@@
 
-Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as ``Working Group'') composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above (NEFSC 2023).
-This assessment included the spatial features and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
-Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.
+Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as ``Working Group'') composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the BSB stock using the multi-stock and multi-region extension of WHAM described above (NEFSC 2023).
+This assessment included the two regions and two stock components and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous BSB assessments.
+Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on BSB.
 
@@
 
-The first year being modeled for the population is 1989 and the fishery and index data used in the model span from 1989 to 2021.
-The north and south stock components are modeled as separate populations that spawn and recruit in respective regions.
+The first year modeled for the population is 1989 and the fishery and index data used in the model span from 1989 to 2021.
+The north and south stock components are modeled as separate populations that spawn and recruit in respective regions and there are eight age classes with the last being a plus group (\(a \in \{1,\ldots,8+\}\)).
 We have observations for each of four total fishing fleets, where two fishing fleets (recreational and commercial) operate in each region.
@@
 The southern stock component is assumed to never move to the northern region.
-For the northern component, a proportion \(\mu_{\text{N}\rightarrow \text{S}}\) can move to the southern region each month during the last five months of the year, but no movement is allows from the south to the north during this period (Figure \ref{fig:migration-diagram}).
-During the first four intervals of the year a proportion \(\mu_{\text{S}\rightarrow \text{N}}\) the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
+For the northern component, a proportion \(\mu_{\text{N}\rightarrow \text{S}}\) can move to the southern region each month during the last five months of the year, but no movement is allowed from the south to the north during this period (Figure \ref{fig:migration-diagram}).
+During the first four intervals of the year a proportion \(\mu_{\text{S}\rightarrow \text{N}}\) of the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
 In the fifth interval (May), all northern component individuals remaining in the south are assumed to move back to the north for the subsequent spawning period.
@@
 for the portion of the year before spawning.
-As noted in the description of the general WHAM model, tagging data are not yet allowed.
+As noted in the description of the general WHAM model, tagging data are not yet included.
 However, the Working Group also fit a Stock Synthesis model (Methot and Wetzel 2013) which included tagging data and provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).
@@
 With movement as configured, the northern origin fish (ages 2+) can occur in the southern region on January 1.
-Estimating initial numbers at age as separate parameters can be challenging even in single-stock models, but for black sea bass the available data cannot distinguish the proportion of northern and southern component fish at each age in the southern region in the initial year of the model.
-Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component: an initial recruitment and an equilibrium fully-selected \(F\) (across all fleets) that determines the initial abundance at age in each region for each stock component.
-
-For subsequent years, all models allow AR1 correlation of recruitment and 2DAR1 correlation for apparent survival random effects for both the northern and southern stock components.
-Models with apparent survival random effects will model the transitions (conditional on both survival and movement) of abundances at age of northern origin fish in each region, but there appears to be little ability of the existing data from the southern region to distinguish apparent survival random effects for each stock component.
+Estimating initial numbers at age as separate parameters can be challenging even in single-stock models, but for BSB the available data cannot distinguish the proportion of northern and southern component fish at each age in the southern region in the initial year of the model.
+Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component (Eq.
+\ref{eq:npraa}).
+
+For subsequent years, all models have AR1 correlation of recruitment and 2DAR1 correlation for apparent survival random effects for both the northern and southern stock components.
+Models with apparent survival random effects will model the transitions (conditional on both survival and movement) of abundances at age of northern origin fish in each region, but there appears to be little information in the existing data from the southern region to distinguish apparent survival random effects for each stock component.
 All models we considered assume a very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model.
 Unique variance and correlation parameters for the recruitment and apparent survival random effects are estimated for the northern and southern components.
-
-\hypertarget{uncertainty-in-recreational-index-observations}{%
-\subsection*{Uncertainty in recreational index observations}\label{uncertainty-in-recreational-index-observations}}
-\addcontentsline{toc}{subsection}{Uncertainty in recreational index observations}
-
-The estimated coefficients of variation (CVs) provided by the analyses by the Working Group to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the Working Group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
-All models we considered allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
-Although lack of model convergence was observed in many fits for self-test simulations, the estimation of critical model output (e.g., SSB, fishing mortality) was reliable and robust the poor estimation of these scalars (Figure \ref{fig:self-test-fig}), and estimating these parameters rather than fixing them in the fit to the actual data allows uncertainty in model output to be more properly conveyed.
-
-\hypertarget{index-and-catch-age-composition-observations}{%
-\subsection*{Index and catch age composition observations}\label{index-and-catch-age-composition-observations}}
-\addcontentsline{toc}{subsection}{Index and catch age composition observations}
-
-The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual patterns and retrospective patterns.
-These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for six sets of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
 
@@
 
-All models include bottom temperature observations for the northern and southern regions from 1963 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius (NEFSC 2023).
+All models include observations of winter bottom temperature anomalies for the northern and southern regions from 1959 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius (NEFSC 2023).
+We provide details about the derivation of these observations in the Supplementary Materials.
 We retained the assumption from the peer-reviewed assessment that treated the latent bottom temperature covariates in each region as AR1 processes.
@@
 We fit 14 models with alternative assumptions about the effects of bottom temperature covariates, ranging from no effects to effects on both regions for either recruitment or natural mortality at age 1 (Table \ref{tab:model-desc-table}).
-These analyses derive from the hypothesis that bottom temperature affects overwinter survival of fish where the fish turn from age 0 to age 1 on January 1 (Miller et al.
+These analyses derive from the hypothesis that bottom temperature affects survival over the first winter where fish turn from age 0 to age 1 on January 1 (Miller et al.
 2016a).
@@
 
-All models treat annual recruitment as AR1 time-varying random effects with mean parameters for each stock component.
-However, Miller et al.
-(2018) showed how inferences of temperature effects on growth or maturity parameters can be very different whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
+We assume the covariate in year \(y\) affects recruitment in the same year because the covariate observations are from winter months.
+The fish are technically already 1 year old, but there are no observations of these individuals until later in the spring except possibly in fishery catches which are accumulated over the whole year.
+Expected log-recruitment for a given stock component is a linear function of bottom temperature
+\begin{equation}\label{eq:expected-recruitment}
+E\left(\log N_{y,1,r_s}|X_{y}\right) = \theta_{R} + \beta_{R} X_y.
+\end{equation}
+and the annual recruitment is
+\begin{equation}\label{eq:Rec-re}
+\log N_{y,1,r_s} = E\left(\log N_{y,1,r_s}|X_y\right) + \varepsilon_{y,1,r_s},
+\end{equation}
+where \(\varepsilon_{y,1,r_s}\) are the normally distributed random effects with AR1 correlation structure.
+
+Similarly, expected log-natural morality rate as a function of bottom temperature is
+\begin{equation}\label{eq:expected-M1}
+E\left(\log M_{y,1,r_s}|X_y\right) = \theta_{M,a,r_s} + \beta_{M} X_y.
+\end{equation}
+The mean log natural mortality rate is assumed to be \(\theta_{M,a,r} = \log(0.4)\) for all ages and regions as recommended by the Working Group.
+Miller et al.
+(2018) showed how inferences of temperature effects on growth or maturity parameters can be very different depending on whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
 Therefore, we also explored whether including temporal random effects on age 1 natural mortality affected inferences on corresponding temperature effects.
+The model for annual natural mortality for a given stock is
+\begin{equation}\label{eq:M-re}
+\log M_{y,1,r_s} = E\left(\log M_{y,1,r_s}|X_y\right) + \varepsilon_{M,y,1,r_s}.
+\end{equation}
+Because the bottom temperature anomalies are centered and \(E(\varepsilon_{M,y,1,r_s}) = 0\), mean log-natural mortality at age 1 over the time series should be approximately equal to \(\theta_{M,a,r_s}\) for models that include those effects.
+
 Initially, we included random effects on age 1 natural mortality for both the northern and southern stock components, but estimates of these random effects and corresponding variance for the southern component converged to 0 so these were not included in models presented here.
-
-We assume the covariate in year \(y\) affects recruitment in the same year because the covariate observations are from months January to March.
-The fish are technically already 1 year old, but there are no observations of these individuals until later in the year except possibly in fishery catches which are accumulated over the whole year.
-Expected log-recruitment for a given stock component is a linear function of bottom temperature
-
-\begin{equation}\label{eq:expected-recruitment}
-E\left(\log N_{y,1}|x_y\right) = \mu_{R} + \beta_{R} x_y.
-\end{equation}
-Similarly, expected log-natural morality as a function of bottom temperature is
-\begin{equation}\label{eq:expected-M1}
-E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y.
-\end{equation}
-Because age 1 fish for the northern component can exist in both regions after January 1 other the spawning period, natural mortality is acting in both regions for this stock.
-For models with covariate and(or) annual random effects for age 1 fish we assume them only in the northern regions for the northern component.
-The corresponding random effects are
-\begin{equation}\label{eq:Rec-re}
-\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \varepsilon_{y,1}
-\end{equation}
-and
-\begin{equation}\label{eq:M-re}
-\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \varepsilon_{M,y}.
-\end{equation}
-The constant or mean log-natural mortality rate is assumed to be \(\mu_{M,a} = \log(0.4)\) for all ages as recommended by the Working Group.
-Because the bottom temperature anomalies and random effects are centered at 0, mean log-natural mortality at age 1 over the time series should be approximately equal to \(\mu_{M,1}\) for models that include those effects.
+Because age 1 fish for the northern component can exist in both regions after January 1, other than during the spawning period, natural mortality is acting in both regions for this stock.
+For models with covariate and(or) annual random effects for age 1 fish, we use them only in the northern region for the northern component.
+
+\hypertarget{uncertainty-in-recreational-index-observations}{%
+\subsection*{Uncertainty in recreational index observations}\label{uncertainty-in-recreational-index-observations}}
+\addcontentsline{toc}{subsection}{Uncertainty in recreational index observations}
+
+The estimated coefficients of variation (CVs) provided by the analyses by the Working Group to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the Working Group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
+All models we considered allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
+Although lack of model convergence was observed in many fits for self-test simulations, the estimation of critical model output (e.g., SSB, fishing mortality) was reliable and robust to the poor estimation of these scalars (Figure \ref{fig:self-test-fig}), and estimating these parameters rather than fixing them in the fit to the actual data allows uncertainty in model output to be more properly conveyed.
+
+\hypertarget{index-and-catch-age-composition-observations}{%
+\subsection*{Index and catch age composition observations}\label{index-and-catch-age-composition-observations}}
+\addcontentsline{toc}{subsection}{Index and catch age composition observations}
+
+The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual and retrospective patterns.
+These analyses resulted in use of selectivity random effects for the northern fleet and indices, logistic-normal likelihoods for six sets of age composition observations, and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
 
@@
 
-We used a development version (\href{https://github.com/timjmiller/wham/tree/fb8b089}{commit fb8b089}) of the WHAM package prior to the release of version 2.0 for all results.
+We used a development version (\href{https://github.com/timjmiller/wham/tree/53e236b}{commit 53e236b}) of the WHAM package after the release of version 2.0 for all results.
 All code for fitting models and generating results can be found at \href{https://github.com/timjmiller/multi_wham_bsb}{github.com/timjmiller/multi\_wham\_bsb}.
@@
 Therefore, there are eight fits of each model with the time series reduced by zero to seven years.
-We calculated Mohn's \(\rho\) for SSB, and average \(F\) at ages 6 and 7.
-Absolute values of Mohn's \(\rho\) near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended.
+We calculated Mohn's \(\rho\) for SSB, and average fishing mortality rate at ages 6 and 7.
+Absolute values of Mohn's \(\rho\) near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended (Mohn 1999).
 As in Miller et al.
@@
 We performed a jitter analysis of the base model \(M_0\) and the best fitting model to investigate whether a local mimimum of the negative log-likelihood surface was obtained by the optimization.
-We used the jitter\_wham function in the WHAM package which by default simulates starting values from a multivariate normal distribution with mean and covariance defined by the MLEs and hessian-based covariance matrix from the fitted model.
-See the Supplementary Materials for more details.
+We used the \verb|jitter_wham| function in the WHAM package which by default simulates starting values from a multivariate normal distribution with mean and covariance defined by the MLEs and hessian-based covariance matrix from the fitted model (Supplementary Materials).
 
@@
 We estimated median relative bias of SSB across these simulations.
-See the Supplementary Materials for more details.
+Finally, we fitted three sensitivity models which differed from the best performing model, by 1) assuming no movement for either stock component, 2) movement parameters are fixed at the means for the prior distributions, and 3) a constant natural mortality rate is estimated across all regions and stocks (Supplementary Materials).
 
 \hypertarget{reference-points-and-projections}{%
-\subsection*{Reference points, and projections}\label{reference-points-and-projections}}
-\addcontentsline{toc}{subsection}{Reference points, and projections}
-
-We produced SPR-based BRP estimates (\(F_{40\%}\) and SSB\(_{40\%}\)) from the best performing model using a variety of options provided in WHAM.
-We generated annual BRP estimates based on the annual weight at age, maturity at age, natural mortality, fleet selectivity, and either the annual recruitment random effect (Eq.
+\subsection*{Reference points and projections}\label{reference-points-and-projections}}
+\addcontentsline{toc}{subsection}{Reference points and projections}
+
+We produced SPR-based reference point estimates (\(F_{40\%}\) and SSB\(_{40\%}\), setting \(p=40\) in eq.
+\ref{eq:newton-obj}) from the best performing model using a variety of options provided in WHAM.
+We generated annual reference point estimates based on the annual weight at age, maturity at age, natural mortality, fleet selectivity, and either the annual recruitment random effect (Eq.
 \ref{eq:Rec-re}) or the annual expected recruitment (Eq.
 \ref{eq:expected-recruitment}).
-We also calculated status as the ratios of annual average F and SSB to the corresponding reference points.
-Finally, we calculated ``prevailing'' BRPs using the average weight at age, maturity at age, M, fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
-All BRPs and status estimates are generated for each region as well as for the total stock area.
-
-We projected the best black sea bass model under three alternative scenarios for the bottom temperature covariate where 1) the AR1 time-series model continues into the projection years, 2) the average of the most recent 5 years is projected, and 3) bottom temperature increases in projection years following a prediction from a simple linear regression of the estimated bottom temperature anomalies over time.
+We also calculated status as the ratios of annual average fishing mortality rates and SSB to the corresponding reference points.
+Finally, we calculated ``prevailing'' reference points using the average weight, maturity, natural mortality rate, and fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
+All reference points and status estimates are generated for each region as well as for the total stock area.
+
+We projected the best BSB model under three alternative scenarios for the bottom temperature covariate where 1) the AR1 time-series model continues into the projection years, 2) the average of the most recent 5 years is projected, and 3) bottom temperature increases in projection years following a prediction from a simple linear regression of the estimated bottom temperature anomalies over time.
 We specified fishing mortality in projection years to be constant at the value in the last model year (2021).
-Average weight, natural mortality, movement, and maturity at age over the last 5 years are used in projection years whereas the AR1 processes for all numbers at age random effects are continued.
+Average weight, natural mortality, movement, and maturity at age over the last 5 years are used in projection years, whereas the AR1 processes for all numbers at age random effects are continued.
 
@@
 There was little evidence for also including random effects on age 1 natural mortality for the northern stock component.
-However, when more than three terminal years of data were removed, the evidence for adding age 1 natural mortality random effects for the northern component (\(M_8\)) increased is similar to adding bottom temperature effects on southern component recruitment (\(M_3\)).
+However, when more than three terminal years of data were removed, the evidence for adding age 1 natural mortality random effects for the northern component (\(M_8\)) increased similar to adding bottom temperature effects on southern component recruitment (\(M_3\)).
 
@@
 The effect of the anomalies on expected recruitment can be better observed by comparing the time series of the northern and southern components (Figure \ref{fig:E-R-SSB-F}).
-The expected recruitment for the southern component without any covariate effects is constant over time whereas the effect of the anomalies on expected recruitment for the northern component induces annual variation in expected recruitment.
+The expected recruitment for the southern component without any covariate effects is constant over time, whereas the effect of the anomalies on expected recruitment for the northern component induces annual variation in expected recruitment.
 The annual variation in expected recruitment (and the temperature anomalies) is correlated with the recruitment random effects of the northern component and therefore reduces the residual variation in the recruitment random effects.
@@
 
+The three sensitivity fits suggest the estimates of stock size and fishing mortality are fairly robust to whether the movement rates are fixed at the values from the companion Stock Synthesis fit or whether no movement is assumed (Figure \ref{fig:sensitivity-plots}).
+A constant natural mortality rate was estimable, but the estimated rate (0.72) is nearly twice that determined by the Working Group and resulted in generally higher stock sizes and lower fishing mortality rates.
+
 \hypertarget{reference-points-1}{%
@@
 
-Using the expected recruitment rather than the random effects as the multiplier with equilibrium spawning biomass per recruit (\(\upphi(\widetilde{F})\), eq.
+Using the expected recruitment rather than the random effects as the multiplier with \(\upphi(\widetilde{F})\) (eq.
 \ref{eq:ssbpr}) results in less variable estimates of annual SSB\(_{40\%}\) (Figure \ref{fig:annual-BRPs}: top row).
-Estimates were less variable for the southern stock component than the northern stock component mainly because the expected recruitment is constant for the former whereas that for the northern component varies via the effect of the bottom temperature anomalies.
-Temporal variation for the southern stock component is mainly caused by annual variation in the inputs to the \(\upphi(\widetilde{F})\) calculation and \(F_{40\%}\), but there is also some effect of the variation in the proportion of total recruitment for each stock component as used in the weighting terms to define the 40\% reduction in unfished spawning biomass per recruit (\(\lambda_s\) in Eq.
+Estimates were less variable for the southern stock component than the northern stock component mainly because the expected recruitment is constant for the former, whereas that for the northern component varies via the effect of the bottom temperature anomalies.
+Temporal variation for the southern stock component is mainly caused by annual variation in the inputs to the \(\upphi(\widetilde{F})\) calculation and \(F_{40\%}\), but there is also some effect of the variation in the proportion of total recruitment for each stock component as used in the weighting terms to define the 40\% reduction in \(\upphi(0)\) (\(\lambda_s\) in Eq.
 \ref{eq:newton-obj}).
-Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB\(_{40\%}\) for the southern stock component whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
+Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB\(_{40\%}\) for the southern stock component, whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
 
@@
 For similar reasons, estimates of uncertainty of \(F_{40\%}\) are also similar between the type of recruitment estimates (Figure \ref{fig:annual-F40-cvs}).
-Annual estimates for the northern stock component were lowest in the earliest years of the time series whereas those for the southern component are relatively stable over time.
+Annual estimates for the northern stock component were lowest in the earliest years of the time series, whereas those for the southern component are relatively stable over time.
 
@@
 When continuing the AR1 model for the bottom temperature anomalies in the projection years, the uncertainty grows and asymptotes to a level determined by the marginal variance of the process and the predicted value approaches the estimated marginal mean value of approximately 0 (Figure \ref{fig:R-BT-proj}).
-There is no uncertainty in the projected linear trend because the values were specified as known whereas there is a small level of uncertainty in the projections of the recent average based on the uncertainty in the estimated mean.
+There is no uncertainty in the projected linear trend because the values were specified as known, whereas there is a small level of uncertainty in the projections of the recent average based on the uncertainty in the estimated mean.
 Comparing projected recruitment for the northern and southern stock components, the projected expected recruitment has much lower uncertainty than the projected random effects regardless of whether bottom temperature affects recruitment.
-For the northern component, CVs of projected expected recruitment are smaller than those of recruitment random effect, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for the future bottom temperature anomalies are made (Figure \ref{fig:R-F-SSB-CVs}).
+For the northern component, CVs of projected expected recruitments are smaller than those of recruitment random effects, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for future bottom temperature anomalies (Figure \ref{fig:R-F-SSB-CVs}).
 
@@
 
-In our investigation of bottom temperature effects on recruitment and natural mortality and time-varying random effects on natural mortality at age 1, we found only including effects on recruitment for the northern stock component to be the best model with respect to AIC.
-Black sea bass in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north (McMahan et al.
+In our investigation of bottom temperature effects on recruitment and natural mortality, and time-varying random effects on natural mortality at age 1, we found only including effects on recruitment for the northern stock component to be the best model with respect to AIC.
+BSB in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north (McMahan et al.
 2020).
@@
 Although this might suggest making inferences on the population by using an ensemble of these and weighting by AIC (Burnham and Anderson 2002), the estimates of assessment outputs relevant for management (e.g., SSB and \(F\)) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
-Nevertheless, continued monitoring of effects on the southern component may be prudent.
-
-We showed that using expected recruitment rather than recruitment random effects to define BRPs can produce the same estimates but with lower uncertainty.
-Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same BRPs and even less uncertainty than those we showed using expected recruitment.
+Continued monitoring of temperature effects on the southern component and investigations of non-linear effects of temperature for both components may be prudent because of the possibility of an optimal temperature range for survival via effects on growth during the pre-recruit phase (Ware and Lambert 1985; Anderson 1988).
+Furthermore, our inferences about effects on natural mortality at age 1 may be influenced by our assumptions of constant median (or just constant, depending on the model) natural mortality rate and movement rates for all ages.
+Further investigations of variation in these parameters with age and effects of bottom temperature could improve our understanding of the demography for this stock.
+
+We showed that using expected recruitment rather than recruitment random effects to define reference points can produce the same estimates but with lower uncertainty.
+Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same reference points and even lower uncertainty.
 These results can be important for management when uncertainty in assessment output is used in making catch advice.
-The typical approach for defining stock and harvest status is to compare BRPs defined under prevailing conditions with current stock and harvest levels, but we could also make the comparisons with projected BRPs like those we demonstrated are possible.
+The typical approach for defining stock and harvest status is to compare reference points defined under prevailing conditions with current stock and harvest levels, but, as we demonstrated, it is also possible to make the comparisons of projected stock and harvest levels with projected reference points.
 The uncertainty in these ratios could also be used in defining catch advice.
 
-For black sea bass, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figure 6).
+Reference points under prevailing conditions use average values for recruitment and other inputs and MSY-based reference points are based on the stock-recruit relationship which is also typically constant over time.
+If reference points should be defined in such a way that they are less variable than the stock over time, expected inputs to the reference points (instead of annual values) would be more appropriate.
+Lower variability in reference points would presumably correlate with lower variability in catch advice which might be desirable (e.g., Cochrane et al.
+1998; Kell et al.
+2005), but the appropriate level of temporal variability in reference points and catch advice will depend on the characteristics of the fish stock, associated fisheries, and management structure (Holland 2010; Punt et al.
+2016).
+
+For BSB, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figures \ref{fig:R-BT-proj} and \ref{fig:F-SSB-proj}).
 In the projection period, the AR1 models for bottom temperature and recuitment revert to the long-term means of those processes.
-This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypothesis of increasing water temperature and changing environmental conditions (Hare et al.
+This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypotheses of increasing water temperature and changing environmental conditions (Hare et al.
 2016; Pershing et al.
 2021).
-Because of these concerns, the most recent stock assessment for yellowtail flounder used a change point analysis on environmental covariates to determine current conditions.
-The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections (NEFSC, 2024).
+Because of these concerns, the most recent stock assessment for yellowtail flounder (NEFSC 2024b) used a change point analysis on environmental covariates to determine current conditions.
+The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections.
 Unfortunately, this methodology still has limitations because it is unlikely that current environmental conditions are stable.
-Similarly, a linear increase of temperature is also not likely in the future (see past trends, top row of Figure 6).
-As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
+Similarly, a linear increase of temperature is also not likely in the future (top row of Figure \ref{fig:E-R-SSB-F}).
+As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in reference points and population projections becomes an important decision in the management process.
 
@@
 
-WHAM provides a large class of models for time- and age-varying random effects for population attributes and treatment of environmental covariates and their effects on populations using state-space methods (Maunder 2024).
+WHAM provides a large class of models for time- and age-varying random effects for population attributes and treatment of environmental covariates and their effects on populations using state-space methods.
 This framework accounts for the magnitude of differing uncertainties in observations and stochastic population dynamics processes.
-Version 2.0 extends these inferences to movement rates for multi-region models and stock (component) specific configuration of random and environmental covariate effects.
+Version 2.0 extends these inferences to movement rates for multi-region models and stock(component)-specific configuration of random and environmental covariate effects.
 
@@
 2019).
-We used estimates of movement parameters from a companion assessment model to define prior distributions for corresponding movement parameters in the state-space black sea bass model, but other than the tagging data, both of the assessment models used much the same data which can lead to inappropriate inferences.
+We used estimates of movement parameters from a companion assessment model to define prior distributions for corresponding movement parameters in the state-space BSB model, but other than the tagging data, both of the assessment models used much the same data which can lead to inappropriate inferences.
 In lieu of integrating an observation model for the tagging data, a better approach would be to estimate the movement parameters externally from just the tagging data.
@@
 Tagging data can also inform mortality rate parameters (Hampton 1991).
-It is well known that natural mortality is seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality from other assessment model parameters (Lee et al.
+It is well known that natural mortality rates are seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality rates from other assessment model parameters (Lee et al.
 2011; Clark 2022).
 Estimation of natural mortality may be even more challenging within state-space assessment models with their greater flexibility from inclusion of time-varying random effects.
-For example, we fixed the mean natural mortality for age 1 fish and Cadigan (2016) and Stock et al.
+For example, we fixed the mean natural mortality for age 1 fish and both Cadigan (2016) and Stock et al.
 (2021) also estimated natural mortality deviations.
@@
 Regardless, tagging observations can provide improved inferences and allowing tagging observations in WHAM should be a high priority even for models with a single stock and region.
-This extension would also be beneficial for black sea bass because the current assessment uses much the same data to estimate movement outside of the model as we are using to estimate the other parameters in the WHAM model.
-
-WHAM version 2.0 only allows models where each stock spawns and recruits in a single region however, some stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region (Kapur et al.
+
+WHAM version 2.0 only allows models where each stock spawns and recruits in a single region; however, some stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region (Kapur et al.
 2021).
@@
 
+Our application to BSB is a particular combination of the spatial modeling options described by Berger et al.
+(2024).
+Other combinations of the described options are possible in WHAM 2.0.
+For example, WHAM 2.0 can accommodate all of fleet structure options described in Table 2 by Berger et al.
+(2024) and we expect more of these options to become available in future versions of WHAM.
+
 \hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
+\section*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{section}{Acknowledgements}
 
 We express our gratitude to the many scientists who have collected and processed the data that went into the assessment model, including survey biologists, port samplers, observers and agers.
-We thank the black sea bass research track assessment working group and peer-reviewers for the work compiling data and determining many of the configurations of the black sea bass assessment model that we used for our analyses.
-We also thank Chris Legault for comments and suggestions that improved the clarity of this paper.
+We thank the BSB research track assessment working group and peer-reviewers for the work compiling data and determining many of the configurations of the BSB assessment model that we used for our analyses.
+We also thank Chris Legault, two anonymous reviewers, and the associate editor for comments and suggestions that improved the clarity of this paper.
+
+\pagebreak
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-anderson88}{}}%
+Anderson, J.T.
+1988.
+A review of size dependent survival during pre-recruit stages of fishes in relation to recruitment.
+Journal of Northwest Atlantic Fishery Science \textbf{8}: 55--66.
+doi:\href{https://doi.org/10.2960/J.v8.a6}{10.2960/J.v8.a6}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-arnason72}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-bergeretal24}{}}%
+Berger, A.M., Barceló, C., Goethel, D.R., Hoyle, S.D., Lynch, P.D., McKenzie, J., Dunn, A., Punt, A.E., Methot, R.D., Hampton, J., Porch, C.E., McGarvey, R., Thorson, J.T., A'mar, Z.T., Deroba, J.J., Elvarsson, B.Þ., Holmes, S.J., Howell, D., Langseth, B.J., Marsh, C., Maunder, M.N., Mormede, S., and Rasmussen, S.
+2024.
+Synthesizing the spatial functionality of contemporary stock assessment software to identify future needs for next generation assessment platforms.
+Fisheries Research \textbf{275}: 107008.
+doi:\href{https://doi.org/10.1016/j.fishres.2024.107008}{10.1016/j.fishres.2024.107008}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-bosleyetal22}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-cochraneetal98}{}}%
+Cochrane, K.L., Butterworth, D.S., De Oliveira, J.A.A., and Roel, B.A.
+1998.
+Management procedures in a fishery based on highly variable stocks and with conflicting objectives: Experiences in the {S}outh {A}frican pelagic fishery.
+Reviews in Fish Biology and Fisheries \textbf{8}(2): 177--214.
+doi:\href{https://doi.org/10.1023/A:1008894011847}{10.1023/A:1008894011847}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-commenges99}{}}%
@@
 doi:\href{https://doi.org/10.1093/icesjms/fsad133}{10.1093/icesjms/fsad133}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dupontaviceetal23}{}}%
+du Pontavice, H., Chen, Z., and Saba, V.S.
+2023.
+A high-resolution ocean bottom temperature product for the northeast {U}.{S}.
+Continental shelf marine ecosystem.
+Progress in Oceanography \textbf{210}: 102948.
+doi:\href{https://doi.org/10.1016/j.pocean.2022.102948}{10.1016/j.pocean.2022.102948}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-holland10}{}}%
+Holland, D.
+2010.
+Management strategy evaluation and management procedures: Tools for rebuilding and sustaining fisheries.
+OECD Publishing, Paris.
+doi:\href{https://doi.org/10.1787/5kmd77jhvkjf-en}{10.1787/5kmd77jhvkjf-en}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-kapuretal21}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-kelletal05}{}}%
+Kell, L.T., Pastoors, M.A., Scott, R.D., Smith, M.T., Van Beek, F.A., O'Brien, C.M., and Pilling, G.M.
+2005.
+Evaluation of multiple management objectives for northeast {A}tlantic flatfish stocks: Sustainability vs.
+Stability of yield.
+ICES Journal of Marine Science \textbf{62}(6): 1104--1117.
+doi:\href{https://doi.org/10.1016/j.icesjms.2005.05.005}{10.1016/j.icesjms.2005.05.005}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-kemenysnell60}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-maunder24}{}}%
-Maunder, M.N.
-2024.
-Towards a comprehensive framework for providing management advice from statistical inference using population dynamics models.
-Ecological Modelling \textbf{498}: 110836.
-doi:\href{https://doi.org/10.1016/j.ecolmodel.2024.110836}{10.1016/j.ecolmodel.2024.110836}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-mcmahanetal20}{}}%
@@
 In review.
-Factors affecting reliablity of state-space age-structured assessment models.
-Canadian Journal of Fisheries Science.
+An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models.
+Canadian Journal of Fisheries and Aquatic Sciences.
 
@@
 A state-space approach to incorporating environmental effects on recruitment in an age-structured assessment model with an application to southern {New England} yellowtail flounder.
-Canadian Journal of Fisheries Science \textbf{73}(8): 1261--1270.
+Canadian Journal of Fisheries and Aquatic Sciences \textbf{73}(8): 1261--1270.
 doi:\href{https://doi.org/10.1139/cjfas-2015-0339}{10.1139/cjfas-2015-0339}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-millerlegault17}{}}%
+Miller, T.J., and Legault, C.M.
+2017.
+Statistical behavior of retrospective patterns and their effects on estimation of stock and harvest status.
+Fisheries Research \textbf{186}: 109--120.
+doi:\href{https://doi.org/10.1016/j.fishres.2016.08.002}{10.1016/j.fishres.2016.08.002}.
 
@@
 Temporal and environmental variation in growth and maturity and effects on management reference points of {Georges Bank Atlantic} cod.
-Canadian Journal of Fisheries Science \textbf{75}(12): 2159--2171.
+Canadian Journal of Fisheries and Aquatic Sciences \textbf{75}(12): 2159--2171.
 doi:\href{https://doi.org/10.1139/cjfas-2017-0124}{10.1139/cjfas-2017-0124}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-millerstock20}{}}%
-Miller, T.J., and Stock, B.C. 2020. The {Woods Hole Assessment Model} ({WHAM}). https://timjmiller.github.io/wham.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nefsc24}{}}%
-NEFSC. 2024. Butterfish research track assessment report. US Dept Commer Northeast Fish Sci Cent Ref Doc. 24-03; 191 p.
+NEFSC. 2024a. Butterfish research track assessment report. US Dept Commer Northeast Fish Sci Cent Ref Doc. 24-03; 191 p.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nefsc24a}{}}%
+NEFSC. 2024b. The yellowtail flounder research track assessment report. {Available} at https://www.fisheries.noaa.gov/inport/item/26539. Accessed on 1/10/2025.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-puntetal16}{}}%
+Punt, A.E., Butterworth, D.S., Moor, C.L.
+de, De Oliveira, J.A.A., and Haddon, M.
+2016.
+Management strategy evaluation: Best practices.
+Fish and Fisheries \textbf{17}(2): 303--334.
+doi:\href{https://doi.org/10.1111/faf.12104}{10.1111/faf.12104}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-puntetal20}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-warelambert85}{}}%
+Ware, D.M., and Lambert, T.C.
+1985.
+Early life history of {A}tlantic mackerel (\emph{{S}comber scombrus}) in the southern {G}ulf of {S}t.
+{L}awrence.
+Canadian Journal of Fisheris and Aquatic Sciences \textbf{42}(3): 577--592.
+doi:\href{https://doi.org/10.1139/f85-075}{10.1139/f85-075}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-yingetal11}{}}%
@@
 
-\caption{The Northwest Atlantic coastal area where the NEUS black sea bass stock occurs.
-Red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations (after Miller et al.
+\caption{The Northwest Atlantic coastal area where the Northeast US black sea bass stock occurs.
+The red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations (after Miller et al.
 2016a) and the bathymetric contour is the 400 m isocline.}\label{fig:map}
@@
 
-\caption{Diagram of intervals within the year and configuation of the dynamics of each component of the BSB population.}\label{fig:migration-diagram}
+\caption{Diagram of intervals within the year and configuation of the dynamics of each component of the black sea bass population.}\label{fig:migration-diagram}
 \end{figure}
@@
 
-\caption{Prior (black) and posterior (red) distributions of movement of the northern stock component from north to south (top) and south to north (bottom).}\label{fig:move-prior-posterior}
+\caption{Prior (black) and posterior (red) distributions of the movement parameter for the northern stock component from north to south (top) and south to north (bottom).
+Vertical lines indicate the prior and posterior estimates.}\label{fig:move-prior-posterior}
 \end{figure}
@@
 \caption{Annual observations and posterior estimates of bottom temperature anomalies (top row), annual expected and random effect recruitment estimates (second row) and SSB and average fishing mortality at ages 6 and 7 from model $M_1$.
-Columns define estimates regionally (spawning region for SSB) and toal SSB and fishing mortality across regions.
+Columns define estimates regionally (spawning region for SSB) and total SSB and fishing mortality across regions.
 Polygons and vertical lines represent 95\% confidence intervals.}\label{fig:E-R-SSB-F}
@@
 
-\caption{Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio, equilibrium SSB at $F_{40\%}$ based on annual inputs to $\upphi(\widetilde{F})$ calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to $\upphi(\widetilde{F})$ averaged over the last 5 years and recruitment is averaged over the time series.
+\caption{Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio ($F_{40\%}$), equilibrium SSB at $F_{40\%}$ based on annual inputs to $\upphi(\widetilde{F})$ calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to $\upphi(\widetilde{F})$ averaged over the last 5 years and recruitment is averaged over the time series.
 Red and blue in first 4 rows and black and gray in final row indicate alternative annual estimates of recruitment (random effect or expected).
@@
 \caption{Annual estimates of bottom temperature in the northern region and alternative recruitment estimates (random effect or expected) by region.
-Estimates from years after 2021 are from projections of model $M1$ under three alternative assumptions for the bottom temperature anomalies in the northern region.
-Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:R-BT-proj}
+Estimates from years after 2021 are from projections of model $M_1$ under three alternative assumptions for the bottom temperature anomalies in the northern region.
+Vertical dotted lines indicate the last year of data and polygons represent 95\% confidence intervals.}\label{fig:R-BT-proj}
 \end{figure}
@@
 Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:F-SSB-proj}
+Vertical dotted lines indicate the last year of data and polygons represent 95\% confidence intervals.}\label{fig:F-SSB-proj}
 \end{figure}
@@
 Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region and average $\upphi(\widetilde{F})$ inputs from the last 5 years of the unprojected model.
-Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:brps-proj}
+Vertical dotted lines indicate the last year of data and polygons represent 95\% confidence intervals.}\label{fig:brps-proj}
 \end{figure}
@@
 \centering
-\caption{\label{tab:model-desc-table}Assumptions for temperature effects and random effects for age 1 natural mortality for each model.}
+\caption{\label{tab:model-desc-table}Assumptions for temperature effects and random effects for age 1 natural mortality rates ($M$) for each model.}
 \centering
@@
 \renewcommand\thetable{A\arabic{table}}
-\input{symboltable.tex}
+\input{symboltable_R2.tex}
+
+\hypertarget{appendix-b}{%
+\section*{Appendix B}\label{appendix-b}}
+\addcontentsline{toc}{section}{Appendix B}
+
+New features of WHAM 2.0 that are not used in the application to BSB are described here..
+
+\hypertarget{simultaneous-movement-and-mortality}{%
+\subsection*{Simultaneous movement and mortality}\label{simultaneous-movement-and-mortality}}
+\addcontentsline{toc}{subsection}{Simultaneous movement and mortality}
+
+When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
+We obtain the PTM over an interval \(i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
+\begin{equation*}
+\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
+\end{equation*}
+The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region.
+Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero.
+For two regions and one fleet operating in each region:
+\begin{equation*}
+ \mathbf{A}_{y,a,i} = \begin{bmatrix}
+ a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{1,y,a,i} & 0 & M_{y,a,1} \\
+ \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{2,y,a,i} & M_{y,a,2} \\
+ 0 & 0 & 0 & 0 & 0 \\
+ 0 & 0 & 0 & 0 & 0 \\
+ 0 & 0 & 0 & 0 & 0
+ \end{bmatrix},
+\end{equation*}
+where \(a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + M_{y,a,r} + \sum^{n_F}_{f:r_f=r}F_{f,y,a,i})\).
+When there is one region, \(n_F\) fleets, and \(\delta_i = 1\), exponentiating the instantaneous rate matrix results in the PTM defined in Eq.
+\ref{eq:ptm_1_region}.
+
+\hypertarget{initial-abundance-at-age-as-random-effects}{%
+\subsection*{Initial abundance at age as random effects}\label{initial-abundance-at-age-as-random-effects}}
+\addcontentsline{toc}{subsection}{Initial abundance at age as random effects}
+
+The initial abundances at age can also be treated as independent or AR1 random effects. Defining the vector of initial abundance at age in region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
+\begin{equation*}
+\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r},
+\end{equation*}
+where
+\begin{equation*}
+Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.
+\end{equation*}
+
+\hypertarget{general-natural-mortality-model}{%
+\subsection*{General natural mortality model}\label{general-natural-mortality-model}}
+\addcontentsline{toc}{subsection}{General natural mortality model}
+
+The general model for natural mortality at age \(a\) for a given stock in region \(r\) is a function of random and covariate effects
+\begin{equation}\label{eq:M_model}
+  \log M_{y,a,r} = \theta_{M,a,r} + \varepsilon_{M,y,a,r} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} X_{k,y} = E\left(\log M_{y,a,r}|X_{k,y}\right) + \varepsilon_{M,y,a,r}.
+\end{equation}
+The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
+\begin{equation*}
+  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
+\end{equation*}
+
+\hypertarget{general-catchability-model}{%
+\subsection*{General catchability model}\label{general-catchability-model}}
+\addcontentsline{toc}{subsection}{General catchability model}
+
+Catchability of each index can be treated as logit functions of normal random effects and(or) environmental covariate effects
+\begin{equation*}
+\log \left(\frac{q_{m,y}-l_m}{u_m-q_{m,y}}\right) = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} X_{k,y},
+\end{equation*}
+where \(\theta_{q,m}\) is an intercept or mean parameter, and \(u_{m}\) and \(l_{m}\) are the upper and lower bounds of catchability for index \(m\) (defaults are 0 and 1000).
+The general covariance structure for the annual random effects is AR1
+\begin{equation*}
+Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.
+\end{equation*}
 
@@
 
-\hypertarget{supplemental-materials}{%
-\section*{Supplemental Materials}\label{supplemental-materials}}
-\addcontentsline{toc}{section}{Supplemental Materials}
+\hypertarget{supplementary-materials}{%
+\section*{Supplementary Materials}\label{supplementary-materials}}
+\addcontentsline{toc}{section}{Supplementary Materials}
 
@@
 The roots of the matrices are calculated using an eigen decomposition of the matrices
-\[ \boldsymbol{\mu}_k =  \left(\boldsymbol{\mu}_k^*\right)^{z_k} = \mathbf{V}_k \mathbf{D}_k^{z_k} \mathbf{V}_k^{-1}\]
+\[ \boldsymbol{\mu}_k =  \left(\boldsymbol{\mu}_k^*\right)^{z_k} = \mathbf{V}_k \mathbf{D}_k^{z_k} \mathbf{V}_k^{-1},\]
 where \(z_1 = 1/5\) for and \(z_2 = 1/4\), and \(\mathbf{V}_{k}\) and \(\mathbf{D}_{k}\) are the matrix of eigenvectors (columnwise) and the diagonal matrix of corresponding eigenvalues of \(\boldsymbol{\mu}^*_k\).
@@
 
+\hypertarget{bottom-temperature-anomalies}{%
+\subsection*{Bottom temperature anomalies}\label{bottom-temperature-anomalies}}
+\addcontentsline{toc}{subsection}{Bottom temperature anomalies}
+
+The Working Group created bottom temperature observations from a high resolution ocean bottom temperature product by du Pontavice et al.
+(2023).
+The annual observations for each region are defined by the average over all spatial bottom temperature values for February and March by region and year.
+Similarly, the Working Group calculated standard errors from the standard deviation of all values in the region and the total number of values for a given year.
+We created regional bottom temperature anomalies by subtracting means for each region across all years.
+
 \hypertarget{diagnostics}{%
@@
 The 50 jitter fits demonstrated that a local minimum was obtained for the original fit of model \(M_0\) (Figure \ref{fig:jitter-M0}).
-Some lower NLLs were obtained with unacceptable gradients, but a slightly lower NLL was found with a satisfactory gradient and with and invertible hessian.
-We therefore refit model \(M_0\) and all remaining models using the better parameter estimates as initial values.
+One lower NLL was obtained with unacceptable gradients (No.~25), but a slightly lower NLL was found with a satisfactory gradient for 3 of the jitters (Nos.
+9, 13, 29).
+However, one of the jitter fits (No.~9) did not provide a non-zero estimate of the variance parameter for one of the indices and the other two provided identical results and we refit model \(M_0\) and all remaining models using the better parameter estimates as initial values.
 
@@
 \caption{Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_0$.
-Black jitters had maximum absolute gradient values < $10^{-10}$, grey jitters had values > $10^{-10}$ and < 1, and red jitters had values > 1.}\label{fig:jitter-M0}
+Black jitters had maximum absolute gradient values < $10^{-2}$ and red jitters had values > 1.}\label{fig:jitter-M0}
 \end{figure}
@@
 \caption{Median relative error of SSB (Total and by stock component) and total fully-selected fishing mortality for estimation models fitted to simulated data from model $M_1$ where the observation variance of log-indices are fixed and estimated.
-Black and Red dashed lines represent the median of the annual medians and the median across all annual relative errors, respectively.}\label{fig:self-test-fig}
+Black and Red dashed lines represent the median of the annual medians and the median across all annual relative errors, respectively.
+Vertical lines represent 95\% confidence intervals.}\label{fig:self-test-fig}
 \end{figure}
@@
 \caption{Coefficients of variation for estimates of alternative recruitment estimates (random effects or expected), average fishing mortality at age 6 and 7, and SSB by region and in total from model $M_1$.
-Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:R-F-SSB-CVs}
+Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data.}\label{fig:R-F-SSB-CVs}
 \end{figure}
@@
 \caption{Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-F40-cvs}
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data.}\label{fig:annual-F40-cvs}
 \end{figure}
@@
 \caption{Coefficients of variation for annual equilibrium SSB$_{40\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations and alternative annual recruitment types.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-SSB40-cvs}
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data.}\label{fig:annual-SSB40-cvs}
 \end{figure}
@@
 Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:F-status-proj}
+Vertical dotted lines indicate the last year of data and polygons represent 95\% confidence intervals.}\label{fig:F-status-proj}
 \end{figure}
@@
 Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:SSB-status-proj}
+Vertical dotted lines indicate the last year of data and polygons represent 95\% confidence intervals.}\label{fig:SSB-status-proj}
 \end{figure}
@@
 \caption{Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\bar{F}_{40\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-F-status-cvs}
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data.}\label{fig:annual-F-status-cvs}
 \end{figure}
@@
 \caption{Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-SSB-status-cvs}
-\end{figure}
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data.}\label{fig:annual-SSB-status-cvs}
+\end{figure}
+
+\begin{landscape}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{SSB_F_sensitivity_plots}
+
+}
+
+\caption{Estimates of annual SSB and fishing mortality rates for the best performing model $M_1$ and models that are otherwise the same except where  1) movement rates are fixed at the means for the prior distribution, 2) a constant natural mortality rate is estimated, or 3) there is no movement for either stock component.}\label{fig:sensitivity-plots}
+\end{figure}
+\end{landscape}
 

### 2025-06-17 paper/symboltable_R2.tex
revisions complete

\begin{longtable}[c]{r p{0.85\textwidth}}
\caption{Definition of terms.\label{symbols}}%
\\ \hline \hline \endfirsthead 
\caption[]{(Continued)} %
\\ \hline \hline \endhead
 \hline \hline \endfoot
$a$ & Age class indicator\\
$A$ & Last age class (``plus group'')\\
$a_{y,a,i,r}$ & Hazard (or negative sum of the instantaneous rates of mortality and movement) from the state corresonding to being alive in region $r$ for year $y$, age $a$ and seasonal interval $i$\\
$\mathbf{A}_{y,a,i}$ & Instantaneous rate matrix for seasonal interval $i$, year $y$, and age $a$ ($(n_R + n_F + 1) \times (n_R + n_F + 1)$)\\
$\mathbf{c}_{a}$ & Vector (length $n_F$) of mean individual weight at age $a$ for each fleet to use for $\upnu(\widetilde{F})$\\
$\mathbf{c}_{y,a}$ & Vector (length $n_F$) of mean individual weight at age $a$ in year $y$ for each fleet\\
$C_{f,y}$ & Observed aggregate catch for fleet $f$ in year $y$\\
$\widehat C_{f,y}$ & Predicted aggregate catch for fleet $f$ in year $y$\\
$\widehat{\mathbf{C}}_{y,a}$ & Vector (length $n_F$) of predicted biomass captured at age $a$ in year $y$ by each fleet across all stocks\\
$\widehat{\mathbf{C}}_y$ & Vector (length $n_F$) of predicted aggregate catch for each fleet in year $y$\\
$D_{y,a,i}$ & Proportion dying from natural mortality over the seasonal interval $i$ given alive (1 region model)\\
$\mathbf{D}_{y,a,i}$ & Vector (length $n_R$) of proportions dying from natural mortality over the seasonal interval $i$ given alive  in each region at the start of the interval\\
$f$ & Fishing fleet indicator\\
$F_{f,1,a}$ & Fishing mortality rate at age $a$ for fleet $f$ in the first year\\
$F_{f,y,a,i}$ & Fishing mortality rate for fleet $f$ in year $y$ at age $a$ in seasonal interval $i$\\
$F_{\text{MSY}}$ & Fully-selected fishing mortality rate that maximizes equilibrium yield\\
$\widetilde{F}$ & Equilibrium fully-selected fishing mortality rate parameter for equilibrium spawning-biomass-per-recruit and yield-per-recruit\\
$\widetilde{F}_1$ & Equilibrium fully-selected fishing mortality rate parameter for equilibrium abundance at age in the first year\\
$\widetilde{F}_{f,a}$ & Equilibrium fully-selected fishing mortality rate at age $a$ for fleet $f$ for equilibrium spawning-biomass-per-recruit and yield-per-recruit\\
$\widetilde{F}_{f,1,a}$ & Equilibrium fishing mortality rate at age $a$ for fleet $f$ for equilibrium abundance at age in the first year\\
$\overline{F}_{f,a}$ & Average fishing mortality rate for fleet $f$ at age $a$ over user-specified set of years\\
$g(\mu_{r\rightarrow r',y,a,i})$ & Link function for movement rate $\mu_{r\rightarrow r',y,a,i}$\\
$h(\log \widetilde{F})$ & Function used in Newton-Raphson method to determine reference points\\
$h^{(1)}(\log \widetilde{F})$ & Derivative of $h(\log \widetilde{F})$ with respect to $\log \widetilde{F}$\\
$H_{y,a,i}(r,f)$ & Proportion captured in fleet $f$ given beginning the seasonal interval alive in region $r$ for year $y$, age $a$, and seasonal interval $i$ \\
$\mathbf{H}_{a}$ & Matrix ($n_R \times n_F$) of equilibrium proportions captured from age $a$ to $a+1$ in each fleet (columns) given alive on January 1 in each region (rows)\\
$\mathbf{H}_{y,a,i}$ &  Matrix ($n_R \times n_F$) of proportions captured in each fishing fleet for year $y$, age $a$, and seasonal interval $i$ given alive in region $r$ at the beginning of the interval (submatrix of $\mathbf{P}_{y,a,i}$)\\
$\mathbf{H}_{s,y,a}$ &  Matrix ($n_R \times n_F$) of proportions captured in each fishing fleet throughout year $y$, age $a$ for stock $s$ (submatrix of $\mathbf{P}_{y,a}$)\\
$i$ & Seasonal interval indicator \\
$\widehat I_{m,y,a}$ & Predicted relative abundance index for survey $m$ in year $y$ at age $a$ \\
$\mathbf{I}$ & $n_R \times n_R$ identity matrix \\
$\mathbf{I}_{H}$ & $n_F \times n_F$ identity matrix \\
$K$ & Number of seasonal intervals in the annual time step\\
$l_{m}$ & Lower bound for index $m$ catchability\\
$m$ & Index observation indicator\\
$M_{y,a}$ &  Natural mortality rate in year $y$ at age $a$ (1 region model)\\
$M_{y,a,r}$ & Natural mortality rate in region $r$ and year $y$ at age $a$\\
$n_E$ & Number of environmental covariates\\
$n_F$ & Number of fishing fleets\\
$n_R$ & Number of regions\\
$N_{y,a,r}$ & Abundance at age $a$ in year $y$ alive in region $r$ on January 1\\
$N_{m,s,y,a}$ & Abundance of stock $s$ in year $y$ at age $a$ at $t_m$ in region $r_m$\\
$\mathbf{N}_{y,a}$ & Vector (length $n_R + n_F + 1$) of abundances at age $a$ in year $y$ in each state (alive in each region, dead in each fleet and dead due to natural mortality)\\
$\mathbf{N}_{O,y,a}$ & Vector (length $n_R$) of abundance at age $a$ in year $y$ alive in each region on January 1 for a given stock\\
$\mathbf{N}_{O,s,y,a}$ & Vector (length $n_R$) of abundance at age $a$ in year $y$ alive in each region on January 1 for stock $s$\\
$\mathbf{N}_{O,1,r}$ & Vector (length $A$) of abundance by age in region $r$ in the first year\\
$\mathbf{N}_{H,s,y,a}$ & Vector (length $n_F$) of numbers of stock $s$ at age $a$ in year $y$ captured by each fleet\\
$\mathbf{N}_{H,y,a}$ & Vector (length $n_F$) of numbers at age $a$ in year $y$ captured by each fleet across all stocks\\
$\widetilde{\mathbf{N}}_{1}$ & Vector (length $n_R$) of equilibrium numbers at age 1 (only non-zero in element corresponding to $r_s$ for stock $s$\\
$O_{y,a,i}(r,r')$ & Proportions surviving and occurring in region $r'$ given beginning the seasonal interval alive in region $r$ for year $y$, age $a$ and seasonal interval $i$\\
$\mathbf{O}_{a}$ & Matrix ($n_R \times n_R$) of equilibrium proportions surviving from age $a$ to $a+1$ and occurring in each region (columns) given alive on January 1 in each region (rows)\\
$\mathbf{O}_{y,a,i}$ & Matrix ($n_R \times n_R$) of proportions surviving seasonal interval $i$ and occurring in each region for year $y$, age $a$ (submatrix of $\mathbf{P}_{y,a,i}$)\\
$\mathbf{O}_{y,a,\cdot,r_s}(t_s)$ & Vector (length $n_R$) of proportions surviving up to time $t_s$ and occurring in region $r_s$ given being alive in each region at the start of the year ($r_s$ column of the upper-left submatrix of Eq. \ref{eq:ptm_spawn})\\
$\mathbf{O}_{s,y,a,\cdot,r_m}(t_m)$ & Vector (length $n_R$) of proportion at age $a$ in year $y$ surviving up to time $t_m$ and occurring in region $r_m$ given being alive in each region at the start of the year for stock $s$ ($r_m$ column of the upper-left submatrix of Eq. \ref{eq:ptm_index})\\
$\mathbf{O}_{a}(t_s)$ & Matrix ($n_R \times n_R$) of equilibrium proportions surviving from age $a$ to $a+t_s$ and occurring in each region (columns) given alive on January 1 in each region (rows)\\
$\mathbf{O}_{a,\cdot,r_s}(t_s)$ & Column $r_s$ of $\mathbf{O}_{a}(t_s)$\\
$\widetilde{\mathbf{O}}_{a}$ & Matrix ($n_R \times n_R$) of equilibrium proportions alive in each region at age $a$ (columns) given recruitment in each region (rows)\\
$\widetilde{\mathbf{O}}_{a,r_s,\cdot}$ & Row vector $r_s$ (length $n_R$) of $\widetilde{\mathbf{O}}_{a}$\\
$p$ & Percent reduction of $\upphi(\widetilde{F})$ from unfished value for SPR-based reference points\\ 
$\mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K)$ & Probability transition matrix for year $y$ and age $a$ over seasonal intervals $\delta_1,\ldots, \delta_K$\\
$\mathbf{P}_{y,a,i}$ & Probability transition matrix for year $y$, age $a$, and seasonal interval $i$\\
$q_{m,y}$ & Catchability of index $m$ in year $y$\\
$r$ & Region indicator\\
$r_f$ & Region where fleet $f$ operates\\
$r_s$ & Region where stock $s$ spawns and recruits\\
$s$ & Stock (component) indicator\\
$S_{y,a,i}$ & Proportion surviving seasonal interval $i$ in year $y$, at age $a$ (1 region model)\\
$\mathbf{S}_{y,a,i}$ & Matrix ($n_R \times n_R$) of of proportions surviving in each region over the seasonal interval $i$, year $y$, age $a$ (prior to movement)\\
$t_m$ & Duration between the beginning of the year and when index $m$ observes the population\\
$t_s$ & Fraction between the beginning of the year and when spawning occurs for stock $s$\\
$u_{m}$ & Upper bound for index $m$ catchability\\
$w_{m,y,a}$ & Average weight of individuals at age $a$ for index $m$ if the index is quantified in biomass, otherwise it is one\\
$w_{y,a}$ & Mean individual weight at spawning for a given stock at age $a$ in year $y$\\
$X_{y}$ & Latent bottom temperature that may affect a black sea bass stock component in year $y$\\
$X_{k,y}$ & Latent environmetal covariate $k$ affecting the population in year $y$\\
$y$ & Year indicator\\
$Z_{y,a,i}$ & Total mortality rate in year $y$ at age $a$ in seasonal interval $i$ (1 region model)\\
$Z_{y,a,i,r}$ & Total mortality rate in region $r$ and year $y$ at age $a$ in seasonal interval $i$\\
$\text{mat}_{y,a}$ & Proportion mature at age $a$ in year $y$ for a given stock\\
$\text{sel}_{f,1,a}$ & Selectivity scalar for fleet $f$ at age $a$ in the first year\\
$\text{sel}_{f,a}$ & Selectivity scalar for fleet $f$ at age $a$ for SPR-based reference points\\
$\text{sel}_{m,y,a}$ &  Selectivity of index $m$ at age $a$ in year $y$ \\
$\text{SSB}_{y}$ & Spawning stock biomass for a given stock in region $r_s$ in year $y$\\
$\beta_M$ & Effect of bottom temperature on age 1 natural mortality rate for a given black sea bass stock component\\
$\beta_{M,0}$ & Median natural mortality rate parameter for a given black sea bass stock component\\
$\beta_{M,r,a,k}$ & Effect of environmental covariate $k$ on natural mortality rate in region $r$ at age $a$\\
$\beta_{r \rightarrow r',a,i,k}$ & Effect of environmental covariate $k$ on movement from region $r$ to $r'$ at age $a$ in seasonal interval $i$\\
$\beta_R$ & Effect of bottom temperature on recruitment for a given black sea bass stock component\\
$\beta_{R,0}$ & Median recruitment parameter for a given black sea bass stock component\\
$\beta_{q,m,k}$ & Effect of environmental covariate $k$ on index $m$ catchability\\
$\gamma_{r\rightarrow r',i}$ & Random effect for link-transformed mean movement from region $r$ to $r'$ in seasonal interval $i$ when a prior distribution is assumed\\
$\delta_i$ & Duration of seasonal interval $i$\\
$\delta_{m,i}$ & Duration between the beginning of seasonal interval $i$ and $t_m$\\
$\delta_{s,i}$ & Duration between the beginning of seasonal interval $i$ and $t_s$ \\
$\varepsilon_{y,a,r}$ & Normally distributed random effect for abundance at age $a$ in year $y$ in region $r$\\
$\varepsilon_{N_1,a,r}$ & Normally distributed random effect for initial number at age $a$ in the first year for region $r$\\ 
$\varepsilon_{y,1}$ & Normally distributed random effect for abundance at age 1 in year $y$ for a given stock\\
$\varepsilon_{r\rightarrow r',y,a,i}$ & Normally distributed random effect for movement from region $r$ to region $r'$ in seasonal interval $i$ and year $y$ at age $a$\\
$\varepsilon_{M,y,a,r}$ & Normally distributed random effect for natural mortality rate in region $r$ and year $y$ at age $a$\\
$\varepsilon_{q,m,y}$ & Index $m$ catchability random effect in year $y$\\
$\boldsymbol{\varepsilon}_{y,a}$ & Vector (length $n_R$) of normally distributed random effects for abundance alive in each region on January 1 of year $y$ at age $a$\\
$\boldsymbol{\varepsilon}_{N_1,r}$ & Vector (length $A$) of normally distributed random effects for initial numbers at age in the first year for region $r$\\ 
$\theta_{r\rightarrow r',i}$ & Mean or intercept parameter across age and year for movement from region $r$ to region $r'$ in seasonal interval $i$\\
$\theta_{N_1,r}$ & Mean parameter for initial numbers at age random effects in the first year for region $r$\\
$\theta_{M,a,r}$ & Mean parameter for natural mortality rate at age $a$ in region $r$\\
$\theta_{R}$ & Mean parameter across year for recruitment\\
$\theta_{q,m}$ & Mean index $m$ catchability parameter\\
$\lambda_s$ & Weighting of equilibrium spawning biomass-per-recruit for stock $s$ to use for defining SPR-based fishing mortality reference point\\
$\mu_{r\rightarrow r',y,a,i}$ & Movement rate from region $r$ to region $r'$ in seasonal interval $i$ and year $y$ at age $a$\\
$\mu_{N\rightarrow S}$ & Proportions of the northern component of the BSB stock moving south from the north\\
$\mu_{S\rightarrow N}$ & Proportions of the northern component of the BSB stock moving north from the south\\
$\boldsymbol{\mu}_{1}$ & Matrix ($2 \times 2$) of proportions of the northern component of the BSB stock moving or staying in north and south region for the seasonal intervals after spawning\\
$\boldsymbol{\mu}_{2}$ & Matrix ($2 \times 2$) of proportions of the northern component of the BSB stock moving or staying in north and south region for the seasonal intervals before spawning\\
$\boldsymbol{\mu}_{y,a,i}$ & Matrix ($n_R \times n_R$) of proportions moving or staying in each region at the end of seasonal interval $i$ in year $y$ and age $a$ \\
$\upnu(\widetilde{F})$ & Matrix of equilibrium yield-per-recruit by region and fleet at fully-selected fishing mortality rate $\widetilde{F}$\\
$\rho_{N,\text{age},r}$ & First order auto-regressive correlation parameter across ages for abundance random effects in region $r$ for age classes $>1$\\
$\rho_{N,\text{year},r}$ & First order auto-regressive correlation parameter across years for abundance at age random effects in region $r$ for age classes $>1$\\
$\rho_{N,1,\text{year}}$ & First order auto-regressive correlation parameter across years for abundance at age 1 random effects\\
$\rho_{N_1,r}$ & First order auto-regressive correlation parameter for initial numbers at age random effects in the first year for region $r$\\
$\rho_{r \rightarrow r',\text{age},i}$ & First order auto-regressive correlation parameter across ages for movement random effects from region $r$ to $r'$ in seasonal interval $i$\\
$\rho_{r \rightarrow r',\text{year},i}$ & First order auto-regressive correlation parameter across years for movement random effects from region $r$ to $r'$ in seasonal interval $i$\\
$\rho_{M,\text{age},r}$ & First order auto-regressive correlation parameter across ages for natural mortality rate random effects in region $r$\\
$\rho_{M,\text{year},r}$ & First order auto-regressive correlation parameter across years for natural mortality rate random effects in region $r$\\
$\rho_{q,m}$ & First order auto-regressive correlation parameter across years for index $m$ catchability random effects\\
$\sigma_{f,y}$ & Standard deviation of observed log-aggregate catch for fleet $f$ in year $y$\\
$\sigma_{N,a,r}$ & Standard deviation parameter for normally distributed random effects of abundance at age $a$ in region $r$\\
$\sigma_{N,1}$ & Standard deviation parameter for normally distributed random effects of abundance at age 1 for a givens stock\\
$\sigma_{N_1,r}$ & Standard deviation parameter for numbers at age normally distributed random effects in the first year for region $r$\\
$\sigma_{r \rightarrow r',i}$ & Standard deviation parameter for normally distributed random effects of movement from region $r$ to $r'$ in seasonal interval $i$\\
$\sigma_{\gamma, r \rightarrow r',i}$ & Standard deviation parameter for prior distribution of $\gamma_{r\rightarrow r',i}$\\
$M_{y,a,r}$ & natural mortality rate for age $a$ in year $y$ in region $r$\\
$\sigma_{M,r}$ & Standard deviation parameter for natural mortality rate random effects in region $r$\\
$\sigma_{q,m}$ & Standard deviation parameter for index $m$ catchability random effects\\
$\upphi(\widetilde{F})$ & Equilibrium spawning biomass per recruit a given stock at fully-selected fishing mortality rate $\widetilde{F}$
\end{longtable}

### 2025-07-01 paper/paper.tex
final

@@
 \title{Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass}
-\author{Timothy J. Miller\textsuperscript{1} \and Kiersten Curti \and Alexander C. Hansell}
-\date{18 March, 2025}
+\author{Timothy J. Miller\textsuperscript{1} \and Kiersten L. Curti \and Alexander C. Hansell}
+\date{19 March, 2025}
 
@@
 2020; Varkey et al.
-2022) and to the south, the Woods Hole Assessment Model (WHAM, Miller and Stock 2020; Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).
-
-WHAM is an R package developed and maintained at NOAA's Northeast Fisheries Science Center {[}\url{https://timjmiller.github.io/wham}{]}.
+2022) and to the south, the Woods Hole Assessment Model (WHAM, Miller and Stock 2020; Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024a).
+
+WHAM is an R package developed and maintained by scientists at NOAA's Northeast Fisheries Science Center (\url{https://timjmiller.github.io/wham}).
 WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and effects of environmental covariates on various population parameters.
@@
 We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) (NEFSC 2023).
-Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age assessment.
+Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age model with results subsequently combined to inform management.
 The stock was split into two regions to account for spatial dynamics and improve model diagnostics (ASMFC 2016).
@@
 
-When modeling multiple stock components occurring in different regions further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al.
+When modeling multiple stock components occurring in different regions, further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al.
 2021).
 Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises.
-Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points.
+Finally, the inclusion of explicit effects of environmental covariates in the model also requires assumptions on how they are treated in projections and reference points.
 We demonstrate some of the options of WHAM version 2.0 to provide biological reference points and short-term projections.
@@
 1987; Commenges 1999; Andersen and Keiding 2002).
-We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval \(\delta\) (columns), conditional on being in each of those state at beginning of the interval (rows).
+We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval \(\delta\) (columns), conditional on being in each of those states at the beginning of the interval (rows).
 For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
@@
 
-WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would be multinomial distributed as in a model for tagging data where fates of individual tagged fish are assumed independent.
+WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would have a multinomial distribution as in a model for tagging data where fates of individual tagged fish are assumed independent.
 The PTMs determine the expected numbers 1) in each state on January 1 of year \(t+1\) at age \(a+1\) given the abundances at age \(a\) on January 1 of year \(t\), 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.
@@
 where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
+
 When there is autocorrelation with age, WHAM now assumes this applies only to ages \(a>1\) by default so that recruitment random effects are independent of those for the apparent survival of older age classes.
@@
 For the plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is a ``fundamental matrix'\,' derived using the matrix version of the geometric series (Kemeny and Snell 1960).
-Recall that recruitment for stock \(s\) only occur in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
+Recall that recruitment for stock \(s\) only occurs in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
 \[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}^T \mathbf{N}_{O,1,1}.\]
@@
 \[\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}^T \mathbf{N}_{O,s,y,a}\]
-(recall each fleet can harvest in only one region as described above). The vector of predicted numbers caught be each fleet across stocks is
+(recall each fleet can harvest in only one region as described above). The vector of predicted numbers caught by each fleet across stocks is
 \[\widehat{\mathbf{N}}_{H,y,a} = \sum^{n_S}_{s=1} \widehat{\mathbf{N}}_{H,s,y,a}\]
@@
 
-Currently a single fishing mortality reference point \(\widetilde F\) is estimated across stocks and regions and the fleet- and age-specific reference points is \(\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}\).
+Currently a single fishing mortality reference point \(\widetilde F\) is estimated across stocks and regions and the fleet- and age-specific reference point is \(\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}\).
 Selectivity is determined across fleets and ages as in the equilibrium assumption for initial abundance at age except here it is a function of average F at age over a user-defined set of years
@@
 
-When a Beverton-Holt or Ricker stock recruit relationship is assumed, an analogous Newton method is used to find \(\log F\) that maximizes yield for MSY-based reference points.
-which are also a functions of the equilibrium yield per recruit (Eq.
+When a Beverton-Holt or Ricker stock recruit relationship is assumed, an analogous Newton method is used to find \(\log F\) that maximizes yield for MSY-based reference points, which are also a functions of the equilibrium yield per recruit (Eq.
 \ref{eq:ypr}) and equilibrium recruitment.
@@
 
-Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging from the Gulf of Mexico to the Gulf of Maine.
+Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging the entire east coast of the United States.
 Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity (Able and Stanton Hales, Jr.
@@
 Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as ``Working Group'') composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above (NEFSC 2023).
-This assessment included the spatial features and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
+This assessment included the two regions and two stock components and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
 Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.
@@
 For the northern component, a proportion \(\mu_{\text{N}\rightarrow \text{S}}\) can move to the southern region each month during the last five months of the year, but no movement is allows from the south to the north during this period (Figure \ref{fig:migration-diagram}).
-During the first four intervals of the year a proportion \(\mu_{\text{S}\rightarrow \text{N}}\) the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
+During the first four intervals of the year a proportion \(\mu_{\text{S}\rightarrow \text{N}}\) of the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
 In the fifth interval (May), all northern component individuals remaining in the south are assumed to move back to the north for the subsequent spawning period.
@@
 However, Miller et al.
-(2018) showed how inferences of temperature effects on growth or maturity parameters can be very different whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
+(2018) showed how inferences of temperature effects on growth or maturity parameters can be very different depending on whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
 Therefore, we also explored whether including temporal random effects on age 1 natural mortality affected inferences on corresponding temperature effects.
@@
 We assume the covariate in year \(y\) affects recruitment in the same year because the covariate observations are from months January to March.
-The fish are technically already 1 year old, but there are no observations of these individuals until later in the year except possibly in fishery catches which are accumulated over the whole year.
+The fish are technically already 1 year old, but there are no observations of these individuals until later in the spring except possibly in fishery catches which are accumulated over the whole year.
 Expected log-recruitment for a given stock component is a linear function of bottom temperature
@@
 We also calculated status as the ratios of annual average F and SSB to the corresponding reference points.
-Finally, we calculated ``prevailing'' BRPs using the average weight at age, maturity at age, M, fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
+Finally, we calculated ``prevailing'' BRPs using the average weight at age, maturity at age, M, and fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
 All BRPs and status estimates are generated for each region as well as for the total stock area.
@@
 
-For black sea bass, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figure 6).
+For black sea bass, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figures \ref{fig:R-BT-proj} and \ref{fig:F-SSB-proj}).
 In the projection period, the AR1 models for bottom temperature and recuitment revert to the long-term means of those processes.
-This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypothesis of increasing water temperature and changing environmental conditions (Hare et al.
+This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypotheses of increasing water temperature and changing environmental conditions (Hare et al.
 2016; Pershing et al.
@@
 Because of these concerns, the most recent stock assessment for yellowtail flounder used a change point analysis on environmental covariates to determine current conditions.
-The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections (NEFSC, 2024).
+The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections (NEFSC 2024b).
 Unfortunately, this methodology still has limitations because it is unlikely that current environmental conditions are stable.
-Similarly, a linear increase of temperature is also not likely in the future (see past trends, top row of Figure 6).
-As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
+Similarly, a linear increase of temperature is also not likely in the future (top row of Figure \ref{fig:E-R-SSB-F}).
+As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in BRPs and population projections becomes an important decision in the management process.
 
@@
 Estimation of natural mortality may be even more challenging within state-space assessment models with their greater flexibility from inclusion of time-varying random effects.
-For example, we fixed the mean natural mortality for age 1 fish and Cadigan (2016) and Stock et al.
+For example, we fixed the mean natural mortality for age 1 fish and both Cadigan (2016) and Stock et al.
 (2021) also estimated natural mortality deviations.

### 2025-07-01 paper/paper_R1.tex
final

--- 2025-07-01 paper/paper.tex
@@
 \author{Timothy J. Miller\textsuperscript{1} \and Kiersten L. Curti \and Alexander C. Hansell}
-\date{19 March, 2025}
+\date{16 May, 2025}
 
@@
 
-However, WHAM has only allowed one stock (component) and a single region and without seasonal changes in stock dynamics.
+However, WHAM has only allowed one stock and a single region and without seasonal changes in stock dynamics.
 Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice (Ying et al.
@@
 We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) (NEFSC 2023).
-Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age model with results subsequently combined to inform management.
-The stock was split into two regions to account for spatial dynamics and improve model diagnostics (ASMFC 2016).
+Black sea bass is a high profile stock in the Northeast U.S.
+and important to both commercial fishing fleets and recreational anglers.
+Black sea bass are targeted from Maine to North Carolina in both state and federal waters.
+Past stock assessments have estimated that the stock is healthy (not overfished and overfishing not occurring).
+Previous assessments assumed two stock components (north and south), each assessed with a separate statistical catch at age model with results subsequently combined to inform management, but these assessments have exhibited strong retrospective patterns.
+The stock was split into two regional components to account for spatial dynamics and improve model diagnostics (ASMFC 2016).
 Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (Moser and Shepherd 2009).
@@
 2016).
-Thus, a focus of the black sea bass research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process.
+Thus, a focus of the black sea bass research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement, and explore environmental drivers of population dynamics (NEFSC 2023).
+We denote the fish in the north as separate stock components rather than stocks because both components are managed as a single stock, but recognizing differences in migration between fish originating in the two areas.
+
+To illustrate the usage of WHAM 2.0 we apply it to the two stock components of black sea bass off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process.
 However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
@@
 1987; Commenges 1999; Andersen and Keiding 2002).
-We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval \(\delta\) (columns), conditional on being in each of those states at the beginning of the interval (rows).
+We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval with duration \(\delta\) (columns), conditional on being in each of those states at the beginning of the interval (rows).
 For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
@@
 is the \(n_R \times n_F\) matrix defining probabilities of being captured in each fleet during the interval, and \(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of dying due to natural mortality during the interval.
-We have the identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).
+We have the \(n_F\times n_F\) identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).
 
@@
 WHAM assumes each fleet \(f\) can harvest in only one region (\(r_f\)) during specified seasons.
-So, for each fleet \(f\), row \(r_f\) and column \(f\) of \(\mathbf{H}_{y,a,i}\) will be \(F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\) when fleet \(f\) is harvesting during the interval \(\delta_i\) and all other elements will be zero.
-Row \(r\) of the single-column matrix \(\mathbf{D}_{y,a,i}\) is \(M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\)
-
-When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
-We obtain the PTM over an interval \(\delta_i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
-\begin{equation*}
-\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
-\end{equation*}
-The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region.
-Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero.
-For two regions and one fleet operating in each region:
-\begin{equation*}
- \mathbf{A}_{y,a,i} = \begin{bmatrix}
- a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
- \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
- 0 & 0 & 0 & 0 & 0 \\
- 0 & 0 & 0 & 0 & 0 \\
- 0 & 0 & 0 & 0 & 0
- \end{bmatrix}
-\end{equation*}
-where \(a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})\).
-When there is one region, \(n_f\) fleets, and \(\delta_i = 1\), exponentiating the instantaneous rate matrix results in the PTM defined in Eq.
-\ref{eq:ptm_1_region}.
+So, for each fleet \(f\), row \(r_f\) and column \(f\) of \(\mathbf{H}_{y,a,i}\) will be \(F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\) when fleet \(f\) is harvesting during the interval \(i\) and all other elements will be zero.
+Row \(r\) of the single-column matrix \(\mathbf{D}_{y,a,i}\) is \(M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\).
+For details defining PTMs with movement and mortality occuring simultaneously see Appendix B.
 
@@
 Seasonality can be configured to accommodate characteristics of spawning, movement, and fleet-specific behavior.
-The annual time step can be divided into \(K\) seasons and the interval size \(\delta_i\) for each season \(i\) does not need to be equal to any other seasonal interval.
-Under the Markov assumption, the PTM of surviving and moving and dying over \(K\) intervals \(\delta_1,\ldots, \delta_K\) (i.e., the entire year) is just the product of the PTMs for each interval:
+The annual time step can be divided into \(K\) seasons and the interval duration \(\delta_i\) for each season \(i\) does not need to be equal to any other seasonal interval.
+Under the Markov assumption the PTMs for each consecutive interval are independent and the PTM of surviving and moving and dying over \(K\) intervals \(1,\ldots, K\) (i.e., the entire year) is just the product of the PTMs for each interval:
 \[ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).\]
 
-For a stock spawning at some fraction of the year \(0<t_s<1\) in interval \(\delta_j\), the fraction of time in season \(j\) is
-\[\delta_{s,j} = t_s-\sum^{j-1}_{i=0}\delta_i
-\]
-and the PTM defining the proportions in each state at time \(t_s\) for age \(a\) is
+For a stock spawning at some fraction of the year \(0<t_s<1\) in interval \(i \in \{1,\ldots,K\}\), the fraction of time in the interval is
+\[\delta_{s,i} = t_s-\sum^{i-1}_{j=0}\delta_j\]
+(where \(\delta_0 = 0\)) and the PTM defining the proportions in each state at time \(t_s\) for age \(a\) is
 \begin{equation}\label{eq:ptm_spawn}
-\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{s,j}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{s,j}).
+\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{s,i}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,j}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{s,i})
 \end{equation}
-Similarly, for an index \(m\) occurring at fraction of the year \(t_m\) in interval \(\delta_j\) the proportions in each state at the time of the observation is
+where \(\mathbf{P}_{y,a,0}\) is an identity matrix.
+Similarly, for an index \(m\) occurring at fraction of the year \(t_m\) in interval \(i\) the proportions in each state at the time of the observation is
 \begin{equation} \label{eq:ptm_index} 
-\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{m,j}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{m,j}).
+\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{m,i}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,i}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{m,i}).
 \end{equation}
@@
 If recruitment is assumed to be a function of SSB, it is only the spawning population in region \(r_s\) at the time of spawning that constitutes the SSB in the stock-recruit function.
-However, models can configure spawning individuals to occur in other regions at the time of spawning.
+However, models can configure individuals to occur in other regions at the time of spawning.
 Aside from treating recruitment as a random walk, the general model for annual recruitment as random effects is
@@
 \[\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}^T(t_s) \mathbf{N}_{O,y,a}\]
-and \(w_{y,a}\) is the mean weight at age of spawning individuals, \(\text{mat}_{y,a}\) is the maturity at age, and \(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and occurring in region \(r_s\) at time \(t_s\) given being alive in each region at the start of the year (the \(r_s\) column of the upper-left submatrix of Eq.
+and \(w_{y,a}\) and \(\text{mat}_{y,a}\) are the stock-specific mean weight at age of spawning individuals and maturity at age, respectively, and \(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and occurring in region \(r_s\) at time \(t_s\) given being alive in each region at the start of the year (the \(r_s\) column of the upper-left submatrix of Eq.
 \ref{eq:ptm_spawn}).
@@
 \[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}^T \mathbf{N}_{O,1,1}.\]
-
-The initial abundances at age can also be treated as independent or AR1 random effects. Defining the vector of initial abundance at age in region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
-\[\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}\]
-where
-\[  Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.\]
+Initial abundance at age can also be estimated as random effects as described in Appendix B.
 
@@
 When random effects are used, the same 2DAR1 structure with age and year as described by Stock and Miller (2021) can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
-So the general model for natural mortality is
-\begin{equation*}
-  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
-\end{equation*}
-The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
-\begin{equation*}
-  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
-\end{equation*}
+See Appendix B for details.
 
@@
 Predicted age composition observations are functions of \(\widehat{I}_{m,y,a}\) as described by Stock and Miller (2021) and the likelihood options are the same as those for catch explained above.
-
-Catchability of the index can also be treated as logit functions of normal random effects and(or) environmental covariate effects
-\[log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}\]
-where \(\theta_{q,m}\) is an intercept or mean parameter, and \(u_{m}\) and \(l_{m}\) are the upper and lower bounds of catchability for index \(m\) (defaults are 0 and 1000).
-The general covariance structure for the annual random effects is AR1
-\[Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.\]
+See Appendix B for options to model temporal variation in catchability.
 
@@
 \begin{equation}\label{eq:newton}
-  \log\widetilde{F}^{j} = \log\widetilde{F}^{j-1} - \frac{g\left(\log\widetilde{F}^{j-1}\right)}{g^{(1)}\left(\log\widetilde{F}^{j-1}\right)}
+  \log\widetilde{F}^{j} = \log\widetilde{F}^{j-1} - \frac{h\left(\log\widetilde{F}^{j-1}\right)}{h^{(1)}\left(\log\widetilde{F}^{j-1}\right)}
 \end{equation}
-where \(g(\log F)\) is the difference between the weighted sums of spawning biomass per recruit at \(F\) and \(X\)\% of unfished spawning biomass per recruit across stocks:
+where \(h(\log F)\) is the difference between the weighted sums of spawning biomass per recruit at \(F\) and \(X\)\% of unfished spawning biomass per recruit across stocks:
 \begin{equation}\label{eq:newton-obj}
-  g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
+  h(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
 \end{equation}
 where \(\upphi_s\left(F=0\right)\) is the equilibrium unfished spawning biomass per recruit.
-\(g^{(1)}(\log F)\) is the derivative of \(g\) with respect to \(\log F\), and the weights to use for each stock \(\lambda_s\) can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate ``static'' equilibrium spawning biomass and yield.
+\(h^{(1)}(\log F)\) is the derivative of \(h\) with respect to \(\log F\), and the weights to use for each stock \(\lambda_s\) can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate ``static'' equilibrium spawning biomass and yield.
 
@@
 \ref{eq:ypr}) and equilibrium recruitment.
-The function \(g(\log F)\) and \(g^{(1)}(\log F)\) in Eq.
+The function \(h(\log F)\) and \(h^{(1)}(\log F)\) in Eq.
 \ref{eq:newton} are the first and second derivatives of the sum of the stock-specific equilibrium yields with respect to \(\log F\) (analogous to that described by Miller et al.
@@
 2013).
-Fish in this northern stock unit perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring (Musick and Mercer 1977).
+Fish in this northern stock component perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring (Musick and Mercer 1977).
 Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude (Moser and Shepherd 2009).
@@
 The first year being modeled for the population is 1989 and the fishery and index data used in the model span from 1989 to 2021.
-The north and south stock components are modeled as separate populations that spawn and recruit in respective regions.
+The north and south stock components are modeled as separate populations that spawn and recruit in respective regions and there are eight age classes with the last being a plus group (\(a \in \{1,\ldots,8+\}\)).
 We have observations for each of four total fishing fleets, where two fishing fleets (recreational and commercial) operate in each region.
@@
 The southern stock component is assumed to never move to the northern region.
-For the northern component, a proportion \(\mu_{\text{N}\rightarrow \text{S}}\) can move to the southern region each month during the last five months of the year, but no movement is allows from the south to the north during this period (Figure \ref{fig:migration-diagram}).
+For the northern component, a proportion \(\mu_{\text{N}\rightarrow \text{S}}\) can move to the southern region each month during the last five months of the year, but no movement is allowed from the south to the north during this period (Figure \ref{fig:migration-diagram}).
 During the first four intervals of the year a proportion \(\mu_{\text{S}\rightarrow \text{N}}\) of the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
@@
 
-All models include bottom temperature observations for the northern and southern regions from 1963 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius (NEFSC 2023).
+All models include observations of winter bottom temperature anomalies for the northern and southern regions from 1959 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius (NEFSC 2023).
+See the Supplementary Materials for details about the derivation of these observations.
 We retained the assumption from the peer-reviewed assessment that treated the latent bottom temperature covariates in each region as AR1 processes.
@@
 
-We assume the covariate in year \(y\) affects recruitment in the same year because the covariate observations are from months January to March.
+We assume the covariate in year \(y\) affects recruitment in the same year because the covariate observations are from winter months.
 The fish are technically already 1 year old, but there are no observations of these individuals until later in the spring except possibly in fishery catches which are accumulated over the whole year.
@@
 We estimated median relative bias of SSB across these simulations.
+Finally, we fitted three sensitivity models which differed from the best performing model, by 1) assuming no movement for either stock component, 2) movement parameters are fixed at the means for the prior distributions, and 3) a constant natural mortality rate is estimated across all regions and stocks.
 See the Supplementary Materials for more details.
@@
 \hypertarget{reference-points-and-projections}{%
-\subsection*{Reference points, and projections}\label{reference-points-and-projections}}
-\addcontentsline{toc}{subsection}{Reference points, and projections}
+\subsection*{Reference points and projections}\label{reference-points-and-projections}}
+\addcontentsline{toc}{subsection}{Reference points and projections}
 
@@
 Although this might suggest making inferences on the population by using an ensemble of these and weighting by AIC (Burnham and Anderson 2002), the estimates of assessment outputs relevant for management (e.g., SSB and \(F\)) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
-Nevertheless, continued monitoring of effects on the southern component may be prudent.
+Continued monitoring of temperature effects on the southern component and investigations of non-linear effects of temperature for both components may be prudent because of the possibility of an optimal temperature range for survival via effects on growth during the pre-recruit phase (Ware and Lambert 1985; Anderson 1988).
+Furthermore, our inferences about effects on natural mortality at age 1 may be influences by our assumptions of constant median (or constant depending on the model) natural mortality rate and movement rates for all ages.
+Further investigations of variation in these parameters with age and effects of bottom temperature could improve our understanding of the demography for this stock.
 
 We showed that using expected recruitment rather than recruitment random effects to define BRPs can produce the same estimates but with lower uncertainty.
-Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same BRPs and even less uncertainty than those we showed using expected recruitment.
+Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same BRPs and even lower uncertainty.
 These results can be important for management when uncertainty in assessment output is used in making catch advice.
@@
 The uncertainty in these ratios could also be used in defining catch advice.
+
+BRPS under prevailing conditions use average values for recruitment and other inputs and MSY-based BRPs are based on the stock-recruit relationship which is also typically constant over time.
+If BRPs should be defined in such a way that they are less variable than the stock over time, expected inputs to the BRPs (instead of annual values) would be more appropriate.
+Lower variability in BRPs would presumably correlate with lower variability in catch advice which might be desirable (e.g., Cochrane et al.
+1998; Kell et al.
+2005), but the appropriate level of temporal variability in BRPs and catch advice will dependent on the characteristics of the fish stock, associated fisheries, and management structure (Holland 2010; Punt et al.
+2016).
 
@@
 
+Our application to black sea bass is a particular combination of the spatial modeling options described by Berger et al.
+(2024).
+Other combinations of the described options are possible in WHAM 2.0.
+For example, WHAM 2.0 can accommodate all of fleet structure options described in Table 2 by Berger et al.
+(2024) and we expect more of these options to become available in future versions of WHAM.
+
 \hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
+\section*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{section}{Acknowledgements}
 
@@
 In review.
-Factors affecting reliablity of state-space age-structured assessment models.
+An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models.
 Canadian Journal of Fisheries Science.
@@
 
-\caption{Prior (black) and posterior (red) distributions of movement of the northern stock component from north to south (top) and south to north (bottom).}\label{fig:move-prior-posterior}
+\caption{Prior (black) and posterior (red) distributions of movement of the northern stock component from north to south (top) and south to north (bottom).
+Vertical lines are the prior and posterior estimates.}\label{fig:move-prior-posterior}
 \end{figure}
@@
 \renewcommand\thetable{A\arabic{table}}
-\input{symboltable.tex}
+\input{symboltable_R1.tex}
+
+\hypertarget{appendix-b}{%
+\section*{Appendix B}\label{appendix-b}}
+\addcontentsline{toc}{section}{Appendix B}
+
+New features of WHAM 2.0 that are not used in the application to black sea bass are described here..
+
+\hypertarget{simultaneous-movement-and-mortality}{%
+\subsection*{Simultaneous movement and mortality}\label{simultaneous-movement-and-mortality}}
+\addcontentsline{toc}{subsection}{Simultaneous movement and mortality}
+
+When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
+We obtain the PTM over an interval \(i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
+\begin{equation*}
+\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
+\end{equation*}
+The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region.
+Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero.
+For two regions and one fleet operating in each region:
+\begin{equation*}
+ \mathbf{A}_{y,a,i} = \begin{bmatrix}
+ a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
+ \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
+ 0 & 0 & 0 & 0 & 0 \\
+ 0 & 0 & 0 & 0 & 0 \\
+ 0 & 0 & 0 & 0 & 0
+ \end{bmatrix}
+\end{equation*}
+where \(a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})\).
+When there is one region, \(n_f\) fleets, and \(\delta_i = 1\), exponentiating the instantaneous rate matrix results in the PTM defined in Eq.
+\ref{eq:ptm_1_region}.
+
+\hypertarget{initial-abundance-at-age-as-random-effects}{%
+\subsection*{Initial abundance at age as random effects}\label{initial-abundance-at-age-as-random-effects}}
+\addcontentsline{toc}{subsection}{Initial abundance at age as random effects}
+
+The initial abundances at age can also be treated as independent or AR1 random effects. Defining the vector of initial abundance at age in region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
+\begin{equation*}
+\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}
+\end{equation*}
+where
+\begin{equation*}
+Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.
+\end{equation*}
+
+\hypertarget{general-natural-mortality-model}{%
+\subsection*{General natural mortality model}\label{general-natural-mortality-model}}
+\addcontentsline{toc}{subsection}{General natural mortality model}
+
+The general model for natural mortality at age \(a\) for a given stock in region \(r\) is a function of random and covariate effects
+\begin{equation*}
+  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
+\end{equation*}
+The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
+\begin{equation*}
+  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
+\end{equation*}
+
+\hypertarget{general-catchability-model}{%
+\subsection*{General catchability model}\label{general-catchability-model}}
+\addcontentsline{toc}{subsection}{General catchability model}
+
+Catchability of each index can be treated as logit functions of normal random effects and(or) environmental covariate effects
+\begin{equation*}
+log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}
+\end{equation*}
+where \(\theta_{q,m}\) is an intercept or mean parameter, and \(u_{m}\) and \(l_{m}\) are the upper and lower bounds of catchability for index \(m\) (defaults are 0 and 1000).
+The general covariance structure for the annual random effects is AR1
+\begin{equation*}
+Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.
+\end{equation*}
 
@@
 
+\begin{landscape}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{SSB_F_sensitivity_plots}
+
+}
+
+\caption{Estimates of annual SSB and fishing mortality rates for the best performing model $M_1$ and models that are otherwise the same except where  1) movement rates are fixed at the means for the prior distribution, 2) a constant natural mortality rate is estimated, or 3) there is no movement for either stock component.}\label{fig:sensitivity-plots}
+\end{figure}
+\end{landscape}
+
 \end{document}

### 2025-07-01 paper/paper_R2.Rmd
final

--- 2025-05-19 paper/paper_R1.Rmd
@@
 library(kableExtra)
-library(wham)
+#library(wham)
 #library(ggplotFL)
-library(viridis)
+#library(viridis)
 type <- "latex"
@@
 The State-space Assessment Model [SAM, @nielsenberg14] continues to be developed and remains widely used within  the International Council for the Exploration of the Sea (ICES) to assess European fish stocks.
-Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada [@varkeyetal22;@perreaultetal20] and to the south, the Woods Hole Assessment Model [WHAM, @millerstock20; @stockmiller21] is now used to assess many fish stocks in the Northwest Atlantic Ocean [@nefsc22;@nefsc22a;@nefsc24].
+Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada [@varkeyetal22;@perreaultetal20] and to the south, the Woods Hole Assessment Model [WHAM, @stockmiller21] is now used to assess many fish stocks in the Northwest Atlantic Ocean [@nefsc22;@nefsc22a;@nefsc24].
 
 WHAM is an R package developed and maintained by scientists at NOAA's Northeast Fisheries Science Center (https://timjmiller.github.io/wham).
-WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and effects of environmental covariates on various population parameters.
+WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects, to models with several time and age varying process errors and effects of environmental covariates on various population parameters.
 Like SAM, WHAM models are built using the Template Model Builder package [TMB, @kristensenetal16] which provides a computationally efficient means of fitting an extremely wide class of models with random effects.
@@
 
-However, WHAM has only allowed one stock and a single region and without seasonal changes in stock dynamics.
+However, WHAM has only allowed consideration of a single stock and region, and no seasonal changes in stock dynamics.
 Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice [@bosleyetal22;@yingetal11;@caoetal14].
@@
 
-We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) [@nefsc23].
-Black sea bass is a high profile stock in the Northeast U.S.
+We developed this extension of WHAM in concert with a research track assessment process for black sea bass [BSB\, \textit{Centropristis striata}\; @nefsc23], which is a high profile stock in the Northeast U.S.
 and important to both commercial fishing fleets and recreational anglers.
-Black sea bass are targeted from Maine to North Carolina in both state and federal waters.
+BSB are targeted from Maine to North Carolina in both state and federal waters.
 Past stock assessments have estimated that the stock is healthy (not overfished and overfishing not occurring).
-Previous assessments assumed two stock components (north and south), each assessed with a separate statistical catch at age model with results subsequently combined to inform management, but these assessments have exhibited strong retrospective patterns.
+Those assessments assumed two stock components (north and south), each assessed with a separate statistical catch at age model, with results subsequently combined to inform management.
+However, the assessments exhibited strong retrospective patterns.
 The stock was split into two regional components to account for spatial dynamics and improve model diagnostics [@asmfc16].
-Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall [@mosershepherd09].
-Black sea bass distribution has been linked to warming waters on the northwest shelf [@belletal15] and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north [@hareetal16].
-Thus, a focus of the black sea bass research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement, and explore environmental drivers of population dynamics [@nefsc23].
+Additionally, BSB have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall [@mosershepherd09].
+BSB distribution has been linked to warming waters on the northwest shelf [@belletal15] and it is hypothesized that the distribution and productivity of the BSB stock is especially susceptible to long-term changes in temperature causing increases in abundance further north [@hareetal16].
+Thus, a focus of the BSB research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement, and explore environmental drivers of population dynamics [@nefsc23].
 We denote the fish in the north as separate stock components rather than stocks because both components are managed as a single stock, but recognizing differences in migration between fish originating in the two areas.
 
-To illustrate the usage of WHAM 2.0 we apply it to the two stock components of black sea bass off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process.
-However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
-
-When modeling multiple stock components occurring in different regions, further dimensionality is added to the class of possible equilbrium reference points [e.g., @kapuretal21].
+To illustrate the usage of WHAM 2.0 we apply it to the two stock components of BSB off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process [@nefsc23].
+However, here we evaluate evidence for alternative hypotheses of temporal variation and effects of bottom temperature, specifically, on recruitment and natural mortality of age 1 individuals.
+
+When modeling multiple stock components and regions, further dimensionality is added to the class of possible equilbrium reference points [e.g., @kapuretal21].
 Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises.
@@
 Many of the options and equations of WHAM version 2.0 are the same as those in @stockmiller21, so we will only describe extensions and differences that have occurred since their first description of WHAM.
-The new version of WHAM can model multiple stocks each with their own movement, harvest, and natural mortality.
+The new version of WHAM can model multiple stocks, each with their own movement, harvest, and natural mortality.
 Much of the description below is for a specific stock $s$, but, for simplicity, this subscript is implicit except when necessary.
@@
 
-Because individuals may be alive in one of several regions or harvested in one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model.
+Because individuals may be alive in one of several regions or harvested by one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model.
 Approaches to modeling transitions among these states may treat time discretely [e.g.,@arnason72;@schwarzetal93] or continuously [e.g., @hearnetal87;@commenges99; @andersenkeiding02].
@@
 For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
-The time interval $i$ with duration $\delta_i$ would be a season and the PTMs would be uniquely defined for each stock by year $y$, age $a$, and season $i$ on January 1.
+The time interval $i$ with duration $\delta_i$ would be a season and the PTMs would be uniquely defined for each stock by year $y$, age $a$, and season $i$.
 Each row and column of the PTM correspond to one of the states: alive in region $r$, dead in fleet $f$, or dead from natural causes.
-The probabilities in each row sum to unity (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval.
+The probabilities in each row sum to one (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval.
 Given $n_R$ regions and $n_F$ fleets, the square PTM ($n_R + n_F + 1$ rows and columns) as a function of sub-matrices is
@@
 \end{equation*}
-is the $n_R \times n_F$ matrix defining probabilities of being captured in each fleet during the interval, and $\mathbf{D}_{y,a,i}$ is the $n_R x 1$ matrix of probabilities of dying due to natural mortality during the interval.
-We have the $n_F\times n_F$ identity matrix $\mathbf{I}_{H}$ for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).
+is the $n_R \times n_F$ matrix defining probabilities of being captured in each fleet during the interval, and $\mathbf{D}_{y,a,i}$ is the $n_R \times 1$ matrix of probabilities of dying due to natural mortality during the interval.
+We have the $n_F\times n_F$ identity matrix $\mathbf{I}_{H}$ for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is one (no zombies allowed).
 
 WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would have a multinomial distribution as in a model for tagging data where fates of individual tagged fish are assumed independent.
-The PTMs determine the expected numbers 1) in each state on January 1 of year $t+1$ at age $a+1$ given the abundances at age $a$ on January 1 of year $t$, 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.
+The PTMs determine the expected numbers 1) in each state on January 1 of year $y+1$ at age $a+1$ given the abundances at age $a$ on January 1 of year $y$, 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.
 
@@
      0 & 0 & 1
-  \end{bmatrix}
-\end{equation}
-where $S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}$, $\mathbf{H}_{y,a,i}$ is a 1 x $n_F$ matrix with elements for each fleet $f$: $\frac{F_{y,a,i,f}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)$, $D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)$, and $Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{y,a,i,f}$ is the total mortality rate.
+  \end{bmatrix},
+\end{equation}
+where $S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}$, $\mathbf{H}_{y,a,i}$ is a 1 x $n_F$ matrix with elements for each fleet $f$: $\frac{F_{f,y,a,i}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)$, $D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)$, and $Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{f,y,a,i}$ is the total mortality rate.
 
@@
 \end{equation*}
-and $\boldsymbol{\mu}_{y,a,i}$ is a $n_R \times n_R$ matrix of probabilities of moving from one region to another or staying in the region they occurred at the beginning of the interval: 
+and $\boldsymbol{\mu}_{y,a,i}$ is a $n_R \times n_R$ matrix of proportions moving from one region to another or staying in the region they occurred at the beginning of the interval: 
 \begin{equation*}
@@
 WHAM assumes each fleet $f$ can harvest in only one region ($r_f$) during specified seasons.
-So, for each fleet $f$, row $r_f$ and column $f$ of $\mathbf{H}_{y,a,i}$ will be $F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$ when fleet $f$ is harvesting during the interval $i$ and all other elements will be zero.
+So, for each fleet $f$, row $r_f$ and column $f$ of $\mathbf{H}_{y,a,i}$ will be $F_{f,y,a,i}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$ when fleet $f$ is harvesting during the interval $i$ and all other elements will be zero.
 Row $r$ of the single-column matrix $\mathbf{D}_{y,a,i}$ is  $M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$.
-For details defining PTMs with movement and mortality occuring simultaneously see Appendix B.
+For details defining PTMs with movement and mortality occurring simultaneously see Appendix B.
 
@@
 The annual time step can be divided into $K$ seasons and the interval duration $\delta_i$ for each season $i$ does not need to be equal to any other seasonal interval.
-Under the Markov assumption the PTMs for each consecutive interval are independent and the PTM of surviving and moving and dying over $K$ intervals $1,\ldots, K$ (i.e., the entire year) is just the product of the PTMs for each interval:
+Under the Markov assumption the PTMs for each consecutive interval are independent and the PTM of surviving, moving, and dying over $K$ intervals $1,\ldots, K$ (i.e., the entire year) is just the product of the PTMs for each interval:
 $$ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).$$
@@
 $$\delta_{s,i} = t_s-\sum^{i-1}_{j=0}\delta_j$$
-(where $\delta_0 = 0$) and the PTM defining the proportions in each state at time $t_s$ for age $a$ is
+(where $\delta_0 = 0$), and the PTM defining the proportions in each state at time $t_s$ for age $a$ is
 \begin{equation}\label{eq:ptm_spawn}
-\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{s,i}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,j}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{s,i})
+\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{s,i}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,j}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{s,i}),
 \end{equation}
@@
 
-When there are $n_R$ regions and $n_F$ fleets, the vector of abundance in each state at age $a>1$ on January 1 is $\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}^T, \mathbf{0}^T)^T$ where $\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})^T$ is the number in the states corresponding to being alive in each region and $\mathbf{0}$ is a vector ($n_F+1$) for the numbers captured in each fleet and dead from natural mortality because no age $a$ fish have died yet on January 1.
+When there are $n_R$ regions and $n_F$ fleets, the vector of abundance for a given stock (component) in each state at age $a>1$ on January 1 is $\mathbf{N}_{y,a}^T = (\mathbf{N}_{O,y,a}^T, \mathbf{0}^T)$, where $\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})^T$ is the number in the states corresponding to being alive in each region and $\mathbf{0}$ is a vector ($n_F+1$) for the numbers captured in each fleet or dead from natural mortality because no age $a$ fish have died yet on January 1.
 
 Each stock $s$ is assumed to spawn and recruit in one region $r_s$.
-So for age $a=1$, $\mathbf{N}_{O,y,1}$ is 0 except for row $r = r_s$.
-Options for configuring recruitment ($N_{y,1,r_s}$) for each stock are the same as previous versions of WHAM [@millerstock20].
+So for age $a=1$, $\mathbf{N}_{O,y,1}$ is 0 except for row $r_s$.
+Options for configuring recruitment ($N_{y,1,r_s}$) for each stock are the same as previous versions of WHAM.
 If recruitment is assumed to be a function of SSB, it is only the spawning population in region $r_s$ at the time of spawning that constitutes the SSB in the stock-recruit function.
@@
 \begin{equation*}
-\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1,r_s} =  f\left(\text{SSB}_{y-1,r_s}\right) + \varepsilon_{y,1,r_s}
+\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1} =  f\left(\text{SSB}_{y-1}\right) + \varepsilon_{y,1,r_s},
 \end{equation*}
 where  
-$$\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}^T(t_s) \mathbf{N}_{O,y,a}$$
-and $w_{y,a}$ and $\text{mat}_{y,a}$ are the stock-specific mean weight at age of spawning individuals and maturity at age, respectively, and $\mathbf{O}_{y,a,r_s}(t_s)$ are the probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year (the $r_s$ column of the upper-left submatrix of Eq.
+$$\text{SSB}_{y} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,\cdot,r_s}^T(t_s) \mathbf{N}_{O,y,a}$$
+and $w_{y,a}$ and $\text{mat}_{y,a}$ are the stock-specific mean weight at age of spawning individuals and maturity at age, respectively, and $\mathbf{O}_{y,a,\cdot,r_s}(t_s)$ are the probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year (column $r_s$ of the upper-left submatrix of Eq.
 \ref{eq:ptm_spawn}).
@@
 \begin{equation*}
-\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,A-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}^T \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}^T \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
+\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,A-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}^T \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}^T \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A},
 \end{equation*}
@@
 When there is autocorrelation with age, WHAM now assumes this applies only to ages $a>1$ by default so that recruitment random effects are independent of those for the apparent survival of older age classes.
-So the general covariance structure for a given stock at ages $a>1$ in region $r$ is that of a two-dimensional first-order autoregressive (2DAR1) process
+The general covariance structure for a given stock at ages $a>1$ in region $r$ is a two-dimensional first-order autoregressive (2DAR1) process
 \begin{equation*}
@@
 \end{equation*}
-When the annual changes in abundance at age are treated deterministically, $\boldsymbol{\varepsilon}_{y,a} = 0$.
+If the annual changes in abundance at age are treated deterministically, then $\boldsymbol{\varepsilon}_{y,a} = \mathbf{0}$.
 Since recruitment for a given stock currently only occurs in one region $r_s$ there is only a single vector of time-varying recruitment random effects for each stock.
@@
 Initial numbers at age for each stock and region can be treated as age-specific fixed effects or with an equilibrium assumption as in previous versions of WHAM.
-For the equilibrium option there are two parameters for each stock: the stock-specific fully-selected fishing mortality rate $\log \widetilde{F}$ and the recruitment in year 1 $\log N_{1,1,r_s}$.
-A stock-specific equilibrium fishing mortality at age by fleet ${\tilde F}_{a,f}$ is the product of $\widetilde{F}$ and the selectivity across fleets in the first year
-\begin{equation*}
-  \text{sel}_{1,a,f} = \frac{F_{1,a,f}}{\max_a \sum_{f=1}^{n_F} F_{1,a,f}}.
-\end{equation*}
-We use $\widetilde{F}_{a,f}$ to define an equilibrium abundance per recruit by region at age $a$ conditional on recruiting to each region
+For the equilibrium option there are two parameters for each stock: the stock-specific equilibrium fully-selected fishing mortality rate ($\widetilde{F}_1$) and the recruitment in year 1 ($N_{1,1,r_s}$).
+A stock-specific equilibrium fishing mortality at age by fleet ($\widetilde{F}_{f,1,a}$) is the product of $\widetilde{F}_1$ and the selectivity across fleets in the first year
+\begin{equation*}
+  \text{sel}_{f,1,a} = \frac{F_{f,1,a}}{\max_a \sum_{f=1}^{n_F} F_{f,1,a}}.
+\end{equation*}
+We use $\widetilde{F}_{f,1,a}$ to define an equilibrium abundance per recruit by region at age $a$ conditional on recruiting to each region
 \begin{equation}\label{eq:npraa}
@@
 \left[\prod^{a-1}_{j=0}\mathbf{O}_{j}\right] \left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1} & a = A
- \end{array}
+ \end{array},
 \right.
 \end{equation}
-where $\mathbf{O}_{j}$ is the equilibrium probability of surviving age $a$ and occurring in each region and $\mathbf{O}_{0} = \mathbf{I}$.
+where $\mathbf{O}_{j}$ is the equilibrium proportions surviving age $a$ and occurring in each region and $\mathbf{O}_{0} = \mathbf{I}$, the identity matrix.
 Natural mortality and movement rates in the first year of the model are also used in Eq.
@@
 For the plus group $a=A$, $\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}$ is a ``fundamental matrix'' derived using the matrix version of the  geometric series [@kemenysnell60].
-Recall that recruitment for stock $s$ only occurs in region $r_s$ so, the equilibrium initial numbers at age $a$ by region are
-$$\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}^T \mathbf{N}_{O,1,1}.$$
+Recall that recruitment for a given stock $s$ only occurs in region $r_s$ so, the equilibrium initial numbers at age $a$ by region are
+$$\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a,r_s,\cdot} N_{1,1,r_s},$$
+where $\widetilde{\mathbf{O}}_{a,r_s,\cdot}$ is the vector of values in row $r_s$ of $\widetilde{\mathbf{O}}_{a}$.
 Initial abundance at age can also be estimated as random effects as described in Appendix B.
@@
 If survival and movement occur simultaneously, the parameters are estimated with a log link function  and if they are separable, an additive logit link function (like a multinomial regression) is used.
-On the transformed scale, the general model for the movement parameter from region $r$ to $r'$ in season $i$ and year $y$ for individuals of age $a$ is a linear function of an intercept or mean parameter ($\theta_{r\rightarrow r',i}$) and both random and environmental effects:
-\begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
-\end{equation*}
-The random effects $\varepsilon_{r\rightarrow r',y,a,i}$ are season-, and region-to-region-specific and modeled most generally as 2DAR1 random effects with age and(or) year where the covariance is
-\begin{equation*}
-  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
-\end{equation*}
-similar to how WHAM models variation in survival, natural mortality, and selectivity.
-Effects of covariate $E_k$ can be age-, season-, and region-to-region-specific $\beta_{r\rightarrow r',a,i,k}$ and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.
+On the transformed scale, the general model for the movement parameter from region $r$ to $r'$ in season $i$ and year $y$ for individuals of age $a$ is a linear function of an intercept or mean parameter ($\theta_{r\rightarrow r',i}$) and both random and covariate effects:
+\begin{equation*}
+  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} X_{k,y}.
+\end{equation*}
+The random effects $\varepsilon_{r\rightarrow r',y,a,i}$ are region-to-region- and season-specific, and modeled most generally as 2DAR1 random effects with age and(or) year, where the covariance is
+\begin{equation*}
+  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}.
+\end{equation*}
+This is similar to how WHAM models variation in survival, natural mortality, and selectivity.
+Covariate $X_k$ effects ($\beta_{r\rightarrow r',a,i,k}$) can be region-to-region-, age-, and season-specific, and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.
 
@@
 Therefore, movement parameters would usually either need to be fixed or assumed to have some prior distribution, possibly based on external parameter estimates.
-WHAM includes prior distributions for the season and region-to-region specific (mean) movement parameters and treats them as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
-  \begin{equation*}
-  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
+WHAM includes prior distributions for the region-to-region- and season-specific (mean) movement parameters and treats them as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
+  \begin{equation*}
+  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{\gamma,r\rightarrow r',i}\right).
   \end{equation*}
@@
   \begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y}.
+  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} X_{k,y}.
   \end{equation*}
@@
 When random effects are used, the same 2DAR1 structure with age and year as described by @stockmiller21 can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
-See Appendix B for details.
+We provide details in Appendix B.
 
@@
 The log-normal distributional assumption for aggregate catch observations is the same as @stockmiller21, but the predicted catch is now a function of catch from each stock starting the year in each region and any seasonal movement.
-For a given stock and age, the vector of numbers captured by each fleet over the year are
-$$\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}^T \mathbf{N}_{O,s,y,a}$$
-(recall each fleet can harvest in only one region as described above). The vector of predicted numbers caught by each fleet across stocks is 
-$$\widehat{\mathbf{N}}_{H,y,a} = \sum^{n_S}_{s=1} \widehat{\mathbf{N}}_{H,s,y,a}$$
+For a given stock $s$ at age $a$, the vector of numbers captured by each fleet over the year $y$ is
+$$\mathbf{N}_{H,s,y,a} = \mathbf{H}_{s,y,a}^T \mathbf{N}_{O,s,y,a}$$
+(recall each fleet can harvest in only one region as described above). The vector of numbers caught by each fleet across stocks is 
+$$\mathbf{N}_{H,y,a} = \sum^{n_S}_{s=1} \mathbf{N}_{H,s,y,a}$$
 and the vector of predicted aggregate catch by each fleet at age $a$ is
-$$\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \widehat{\mathbf{N}}_{H,y,a}$$
+$$\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \mathbf{N}_{H,y,a},$$
 where $\mathbf{c}_{y,a}$ is the vector of mean individual weight at age $a$ for each fleet and the aggregate catch by fleet is 
@@
 The log-aggregate catch observations for fleet $f$ are normally distributed
-$$ \log C_{y,f} \sim \text{N}\left(\log \widehat {C}_{y,f}, \sigma^2_{y,f}\right).$$
-
-The predicted numbers caught for each fleet $f$ (row $f$ of $\widehat{\mathbf{N}}_{H,y,a}$) are used to make predicted age composition observations as described by @stockmiller21.
+$$ \log C_{f,y} \sim \text{N}\left(\log \widehat {C}_{f,y}, \sigma^2_{f,y}\right),$$
+
+where $\widehat C_{f,y}$ is an element of $\widehat{\mathbf{C}}_y$.
+The predicted numbers caught for each fleet $f$ (row $f$ of $\mathbf{N}_{H,y,a}$) are used to make predicted age composition observations as described by @stockmiller21.
 Since then, three additional likelihood options for age composition observations have been added: a logistic-normal with AR(1) correlation structure [@francis14], the alternative Dirichlet-multinomial parameterization described by @thorsonetal17, and the multivariate Tweedie [@thorsonetal23].
@@
 
-Like fishing fleets, an index $m$ can observe the population in a single region $r_m$ at fraction of the year $t_m$, the predicted abundance at $t_m$ in region $r_m$ is 
-$$\widehat{N}_{s,y,a,m} = \mathbf{O}_{s,y,a,r_m}^T(t_m) \mathbf{N}_{O,s,y,a}$$
-where $\mathbf{O}_{s,y,a,r_m}(t_m)$, the $r_m$ column of the upper-left submatrix of Eq.
-\ref{eq:ptm_index}, are the probabilities of surviving and occurring in region $r_m$ at time $t_m$ given being alive in each region at the start of the year.
+An index $m$ can observe any stocks that occur in a single region $r_m$ at fraction of the year $t_m$. The abundance at $t_m$ in region $r_m$ is 
+$$N_{m,s,y,a} = \mathbf{O}_{s,y,a,\cdot,r_m}^T(t_m) \mathbf{N}_{O,s,y,a},$$
+where $\mathbf{O}_{s,y,a,\cdot,r_m}(t_m)$, column $r_m$ of the upper-left submatrix of Eq.
+\ref{eq:ptm_index}, is the vector of proportions surviving and occurring in region $r_m$ at time $t_m$ given being alive in each region at the start of the year.
 The predicted index at age is
-$$\widehat{I}_{m,y,a} = q_{m,y} \text{sel}_{m,y,a}w_{m,y,a}\sum^{n_S}_{s = 1}\widehat{N}_{s,y,a,m}$$
-where $q_{m,y}$ is the catchability of the index in year $y$, $\text{sel}_{m,y,a}$ is the selectivity and $w_{m,y,a}$ is the average weight of individuals at age $a$ if the index is quantified in biomass and $w_{m,y,a} = 1$ if the index is quantified in numbers.
+$$\widehat{I}_{m,y,a} = q_{m,y} \text{sel}_{m,y,a}w_{m,y,a}\sum^{n_S}_{s = 1}N_{m,s,y,a},$$
+where $q_{m,y}$ is the catchability of the index in year $y$, $\text{sel}_{m,y,a}$ is the selectivity, and $w_{m,y,a}$ is the average weight of individuals at age $a$ if the index is quantified in biomass and $w_{m,y,a} = 1$ if the index is quantified in numbers.
 Predicted age composition observations are functions of $\widehat{I}_{m,y,a}$ as described by @stockmiller21 and the likelihood options are the same as those for catch explained above.
-See Appendix B for options to model temporal variation in catchability.
+We describe options to model temporal variation in catchability in Appendix B.
 
@@
 
-Currently a single fishing mortality reference point $\widetilde F$ is estimated across stocks and regions and the fleet- and age-specific reference point is $\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}$.
-Selectivity is determined across fleets and ages as in the equilibrium assumption for initial abundance at age except here it is a function of average F at age over a user-defined set of years
+WHAM calculates annual reference points using biological inputs each year and it also provides the same reference points under prevailing conditions where biological inputs are averaged over a user-defined set of recent years.
+A single fishing mortality reference point $\widetilde{F}$ is estimated across stocks and regions and the fleet- and age-specific reference point is $\widetilde{F}_{f,a} = \widetilde{F} \text{sel}_{f,a}$.
+Selectivity is determined across fleets and ages as in the equilibrium assumption for initial abundance at age except here it is a function of fishing mortality at age, possibly averaged over years,
 \begin{equation}\label{eq:spr_sel}
-  \text{sel}_{a,f} = \frac{\overline F_{a,f}}{\max_a \sum^{n_F}_{f=1}{\overline F}_{a,f}}
-\end{equation}
-
-The equilibrium spawning stock biomass per recruit for a given stock in region $r_s$ is defined as
+  \text{sel}_{f,a} = \frac{\overline{F}_{f,a}}{\max_a \sum^{n_F}_{f=1}{\overline{F}}_{f,a}}.
+\end{equation}
+
+The equilibrium spawning stock biomass-per-recruit for a given stock in region $r_s$ is defined as
 \begin{equation}\label{eq:ssbpr}
- \upphi(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{O}_{a,\cdot,r_s}(t_s) w_{a} m_{a}
-\end{equation}
-where $w_{a}$ and $m_{a}$ are the mean individual weight and probability of maturity at age $a$, $\widetilde{\mathbf{O}}_{a}$ are as described in Eq.
-\ref{eq:npraa}, and $\mathbf{O}_{a}(t_s)$ is the $n_R \times n_R$ upper-left sub-matrix of eq.
-\ref{eq:ptm_spawn} with the probabilities of surviving and occurring in each region $r'$ at age $a+t_s$ given starting in region $r$ at age $a$.
-The further subscripts $r_s,\cdot$ and $\cdot,r_s$ indicate row or column $r_s$, respectively.
+ \upphi(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{O}_{a,\cdot,r_s}(t_s) w_{a} m_{a},
+\end{equation}
+where $w_{a}$ and $m_{a}$ are the mean individual weight and proportion mature at age $a$, $\widetilde{\mathbf{O}}_{a,r_s,\cdot}$ is row $r_s$ of $\widetilde{\mathbf{O}}_a$ in Eq.
+\ref{eq:npraa}, and $\mathbf{O}_{a,\cdot,r_s}(t_s)$ is column $r_s$ of the matrix of proportions surviving and occurring in each region at age $a+t_s$ (the upper-left sub-matrix of eq.
+\ref{eq:ptm_spawn}).
 Using these rows and columns is required because of the assumption that spawning and recruitment only occur in region $r_s$.
-
-The equilibrium spawning biomass per recruit (eq.
-\ref{eq:ssbpr}) is conditional on the region of recruitment $r_s$.
-The equilibrium recruitment depends on the stock dynamics.
-WHAM currently only allows complete spawning region fidelity so that a stock only spawns and recruits in a single region.
-Therefore, the equilibrium vector of numbers at age 1 for each state, $\widetilde {\mathbf{N}}_{1}$, will only be nonzero in the row corresponding to the spawning region ($r_s$).
-The matrices of probabilities of surviving and occurring in each region, $\widetilde{\mathbf{O}}_{s,a}$ and $\mathbf{O}_{s,a}(t_s)$, are functions of the fishing mortality rates for fleets in each region $\widetilde F_{f,a}$.
-
-The matrix of equilibrium yield per recruit by region and fleet as a function of $\widetilde F$ is calculated as
+For prevailing conditions, the years to average for $w_a$, $m_a$, and natural mortality and movement rates to parameterize $\mathbf{O}_{a}(t_s)$ and $\widetilde{\mathbf{O}}_{a}$, are defined by the user.
+
+The equilibrium yield-per-recruit as a function of $\widetilde{F}$ for a given stock $s$ is calculated as
 \begin{equation}\label{eq:ypr}
- \widetilde{\nu}({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{H}_{a} \mathbf{c}_{s,a}
-\end{equation}
-where $\mathbf{c}_a$ is the vector of mean individual weight at age for each fleet, and $\mathbf{H}_{s,a}$ is the submatrix of the probabilities of being captured in each fleet over the interval from $a$ to $a+1$, defined in eq.
+ \upnu({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{H}_{a} \mathbf{c}_{a},
+\end{equation}
+where $\mathbf{c}_{a}$ is the vector of mean individual weight at age for each fleet, and $\mathbf{H}_{a}$ is the submatrix of the proportions captured in each fleet over the interval from $a$ to $a+1$, defined in eq.
 \ref{eq:ptm}.
-
-As in previous versions of the WHAM package, reference points for prevailing conditions, average all of the inputs to the spawning biomass and yield per recruit calculations (weight, maturity, natural mortality, and movement) over a set of user-specified years as shown in eq.
-\ref{eq:spr_sel} for selectivity (e.g., last 5 years of the model).
-
-For $X\%$ SPR-based reference points, we use a Newton method and iterate for steps $j = 1,\ldots,10$
+Other than mean individual weight, $\upnu({\widetilde{F}})$ uses all of the same (averaged) inputs as $\upphi({\widetilde{F}})$ in Eq.
+\ref{eq:ssbpr}.
+
+For reference points based on reductions of $\upphi({\widetilde{F}})$ to $p\%$ of the unfished value, $\upphi(0)$ ($p/100$ is known as the spawning potential ratio; SPR), we use a Newton-Raphson method and iterate for steps $j = 1,\ldots,10$
 \begin{equation}\label{eq:newton}
-  \log\widetilde{F}^{j} = \log\widetilde{F}^{j-1} - \frac{h\left(\log\widetilde{F}^{j-1}\right)}{h^{(1)}\left(\log\widetilde{F}^{j-1}\right)}
-\end{equation}
-where $h(\log F)$ is the difference between the weighted sums of spawning biomass per recruit at $F$ and $X$\% of unfished spawning biomass per recruit across stocks:
+  \log\widetilde{F}^{j} = \log\widetilde{F}^{j-1} - \frac{h\left(\log\widetilde{F}^{j-1}\right)}{h^{(1)}\left(\log\widetilde{F}^{j-1}\right)},
+\end{equation}
+where $h(\log \widetilde{F})$ is the sum of weighted differences between $\upphi({\widetilde{F}})$ and $p$\% of $\upphi_s\left(0\right)$ across stocks:
 \begin{equation}\label{eq:newton-obj}
-  h(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
-\end{equation}
-where $\upphi_s\left(F=0\right)$ is the equilibrium unfished spawning biomass per recruit.
-$h^{(1)}(\log F)$ is the derivative of $h$ with respect to $\log F$, and the weights to use for each stock $\lambda_s$ can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate "static" equilibrium spawning biomass and yield.
-
-When a Beverton-Holt or Ricker stock recruit relationship is assumed, an analogous Newton method is used to find $\log F$ that maximizes yield for MSY-based reference points, which are also a functions of the equilibrium yield per recruit (Eq.
-\ref{eq:ypr}) and equilibrium recruitment.
-The function $h(\log F)$ and $h^{(1)}(\log F)$ in Eq.
-\ref{eq:newton} are the first and second derivatives of the sum of the stock-specific equilibrium yields with respect to $\log F$ [analogous to that described by @milleretal16].
+  h(\log \widetilde{F}) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(e^{\log \widetilde{F}}\right) - \frac{p}{100}\upphi_s\left(0\right)\right],
+\end{equation}
+and $h^{(1)}(\log \widetilde{F})$ is the derivative of $h(\log \widetilde{F})$ with respect to $\log \widetilde{F}$ [analogous to the approach for SPR-based reference points in @millerlegault17].
+The weights for each stock ($\lambda_s$) can be specific values supplied by the user or defined to be the proportions of average recruitment for each stock where the years used to average are those the user defines to calculate prevailing equilibrium spawning biomass and yield reference points, which are the product of the defined average recruitment and $\upphi({\widetilde{F}})$ and $\upnu({\widetilde{F}})$, respectively.
+The Newton-Raphson method is used by WHAM internally to allow uncertainty in all estimated parameters to be propagated into uncertainty estimates for the reference points.
+
+When a Beverton-Holt or Ricker stock-recruit relationship is assumed, we use the derivative (with respect to $\widetilde{F}$) of the weighted sum of stock-specific equilibrium yields (the product of $\upnu({\widetilde{F}})$ and equilibrium recruitment from the stock-recruit relationship) as $h(\log \widetilde{F})$ to find $\widetilde{F}$ which maximizes the weighted total yield across stocks for MSY-based reference points [$F_{\text{MSY}}$, analogous to the approach for MSY-based reference points in @milleretal16].
 
@@
 
-The projection options are generally the same as those for previous versions of WHAM [@millerstock20].
+The projection options are generally the same as those for previous versions of WHAM.
 When there is movement of any stocks, the user has the option to project and use any random effects for time-varying movement or use the average over user specified years, analogous to how natural mortality can be treated in the projection period.
@@
 
-Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging the entire east coast of the United States.
+BSB are a temperate reef fish in the western Atlantic Ocean ranging the entire east coast of the United States.
 Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity [@ablehales97;@fabrizioetal13].
-Fish in this northern stock component perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring [@musickmercer77].
+Fish in this NEUS stock perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring [@musickmercer77].
 Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude [@mosershepherd09].
 
-Prior to its 2023 peer-reviewed assessment, the NEUS black sea bass stock was assessed using the Age-Structured Assessment Program (ASAP) model [@legaultrestrepo99], a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
-Northern and southern components of the NEUS black sea bass stock ascribed to regions divided by the Hudson Canyon were separately modeled in ASAP (Figure \ref{fig:map}).
+Prior to its 2023 peer-reviewed assessment, the NEUS BSB stock was assessed using the Age-Structured Assessment Program (ASAP) model [@legaultrestrepo99], a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
+Northern and southern components of the NEUS BSB stock ascribed to regions divided by the Hudson Canyon were separately modeled in ASAP (Figure \ref{fig:map}).
 Results from the separate ASAP models were combined for a unit-stock assessment.
@@
 
-Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as "Working Group") composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above [@nefsc23].
-This assessment included the two regions and two stock components and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
-Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.
+Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as "Working Group") composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the BSB stock using the multi-stock and multi-region extension of WHAM described above [@nefsc23].
+This assessment included the two regions and two stock components and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous BSB assessments.
+Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on BSB.
 
@@
 
-The first year being modeled for the population is 1989 and the fishery and index data used in the model span from 1989 to 2021.
+The first year modeled for the population is 1989 and the fishery and index data used in the model span from 1989 to 2021.
 The north and south stock components are modeled as separate populations that spawn and recruit in respective regions and there are eight age classes with the last being a plus group ($a \in \{1,\ldots,8+\}$).
@@
 for the portion of the year before spawning.
-As noted in the description of the general WHAM model, tagging data are not yet allowed.
+As noted in the description of the general WHAM model, tagging data are not yet included.
 However, the Working Group also fit a Stock Synthesis model [@methotwetzel13] which included tagging data and provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).
@@
 With movement as configured, the northern origin fish (ages 2+) can occur in the southern region on January 1.
-Estimating initial numbers at age as separate parameters can be challenging even in single-stock models, but for black sea bass the available data cannot distinguish the proportion of northern and southern component fish at each age in the southern region in the initial year of the model.
-Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component: an initial recruitment and an equilibrium fully-selected $F$ (across all fleets) that determines the initial abundance at age in each region for each stock component.
-
-For subsequent years, all models allow AR1 correlation of recruitment and 2DAR1 correlation for apparent survival random effects for both the northern and southern stock components.
-Models with apparent survival random effects will model the transitions (conditional on both survival and movement) of abundances at age of northern origin fish in each region, but there appears to be little ability of the existing data from the southern region to distinguish apparent survival random effects for each stock component.
+Estimating initial numbers at age as separate parameters can be challenging even in single-stock models, but for BSB the available data cannot distinguish the proportion of northern and southern component fish at each age in the southern region in the initial year of the model.
+Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component (Eq.
+\ref{eq:npraa}).
+
+For subsequent years, all models have AR1 correlation of recruitment and 2DAR1 correlation for apparent survival random effects for both the northern and southern stock components.
+Models with apparent survival random effects will model the transitions (conditional on both survival and movement) of abundances at age of northern origin fish in each region, but there appears to be little information in the existing data from the southern region to distinguish apparent survival random effects for each stock component.
 All models we considered assume a very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model.
@@
 
-## Uncertainty in recreational index observations {-}
-
-The estimated coefficients of variation (CVs) provided by the analyses by the Working Group to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the Working Group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
-All models we considered allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
-Although lack of model convergence was observed in many fits for self-test simulations, the estimation of critical model output (e.g., SSB, fishing mortality) was reliable and robust to the poor estimation of these scalars (Figure \ref{fig:self-test-fig}), and estimating these parameters rather than fixing them in the fit to the actual data allows uncertainty in model output to be more properly conveyed.
-
-## Index and catch age composition observations {-}
-
-The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual and retrospective patterns.
-These analyses resulted in use of selectivity random effects for the northern fleet and indices, logistic-normal likelihoods for six sets of age composition observations, and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
-
 ## Bottom temperature effects {-}
@@
 All models include observations of winter bottom temperature anomalies for the northern and southern regions from 1959 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius [@nefsc23].
-See the Supplementary Materials for details about the derivation of these observations.
+We provide details about the derivation of these observations in the Supplementary Materials.
 We retained the assumption from the peer-reviewed assessment that treated the latent bottom temperature covariates in each region as AR1 processes.
@@
 
-All models treat annual recruitment as AR1 time-varying random effects with mean parameters for each stock component.
-However, @milleretal18 showed how inferences of temperature effects on growth or maturity parameters can be very different depending on whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
-Therefore, we also explored whether including temporal random effects on age 1 natural mortality affected inferences on corresponding temperature effects.
-Initially, we included random effects on age 1 natural mortality for both the northern and southern stock components, but estimates of these random effects and corresponding variance for the southern component converged to 0 so these were not included in models presented here.
-
 We assume the covariate in year $y$ affects recruitment in the same year because the covariate observations are from winter months.
@@
 Expected log-recruitment for a given stock component is a linear function of bottom temperature
-
 \begin{equation}\label{eq:expected-recruitment}
-E\left(\log N_{y,1}|x_y\right) = \mu_{R} + \beta_{R} x_y.
-\end{equation}
-Similarly, expected log-natural morality as a function of bottom temperature is
+E\left(\log N_{y,1,r_s}|X_{y}\right) = \theta_{R} + \beta_{R} X_y.
+\end{equation}
+and the annual recruitment is
+\begin{equation}\label{eq:Rec-re}
+\log N_{y,1,r_s} = E\left(\log N_{y,1,r_s}|X_y\right) + \varepsilon_{y,1,r_s},
+\end{equation}
+where $\varepsilon_{y,1,r_s}$ are the normally distributed random effects with AR1 correlation structure.
+
+Similarly, expected log-natural morality rate as a function of bottom temperature is
 \begin{equation}\label{eq:expected-M1}
-E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y.
-\end{equation}
-Because age 1 fish for the northern component can exist in both regions after January 1 other the spawning period, natural mortality is acting in both regions for this stock.
-For models with covariate and(or) annual random effects for age 1 fish we assume them only in the northern region for the northern component.
-The corresponding random effects are
-\begin{equation}\label{eq:Rec-re}
-\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \varepsilon_{y,1}
-\end{equation}
-and
+E\left(\log M_{y,1,r_s}|X_y\right) = \theta_{M,a,r_s} + \beta_{M} X_y.
+\end{equation}
+The mean log natural mortality rate is assumed to be $\theta_{M,a,r} = \log(0.4)$ for all ages and regions as recommended by the Working Group.
+@milleretal18 showed how inferences of temperature effects on growth or maturity parameters can be very different depending on whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
+Therefore, we also explored whether including temporal random effects on age 1 natural mortality affected inferences on corresponding temperature effects.
+The model for annual natural mortality for a given stock is
 \begin{equation}\label{eq:M-re}
-\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \varepsilon_{M,y}.
-\end{equation}
-The constant or mean log-natural mortality rate is assumed to be $\mu_{M,a} = \log(0.4)$ for all ages as recommended by the Working Group.
-Because the bottom temperature anomalies and random effects are centered at 0, mean log-natural mortality at age 1 over the time series should be approximately equal to $\mu_{M,1}$ for models that include those effects.
+\log M_{y,1,r_s} = E\left(\log M_{y,1,r_s}|X_y\right) + \varepsilon_{M,y,1,r_s}.
+\end{equation}
+Because the bottom temperature anomalies are centered and $E(\varepsilon_{M,y,1,r_s}) = 0$, mean log-natural mortality at age 1 over the time series should be approximately equal to $\theta_{M,a,r_s}$ for models that include those effects.
+
+Initially, we included random effects on age 1 natural mortality for both the northern and southern stock components, but estimates of these random effects and corresponding variance for the southern component converged to 0 so these were not included in models presented here.
+Because age 1 fish for the northern component can exist in both regions after January 1, other than during the spawning period, natural mortality is acting in both regions for this stock.
+For models with covariate and(or) annual random effects for age 1 fish, we use them only in the northern region for the northern component.
+
+## Uncertainty in recreational index observations {-}
+
+The estimated coefficients of variation (CVs) provided by the analyses by the Working Group to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the Working Group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
+All models we considered allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
+Although lack of model convergence was observed in many fits for self-test simulations, the estimation of critical model output (e.g., SSB, fishing mortality) was reliable and robust to the poor estimation of these scalars (Figure \ref{fig:self-test-fig}), and estimating these parameters rather than fixing them in the fit to the actual data allows uncertainty in model output to be more properly conveyed.
+
+## Index and catch age composition observations {-}
+
+The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual and retrospective patterns.
+These analyses resulted in use of selectivity random effects for the northern fleet and indices, logistic-normal likelihoods for six sets of age composition observations, and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
 
@@
 Therefore, there are eight fits of each model with the time series reduced by zero to seven years.
-We  calculated Mohn's $\rho$ for SSB, and average $F$ at ages 6 and 7.
-Absolute values of Mohn's $\rho$ near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended.
+We  calculated Mohn's $\rho$ for SSB, and average fishing mortality rate at ages 6 and 7.
+Absolute values of Mohn's $\rho$ near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended [@mohn99].
 As in @milleretal16, we also assessed the consistency of the AIC-based model selection over retrospective peels to guard against previously noted changes in perception of covariate effects on recruitment with increased length of the time series of observations [@myers98].
@@
 We performed a jitter analysis of the base model $M_0$ and the best fitting model to investigate whether a local mimimum of the negative log-likelihood surface was obtained by the optimization.
-We used the jitter_wham function in the WHAM package which by default simulates starting values from a multivariate normal distribution with mean and covariance defined by the MLEs and hessian-based covariance matrix from the fitted model.
-See the Supplementary Materials for more details.
+We used the \verb|jitter_wham| function in the WHAM package which by default simulates starting values from a multivariate normal distribution with mean and covariance defined by the MLEs and hessian-based covariance matrix from the fitted model (Supplementary Materials).
 
@@
 We estimated median relative bias of SSB across these simulations.
-Finally, we fitted three sensitivity models which differed from the best performing model, by 1) assuming no movement for either stock component, 2) movement parameters are fixed at the means for the prior distributions, and 3) a constant natural mortality rate is estimated across all regions and stocks.
-See the Supplementary Materials for more details.
+Finally, we fitted three sensitivity models which differed from the best performing model, by 1) assuming no movement for either stock component, 2) movement parameters are fixed at the means for the prior distributions, and 3) a constant natural mortality rate is estimated across all regions and stocks (Supplementary Materials).
 
@@
 
-We produced SPR-based BRP estimates ($F_{40\%}$ and SSB$_{40\%}$) from the best performing model using a variety of options provided in WHAM.
-We generated annual BRP estimates based on the annual weight at age, maturity at age, natural mortality, fleet selectivity, and either the annual recruitment random effect (Eq.
+We produced SPR-based reference point estimates ($F_{40\%}$ and SSB$_{40\%}$, setting $p=40$ in eq.
+\ref{eq:newton-obj}) from the best performing model using a variety of options provided in WHAM.
+We generated annual reference point estimates based on the annual weight at age, maturity at age, natural mortality, fleet selectivity, and either the annual recruitment random effect (Eq.
 \ref{eq:Rec-re}) or the annual expected recruitment (Eq.
 \ref{eq:expected-recruitment}).
-We also calculated status as the ratios of annual average F and SSB to the corresponding reference points.
-Finally, we calculated "prevailing" BRPs using the average weight at age, maturity at age, M, and fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
-All BRPs and status estimates are generated for each region as well as for the total stock area.
-
-We projected the best black sea bass model under three alternative scenarios for the bottom temperature covariate where 1) the AR1 time-series model continues into the projection years, 2) the average of the most recent 5 years is projected, and 3) bottom temperature increases in projection years following a prediction from a simple linear regression of the estimated bottom temperature anomalies over time.
+We also calculated status as the ratios of annual average fishing mortality rates and SSB to the corresponding reference points.
+Finally, we calculated "prevailing" reference points using the average weight, maturity, natural mortality rate, and fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
+All reference points and status estimates are generated for each region as well as for the total stock area.
+
+We projected the best BSB model under three alternative scenarios for the bottom temperature covariate where 1) the AR1 time-series model continues into the projection years, 2) the average of the most recent 5 years is projected, and 3) bottom temperature increases in projection years following a prediction from a simple linear regression of the estimated bottom temperature anomalies over time.
 We specified fishing mortality in projection years to be constant at the value in the last model year (2021).
-Average weight, natural mortality, movement, and maturity at age over the last 5 years are used in projection years whereas the AR1 processes for all numbers at age random effects are continued.
+Average weight, natural mortality, movement, and maturity at age over the last 5 years are used in projection years, whereas the AR1 processes for all numbers at age random effects are continued.
 
@@
 
-Using the expected recruitment rather than the random effects as the multiplier with equilibrium spawning biomass per recruit ($\upphi(\widetilde{F})$, eq.
+Using the expected recruitment rather than the random effects as the multiplier with $\upphi(\widetilde{F})$ (eq.
 \ref{eq:ssbpr}) results in less variable estimates of annual SSB$_{40\%}$ (Figure \ref{fig:annual-BRPs}: top row).
-Estimates were less variable for the southern stock component than the northern stock component mainly because the expected recruitment is constant for the former whereas that for the northern component varies via the effect of the bottom temperature anomalies.
-Temporal variation for the southern stock component is mainly caused by annual variation in the inputs to the $\upphi(\widetilde{F})$ calculation and $F_{40\%}$, but there is also some effect of the variation in the proportion of total recruitment for each stock component as used in the weighting terms to define the 40\% reduction in unfished spawning biomass per recruit ($\lambda_s$ in Eq.
+Estimates were less variable for the southern stock component than the northern stock component mainly because the expected recruitment is constant for the former, whereas that for the northern component varies via the effect of the bottom temperature anomalies.
+Temporal variation for the southern stock component is mainly caused by annual variation in the inputs to the $\upphi(\widetilde{F})$ calculation and $F_{40\%}$, but there is also some effect of the variation in the proportion of total recruitment for each stock component as used in the weighting terms to define the 40\% reduction in $\upphi(0)$ ($\lambda_s$ in Eq.
 \ref{eq:newton-obj}).
-Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB$_{40\%}$ for the southern stock component whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
+Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB$_{40\%}$ for the southern stock component, whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
 
@@
 For similar reasons, estimates of uncertainty of $F_{40\%}$ are also similar between the type of recruitment estimates (Figure \ref{fig:annual-F40-cvs}).
-Annual estimates for the northern stock component were lowest in the earliest years of the time series whereas those for the southern component are relatively stable over time.
+Annual estimates for the northern stock component were lowest in the earliest years of the time series, whereas those for the southern component are relatively stable over time.
 
@@
 When continuing the AR1 model for the bottom temperature anomalies in the projection years, the uncertainty grows and asymptotes to a level determined by the marginal variance of the process and the predicted value approaches the estimated marginal mean value of approximately 0 (Figure \ref{fig:R-BT-proj}).
-There is no uncertainty in the projected linear trend because the values were specified as known whereas there is a small level of uncertainty in the projections of the recent average based on the uncertainty in the estimated mean.
+There is no uncertainty in the projected linear trend because the values were specified as known, whereas there is a small level of uncertainty in the projections of the recent average based on the uncertainty in the estimated mean.
 Comparing projected recruitment for the northern and southern stock components, the projected expected recruitment has much lower uncertainty than the projected random effects regardless of whether bottom temperature affects recruitment.
-For the northern component, CVs of projected expected recruitments are smaller than those of recruitment random effects, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for the future bottom temperature anomalies are made (Figure \ref{fig:R-F-SSB-CVs}).
+For the northern component, CVs of projected expected recruitments are smaller than those of recruitment random effects, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for future bottom temperature anomalies (Figure \ref{fig:R-F-SSB-CVs}).
 
@@
 
-In our investigation of bottom temperature effects on recruitment and natural mortality and time-varying random effects on natural mortality at age 1, we found only including effects on recruitment for the northern stock component to be the best model with respect to AIC.
-Black sea bass in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north [@mcmahanetal20].
+In our investigation of bottom temperature effects on recruitment and natural mortality, and time-varying random effects on natural mortality at age 1, we found only including effects on recruitment for the northern stock component to be the best model with respect to AIC.
+BSB in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north [@mcmahanetal20].
 Opposite effects of temperature on recruitment or population size would be expected for species in the same general area that are at the southern extent of their range [@gabriel92].
@@
 Continued monitoring of temperature effects on the southern component and investigations of non-linear effects of temperature for both components may be prudent because of the possibility of an optimal temperature range for survival via effects on growth during the pre-recruit phase [@warelambert85;@anderson88].
-Furthermore, our inferences about effects on natural mortality at age 1 may be influences by our assumptions of constant median (or constant depending on the model) natural mortality rate and movement rates for all ages.
+Furthermore, our inferences about effects on natural mortality at age 1 may be influenced by our assumptions of constant median (or just constant, depending on the model) natural mortality rate and movement rates for all ages.
 Further investigations of variation in these parameters with age and effects of bottom temperature could improve our understanding of the demography for this stock.
 
-We showed that using expected recruitment rather than recruitment random effects to define BRPs can produce the same estimates but with lower uncertainty.
-Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same BRPs and even lower uncertainty.
+We showed that using expected recruitment rather than recruitment random effects to define reference points can produce the same estimates but with lower uncertainty.
+Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same reference points and even lower uncertainty.
 These results can be important for management when uncertainty in assessment output is used in making catch advice.
-The typical approach for defining stock and harvest status is to compare BRPs defined under prevailing conditions with current stock and harvest levels, but we could also make the comparisons with projected BRPs like those we demonstrated are possible.
+The typical approach for defining stock and harvest status is to compare reference points defined under prevailing conditions with current stock and harvest levels, but, as we demonstrated, it is also possible to make the comparisons of projected stock and harvest levels with projected reference points.
 The uncertainty in these ratios could also be used in defining catch advice.
 
-BRPS under prevailing conditions use average values for recruitment and other inputs and MSY-based BRPs are based on the stock-recruit relationship which is also typically constant over time.
-If BRPs should be defined in such a way that they are less variable than the stock over time, expected inputs to the BRPs (instead of annual values) would be more appropriate.
-Lower variability in BRPs would presumably correlate with lower variability in catch advice which might be desirable [e.g., @cochraneetal98;@kelletal05], but the appropriate level of temporal variability in BRPs and catch advice will dependent on the characteristics of the fish stock, associated fisheries, and management structure [@holland10;@puntetal16].
-
-For black sea bass, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figures \ref{fig:R-BT-proj} and \ref{fig:F-SSB-proj}).
+Reference points under prevailing conditions use average values for recruitment and other inputs and MSY-based reference points are based on the stock-recruit relationship which is also typically constant over time.
+If reference points should be defined in such a way that they are less variable than the stock over time, expected inputs to the reference points (instead of annual values) would be more appropriate.
+Lower variability in reference points would presumably correlate with lower variability in catch advice which might be desirable [e.g., @cochraneetal98;@kelletal05], but the appropriate level of temporal variability in reference points and catch advice will depend on the characteristics of the fish stock, associated fisheries, and management structure [@holland10;@puntetal16].
+
+For BSB, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figures \ref{fig:R-BT-proj} and \ref{fig:F-SSB-proj}).
 In the projection period, the AR1 models for bottom temperature and recuitment revert to the long-term means of those processes.
 This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypotheses of increasing water temperature and changing environmental conditions [@hareetal16;@pershingetal21].
-Because of these concerns, the most recent stock assessment for yellowtail flounder used a change point analysis on environmental covariates to determine current conditions.
-The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections [@nefsc24a].
+Because of these concerns, the most recent stock assessment for yellowtail flounder [@nefsc24a] used a change point analysis on environmental covariates to determine current conditions.
+The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections.
 Unfortunately, this methodology still has limitations because it is unlikely that current environmental conditions are stable.
 Similarly, a linear increase of temperature is also not likely in the future (top row of Figure \ref{fig:E-R-SSB-F}).
-As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in BRPs and population projections becomes an important decision in the management process.
+As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in reference points and population projections becomes an important decision in the management process.
 
@@
 
-WHAM provides a large class of models for time- and age-varying random effects for population attributes and treatment of environmental covariates and their effects on populations using state-space methods [@maunder24].
+WHAM provides a large class of models for time- and age-varying random effects for population attributes and treatment of environmental covariates and their effects on populations using state-space methods.
 This framework accounts for the magnitude of differing uncertainties in observations and stochastic population dynamics processes.
-Version 2.0 extends these inferences to movement rates for multi-region models and stock (component) specific configuration of random and environmental covariate effects.
+Version 2.0 extends these inferences to movement rates for multi-region models and stock(component)-specific configuration of random and environmental covariate effects.
 
@@
 Such observations are critical to estimation of movement parameters [e.g., @goetheletal19].
-We used estimates of movement parameters from a companion assessment model to define prior distributions for corresponding movement parameters in the state-space black sea bass model, but other than the tagging data, both of the assessment models used much the same data which can lead to inappropriate inferences.
+We used estimates of movement parameters from a companion assessment model to define prior distributions for corresponding movement parameters in the state-space BSB model, but other than the tagging data, both of the assessment models used much the same data which can lead to inappropriate inferences.
 In lieu of integrating an observation model for the tagging data, a better approach would be to estimate the movement parameters externally from just the tagging data.
@@
 Tagging data can also inform mortality rate parameters [@hampton91].
-It is well known that natural mortality is seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality from other assessment model parameters [@leeetal11;@clark22].
+It is well known that natural mortality rates are seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality rates from other assessment model parameters [@leeetal11;@clark22].
 Estimation of natural mortality may be even more challenging within state-space assessment models with their greater flexibility from inclusion of time-varying random effects.
@@
 Regardless, tagging observations can provide improved inferences and allowing tagging observations in WHAM should be a high priority even for models with a single stock and region.
-This extension would also be beneficial for black sea bass because the current assessment uses much the same data to estimate movement outside of the model as we are using to estimate the other parameters in the WHAM model.
-
-WHAM version 2.0 only allows models where each stock spawns and recruits in a single region however, some stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region [@kapuretal21].
+
+WHAM version 2.0 only allows models where each stock spawns and recruits in a single region; however, some stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region [@kapuretal21].
 Therefore, extending WHAM to allow such models would be beneficial.
@@
 
-Our application to black sea bass is a particular combination of the spatial modeling options described by @bergeretal24.
+Our application to BSB is a particular combination of the spatial modeling options described by @bergeretal24.
 Other combinations of the described options are possible in WHAM 2.0.
@@
 We express our gratitude to the many scientists who have collected and processed the data that went into the assessment model, including survey biologists, port samplers, observers and agers.
-We thank the black sea bass research track assessment working group and peer-reviewers for the work compiling data and determining many of the configurations of the black sea bass assessment model that we used for our analyses.
+We thank the BSB research track assessment working group and peer-reviewers for the work compiling data and determining many of the configurations of the BSB assessment model that we used for our analyses.
 We also thank Chris Legault, two anonymous reviewers, and the associate editor for comments and suggestions that improved the clarity of this paper.
 
+\pagebreak
+
 # References {-}
@@
 
-(ref:caption) The Northwest Atlantic coastal area where the NEUS black sea bass stock occurs.
-Red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations [after @milleretal16_yoy_survival] and the bathymetric contour is the 400 m isocline.
+(ref:caption) The Northwest Atlantic coastal area where the Northeast US black sea bass stock occurs.
+The red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations [after @milleretal16_yoy_survival] and the bathymetric contour is the 400 m isocline.
 
@@
 
-```{r migration-diagram, echo = FALSE, out.width="80%", fig.align="center", fig.cap = "Diagram of intervals within the year and configuation of the dynamics of each component of the BSB population."}
+```{r migration-diagram, echo = FALSE, out.width="80%", fig.align="center", fig.cap = "Diagram of intervals within the year and configuation of the dynamics of each component of the black sea bass population."}
 knitr::include_graphics(here::here("paper","bsb_movement_diagram.pdf"))
@@
 
-```{r move-prior-posterior, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Prior (black) and posterior (red) distributions of movement of the northern stock component from north to south (top) and south to north (bottom).
-Vertical lines are the prior and posterior estimates."}
+```{r move-prior-posterior, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Prior (black) and posterior (red) distributions of the movement parameter for the northern stock component from north to south (top) and south to north (bottom).
+Vertical lines indicate the prior and posterior estimates."}
 knitr::include_graphics(here::here("paper", "move_prior_post.pdf"))
@@
 ```{r E-R-SSB-F, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual observations and posterior estimates of bottom temperature anomalies (top row), annual expected and random effect recruitment estimates (second row) and SSB and average fishing mortality at ages 6 and 7 from model $M_1$.
-Columns define estimates regionally (spawning region for SSB) and toal SSB and fishing mortality across regions.
+Columns define estimates regionally (spawning region for SSB) and total SSB and fishing mortality across regions.
 Polygons and vertical lines represent 95\\% confidence intervals."}
@@
 
-```{r annual-BRPs, echo = FALSE, eval= TRUE, out.height = "85%", fit.align = "center", fig.cap = "Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio, equilibrium SSB at $F_{40\\%}$ based on annual inputs to $\\upphi(\\widetilde{F})$ calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to $\\upphi(\\widetilde{F})$ averaged over the last 5 years and recruitment is averaged over the time series.
+```{r annual-BRPs, echo = FALSE, eval= TRUE, out.height = "85%", fit.align = "center", fig.cap = "Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio ($F_{40\\%}$), equilibrium SSB at $F_{40\\%}$ based on annual inputs to $\\upphi(\\widetilde{F})$ calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to $\\upphi(\\widetilde{F})$ averaged over the last 5 years and recruitment is averaged over the time series.
 Red and blue in first 4 rows and black and gray in final row indicate alternative annual estimates of recruitment (random effect or expected).
@@
 ```{r R-BT-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of bottom temperature in the northern region and alternative recruitment estimates (random effect or expected) by region.
-Estimates from years after 2021 are from projections of model $M1$ under three alternative assumptions for the bottom temperature anomalies in the northern region.
-Vertical dotted line is the last year of data and polygons represent 95\\% confidence intervals."}
+Estimates from years after 2021 are from projections of model $M_1$ under three alternative assumptions for the bottom temperature anomalies in the northern region.
+Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."}
 knitr::include_graphics(here::here("paper", "proj_ecov_Recruit_results.pdf"))
@@
 Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted line is the last year of data and polygons represent 95\\% confidence intervals."}
+Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."}
 knitr::include_graphics(here::here("paper", "proj_F_SSB.pdf"))
@@
 Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region and average $\\upphi(\\widetilde{F})$ inputs from the last 5 years of the unprojected model.
-Vertical dotted line is the last year of data and polygons represent 95\\% confidence intervals."}
+Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."}
 knitr::include_graphics(here::here("paper", "proj_brps.pdf"))
@@
 mod_table %>% kable(format = type, booktabs = T, escape=F, row.names = F, label = "model-desc-table",
-    caption="Assumptions for temperature effects and random effects for age 1 natural mortality for each model.", align = c(rep("l",4))) %>%
+    caption="Assumptions for temperature effects and random effects for age 1 natural mortality rates ($M$) for each model.", align = c(rep("l",4))) %>%
   add_header_above(c(" " = 1, "Temperature Effect" = 2, " " = 1 )) %>% kable_styling(latex_options = "hold_position")
@@
 \renewcommand\thetable{A\arabic{table}}
-\input{symboltable_R1.tex}
+\input{symboltable_R2.tex}
 
@@
 
-New features of WHAM 2.0 that are not used in the application to black sea bass are described here..
+New features of WHAM 2.0 that are not used in the application to BSB are described here..
 
@@
  \mathbf{A}_{y,a,i} = \begin{bmatrix}
- a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
- \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
+ a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{1,y,a,i} & 0 & M_{y,a,1} \\
+ \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{2,y,a,i} & M_{y,a,2} \\
  0 & 0 & 0 & 0 & 0 \\
@@
  0 & 0 & 0 & 0 & 0
- \end{bmatrix}
-\end{equation*}
-where $a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})$.
-When there is one region, $n_f$ fleets, and $\delta_i = 1$, exponentiating the instantaneous rate matrix results in the PTM defined in Eq.
+ \end{bmatrix},
+\end{equation*}
+where $a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + M_{y,a,r} + \sum^{n_F}_{f:r_f=r}F_{f,y,a,i})$.
+When there is one region, $n_F$ fleets, and $\delta_i = 1$, exponentiating the instantaneous rate matrix results in the PTM defined in Eq.
 \ref{eq:ptm_1_region}.
@@
 \begin{equation*}
-\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}
+\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r},
 \end{equation*}
@@
 The general model for natural mortality at age $a$ for a given stock in region $r$ is a function of random and covariate effects
-\begin{equation*}
-  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
-\end{equation*}
+\begin{equation}\label{eq:M_model}
+  \log M_{y,a,r} = \theta_{M,a,r} + \varepsilon_{M,y,a,r} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} X_{k,y} = E\left(\log M_{y,a,r}|X_{k,y}\right) + \varepsilon_{M,y,a,r}.
+\end{equation}
 The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
@@
 \begin{equation*}
-log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}
+\log \left(\frac{q_{m,y}-l_m}{u_m-q_{m,y}}\right) = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} X_{k,y},
 \end{equation*}
@@
 The roots of the matrices are calculated using an eigen decomposition of the matrices
-$$ \boldsymbol{\mu}_k =  \left(\boldsymbol{\mu}_k^*\right)^{z_k} = \mathbf{V}_k \mathbf{D}_k^{z_k} \mathbf{V}_k^{-1}$$
+$$ \boldsymbol{\mu}_k =  \left(\boldsymbol{\mu}_k^*\right)^{z_k} = \mathbf{V}_k \mathbf{D}_k^{z_k} \mathbf{V}_k^{-1},$$
 where $z_1 = 1/5$ for and $z_2 = 1/4$, and $\mathbf{V}_{k}$ and $\mathbf{D}_{k}$ are the matrix of eigenvectors (columnwise) and the diagonal matrix of corresponding eigenvalues of $\boldsymbol{\mu}^*_k$.
@@
 ```{r R-F-SSB-CVs, echo = FALSE, eval= TRUE, out.height = "90%", fit.align = "center", fig.cap = "Coefficients of variation for estimates of alternative recruitment estimates (random effects or expected), average fishing mortality at age 6 and 7, and SSB by region and in total from model $M_1$.
-Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
+Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."}
 knitr::include_graphics(here::here("paper", "R_SSB_F_cv_results.pdf"))
@@
 ```{r annual-F40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."}
 knitr::include_graphics(here::here("paper", "proj_F40_CV.pdf"))
@@
 ```{r annual-SSB40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium SSB$_{40\\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations and alternative annual recruitment types.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."}
 knitr::include_graphics(here::here("paper", "proj_SSB40_CV.pdf"))
@@
 Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted line is the last year of data and polygons represent 95\\% confidence intervals."}
+Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."}
 knitr::include_graphics(here::here("paper", "proj_F_status_results.pdf"))
@@
 Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted line is the last year of data and polygons represent 95\\% confidence intervals."}
+Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."}
 knitr::include_graphics(here::here("paper", "proj_SSB_status_results.pdf"))
@@
 ```{r annual-F-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\\bar{F}_{40\\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."}
 knitr::include_graphics(here::here("paper", "proj_F_status_CV.pdf"))
@@
 ```{r annual-SSB-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies."}
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."}
 knitr::include_graphics(here::here("paper", "proj_SSB_status_CV.pdf"))

### 2025-07-01 paper/paper_final.Rmd
final

--- 2025-07-01 paper/paper_R2.Rmd
@@
 author: 
-  - Timothy J. Miller^1^
-  - Kiersten L. Curti
-  - Alexander C. Hansell
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+  - Timothy J. Miller^1^^,2^
+  - Kiersten L. Curti^1^^,3^
+  - Alexander C. Hansell^1^^,4^
+# date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
@@
     includes:
-      in_header: options.sty
-#    citation_package: biblatex
-#biblio-style: authoryear
-#  pdf_document:
-#    keep_tex: yes
-#    fig_caption: yes
-#    number_sections: yes
-#    includes:
-#      in_header: options.sty
-#  html_document:
-#    df_print: paged
+      in_header: options_main_final.sty
 csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
@@
 ```
-
-$^1$timothy.j.miller@noaa.gov, Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\
+$1$ Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\
+$2$ orcid: 0000-0003-1411-1206, timothy.j.miller@noaa.gov\
+$3$ orcid: 0009-0009-1676-7269\
+$4$ orcid: 0000-0001-7827-1749\
 
@@
 Much of the description below is for a specific stock $s$, but, for simplicity, this subscript is implicit except when necessary.
+We have provided an Appendix A Table \ref{symbols} containing definitions of all mathematical notation in this paper.
 
@@
 Estimated reference points in the first few projection years can differ depending on whether the expected recruitment or the random effects are used in their calculation, but longer term projected values are identical because the random effect estimates approach the expected values as the influence of observations diminishes.
-However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures \ref{fig:annual-F40-cvs} and \ref{fig:annual-SSB40-cvs}).
+However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures \ref{fig:annual-SSB40-cvs} and \ref{fig:annual-F40-cvs}).
 The reduction in uncertainty using expected recruitment is less for $F_{40\%}$ than SSB$_{40\%}$ and for the northern component which includes bottom temperature effects than the southern component.
@@
 
+# Competing Interests {-}
+
+The authors declare there are no competing interests.
+
+# Data and Code Availability {-}
+
+As described in the methods section, all data and code for conducting analyses and creating the manuscript are available at [DOI].
+
 \pagebreak
@@
 The red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations [after @milleretal16_yoy_survival] and the bathymetric contour is the 400 m isocline.
+Map created with datum NAD83 and Latitude-Longitude coordinate system.
 
@@
 knitr::include_graphics(here::here("paper","bsb_movement_diagram.pdf"))
-```
-\pagebreak
-
-```{r move-prior-posterior, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Prior (black) and posterior (red) distributions of the movement parameter for the northern stock component from north to south (top) and south to north (bottom).
-Vertical lines indicate the prior and posterior estimates."}
-knitr::include_graphics(here::here("paper", "move_prior_post.pdf"))
 ```
@@
 knitr::include_graphics(here::here("paper", "E_R_SSB_F_fig.pdf"))
+```
+\pagebreak
+
+```{r move-prior-posterior, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Prior (black) and posterior (red) distributions of the movement parameter for the northern stock component from north to south (top) and south to north (bottom).
+Vertical lines indicate the prior and posterior estimates."}
+knitr::include_graphics(here::here("paper", "move_prior_post.pdf"))
 ```
@@
 
-\pagebreak
-
-# Supplementary Materials {-}
-
-## Deriving the prior distribution for movement parameters {-}
-
-The Working Group fit a Stock Synthesis model [@methotwetzel13] that included tagging data with 2 seasons (6 months each) and 2 regions where a proportion $\mu^*_1$ of the northern component moves to the south in one season and some proportion  $\mu^*_{2\rightarrow 1}$ move back to the south in the second season [@nefsc23].
-The seasonal movement matrices for each season are
-\begin{equation*}
-\boldsymbol{\mu}^*_{1} = 
-  \begin{bmatrix}
-     1-\mu^*_{1\rightarrow 2} & \mu^*_{1\rightarrow 2} \\
-     0 & 1 \\
-  \end{bmatrix}
-\end{equation*}
-and
-\begin{equation*}
-\boldsymbol{\mu}_{2} = 
-  \begin{bmatrix}
-     1 &  0 \\
-     \mu^*_{2\rightarrow 1} & 1-\mu^*_{2\rightarrow 1} \\
-  \end{bmatrix}.
-\end{equation*}
-
-To obtain estimates of movement proportions for the monthly intervals in the WHAM model, the half-year movement matrices were converted to monthly movement matrices by taking the root $z_k$ of $\boldsymbol{\mu}^*_{k}$ which are defined by the number of months of movement for each season (5 and 4, respectively).
-The roots of the matrices are calculated using an eigen decomposition of the matrices
-$$ \boldsymbol{\mu}_k =  \left(\boldsymbol{\mu}_k^*\right)^{z_k} = \mathbf{V}_k \mathbf{D}_k^{z_k} \mathbf{V}_k^{-1},$$
-where $z_1 = 1/5$ for and $z_2 = 1/4$, and $\mathbf{V}_{k}$ and $\mathbf{D}_{k}$ are the matrix of eigenvectors (columnwise) and the diagonal matrix of corresponding eigenvalues of $\boldsymbol{\mu}^*_k$.
+<!-- \pagebreak -->
+
+<!-- # Supplementary Materials {-} -->
+
+<!-- ## Deriving the prior distribution for movement parameters {-} -->
+
+<!-- The Working Group fit a Stock Synthesis model [@methotwetzel13] that included tagging data with 2 seasons (6 months each) and 2 regions where a proportion $\mu^*_1$ of the northern component moves to the south in one season and some proportion  $\mu^*_{2\rightarrow 1}$ move back to the south in the second season [@nefsc23].
+The seasonal movement matrices for each season are -->
+<!-- \begin{equation*} -->
+<!-- \boldsymbol{\mu}^*_{1} =  -->
+<!--   \begin{bmatrix} -->
+<!--      1-\mu^*_{1\rightarrow 2} & \mu^*_{1\rightarrow 2} \\ -->
+<!--      0 & 1 \\ -->
+<!--   \end{bmatrix} -->
+<!-- \end{equation*} -->
+<!-- and -->
+<!-- \begin{equation*} -->
+<!-- \boldsymbol{\mu}_{2} =  -->
+<!--   \begin{bmatrix} -->
+<!--      1 &  0 \\ -->
+<!--      \mu^*_{2\rightarrow 1} & 1-\mu^*_{2\rightarrow 1} \\ -->
+<!--   \end{bmatrix}. -->
+<!-- \end{equation*} -->
+
+<!-- To obtain estimates of movement proportions for the monthly intervals in the WHAM model, the half-year movement matrices were converted to monthly movement matrices by taking the root $z_k$ of $\boldsymbol{\mu}^*_{k}$ which are defined by the number of months of movement for each season (5 and 4, respectively).
+The roots of the matrices are calculated using an eigen decomposition of the matrices -->
+<!-- $$ \boldsymbol{\mu}_k =  \left(\boldsymbol{\mu}_k^*\right)^{z_k} = \mathbf{V}_k \mathbf{D}_k^{z_k} \mathbf{V}_k^{-1},$$ -->
+<!-- where $z_1 = 1/5$ for and $z_2 = 1/4$, and $\mathbf{V}_{k}$ and $\mathbf{D}_{k}$ are the matrix of eigenvectors (columnwise) and the diagonal matrix of corresponding eigenvalues of $\boldsymbol{\mu}^*_k$.
 The Working Group used a parametric bootstrap approach to determine an appropriate standard deviation for the prior distribution for the movement parameters.
-Stock Synthesis also estimates parameters on a transformed scale, but different from WHAM:
-$$\mu^*_{r\rightarrow r'} = \frac{1}{1 + 2e^{-x_{r\rightarrow r'}}}$$
-The estimated parameters and standard errors from the Stock Synthesis model were $x_{1\rightarrow 2}=-1.44$ and $x_{2\rightarrow 1}=1.94$ and $SE(x_{1\rightarrow 2})) = 0.21$ and $SE(x_{2\rightarrow 1})) = 0.37$.
+Stock Synthesis also estimates parameters on a transformed scale, but different from WHAM:  -->
+<!-- $$\mu^*_{r\rightarrow r'} = \frac{1}{1 + 2e^{-x_{r\rightarrow r'}}}$$ -->
+<!-- The estimated parameters and standard errors from the Stock Synthesis model were $x_{1\rightarrow 2}=-1.44$ and $x_{2\rightarrow 1}=1.94$ and $SE(x_{1\rightarrow 2})) = 0.21$ and $SE(x_{2\rightarrow 1})) = 0.37$.
 The resulting in the estimated proportions were $\mu^*_{1\rightarrow 2}=0.11$ and $\mu^*_{2\rightarrow 1}=0.78$.
-
-In WHAM, an additive logit transformation is used which is simply a logit transformation when there are only two regions: 
-$$
-\mu_{r\rightarrow r'} = \frac{1}{1+e^{-y_{r\rightarrow r'}}}.
-$$
-We simulated 1000 values from a normal distribution with mean and standard deviation defined by the parameter estimate and standard error $\tilde x_{{r\rightarrow r'},b} \sim N(x_{r\rightarrow r'}, SE(x_{r\rightarrow r'}))$ from the Stock Synthesis model.
+-->
+
+<!-- In WHAM, an additive logit transformation is used which is simply a logit transformation when there are only two regions:  -->
+<!-- $$ -->
+<!-- \mu_{r\rightarrow r'} = \frac{1}{1+e^{-y_{r\rightarrow r'}}}. -->
+<!-- $$ -->
+<!-- We simulated 1000 values from a normal distribution with mean and standard deviation defined by the parameter estimate and standard error $\tilde x_{{r\rightarrow r'},b} \sim N(x_{r\rightarrow r'}, SE(x_{r\rightarrow r'}))$ from the Stock Synthesis model.
 For each simulated value we constructed $\tilde {\boldsymbol{\mu}}^*_{{r\rightarrow r'},b}$, took the appropriate root and calculated inverse logit for $\tilde y_{{r\rightarrow r'},b}$.
@@
 The mean values did not differ meaningfully from the transformation of the original estimates ($y_{1\rightarrow 2} = -3.79$ and $y_{2\rightarrow 1} = -0.79$) and the standard deviation was approximately 0.2 for both parameters.
-
-## Bottom temperature anomalies {-}
-
-The Working Group created bottom temperature observations from a high resolution ocean bottom temperature product by @dupontaviceetal23.
+-->
+
+<!-- ## Bottom temperature anomalies {-} -->
+
+<!-- The Working Group created bottom temperature observations from a high resolution ocean bottom temperature product by @dupontaviceetal23.
 The annual observations for each region are defined by the average over all spatial bottom temperature values for February and March by region and year.
@@
 We created regional bottom temperature anomalies by subtracting means for each region across all years.
-
-## Diagnostics {-}
-
-### Jitter fits for model $M_0$ {-}
-
-WHAM by default completes three newton steps after the stats::nlminb minimization function completes to reduce the gradient at the minimized NLL.
+-->
+
+<!-- ## Diagnostics {-} -->
+
+<!-- ### Jitter fits for model $M_0$ {-} -->
+
+<!-- WHAM by default completes three newton steps after the stats::nlminb minimization function completes to reduce the gradient at the minimized NLL.
 However, this generally has negligible effects on model estimates and the NLL.
@@
 Without the Newton steps, the maximum (absolute) gradient sizes are generally less than 0.01 for models that converge satisfactorily.
-
-The 50 jitter fits demonstrated that a local minimum was obtained for the original fit of model $M_0$ (Figure \ref{fig:jitter-M0}).
+-->
+
+<!-- The 50 jitter fits demonstrated that a local minimum was obtained for the original fit of model $M_0$ (Figure \ref{fig:jitter-M0}).
 One lower NLL was obtained with unacceptable gradients (No.
@@
 9) did not provide a non-zero estimate of the variance parameter for one of the indices and the other two provided identical results and we refit model $M_0$ and all remaining models using the better parameter estimates as initial values.
-
-### Jitter fits for model $M_1$ {-}
-
-The 50 jitter fits gave no evidence of a better minimization of the NLL.
+-->
+
+<!-- ### Jitter fits for model $M_1$ {-} -->
+
+<!-- The 50 jitter fits gave no evidence of a better minimization of the NLL.
 Three lower NLLs were obtained, but with unacceptably large gradients (Figure \ref{fig:jitter-M1}).
 The largest differences in parameter estimates for these three jitters were for numbers at age and selectivity random effects variance and correlation parameters.
-
-### Self test for model $M_1$ {-}
-
-Initial fits to simulated data from model $M_1$ showed estimation of the observation error standard deviation multiplier for the recreational catch-per angler indices in the north and south regions was unstable.
+-->
+
+<!-- ### Self test for model $M_1$ {-} -->
+
+<!-- Initial fits to simulated data from model $M_1$ showed estimation of the observation error standard deviation multiplier for the recreational catch-per angler indices in the north and south regions was unstable.
 Many of the fits to the simulated data produced implausible estimates at the 0 boundary for these parameters (very negative values on log-scale).
@@
 We also fit analogous models with the multiplier parameters fixed at the true values, which did improve convergence, but larger bias was estimated for fishing mortality and SSB for the northern component.
-
-\setcounter{table}{0}
-\renewcommand\thetable{S\arabic{table}}
-
-```{r, echo = F, eval = T}
-age_comp_names <- c("North commercial fleet", "North recreational fleet", "South commercial fleet", "South recreational fleet",
-  "North recreational CPA index", "North VAST index", "South recreational CPA index", "South VAST index")
-age_comp_mod <- c("Dirichlet-Multinomial", "Logistic-normal (Independent)", "Logistic-normal (AR1 correlation)","Logistic-normal (AR1 correlation)", "Logistic-normal (Independent)","Dirichlet-Multinomial","Logistic-normal (AR1 correlation)","Logistic-normal (AR1 correlation)")
-mean_sel_mod <- c("age-specific (ages > 3 fully selected)",
-  "age-specific (ages > 6 fully selected)",
-  "logistic", "logistic",
-  "age-specific (ages > 1 fully selected)",
-  "age-specific (ages > 4 fully selected)",
-  "age-specific (ages > 2 fully selected)",
-  "age-specific (ages > 1 fully selected)")
-sel_re_mod <- c("AR1 correlation by age and year", 
-  "AR1 correlation by age and year",
-  "None",
-  "None",
-  "AR1 correlation by year",
-  "AR1 correlation by age and year",
-  "None",
-  "None")
-out <- cbind.data.frame("Data component" = age_comp_names,  "Age Composition Likelihood" = age_comp_mod,  
-  "Mean Selectivity model" = mean_sel_mod, "Random effects Model" = sel_re_mod)
-
-out %>% kable(format = type, booktabs = T, escape=F, row.names = F, label = "age-comp-sel-table",
-    caption="Configuration of age composition likelihoods, mean selectivity models, and selectivity random effects models for each age composition data component.
-For all logistic-normal likelihoods, any ages observed as zeros are treated as missing.", align = c(rep("l",4))) %>% landscape()
-#knitr::knit_print(kableExtra::landscape(out))
-
-```
-
-```{r, echo = F, eval = T}
-aic_wts <- readRDS("../results/aic_wts.RDS")
-aic_wts <- data.frame(round(aic_wts, 2))
-aic_wts <- cbind(Model = row.names(aic_wts), aic_wts)
-row.names(aic_wts) <- NULL
-colnames(aic_wts) <- c("Model", as.character(0:7))
-aic_wts %>%
-  kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "aic-wts-table",
-  caption = "Model AIC weights for each retrospective peel.") %>%
-  add_header_above(c(" " = 1, "Peel" = 8))
-```
-
-\setcounter{figure}{0}
-\renewcommand\thefigure{S\arabic{figure}}
-
-\pagebreak
-
-```{r jitter-M0, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_0$.
-Black jitters had maximum absolute gradient values < $10^{-2}$ and red jitters had values > 1."}
-knitr::include_graphics(here::here("paper", "fit_0_jitter_plt.pdf"))
-```
-\pagebreak
-
-```{r jitter-M1, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_1$.
-Fits with black dots had maximum absolute gradient value < 0.01 and fits with red dots had values > 10."}
-knitr::include_graphics(here::here("paper", "fit_1_jitter_plt.pdf"))
-```
-\pagebreak
-
-```{r self-test-fig, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Median relative error of SSB (Total and by stock component) and total fully-selected fishing mortality for estimation models fitted to simulated data from model $M_1$ where the observation variance of log-indices are fixed and estimated.
+-->
+
+<!-- \setcounter{table}{0} -->
+<!-- \renewcommand\thetable{S\arabic{table}} -->
+
+<!-- ```{r, echo = F, eval = T} -->
+<!-- age_comp_names <- c("North commercial fleet", "North recreational fleet", "South commercial fleet", "South recreational fleet", -->
+<!--   "North recreational CPA index", "North VAST index", "South recreational CPA index", "South VAST index") -->
+<!-- age_comp_mod <- c("Dirichlet-Multinomial", "Logistic-normal (Independent)", "Logistic-normal (AR1 correlation)","Logistic-normal (AR1 correlation)", "Logistic-normal (Independent)","Dirichlet-Multinomial","Logistic-normal (AR1 correlation)","Logistic-normal (AR1 correlation)") -->
+<!-- mean_sel_mod <- c("age-specific (ages > 3 fully selected)", -->
+<!--   "age-specific (ages > 6 fully selected)", -->
+<!--   "logistic", "logistic", -->
+<!--   "age-specific (ages > 1 fully selected)", -->
+<!--   "age-specific (ages > 4 fully selected)", -->
+<!--   "age-specific (ages > 2 fully selected)", -->
+<!--   "age-specific (ages > 1 fully selected)") -->
+<!-- sel_re_mod <- c("AR1 correlation by age and year",  -->
+<!--   "AR1 correlation by age and year", -->
+<!--   "None", -->
+<!--   "None", -->
+<!--   "AR1 correlation by year", -->
+<!--   "AR1 correlation by age and year", -->
+<!--   "None", -->
+<!--   "None") -->
+<!-- out <- cbind.data.frame("Data component" = age_comp_names,  "Age Composition Likelihood" = age_comp_mod,   -->
+<!--   "Mean Selectivity model" = mean_sel_mod, "Random effects Model" = sel_re_mod) -->
+
+<!-- out %>% kable(format = type, booktabs = T, escape=F, row.names = F, label = "age-comp-sel-table", -->
+<!--     caption="Configuration of age composition likelihoods, mean selectivity models, and selectivity random effects models for each age composition data component.
+For all logistic-normal likelihoods, any ages observed as zeros are treated as missing.", align = c(rep("l",4))) %>% landscape() -->
+<!-- #knitr::knit_print(kableExtra::landscape(out)) -->
+
+<!-- ``` -->
+
+<!-- ```{r, echo = F, eval = T} -->
+<!-- aic_wts <- readRDS("../results/aic_wts.RDS") -->
+<!-- aic_wts <- data.frame(round(aic_wts, 2)) -->
+<!-- aic_wts <- cbind(Model = row.names(aic_wts), aic_wts) -->
+<!-- row.names(aic_wts) <- NULL -->
+<!-- colnames(aic_wts) <- c("Model", as.character(0:7)) -->
+<!-- aic_wts %>% -->
+<!--   kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "aic-wts-table", -->
+<!--   caption = "Model AIC weights for each retrospective peel.") %>% -->
+<!--   add_header_above(c(" " = 1, "Peel" = 8)) -->
+<!-- ``` -->
+
+<!-- \setcounter{figure}{0} -->
+<!-- \renewcommand\thefigure{S\arabic{figure}} -->
+
+<!-- \pagebreak -->
+
+<!-- ```{r self-test-fig, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Median relative error of SSB (Total and by stock component) and total fully-selected fishing mortality for estimation models fitted to simulated data from model $M_1$ where the observation variance of log-indices are fixed and estimated.
 Black and Red dashed lines represent the median of the annual medians and the median across all annual relative errors, respectively.
-Vertical lines represent 95\\% confidence intervals."}
-knitr::include_graphics(here::here("paper", "self_test_results.pdf"))
-```
-
-\begin{landscape}
-
-```{r bottom-temperature, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Observations with 95\\% confidence intervals (points with vertical lines) and posterior estimates with 95\\% confidence intervals (lines with polygons) of bottom temperature anomalies in the north and south regions from model $M_1$.
-Gray vertical line defines the first year that the black sea bass stock is modeled."}
-knitr::include_graphics(here::here("paper", "BTA_full_fig.pdf"))
-```
-
-\end{landscape}
-\pagebreak
-
-```{r BT-Ecov-R, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Expected and random effect recruitment estimates for the northern stock component.
-Color of points defined by the corresponding annual bottom temperature anomaly."}
-knitr::include_graphics(here::here("paper", "best_R_Ecov.pdf"))
-```
-\pagebreak
-
-```{r SSB-F-R-rel-M1, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Estimates of SSB, F, and recruitment relative to those of the best performing model, $M_1$."}
-knitr::include_graphics(here::here("paper", "SSB_F_R_rel_M1.pdf"))
-```
-
-```{r Ecov-M1-rel-M0, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Relative differences in posterior estimates of northern region bottom temperature anomalies ($\\widehat X$) from the null model without effects on recruitment ($M_0$) and with effects on the northern stock component ($M_1$)."}
-knitr::include_graphics(here::here("paper", "Ecov_M1_rel_M0.pdf"))
-```
-
-```{r selectivity-south, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Selectivty for fleets and indices in the southern region."}
-knitr::include_graphics(here::here("paper", "selectivity_south_plot.pdf"))
-```
-\pagebreak
-
-\begin{landscape}
-
-```{r R-F-SSB-CVs, echo = FALSE, eval= TRUE, out.height = "90%", fit.align = "center", fig.cap = "Coefficients of variation for estimates of alternative recruitment estimates (random effects or expected), average fishing mortality at age 6 and 7, and SSB by region and in total from model $M_1$.
+Vertical lines represent 95\\% confidence intervals."} -->
+<!-- knitr::include_graphics(here::here("paper", "self_test_results.pdf")) -->
+<!-- ``` -->
+
+<!-- ```{r BT-Ecov-R, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Expected and random effect recruitment estimates for the northern stock component.
+Color of points defined by the corresponding annual bottom temperature anomaly."} -->
+<!-- knitr::include_graphics(here::here("paper", "best_R_Ecov.pdf")) -->
+<!-- ``` -->
+
+<!-- \begin{landscape} -->
+<!-- ```{r bottom-temperature, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Observations with 95\\% confidence intervals (points with vertical lines) and posterior estimates with 95\\% confidence intervals (lines with polygons) of bottom temperature anomalies in the north and south regions from model $M_1$.
+Gray vertical line defines the first year that the black sea bass stock is modeled."} -->
+<!-- knitr::include_graphics(here::here("paper", "BTA_full_fig.pdf")) -->
+<!-- ``` -->
+<!-- \end{landscape} -->
+<!-- \pagebreak -->
+
+<!-- ```{r Ecov-M1-rel-M0, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Relative differences in posterior estimates of northern region bottom temperature anomalies ($\\widehat X$) from the null model without effects on recruitment ($M_0$) and with effects on the northern stock component ($M_1$)."} -->
+<!-- knitr::include_graphics(here::here("paper", "Ecov_M1_rel_M0.pdf")) -->
+<!-- ``` -->
+
+<!-- ```{r selectivity-south, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Selectivty for fleets and indices in the southern region."} -->
+<!-- knitr::include_graphics(here::here("paper", "selectivity_south_plot.pdf")) -->
+<!-- ``` -->
+
+<!-- \begin{landscape} -->
+
+<!-- ```{r sensitivity-plots, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Estimates of annual SSB and fishing mortality rates for the best performing model $M_1$ and models that are otherwise the same except where  1) movement rates are fixed at the means for the prior distribution, 2) a constant natural mortality rate is estimated, or 3) there is no movement for either stock component."} -->
+<!-- knitr::include_graphics(here::here("paper", "SSB_F_sensitivity_plots.pdf")) -->
+<!-- ``` -->
+<!-- \end{landscape} -->
+
+<!-- ```{r annual-SSB40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium SSB$_{40\\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations and alternative annual recruitment types.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."} -->
+<!-- knitr::include_graphics(here::here("paper", "proj_SSB40_CV.pdf")) -->
+<!-- ``` -->
+<!-- ```{r annual-SSB-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."} -->
+<!-- knitr::include_graphics(here::here("paper", "proj_SSB_status_CV.pdf")) -->
+<!-- ``` -->
+
+<!-- ```{r annual-F40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."} -->
+<!-- knitr::include_graphics(here::here("paper", "proj_F40_CV.pdf")) -->
+<!-- ``` -->
+
+<!-- ```{r annual-F-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\\bar{F}_{40\\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."} -->
+<!-- knitr::include_graphics(here::here("paper", "proj_F_status_CV.pdf")) -->
+<!-- ``` -->
+
+<!-- \begin{landscape} -->
+
+<!-- ```{r R-F-SSB-CVs, echo = FALSE, eval= TRUE, out.height = "90%", fit.align = "center", fig.cap = "Coefficients of variation for estimates of alternative recruitment estimates (random effects or expected), average fishing mortality at age 6 and 7, and SSB by region and in total from model $M_1$.
 Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data."}
-knitr::include_graphics(here::here("paper", "R_SSB_F_cv_results.pdf"))
-```
-\end{landscape}
-
-```{r annual-F40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data."}
-knitr::include_graphics(here::here("paper", "proj_F40_CV.pdf"))
-```
-
-```{r annual-SSB40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium SSB$_{40\\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations and alternative annual recruitment types.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data."}
-knitr::include_graphics(here::here("paper", "proj_SSB40_CV.pdf"))
-```
-
-```{r F-status-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of ratios of fishing mortality to $F_{40\\%}$ by region and in total.
+Vertical dotted lines indicate the last year of data."} -->
+<!-- knitr::include_graphics(here::here("paper", "R_SSB_F_cv_results.pdf")) -->
+<!-- ``` -->
+<!-- \end{landscape} -->
+
+<!-- \pagebreak -->
+
+<!-- ```{r jitter-M0, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_0$.
+Black jitters had maximum absolute gradient values < $10^{-2}$ and red jitters had values > 1."} -->
+<!-- knitr::include_graphics(here::here("paper", "fit_0_jitter_plt.pdf")) -->
+<!-- ``` -->
+<!-- \pagebreak -->
+
+<!-- ```{r jitter-M1, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_1$.
+Fits with black dots had maximum absolute gradient value < 0.01 and fits with red dots had values > 10."} -->
+<!-- knitr::include_graphics(here::here("paper", "fit_1_jitter_plt.pdf")) -->
+<!-- ``` -->
+<!-- \pagebreak -->
+
+<!-- ```{r SSB-F-R-rel-M1, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Estimates of SSB, F, and recruitment relative to those of the best performing model, $M_1$."} -->
+<!-- knitr::include_graphics(here::here("paper", "SSB_F_R_rel_M1.pdf")) -->
+<!-- ``` -->
+
+<!-- ```{r F-status-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of ratios of fishing mortality to $F_{40\\%}$ by region and in total.
 Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."}
-knitr::include_graphics(here::here("paper", "proj_F_status_results.pdf"))
-```
-```{r SSB-status-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of ratios of SSB to SSB$_{40\\%}$ by region and in total.
+Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."} -->
+<!-- knitr::include_graphics(here::here("paper", "proj_F_status_results.pdf")) -->
+<!-- ``` -->
+<!-- ```{r SSB-status-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of ratios of SSB to SSB$_{40\\%}$ by region and in total.
 Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."}
-knitr::include_graphics(here::here("paper", "proj_SSB_status_results.pdf"))
-```
-
-```{r annual-F-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\\bar{F}_{40\\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data."}
-knitr::include_graphics(here::here("paper", "proj_F_status_CV.pdf"))
-```
-
-```{r annual-SSB-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data."}
-knitr::include_graphics(here::here("paper", "proj_SSB_status_CV.pdf"))
-```
-
-\begin{landscape}
-
-```{r sensitivity-plots, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Estimates of annual SSB and fishing mortality rates for the best performing model $M_1$ and models that are otherwise the same except where  1) movement rates are fixed at the means for the prior distribution, 2) a constant natural mortality rate is estimated, or 3) there is no movement for either stock component."}
-knitr::include_graphics(here::here("paper", "SSB_F_sensitivity_plots.pdf"))
-```
-\end{landscape}
+Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."} -->
+<!-- knitr::include_graphics(here::here("paper", "proj_SSB_status_results.pdf")) -->
+<!-- ``` -->

### 2025-07-01 paper/paper_final.tex
final

--- 2025-06-17 paper/paper_R2.tex
@@
   pdftitle={Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass},
-  pdfauthor={Timothy J. Miller1; Kiersten L. Curti; Alexander C. Hansell},
+  pdfauthor={Timothy J. Miller1,2; Kiersten L. Curti1,3; Alexander C. Hansell1,4},
   hidelinks,
@@
 %\onehalfspacing
-\doublespacing
+%\doublespacing
 \usepackage{lineno}
-\linenumbers
+%\linenumbers
 \usepackage[belowskip=0pt,aboveskip=0pt]{caption}
@@
 \usepackage{upgreek}
-
+\usepackage{xr}
+\externaldocument{supp_final}
 \newcommand{\blandscape}{\begin{landscape}}
@@
 \title{Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass}
-\author{Timothy J. Miller\textsuperscript{1} \and Kiersten L. Curti \and Alexander C. Hansell}
-\date{16 June, 2025}
+\author{Timothy J. Miller\textsuperscript{1,2} \and Kiersten L. Curti\textsuperscript{1,3} \and Alexander C. Hansell\textsuperscript{1,4}}
+\date{}
 
@@
 
-\(^1\)\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}, Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\\
+\(1\) Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\\
+\(2\) orcid: 0000-0003-1411-1206, \href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
+\(3\) orcid: 0009-0009-1676-7269\\
+\(4\) orcid: 0000-0001-7827-1749\\
 
@@
 Much of the description below is for a specific stock \(s\), but, for simplicity, this subscript is implicit except when necessary.
+We have provided an Appendix A Table \ref{symbols} containing definitions of all mathematical notation in this paper.
 
@@
 Estimated reference points in the first few projection years can differ depending on whether the expected recruitment or the random effects are used in their calculation, but longer term projected values are identical because the random effect estimates approach the expected values as the influence of observations diminishes.
-However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures \ref{fig:annual-F40-cvs} and \ref{fig:annual-SSB40-cvs}).
+However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures \ref{fig:annual-SSB40-cvs} and \ref{fig:annual-F40-cvs}).
 The reduction in uncertainty using expected recruitment is less for \(F_{40\%}\) than SSB\(_{40\%}\) and for the northern component which includes bottom temperature effects than the southern component.
@@
 
+\hypertarget{competing-interests}{%
+\section*{Competing Interests}\label{competing-interests}}
+\addcontentsline{toc}{section}{Competing Interests}
+
+The authors declare there are no competing interests.
+
+\hypertarget{data-and-code-availability}{%
+\section*{Data and Code Availability}\label{data-and-code-availability}}
+\addcontentsline{toc}{section}{Data and Code Availability}
+
+As described in the methods section, all data and code for conducting analyses and creating the manuscript are available at {[}DOI{]}.
+
 \pagebreak
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-dupontaviceetal23}{}}%
-du Pontavice, H., Chen, Z., and Saba, V.S.
-2023.
-A high-resolution ocean bottom temperature product for the northeast {U}.{S}.
-Continental shelf marine ecosystem.
-Progress in Oceanography \textbf{210}: 102948.
-doi:\href{https://doi.org/10.1016/j.pocean.2022.102948}{10.1016/j.pocean.2022.102948}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-fabrizioetal13}{}}%
@@
 The red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations (after Miller et al.
-2016a) and the bathymetric contour is the 400 m isocline.}\label{fig:map}
+2016a) and the bathymetric contour is the 400 m isocline.
+Map created with datum NAD83 and Latitude-Longitude coordinate system.}\label{fig:map}
 \end{figure}
@@
 
-{\centering \includegraphics[width=1\linewidth]{move_prior_post}
-
-}
-
-\caption{Prior (black) and posterior (red) distributions of the movement parameter for the northern stock component from north to south (top) and south to north (bottom).
-Vertical lines indicate the prior and posterior estimates.}\label{fig:move-prior-posterior}
-\end{figure}
-\pagebreak
-
-\begin{figure}
-
 {\centering \includegraphics[height=0.95\textheight]{E_R_SSB_F_fig}
@@
 Polygons and vertical lines represent 95\% confidence intervals.}\label{fig:E-R-SSB-F}
+\end{figure}
+\pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{move_prior_post}
+
+}
+
+\caption{Prior (black) and posterior (red) distributions of the movement parameter for the northern stock component from north to south (top) and south to north (bottom).
+Vertical lines indicate the prior and posterior estimates.}\label{fig:move-prior-posterior}
 \end{figure}
@@
 
-\pagebreak
-
-\hypertarget{supplementary-materials}{%
-\section*{Supplementary Materials}\label{supplementary-materials}}
-\addcontentsline{toc}{section}{Supplementary Materials}
-
-\hypertarget{deriving-the-prior-distribution-for-movement-parameters}{%
-\subsection*{Deriving the prior distribution for movement parameters}\label{deriving-the-prior-distribution-for-movement-parameters}}
-\addcontentsline{toc}{subsection}{Deriving the prior distribution for movement parameters}
-
-The Working Group fit a Stock Synthesis model (Methot and Wetzel 2013) that included tagging data with 2 seasons (6 months each) and 2 regions where a proportion \(\mu^*_1\) of the northern component moves to the south in one season and some proportion \(\mu^*_{2\rightarrow 1}\) move back to the south in the second season (NEFSC 2023).
-The seasonal movement matrices for each season are
-\begin{equation*}
-\boldsymbol{\mu}^*_{1} = 
-  \begin{bmatrix}
-     1-\mu^*_{1\rightarrow 2} & \mu^*_{1\rightarrow 2} \\
-     0 & 1 \\
-  \end{bmatrix}
-\end{equation*}
-and
-\begin{equation*}
-\boldsymbol{\mu}_{2} = 
-  \begin{bmatrix}
-     1 &  0 \\
-     \mu^*_{2\rightarrow 1} & 1-\mu^*_{2\rightarrow 1} \\
-  \end{bmatrix}.
-\end{equation*}
-
-To obtain estimates of movement proportions for the monthly intervals in the WHAM model, the half-year movement matrices were converted to monthly movement matrices by taking the root \(z_k\) of \(\boldsymbol{\mu}^*_{k}\) which are defined by the number of months of movement for each season (5 and 4, respectively).
-The roots of the matrices are calculated using an eigen decomposition of the matrices
-\[ \boldsymbol{\mu}_k =  \left(\boldsymbol{\mu}_k^*\right)^{z_k} = \mathbf{V}_k \mathbf{D}_k^{z_k} \mathbf{V}_k^{-1},\]
-where \(z_1 = 1/5\) for and \(z_2 = 1/4\), and \(\mathbf{V}_{k}\) and \(\mathbf{D}_{k}\) are the matrix of eigenvectors (columnwise) and the diagonal matrix of corresponding eigenvalues of \(\boldsymbol{\mu}^*_k\).
-The Working Group used a parametric bootstrap approach to determine an appropriate standard deviation for the prior distribution for the movement parameters.
-Stock Synthesis also estimates parameters on a transformed scale, but different from WHAM:
-\[\mu^*_{r\rightarrow r'} = \frac{1}{1 + 2e^{-x_{r\rightarrow r'}}}\]
-The estimated parameters and standard errors from the Stock Synthesis model were \(x_{1\rightarrow 2}=-1.44\) and \(x_{2\rightarrow 1}=1.94\) and \(SE(x_{1\rightarrow 2})) = 0.21\) and \(SE(x_{2\rightarrow 1})) = 0.37\).
-The resulting in the estimated proportions were \(\mu^*_{1\rightarrow 2}=0.11\) and \(\mu^*_{2\rightarrow 1}=0.78\).
-
-In WHAM, an additive logit transformation is used which is simply a logit transformation when there are only two regions:
-\[
-\mu_{r\rightarrow r'} = \frac{1}{1+e^{-y_{r\rightarrow r'}}}.
-\]
-We simulated 1000 values from a normal distribution with mean and standard deviation defined by the parameter estimate and standard error \(\tilde x_{{r\rightarrow r'},b} \sim N(x_{r\rightarrow r'}, SE(x_{r\rightarrow r'}))\) from the Stock Synthesis model.
-For each simulated value we constructed \(\tilde {\boldsymbol{\mu}}^*_{{r\rightarrow r'},b}\), took the appropriate root and calculated inverse logit for \(\tilde y_{{r\rightarrow r'},b}\).
-We calculated the mean and standard deviation of the values \(y_{i,b}\).
-The mean values did not differ meaningfully from the transformation of the original estimates (\(y_{1\rightarrow 2} = -3.79\) and \(y_{2\rightarrow 1} = -0.79\)) and the standard deviation was approximately 0.2 for both parameters.
-
-\hypertarget{bottom-temperature-anomalies}{%
-\subsection*{Bottom temperature anomalies}\label{bottom-temperature-anomalies}}
-\addcontentsline{toc}{subsection}{Bottom temperature anomalies}
-
-The Working Group created bottom temperature observations from a high resolution ocean bottom temperature product by du Pontavice et al.
-(2023).
-The annual observations for each region are defined by the average over all spatial bottom temperature values for February and March by region and year.
-Similarly, the Working Group calculated standard errors from the standard deviation of all values in the region and the total number of values for a given year.
-We created regional bottom temperature anomalies by subtracting means for each region across all years.
-
-\hypertarget{diagnostics}{%
-\subsection*{Diagnostics}\label{diagnostics}}
-\addcontentsline{toc}{subsection}{Diagnostics}
-
-\hypertarget{jitter-fits-for-model-m_0}{%
-\subsubsection*{\texorpdfstring{Jitter fits for model \(M_0\)}{Jitter fits for model M\_0}}\label{jitter-fits-for-model-m_0}}
-\addcontentsline{toc}{subsubsection}{Jitter fits for model \(M_0\)}
-
-WHAM by default completes three newton steps after the stats::nlminb minimization function completes to reduce the gradient at the minimized NLL.
-However, this generally has negligible effects on model estimates and the NLL.
-To reduce computation time, we did not complete these newton steps when performing jitter fits of the model.
-Without the Newton steps, the maximum (absolute) gradient sizes are generally less than 0.01 for models that converge satisfactorily.
-
-The 50 jitter fits demonstrated that a local minimum was obtained for the original fit of model \(M_0\) (Figure \ref{fig:jitter-M0}).
-One lower NLL was obtained with unacceptable gradients (No.~25), but a slightly lower NLL was found with a satisfactory gradient for 3 of the jitters (Nos.
-9, 13, 29).
-However, one of the jitter fits (No.~9) did not provide a non-zero estimate of the variance parameter for one of the indices and the other two provided identical results and we refit model \(M_0\) and all remaining models using the better parameter estimates as initial values.
-
-\hypertarget{jitter-fits-for-model-m_1}{%
-\subsubsection*{\texorpdfstring{Jitter fits for model \(M_1\)}{Jitter fits for model M\_1}}\label{jitter-fits-for-model-m_1}}
-\addcontentsline{toc}{subsubsection}{Jitter fits for model \(M_1\)}
-
-The 50 jitter fits gave no evidence of a better minimization of the NLL.
-Three lower NLLs were obtained, but with unacceptably large gradients (Figure \ref{fig:jitter-M1}).
-The largest differences in parameter estimates for these three jitters were for numbers at age and selectivity random effects variance and correlation parameters.
-
-\hypertarget{self-test-for-model-m_1}{%
-\subsubsection*{\texorpdfstring{Self test for model \(M_1\)}{Self test for model M\_1}}\label{self-test-for-model-m_1}}
-\addcontentsline{toc}{subsubsection}{Self test for model \(M_1\)}
-
-Initial fits to simulated data from model \(M_1\) showed estimation of the observation error standard deviation multiplier for the recreational catch-per angler indices in the north and south regions was unstable.
-Many of the fits to the simulated data produced implausible estimates at the 0 boundary for these parameters (very negative values on log-scale).
-However, across all fits including those with poor convergence, estimation of SSB and fishing mortality was reliable (Figure \ref{fig:self-test-fig}).
-We also fit analogous models with the multiplier parameters fixed at the true values, which did improve convergence, but larger bias was estimated for fishing mortality and SSB for the northern component.
-
-\setcounter{table}{0}
-\renewcommand\thetable{S\arabic{table}}
-
-\begin{landscape}\begin{table}
-
-\caption{\label{tab:age-comp-sel-table}Configuration of age composition likelihoods, mean selectivity models, and selectivity random effects models for each age composition data component.
-For all logistic-normal likelihoods, any ages observed as zeros are treated as missing.}
-\centering
-\begin{tabular}[t]{llll}
-\toprule
-Data component & Age Composition Likelihood & Mean Selectivity model & Random effects Model\\
-\midrule
-North commercial fleet & Dirichlet-Multinomial & age-specific (ages > 3 fully selected) & AR1 correlation by age and year\\
-North recreational fleet & Logistic-normal (Independent) & age-specific (ages > 6 fully selected) & AR1 correlation by age and year\\
-South commercial fleet & Logistic-normal (AR1 correlation) & logistic & None\\
-South recreational fleet & Logistic-normal (AR1 correlation) & logistic & None\\
-North recreational CPA index & Logistic-normal (Independent) & age-specific (ages > 1 fully selected) & AR1 correlation by year\\
-\addlinespace
-North VAST index & Dirichlet-Multinomial & age-specific (ages > 4 fully selected) & AR1 correlation by age and year\\
-South recreational CPA index & Logistic-normal (AR1 correlation) & age-specific (ages > 2 fully selected) & None\\
-South VAST index & Logistic-normal (AR1 correlation) & age-specific (ages > 1 fully selected) & None\\
-\bottomrule
-\end{tabular}
-\end{table}
-\end{landscape}
-
-\begin{table}
-
-\caption{\label{tab:aic-wts-table}Model AIC weights for each retrospective peel.}
-\centering
-\begin{tabular}[t]{lrrrrrrrr}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Peel} \\
-\cmidrule(l{3pt}r{3pt}){2-9}
-Model & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
-\midrule
-$M_{0}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.01 & 0.01\\
-$M_{1}$ & 0.45 & 0.43 & 0.42 & 0.38 & 0.41 & 0.37 & 0.36 & 0.38\\
-$M_{2}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{3}$ & 0.30 & 0.31 & 0.25 & 0.23 & 0.24 & 0.23 & 0.25 & 0.23\\
-$M_{4}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-\addlinespace
-$M_{5}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{6}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{7}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{8}$ & 0.15 & 0.15 & 0.20 & 0.24 & 0.21 & 0.24 & 0.22 & 0.23\\
-$M_{9}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-\addlinespace
-$M_{10}$ & 0.10 & 0.11 & 0.12 & 0.14 & 0.13 & 0.15 & 0.15 & 0.14\\
-$M_{11}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{12}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{13}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\setcounter{figure}{0}
-\renewcommand\thefigure{S\arabic{figure}}
-
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{fit_0_jitter_plt}
-
-}
-
-\caption{Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_0$.
-Black jitters had maximum absolute gradient values < $10^{-2}$ and red jitters had values > 1.}\label{fig:jitter-M0}
-\end{figure}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{fit_1_jitter_plt}
-
-}
-
-\caption{Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_1$.
-Fits with black dots had maximum absolute gradient value < 0.01 and fits with red dots had values > 10.}\label{fig:jitter-M1}
-\end{figure}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{self_test_results}
-
-}
-
-\caption{Median relative error of SSB (Total and by stock component) and total fully-selected fishing mortality for estimation models fitted to simulated data from model $M_1$ where the observation variance of log-indices are fixed and estimated.
-Black and Red dashed lines represent the median of the annual medians and the median across all annual relative errors, respectively.
-Vertical lines represent 95\% confidence intervals.}\label{fig:self-test-fig}
-\end{figure}
-
-\begin{landscape}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{BTA_full_fig}
-
-}
-
-\caption{Observations with 95\% confidence intervals (points with vertical lines) and posterior estimates with 95\% confidence intervals (lines with polygons) of bottom temperature anomalies in the north and south regions from model $M_1$.
-Gray vertical line defines the first year that the black sea bass stock is modeled.}\label{fig:bottom-temperature}
-\end{figure}
-
-\end{landscape}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{best_R_Ecov}
-
-}
-
-\caption{Expected and random effect recruitment estimates for the northern stock component. Color of points defined by the corresponding annual bottom temperature anomaly.}\label{fig:BT-Ecov-R}
-\end{figure}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{SSB_F_R_rel_M1}
-
-}
-
-\caption{Estimates of SSB, F, and recruitment relative to those of the best performing model, $M_1$.}\label{fig:SSB-F-R-rel-M1}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{Ecov_M1_rel_M0}
-
-}
-
-\caption{Relative differences in posterior estimates of northern region bottom temperature anomalies ($\widehat X$) from the null model without effects on recruitment ($M_0$) and with effects on the northern stock component ($M_1$).}\label{fig:Ecov-M1-rel-M0}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{selectivity_south_plot}
-
-}
-
-\caption{Selectivty for fleets and indices in the southern region.}\label{fig:selectivity-south}
-\end{figure}
-\pagebreak
-
-\begin{landscape}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.9\textheight]{R_SSB_F_cv_results}
-
-}
-
-\caption{Coefficients of variation for estimates of alternative recruitment estimates (random effects or expected), average fishing mortality at age 6 and 7, and SSB by region and in total from model $M_1$.
-Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data.}\label{fig:R-F-SSB-CVs}
-\end{figure}
-\end{landscape}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_F40_CV}
-
-}
-
-\caption{Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data.}\label{fig:annual-F40-cvs}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_SSB40_CV}
-
-}
-
-\caption{Coefficients of variation for annual equilibrium SSB$_{40\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations and alternative annual recruitment types.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data.}\label{fig:annual-SSB40-cvs}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_F_status_results}
-
-}
-
-\caption{Annual estimates of ratios of fishing mortality to $F_{40\%}$ by region and in total.
-Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted lines indicate the last year of data and polygons represent 95\% confidence intervals.}\label{fig:F-status-proj}
-\end{figure}
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_SSB_status_results}
-
-}
-
-\caption{Annual estimates of ratios of SSB to SSB$_{40\%}$ by region and in total.
-Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted lines indicate the last year of data and polygons represent 95\% confidence intervals.}\label{fig:SSB-status-proj}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_F_status_CV}
-
-}
-
-\caption{Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\bar{F}_{40\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data.}\label{fig:annual-F-status-cvs}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_SSB_status_CV}
-
-}
-
-\caption{Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data.}\label{fig:annual-SSB-status-cvs}
-\end{figure}
-
-\begin{landscape}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{SSB_F_sensitivity_plots}
-
-}
-
-\caption{Estimates of annual SSB and fishing mortality rates for the best performing model $M_1$ and models that are otherwise the same except where  1) movement rates are fixed at the means for the prior distribution, 2) a constant natural mortality rate is estimated, or 3) there is no movement for either stock component.}\label{fig:sensitivity-plots}
-\end{figure}
-\end{landscape}
-
 \end{document}

### 2025-07-01 paper/supp_final.Rmd
final

--- 2025-07-01 paper/paper_R2.Rmd
@@
 ---
-title: "Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass"
+title: |
+  |  Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass:
+  |  Supplementary Materials
 author: 
-  - Timothy J. Miller^1^
-  - Kiersten L. Curti
-  - Alexander C. Hansell
-date: "`r format(Sys.time(), '%d %B, %Y')`"
+  - Timothy J. Miller^1^^,2^
+  - Kiersten L. Curti^1^^,3^
+  - Alexander C. Hansell^1^^,4^
 output:
@@
     includes:
-      in_header: options.sty
-#    citation_package: biblatex
-#biblio-style: authoryear
-#  pdf_document:
-#    keep_tex: yes
-#    fig_caption: yes
-#    number_sections: yes
-#    includes:
-#      in_header: options.sty
-#  html_document:
-#    df_print: paged
+      in_header: options_final.sty
 csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
@@
 ```{r setup, include=FALSE, echo=FALSE}
+#knit the supplementary file first, then the main file so that .aux file is available for cross-referencing (using the latex xr package)
 knitr::opts_chunk$set(echo = TRUE, tidy=FALSE, tidy.opts=list(width.cutoff=60), warning = FALSE, message = FALSE, fig.align = "center")
 #knitr::opts_knit$set(root.dir = '../')
+options(tinytex.clean = FALSE) #need .aux files for cross-referencing between main and supp documents, 
 library(knitr)
@@
 library(kableExtra)
-#library(wham)
-#library(ggplotFL)
-#library(viridis)
 type <- "latex"
@@
 
-$^1$timothy.j.miller@noaa.gov, Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\
+$1$ Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\
+$2$ orcid: 0000-0003-1411-1206, timothy.j.miller@noaa.gov\
+$3$ orcid: 0009-0009-1676-7269\
+$4$ orcid: 0000-0001-7827-1749\
 
 \pagebreak
-
-**keywords:** stock assessment, state-space, multi-stock, multi-region, age-structured, black sea bass
-
-## Abstract {-}
-
-The Woods Hole Assessment Model is a general state-space age-structured assessment model that is used to assess and manage many stocks in the Northeast US.
-We first describe an extension of the model allowing any number of stocks (or stock components) and regions with movement among regions as well as seasonal variation in stock and fleet dynamics.
-Movement rates can be functions of time- and age-varying random effects and environmental covariates.
-We then illustrate the model by applying it to data for the northern and southern components of the Northeast United States black sea bass stock and evaluate alternative hypotheses of bottom temperature and time-varying random effects on recruitment and natural mortality.
-We show strong evidence for temperature effects on recruitment, primarily for the northern stock component, and no evidence for including random effects or temperature effects on age 1 natural mortality.
-
-\pagebreak
-
-# Introduction {-}
-
-State-space age-structured stock assessment models can be used to estimate time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures [@nielsenberg14;@cadigan16;@milleretal16].
-This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks [@puntetal20].
-The State-space Assessment Model [SAM, @nielsenberg14] continues to be developed and remains widely used within  the International Council for the Exploration of the Sea (ICES) to assess European fish stocks.
-Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada [@varkeyetal22;@perreaultetal20] and to the south, the Woods Hole Assessment Model [WHAM, @stockmiller21] is now used to assess many fish stocks in the Northwest Atlantic Ocean [@nefsc22;@nefsc22a;@nefsc24].
-
-WHAM is an R package developed and maintained by scientists at NOAA's Northeast Fisheries Science Center (https://timjmiller.github.io/wham).
-WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects, to models with several time and age varying process errors and effects of environmental covariates on various population parameters.
-Like SAM, WHAM models are built using the Template Model Builder package [TMB, @kristensenetal16] which provides a computationally efficient means of fitting an extremely wide class of models with random effects.
-WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as "apparent survival"), fishery and index selectivity, natural mortality, and catchability.
-
-However, WHAM has only allowed consideration of a single stock and region, and no seasonal changes in stock dynamics.
-Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice [@bosleyetal22;@yingetal11;@caoetal14].
-Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks [@puntetal20].
-We describe here the implementation of these features and other extensions since @stockmiller21 in WHAM version 2.0.
-Many of these new configuration options can also be useful when modeling a single stock and region.
-
-We developed this extension of WHAM in concert with a research track assessment process for black sea bass [BSB\, \textit{Centropristis striata}\; @nefsc23], which is a high profile stock in the Northeast U.S.
-and important to both commercial fishing fleets and recreational anglers.
-BSB are targeted from Maine to North Carolina in both state and federal waters.
-Past stock assessments have estimated that the stock is healthy (not overfished and overfishing not occurring).
-Those assessments assumed two stock components (north and south), each assessed with a separate statistical catch at age model, with results subsequently combined to inform management.
-However, the assessments exhibited strong retrospective patterns.
-The stock was split into two regional components to account for spatial dynamics and improve model diagnostics [@asmfc16].
-Additionally, BSB have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall [@mosershepherd09].
-BSB distribution has been linked to warming waters on the northwest shelf [@belletal15] and it is hypothesized that the distribution and productivity of the BSB stock is especially susceptible to long-term changes in temperature causing increases in abundance further north [@hareetal16].
-Thus, a focus of the BSB research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement, and explore environmental drivers of population dynamics [@nefsc23].
-We denote the fish in the north as separate stock components rather than stocks because both components are managed as a single stock, but recognizing differences in migration between fish originating in the two areas.
-
-To illustrate the usage of WHAM 2.0 we apply it to the two stock components of BSB off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process [@nefsc23].
-However, here we evaluate evidence for alternative hypotheses of temporal variation and effects of bottom temperature, specifically, on recruitment and natural mortality of age 1 individuals.
-
-When modeling multiple stock components and regions, further dimensionality is added to the class of possible equilbrium reference points [e.g., @kapuretal21].
-Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises.
-Finally, the inclusion of explicit effects of environmental covariates in the model also requires assumptions on how they are treated in projections and reference points.
-We demonstrate some of the options of WHAM version 2.0 to provide biological reference points and short-term projections.
-
-# Methods {-}
-
-## WHAM description {-}
-
-Many of the options and equations of WHAM version 2.0 are the same as those in @stockmiller21, so we will only describe extensions and differences that have occurred since their first description of WHAM.
-The new version of WHAM can model multiple stocks, each with their own movement, harvest, and natural mortality.
-Much of the description below is for a specific stock $s$, but, for simplicity, this subscript is implicit except when necessary.
-
-### The probability transition matrix {-}
-
-Because individuals may be alive in one of several regions or harvested by one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model.
-Approaches to modeling transitions among these states may treat time discretely [e.g.,@arnason72;@schwarzetal93] or continuously [e.g., @hearnetal87;@commenges99; @andersenkeiding02].
-We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval with duration $\delta$ (columns), conditional on being in each of those states at the beginning of the interval (rows).
-For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
-The time interval $i$ with duration $\delta_i$ would be a season and the PTMs would be uniquely defined for each stock by year $y$, age $a$, and season $i$.
-Each row and column of the PTM correspond to one of the states: alive in region $r$, dead in fleet $f$, or dead from natural causes.
-The probabilities in each row sum to one (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval.
-Given $n_R$ regions and $n_F$ fleets, the square PTM ($n_R + n_F + 1$ rows and columns) as a function of sub-matrices is
-\begin{equation}\label{eq:ptm}
-  \mathbf{P}_{y,a,i} = \begin{bmatrix}
-    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
-    0 & \mathbf{I}_{H} & 0\\
-    0 & 0 & 1
-  \end{bmatrix},
-\end{equation}
-where 
-\begin{equation*}
-  \mathbf{O}_{y,a,i} = 
-  \begin{bmatrix}
-    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
-    \vdots & \ddots & \vdots \\
-    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
-  \end{bmatrix}
-\end{equation*}
-is the $n_R \times n_R$ matrix defining survival and occurring in each region at the end of the interval,
-\begin{equation*} 
-  \mathbf{H}_{y,a,i} = 
-  \begin{bmatrix}
-    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
-    \vdots & \ddots & \vdots \\
-    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
-  \end{bmatrix}
-\end{equation*}
-is the $n_R \times n_F$ matrix defining probabilities of being captured in each fleet during the interval, and $\mathbf{D}_{y,a,i}$ is the $n_R \times 1$ matrix of probabilities of dying due to natural mortality during the interval.
-We have the $n_F\times n_F$ identity matrix $\mathbf{I}_{H}$ for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is one (no zombies allowed).
-
-WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would have a multinomial distribution as in a model for tagging data where fates of individual tagged fish are assumed independent.
-The PTMs determine the expected numbers 1) in each state on January 1 of year $y+1$ at age $a+1$ given the abundances at age $a$ on January 1 of year $y$, 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.
-
-### Single region PTMs {-}
-
-When there is only one region,
-\begin{equation}\label{eq:ptm_1_region}
-\mathbf{P}_{y,a,i} = 
-  \begin{bmatrix}
-     S_{y,a,i} & \mathbf{H}_{y,a,i} & D_{y,a,i} \\
-     0 & \mathbf{I}_{H} & 0\\
-     0 & 0 & 1
-  \end{bmatrix},
-\end{equation}
-where $S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}$, $\mathbf{H}_{y,a,i}$ is a 1 x $n_F$ matrix with elements for each fleet $f$: $\frac{F_{f,y,a,i}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)$, $D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)$, and $Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{f,y,a,i}$ is the total mortality rate.
-
-### Multi-region PTMs {-}
-
-When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously.
-The sequential assumption is used widely in spatially explicit models [e.g., Stock Synthesis, @methotwetzel13].
-Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval.
-WHAM is configured to have movement occur after survival:
-\begin{equation*}
-  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i},
-\end{equation*}
-where $\mathbf{S}_{y,a,i}$ is a $n_R \times n_R$ diagonal matrix of proportions surviving in each region (given they start in that region) 
-\begin{equation*}
-\mathbf{S}_{y,a,i} = 
-  \begin{bmatrix}
-    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
-    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
-    \vdots & \vdots & \ddots & \vdots \\
-    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
-  \end{bmatrix}
-\end{equation*}
-and $\boldsymbol{\mu}_{y,a,i}$ is a $n_R \times n_R$ matrix of proportions moving from one region to another or staying in the region they occurred at the beginning of the interval: 
-\begin{equation*}
-\boldsymbol{\mu}_{y,a,i} = 
-  \begin{bmatrix}
-    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
-    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
-    \vdots & \vdots & \ddots & \vdots \\
-    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
-  \end{bmatrix}
-\end{equation*}
-
-WHAM assumes each fleet $f$ can harvest in only one region ($r_f$) during specified seasons.
-So, for each fleet $f$, row $r_f$ and column $f$ of $\mathbf{H}_{y,a,i}$ will be $F_{f,y,a,i}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$ when fleet $f$ is harvesting during the interval $i$ and all other elements will be zero.
-Row $r$ of the single-column matrix $\mathbf{D}_{y,a,i}$ is  $M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}$.
-For details defining PTMs with movement and mortality occurring simultaneously see Appendix B.
-
-### Seasonality {-}
-
-Seasonality can be configured to accommodate characteristics of spawning, movement, and fleet-specific behavior.
-The annual time step can be divided into $K$ seasons and the interval duration $\delta_i$ for each season $i$ does not need to be equal to any other seasonal interval.
-Under the Markov assumption the PTMs for each consecutive interval are independent and the PTM of surviving, moving, and dying over $K$ intervals $1,\ldots, K$ (i.e., the entire year) is just the product of the PTMs for each interval:
-$$ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).$$
-
-For a stock spawning at some fraction of the year $0<t_s<1$ in interval $i \in \{1,\ldots,K\}$, the fraction of time in the interval is 
-$$\delta_{s,i} = t_s-\sum^{i-1}_{j=0}\delta_j$$
-(where $\delta_0 = 0$), and the PTM defining the proportions in each state at time $t_s$ for age $a$ is
-\begin{equation}\label{eq:ptm_spawn}
-\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{s,i}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,j}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{s,i}),
-\end{equation}
-where $\mathbf{P}_{y,a,0}$ is an identity matrix. Similarly, for an index $m$ occurring  at fraction of the year $t_m$ in interval $i$ the proportions in each state at the time of the observation is
-\begin{equation} \label{eq:ptm_index} 
-\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{m,i}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,i}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{m,i}).
-\end{equation}
-
-### Numbers at age {-}
-
-When there are $n_R$ regions and $n_F$ fleets, the vector of abundance for a given stock (component) in each state at age $a>1$ on January 1 is $\mathbf{N}_{y,a}^T = (\mathbf{N}_{O,y,a}^T, \mathbf{0}^T)$, where $\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})^T$ is the number in the states corresponding to being alive in each region and $\mathbf{0}$ is a vector ($n_F+1$) for the numbers captured in each fleet or dead from natural mortality because no age $a$ fish have died yet on January 1.
-
-Each stock $s$ is assumed to spawn and recruit in one region $r_s$.
-So for age $a=1$, $\mathbf{N}_{O,y,1}$ is 0 except for row $r_s$.
-Options for configuring recruitment ($N_{y,1,r_s}$) for each stock are the same as previous versions of WHAM.
-If recruitment is assumed to be a function of SSB, it is only the spawning population in region $r_s$ at the time of spawning that constitutes the SSB in the stock-recruit function.
-However, models can configure individuals to occur in other regions at the time of spawning.
-Aside from treating recruitment as a random walk, the general model for annual recruitment as random effects is
-\begin{equation*}
-\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1} =  f\left(\text{SSB}_{y-1}\right) + \varepsilon_{y,1,r_s},
-\end{equation*}
-where  
-$$\text{SSB}_{y} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,\cdot,r_s}^T(t_s) \mathbf{N}_{O,y,a}$$
-and $w_{y,a}$ and $\text{mat}_{y,a}$ are the stock-specific mean weight at age of spawning individuals and maturity at age, respectively, and $\mathbf{O}_{y,a,\cdot,r_s}(t_s)$ are the probabilities of surviving and occurring in region $r_s$ at time $t_s$ given being alive in each region at the start of the year (column $r_s$ of the upper-left submatrix of Eq.
-\ref{eq:ptm_spawn}).
-
-As in previous versions of WHAM, the apparent survival (and movement) from one year to another after recruitment can be treated deterministically or as functions of random effects.
-The predicted numbers at age $a$ in year $y$ for a given stock are vector analogs ($\mathbf{N}_{O,y,a}$) of the equations for numbers at age in the standard WHAM model [@stockmiller21].
-For ages $a = 2,\ldots, A-1$, where $A$ is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first $n_R$ elements of the vector
-$$\mathbf{P}_{y-1,a-1}^T \mathbf{N}_{y-1,a-1}.$$
-The numbers alive in each region can also be modeled more simply using the sub-matrix $\mathbf{O}_{y,a}$.
-The general model for the vector of abundance at age in year $y$ at age $a$ given the vector in year $y-1$ at age $a-1$ is
-\begin{equation*}
-\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}^T \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
-\end{equation*}
-for ages $a = 2,\ldots, A-1$, and for the plus group
-\begin{equation*}
-\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,A-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}^T \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}^T \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A},
-\end{equation*}
-where $\boldsymbol{\varepsilon}_{y,a}$ is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
-
-When there is autocorrelation with age, WHAM now assumes this applies only to ages $a>1$ by default so that recruitment random effects are independent of those for the apparent survival of older age classes.
-The general covariance structure for a given stock at ages $a>1$ in region $r$ is a two-dimensional first-order autoregressive (2DAR1) process
-\begin{equation*}
-  Cov\left(\varepsilon_{y,a,r},\varepsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}
-\end{equation*}
-and that for age 1 is just AR1 across years 
-\begin{equation*}
-  Cov\left(\varepsilon_{y,1},\varepsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
-\end{equation*}
-If the annual changes in abundance at age are treated deterministically, then $\boldsymbol{\varepsilon}_{y,a} = \mathbf{0}$.
-Since recruitment for a given stock currently only occurs in one region $r_s$ there is only a single vector of time-varying recruitment random effects for each stock.
-
-### Initial numbers at age {-}
-
-Initial numbers at age for each stock and region can be treated as age-specific fixed effects or with an equilibrium assumption as in previous versions of WHAM.
-For the equilibrium option there are two parameters for each stock: the stock-specific equilibrium fully-selected fishing mortality rate ($\widetilde{F}_1$) and the recruitment in year 1 ($N_{1,1,r_s}$).
-A stock-specific equilibrium fishing mortality at age by fleet ($\widetilde{F}_{f,1,a}$) is the product of $\widetilde{F}_1$ and the selectivity across fleets in the first year
-\begin{equation*}
-  \text{sel}_{f,1,a} = \frac{F_{f,1,a}}{\max_a \sum_{f=1}^{n_F} F_{f,1,a}}.
-\end{equation*}
-We use $\widetilde{F}_{f,1,a}$ to define an equilibrium abundance per recruit by region at age $a$ conditional on recruiting to each region
-\begin{equation}\label{eq:npraa}
- \widetilde{\mathbf{O}}_{a} = \left\{
- \begin{array}{ll}
-\prod^{a-1}_{j=0}\mathbf{O}_{j}  & 1\leq a<A\\
-\left[\prod^{a-1}_{j=0}\mathbf{O}_{j}\right] \left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1} & a = A
- \end{array},
-\right.
-\end{equation}
-where $\mathbf{O}_{j}$ is the equilibrium proportions surviving age $a$ and occurring in each region and $\mathbf{O}_{0} = \mathbf{I}$, the identity matrix.
-Natural mortality and movement rates in the first year of the model are also used in Eq.
-\ref{eq:npraa}.
-For the plus group $a=A$, $\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}$ is a ``fundamental matrix'' derived using the matrix version of the  geometric series [@kemenysnell60].
-Recall that recruitment for a given stock $s$ only occurs in region $r_s$ so, the equilibrium initial numbers at age $a$ by region are
-$$\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a,r_s,\cdot} N_{1,1,r_s},$$
-where $\widetilde{\mathbf{O}}_{a,r_s,\cdot}$ is the vector of values in row $r_s$ of $\widetilde{\mathbf{O}}_{a}$.
-Initial abundance at age can also be estimated as random effects as described in Appendix B.
-
-### Parametizing movement {-}
-
-For each season, there are at most $n_R-1$ parameters determining movement for a given stock among regions given starting the season in region $r$ in either the sequential or simultaneous configurations.
-Movement parameters are estimated on a transformed scale via a link function $g(\cdot)$.
-If survival and movement occur simultaneously, the parameters are estimated with a log link function  and if they are separable, an additive logit link function (like a multinomial regression) is used.
-On the transformed scale, the general model for the movement parameter from region $r$ to $r'$ in season $i$ and year $y$ for individuals of age $a$ is a linear function of an intercept or mean parameter ($\theta_{r\rightarrow r',i}$) and both random and covariate effects:
-\begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} X_{k,y}.
-\end{equation*}
-The random effects $\varepsilon_{r\rightarrow r',y,a,i}$ are region-to-region- and season-specific, and modeled most generally as 2DAR1 random effects with age and(or) year, where the covariance is
-\begin{equation*}
-  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}.
-\end{equation*}
-This is similar to how WHAM models variation in survival, natural mortality, and selectivity.
-Covariate $X_k$ effects ($\beta_{r\rightarrow r',a,i,k}$) can be region-to-region-, age-, and season-specific, and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.
-
-There is currently no likelihood component for tagging data.
-Therefore, movement parameters would usually either need to be fixed or assumed to have some prior distribution, possibly based on external parameter estimates.
-WHAM includes prior distributions for the region-to-region- and season-specific (mean) movement parameters and treats them as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
-  \begin{equation*}
-  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{\gamma,r\rightarrow r',i}\right).
-  \end{equation*}
-When priors are used, the movement is defined instead as
-  \begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} X_{k,y}.
-  \end{equation*}
-
-### Natural mortality {-}
-
-When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific, but currently are constant across seasons.
-When random effects are used, the same 2DAR1 structure with age and year as described by @stockmiller21 can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
-We provide details in Appendix B.
-
-### Catch observations {-}
-
-The log-normal distributional assumption for aggregate catch observations is the same as @stockmiller21, but the predicted catch is now a function of catch from each stock starting the year in each region and any seasonal movement.
-For a given stock $s$ at age $a$, the vector of numbers captured by each fleet over the year $y$ is
-$$\mathbf{N}_{H,s,y,a} = \mathbf{H}_{s,y,a}^T \mathbf{N}_{O,s,y,a}$$
-(recall each fleet can harvest in only one region as described above). The vector of numbers caught by each fleet across stocks is 
-$$\mathbf{N}_{H,y,a} = \sum^{n_S}_{s=1} \mathbf{N}_{H,s,y,a}$$
-and the vector of predicted aggregate catch by each fleet at age $a$ is
-$$\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \mathbf{N}_{H,y,a},$$
-where $\mathbf{c}_{y,a}$ is the vector of mean individual weight at age $a$ for each fleet and the aggregate catch by fleet is 
-$$\widehat{\mathbf{C}}_y = \sum^{A}_{a=1} \widehat{\mathbf{C}}_{y,a}.$$
-The log-aggregate catch observations for fleet $f$ are normally distributed
-$$ \log C_{f,y} \sim \text{N}\left(\log \widehat {C}_{f,y}, \sigma^2_{f,y}\right),$$
-
-where $\widehat C_{f,y}$ is an element of $\widehat{\mathbf{C}}_y$.
-The predicted numbers caught for each fleet $f$ (row $f$ of $\mathbf{N}_{H,y,a}$) are used to make predicted age composition observations as described by @stockmiller21.
-Since then, three additional likelihood options for age composition observations have been added: a logistic-normal with AR(1) correlation structure [@francis14], the alternative Dirichlet-multinomial parameterization described by @thorsonetal17, and the multivariate Tweedie [@thorsonetal23].
-
-### Index observations {-}
-
-An index $m$ can observe any stocks that occur in a single region $r_m$ at fraction of the year $t_m$. The abundance at $t_m$ in region $r_m$ is 
-$$N_{m,s,y,a} = \mathbf{O}_{s,y,a,\cdot,r_m}^T(t_m) \mathbf{N}_{O,s,y,a},$$
-where $\mathbf{O}_{s,y,a,\cdot,r_m}(t_m)$, column $r_m$ of the upper-left submatrix of Eq.
-\ref{eq:ptm_index}, is the vector of proportions surviving and occurring in region $r_m$ at time $t_m$ given being alive in each region at the start of the year.
-The predicted index at age is
-$$\widehat{I}_{m,y,a} = q_{m,y} \text{sel}_{m,y,a}w_{m,y,a}\sum^{n_S}_{s = 1}N_{m,s,y,a},$$
-where $q_{m,y}$ is the catchability of the index in year $y$, $\text{sel}_{m,y,a}$ is the selectivity, and $w_{m,y,a}$ is the average weight of individuals at age $a$ if the index is quantified in biomass and $w_{m,y,a} = 1$ if the index is quantified in numbers.
-Predicted age composition observations are functions of $\widehat{I}_{m,y,a}$ as described by @stockmiller21 and the likelihood options are the same as those for catch explained above.
-We describe options to model temporal variation in catchability in Appendix B.
-
-### Reference points {-}
-
-WHAM calculates annual reference points using biological inputs each year and it also provides the same reference points under prevailing conditions where biological inputs are averaged over a user-defined set of recent years.
-A single fishing mortality reference point $\widetilde{F}$ is estimated across stocks and regions and the fleet- and age-specific reference point is $\widetilde{F}_{f,a} = \widetilde{F} \text{sel}_{f,a}$.
-Selectivity is determined across fleets and ages as in the equilibrium assumption for initial abundance at age except here it is a function of fishing mortality at age, possibly averaged over years,
-\begin{equation}\label{eq:spr_sel}
-  \text{sel}_{f,a} = \frac{\overline{F}_{f,a}}{\max_a \sum^{n_F}_{f=1}{\overline{F}}_{f,a}}.
-\end{equation}
-
-The equilibrium spawning stock biomass-per-recruit for a given stock in region $r_s$ is defined as
-\begin{equation}\label{eq:ssbpr}
- \upphi(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{O}_{a,\cdot,r_s}(t_s) w_{a} m_{a},
-\end{equation}
-where $w_{a}$ and $m_{a}$ are the mean individual weight and proportion mature at age $a$, $\widetilde{\mathbf{O}}_{a,r_s,\cdot}$ is row $r_s$ of $\widetilde{\mathbf{O}}_a$ in Eq.
-\ref{eq:npraa}, and $\mathbf{O}_{a,\cdot,r_s}(t_s)$ is column $r_s$ of the matrix of proportions surviving and occurring in each region at age $a+t_s$ (the upper-left sub-matrix of eq.
-\ref{eq:ptm_spawn}).
-Using these rows and columns is required because of the assumption that spawning and recruitment only occur in region $r_s$.
-For prevailing conditions, the years to average for $w_a$, $m_a$, and natural mortality and movement rates to parameterize $\mathbf{O}_{a}(t_s)$ and $\widetilde{\mathbf{O}}_{a}$, are defined by the user.
-
-The equilibrium yield-per-recruit as a function of $\widetilde{F}$ for a given stock $s$ is calculated as
-\begin{equation}\label{eq:ypr}
- \upnu({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{H}_{a} \mathbf{c}_{a},
-\end{equation}
-where $\mathbf{c}_{a}$ is the vector of mean individual weight at age for each fleet, and $\mathbf{H}_{a}$ is the submatrix of the proportions captured in each fleet over the interval from $a$ to $a+1$, defined in eq.
-\ref{eq:ptm}.
-Other than mean individual weight, $\upnu({\widetilde{F}})$ uses all of the same (averaged) inputs as $\upphi({\widetilde{F}})$ in Eq.
-\ref{eq:ssbpr}.
-
-For reference points based on reductions of $\upphi({\widetilde{F}})$ to $p\%$ of the unfished value, $\upphi(0)$ ($p/100$ is known as the spawning potential ratio; SPR), we use a Newton-Raphson method and iterate for steps $j = 1,\ldots,10$
-\begin{equation}\label{eq:newton}
-  \log\widetilde{F}^{j} = \log\widetilde{F}^{j-1} - \frac{h\left(\log\widetilde{F}^{j-1}\right)}{h^{(1)}\left(\log\widetilde{F}^{j-1}\right)},
-\end{equation}
-where $h(\log \widetilde{F})$ is the sum of weighted differences between $\upphi({\widetilde{F}})$ and $p$\% of $\upphi_s\left(0\right)$ across stocks:
-\begin{equation}\label{eq:newton-obj}
-  h(\log \widetilde{F}) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(e^{\log \widetilde{F}}\right) - \frac{p}{100}\upphi_s\left(0\right)\right],
-\end{equation}
-and $h^{(1)}(\log \widetilde{F})$ is the derivative of $h(\log \widetilde{F})$ with respect to $\log \widetilde{F}$ [analogous to the approach for SPR-based reference points in @millerlegault17].
-The weights for each stock ($\lambda_s$) can be specific values supplied by the user or defined to be the proportions of average recruitment for each stock where the years used to average are those the user defines to calculate prevailing equilibrium spawning biomass and yield reference points, which are the product of the defined average recruitment and $\upphi({\widetilde{F}})$ and $\upnu({\widetilde{F}})$, respectively.
-The Newton-Raphson method is used by WHAM internally to allow uncertainty in all estimated parameters to be propagated into uncertainty estimates for the reference points.
-
-When a Beverton-Holt or Ricker stock-recruit relationship is assumed, we use the derivative (with respect to $\widetilde{F}$) of the weighted sum of stock-specific equilibrium yields (the product of $\upnu({\widetilde{F}})$ and equilibrium recruitment from the stock-recruit relationship) as $h(\log \widetilde{F})$ to find $\widetilde{F}$ which maximizes the weighted total yield across stocks for MSY-based reference points [$F_{\text{MSY}}$, analogous to the approach for MSY-based reference points in @milleretal16].
-
-### Projections {-}
-
-The projection options are generally the same as those for previous versions of WHAM.
-When there is movement of any stocks, the user has the option to project and use any random effects for time-varying movement or use the average over user specified years, analogous to how natural mortality can be treated in the projection period.
-The projection of any environmental covariates has been revised to better include error in the estimated latent covariate in any effects on the population in projection years.
-Users can also specify a catch or fully-selected fishing mortality in each projection year and they can be fleet-specific.
-
-# Application to black sea bass {-}
-
-BSB are a temperate reef fish in the western Atlantic Ocean ranging the entire east coast of the United States.
-Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity [@ablehales97;@fabrizioetal13].
-Fish in this NEUS stock perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring [@musickmercer77].
-Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude [@mosershepherd09].
-
-Prior to its 2023 peer-reviewed assessment, the NEUS BSB stock was assessed using the Age-Structured Assessment Program (ASAP) model [@legaultrestrepo99], a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
-Northern and southern components of the NEUS BSB stock ascribed to regions divided by the Hudson Canyon were separately modeled in ASAP (Figure \ref{fig:map}).
-Results from the separate ASAP models were combined for a unit-stock assessment.
-The ASAP-based assessments exhibited strong retrospective patterns [@mohn99], and exploring alternative modeling approaches for the northern and southern stock components has been a high priority for management.
-
-Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as "Working Group") composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the BSB stock using the multi-stock and multi-region extension of WHAM described above [@nefsc23].
-This assessment included the two regions and two stock components and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous BSB assessments.
-Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on BSB.
-
-## Basic structure {-}
-
-The first year modeled for the population is 1989 and the fishery and index data used in the model span from 1989 to 2021.
-The north and south stock components are modeled as separate populations that spawn and recruit in respective regions and there are eight age classes with the last being a plus group ($a \in \{1,\ldots,8+\}$).
-We have observations for each of four total fishing fleets, where two fishing fleets (recreational and commercial) operate in each region.
-
-There are 11 "seasonal" intervals within each calendar year: five monthly time intervals from Jan 1 to May 31, a spawning season from June 1 to July 31, and five monthly intervals from August 1 to December 31.
-The southern stock component is assumed to never move to the northern region.
-For the northern component, a proportion $\mu_{\text{N}\rightarrow \text{S}}$ can move to the southern region each month during the last five months of the year, but no movement is allowed from the south to the north during this period (Figure \ref{fig:migration-diagram}).
-During the first four intervals of the year a proportion $\mu_{\text{S}\rightarrow \text{N}}$ of the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
-In the fifth interval (May), all northern component individuals remaining in the south are assumed to move back to the north for the subsequent spawning period.
-Survival and movement occur sequentially in each interval and each of the two movement proportions are assumed constant across intervals, ages, and years.
-
-The two monthly movement matrices are
-\begin{equation*}
-\boldsymbol{\mu}_{1} = 
-  \begin{bmatrix}
-     1-\mu_{\text{N}\rightarrow \text{S}} & \mu_{\text{N}\rightarrow \text{S}} \\
-     0 & 1 \\
-  \end{bmatrix}
-\end{equation*}
-for the portion of the year after spawning and
-\begin{equation*}
-\boldsymbol{\mu}_{2} = 
-  \begin{bmatrix}
-     1 &  0 \\
-     \mu_{\text{S}\rightarrow \text{N}} & 1-\mu_{\text{S}\rightarrow \text{N}} \\
-  \end{bmatrix}
-\end{equation*}
-for the portion of the year before spawning.
-As noted in the description of the general WHAM model, tagging data are not yet included.
-However, the Working Group also fit a Stock Synthesis model [@methotwetzel13] which included tagging data and provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).
-
-With movement as configured, the northern origin fish (ages 2+) can occur in the southern region on January 1.
-Estimating initial numbers at age as separate parameters can be challenging even in single-stock models, but for BSB the available data cannot distinguish the proportion of northern and southern component fish at each age in the southern region in the initial year of the model.
-Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component (Eq.
-\ref{eq:npraa}).
-
-For subsequent years, all models have AR1 correlation of recruitment and 2DAR1 correlation for apparent survival random effects for both the northern and southern stock components.
-Models with apparent survival random effects will model the transitions (conditional on both survival and movement) of abundances at age of northern origin fish in each region, but there appears to be little information in the existing data from the southern region to distinguish apparent survival random effects for each stock component.
-All models we considered assume a very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model.
-Unique variance and correlation parameters for the recruitment and apparent survival random effects are estimated for the northern and southern components.
-
-## Bottom temperature effects {-}
-
-All models include observations of winter bottom temperature anomalies for the northern and southern regions from 1959 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius [@nefsc23].
-We provide details about the derivation of these observations in the Supplementary Materials.
-We retained the assumption from the peer-reviewed assessment that treated the latent bottom temperature covariates in each region as AR1 processes.
-
-We fit 14 models with alternative assumptions about the effects of bottom temperature covariates, ranging from no effects to effects on both regions for either recruitment or natural mortality at age 1 (Table \ref{tab:model-desc-table}).
-These analyses derive from the hypothesis that bottom temperature affects survival over the first winter where fish turn from age 0 to age 1 on January 1 [@milleretal16_yoy_survival].
-This covariate may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of the year after becoming age 1.
-Furthermore, we have no direct observations of age 1 individuals from surveys until the spring season each year.
-Therefore, we fit models with effects of temperature on recruitment or natural mortality at age 1.
-
-We assume the covariate in year $y$ affects recruitment in the same year because the covariate observations are from winter months.
-The fish are technically already 1 year old, but there are no observations of these individuals until later in the spring except possibly in fishery catches which are accumulated over the whole year.
-Expected log-recruitment for a given stock component is a linear function of bottom temperature
-\begin{equation}\label{eq:expected-recruitment}
-E\left(\log N_{y,1,r_s}|X_{y}\right) = \theta_{R} + \beta_{R} X_y.
-\end{equation}
-and the annual recruitment is
-\begin{equation}\label{eq:Rec-re}
-\log N_{y,1,r_s} = E\left(\log N_{y,1,r_s}|X_y\right) + \varepsilon_{y,1,r_s},
-\end{equation}
-where $\varepsilon_{y,1,r_s}$ are the normally distributed random effects with AR1 correlation structure.
-
-Similarly, expected log-natural morality rate as a function of bottom temperature is
-\begin{equation}\label{eq:expected-M1}
-E\left(\log M_{y,1,r_s}|X_y\right) = \theta_{M,a,r_s} + \beta_{M} X_y.
-\end{equation}
-The mean log natural mortality rate is assumed to be $\theta_{M,a,r} = \log(0.4)$ for all ages and regions as recommended by the Working Group.
-@milleretal18 showed how inferences of temperature effects on growth or maturity parameters can be very different depending on whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
-Therefore, we also explored whether including temporal random effects on age 1 natural mortality affected inferences on corresponding temperature effects.
-The model for annual natural mortality for a given stock is
-\begin{equation}\label{eq:M-re}
-\log M_{y,1,r_s} = E\left(\log M_{y,1,r_s}|X_y\right) + \varepsilon_{M,y,1,r_s}.
-\end{equation}
-Because the bottom temperature anomalies are centered and $E(\varepsilon_{M,y,1,r_s}) = 0$, mean log-natural mortality at age 1 over the time series should be approximately equal to $\theta_{M,a,r_s}$ for models that include those effects.
-
-Initially, we included random effects on age 1 natural mortality for both the northern and southern stock components, but estimates of these random effects and corresponding variance for the southern component converged to 0 so these were not included in models presented here.
-Because age 1 fish for the northern component can exist in both regions after January 1, other than during the spawning period, natural mortality is acting in both regions for this stock.
-For models with covariate and(or) annual random effects for age 1 fish, we use them only in the northern region for the northern component.
-
-## Uncertainty in recreational index observations {-}
-
-The estimated coefficients of variation (CVs) provided by the analyses by the Working Group to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the Working Group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
-All models we considered allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
-Although lack of model convergence was observed in many fits for self-test simulations, the estimation of critical model output (e.g., SSB, fishing mortality) was reliable and robust to the poor estimation of these scalars (Figure \ref{fig:self-test-fig}), and estimating these parameters rather than fixing them in the fit to the actual data allows uncertainty in model output to be more properly conveyed.
-
-## Index and catch age composition observations {-}
-
-The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual and retrospective patterns.
-These analyses resulted in use of selectivity random effects for the northern fleet and indices, logistic-normal likelihoods for six sets of age composition observations, and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
-
-## Model fitting and diagnostics {-}
-
-We used a development version ([commit 53e236b](https://github.com/timjmiller/wham/tree/53e236b)) of the WHAM package after the release of version 2.0 for all results.
-All code for fitting models and generating results can be found at [github.com/timjmiller/multi_wham_bsb](https://github.com/timjmiller/multi_wham_bsb).
-
-We examined retrospective patterns for all models by fitting corresponding models where the terminal year is reduced sequentially by one year (peel) for seven years.
-Therefore, there are eight fits of each model with the time series reduced by zero to seven years.
-We  calculated Mohn's $\rho$ for SSB, and average fishing mortality rate at ages 6 and 7.
-Absolute values of Mohn's $\rho$ near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended [@mohn99].
-As in @milleretal16, we also assessed the consistency of the AIC-based model selection over retrospective peels to guard against previously noted changes in perception of covariate effects on recruitment with increased length of the time series of observations [@myers98].
-This retrospective examination was also recommended by @brooks24.
-
-We performed a jitter analysis of the base model $M_0$ and the best fitting model to investigate whether a local mimimum of the negative log-likelihood surface was obtained by the optimization.
-We used the \verb|jitter_wham| function in the WHAM package which by default simulates starting values from a multivariate normal distribution with mean and covariance defined by the MLEs and hessian-based covariance matrix from the fitted model (Supplementary Materials).
-
-For the best performing model, we also performed simulation self-tests where new observations were simulated conditional on all estimated fixed and random effects and the same model configuration was fit to each of the simulated data sets.
-We estimated median relative bias of SSB across these simulations.
-Finally, we fitted three sensitivity models which differed from the best performing model, by 1) assuming no movement for either stock component, 2) movement parameters are fixed at the means for the prior distributions, and 3) a constant natural mortality rate is estimated across all regions and stocks (Supplementary Materials).
-
-## Reference points and projections {-}
-
-We produced SPR-based reference point estimates ($F_{40\%}$ and SSB$_{40\%}$, setting $p=40$ in eq.
-\ref{eq:newton-obj}) from the best performing model using a variety of options provided in WHAM.
-We generated annual reference point estimates based on the annual weight at age, maturity at age, natural mortality, fleet selectivity, and either the annual recruitment random effect (Eq.
-\ref{eq:Rec-re}) or the annual expected recruitment (Eq.
-\ref{eq:expected-recruitment}).
-We also calculated status as the ratios of annual average fishing mortality rates and SSB to the corresponding reference points.
-Finally, we calculated "prevailing" reference points using the average weight, maturity, natural mortality rate, and fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
-All reference points and status estimates are generated for each region as well as for the total stock area.
-
-We projected the best BSB model under three alternative scenarios for the bottom temperature covariate where 1) the AR1 time-series model continues into the projection years, 2) the average of the most recent 5 years is projected, and 3) bottom temperature increases in projection years following a prediction from a simple linear regression of the estimated bottom temperature anomalies over time.
-We specified fishing mortality in projection years to be constant at the value in the last model year (2021).
-Average weight, natural mortality, movement, and maturity at age over the last 5 years are used in projection years, whereas the AR1 processes for all numbers at age random effects are continued.
-
-# Results {-}
-
-We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock component and no random effects on age 1 natural mortality ($M_1$, Tables \ref{tab:diff-aic-table} and  \ref{tab:aic-wts-table}).
-Model $M_1$ consistently had the lowest AIC across retrospective peels where the terminal years of data are removed from the model.
-The difference in AIC for the model that also included bottom temperature effects on recruitment for the southern component ($M_3$) suggested some evidence for this hypothesis as well.
-There was little evidence for also including random effects on age 1 natural mortality for the northern stock component.
-However, when more than three terminal years of data were removed, the evidence for adding age 1 natural mortality random effects for the northern component ($M_8$) increased similar to adding bottom temperature effects on southern component recruitment ($M_3$).
-
-Comparisons of estimates from retrospective peels did not show any indication of differences in model performance.
-There was little variation in measures of retrospective patterns for SSB or fishing mortality across models.
-Mohn's $\rho \approx -0.03$ for SSB of both stock components and Mohn's $\rho \approx 0.03$ and $\approx -0.04$ for average $F$ at ages 6 and 7 in north and south regions, respectively (Table \ref{tab:rho-table}).
-
-Estimated positive effects of bottom temperature on recruitment for the northern or southern components were stable over the retrospective peels and differed negligibly whether effects for each component were estimated in isolation or together (Table \ref{tab:beta-sig-peel-table}).
-The estimates for the effects on each stock component were essentially equivalent whether the effect was on one component or both.
-Estimates of residual variability in northern component recruitment, as measured by the conditional or marginal standard deviation of the recruitment random effects, increased slightly with the number of peeled years.
-However, the ratio of standard deviations of models with and without temperature effects was stable with approximately 20% reduction in residual standard deviation when temperature effects were included.
-The residual variation is reduced because the expected recruitment (Eq.
-\ref{eq:expected-recruitment}) is a function of the covariate (Figure \ref{fig:BT-Ecov-R}).
-
-The posterior estimates of the bottom temperature covariate match the observations well because of the high precision of the observed anomalies (Figure \ref{fig:bottom-temperature}).
-The anomalies for the north and south regions appear highly correlated presumably because of the proximity of the two regions.
-Because the temperature anomalies are treated as latent variables, other data components in the model can affect the estimated anomalies when effects on recruitment are included [e.g., @milleretal18], but in this case the estimates of anomalies are altered negligibly during the population model years when the effect on recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
-
-The effect of the anomalies on expected recruitment can be better observed by comparing the time series of the northern and southern components (Figure \ref{fig:E-R-SSB-F}).
-The expected recruitment for the southern component without any covariate effects is constant over time, whereas the effect of the anomalies on expected recruitment for the northern component induces annual variation in expected recruitment.
-The annual variation in expected recruitment (and the temperature anomalies) is correlated with the recruitment random effects of the northern component and therefore reduces the residual variation in the recruitment random effects.
-For the northern stock component, most of the lowest estimated recruitment random effects occurred prior to 2000 and several large recruitment estimates occurred after 2010.
-For the southern stock component, recruitment random effect estimates have been less variable and similar to the estimated median particularly since 2005.
-The estimated size of the northern stock component, as measured by SSB, increased substantially since 2005, whereas that for the southern component peaked in 2002.
-The increase in the northern component is also reflected in the total size of the combined components.
-Estimated average total fishing mortality rates  across ages 6 and 7 (combining the commercial and recreational fleets) were highest in the southern region prior to 1998 and have declined substantially since then.
-Estimated fishing mortality rates in the northern region have been less variable, but had largest peaks in 1996, 1999, and 2005.
-The decline in estimates in the southern component is also reflected in the decline in total fishing mortality across regions.
-
-The estimated movement from the north to the south for the northern stock component is much less than anticipated from the prior distribution that was parameterized based on the companion Stock Synthesis model fit completed by the Working Group (Figure \ref{fig:move-prior-posterior}).
-Conversely, the estimated movement rate from south to north was more consistent and slightly greater than that from the companion model.
-The estimated movement rates varied little across all fitted models implying that the estimation of movement was not sensitive to the alternative assumptions we considered.
-Furthermore, the very low estimated movement rate from north to south is also consistent with preliminary fits of the WHAM model without the prior distribution where the movement rate was estimated at the lower bound of 0 resulting in a lack of convergence.
-The estimated movement rates imply that a very small fraction of northern fish are ever in the southern region.
-
-The 2DAR1 random effects for the selectivity parameters of the northern recreational fleet indicate a decrease in selectivity at ages 2 to 6 over the time series which corresponds to increases in the size limits in regulations by states where these catches occurred (Figure \ref{fig:selectivity-re}).
-The model also estimated a more modest decrease in selectivity at ages 2 to 3 for the northern commercial fleet.
-The selectivity at the youngest ages for the two fishery independent indices varied over time but without notable trend.
-The selectivity of fleets and indices in the south increased with age, but were constant over time (Figure \ref{fig:selectivity-south}).
-
-The three sensitivity fits suggest the estimates of stock size and fishing mortality are fairly robust to whether the movement rates are fixed at the values from the companion Stock Synthesis fit or whether no movement is assumed (Figure \ref{fig:sensitivity-plots}).
-A constant natural mortality rate was estimable, but the estimated rate (0.72) is nearly twice that determined by the Working Group and resulted in generally higher stock sizes and lower fishing mortality rates.
-
-## Reference points {-}
-
-Using the expected recruitment rather than the random effects as the multiplier with $\upphi(\widetilde{F})$ (eq.
-\ref{eq:ssbpr}) results in less variable estimates of annual SSB$_{40\%}$ (Figure \ref{fig:annual-BRPs}: top row).
-Estimates were less variable for the southern stock component than the northern stock component mainly because the expected recruitment is constant for the former, whereas that for the northern component varies via the effect of the bottom temperature anomalies.
-Temporal variation for the southern stock component is mainly caused by annual variation in the inputs to the $\upphi(\widetilde{F})$ calculation and $F_{40\%}$, but there is also some effect of the variation in the proportion of total recruitment for each stock component as used in the weighting terms to define the 40\% reduction in $\upphi(0)$ ($\lambda_s$ in Eq.
-\ref{eq:newton-obj}).
-Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB$_{40\%}$ for the southern stock component, whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
-
-Ratios of SSB to SSB$_{40\%}$ should be viewed with caution because the SSB reference point is being defined using annual recruitments rather than some form of average recruitment and therefore, ratios larger than 2.5 which would correspond to unfished population sizes are possible (Figure \ref{fig:annual-BRPs}: second row).
-However, for the southern component, using the constant expected recruitment is comparable to using median recruitment and the annual variation in the inputs to $\upphi(\widetilde{F})$ indicates no years where SSB was greater than unfished levels throughout the time series.
-The patterns in uncertainty of SSB ratios are the same as those for the SSB reference points (Figure \ref{fig:annual-SSB-status-cvs}).
-
-Differences between annual $F_{40\%}$ estimates using the alternative types of recruitment estimates are much less than those for SSB$_{40\%}$ because the recruitment estimates are only used in the weighting of the stock component-specific $\upphi(\widetilde{F})$ to determine $F_{40\%}$ ($\lambda_s$ in Eq.
-\ref{eq:newton-obj}, Figure \ref{fig:annual-BRPs}: third row).
-For similar reasons, estimates of uncertainty of $F_{40\%}$ are also similar between the type of recruitment estimates (Figure \ref{fig:annual-F40-cvs}).
-Annual estimates for the northern stock component were lowest in the earliest years of the time series, whereas those for the southern component are relatively stable over time.
-
-For a given type of recruitment used to calculate $F_{40\%}$, the annual ratios of average $F$ to the corresponding reference point are identical for each stock component and in total because of the selectivity used to define a single F reference point across both regions is exactly the same as that for the annual estimate of fishing mortality at age by fleet (Figure \ref{fig:annual-BRPs}: fourth row).
-As for annual $F_{40\%}$, estimates of uncertainty in the ratios are similar whether expected recruitment or the random effects are used to calculate the reference points (Figure \ref{fig:annual-F-status-cvs}).
-
-Unlike the annual ratios of $F$ to $F_{40\%}$, the $F_{40\%}$ under prevailing conditions uses a selectivity defined by the average $F$ at age and fleet over the last 5 years, which is different from that of the selectivity in the terminal year and therefore those ratios for each region and in total are different (Figure \ref{fig:annual-BRPs}: bottom row).
-Although the alternative types of recruitment estimates result in different points on the Kobe plots for joint status, the uncertainty in the estimates suggest similar probabilities of alternative quadrants.
-However, the usage of the confidence regions of the ratios to define probabilities is an approximation because these ratios are functions of both empirical Bayes estimates of random effects as well as maximum likelihood estimates of fixed effects.
-
-## Projections {-}
-
-When continuing the AR1 model for the bottom temperature anomalies in the projection years, the uncertainty grows and asymptotes to a level determined by the marginal variance of the process and the predicted value approaches the estimated marginal mean value of approximately 0 (Figure \ref{fig:R-BT-proj}).
-There is no uncertainty in the projected linear trend because the values were specified as known, whereas there is a small level of uncertainty in the projections of the recent average based on the uncertainty in the estimated mean.
-Comparing projected recruitment for the northern and southern stock components, the projected expected recruitment has much lower uncertainty than the projected random effects regardless of whether bottom temperature affects recruitment.
-For the northern component, CVs of projected expected recruitments are smaller than those of recruitment random effects, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for future bottom temperature anomalies (Figure \ref{fig:R-F-SSB-CVs}).
-
-Because projected fishing mortality is specified to be the same as that in the terminal year the same uncertainty is also propagated in the projections (Figure \ref{fig:F-SSB-proj}).
-As we would expect, there is no effect of alternative assumptions about bottom temperature in the projections for the southern stock component because the effect of the bottom temperature anomaly is only on the northern component.
-However, the alternative bottom temperature assumptions do affect projections of northern component SSB and the projections do not stabilize when the anomalies are projected to increase linearly.
-When the same AR1 assumption for the bottom temperature anomalies is used in projections, CVs of northern and total SSB projections are larger than when the recent average or linear trend are assumed (Figure \ref{fig:R-F-SSB-CVs}).
-
-Unlike the annual reference points during the model years (prior to 2022), components of the $\upphi(\widetilde{F})$ calculations are constant in the projection years where the only changing annual values are the recruitments used in the weighting for $F_{40\%}$ and the recruitments used to calculate the resulting SSB$_{40\%}$ (Figure \ref{fig:brps-proj}).
-Estimated reference points in the first few projection years can differ depending on whether the expected recruitment or the random effects are used in their calculation, but longer term projected values are identical because the random effect estimates approach the expected values as the influence of observations diminishes.
-However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures \ref{fig:annual-F40-cvs} and \ref{fig:annual-SSB40-cvs}).
-The reduction in uncertainty using expected recruitment is less for $F_{40\%}$ than SSB$_{40\%}$ and for the northern component which includes bottom temperature effects than the southern component.
-
-# Discussion {-}
-
-## Black sea bass {-}
-
-In our investigation of bottom temperature effects on recruitment and natural mortality, and time-varying random effects on natural mortality at age 1, we found only including effects on recruitment for the northern stock component to be the best model with respect to AIC.
-BSB in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north [@mcmahanetal20].
-Opposite effects of temperature on recruitment or population size would be expected for species in the same general area that are at the southern extent of their range [@gabriel92].
-For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer and fall on the Northeast US shelf [@sullivanetal05;@milleretal16].
-
-Our finding regarding bottom temperature effects is consistent with the bottom temperature effects on recruitment in the assessment model accepted during the peer-review process and which is currently used to assess the stock [@nefsc23].
-However, AIC suggested some weight for models with temperature effects on recruitment for the southern component and random effects on M at age 1.
-Although this might suggest making inferences on the population by using an ensemble of these and weighting by AIC [@burnhamanderson02], the estimates of assessment outputs relevant for management (e.g., SSB and $F$) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
-Continued monitoring of temperature effects on the southern component and investigations of non-linear effects of temperature for both components may be prudent because of the possibility of an optimal temperature range for survival via effects on growth during the pre-recruit phase [@warelambert85;@anderson88].
-Furthermore, our inferences about effects on natural mortality at age 1 may be influenced by our assumptions of constant median (or just constant, depending on the model) natural mortality rate and movement rates for all ages.
-Further investigations of variation in these parameters with age and effects of bottom temperature could improve our understanding of the demography for this stock.
-
-We showed that using expected recruitment rather than recruitment random effects to define reference points can produce the same estimates but with lower uncertainty.
-Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same reference points and even lower uncertainty.
-These results can be important for management when uncertainty in assessment output is used in making catch advice.
-The typical approach for defining stock and harvest status is to compare reference points defined under prevailing conditions with current stock and harvest levels, but, as we demonstrated, it is also possible to make the comparisons of projected stock and harvest levels with projected reference points.
-The uncertainty in these ratios could also be used in defining catch advice.
-
-Reference points under prevailing conditions use average values for recruitment and other inputs and MSY-based reference points are based on the stock-recruit relationship which is also typically constant over time.
-If reference points should be defined in such a way that they are less variable than the stock over time, expected inputs to the reference points (instead of annual values) would be more appropriate.
-Lower variability in reference points would presumably correlate with lower variability in catch advice which might be desirable [e.g., @cochraneetal98;@kelletal05], but the appropriate level of temporal variability in reference points and catch advice will depend on the characteristics of the fish stock, associated fisheries, and management structure [@holland10;@puntetal16].
-
-For BSB, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figures \ref{fig:R-BT-proj} and \ref{fig:F-SSB-proj}).
-In the projection period, the AR1 models for bottom temperature and recuitment revert to the long-term means of those processes.
-This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypotheses of increasing water temperature and changing environmental conditions [@hareetal16;@pershingetal21].
-Because of these concerns, the most recent stock assessment for yellowtail flounder [@nefsc24a] used a change point analysis on environmental covariates to determine current conditions.
-The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections.
-Unfortunately, this methodology still has limitations because it is unlikely that current environmental conditions are stable.
-Similarly, a linear increase of temperature is also not likely in the future (top row of Figure \ref{fig:E-R-SSB-F}).
-As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in reference points and population projections becomes an important decision in the management process.
-
-## General model {-}
-
-WHAM provides a large class of models for time- and age-varying random effects for population attributes and treatment of environmental covariates and their effects on populations using state-space methods.
-This framework accounts for the magnitude of differing uncertainties in observations and stochastic population dynamics processes.
-Version 2.0 extends these inferences to movement rates for multi-region models and stock(component)-specific configuration of random and environmental covariate effects.
-
-An obvious limitation to WHAM is the inability to include tagging observations of any type.
-Such observations are critical to estimation of movement parameters [e.g., @goetheletal19].
-We used estimates of movement parameters from a companion assessment model to define prior distributions for corresponding movement parameters in the state-space BSB model, but other than the tagging data, both of the assessment models used much the same data which can lead to inappropriate inferences.
-In lieu of integrating an observation model for the tagging data, a better approach would be to estimate the movement parameters externally from just the tagging data.
-
-Tagging data can also inform mortality rate parameters [@hampton91].
-It is well known that natural mortality rates are seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality rates from other assessment model parameters [@leeetal11;@clark22].
-Estimation of natural mortality may be even more challenging within state-space assessment models with their greater flexibility from inclusion of time-varying random effects.
-For example, we fixed the mean natural mortality for age 1 fish and both @cadigan16 and @stocketal21 also estimated natural mortality deviations.
-However, the simulation studies of @milleretalinreview1 suggest estimating natural mortality in WHAM is feasible in some situations with typical assessment data.
-Regardless, tagging observations can provide improved inferences and allowing tagging observations in WHAM should be a high priority even for models with a single stock and region.
-
-WHAM version 2.0 only allows models where each stock spawns and recruits in a single region; however, some stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region [@kapuretal21].
-Therefore, extending WHAM to allow such models would be beneficial.
-Finally, because age composition data are scarce for many stocks, incorporating the work by @correaetal23 on allowing length composition and modeling growth within WHAM would provide an assessment tool to include process errors in assessment with such data and possibly with multiple stocks and regions.
-
-Our application to BSB is a particular combination of the spatial modeling options described by @bergeretal24.
-Other combinations of the described options are possible in WHAM 2.0.
-For example, WHAM 2.0 can accommodate all of fleet structure options described in Table 2 by @bergeretal24 and we expect more of these options to become available in future versions of WHAM.
-
-# Acknowledgements {-}
-
-We express our gratitude to the many scientists who have collected and processed the data that went into the assessment model, including survey biologists, port samplers, observers and agers.
-We thank the BSB research track assessment working group and peer-reviewers for the work compiling data and determining many of the configurations of the BSB assessment model that we used for our analyses.
-We also thank Chris Legault, two anonymous reviewers, and the associate editor for comments and suggestions that improved the clarity of this paper.
-
-\pagebreak
-
-# References {-}
-
-<div id="refs"></div>
-
-\pagebreak
-
-\pagebreak
-
-# Figures {-}
-
-(ref:caption) The Northwest Atlantic coastal area where the Northeast US black sea bass stock occurs.
-The red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations [after @milleretal16_yoy_survival] and the bathymetric contour is the 400 m isocline.
-
-```{r map, echo = FALSE, out.width="100%", fig.align="center", fig.cap = "(ref:caption)"}
-knitr::include_graphics(here::here("paper","map.pdf"))
-```
-\pagebreak
-
-```{r migration-diagram, echo = FALSE, out.width="80%", fig.align="center", fig.cap = "Diagram of intervals within the year and configuation of the dynamics of each component of the black sea bass population."}
-knitr::include_graphics(here::here("paper","bsb_movement_diagram.pdf"))
-```
-\pagebreak
-
-```{r move-prior-posterior, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Prior (black) and posterior (red) distributions of the movement parameter for the northern stock component from north to south (top) and south to north (bottom).
-Vertical lines indicate the prior and posterior estimates."}
-knitr::include_graphics(here::here("paper", "move_prior_post.pdf"))
-```
-\pagebreak
-
-```{r E-R-SSB-F, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual observations and posterior estimates of bottom temperature anomalies (top row), annual expected and random effect recruitment estimates (second row) and SSB and average fishing mortality at ages 6 and 7 from model $M_1$.
-Columns define estimates regionally (spawning region for SSB) and total SSB and fishing mortality across regions.
-Polygons and vertical lines represent 95\\% confidence intervals."}
-knitr::include_graphics(here::here("paper", "E_R_SSB_F_fig.pdf"))
-```
-\pagebreak
-
-```{r selectivity-re, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Time and age-varying selectivty for fleets and indices in the northern region with autoregressive random effects."}
-knitr::include_graphics(here::here("paper", "selectivity_re_plot.pdf"))
-```
-\pagebreak
-
-```{r annual-BRPs, echo = FALSE, eval= TRUE, out.height = "85%", fit.align = "center", fig.cap = "Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio ($F_{40\\%}$), equilibrium SSB at $F_{40\\%}$ based on annual inputs to $\\upphi(\\widetilde{F})$ calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to $\\upphi(\\widetilde{F})$ averaged over the last 5 years and recruitment is averaged over the time series.
-Red and blue in first 4 rows and black and gray in final row indicate alternative annual estimates of recruitment (random effect or expected).
-All results are based on model $M_1$.
-Polygons represent 95\\% confidence intervals or regions."}
-knitr::include_graphics(here::here("paper", "brp_status_results.pdf"))
-```
-
-```{r R-BT-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of bottom temperature in the northern region and alternative recruitment estimates (random effect or expected) by region.
-Estimates from years after 2021 are from projections of model $M_1$ under three alternative assumptions for the bottom temperature anomalies in the northern region.
-Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."}
-knitr::include_graphics(here::here("paper", "proj_ecov_Recruit_results.pdf"))
-```
-
-```{r F-SSB-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates of average fishing mortality and SSB by region and in total.
-Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."}
-knitr::include_graphics(here::here("paper", "proj_F_SSB.pdf"))
-```
-
-\begin{landscape}
-
-```{r brps-proj, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Annual estimates by region and in total of average equilibrium $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio and SSB at $F_{40\\%}$.
-Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region and average $\\upphi(\\widetilde{F})$ inputs from the last 5 years of the unprojected model.
-Vertical dotted lines indicate the last year of data and polygons represent 95\\% confidence intervals."}
-knitr::include_graphics(here::here("paper", "proj_brps.pdf"))
-```
-
-\end{landscape}
-
-\clearpage
-
-\setcounter{table}{0}
-\renewcommand\thetable{\arabic{table}}
-
-# Tables {-}
-
-```{r, echo = F, eval = T}
-mod_names <- paste0("$M_{", 0:13,"}$")
-col_names <- c("Model", "North", "South", "$M$ at age 1 random effects")
-
-mod_table <- matrix(nrow = length(mod_names), ncol = length(col_names), dimnames = list(mod_names,col_names))
-
-mod_table[,1] <- mod_names
-mod_table[,2:4] <- "--"
-mod_table[,4] <- rep(c("none", "time-varying"), each = 7)
-mod_table[c(2,4),2] <- "Recruitment"
-mod_table[c(3,4),3] <- "Recruitment"
-mod_table[c(5,7),2] <- "$M$ at age 1"
-mod_table[c(6,7),3] <- "$M$ at age 1"
-mod_table[8:14,2:3] <- mod_table[1:7,2:3]
-
-mod_table %>% kable(format = type, booktabs = T, escape=F, row.names = F, label = "model-desc-table",
-    caption="Assumptions for temperature effects and random effects for age 1 natural mortality rates ($M$) for each model.", align = c(rep("l",4))) %>%
-  add_header_above(c(" " = 1, "Temperature Effect" = 2, " " = 1 )) %>% kable_styling(latex_options = "hold_position")
-```
-
-```{r, echo = F, eval = T}
-diff_aic <- readRDS(here::here("results","diff_aic.RDS"))
-diff_aic <- data.frame(round(diff_aic, 2))
-diff_aic <- cbind(Model = row.names(diff_aic), diff_aic)
-row.names(diff_aic) <- NULL
-colnames(diff_aic) <- c("Model", as.character(0:7))
-diff_aic %>%
-  kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "diff-aic-table",
-  caption = "Difference between AIC and the lowest AIC for each model by retrospective peel.") %>%
-  add_header_above(c(" " = 1, "Peel" = 8))
-```
-
-```{r, echo = F, eval = T}
-rho_table <- readRDS("../results/rho_table.RDS")
-row.names(rho_table) <- NULL
-colnames(rho_table) <- c("Model", rep(c("North","South"), 2))
-rho_table %>%
-  kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "rho-table",
-  caption = "Mohn's $\\rho$ for SSB, and average F at ages 6 and 7 in northern and southern regions.") %>%
-  add_header_above(c(" " = 1, "SSB" = 2, "Average F" = 2))
-```
-```{r, echo = F, eval = T}
-par_names <- paste0("$M_{", c(1,2,3,3),"}$ ", "$\\widehat{\\beta}_{R,", c("N","S"), "}$")
-par_names <- c(par_names, paste0("$M_{", rep(0:1, 3),"}$ ", rep(c("Conditional ", "", "Marginal "),each = 2), "$\\widehat{\\", rep(c("sigma","rho","sigma"), each = 2),"}_{R,N}$"))
-peel_par_table <- t(readRDS(here::here("results/beta_sigma_par_peel_table.RDS"))[,1:10])
-# peel_par_table <- readRDS("../results/beta_sigma_par_peel_table.RDS")[,1:10]
-#colnames(peel_par_table) <- par_names
-peel_par_table <- cbind.data.frame(par_names, round(peel_par_table,3))
-colnames(peel_par_table) <- c("Parameter", as.character(0:7))
-row.names(peel_par_table) <- NULL
-peel_par_table %>%
-  kable(format = type, booktabs = T, escape=F, row.names = FALSE, label = "beta-sig-peel-table",
-  caption = "Estimates of temperature effects on recruitment and variance and autocorrelation parameters for recruitment for northern ($N$) and southern ($S$) components for models with no effects ($M_0$), effects on specific components ($M_1$ and $M_2$) or both components simultaneously ($M_3$) from the full model (Peel 0) and each retrospective peel.")%>%
-  add_header_above(c(" " = 1, "Peel" = 8))
-
-```
-
-\clearpage
-
-# Appendix A {-}
-
-\setcounter{table}{0}
-\renewcommand\thetable{A\arabic{table}}
-\input{symboltable_R2.tex}
-
-# Appendix B {-}
-
-New features of WHAM 2.0 that are not used in the application to BSB are described here..
-
-## Simultaneous movement and mortality {-}
-
-When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
-We obtain the PTM over an interval $i$ by exponentiating the instantaneous rate matrix [@millerandersen08]
-\begin{equation*}
-\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
-\end{equation*}
-The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region.
-Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero.
-For two regions and one fleet operating in each region:
- \begin{equation*}
- \mathbf{A}_{y,a,i} = \begin{bmatrix}
- a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{1,y,a,i} & 0 & M_{y,a,1} \\
- \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{2,y,a,i} & M_{y,a,2} \\
- 0 & 0 & 0 & 0 & 0 \\
- 0 & 0 & 0 & 0 & 0 \\
- 0 & 0 & 0 & 0 & 0
- \end{bmatrix},
-\end{equation*}
-where $a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + M_{y,a,r} + \sum^{n_F}_{f:r_f=r}F_{f,y,a,i})$.
-When there is one region, $n_F$ fleets, and $\delta_i = 1$, exponentiating the instantaneous rate matrix results in the PTM defined in Eq.
-\ref{eq:ptm_1_region}.
-
-## Initial abundance at age as random effects {-}
-
-The initial abundances at age can also be treated as independent or AR1 random effects. Defining the vector of initial abundance at age in region $r$ as $\mathbf{N}_{O,1,r}$, the general model is 
-\begin{equation*}
-\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r},
-\end{equation*}
-where 
-\begin{equation*}
-Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.
-\end{equation*}
-
-## General natural mortality model {-}
-
-The general model for natural mortality at age $a$ for a given stock in region $r$ is a function of random and covariate effects
-\begin{equation}\label{eq:M_model}
-  \log M_{y,a,r} = \theta_{M,a,r} + \varepsilon_{M,y,a,r} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} X_{k,y} = E\left(\log M_{y,a,r}|X_{k,y}\right) + \varepsilon_{M,y,a,r}.
-\end{equation}
-The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
-\begin{equation*}
-  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
-\end{equation*}
-
-## General catchability model {-}
-
-Catchability of each index can be treated as logit functions of normal random effects and(or) environmental covariate effects
-\begin{equation*}
-\log \left(\frac{q_{m,y}-l_m}{u_m-q_{m,y}}\right) = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} X_{k,y},
-\end{equation*}
-where $\theta_{q,m}$ is an intercept or mean parameter, and $u_{m}$ and $l_{m}$ are the upper and lower bounds of catchability for index $m$ (defaults are 0 and 1000).
-The general covariance structure for the annual random effects is AR1
-\begin{equation*}
-Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.
-\end{equation*}
-
-\pagebreak
-
-# Supplementary Materials {-}
 
@@
 
+# References {-}
+
+<div id="refs"></div>
+
+\pagebreak
+
 \setcounter{table}{0}
@@
 
+```{r self-test-fig, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Median relative error of SSB (Total and by stock component) and total fully-selected fishing mortality for estimation models fitted to simulated data from model $M_1$ where the observation variance of log-indices are fixed and estimated.
+Black and Red dashed lines represent the median of the annual medians and the median across all annual relative errors, respectively.
+Vertical lines represent 95\\% confidence intervals."}
+knitr::include_graphics(here::here("paper", "self_test_results.pdf"))
+```
+
+```{r BT-Ecov-R, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Expected and random effect recruitment estimates for the northern stock component.
+Color of points defined by the corresponding annual bottom temperature anomaly."}
+knitr::include_graphics(here::here("paper", "best_R_Ecov.pdf"))
+```
+
+\begin{landscape}
+```{r bottom-temperature, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Observations with 95\\% confidence intervals (points with vertical lines) and posterior estimates with 95\\% confidence intervals (lines with polygons) of bottom temperature anomalies in the north and south regions from model $M_1$.
+Gray vertical line defines the first year that the black sea bass stock is modeled."}
+knitr::include_graphics(here::here("paper", "BTA_full_fig.pdf"))
+```
+\end{landscape}
+\pagebreak
+
+```{r Ecov-M1-rel-M0, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Relative differences in posterior estimates of northern region bottom temperature anomalies ($\\widehat X$) from the null model without effects on recruitment ($M_0$) and with effects on the northern stock component ($M_1$)."}
+knitr::include_graphics(here::here("paper", "Ecov_M1_rel_M0.pdf"))
+```
+
+```{r selectivity-south, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Selectivty for fleets and indices in the southern region."}
+knitr::include_graphics(here::here("paper", "selectivity_south_plot.pdf"))
+```
+
+\begin{landscape}
+
+```{r sensitivity-plots, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Estimates of annual SSB and fishing mortality rates for the best performing model $M_1$ and models that are otherwise the same except where  1) movement rates are fixed at the means for the prior distribution, 2) a constant natural mortality rate is estimated, or 3) there is no movement for either stock component."}
+knitr::include_graphics(here::here("paper", "SSB_F_sensitivity_plots.pdf"))
+```
+\end{landscape}
+
+```{r annual-SSB40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium SSB$_{40\\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations and alternative annual recruitment types.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."}
+knitr::include_graphics(here::here("paper", "proj_SSB40_CV.pdf"))
+```
+```{r annual-SSB-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."}
+knitr::include_graphics(here::here("paper", "proj_SSB_status_CV.pdf"))
+```
+
+```{r annual-F40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."}
+knitr::include_graphics(here::here("paper", "proj_F40_CV.pdf"))
+```
+
+```{r annual-F-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\\bar{F}_{40\\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."}
+knitr::include_graphics(here::here("paper", "proj_F_status_CV.pdf"))
+```
+
+\begin{landscape}
+
+```{r R-F-SSB-CVs, echo = FALSE, eval= TRUE, out.height = "90%", fit.align = "center", fig.cap = "Coefficients of variation for estimates of alternative recruitment estimates (random effects or expected), average fishing mortality at age 6 and 7, and SSB by region and in total from model $M_1$.
+Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data."}
+knitr::include_graphics(here::here("paper", "R_SSB_F_cv_results.pdf"))
+```
+\end{landscape}
+
+\pagebreak
+
 ```{r jitter-M0, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Minimized negative log-likelihood for 50 fits where minimization used initial parameter values jittered from those provided by an initial fit for model $M_0$.
@@
 
-```{r self-test-fig, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Median relative error of SSB (Total and by stock component) and total fully-selected fishing mortality for estimation models fitted to simulated data from model $M_1$ where the observation variance of log-indices are fixed and estimated.
-Black and Red dashed lines represent the median of the annual medians and the median across all annual relative errors, respectively.
-Vertical lines represent 95\\% confidence intervals."}
-knitr::include_graphics(here::here("paper", "self_test_results.pdf"))
-```
-
-\begin{landscape}
-
-```{r bottom-temperature, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Observations with 95\\% confidence intervals (points with vertical lines) and posterior estimates with 95\\% confidence intervals (lines with polygons) of bottom temperature anomalies in the north and south regions from model $M_1$.
-Gray vertical line defines the first year that the black sea bass stock is modeled."}
-knitr::include_graphics(here::here("paper", "BTA_full_fig.pdf"))
-```
-
-\end{landscape}
-\pagebreak
-
-```{r BT-Ecov-R, echo = FALSE, eval= TRUE, out.width = "100%", fit.align = "center", fig.cap = "Expected and random effect recruitment estimates for the northern stock component.
-Color of points defined by the corresponding annual bottom temperature anomaly."}
-knitr::include_graphics(here::here("paper", "best_R_Ecov.pdf"))
-```
-\pagebreak
-
 ```{r SSB-F-R-rel-M1, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Estimates of SSB, F, and recruitment relative to those of the best performing model, $M_1$."}
 knitr::include_graphics(here::here("paper", "SSB_F_R_rel_M1.pdf"))
-```
-
-```{r Ecov-M1-rel-M0, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Relative differences in posterior estimates of northern region bottom temperature anomalies ($\\widehat X$) from the null model without effects on recruitment ($M_0$) and with effects on the northern stock component ($M_1$)."}
-knitr::include_graphics(here::here("paper", "Ecov_M1_rel_M0.pdf"))
-```
-
-```{r selectivity-south, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Selectivty for fleets and indices in the southern region."}
-knitr::include_graphics(here::here("paper", "selectivity_south_plot.pdf"))
-```
-\pagebreak
-
-\begin{landscape}
-
-```{r R-F-SSB-CVs, echo = FALSE, eval= TRUE, out.height = "90%", fit.align = "center", fig.cap = "Coefficients of variation for estimates of alternative recruitment estimates (random effects or expected), average fishing mortality at age 6 and 7, and SSB by region and in total from model $M_1$.
-Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data."}
-knitr::include_graphics(here::here("paper", "R_SSB_F_cv_results.pdf"))
-```
-\end{landscape}
-
-```{r annual-F40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data."}
-knitr::include_graphics(here::here("paper", "proj_F40_CV.pdf"))
-```
-
-```{r annual-SSB40-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual equilibrium SSB$_{40\\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations and alternative annual recruitment types.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data."}
-knitr::include_graphics(here::here("paper", "proj_SSB40_CV.pdf"))
 ```
@@
 ```
-
-```{r annual-F-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\\bar{F}_{40\\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data."}
-knitr::include_graphics(here::here("paper", "proj_F_status_CV.pdf"))
-```
-
-```{r annual-SSB-status-cvs, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\\upphi(\\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
-Vertical dotted lines indicate the last year of data."}
-knitr::include_graphics(here::here("paper", "proj_SSB_status_CV.pdf"))
-```
-
-\begin{landscape}
-
-```{r sensitivity-plots, echo = FALSE, eval= TRUE, out.height = "95%", fit.align = "center", fig.cap = "Estimates of annual SSB and fishing mortality rates for the best performing model $M_1$ and models that are otherwise the same except where  1) movement rates are fixed at the means for the prior distribution, 2) a constant natural mortality rate is estimated, or 3) there is no movement for either stock component."}
-knitr::include_graphics(here::here("paper", "SSB_F_sensitivity_plots.pdf"))
-```
-\end{landscape}

### 2025-07-01 paper/supp_final.tex
final

--- 2025-03-04 paper/paper.tex
@@
 \hypersetup{
-  pdftitle={Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass},
-  pdfauthor={Timothy J. Miller1; Kierstin Curti; Alexander C. Hansell},
+  pdfauthor={Timothy J. Miller1,2; Kiersten L. Curti1,3; Alexander C. Hansell1,4},
   hidelinks,
@@
 \usepackage{setspace}
-%\singlespacing
-%\onehalfspacing
-\doublespacing
+%\doublespacing
 \usepackage{lineno}
-\linenumbers
+%\linenumbers
 \usepackage[belowskip=0pt,aboveskip=0pt]{caption}
@@
 
-\title{Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass}
-\author{Timothy J. Miller\textsuperscript{1} \and Kierstin Curti \and Alexander C. Hansell}
-\date{04 March, 2025}
+\title{~Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass:\\
+\hspace*{0.333em}Supplementary Materials}
+\author{Timothy J. Miller\textsuperscript{1,2} \and Kiersten L. Curti\textsuperscript{1,3} \and Alexander C. Hansell\textsuperscript{1,4}}
+\date{}
 
@@
 
-\(^1\)\href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}, Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\\
+\(1\) Northeast Fisheries Science Center, National Marine Fisheries Service, 166 Water Street, Woods Hole, MA 02543, USA\\
+\(2\) orcid: 0000-0003-1411-1206, \href{mailto:timothy.j.miller@noaa.gov}{\nolinkurl{timothy.j.miller@noaa.gov}}\\
+\(3\) orcid: 0009-0009-1676-7269\\
+\(4\) orcid: 0000-0001-7827-1749\\
 
 \pagebreak
-
-\hypertarget{main-message}{%
-\subsection*{Main Message}\label{main-message}}
-\addcontentsline{toc}{subsection}{Main Message}
-
-Describes the multi-region, multi-stock generalization of WHAM and its usage to evaluate evidence of alternative hypotheses about temperature effects on recruitment and age 1 natural mortality of black sea bass stock components.
-
-\hypertarget{abstract}{%
-\subsection*{Abstract}\label{abstract}}
-\addcontentsline{toc}{subsection}{Abstract}
-
-The Woods Hole Assessment Model is a general state-space age-structured assessment model that is used to assess and manage many stocks in the Northeast US.
-We first describe an extension of the model allowing any number of stocks (or stock components) and and regions with movement among regions as well as seasonal variation in stock and fleet dynamics.
-Movement rates can be a be functions of time- and age-varying random effects and environmental covariates.
-We then illustrate the model by applying it to data for the northern and southern components of the Northeast US black sea bass stock and evaluate alternative hypotheses of bottom temperature and time-varying random effects on recruitment and natural mortality.
-We show strong evidence for temperature effects on recruitment, primarily for the northern stock component, and no evidence for including random effects or temperature effects on age 1 natural mortality.
-
-\pagebreak
-
-\hypertarget{introduction}{%
-\section*{Introduction}\label{introduction}}
-\addcontentsline{toc}{section}{Introduction}
-
-A state-space age-structured stock assessment models can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al.
-2016b).
-This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks (Punt et al.
-2020).
-The State-space Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed and remains widely used within the International Council for the Exploration of the Sea (ICES) to assess European fish stocks.
-Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada (Perreault et al.
-2020; Varkey et al.
-2022) and to the south, the Woods Hole Assessment Model (WHAM, Miller and Stock 2020; Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).
-
-WHAM is an R package developed and maintained at NOAA's Northeast Fisheries Science Center {[}\url{https://timjmiller.github.io/wham}{]}.
-WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and effects of environmental covariates on various population parameters.
-Like SAM, WHAM models are built using the Template Model Builder package (TMB, Kristensen et al.
-2016) which provides a computationally efficient means of fitting an extremely wide class of models with random effects.
-WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as ``apparent survival''), fishery and index selectivity, natural mortality, and catchability.
-
-However, WHAM has only allowed one stock (component) and a single region and without seasonal changes in stock dynamics.
-Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice (Ying et al.
-2011; Cao et al.
-2014; Bosley et al.
-2022).
-Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks (Punt et al.
-2020).
-We describe here the implementation of these features and other extensions since Stock and Miller (2021) in WHAM version 2.0.
-Many of these new configuration options can also be useful when modeling a single stock and region.
-
-We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) (NEFSC 2023).
-Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age assessment.
-The stock was split into two regions to account for spatial dynamics and improve model diagnostics (ASMFC 2016).
-Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (Moser and Shepherd 2009).
-Black sea bass distribution has been linked to warming waters on the northwest shelf (Bell et al.
-2015) and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare 2016).
-Thus, a focus of the research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the NEUS using most of the model assumptions that were accepted during the peer-review process.
-However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.
-
-When modeling multiple stock components occurring in different regions further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al.
-2021).
-Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises.
-Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points.
-We demonstrate some of options of WHAM version 2.0 to provide biological reference points and short-term projections.
-
-\hypertarget{methods}{%
-\section*{Methods}\label{methods}}
-\addcontentsline{toc}{section}{Methods}
-
-\hypertarget{wham-description}{%
-\subsection*{WHAM description}\label{wham-description}}
-\addcontentsline{toc}{subsection}{WHAM description}
-
-Many of the options and equations of WHAM version 2.0 are the same as those in Stock and Miller (2021), so we will only describe extensions and differences that have occurred since their first description of WHAM.
-The new version of WHAM can model multiple stocks each with their own movement, harvest, and natural mortality.
-Much of the description below is for a specific stock \(s\), but, for simplicity, this subscript is implicit except when necessary.
-
-\hypertarget{the-probability-transition-matrix}{%
-\subsubsection*{The probability transition matrix}\label{the-probability-transition-matrix}}
-\addcontentsline{toc}{subsubsection}{The probability transition matrix}
-
-Because individuals may be alive in one of several regions or harvested in one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model.
-Approaches to modeling transitions among these states may treat time discretely (e.g., Arnason 1972; Schwarz et al.
-1993) or continuously (e.g., Hearn et al.
-1987; Commenges 1999; Andersen and Keiding 2002).
-We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval \(\delta\) (columns), conditional on being in each of those state at beginning of the interval (rows).
-For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality.
-The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by season \(i\), year \(y\), and age \(a\) on January 1.
-Each row and column of the PTM correspond to one the states: alive in region \(r\), dead in fleet \(f\), or dead from natural causes.
-The probabilities in each row sum to unity (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval.
-Given \(n_R\) regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
-\begin{equation}\label{eq:ptm}
-  \mathbf{P}_{y,a,i} = \begin{bmatrix}
-    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
-    0 & \mathbf{I}_{H} & 0\\
-    0 & 0 & 1
-  \end{bmatrix}
-\end{equation}
-where
-\begin{equation*}
-  \mathbf{O}_{y,a,i} = 
-  \begin{bmatrix}
-    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
-    \vdots & \ddots & \vdots \\
-    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
-  \end{bmatrix}
-\end{equation*}
-is the \(n_R \times n_R\) matrix defining survival and occurring in each region at the end of the interval,
-\begin{equation*} 
-  \mathbf{H}_{y,a,i} = 
-  \begin{bmatrix}
-    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
-    \vdots & \ddots & \vdots \\
-    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
-  \end{bmatrix}
-\end{equation*}
-is the \(n_R \times n_F\) matrix defining probabilities of being captured in each fleet during the interval, and \(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of dying due to natural mortality during the interval.
-We have the identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).
-
-WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would be multinomial distributed as in a model for tagging data where fates of individual tagged fish are assumed independent.
-The PTMs determine the expected numbers 1) in each state on January 1 of year \(t+1\) at age \(a+1\) given the abundances at age \(a\) on January 1 of year \(t\), 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.
-
-\hypertarget{single-region-ptms}{%
-\subsubsection*{Single region PTMs}\label{single-region-ptms}}
-\addcontentsline{toc}{subsubsection}{Single region PTMs}
-
-When there is only one region,
-\begin{equation}\label{eq:ptm_1_region}
-\mathbf{P}_{y,a,i} = 
-  \begin{bmatrix}
-     S_{y,a,i} & \mathbf{H}_{y,a,i} & D_{y,a,i} \\
-     0 & \mathbf{I}_{H} & 0\\
-     0 & 0 & 1
-  \end{bmatrix}
-\end{equation}
-where \(S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}\), \(\mathbf{H}_{y,a,i}\) is a 1 x \(n_F\) matrix with elements for each fleet \(f\): \(\frac{F_{y,a,i,f}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), \(D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), and \(Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{y,a,i,f}\) is the total mortality rate.
-
-\hypertarget{multi-region-ptms}{%
-\subsubsection*{Multi-region PTMs}\label{multi-region-ptms}}
-\addcontentsline{toc}{subsubsection}{Multi-region PTMs}
-
-When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously.
-The sequential assumption is used widely in spatially explicit model (e.g., Stock Synthesis, Methot and Wetzel 2013).
-Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval.
-WHAM is configured to have movement occur after survival:
-\begin{equation*}
-  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i}
-\end{equation*}
-where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\) diagonal matrix of proportions surviving in each region (given they start in that region)
-\begin{equation*}
-\mathbf{S}_{y,a,i} = 
-  \begin{bmatrix}
-    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
-    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
-    \vdots & \vdots & \ddots & \vdots \\
-    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
-  \end{bmatrix}
-\end{equation*}
-and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\) matrix of probabilities of moving from one region to another or staying in the region they occurred at the beginning of the interval:
-\begin{equation*}
-\boldsymbol{\mu}_{y,a,i} = 
-  \begin{bmatrix}
-    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
-    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
-    \vdots & \vdots & \ddots & \vdots \\
-    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
-  \end{bmatrix}
-\end{equation*}
-
-WHAM assumes each fleet \(f\) can harvest in only one region (\(r_f\)) during specified seasons.
-So, for each fleet \(f\), row \(r_f\) and column \(f\) of \(\mathbf{H}_{y,a,i}\) will be \(F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\) when fleet \(f\) is harvesting during the interval \(\delta_i\) and all other elements will be zero.
-Row \(r\) of the single-column matrix \(\mathbf{D}_{y,a,i}\) is \(M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\)
-
-When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
-We obtain the probability transition matrix over an interval \(\delta_i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
-\begin{equation*}
-\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
-\end{equation*}
-The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region.
-Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero.
-For two regions and one fleet operating in each region:
-\begin{equation*}
- \mathbf{A}_{y,a,i} = \begin{bmatrix}
- a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
- \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
- 0 & 0 & 0 & 0 & 0 \\
- 0 & 0 & 0 & 0 & 0 \\
- 0 & 0 & 0 & 0 & 0
- \end{bmatrix}.
-\end{equation*}
-where \(a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})\).
-When there is one region, \(n_f\) fleets, and \(\delta_i = 1\), exponentiating the instantaneous rate matrix results in the PTM defined in Eq.
-\ref{eq:ptm_1_region}.
-
-\hypertarget{seasonality}{%
-\subsubsection*{Seasonality}\label{seasonality}}
-\addcontentsline{toc}{subsubsection}{Seasonality}
-
-Seasonality can be configured to accommodate characteristics of spawning, movement, and fleet-specific behavior.
-The annual time step can be divided into \(K\) seasons and the interval size \(\delta_i\) for each season \(i\) does not need to be equal to any other seasonal interval.
-Under the Markov assumption, the PTM of surviving and moving and dying over \(K\) intervals \(\delta_1,\ldots, \delta_K\) (i.e., the entire year) is just the product of the PTMs for each interval:
-\[ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).\]
-
-For a stock spawning at some fraction of the year \(0<t_s<1\) in interval \(\delta_j\), the fraction of time in season \(j\) is
-\[\delta_{s,j} = t_s-\sum^{j-1}_{i=0}\delta_i
-\]
-and the PTM defining the proportions in each state at time \(t_s\) for age \(a\) is
-\begin{equation}\label{eq:ptm_spawn}
-\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{s,j}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{s,j}).
-\end{equation}
-Similarly, for an index \(m\) occurring at fraction of the year \(t_m\) in interval \(\delta_j\) the proportions in each state at the time of the observation is
-\begin{equation} \label{eq:ptm_index} 
-\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{m,j}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{m,j}).
-\end{equation}
-
-\hypertarget{numbers-at-age}{%
-\subsubsection*{Numbers at age}\label{numbers-at-age}}
-\addcontentsline{toc}{subsubsection}{Numbers at age}
-
-When there are \(n_R\) regions and \(n_F\) fleets, the vector of abundance in each state at age \(a>1\) on January 1 is \(\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}', \mathbf{0}')'\) where \(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})'\) is the number in the states corresponding to being alive in each region and \(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each fleet and dead from natural mortality because no age \(a\) fish have died yet on January 1.
-
-Each stock \(s\) is assumed to spawn and recruit in one region \(r_s\).
-So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row \(r = r_s\).
-Options for configuring recruitment (\(N_{y,1,r_s}\)) for each stock are the same as previous versions of WHAM (Miller and Stock 2020).
-If recruitment is assumed to be a function of SSB, it is only the spawning population in region \(r_s\) at the time of spawning that constitutes the SSB in the stock-recruit function.
-However, models can configure spawning individuals to occur in other regions at the time of spawning.
-Aside from treating recruitment as a random walk, the general model for annual recruitment as random effects is
-\begin{equation*}
-\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1,r_s} =  f\left(\text{SSB}_{y-1,r_s}\right) + \varepsilon_{y,1,r_s}
-\end{equation*}
-where\\
-\[\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}(t_s)' \mathbf{N}_{O,y,a}\]
-and \(w_{y,a}\) is the mean weight at age of spawning individuals, \(\text{mat}_{y,a}\) is the maturity at age, and \(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and occurring in region \(r_s\) at time \(t_s\) given being alive in each region at the start of the year (the \(r_s\) column of the upper-left submatrix of Eq.
-\ref{eq:ptm_spawn}).
-
-As in previous versions of WHAM, the transitions in numbers at age from one year to another after recruitment can be treated deterministically or as functions of random effects.
-The predicted numbers at age in year \(y\) at age \(a\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (Stock and Miller 2021).
-For ages \(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first \(n_R\) elements of the vector
-\[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\]
-The numbers alive in each region can also be modeled more simply using the sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions in abundance at age is
-\begin{equation*}
-\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
-\end{equation*}
-for ages \(a = 2,\ldots, A-1\), and for the plus group
-\begin{equation*}
-\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
-\end{equation*}
-where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
-When there is autocorrelation with age, WHAM now assumes this applies only to ages \(a>1\) by default so that recruitment random effects are independent of those for for the annual transitions of older age classes.
-So the general covariance structure for a given stock at ages \(a>1\) in region \(r\) is that of a two-dimensional first-order autoregressive (2DAR1) process
-\begin{equation*}
-  Cov\left(\varepsilon_{y,a,r},\varepsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)}
-\end{equation*}
-and that for age 1 is just AR1 across years
-\begin{equation*}
-  Cov\left(\varepsilon_{y,1},\varepsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
-\end{equation*}
-When the transitions in abundance at age are treated deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\).
-Since recruitment for a given stock currently only occurs in one region \(r_s\) there is only a single vector of time-varying recruitment random effects for each stock.
-
-\hypertarget{initial-numbers-at-age}{%
-\subsubsection*{Initial numbers at age}\label{initial-numbers-at-age}}
-\addcontentsline{toc}{subsubsection}{Initial numbers at age}
-
-Initial numbers at age for each stock and region can be treated as age-specific fixed effects or with an equilibrium assumption as in previous versions of WHAM.
-For the equilibrium option there are two parameters for each stock: the stock-specific fully-selected fishing mortality rate \(\log \widetilde{F}\) and the recruitment in year 1 \(\log N_{1,1,r_s}\).
-A stock-specific equilibrium fishing mortality at age by fleet \({\tilde F}_{a,f}\) is the product of \(\widetilde{F}\) and the selectivity across fleets in the first year
-\begin{equation*}
-  \text{sel}_{1,a,f} = \frac{F_{1,a,f}}{\max_a \sum_{f=1}^{n_F} F_{1,a,f}}.
-\end{equation*}
-We use \(\widetilde{F}_{a,f}\) to define an equilibrium abundance per recruit by region at age \(a\) conditional on recruiting to each region
-\begin{equation}\label{eq:npraa}
- \widetilde{\mathbf{O}}_{a} = \left\{
- \begin{array}{ll}
-\prod^{a-1}_{j=0}\mathbf{O}_{j}  & 1\leq a<A\\
-\left[\prod^{a-1}_{j=0}\mathbf{O}_{j}\right] \left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1} & a = A
- \end{array}
-\right.
-\end{equation}
-where \(\mathbf{O}_{j}\) is the equilibrium probability of surviving age \(a\) and occurring in each region and \(\mathbf{O}_{0} = \mathbf{I}\).
-Natural mortality and movement rates in the first year of the model are also used in Eq.
-\ref{eq:npraa}.
-For the plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is a ``fundamental matrix'\,' derived using the matrix version of the geometric series (Kemeny and Snell 1960).
-Recall that recruitment for stock \(s\) only occur in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
-\[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}' \mathbf{N}_{O,1,1}.\]
-
-The initial abundances at age can also be treated as independent or AR1 random effects. Defining the vector of initial abundance at age in region \(r\) as \(\mathbf{N}_{O,1,r}\), the general model is
-\[\log \mathbf{N}_{O,1,r} = \theta_{N_1,r} + \boldsymbol{\varepsilon}_{N_1,r}\]
-where
-\[  Cov\left(\varepsilon_{N_1,a,r},\varepsilon_{N_1,a',r}\right) = \frac{\rho_{N_1,r}^{|a-a'|}\sigma^2_{N_1,r}} {\left(1-\rho_{N_1,r}^2\right)}.\]
-
-\hypertarget{parametizing-movement}{%
-\subsubsection*{Parametizing movement}\label{parametizing-movement}}
-\addcontentsline{toc}{subsubsection}{Parametizing movement}
-
-For each season, there are at most \(n_R-1\) parameters determining movement among regions given starting the season in region \(r\) in either the sequential or simultaneous configurations.
-Movement parameters are estimated on a transformed scale via a link function \(g(\cdot)\).
-If survival and movement occur simultaneously, the parameters are estimated with a log link function and if they are separable, an additive logit link function (like a multinomial regression) is used.
-On the transformed scale, the general model for the movement parameter from region \(r\) to \(r'\) in season \(i\) and year \(y\) for individuals of age \(a\) is a linear function of an intercept or mean parameter (\(\theta_{r\rightarrow r',i}\)) and both random and environmental effects:
-\begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \theta_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r \rightarrow r',a,i,k} E_{k,y}.
-\end{equation*}
-The random effects \(\varepsilon_{r\rightarrow r',y,a,i}\) are season-, and region-to-region-specific and modeled most generally as 2DAR1 random effects with age and(or) year where the covariance is
-\begin{equation*}
-  Cov\left(\varepsilon_{r\rightarrow r',y,a,i},\varepsilon_{r\rightarrow r',y',a',i}\right) =   \frac{\rho_{r\rightarrow r',\text{age},i}^{|a-a'|}\rho_{r\rightarrow r',\text{year},i}^{|y-y'|}\sigma^2_{r\rightarrow r',i}}{\left(1 -  \rho_{r\rightarrow r',\text{age},i}^2\right)\left(1 - \rho_{r\rightarrow r',\text{year},i}^2\right)}
-\end{equation*}
-similar to how WHAM models variation in survival, natural mortality, and selectivity.
-Effects of covariate \(E_k\) can be age-, season-, and region-to-region-specific \(\beta_{r\rightarrow r',a,i,k}\) and the same orthogonal polynomial options in the previous versions of WHAM for effects on recruitment and natural mortality are available.
-
-There is currently no likelihood component for tagging data.
-Therefore, movement parameters would usually either need to be fixed or assumed to have some prior distribution, possibly based on external parameter estimates.
-WHAM includes prior distributions for the season and region-to-region specific (mean) movement parameters and treats them as random effects with the mean defined by the initial value of the fixed effect counterpart and standard deviation
-\begin{equation*}
-  \gamma_{r\rightarrow r',i} \sim \text{N}\left(\theta_{r\rightarrow r',i}, \sigma^2_{r\rightarrow r',i}\right).
-  \end{equation*}
-When priors are used, the movement is defined instead as
-\begin{equation*}
-  g(\mu_{r\rightarrow r',y,a,i}) = \gamma_{r\rightarrow r',i} + \varepsilon_{r\rightarrow r',y,a,i} + \sum^{n_E}_{k=1} \beta_{r\rightarrow r',a,i,k} E_{k,y} 
-  \end{equation*}
-
-\hypertarget{natural-mortality}{%
-\subsubsection*{Natural mortality}\label{natural-mortality}}
-\addcontentsline{toc}{subsubsection}{Natural mortality}
-
-When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific.
-When random effects are used, the same 2DAR1 structure with age and year as described by Stock and Miller (2021) can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
-So the general model for natural mortality is
-\begin{equation*}
-  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
-\end{equation*}
-The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
-\begin{equation*}
-  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
-\end{equation*}
-
-\hypertarget{catch-observations}{%
-\subsubsection*{Catch observations}\label{catch-observations}}
-\addcontentsline{toc}{subsubsection}{Catch observations}
-
-The log-normal distributional assumption for aggregate catch observations is the same as Stock and Miller (2021), but the predicted catch is now a function of catch from each stock starting the year in each region.
-For a given stock and age, the numbers captured in each fleet over the year are
-\[\widehat{\mathbf{N}}_{H,s,y,a} = \mathbf{H}_{s,y,a}' \mathbf{N}_{O,s,y,a}\]
-The predicted numbers caught be each fleet across stocks is
-\[\widehat{\mathbf{N}}_{H,y,a} = \sum^{n_S}_{s=1} \widehat{\mathbf{N}}_{H,s,y,a}\]
-and the predicted aggregate catch at age \(a\) is
-\[\widehat{\mathbf{C}}_{y,a} = \text{diag}\left(\mathbf{c}_{y,a}\right) \widehat{\mathbf{N}}_{H,y,a}\]
-where \(\mathbf{c}_{y,a}\) is the vector of mean individual weight at age \(a\) for each fleet and the aggregate catch by fleet is
-\[\widehat{\mathbf{C}}_y = \sum^{A}_{a=1} \widehat{\mathbf{C}}_{y,a}\]
-The log-aggregate catch observations for fleet \(f\) are normally distributed
-\[ \log C_{y,f} \sim \text{N}\left(\log \widehat {C}_{y,f}, \sigma^2_{y,f}\right).\]
-
-The predicted numbers caught for each fleet \(f\) (row \(f\) of \(\widehat{\mathbf{N}}_{H,y,a}\)) are used to make predicted age composition observations as described by Stock and Miller (2021).
-Since then, three additional likelihood options for age composition observations have been added: a logistic-normal with AR(1) correlation structure (Francis 2014), the alternative Dirichlet-multinomial parameterization described by Thorson et al.
-(2017), and the multivariate Tweedie (Thorson et al.
-2023).
-
-\hypertarget{index-observations}{%
-\subsubsection*{Index observations}\label{index-observations}}
-\addcontentsline{toc}{subsubsection}{Index observations}
-
-For index \(m\) occurring in region \(r_m\) at fraction of the year \(t_m\), the predicted abundance at \(t_m\) in region \(r_m\) is
-\[\widehat{N}_{s,y,a,m} = \mathbf{O}_{s,y,a,r_m}(t_m)' \mathbf{N}_{O,s,y,a}\]
-where \(\mathbf{O}_{s,y,a,r_m}(t_m)\), the \(r_m\) column of the upper-left submatrix of Eq.
-\ref{eq:ptm_index}, are the probabilities of surviving and occurring in region \(r_m\) at time \(t_m\) given being alive in each region at the start of the year.
-The predicted index at age is
-\[\widehat{I}_{m,y,a} = q_{m,y} \text{sel}_{m,y,a}w_{m,y,a}\sum^{n_S}_{s = 1}\widehat{N}_{s,y,a,m}\]
-where \(q_{m,y}\) is the catchability of the index in year \(y\), \(\text{sel}_{m,y,a}\) is the selectivity and \(w_{m,y,a}\) is the average weight of individuals at age \(a\) if the index is quantified in biomass and \(w_{m,y,a} = 1\) if the index is quantified in numbers.
-Predicted age composition observations are functions of \(\widehat{I}_{m,y,a}\) as described by Stock and Miller (2021) and the likelihood options are the same as those for catch explained above.
-
-Catchability of the index can also be treated as logit functions of normal random effects and(or) environmental covariate effects
-\[log \frac{q_{m,y}-l_m}{u_m-q_{m,y}} = \theta_{q,m} + \varepsilon_{q,m,y}  + \sum^{n_E}_{k=1} \beta_{q,m,k} E_{k,y}\]
-where \(\theta_{q,m}\) is an intercept or mean parameter, and \(u_{m}\) and \(l_{m}\) are the upper and lower bounds of catchability for index \(m\) (defaults are 0 and 1000).
-The general covariance structure for the annual random effects is AR1
-\[Cov\left(\varepsilon_{q,m,y},\varepsilon_{q,m,y'}\right) =   \frac{\rho_{q,m}^{|y-y'|}\sigma^2_{q,m}}{1 - \rho_{q,m}^2}.\]
-
-\hypertarget{reference-points}{%
-\subsubsection*{Reference points}\label{reference-points}}
-\addcontentsline{toc}{subsubsection}{Reference points}
-
-Currently a single fishing mortality reference point \(\widetilde F\) is estimated across stocks and regions and the fleet- and age-specific reference points is \(\widetilde F_{f,a} = \widetilde F \text{sel}_{f,a}\).
-Selectivity is determined across fleets and ages as in the equilibrium assumption for initial abundance at age except here it is a function of average F at age over a user-defined set of years
-\begin{equation}\label{eq:spr_sel}
-  \text{sel}_{a,f} = \frac{\overline F_{a,f}}{\max_a \sum^{n_F}_{f=1}{\overline F}_{a,f}}
-\end{equation}
-
-The equilibrium spawning stock biomass per recruit for a given stock in region \(r_s\) is defined as
-\begin{equation}\label{eq:ssbpr}
- \upphi(\widetilde{F}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{O}_{a,\cdot,r_s}(t_s) w_{a} m_{a}
-\end{equation}
-where \(w_{a}\) and \(m_{a}\) are the mean individual weight and probability of maturity at age \(a\), \(\widetilde{\mathbf{O}}_{a}\) are as described in Eq.
-\ref{eq:npraa}, and \(\mathbf{O}_{a}(t_s)\) is the \(n_R \times n_R\) upper-left sub-matrix of eq.
-\ref{eq:ptm_spawn} with the probabilities of surviving and occurring in each region \(r'\) at age \(a+t_s\) given starting in region \(r\) at age \(a\).
-The further subscripts \(r_s,\cdot\) and \(\cdot,r_s\) indicate row or column \(r_s\), respectively.
-Using these rows and columns is required because of the assumption that spawning and recruitment only occur in region \(r_s\).
-
-The equilibrium spawning biomass per recruit (eq.
-\ref{eq:ssbpr}) is conditional on the region of recruitment \(r_s\).
-The equilibrium recruitment depends on the stock dynamics.
-WHAM currently only allows complete spawning region fidelity so that a stock only spawns and recruits in a single region.
-Therefore, the equilibrium vector of numbers at age 1 for each state, \(\widetilde {\mathbf{N}}_{1}\), will only be nonzero in the row corresponding to the spawning region (\(r_s\)).
-The matrices of probabilities of surviving and occurring in each region, \(\widetilde{\mathbf{O}}_{s,a}\) and \(\mathbf{O}_{s,a}(t_s)\), are functions of the fishing mortality rates for fleets in each region \(\widetilde F_{f,a}\).
-
-The matrix of equilibrium yield per recruit by region and fleet as a function of \(\widetilde F\) is calculated as
-\begin{equation}\label{eq:ypr}
- \widetilde{\nu}({\widetilde{F}}) = \sum^{A}_{a=1} \widetilde{\mathbf{O}}_{a,r_s,\cdot} \mathbf{H}_{a} \mathbf{c}_{s,a}
-\end{equation}
-where \(\mathbf{c}_a\) is the vector of mean individual weight at age for each fleet, and \(\mathbf{H}_{s,a}\) is the submatrix of the probabilities of being captured in each fleet over the interval from \(a\) to \(a+1\), defined in eq.
-\ref{eq:ptm}.
-
-As in previous versions of WHAM package, reference points for prevailing conditions, average all of the inputs to the spawning biomass and yield per recruit calculations (weight, maturity, natural mortality, and movement) over a set of user-specified years as shown in eq.
-\ref{eq:spr_sel} for selectivity (e.g., last 5 years of the model).
-
-For \(X\%\) SPR-based reference points, we use a Newton method and iterate
-\begin{equation}\label{eq:newton}
-  \log\widetilde{F}^{(i)} = \log\widetilde{F}^{(i-1)} - \frac{g\left(\log\widetilde{F}^{(i-1)}\right)}{g'\left(\log\widetilde{F}^{(i-1)}\right)}
-\end{equation}
-where \(g(\log F)\) is the difference between the weighted sums of spawning biomass per recruit at \(F\) and \(X\)\% of unfished spawning biomass per recruit across stocks:
-\begin{equation}\label{eq:newton-obj}
-  g(\log F) = \sum^{n_s}_{s=1} \lambda_s\left[\upphi_s\left(F = e^{\log F}\right) - \frac{X}{100}\upphi_s\left(F=0\right)\right].
-\end{equation}
-where \(\upphi_s\left(F=0\right)\) is the equilibrium unfished spawning biomass per recruit.
-\(g'(\log F)\) is the derivative of \(g\) with respect to \(\log F\), and the weights to use for each stock \(\lambda_s\) can be specified by the user or relative to the average of recruitment for each stock over the same years the user defines to calculate ``static'\,' equilibrium spawning biomass and yield.
-
-When a Beverton-Holt or Ricker stock recruit relationship is assumed, an analogous Newton method is used to find \(\log F\) that maximizes yield for MSY-based reference points.
-which are also a functions of the equilibrium yield per recruit (Eq.
-\ref{eq:ypr}) and equilibrium recruitment.
-The function \(g(\log F)\) in Eq.
-\ref{eq:newton} is the first derivative of the yield curve with respect to \(\log F\).
-
-\hypertarget{projections}{%
-\subsubsection*{Projections}\label{projections}}
-\addcontentsline{toc}{subsubsection}{Projections}
-
-The projection options are generally the same as those for previous versions of WHAM (Miller and Stock 2020).
-When there is movement of any stocks, the user has the option to project and use any random effects for time-varying movement or use the average over user specified years, analogous to how natural mortality can be treated in the projection period.
-The projection of any environmental covariates has been revised to better include error in the estimated latent covariate in any effects on the population in projection years.
-Users can also specify a catch or fully-selected fishing mortality in each projection year and they can be fleet-specific.
-
-\hypertarget{application-to-black-sea-bass}{%
-\section*{Application to black sea bass}\label{application-to-black-sea-bass}}
-\addcontentsline{toc}{section}{Application to black sea bass}
-
-Black sea bass are a temperate reef fish in the western Atlantic Ocean ranging from the Gulf of Mexico to the Gulf of Maine.
-Fish north of Cape Hatteras, NC, are considered to comprise a single stock unit, but individual populations exhibit spawning site fidelity
-(Able and Stanton Hales, Jr.
-1997; Fabrizio et al.
-2013).
-Fish in this northern stock unit perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring (Musick and Mercer 1977).
-Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude (Moser and Shepherd 2009).
-
-Prior to its 2023 peer-reviewed assessment, the NEUS black sea bass stock was assessed using the Age-Structured Assessment Program (ASAP) model (Legault and Restrepo 1999), a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
-Northern and southern components of the NEUS black sea bass stock ascribed to regions divided by the Hudson Canyon were separately modeled in ASAP (Figure \ref{fig:map}).
-Results from the separate ASAP models were combined for a unit-stock assessment.
-The ASAP-based assessments exhibited strong retrospective patterns (Mohn 1999), and exploring alternative modeling approaches for the northern and southern stock components has been a high priority for management.
-
-Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as ``Working Group'') composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the black sea bass stock using the multi-stock and multi-region extension of WHAM described above (NEFSC 2023).
-This assessment included the spatial features and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous black sea bass assessments.
-Below we describe the assumptions and configuration of the assessment model as determined by the Working Group as well as the alternative assumptions for recruitment and natural mortality in the models we fit to evaluate alternative hypotheses of bottom temperature effects on black sea bass.
-
-\hypertarget{basic-structure}{%
-\subsection*{Basic structure}\label{basic-structure}}
-\addcontentsline{toc}{subsection}{Basic structure}
-
-The first year being modeled for the population is 1989 and the fishery and index data used in the model span from 1989 to 2021.
-The north and south stock components are modeled as separate populations that spawn and recruit in respective regions.
-We have observations for each of four total fishing fleets, where two fishing fleets (recreational and commercial) operate in each region.
-
-There are 11 ``seasonal'' intervals within each calendar year: five monthly time intervals from Jan 1 to May 31, a spawning season from June 1 to July 31, and five monthly intervals from August 1 to December 31.
-The southern stock component is assumed to never move to the northern region.
-For the northern component, a proportion \(\mu_{\text{N}\rightarrow \text{S}}\) can move to the southern region each month during the last five months of the year, but no movement is allows from the south to the north during this period (Figure \ref{fig:migration-diagram}).
-During the first four intervals of the year a proportion \(\mu_{\text{S}\rightarrow \text{N}}\) the northern component individuals in the south can move back to the north, but no movement from the north to south is allowed during this period.
-In the fifth interval (May), all northern component individuals remaining in the south are assumed to move back to the north for the subsequent spawning period.
-Survival and movement occur sequentially in each interval and each of the two movement proportions are assumed constant across intervals, ages, and years.
-
-The two monthly movement matrices are
-\begin{equation*}
-\boldsymbol{\mu}_{1} = 
-  \begin{bmatrix}
-     1-\mu_{\text{N}\rightarrow \text{S}} & \mu_{\text{N}\rightarrow \text{S}} \\
-     0 & 1 \\
-  \end{bmatrix}
-\end{equation*}
-for the portion of the year after spawning and
-\begin{equation*}
-\boldsymbol{\mu}_{2} = 
-  \begin{bmatrix}
-     1 &  0 \\
-     \mu_{\text{S}\rightarrow \text{N}} & 1-\mu_{\text{S}\rightarrow \text{N}} \\
-  \end{bmatrix}
-\end{equation*}
-for the portion of the year before spawning.
-As noted in the description of the general WHAM model, tagging data are not yet allowed.
-However, the Working Group also fit a Stock Synthesis model (Methot and Wetzel 2013) which included tagging data and provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).
-
-With movement as configured, the northern origin fish (ages 2+) can occur in the southern region on January 1.
-Estimating initial numbers at age as separate parameters can be challenging even in single-stock models, but for black sea bass the available data cannot distinguish the proportion of northern and southern component fish at each age in the southern region in the initial year of the model.
-Therefore, we used the simplifying equilibrium assumption where there are two parameters estimated for each stock component: an initial recruitment and an equilibrium fully-selected \(F\) that determines the abundance at age in each region for each stock component.
-
-For subsequent years, all models allow AR1 correlation of recruitment and 2DAR1 correlation for survival for both the northern and southern stock components.
-Models with apparent survival random effects will model the transitions (conditional on both survival and movement) of abundances at age of northern origin fish in each region, but there appears to be little ability of the existing data from the southern region to distinguish apparent survival random effects for each stock component.
-All models we considered assume very small variance for the transitions of northern fish in the southern region, which is essentially the same as the deterministic transition assumptions of a statistical catch at age model.
-Unique variance and correlation parameters for the recruitment and apparent survival random effects are estimated for the northern and southern components.
-
-\hypertarget{uncertainty-in-recreational-index-observations}{%
-\subsection*{Uncertainty in recreational index observations}\label{uncertainty-in-recreational-index-observations}}
-\addcontentsline{toc}{subsection}{Uncertainty in recreational index observations}
-
-The estimated coefficients of variation (CVs) provided by the analyses by the Working Group to generate the recreational catch per angler (recreational CPA) index ranged between 0.02 and 0.06 which the Working Group felt did not capture the true observation uncertainty in the index with regard to its relationship to stock abundance.
-All models we considered allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
-However, lack of model convergence was observed in many fits for self-test simulations.
-The model estimates are negligibly affected by estimating these scalars, but estimating rather than fixing these parameters allowed uncertainty in model output to be more properly conveyed.
-
-\hypertarget{index-and-catch-age-composition-observations}{%
-\subsection*{Index and catch age composition observations}\label{index-and-catch-age-composition-observations}}
-\addcontentsline{toc}{subsection}{Index and catch age composition observations}
-
-The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual patterns and retrospective patterns.
-These analyses resulted in use of selectivity random effects for the northern fleet and indices and logistic-normal likelihoods for six sets of of age composition observations and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
-
-\hypertarget{bottom-temperature-effects}{%
-\subsection*{Bottom temperature effects}\label{bottom-temperature-effects}}
-\addcontentsline{toc}{subsection}{Bottom temperature effects}
-
-All models include bottom temperature observations for the northern and southern regions from 1963 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius (NEFSC 2023).
-We retained the assumption from the peer-reviewed assessment that treated the latent bottom temperature covariates in each region as AR1 processes.
-
-We fit 14 models with alternative assumptions about the effects of bottom temperature covariates, ranging from no effects to effects on both regions for either recruitment or natural mortality at age 1 (Table \ref{tab:model-desc-table}).
-These analyses derive from the hypothesis that bottom temperature affects overwinter survival of fish where the fish turn from age 0 to age 1 on January 1 (Miller et al.
-2016a).
-This covariate may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of of the year after becoming age 1.
-Furthermore, we have no direct observations of age 1 individuals from surveys until the spring season each year.
-Therefore, we fit models with effects of temperature on recruitment or natural mortality at age 1.
-
-All models treat annual recruitment as AR1 time-varying random effects with mean parameters for each stock component.
-However, Miller et al.
-(2018) showed how inferences of temperature effects on growth or maturity parameters can be very different whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
-Therefore, we also explored whether including temporal random effects on age 1 natural mortality affected inferences on corresponding temperature effects.
-Initially, we included random effects on age 1 natural mortality for both the northern and southern stock components, but estimates of these random effects and corresponding variance for the southern component converged to 0 so these were not included in models presented here.
-
-We assume the covariate in year \(y\) affects recruitment in the same years because the covariate observations are from months January to March.
-The fish are technically already 1 year old, but there are no observations of these individuals until later in the year except possibly in fishery catches which are accumulated over the whole year.
-Expected log-recruitment for a given stock is a linear function of bottom temperature
-
-\begin{equation}\label{eq:expected-recruitment}
-E\left(\log N_{y,1}|x_y\right) = \mu_{R} + \beta_{R} x_y.
-\end{equation}
-Similarly, expected log-natural morality as a function of bottom temperature is
-\begin{equation}\label{eq:expected-M1}
-E\left(\log M_{y,1}|x_y\right) = \mu_{M,1} + \beta_{M} x_y
-\end{equation}
-Because age 1 fish for the northern component can exist in both regions after January 1, natural mortality is acting in both regions for this stock.
-For models with covariate and(or) annual random effects for age 1 fish we assume them only in the northern regions for the northern component.
-The corresponding random effects are
-\begin{equation}\label{eq:Rec-re}
-\log N_{y,1} = E\left(\log N_{y,1}|x_y\right) + \varepsilon_{y,1}
-\end{equation}
-and
-\begin{equation}\label{eq:M-re}
-\log M_{y,1} = E\left(\log M_{y,1}|x_y\right) + \varepsilon_{M,y}.
-\end{equation}
-The constant or mean log-natural mortality rate is assumed to be \(\mu_{M,a} = \log(0.4)\) for all ages as recommended by the Working Group.
-Because the bottom temperature anomalies and random effects are centered at 0, mean log-natural mortality at age 1 over the time series should be approximately equal to \(\mu_{M,1}\) for models that include those effects.
-
-\hypertarget{model-fitting-and-diagnostics}{%
-\subsection*{Model fitting and diagnostics}\label{model-fitting-and-diagnostics}}
-\addcontentsline{toc}{subsection}{Model fitting and diagnostics}
-
-We used a development version (\href{https://github.com/timjmiller/wham/tree/fb8b089}{commit fb8b089}) of the WHAM package prior to the release of version 2.0 for all results.
-All code for fitting models and generating results can be found at \href{https://github.com/timjmiller/wham_bsb_paper}{github.com/timjmiller/wham\_bsb\_paper}.
-
-We examined retrospective patterns for all models by fitting corresponding models where the terminal year is reduced sequentially by one year (peel) for seven years.
-Therefore, there are eight fits of each model with the time series reduced by zero to seven years.
-We calculated Mohn's \(\rho\) for SSB, and average \(F\) at ages 6 and 7.
-Absolute values of Mohn's \(\rho\) near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended.
-As in Miller et al.
-(2016b), we also assessed the consistency of the AIC-based model selection over retrospective peels to guard against previously noted changes in perception of covariate effects on recruitment with increased length of the time series of observations (Myers 1998).
-This retrospective examination was also recommended by Brooks (2024).
-
-We performed a jitter analysis of the base model \(M_0\) and the best fitting model to investigate whether a local mimimum of the negative log-likelihood surface was obtained by the optimization.
-We used the jitter\_wham function in the WHAM package which by default simulates starting values from a multivariate normal distribution with mean and covariance defined by the MLEs and hessian-based covariance matrix from the fitted model.
-See the Supplementary Materials for more details.
-
-For the best performing model, we also performed simulation self-tests where new observations were simulated conditional on all estimated fixed and random effects and the same model configuration was fit to each of the simulated data sets.
-We estimated median relative bias of SSB across these simulations.
-See the Supplementary Materials for more details.
-
-\hypertarget{reference-points-and-projections}{%
-\subsection*{Reference points, and projections}\label{reference-points-and-projections}}
-\addcontentsline{toc}{subsection}{Reference points, and projections}
-
-We produced SPR-based BRP estimates (\(F_{40\%}\) and SSB\(_{40\%}\)) from the best performing model using a variety of options provided in WHAM.
-We generated annual BRP estimates based on the annual weight at age, maturity at age, natural mortality, fleet selectivity, and either the annual recruitment random effect (Eq.
-\ref{eq:Rec-re}) or the annual expected recruitment (Eq.
-\ref{eq:expected-recruitment}).
-We also calculated status as the ratios of annual average F and SSB to the corresponding reference points.
-Finally, we calculated ``prevailing'' BRPs using the average weight at age, maturity at age, M, fleet selectivity from the last 5 years (2017 to 2021) and average recruitment over years 2000 to 2021, and made Kobe plots of joint status for the terminal year of the model (2021).
-All BRPs and status estimates are generated for each region as well as for the total stock.
-
-We projected the best black sea bass model under three alternative scenarios for the bottom temperature covariate where 1) the AR1 time-series model continues into the projection years, 2) the average of the most recent 5 years is projected, and 3) bottom temperature increases in projection years following a prediction from a simple linear regression of the estimated bottom temperature anomalies over time.
-We specified fishing mortality in projection years to be constant at the value in the last model year (2021).
-Average weight, natural mortality, movement, and maturity at age over the last 5 years are used in projection years whereas the AR1 processes for all numbers at age random effects are continued.
-
-\hypertarget{results}{%
-\section*{Results}\label{results}}
-\addcontentsline{toc}{section}{Results}
-
-We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock and no random effects on age 1 natural mortality (\(M_1\), Tables \ref{tab:diff-aic-table} and \ref{tab:aic-wts-table}).
-Model \(M_1\) consistently had the lowest AIC across retrospective peels where the terminal years of data are removed from the model.
-The difference in AIC for the model that also included bottom temperature effects on recruitment for the southern component (\(M_3\)) suggested some evidence for this hypothesis as well.
-There was little evidence for also including random effects on age 1 natural mortality for the northern stock component.
-However, when more than three terminal years of data were removed, the evidence for adding age 1 natural mortality random effects for the northern component (\(M_8\)) increased is similar to adding bottom temperature effects on southern component recruitment (\(M_3\)).
-
-Comparisons of estimates from retrospective peels did not show any indication of differences in model performance.
-There was very little variation in measures of retrospective patterns for SSB or fishing mortality across models.
-Mohn's \(\rho \approx -0.03\) for SSB of both stock components and Mohn's \(\rho \approx 0.03\) and \(\approx -0.04\) for average \(F\) at ages 6 and 7 in north and south regions, respectively (Table \ref{tab:rho-table}).
-
-Estimated effects of bottom temperature on recruitment for either the northern or southern component were stable over the retrospective peels and differed negligibly whether effects for each component were estimated in isolation or together (Table \ref{tab:beta-sig-peel-table}).
-The estimates for the effects on each stock component were essentially equivalent whether the effect was on one component or both.
-Estimates of residual variability in northern component recruitment, as measured by the conditional or marginal standard deviation of the recruitment random effects, increased slightly with the number of peeled years.
-However, the ratio of standard deviations of models with and without temperature effects was stable with approximately 20\% reduction in residual standard deviation when temperature effects were included.
-The residual variation is reduced because the expected recruitment (Eq.
-\ref{eq:expected-recruitment}) is a function of the covariate (Figure \ref{fig:BT-Ecov-R}).
-
-The posterior estimates of the bottom temperature covariate match the observations well because of the high precision of the observed anomalies (Figure \ref{fig:bottom-temperature}).
-The anomalies for the north and south regions appear highly correlated presumably because of the proximity of the two regions.
-Because the temperature anomalies are treated as latent variables, other data components in the model can affect the estimated anomalies when effects on recruitment are included (e.g., Miller et al.
-2018), but in this case the estimates of anomalies are altered negligibly during the population model years when the effect on recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
-
-The effect of the anomalies on expected recruitment can be better observed by comparing the time series of the northern and southern components (Figure \ref{fig:E-R-SSB-F}).
-The expected recruitment for the southern component without any covariate effects is constant over time whereas the effect on expected recruitment for the northern component induces correlation of the bottom temperature anomalies and expected recruitment.
-For the northern stock component, most of the lowest estimated recruitment random effects occurred prior to 2000 and several large recruitment estimates occurred after 2010.
-For the southern stock, recruitment random effect estimates have been less variable and similar to the estimated median particularly since 2005.
-The estimated stock size of the northern component, as measured by SSB, increased substantially since 2005, whereas that for the southern component peaked in 2002.
-The increase in the northern component is also reflected in the total size of the combined components.
-Estimated average total fishing mortality rates across ages 6 and 7 (combining the commercial and recreational fleets) were highest in the southern region prior to 1998 and have declined substantially since then.
-Estimated fishing mortality rates in the northern region have been less variable, but had largest peaks in 1996, 1999, and 2005.
-The decline in estimates in the southern component is also reflected in the decline in total fishing mortality across regions.
-
-The estimated movement from the north to the south for the northern stock component is much less than anticipated from the prior distribution that was parameterized based on the companion Stock Synthesis model fit completed by the Working Group (Figure \ref{fig:move-prior-posterior}).
-Conversely, the estimated movement rate from south to north was more consistent and slightly greater than that from the companion model.
-The estimated movement rates varied little across all fitted models implying that the estimation of movement was not sensitive to the alternative assumptions we considered.
-Furthermore, the very low estimated movement rate from north to south is also consistent with preliminary fits of the WHAM model without the prior distribution where the movement rate was estimated at the lower bound of 0 resulting in a lack of convergence.
-The estimated movement rates imply that a very small fraction of northern fish are ever in the southern region.
-
-The 2DAR1 random effects for the selectivity parameters of the northern recreational fleet indicate a decrease in selectivity at ages 2 to 6 over the time series which corresponds to increases in the size limits in regulations by states where these catches occurred (Figure \ref{fig:selectivity-re}).
-The model also estimated a more modest decrease in selectivity at ages 2 to 3 for the northern commercial fleet.
-The selectivity at the youngest ages for the two fishery independent indices varied over time but without notable trend.
-
-\hypertarget{reference-points-1}{%
-\subsection*{Reference points}\label{reference-points-1}}
-\addcontentsline{toc}{subsection}{Reference points}
-
-Using the expected recruitment rather than the random effects as the multiplier with equilibrium spawning biomass per recruit (\(\upphi(\widetilde{F})\), eq.
-\ref{eq:ssbpr}) results in less variable estimates of annual SSB\(_{40\%}\) (Figure \ref{fig:annual-BRPs}: top row).
-Estimates were less variable for the southern stock component than the northern stock component mainly because the expected recruitment is constant for the former whereas that for the northern component varies via the effect of the bottom temperature anomalies.
-Temporal variation for the southern stock component is mainly caused by annual variation in the inputs to the \(\upphi(\widetilde{F})\) calculation and \(F_{40\%}\), but there is also some effect of the variation in the proportion of total recruitment for each stock component as used in the weighting terms to define the 40\% reduction in unfished spawning biomass per recruit (\(\lambda_s\) in Eq.
-\ref{eq:newton-obj}).
-Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB\(_{40\%}\) for the southern stock component whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
-
-Ratios of SSB to SSB\(_{40\%}\) should be viewed with caution because the SSB reference point is being defined using annual recruitments rather than some form of average recruitment and therefore, ratios larger than 2.5 which would correspond to unfished population sizes are possible (Figure \ref{fig:annual-BRPs}: second row).
-However, for the southern component, using the constant expected recruitment is comparable to using median recruitment and the annual variation in the inputs to \(\upphi(\widetilde{F})\) indicates no years where SSB was greater than unfished levels throughout the time series.
-The patterns in uncertainty of SSB ratios are the same as those for the SSB reference points (Figure \ref{fig:annual-SSB-status-cvs}).
-
-Differences between annual \(F_{40\%}\) estimates using the alternative types of recruitment estimates are much less than those for SSB\(_{40\%}\) because the recruitment estimates are only used in the weighting of the stock-specific \(\upphi(\widetilde{F})\) to determine \(F_{40\%}\) (\(\lambda_s\) in Eq.
-\ref{eq:newton-obj}, Figure \ref{fig:annual-BRPs}: third row).
-For similar reasons, estimates of uncertainty of \(F_{40\%}\) are also similar between the type of recruitment estimates (Figure \ref{fig:annual-F40-cvs}).
-Annual estimates for the northern stock component were lowest in the earliest years of the time series whereas those for the soutern component are relatively stable over time.
-
-For a given type of recruitment used to calculate \(F_{40\%}\), the annual ratios of average \(F\) to the corresponding reference point are identical for each stock component and in total because of the selectivity used to define a single F reference point across both regions is exactly the same as that for the annual estimate of fishing mortality at age by fleet (Figure \ref{fig:annual-BRPs}: fourth row).
-As for annual \(F_{40\%}\), estimates of uncertainty in the ratios are similar whether expected recruitment or the random effects are used to calculate the reference points (Figure \ref{fig:annual-F-status-cvs}).
-
-Unlike the annual ratios of \(F\) to \(F_{40\%}\), the \(F_{40\%}\) under prevailing conditions uses a selectivity defined by the average \(F\) at age and fleet over the last 5 years, which is different from that of the selectivity in the terminal year and therefore those ratios for each region and in total are different (Figure \ref{fig:annual-BRPs}: bottom row).
-Although the alternative types of recruitment estimates result in different points on the Kobe plots for joint status, the uncertainty in the estimates suggest similar probabilities of alternative quadrants.
-However, the usage of the confidence regions of the ratios to defined probabilities is an approximation because these ratios are functions of both empirical Bayes estimates of random effects as well as maximum likelihood estimates of fixed effects.
-
-\hypertarget{projections-1}{%
-\subsection*{Projections}\label{projections-1}}
-\addcontentsline{toc}{subsection}{Projections}
-
-When continuing the AR1 model for the bottom temperature anomalies in the projection years, the uncertainty grows and asymptotes to a level determined by the marginal variance of the process and the predicted value approaches the estimated marginal mean value of approximately 0 (Figure \ref{fig:R-BT-proj}).
-There is no uncertainty in the projected linear trend because the values were specified as known whereas there is a small level of uncertainty in the projections of the recent average based on the uncertainty in the estimated mean.
-Comparing projected recruitment for the northern and southern stock components, the projected expected recruitment has much lower uncertainty than the projected random effects regardless of whether bottom temperature affects recruitment.
-For the northern component, CVs of projected expected recruitment are smaller than those of recruitment random effect, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for the future bottom temperature anomalies are made (Figure \ref{fig:R-F-SSB-CVs}).
-
-Because projected fishing mortality is specified to be the same as that in the terminal year the same uncertainty is also propagated in the projections (Figure \ref{fig:F-SSB-proj}).
-As we would expect, there is no effect of alternative assumptions about bottom temperature in the projections for the southern stock component because the effect of the bottom temperature anomaly is only on the northern component.
-However, the alternative bottom temperature assumptions do affect projections of northern component SSB and the projections do not stabilize when the anomalies are projected to increase linearly.
-When the same AR1 assumption for the bottom temperature anomalies is used in projections, CVs of northern and total SSB projections are larger than when the recent average or linear trend are assumed (Figure \ref{fig:R-F-SSB-CVs}).
-
-Unlike the annual reference points during the model years (prior to 2022), components of the \(\upphi(\widetilde{F})\) calculations are constant in the projection years where the only changing annual values are the recruitments used in the weighting for \(F_{40\%}\) and the recruitments used to calculate the resulting SSB\(_{40\%}\) (Figure \ref{fig:brps-proj}).
-Estimated reference points in the first few projection years can differ depending on whether the expected recruitment or the random effects are used in their calculation, but longer term projected values are identical because the random effect estimates approach the expected values as the influence of observations diminishes.
-However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures \ref{fig:annual-F40-cvs} and \ref{fig:annual-SSB40-cvs}).
-The reduction in uncertainty using expected recruitment is less for \(F_{40\%}\) than SSB\(_{40\%}\) and for the northern component which includes bottom temperature effects than the southern component.
-
-\hypertarget{discussion}{%
-\section*{Discussion}\label{discussion}}
-\addcontentsline{toc}{section}{Discussion}
-
-\hypertarget{black-sea-bass}{%
-\subsection*{Black sea bass}\label{black-sea-bass}}
-\addcontentsline{toc}{subsection}{Black sea bass}
-
-In our investigation of bottom temperature effects on recruitment and natural mortality and time-varying random effects on natural mortality at age 1, we found only including effects on recruitment for the northern stock component to be the best model with respect to AIC.
-Black sea bass in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north (McMahan et al.
-2020).
-Opposite effects of temperature on recruitment or population size would be expected for species in the same general area that are at the southern extent of their range (Gabriel 1992).
-For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer fall on the Northeast US shelf (Sullivan et al.
-2005; Miller et al.
-2016b).
-
-Our finding regarding bottom temperature effects is consistent with the bottom temperature effects on recruitment in the assessment model accepted during the peer-review process and which is currently used to assess the stock (NEFSC 2023).
-However, AIC suggested some weight for models with temperature effects on recruitment for the southern component and random effects on M at age 1.
-Although this might suggest making inferences on the population by using an ensemble of these and weighting by AIC (Burnham and Anderson 2002), the estimates of assessment outputs relevant for management (e.g., SSB and \(F\)) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
-Nevertheless, continued monitoring of effects on the southern component may be prudent.
-
-We showed that using expected recruitment rather than recruitment random effects to define BRPs can produce the same estimates but with lower uncertainty.
-Although the option is unavailable currently in WHAM, we expect using expected bottom temperature anomalies rather than the random effects would also result in the same BRPs and even less uncertainty than those we showed using expected recruitment.
-These results can be important for management when uncertainty in assessment output is used in making catch advice.
-The typical approach for defining stock and harvest status is to compare BRPs defined under prevailing conditions with current stock and harvest levels, but we could also make the comparisons with projected BRPs like those we demonstrated are possible.
-The uncertainty in these ratios could also be used in defining catch advice.
-
-For black sea bass, it is evident that the projection methodology of the environmental covariate has an impact on short-term projection estimates of recruitment and SSB (Figure 6).
-In the projection period, the AR1 models for bottom temperature and recuitment revert to the long-term means of those processes.
-This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypothesis of increasing water temperature and changing environmental conditions (Hare 2016; Pershing et al.
-2021).
-Because of these concerns, the most recent stock assessment for yellowtail flounder used a change point analysis on environmental covariates to determine current conditions.
-The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections (NEFSC, 2024).
-Unfortunately, this methodology still has limitations because it is unlikely that current environmental conditions are stable.
-Similarly, a linear increase of temperature is also not likely in the future (see past trends, Figure 6 top row).As more assessments begin including effects of environmental covariates explicitly in assessment models, assumptions about these covariates in calculated BRPs and population projections becomes an important decision in the management process.
-
-\hypertarget{general-model}{%
-\subsection*{General model}\label{general-model}}
-\addcontentsline{toc}{subsection}{General model}
-
-WHAM provides a large class of models for time- and age-varying random effects for population attributes and treatment of environmental covariates and their effects on populations using state-space methods (Maunder 2024).
-This framework accounts for the magnitude of differing uncertainties in observations and stochastic population dynamics processes.
-Version 2.0 extends these inferences to movement rates for multi-region models and stock (component) specific configuration of random and environmental covariate effects.
-
-An obvious limitation to WHAM is the inability to include tagging observations of any type.
-Such observations are critical to estimation of movement parameters (e.g., Goethel et al.
-2019).
-We used estimates of movement parameters from a companion assessment model to define prior distributions for corresponding movement parameters in the state-space black sea bass model, but other than the tagging data, both of the assessment models used much the same data which can lead to inappropriate inferences.
-In lieu of integrating an observation model for the tagging data, a better approach would be to estimate the movement parameters externally from just the tagging data.
-
-Tagging data can also inform mortality rate parameters (Hampton 1991).
-It is well known that natural mortality is seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality from other assessment model parameters (Lee et al.
-2011; Clark 2022).
-Estimation of natural mortality may be even more challenging within state-space assessment models with their greater flexibility from inclusion of time-varying random effects.
-For example, we fixed the mean natural mortality for age 1 fish and Cadigan (2016) and Stock et al.
-(2021) also estimated natural mortality deviations.
-However, the simulation studies of Miller et al.
-(In review) suggest estimating natural mortality in WHAM is feasible in some situations with typical assessment data.
-Regardless, tagging observations can provide improved inferences and allowing tagging observations in WHAM should be a high priority even for models with a single stock and region.
-This extension would also be beneficial for black sea bass because the current assessment uses much the same data to estimate movement outside of the model as we are using to estimate the other parameters in the WHAM model.
-
-WHAM version 2.0 only allows models where each stock spawns and recruits in a single region however, some stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region (Kapur et al.
-2021).
-Therefore, extending WHAM to allow such models would be beneficial.
-Similarly age composition data are scarce for many stocks incorporating the work by Correa et al.
-(2023) on allowing length composition and modeling growth within WHAM would provide an assessment tool to include process errors in assessment with such data and possibly with multiple stocks and regions.
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-ablehales97}{}}%
-Able, K.W., and Stanton Hales, Jr., L.
-1997.
-Movements of juvenile black sea bass \emph{{C}entropristis striata} (linnaeus) in a southern {N}ew {J}ersey estuary.
-Journal of Experimental Marine Biology and Ecology \textbf{213}(2): 153--167.
-doi:\href{https://doi.org/10.1016/S0022-0981(96)02743-8}{10.1016/S0022-0981(96)02743-8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-andersenkeiding02}{}}%
-Andersen, P.K., and Keiding, N.
-2002.
-Multi-state models for event history analysis.
-Statistical Methods in Medical Research \textbf{11}: 91--115.
-doi:\href{https://doi.org/10.1191/0962280202SM276ra}{10.1191/0962280202SM276ra}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-arnason72}{}}%
-Arnason, A.N.
-1972.
-Parameter estimates from mark-recapture experiments on two populations subject to migration and death.
-Researches on Population Ecology \textbf{13}: 97--113.
-doi:\href{https://doi.org/10.1007/BF02521971}{10.1007/BF02521971}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-asmfc16}{}}%
-ASMFC. 2016. Black sea bass stock assessment. {Available} at http://www.asmfc.org/uploads/file/5953f11d2016BlackSeaBassStockAssmt.pdf.
-
-\leavevmode\vadjust pre{\hypertarget{ref-belletal15}{}}%
-Bell, R.J., Richardson, D.E., Hare, J.A., Lynch, P.D., and Fratantoni, P.S.
-2015.
-Disentangling the effects of climate, abundance, and size on the distribution of marine fish: An example based on four stocks from the {N}ortheast {US} shelf.
-ICES Journal of Marine Science \textbf{72}(5): 1311--1322.
-doi:\href{https://doi.org/10.1093/icesjms/fsu217}{10.1093/icesjms/fsu217}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bosleyetal22}{}}%
-Bosley, K.M., Schueller, A.M., Goethel, D.R., Hanselman, D.H., Fenske, K.H., Berger, A.M., Deroba, J.J., and Langseth, B.J.
-2022.
-Finding the perfect mismatch: Evaluating misspecification of population structure within spatially explicit integrated population models.
-Fish and Fisheries \textbf{23}(2): 294--315.
-doi:\href{https://doi.org/10.1111/faf.12616}{10.1111/faf.12616}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-brooks24}{}}%
-Brooks, E.N.
-2024.
-Pragmatic approaches to modeling recruitment in fisheries stock assessment: A perspective.
-Fisheries Research \textbf{270}: 106896.
-doi:\href{https://doi.org/10.1016/j.fishres.2023.106896}{10.1016/j.fishres.2023.106896}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-burnhamanderson02}{}}%
-Burnham, K.P., and Anderson, D.R. 2002. Model selection and multimodel inference: A practical information-theoretic approach. Springer-Verlag, New York.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cadigan16}{}}%
-Cadigan, N.G.
-2016.
-A state-space stock assessment model for northern cod, including under-reported catches and variable natural mortality rates.
-Canadian Journal of Fisheries and Aquatic Sciences \textbf{73}(2): 296--308.
-doi:\href{https://doi.org/10.1139/cjfas-2015-0047}{10.1139/cjfas-2015-0047}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-caoetal14}{}}%
-Cao, J., Truesdell, S.B., and Chen, Y.
-2014.
-Impacts of seasonal stock mixing on the assessment of {A}tlantic cod in the {G}ulf of {M}aine.
-ICES Journal of Marine Science \textbf{71}(6): 1443--1457.
-doi:\href{https://doi.org/10.1093/icesjms/fsu066}{10.1093/icesjms/fsu066}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-clark22}{}}%
-Clark, W.G.
-2022.
-Why natural mortality is estimable, in theory if not in practice, in a data-rich stock assessment.
-Fisheries Research \textbf{248}: 106203.
-doi:\href{https://doi.org/10.1016/j.fishres.2021.106203}{10.1016/j.fishres.2021.106203}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-commenges99}{}}%
-Commenges, D. 1999. Multi-state models in epidemiology. Lifetime Data Analysis \textbf{5}: 315--327. doi:\href{https://doi.org/10.1023/A:1009636125294}{10.1023/A:1009636125294}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-correaetal23}{}}%
-Correa, G.M., Monnahan, C.C., Sullivan, J.Y., Thorson, J.T., and Punt, A.E.
-2023.
-Modelling time-varying growth in state-space stock assessments.
-ICES Journal of Marine Science \textbf{80}(7): 2036--2049.
-doi:\href{https://doi.org/10.1093/icesjms/fsad133}{10.1093/icesjms/fsad133}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-fabrizioetal13}{}}%
-Fabrizio, M.C., Manderson, J.P., and Pessutti, J.P.
-2013.
-Habitat associations and dispersal of black sea bass from a {M}id-{A}tlantic {B}ight reef.
-Marine Ecology Progress Series \textbf{482}: 241--253.
-doi:\href{https://doi.org/10.3354/meps10302}{10.3354/meps10302}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-francis14}{}}%
-Francis, R.I.C.C.
-2014.
-Replacing the multinomial in stock assessment models: A first step.
-Fisheries Research \textbf{151}: 70--84.
-doi:\href{https://doi.org/10.1016/j.fishres.2013.12.015}{10.1016/j.fishres.2013.12.015}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gabriel92}{}}%
-Gabriel, W.L.
-1992.
-Persistence of demersal fish assemblages between {C}ape {H}atteras and {N}ova {S}cotia, {N}orthwest {A}tlantic.
-Journal of the Northwest Atlantic Fisheries Science \textbf{14}(1): 29--46.
-
-\leavevmode\vadjust pre{\hypertarget{ref-goetheletal19}{}}%
-Goethel, D.R., Bosley, K.M., Hanselman, D.H., Berger, A.M., Deroba, J.J., Langseth, B.J., and Schueller, A.M.
-2019.
-Exploring the utility of different tag-recovery experimental designs for use in spatially explicit, tag-integrated stock assessment models.
-Fisheries Research \textbf{219}: 105320.
-doi:\href{https://doi.org/10.1016/j.fishres.2019.105320}{10.1016/j.fishres.2019.105320}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hampton91}{}}%
-Hampton, J. 1991. Estimation of southern bluefin tuna \emph{{T}hunnus maccoyii} natural mortality and movement rates from tagging experiments. Fishery Bulletin \textbf{89}(4): 591--610.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hareetal16}{}}%
-Hare, W.E.A.N., Jonathan A.
-AND Morrison.
-2016.
-A vulnerability assessment of fish and invertebrates to climate change on the northeast u.s.
-Continental shelf.
-PLOS ONE \textbf{11}: 1--30.
-doi:\href{https://doi.org/10.1371/journal.pone.0146756}{10.1371/journal.pone.0146756}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hearnetal87}{}}%
-Hearn, W.S., Sundland, R.L., and Hampton, J.
-1987.
-Robust estimation of the natural mortality rate in a completed tagging experiment with variable fishing intensity.
-Journal Du Conseil International Pour L'exploration De La Mer \textbf{43}: 107--117.
-doi:\href{https://doi.org/10.1093/icesjms/43.2.107}{10.1093/icesjms/43.2.107}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kapuretal21}{}}%
-Kapur, M.S., Siple, M.C., Olmos, M., Privitera-Johnson, K.M., Adams, G., Best, J., Castillo-Jordán, C., Cronin-Fine, L., Havron, A.M., Lee, Q., Methot, R.D., and Punt, A.E.
-2021.
-Equilibrium reference point calculations for the next generation of spatial assessments.
-Fisheries Research \textbf{244}: 106132.
-doi:\href{https://doi.org/10.1016/j.fishres.2021.106132}{10.1016/j.fishres.2021.106132}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kemenysnell60}{}}%
-Kemeny, J.G., and Snell, J.L. 1960. Finite markov chains. D. Van Nostrand Company, Princeton, New Jersey.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kristensenetal16}{}}%
-Kristensen, K., Nielsen, A., Berg, C., Skaug, H., and Bell, B.M.
-2016.
-{TMB}: Automatic differentiation and {Laplace} approximation.
-Journal of Statistical Software \textbf{70}: 1--21.
-doi:\href{https://doi.org/10.18637/jss.v070.i05}{10.18637/jss.v070.i05}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-leeetal11}{}}%
-Lee, H.-H., Maunder, M.N., Piner, K.R., and Methot, R.D.
-2011.
-Estimating natural mortality within a fisheries stock assessment model: An evaluation using simulation analysis based on twelve stock assessments.
-Fisheries Research \textbf{109}(1): 89--94.
-doi:\href{https://doi.org/10.1016/j.fishres.2011.01.021}{10.1016/j.fishres.2011.01.021}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-legaultrestrepo99}{}}%
-Legault, C.M., and Restrepo, V.R. 1999. A flexible forward age-structured assessment program. Col. Vol. Sci. Pap. ICCAT \textbf{49}(2): 246--253.
-
-\leavevmode\vadjust pre{\hypertarget{ref-maunder24}{}}%
-Maunder, M.N.
-2024.
-Towards a comprehensive framework for providing management advice from statistical inference using population dynamics models.
-Ecological Modelling \textbf{498}: 110836.
-doi:\href{https://doi.org/10.1016/j.ecolmodel.2024.110836}{10.1016/j.ecolmodel.2024.110836}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mcmahanetal20}{}}%
-McMahan, M.D., Sherwood, G.D., and Grabowski, J.H.
-2020.
-Geographic variation in life-history traits of black sea bass (\emph{{C}entropristis striata}) during a rapid range expansion.
-Frontiers in Marine Science \textbf{7}: 567758.
-doi:\href{https://doi.org/10.3389/fmars.2020.567758}{10.3389/fmars.2020.567758}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-methotwetzel13}{}}%
-Methot, R.D., and Wetzel, C.R.
-2013.
-Stock {S}ynthesis: A biological and statistical framework for fish stock assessment and fishery management.
-Fisheries Research \textbf{142}(1): 86--99.
-doi:\href{https://doi.org/10.1016/j.fishres.2012.10.012}{10.1016/j.fishres.2012.10.012}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-milleretal16_yoy_survival}{}}%
-Miller, A.S., Shepherd, G.R., and Fratantoni, P.S.
-2016a.
-Offshore habitat preference of overwintering juvenile and adult black sea bass, \emph{{C}entropristis} \emph{striata}, and the relationship to year-class success.
-{PLOS} {ONE} \textbf{11}(1): e0147627.
-doi:\href{https://doi.org/10.1371/journal.pone.0147627}{10.1371/journal.pone.0147627}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-millerandersen08}{}}%
-Miller, T.J., and Andersen, P.K.
-2008.
-A finite-state continuous-time approach for inferring regional migration and mortality rates from archival tagging and conventional tag-recovery experiments.
-Biometrics \textbf{64}(4): 1196--1206.
-doi:\href{https://doi.org/10.1111/j.1541-0420.2008.00996.x}{10.1111/j.1541-0420.2008.00996.x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-milleretalinreview1}{}}%
-Miller, T.J., Britten, G., Brooks, E.N., Fay, G., Hansell, A., Legault, C.M., Li, C., Muffley, B., Stock, B.C., and Wiedenmann, J.
-In review.
-Factors affecting reliablity of state-space age-structured assessment models.
-Canadian Journal of Fisheries Science.
-
-\leavevmode\vadjust pre{\hypertarget{ref-milleretal16}{}}%
-Miller, T.J., Hare, J.A., and Alade, L.A.
-2016b.
-A state-space approach to incorporating environmental effects on recruitment in an age-structured assessment model with an application to southern {New England} yellowtail flounder.
-Canadian Journal of Fisheries Science \textbf{73}(8): 1261--1270.
-doi:\href{https://doi.org/10.1139/cjfas-2015-0339}{10.1139/cjfas-2015-0339}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-milleretal18}{}}%
-Miller, T.J., O'Brien, L., and Fratantoni, P.S.
-2018.
-Temporal and environmental variation in growth and maturity and effects on management reference points of {Georges Bank Atlantic} cod.
-Canadian Journal of Fisheries Science \textbf{75}(12): 2159--2171.
-doi:\href{https://doi.org/10.1139/cjfas-2017-0124}{10.1139/cjfas-2017-0124}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-millerstock20}{}}%
-Miller, T.J., and Stock, B.C. 2020. The {Woods Hole Assessment Model} ({WHAM}). https://timjmiller.github.io/wham.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mohn99}{}}%
-Mohn, R.
-1999.
-The retrospective problem in sequential population analysis: An investigation using cod fishery and simulated data.
-ICES Journal of Marine Science \textbf{56}(4): 473--488.
-doi:\href{https://doi.org/10.1006/jmsc.1999.0481}{10.1006/jmsc.1999.0481}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mosershepherd09}{}}%
-Moser, J., and Shepherd, G.R.
-2009.
-Seasonal distribution and movement of black sea bass (\emph{{C}entropristis striata}) in the {N}orthwest {A}tlantic as determined from a mark-recapture experiment.
-Journal of Northwest Atlantic Fishery Science \textbf{40}: 17--28.
-
-\leavevmode\vadjust pre{\hypertarget{ref-musickmercer77}{}}%
-Musick, J.A., and Mercer, L.P.
-1977.
-Seasonal distribution of black sea bass, \emph{{C}entropristis striata}, in the {M}id-{A}tlantic {B}ight with comments on the ecology and fisheries of the species.
-Transactions of the American Fisheries Society \textbf{106}(1): 12--25.
-doi:\href{https://doi.org/10.1577/1548-8659(1977)106\%3C12:SDOBSB\%3E2.0.CO;2}{10.1577/1548-8659(1977)106\textless12:SDOBSB\textgreater2.0.CO;2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-myers98}{}}%
-Myers, R.A.
-1998.
-When do environment--recruitment correlations work?
-Reviews in Fish Biology and Fisheries \textbf{8}(3): 229--249.
-doi:\href{https://doi.org/10.1023/A:1008828730759}{10.1023/A:1008828730759}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nefsc22}{}}%
-NEFSC.
-2022a.
-Final report of the haddock research track assessment working group.
-{Available} at https://s3.us-east-1.amazonaws.com/nefmc.org/14b\_EGB\_Research\_Track\_Haddock\_WG\_Report\_DRAFT.pdf.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nefsc22a}{}}%
-NEFSC. 2022b. Report of the american plaice research track working group. {Available} at https://s3.us-east-1.amazonaws.com/nefmc.org/2\_American-Plaice-WG-Report.pdf.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nefsc23}{}}%
-NEFSC.
-2023.
-Report of the black sea bass (\emph{{C}entropristis} \emph{striata}) research track stock assessment working group.
-{Available} at https://www.mafmc.org/s/a\_2023\_BSB\_UNIT\_RTWG\_Report\_V2\_12\_2\_2023-1.pdf.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nefsc24}{}}%
-NEFSC. 2024. Butterfish research track assessment report. US Dept Commer Northeast Fish Sci Cent Ref Doc. 24-03; 191 p.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nielsenberg14}{}}%
-Nielsen, A., and Berg, C.W.
-2014.
-Estimation of time-varying selectivity in stock assessments using state-space models.
-Fisheries Research \textbf{158}: 96--101.
-doi:\href{https://doi.org/10.1016/j.fishres.2014.01.014}{10.1016/j.fishres.2014.01.014}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-perreaultetal20}{}}%
-Perreault, A.M.J., Wheeland, L.J., Morgan, M.J., and Cadigan, N.G.
-2020.
-A state-space stock assessment model for {American} plaice on the {Grand} {Bank} of {Newfoundland}.
-Journal of Northwest Atlantic Fishery Science \textbf{51}: 45--104.
-doi:\href{https://doi.org/10.2960/j.v51.m727}{10.2960/j.v51.m727}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pershingetal21}{}}%
-Pershing, A.J., Alexander, M.A., Brady, D.C., Brickman, D., Curchitser, E.N., Diamond, A.W., McClenachan, L., Mills, K.E., Nichols, O.C., Pendleton, D.E., Record, N.R., Scott, J.D., Staudinger, M.D., and Wang, Y.
-2021.
-Climate impacts on the {G}ulf of {M}aine ecosystem: A review of observed and expected changes in 2050 from rising temperatures.
-Elementa: Science of the Anthropocene \textbf{9}(1): 00076.
-doi:\href{https://doi.org/10.1525/elementa.2020.00076}{10.1525/elementa.2020.00076}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-puntetal20}{}}%
-Punt, A.E., Dunn, A., Elvarsson, B., Hampton, J., Hoyle, S.D., Maunder, M.N., Methot, R.D., and Nielsen, A.
-2020.
-Essential features of the next-generation integrated fisheries stock assessment package: A perspective.
-Fisheries Research \textbf{229}: 105617.
-doi:\href{https://doi.org/10.1016/j.fishres.2020.105617}{10.1016/j.fishres.2020.105617}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-schwarzetal93}{}}%
-Schwarz, C.J., Schweigert, J.F., and Arnason, A.N.
-1993.
-Estimating migration rates using tag-recovery data.
-Biometrics \textbf{49}: 177--193.
-doi:\href{https://doi.org/10.2307/2532612}{10.2307/2532612}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-stockmiller21}{}}%
-Stock, B.C., and Miller, T.J.
-2021.
-The {Woods Hole Assessment Model} ({WHAM}): A general state-space assessment framework that incorporates time- and age-varying processes via random effects and links to environmental covariates.
-Fisheries Research \textbf{240}: 105967.
-doi:\href{https://doi.org/10.1016/j.fishres.2021.105967}{10.1016/j.fishres.2021.105967}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-stocketal21}{}}%
-Stock, B.C., Xu, H., Miller, T.J., Thorson, J.T., and Nye, J.A.
-2021.
-Implementing two-dimensional autocorrelation in either survival or natural mortality improves a state-space assessment model for {Southern New England}-{Mid Atlantic} yellowtail flounder.
-Fisheries Research \textbf{237}: 105873.
-doi:\href{https://doi.org/10.1016/j.fishres.2021.105873}{10.1016/j.fishres.2021.105873}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sullivanetal05}{}}%
-Sullivan, M.C., Cowen, R.K., and Steves, B.P.
-2005.
-Evidence for atmosphere{--}ocean forcing of yellowtail flounder (\emph{{L}imanda} \emph{ferruginea}) recruitment in the {M}iddle {A}tlantic {B}ight.
-Fisheries Oceanography \textbf{14}(5): 386--399.
-doi:\href{https://doi.org/10.1111/j.1365-2419.2005.00343.x}{10.1111/j.1365-2419.2005.00343.x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-thorsonetal17}{}}%
-Thorson, J.T., Johnson, K.F., Methot, R.D., and Taylor, I.G.
-2017.
-Model-based estimates of effective sample size in stock assessment models using the {D}irichlet-multinomial distribution.
-Fisheries Research \textbf{192}: 84--93.
-doi:\href{https://doi.org/10.1016/j.fishres.2016.06.005}{10.1016/j.fishres.2016.06.005}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-thorsonetal23}{}}%
-Thorson, J.T., Miller, T.J., and Stock, B.C.
-2023.
-The multivariate-tweedie: A self-weighting likelihood for age and length composition data arising from hierarchical sampling designs.
-ICES Journal of Marine Science \textbf{80}(10): 2630--2641.
-doi:\href{https://doi.org/10.1093/icesjms/fsac159}{10.1093/icesjms/fsac159}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-varkeyetal22}{}}%
-Varkey, D.A., Babyn, J., Regular, P., Ings, D.W., Kumar, R., Rogers, B., Champagnat, J., and Morgan, M.J.
-2022.
-A state-space model for stock assessment of cod (gadus morhua) stock in NAFO subdivision 3Ps.
-{DFO} {Can.} {Sci.} {Advis.} {Sec.} {Res.} {Doc.} 2022/022.
-v + 78 p.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yingetal11}{}}%
-Ying, Y., Chen, Y., Lin, L., and Gao, T.
-2011.
-Risks of ignoring fish population spatial structure in fisheries management.
-Canadian Journal of Fisheries and Aquatic Sciences \textbf{68}(12): 2101--2120.
-doi:\href{https://doi.org/10.1139/f2011-116}{10.1139/f2011-116}.
-
-\end{CSLReferences}
-
-\pagebreak
-
-\pagebreak
-
-\hypertarget{figures}{%
-\section*{Figures}\label{figures}}
-\addcontentsline{toc}{section}{Figures}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{map}
-
-}
-
-\caption{The Northwest Atlantic coastal area where the NEUS black sea bass stock occurs.
-Red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations (after Miller et al.
-2016a) and the bathymetric contour is the 400 m isocline.}\label{fig:map}
-\end{figure}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.8\linewidth]{bsb_movement_diagram}
-
-}
-
-\caption{Diagram of intervals within the year and configuation of the dynamics of each component of the BSB population.}\label{fig:migration-diagram}
-\end{figure}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{move_prior_post}
-
-}
-
-\caption{Prior (black) and posterior (red) distributions of movement of northern component stock from north to south (top) and south to north (bottom).}\label{fig:move-prior-posterior}
-\end{figure}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{E_R_SSB_F_fig}
-
-}
-
-\caption{Annual observations and posterior estimates of bottom temperature anomalies (top row), annual expected and random effect recruitment estimates (second row) and SSB and average fishing mortality at ages 6 and 7 from model $M_1$.
-Columns define estimates regionally (spawning region for SSB) and toal SSB and fishing mortality across regions.
-Polygons and vertical lines represent 95\% confidence intervals.}\label{fig:E-R-SSB-F}
-\end{figure}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{selectivity_re_plot}
-
-}
-
-\caption{Time and age-varying selectivty for fleets and indices in the northern region with autoregressive random effects.}\label{fig:selectivity-re}
-\end{figure}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.85\textheight]{brp_status_results}
-
-}
-
-\caption{Estimates of average equilibrium $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio, equilibrium SSB at $F_{40\%}$ based on annual inputs to $\upphi(\widetilde{F})$ calculations, annual fishing and biomass status (ratios), and bivariate kobe plots of status in 2021 where reference points represent prevailing conditions with inputs to $\upphi(\widetilde{F})$ averaged over the last 5 years and recruitment is averaged over the time series.
-Red and blue in first 4 rows and black and gray in final row indicate alternative annual estimates of recruitment (random effect or expected).
-All results are based on model $M_1$.
-Polygons represent 95\% confidence intervals or regions.}\label{fig:annual-BRPs}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_ecov_Recruit_results}
-
-}
-
-\caption{Annual estimates of bottom temperature in the northern region and alternative recruitment estimates (random effect or expected) by region.
-Estimates from years after 2021 are from projections of model $M1$ under three alternative assumptions for the bottom temperature anomalies in the northern region.
-Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:R-BT-proj}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_F_SSB}
-
-}
-
-\caption{Annual estimates of average fishing mortality and SSB by region and in total.
-Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:F-SSB-proj}
-\end{figure}
-
-\begin{landscape}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_brps}
-
-}
-
-\caption{Annual estimates by region and in total of average equilibrium $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio and SSB at $F_{40\%}$.
-Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region and average $\upphi(\widetilde{F})$ inputs from the last 5 years of the unprojected model.
-Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:brps-proj}
-\end{figure}
-
-\end{landscape}
-
-\clearpage
-
-\setcounter{table}{0}
-\renewcommand\thetable{\arabic{table}}
-
-\hypertarget{tables}{%
-\section*{Tables}\label{tables}}
-\addcontentsline{toc}{section}{Tables}
-
-\begin{table}[!h]
-\centering
-\caption{\label{tab:model-desc-table}Assumptions for temperature effects and random effects for age 1 natural mortality for each model.}
-\centering
-\begin{tabular}[t]{llll}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Temperature Effect} & \multicolumn{1}{c}{ } \\
-\cmidrule(l{3pt}r{3pt}){2-3}
-Model & North & South & $M$ at age 1 random effects\\
-\midrule
-$M_{0}$ & -- & -- & none\\
-$M_{1}$ & Recruitment & -- & none\\
-$M_{2}$ & -- & Recruitment & none\\
-$M_{3}$ & Recruitment & Recruitment & none\\
-$M_{4}$ & $M$ at age 1 & -- & none\\
-\addlinespace
-$M_{5}$ & -- & $M$ at age 1 & none\\
-$M_{6}$ & $M$ at age 1 & $M$ at age 1 & none\\
-$M_{7}$ & -- & -- & time-varying\\
-$M_{8}$ & Recruitment & -- & time-varying\\
-$M_{9}$ & -- & Recruitment & time-varying\\
-\addlinespace
-$M_{10}$ & Recruitment & Recruitment & time-varying\\
-$M_{11}$ & $M$ at age 1 & -- & time-varying\\
-$M_{12}$ & -- & $M$ at age 1 & time-varying\\
-$M_{13}$ & $M$ at age 1 & $M$ at age 1 & time-varying\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:diff-aic-table}Difference between AIC and the lowest AIC for each model by retrospective peel.}
-\centering
-\begin{tabular}[t]{lrrrrrrrr}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Peel} \\
-\cmidrule(l{3pt}r{3pt}){2-9}
-Model & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
-\midrule
-$M_{0}$ & 11.83 & 11.39 & 10.41 & 10.05 & 9.86 & 9.41 & 8.47 & 8.22\\
-$M_{1}$ & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00\\
-$M_{2}$ & 12.63 & 12.03 & 11.41 & 11.06 & 10.91 & 10.34 & 9.20 & 9.22\\
-$M_{3}$ & 0.80 & 0.64 & 1.00 & 1.01 & 1.05 & 0.93 & 0.73 & 1.00\\
-$M_{4}$ & 13.81 & 13.35 & 12.26 & 11.82 & 11.71 & 11.18 & 10.08 & 9.85\\
-\addlinespace
-$M_{5}$ & 13.25 & 12.55 & 11.68 & 11.21 & 11.07 & 10.41 & 9.86 & 9.71\\
-$M_{6}$ & 15.22 & 14.51 & 13.52 & 12.97 & 12.91 & 12.17 & 11.46 & 11.33\\
-$M_{7}$ & 14.32 & 13.75 & 12.17 & 11.30 & 11.43 & 10.55 & 9.71 & 9.42\\
-$M_{8}$ & 2.25 & 2.10 & 1.47 & 0.96 & 1.29 & 0.91 & 0.98 & 0.99\\
-$M_{9}$ & 15.12 & 14.39 & 13.17 & 12.31 & 12.48 & 11.49 & 10.44 & 10.43\\
-\addlinespace
-$M_{10}$ & 3.05 & 2.74 & 2.47 & 1.97 & 2.34 & 1.84 & 1.71 & 1.99\\
-$M_{11}$ & 16.29 & 15.74 & 14.17 & 13.30 & 13.43 & 12.55 & 11.68 & 11.39\\
-$M_{12}$ & 15.73 & 14.91 & 13.43 & 12.46 & 12.64 & 11.55 & 11.10 & 10.92\\
-$M_{13}$ & 17.70 & 16.91 & 15.43 & 14.46 & 14.64 & 13.55 & 13.07 & 12.88\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:rho-table}Mohn's $\rho$ for SSB, and average F at ages 6 and 7 in northern and southern regions.}
-\centering
-\begin{tabular}[t]{lrrrr}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{SSB} & \multicolumn{2}{c}{Average F} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-Model & North & South & North & South\\
-\midrule
-$M_{0}$ & -0.040 & -0.023 & 0.041 & -0.048\\
-$M_{1}$ & -0.040 & -0.023 & 0.041 & -0.048\\
-$M_{2}$ & -0.040 & -0.024 & 0.041 & -0.047\\
-$M_{3}$ & -0.040 & -0.024 & 0.041 & -0.047\\
-$M_{4}$ & -0.041 & -0.022 & 0.042 & -0.048\\
-\addlinespace
-$M_{5}$ & -0.041 & -0.023 & 0.042 & -0.048\\
-$M_{6}$ & -0.042 & -0.023 & 0.042 & -0.048\\
-$M_{7}$ & -0.042 & -0.022 & 0.043 & -0.048\\
-$M_{8}$ & -0.044 & -0.023 & 0.044 & -0.048\\
-$M_{9}$ & -0.042 & -0.024 & 0.043 & -0.048\\
-\addlinespace
-$M_{10}$ & -0.044 & -0.024 & 0.044 & -0.047\\
-$M_{11}$ & -0.041 & -0.022 & 0.042 & -0.048\\
-$M_{12}$ & -0.042 & -0.023 & 0.043 & -0.048\\
-$M_{13}$ & -0.041 & -0.023 & 0.042 & -0.048\\
-\bottomrule
-\end{tabular}
-\end{table}
-\begin{table}
-
-\caption{\label{tab:beta-sig-peel-table}Estimates of temperature effects on recruitment and variance and autocorrelation parameters for recruitment for northern ($N$) and southern ($S$) components for models with no effects ($M_0$), effects on specific components ($M_1$ and $M_2$) or both components simultaneously ($M_3$) from the full model (Peel 0) and each retrospective peel.}
-\centering
-\begin{tabular}[t]{lrrrrrrrr}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{8}{c}{Peel} \\
-\cmidrule(l{3pt}r{3pt}){2-9}
-Parameter & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
-\midrule
-$M_{1}$ $\widehat{\beta}_{R,N}$ & 0.474 & 0.480 & 0.485 & 0.476 & 0.464 & 0.468 & 0.439 & 0.445\\
-$M_{2}$ $\widehat{\beta}_{R,S}$ & 0.099 & 0.105 & 0.094 & 0.095 & 0.092 & 0.096 & 0.099 & 0.091\\
-$M_{3}$ $\widehat{\beta}_{R,N}$ & 0.474 & 0.480 & 0.485 & 0.476 & 0.464 & 0.468 & 0.439 & 0.445\\
-$M_{3}$ $\widehat{\beta}_{R,S}$ & 0.099 & 0.105 & 0.094 & 0.095 & 0.092 & 0.096 & 0.099 & 0.091\\
-$M_{0}$ Conditional $\widehat{\sigma}_{R,N}$ & 0.925 & 0.953 & 0.978 & 0.988 & 0.978 & 1.010 & 0.981 & 1.007\\
-\addlinespace
-$M_{1}$ Conditional $\widehat{\sigma}_{R,N}$ & 0.730 & 0.752 & 0.780 & 0.786 & 0.775 & 0.801 & 0.785 & 0.800\\
-$M_{0}$ $\widehat{\rho}_{R,N}$ & 0.362 & 0.375 & 0.385 & 0.405 & 0.424 & 0.432 & 0.427 & 0.431\\
-$M_{1}$ $\widehat{\rho}_{R,N}$ & 0.296 & 0.307 & 0.334 & 0.373 & 0.394 & 0.406 & 0.428 & 0.429\\
-$M_{0}$ Marginal $\widehat{\sigma}_{R,N}$ & 0.992 & 1.028 & 1.060 & 1.080 & 1.081 & 1.120 & 1.084 & 1.117\\
-$M_{1}$ Marginal $\widehat{\sigma}_{R,N}$ & 0.764 & 0.791 & 0.827 & 0.848 & 0.843 & 0.877 & 0.869 & 0.885\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\clearpage
-
-\hypertarget{appendix-a}{%
-\section*{Appendix A}\label{appendix-a}}
-\addcontentsline{toc}{section}{Appendix A}
-
-\setcounter{table}{0}
-\renewcommand\thetable{A\arabic{table}}
-\input{symboltable.tex}
-
-\pagebreak
-
-\hypertarget{supplemental-materials}{%
-\section*{Supplemental Materials}\label{supplemental-materials}}
-\addcontentsline{toc}{section}{Supplemental Materials}
 
@@
 Many of the fits to the simulated data produced implausible estimates at the 0 boundary for these parameters (very negative values on log-scale).
-Therefore, we completed self-tests with these parameters fixed at the true values.
+However, across all fits including those with poor convergence, estimation of SSB and fishing mortality was reliable (Figure \ref{fig:self-test-fig}).
+We also fit analogous models with the multiplier parameters fixed at the true values, which did improve convergence, but larger bias was estimated for fishing mortality and SSB for the northern component.
 
-For 7 of the the simulated data sets the model failed to optimize.
-The maximum absolute gradient was \(<10^{-6}\) for only 9 and \(<10^{-4}\) for 52 of the 93 successfully fitted models.
-The poor convergence appeared to be attributable to the estimation of the scalar for the standard errors of the log-transformed northern Recreational CPA index for which estimates tended to 0 for nearly all of the fits (\(<0.01\) for 83 fits).
-However, even across all fits including those with poor convergence, the SSB estimates appeared to be reliable (Figure \ref{fig:self-test-fig}).
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-dupontaviceetal23}{}}%
+du Pontavice, H., Chen, Z., and Saba, V.S.
+2023.
+A high-resolution ocean bottom temperature product for the northeast {U}.{S}.
+Continental shelf marine ecosystem.
+Progress in Oceanography \textbf{210}: 102948.
+doi:\href{https://doi.org/10.1016/j.pocean.2022.102948}{10.1016/j.pocean.2022.102948}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-methotwetzel13}{}}%
+Methot, R.D., and Wetzel, C.R.
+2013.
+Stock {S}ynthesis: A biological and statistical framework for fish stock assessment and fishery management.
+Fisheries Research \textbf{142}(1): 86--99.
+doi:\href{https://doi.org/10.1016/j.fishres.2012.10.012}{10.1016/j.fishres.2012.10.012}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nefsc23}{}}%
+NEFSC.
+2023.
+Report of the black sea bass (\emph{{C}entropristis} \emph{striata}) research track stock assessment working group.
+{Available} at https://www.mafmc.org/s/a\_2023\_BSB\_UNIT\_RTWG\_Report\_V2\_12\_2\_2023-1.pdf.
+
+\end{CSLReferences}
+
+\pagebreak
 
@@
 
+{\centering \includegraphics[width=1\linewidth]{self_test_results}
+
+}
+
+\caption{Median relative error of SSB (Total and by stock component) and total fully-selected fishing mortality for estimation models fitted to simulated data from model $M_1$ where the observation variance of log-indices are fixed and estimated.
+Black and Red dashed lines represent the median of the annual medians and the median across all annual relative errors, respectively.
+Vertical lines represent 95\% confidence intervals.}\label{fig:self-test-fig}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{best_R_Ecov}
+
+}
+
+\caption{Expected and random effect recruitment estimates for the northern stock component. Color of points defined by the corresponding annual bottom temperature anomaly.}\label{fig:BT-Ecov-R}
+\end{figure}
+
+\begin{landscape}
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{BTA_full_fig}
+
+}
+
+\caption{Observations with 95\% confidence intervals (points with vertical lines) and posterior estimates with 95\% confidence intervals (lines with polygons) of bottom temperature anomalies in the north and south regions from model $M_1$.
+Gray vertical line defines the first year that the black sea bass stock is modeled.}\label{fig:bottom-temperature}
+\end{figure}
+\end{landscape}
+\pagebreak
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{Ecov_M1_rel_M0}
+
+}
+
+\caption{Relative differences in posterior estimates of northern region bottom temperature anomalies ($\widehat X$) from the null model without effects on recruitment ($M_0$) and with effects on the northern stock component ($M_1$).}\label{fig:Ecov-M1-rel-M0}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{selectivity_south_plot}
+
+}
+
+\caption{Selectivty for fleets and indices in the southern region.}\label{fig:selectivity-south}
+\end{figure}
+
+\begin{landscape}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{SSB_F_sensitivity_plots}
+
+}
+
+\caption{Estimates of annual SSB and fishing mortality rates for the best performing model $M_1$ and models that are otherwise the same except where  1) movement rates are fixed at the means for the prior distribution, 2) a constant natural mortality rate is estimated, or 3) there is no movement for either stock component.}\label{fig:sensitivity-plots}
+\end{figure}
+\end{landscape}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_SSB40_CV}
+
+}
+
+\caption{Coefficients of variation for annual equilibrium SSB$_{40\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations and alternative annual recruitment types.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data.}\label{fig:annual-SSB40-cvs}
+\end{figure}
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_SSB_status_CV}
+
+}
+
+\caption{Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data.}\label{fig:annual-SSB-status-cvs}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_F40_CV}
+
+}
+
+\caption{Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data.}\label{fig:annual-F40-cvs}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.95\textheight]{proj_F_status_CV}
+
+}
+
+\caption{Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\bar{F}_{40\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
+Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data.}\label{fig:annual-F-status-cvs}
+\end{figure}
+
+\begin{landscape}
+
+\begin{figure}
+
+{\centering \includegraphics[height=0.9\textheight]{R_SSB_F_cv_results}
+
+}
+
+\caption{Coefficients of variation for estimates of alternative recruitment estimates (random effects or expected), average fishing mortality at age 6 and 7, and SSB by region and in total from model $M_1$.
+Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.
+Vertical dotted lines indicate the last year of data.}\label{fig:R-F-SSB-CVs}
+\end{figure}
+\end{landscape}
+
+\pagebreak
+
+\begin{figure}
+
 {\centering \includegraphics[width=1\linewidth]{fit_0_jitter_plt}
@@
 
-\begin{landscape}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{self_test_results}
-
-}
-
-\caption{Median relative bias of SSB for the North and South stock components for estimation models where the observation variance of log-indices are fixed and estimation is by maximum marginal likelihood (ML) or Restricted Maximum Likelihood (REML) or where the standard deviation of logistic-normal age composition observations are assumed to be 0.1 (over sqrt(1000)) (Low Error).
-Black and Red dashed lines represent the median of the yeaerly medians and the median across all yearly relative errors, respectively.}\label{fig:self-test-fig}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{BTA_full_fig}
-
-}
-
-\caption{Observations with 95\% confidence intervals (points with vertical lines) and posterior estimates with 95\% confidence intervals (lines with polygons) of bottom temperature anomalies in the north and south regions from model $M_1$.
-Gray vertical line defines the first year that the black sea bass stock is modeled.}\label{fig:bottom-temperature}
-\end{figure}
-
-\end{landscape}
-\pagebreak
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{best_R_Ecov}
-
-}
-
-\caption{Expected and random effect recruitment estimates for the northern stock component. Color of points defined by the corresponding annual bottom temperature anomaly.}\label{fig:BT-Ecov-R}
-\end{figure}
-\pagebreak
-
 \begin{figure}
@@
 
-{\centering \includegraphics[height=0.95\textheight]{Ecov_M1_rel_M0}
-
-}
-
-\caption{Relative differences in posterior estimates of northern region bottom temperature anomalies ($\widehat X$) from the null model without effects on recruitment ($M_0$) and with effects on the northern stock component ($M_1$).}\label{fig:Ecov-M1-rel-M0}
-\end{figure}
-
-\begin{landscape}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.9\textheight]{R_SSB_F_cv_results}
-
-}
-
-\caption{Coefficients of variation for estimates of alternative recruitment estimates (random effects or expected), average fishing mortality at age 6 and 7, and SSB by region and in total from model $M_1$.
-Values in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:R-F-SSB-CVs}
-\end{figure}
-\end{landscape}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_F40_CV}
-
-}
-
-\caption{Coefficients of variation for annual equilibrium average $F$ at ages 6 and 7 that produces the 40\% spawning potential ratio as a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-F40-cvs}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_SSB40_CV}
-
-}
-
-\caption{Coefficients of variation for annual equilibrium SSB$_{40\%}$ as a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations and alternative annual recruitment types.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-SSB40-cvs}
-\end{figure}
-
-\begin{figure}
-
 {\centering \includegraphics[height=0.95\textheight]{proj_F_status_results}
@@
 Estimates in years beyond 2021 are from projecting model $M_1$ under alternative assumptions for bottom temperature anomalies in the northern region.
-Vertical dotted line is the last year of data and polygons represent 95\% confidence intervals.}\label{fig:SSB-status-proj}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_SSB_status_CV}
-
-}
-
-\caption{Coefficients of variation for annual ratios of SSB and equilibrium SSB$_{40\%}$ where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-SSB-status-cvs}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[height=0.95\textheight]{proj_F_status_CV}
-
-}
-
-\caption{Coefficients of variation for annual ratios of average fishing mortality and equilibrium $\bar{F}_{40\%}$ at ages 6 and 7 where the latter is a function of annual expected recruitment or recruitment random effects and annual inputs to $\upphi(\widetilde{F})$ calculations.
-Estimates in years after 2021 are from projecting model $M_1$ under three alternative assumptions for the bottom temperature anomolies.}\label{fig:annual-F-status-cvs}
+Vertical dotted lines indicate the last year of data and polygons represent 95\% confidence intervals.}\label{fig:SSB-status-proj}
 \end{figure}

### 2025-07-01 paper/symboltable.tex
final

@@
 $t_s$ & fraction of the annual time step when spawning occurs for stock $s$\\
-$\delta_{s,j}$ & fraction of the annual time step between $t_s$ and the end of season $j-1$\\
+$\delta_{s,j}$ & fraction of the annual time step between $t_s$ and the beginning of season $j$\\
 $t_m$ & fraction of the annual time step when index $m$ observers the population\\
-$\delta_{m,j}$ & fraction of the annual time step between $t_m$ and the end of season $j-1$\\
+$\delta_{m,j}$ & fraction of the annual time step between $t_m$ and the beginning of season $j$\\
 $\mathbf{N}_{y,a}$ & Abundance at age $a$ in year $y$ in each of the living and mortality states on January 1\\

### 2025-07-24 paper/paper_accepted_crossref_supp.Rmd
separate supp materials

--- 2025-07-01 paper/paper_final.Rmd
@@
     includes:
-      in_header: options_main_final.sty
+      in_header: options_accepted_crossref_supp.sty
 csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
 bibliography: bsb-wham.bib
+link-citations: yes
 ---
@@
 We used a development version ([commit 53e236b](https://github.com/timjmiller/wham/tree/53e236b)) of the WHAM package after the release of version 2.0 for all results.
-All code for fitting models and generating results can be found at [github.com/timjmiller/multi_wham_bsb](https://github.com/timjmiller/multi_wham_bsb).
+All code for fitting models and generating results can be found at [github.com/timjmiller/multi_wham_bsb](https://github.com/timjmiller/multi_wham_bsb), which is permanently archived at [doi:10.5281/zenodo.15785564](https://doi.org/10.5281/zenodo.15785564).
 
@@
 
-As described in the methods section, all data and code for conducting analyses and creating the manuscript are available at [DOI].
+As described in the methods section, all data and code for conducting analyses and creating the manuscript are available at [doi:10.5281/zenodo.15785564](https://doi.org/10.5281/zenodo.15785564).
 

### 2025-07-24 paper/paper_accepted_crossref_supp.tex
separate supp materials

--- 2025-07-01 paper/paper_final.tex
@@
 \usepackage{xr}
-\externaldocument{supp_final}
+\externaldocument{supp_accepted} %allows references from supp_accepted.aux
 \newcommand{\blandscape}{\begin{landscape}}
@@
 
-State-space age-structured stock assessment models can be used to estimate time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al.
-2016b).
-This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks (Punt et al.
-2020).
-The State-space Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed and remains widely used within the International Council for the Exploration of the Sea (ICES) to assess European fish stocks.
-Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada (Perreault et al.
-2020; Varkey et al.
-2022) and to the south, the Woods Hole Assessment Model (WHAM, Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024a).
+State-space age-structured stock assessment models can be used to estimate time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (\protect\hyperlink{ref-nielsenberg14}{Nielsen and Berg 2014}; \protect\hyperlink{ref-cadigan16}{Cadigan 2016}; \protect\hyperlink{ref-milleretal16}{Miller et al.
+2016b}).
+This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks (\protect\hyperlink{ref-puntetal20}{Punt et al.
+2020}).
+The State-space Assessment Model (SAM, \protect\hyperlink{ref-nielsenberg14}{Nielsen and Berg 2014}) continues to be developed and remains widely used within the International Council for the Exploration of the Sea (ICES) to assess European fish stocks.
+Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada (\protect\hyperlink{ref-perreaultetal20}{Perreault et al.
+2020}; \protect\hyperlink{ref-varkeyetal22}{Varkey et al.
+2022}) and to the south, the Woods Hole Assessment Model (WHAM, \protect\hyperlink{ref-stockmiller21}{Stock and Miller 2021}) is now used to assess many fish stocks in the Northwest Atlantic Ocean (\protect\hyperlink{ref-nefsc22}{NEFSC 2022a}, \protect\hyperlink{ref-nefsc22a}{2022b}; \protect\hyperlink{ref-nefsc24}{NEFSC 2024a}).
 
@@
 WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects, to models with several time and age varying process errors and effects of environmental covariates on various population parameters.
-Like SAM, WHAM models are built using the Template Model Builder package (TMB, Kristensen et al.
-2016) which provides a computationally efficient means of fitting an extremely wide class of models with random effects.
+Like SAM, WHAM models are built using the Template Model Builder package (TMB, \protect\hyperlink{ref-kristensenetal16}{Kristensen et al.
+2016}) which provides a computationally efficient means of fitting an extremely wide class of models with random effects.
 WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as ``apparent survival''), fishery and index selectivity, natural mortality, and catchability.
@@
 However, WHAM has only allowed consideration of a single stock and region, and no seasonal changes in stock dynamics.
-Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice (Ying et al.
-2011; Cao et al.
-2014; Bosley et al.
-2022).
-Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks (Punt et al.
-2020).
-We describe here the implementation of these features and other extensions since Stock and Miller (2021) in WHAM version 2.0.
+Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice (\protect\hyperlink{ref-yingetal11}{Ying et al.
+2011}; \protect\hyperlink{ref-caoetal14}{Cao et al.
+2014}; \protect\hyperlink{ref-bosleyetal22}{Bosley et al.
+2022}).
+Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks (\protect\hyperlink{ref-puntetal20}{Punt et al.
+2020}).
+We describe here the implementation of these features and other extensions since Stock and Miller (\protect\hyperlink{ref-stockmiller21}{2021}) in WHAM version 2.0.
 Many of these new configuration options can also be useful when modeling a single stock and region.
 
-We developed this extension of WHAM in concert with a research track assessment process for black sea bass (BSB, \textit{Centropristis striata}; NEFSC 2023), which is a high profile stock in the Northeast U.S.
+We developed this extension of WHAM in concert with a research track assessment process for black sea bass (BSB, \textit{Centropristis striata}; \protect\hyperlink{ref-nefsc23}{NEFSC 2023}), which is a high profile stock in the Northeast U.S.
 and important to both commercial fishing fleets and recreational anglers.
@@
 However, the assessments exhibited strong retrospective patterns.
-The stock was split into two regional components to account for spatial dynamics and improve model diagnostics (ASMFC 2016).
-Additionally, BSB have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (Moser and Shepherd 2009).
-BSB distribution has been linked to warming waters on the northwest shelf (Bell et al.
-2015) and it is hypothesized that the distribution and productivity of the BSB stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare et al.
-2016).
-Thus, a focus of the BSB research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement, and explore environmental drivers of population dynamics (NEFSC 2023).
+The stock was split into two regional components to account for spatial dynamics and improve model diagnostics (\protect\hyperlink{ref-asmfc16}{ASMFC 2016}).
+Additionally, BSB have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (\protect\hyperlink{ref-mosershepherd09}{Moser and Shepherd 2009}).
+BSB distribution has been linked to warming waters on the northwest shelf (\protect\hyperlink{ref-belletal15}{Bell et al.
+2015}) and it is hypothesized that the distribution and productivity of the BSB stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (\protect\hyperlink{ref-hareetal16}{Hare et al.
+2016}).
+Thus, a focus of the BSB research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement, and explore environmental drivers of population dynamics (\protect\hyperlink{ref-nefsc23}{NEFSC 2023}).
 We denote the fish in the north as separate stock components rather than stocks because both components are managed as a single stock, but recognizing differences in migration between fish originating in the two areas.
 
-To illustrate the usage of WHAM 2.0 we apply it to the two stock components of BSB off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process (NEFSC 2023).
+To illustrate the usage of WHAM 2.0 we apply it to the two stock components of BSB off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process (\protect\hyperlink{ref-nefsc23}{NEFSC 2023}).
 However, here we evaluate evidence for alternative hypotheses of temporal variation and effects of bottom temperature, specifically, on recruitment and natural mortality of age 1 individuals.
 
-When modeling multiple stock components and regions, further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al.
-2021).
+When modeling multiple stock components and regions, further dimensionality is added to the class of possible equilbrium reference points (e.g., \protect\hyperlink{ref-kapuretal21}{Kapur et al.
+2021}).
 Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises.
@@
 
-Many of the options and equations of WHAM version 2.0 are the same as those in Stock and Miller (2021), so we will only describe extensions and differences that have occurred since their first description of WHAM.
+Many of the options and equations of WHAM version 2.0 are the same as those in Stock and Miller (\protect\hyperlink{ref-stockmiller21}{2021}), so we will only describe extensions and differences that have occurred since their first description of WHAM.
 The new version of WHAM can model multiple stocks, each with their own movement, harvest, and natural mortality.
@@
 Because individuals may be alive in one of several regions or harvested by one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model.
-Approaches to modeling transitions among these states may treat time discretely (e.g., Arnason 1972; Schwarz et al.
-1993) or continuously (e.g., Hearn et al.
-1987; Commenges 1999; Andersen and Keiding 2002).
+Approaches to modeling transitions among these states may treat time discretely (e.g., \protect\hyperlink{ref-arnason72}{Arnason 1972}; \protect\hyperlink{ref-schwarzetal93}{Schwarz et al.
+1993}) or continuously (e.g., \protect\hyperlink{ref-hearnetal87}{Hearn et al.
+1987}; \protect\hyperlink{ref-commenges99}{Commenges 1999}; \protect\hyperlink{ref-andersenkeiding02}{Andersen and Keiding 2002}).
 We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval with duration \(\delta\) (columns), conditional on being in each of those states at the beginning of the interval (rows).
@@
 When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously.
-The sequential assumption is used widely in spatially explicit models (e.g., Stock Synthesis, Methot and Wetzel 2013).
+The sequential assumption is used widely in spatially explicit models (e.g., Stock Synthesis, \protect\hyperlink{ref-methotwetzel13}{Methot and Wetzel 2013}).
 Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval.
@@
 As in previous versions of WHAM, the apparent survival (and movement) from one year to another after recruitment can be treated deterministically or as functions of random effects.
-The predicted numbers at age \(a\) in year \(y\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (Stock and Miller 2021).
+The predicted numbers at age \(a\) in year \(y\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (\protect\hyperlink{ref-stockmiller21}{Stock and Miller 2021}).
 For ages \(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first \(n_R\) elements of the vector
@@
 \ref{eq:npraa}.
-For the plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is a ``fundamental matrix'\,' derived using the matrix version of the geometric series (Kemeny and Snell 1960).
+For the plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is a ``fundamental matrix'\,' derived using the matrix version of the geometric series (\protect\hyperlink{ref-kemenysnell60}{Kemeny and Snell 1960}).
 Recall that recruitment for a given stock \(s\) only occurs in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
@@
 When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific, but currently are constant across seasons.
-When random effects are used, the same 2DAR1 structure with age and year as described by Stock and Miller (2021) can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
+When random effects are used, the same 2DAR1 structure with age and year as described by Stock and Miller (\protect\hyperlink{ref-stockmiller21}{2021}) can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific.
 We provide details in Appendix B.
@@
 
-The log-normal distributional assumption for aggregate catch observations is the same as Stock and Miller (2021), but the predicted catch is now a function of catch from each stock starting the year in each region and any seasonal movement.
+The log-normal distributional assumption for aggregate catch observations is the same as Stock and Miller (\protect\hyperlink{ref-stockmiller21}{2021}), but the predicted catch is now a function of catch from each stock starting the year in each region and any seasonal movement.
 For a given stock \(s\) at age \(a\), the vector of numbers captured by each fleet over the year \(y\) is
@@
 where \(\widehat C_{f,y}\) is an element of \(\widehat{\mathbf{C}}_y\).
-The predicted numbers caught for each fleet \(f\) (row \(f\) of \(\mathbf{N}_{H,y,a}\)) are used to make predicted age composition observations as described by Stock and Miller (2021).
-Since then, three additional likelihood options for age composition observations have been added: a logistic-normal with AR(1) correlation structure (Francis 2014), the alternative Dirichlet-multinomial parameterization described by Thorson et al.
-(2017), and the multivariate Tweedie (Thorson et al.
-2023).
+The predicted numbers caught for each fleet \(f\) (row \(f\) of \(\mathbf{N}_{H,y,a}\)) are used to make predicted age composition observations as described by Stock and Miller (\protect\hyperlink{ref-stockmiller21}{2021}).
+Since then, three additional likelihood options for age composition observations have been added: a logistic-normal with AR(1) correlation structure (\protect\hyperlink{ref-francis14}{Francis 2014}), the alternative Dirichlet-multinomial parameterization described by Thorson et al.
+(\protect\hyperlink{ref-thorsonetal17}{2017}), and the multivariate Tweedie (\protect\hyperlink{ref-thorsonetal23}{Thorson et al.
+2023}).
 
@@
 where \(q_{m,y}\) is the catchability of the index in year \(y\), \(\text{sel}_{m,y,a}\) is the selectivity, and \(w_{m,y,a}\) is the average weight of individuals at age \(a\) if the index is quantified in biomass and \(w_{m,y,a} = 1\) if the index is quantified in numbers.
-Predicted age composition observations are functions of \(\widehat{I}_{m,y,a}\) as described by Stock and Miller (2021) and the likelihood options are the same as those for catch explained above.
+Predicted age composition observations are functions of \(\widehat{I}_{m,y,a}\) as described by Stock and Miller (\protect\hyperlink{ref-stockmiller21}{2021}) and the likelihood options are the same as those for catch explained above.
 We describe options to model temporal variation in catchability in Appendix B.
@@
 \end{equation}
-and \(h^{(1)}(\log \widetilde{F})\) is the derivative of \(h(\log \widetilde{F})\) with respect to \(\log \widetilde{F}\) (analogous to the approach for SPR-based reference points in Miller and Legault 2017).
+and \(h^{(1)}(\log \widetilde{F})\) is the derivative of \(h(\log \widetilde{F})\) with respect to \(\log \widetilde{F}\) (analogous to the approach for SPR-based reference points in \protect\hyperlink{ref-millerlegault17}{Miller and Legault 2017}).
 The weights for each stock (\(\lambda_s\)) can be specific values supplied by the user or defined to be the proportions of average recruitment for each stock where the years used to average are those the user defines to calculate prevailing equilibrium spawning biomass and yield reference points, which are the product of the defined average recruitment and \(\upphi({\widetilde{F}})\) and \(\upnu({\widetilde{F}})\), respectively.
@@
 
-When a Beverton-Holt or Ricker stock-recruit relationship is assumed, we use the derivative (with respect to \(\widetilde{F}\)) of the weighted sum of stock-specific equilibrium yields (the product of \(\upnu({\widetilde{F}})\) and equilibrium recruitment from the stock-recruit relationship) as \(h(\log \widetilde{F})\) to find \(\widetilde{F}\) which maximizes the weighted total yield across stocks for MSY-based reference points (\(F_{\text{MSY}}\), analogous to the approach for MSY-based reference points in Miller et al.
-2016b).
+When a Beverton-Holt or Ricker stock-recruit relationship is assumed, we use the derivative (with respect to \(\widetilde{F}\)) of the weighted sum of stock-specific equilibrium yields (the product of \(\upnu({\widetilde{F}})\) and equilibrium recruitment from the stock-recruit relationship) as \(h(\log \widetilde{F})\) to find \(\widetilde{F}\) which maximizes the weighted total yield across stocks for MSY-based reference points (\(F_{\text{MSY}}\), analogous to the approach for MSY-based reference points in \protect\hyperlink{ref-milleretal16}{Miller et al.
+2016b}).
 
@@
 BSB are a temperate reef fish in the western Atlantic Ocean ranging the entire east coast of the United States.
-Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity (Able and Stanton Hales, Jr.
-1997; Fabrizio et al.
-2013).
-Fish in this NEUS stock perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring (Musick and Mercer 1977).
-Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude (Moser and Shepherd 2009).
-
-Prior to its 2023 peer-reviewed assessment, the NEUS BSB stock was assessed using the Age-Structured Assessment Program (ASAP) model (Legault and Restrepo 1999), a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
+Fish north of Cape Hatteras, NC, are considered to comprise a single NEUS stock unit, but individual populations exhibit spawning site fidelity (\protect\hyperlink{ref-ablehales97}{Able and Stanton Hales, Jr.
+1997}; \protect\hyperlink{ref-fabrizioetal13}{Fabrizio et al.
+2013}).
+Fish in this NEUS stock perform seasonal migrations out on the continental shelf in the fall and back to their inshore spawning areas during the spring (\protect\hyperlink{ref-musickmercer77}{Musick and Mercer 1977}).
+Analyses of tagging studies have demonstrated that the extent of seasonal migration varies along the coast such that fish from populations off of southern New England and further north move offshore and as far south as the coasts of Virgina and North Carolina, whereas fish in the southern portion of the stock range generally move shorter distances between inshore and offshore areas of similar latitude (\protect\hyperlink{ref-mosershepherd09}{Moser and Shepherd 2009}).
+
+Prior to its 2023 peer-reviewed assessment, the NEUS BSB stock was assessed using the Age-Structured Assessment Program (ASAP) model (\protect\hyperlink{ref-legaultrestrepo99}{Legault and Restrepo 1999}), a single-stock and -region statistical catch-at-age model that estimates all model parameters as fixed effects.
 Northern and southern components of the NEUS BSB stock ascribed to regions divided by the Hudson Canyon were separately modeled in ASAP (Figure \ref{fig:map}).
 Results from the separate ASAP models were combined for a unit-stock assessment.
-The ASAP-based assessments exhibited strong retrospective patterns (Mohn 1999), and exploring alternative modeling approaches for the northern and southern stock components has been a high priority for management.
-
-Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as ``Working Group'') composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the BSB stock using the multi-stock and multi-region extension of WHAM described above (NEFSC 2023).
+The ASAP-based assessments exhibited strong retrospective patterns (\protect\hyperlink{ref-mohn99}{Mohn 1999}), and exploring alternative modeling approaches for the northern and southern stock components has been a high priority for management.
+
+Leading up to the 2023 peer-reviewed assessment, a working group (hereafter referred to as ``Working Group'') composed of scientists from federal, state, and academic institutions determined an optimal data and model configuration for the BSB stock using the multi-stock and multi-region extension of WHAM described above (\protect\hyperlink{ref-nefsc23}{NEFSC 2023}).
 This assessment included the two regions and two stock components and investigated inclusion of hypothesized environmental drivers that were prioritized research recommendations from previous BSB assessments.
@@
 As noted in the description of the general WHAM model, tagging data are not yet included.
-However, the Working Group also fit a Stock Synthesis model (Methot and Wetzel 2013) which included tagging data and provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).
+However, the Working Group also fit a Stock Synthesis model (\protect\hyperlink{ref-methotwetzel13}{Methot and Wetzel 2013}) which included tagging data and provided estimated movement rate parameters and standard errors that were used to configure the priors for WHAM (see Supplementary Materials).
 
@@
 
-All models include observations of winter bottom temperature anomalies for the northern and southern regions from 1959 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius (NEFSC 2023).
+All models include observations of winter bottom temperature anomalies for the northern and southern regions from 1959 to 2021 and standard errors of observations ranged between 0.03 and 0.09 degrees Celsius (\protect\hyperlink{ref-nefsc23}{NEFSC 2023}).
 We provide details about the derivation of these observations in the Supplementary Materials.
@@
 We fit 14 models with alternative assumptions about the effects of bottom temperature covariates, ranging from no effects to effects on both regions for either recruitment or natural mortality at age 1 (Table \ref{tab:model-desc-table}).
-These analyses derive from the hypothesis that bottom temperature affects survival over the first winter where fish turn from age 0 to age 1 on January 1 (Miller et al.
-2016a).
+These analyses derive from the hypothesis that bottom temperature affects survival over the first winter where fish turn from age 0 to age 1 on January 1 (\protect\hyperlink{ref-milleretal16_yoy_survival}{Miller et al.
+2016a}).
 This covariate may be a proxy for temperature prior to January 1 and affect survival during the end of the pre-recruit phase or natural mortality in the early part of the year after becoming age 1.
@@
 Miller et al.
-(2018) showed how inferences of temperature effects on growth or maturity parameters can be very different depending on whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
+(\protect\hyperlink{ref-milleretal18}{2018}) showed how inferences of temperature effects on growth or maturity parameters can be very different depending on whether the compared models with and without the effect also include random effects representing residual temporal variation in parameters.
 Therefore, we also explored whether including temporal random effects on age 1 natural mortality affected inferences on corresponding temperature effects.
@@
 We used a development version (\href{https://github.com/timjmiller/wham/tree/53e236b}{commit 53e236b}) of the WHAM package after the release of version 2.0 for all results.
-All code for fitting models and generating results can be found at \href{https://github.com/timjmiller/multi_wham_bsb}{github.com/timjmiller/multi\_wham\_bsb}.
+All code for fitting models and generating results can be found at \href{https://github.com/timjmiller/multi_wham_bsb}{github.com/timjmiller/multi\_wham\_bsb}, which is permanently archived at \href{https://doi.org/10.5281/zenodo.15785564}{doi:10.5281/zenodo.15785564}.
 
@@
 We calculated Mohn's \(\rho\) for SSB, and average fishing mortality rate at ages 6 and 7.
-Absolute values of Mohn's \(\rho\) near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended (Mohn 1999).
+Absolute values of Mohn's \(\rho\) near 0 imply no pattern in estimation of these quantities as the time-series is sequentially extended (\protect\hyperlink{ref-mohn99}{Mohn 1999}).
 As in Miller et al.
-(2016b), we also assessed the consistency of the AIC-based model selection over retrospective peels to guard against previously noted changes in perception of covariate effects on recruitment with increased length of the time series of observations (Myers 1998).
-This retrospective examination was also recommended by Brooks (2024).
+(\protect\hyperlink{ref-milleretal16}{2016b}), we also assessed the consistency of the AIC-based model selection over retrospective peels to guard against previously noted changes in perception of covariate effects on recruitment with increased length of the time series of observations (\protect\hyperlink{ref-myers98}{Myers 1998}).
+This retrospective examination was also recommended by Brooks (\protect\hyperlink{ref-brooks24}{2024}).
 
@@
 The anomalies for the north and south regions appear highly correlated presumably because of the proximity of the two regions.
-Because the temperature anomalies are treated as latent variables, other data components in the model can affect the estimated anomalies when effects on recruitment are included (e.g., Miller et al.
-2018), but in this case the estimates of anomalies are altered negligibly during the population model years when the effect on recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
+Because the temperature anomalies are treated as latent variables, other data components in the model can affect the estimated anomalies when effects on recruitment are included (e.g., \protect\hyperlink{ref-milleretal18}{Miller et al.
+2018}), but in this case the estimates of anomalies are altered negligibly during the population model years when the effect on recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
 
@@
 In our investigation of bottom temperature effects on recruitment and natural mortality, and time-varying random effects on natural mortality at age 1, we found only including effects on recruitment for the northern stock component to be the best model with respect to AIC.
-BSB in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north (McMahan et al.
-2020).
-Opposite effects of temperature on recruitment or population size would be expected for species in the same general area that are at the southern extent of their range (Gabriel 1992).
-For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer and fall on the Northeast US shelf (Sullivan et al.
-2005; Miller et al.
-2016b).
-
-Our finding regarding bottom temperature effects is consistent with the bottom temperature effects on recruitment in the assessment model accepted during the peer-review process and which is currently used to assess the stock (NEFSC 2023).
+BSB in the NEUS is at the northern extent of its range and the estimated increase in expected recruitment with temperature for the northern stock component is consistent with its range extension to the north (\protect\hyperlink{ref-mcmahanetal20}{McMahan et al.
+2020}).
+Opposite effects of temperature on recruitment or population size would be expected for species in the same general area that are at the southern extent of their range (\protect\hyperlink{ref-gabriel92}{Gabriel 1992}).
+For example, higher recruitment of the Southern New England-Mid-Atlantic stock of yellowtail flounder is correlated with more cold water persistence into summer and fall on the Northeast US shelf (\protect\hyperlink{ref-sullivanetal05}{Sullivan et al.
+2005}; \protect\hyperlink{ref-milleretal16}{Miller et al.
+2016b}).
+
+Our finding regarding bottom temperature effects is consistent with the bottom temperature effects on recruitment in the assessment model accepted during the peer-review process and which is currently used to assess the stock (\protect\hyperlink{ref-nefsc23}{NEFSC 2023}).
 However, AIC suggested some weight for models with temperature effects on recruitment for the southern component and random effects on M at age 1.
-Although this might suggest making inferences on the population by using an ensemble of these and weighting by AIC (Burnham and Anderson 2002), the estimates of assessment outputs relevant for management (e.g., SSB and \(F\)) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
-Continued monitoring of temperature effects on the southern component and investigations of non-linear effects of temperature for both components may be prudent because of the possibility of an optimal temperature range for survival via effects on growth during the pre-recruit phase (Ware and Lambert 1985; Anderson 1988).
+Although this might suggest making inferences on the population by using an ensemble of these and weighting by AIC (\protect\hyperlink{ref-burnhamanderson02}{Burnham and Anderson 2002}), the estimates of assessment outputs relevant for management (e.g., SSB and \(F\)) were similar among the models and weighted estimates would differ very little from the estimates with the lowest AIC.
+Continued monitoring of temperature effects on the southern component and investigations of non-linear effects of temperature for both components may be prudent because of the possibility of an optimal temperature range for survival via effects on growth during the pre-recruit phase (\protect\hyperlink{ref-warelambert85}{Ware and Lambert 1985}; \protect\hyperlink{ref-anderson88}{Anderson 1988}).
 Furthermore, our inferences about effects on natural mortality at age 1 may be influenced by our assumptions of constant median (or just constant, depending on the model) natural mortality rate and movement rates for all ages.
@@
 If reference points should be defined in such a way that they are less variable than the stock over time, expected inputs to the reference points (instead of annual values) would be more appropriate.
-Lower variability in reference points would presumably correlate with lower variability in catch advice which might be desirable (e.g., Cochrane et al.
-1998; Kell et al.
-2005), but the appropriate level of temporal variability in reference points and catch advice will depend on the characteristics of the fish stock, associated fisheries, and management structure (Holland 2010; Punt et al.
-2016).
+Lower variability in reference points would presumably correlate with lower variability in catch advice which might be desirable (e.g., \protect\hyperlink{ref-cochraneetal98}{Cochrane et al.
+1998}; \protect\hyperlink{ref-kelletal05}{Kell et al.
+2005}), but the appropriate level of temporal variability in reference points and catch advice will depend on the characteristics of the fish stock, associated fisheries, and management structure (\protect\hyperlink{ref-holland10}{Holland 2010}; \protect\hyperlink{ref-puntetal16}{Punt et al.
+2016}).
 
@@
 In the projection period, the AR1 models for bottom temperature and recuitment revert to the long-term means of those processes.
-This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypotheses of increasing water temperature and changing environmental conditions (Hare et al.
-2016; Pershing et al.
-2021).
-Because of these concerns, the most recent stock assessment for yellowtail flounder (NEFSC 2024b) used a change point analysis on environmental covariates to determine current conditions.
+This is most likely not an appropriate assumption for stocks in the NEUS given observations and hypotheses of increasing water temperature and changing environmental conditions (\protect\hyperlink{ref-hareetal16}{Hare et al.
+2016}; \protect\hyperlink{ref-pershingetal21}{Pershing et al.
+2021}).
+Because of these concerns, the most recent stock assessment for yellowtail flounder (\protect\hyperlink{ref-nefsc24a}{NEFSC 2024b}) used a change point analysis on environmental covariates to determine current conditions.
 The mean environmental covariate from the current condition was then used as a data input in WHAM to inform short-term projections.
@@
 An obvious limitation to WHAM is the inability to include tagging observations of any type.
-Such observations are critical to estimation of movement parameters (e.g., Goethel et al.
-2019).
+Such observations are critical to estimation of movement parameters (e.g., \protect\hyperlink{ref-goetheletal19}{Goethel et al.
+2019}).
 We used estimates of movement parameters from a companion assessment model to define prior distributions for corresponding movement parameters in the state-space BSB model, but other than the tagging data, both of the assessment models used much the same data which can lead to inappropriate inferences.
@@
 
-Tagging data can also inform mortality rate parameters (Hampton 1991).
-It is well known that natural mortality rates are seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality rates from other assessment model parameters (Lee et al.
-2011; Clark 2022).
+Tagging data can also inform mortality rate parameters (\protect\hyperlink{ref-hampton91}{Hampton 1991}).
+It is well known that natural mortality rates are seldom estimated in assessment models because the observed data often provide little information to distinguish natural mortality rates from other assessment model parameters (\protect\hyperlink{ref-leeetal11}{Lee et al.
+2011}; \protect\hyperlink{ref-clark22}{Clark 2022}).
 Estimation of natural mortality may be even more challenging within state-space assessment models with their greater flexibility from inclusion of time-varying random effects.
-For example, we fixed the mean natural mortality for age 1 fish and both Cadigan (2016) and Stock et al.
-(2021) also estimated natural mortality deviations.
+For example, we fixed the mean natural mortality for age 1 fish and both Cadigan (\protect\hyperlink{ref-cadigan16}{2016}) and Stock et al.
+(\protect\hyperlink{ref-stocketal21}{2021}) also estimated natural mortality deviations.
 However, the simulation studies of Miller et al.
-(In review) suggest estimating natural mortality in WHAM is feasible in some situations with typical assessment data.
+(\protect\hyperlink{ref-milleretalinreview1}{In review}) suggest estimating natural mortality in WHAM is feasible in some situations with typical assessment data.
 Regardless, tagging observations can provide improved inferences and allowing tagging observations in WHAM should be a high priority even for models with a single stock and region.
 
-WHAM version 2.0 only allows models where each stock spawns and recruits in a single region; however, some stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region (Kapur et al.
-2021).
+WHAM version 2.0 only allows models where each stock spawns and recruits in a single region; however, some stocks are assessed with spatial units that are assumed to have a global stock and recruitment model that distributes recruitment to each region (\protect\hyperlink{ref-kapuretal21}{Kapur et al.
+2021}).
 Therefore, extending WHAM to allow such models would be beneficial.
 Finally, because age composition data are scarce for many stocks, incorporating the work by Correa et al.
-(2023) on allowing length composition and modeling growth within WHAM would provide an assessment tool to include process errors in assessment with such data and possibly with multiple stocks and regions.
+(\protect\hyperlink{ref-correaetal23}{2023}) on allowing length composition and modeling growth within WHAM would provide an assessment tool to include process errors in assessment with such data and possibly with multiple stocks and regions.
 
 Our application to BSB is a particular combination of the spatial modeling options described by Berger et al.
-(2024).
+(\protect\hyperlink{ref-bergeretal24}{2024}).
 Other combinations of the described options are possible in WHAM 2.0.
 For example, WHAM 2.0 can accommodate all of fleet structure options described in Table 2 by Berger et al.
-(2024) and we expect more of these options to become available in future versions of WHAM.
+(\protect\hyperlink{ref-bergeretal24}{2024}) and we expect more of these options to become available in future versions of WHAM.
 
@@
 
-As described in the methods section, all data and code for conducting analyses and creating the manuscript are available at {[}DOI{]}.
+As described in the methods section, all data and code for conducting analyses and creating the manuscript are available at \href{https://doi.org/10.5281/zenodo.15785564}{doi:10.5281/zenodo.15785564}.
 
@@
 \caption{The Northwest Atlantic coastal area where the Northeast US black sea bass stock occurs.
-The red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations (after Miller et al.
-2016a) and the bathymetric contour is the 400 m isocline.
+The red line indicates the Hudson Canyon delineation of northern and southern regions and spawning populations (after \protect\hyperlink{ref-milleretal16_yoy_survival}{Miller et al.
+2016a}) and the bathymetric contour is the 400 m isocline.
 Map created with datum NAD83 and Latitude-Longitude coordinate system.}\label{fig:map}
@@
 When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates.
-We obtain the PTM over an interval \(i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
+We obtain the PTM over an interval \(i\) by exponentiating the instantaneous rate matrix (\protect\hyperlink{ref-millerandersen08}{Miller and Andersen 2008})
 \begin{equation*}

### 2025-07-24 paper/paper_accepted_submit.Rmd
separate supp materials

--- 2025-07-24 paper/paper_accepted_crossref_supp.Rmd
@@
     includes:
-      in_header: options_accepted_crossref_supp.sty
+      in_header: options_accepted_submit.sty
 csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
@@
 All models we considered allowed a scalar multiple of the standard deviation of the log aggregate index to be estimated for these indices in the northern and southern regions.
-Although lack of model convergence was observed in many fits for self-test simulations, the estimation of critical model output (e.g., SSB, fishing mortality) was reliable and robust to the poor estimation of these scalars (Figure \ref{fig:self-test-fig}), and estimating these parameters rather than fixing them in the fit to the actual data allows uncertainty in model output to be more properly conveyed.
+Although lack of model convergence was observed in many fits for self-test simulations, the estimation of critical model output (e.g., SSB, fishing mortality) was reliable and robust to the poor estimation of these scalars (Figure S1), and estimating these parameters rather than fixing them in the fit to the actual data allows uncertainty in model output to be more properly conveyed.
 
@@
 The Working Group investigated many alternative assumptions for the probability models and selectivity forms for the eight different sets of age composition observations to reduce residual and retrospective patterns.
-These analyses resulted in use of selectivity random effects for the northern fleet and indices, logistic-normal likelihoods for six sets of age composition observations, and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table \ref{tab:age-comp-sel-table}).
+These analyses resulted in use of selectivity random effects for the northern fleet and indices, logistic-normal likelihoods for six sets of age composition observations, and Dirichlet-multinomial likelihoods for one index and one fleet in the northern region (Table S1).
 
@@
 
-We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock component and no random effects on age 1 natural mortality ($M_1$, Tables \ref{tab:diff-aic-table} and  \ref{tab:aic-wts-table}).
+We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock component and no random effects on age 1 natural mortality ($M_1$, Tables \ref{tab:diff-aic-table} and S2).
 Model $M_1$ consistently had the lowest AIC across retrospective peels where the terminal years of data are removed from the model.
@@
 The residual variation is reduced because the expected recruitment (Eq.
-\ref{eq:expected-recruitment}) is a function of the covariate (Figure \ref{fig:BT-Ecov-R}).
-
-The posterior estimates of the bottom temperature covariate match the observations well because of the high precision of the observed anomalies (Figure \ref{fig:bottom-temperature}).
+\ref{eq:expected-recruitment}) is a function of the covariate (Figure S2).
+
+The posterior estimates of the bottom temperature covariate match the observations well because of the high precision of the observed anomalies (Figure S3).
 The anomalies for the north and south regions appear highly correlated presumably because of the proximity of the two regions.
-Because the temperature anomalies are treated as latent variables, other data components in the model can affect the estimated anomalies when effects on recruitment are included [e.g., @milleretal18], but in this case the estimates of anomalies are altered negligibly during the population model years when the effect on recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
+Because the temperature anomalies are treated as latent variables, other data components in the model can affect the estimated anomalies when effects on recruitment are included [e.g., @milleretal18], but in this case the estimates of anomalies are altered negligibly during the population model years when the effect on recruitment is estimated (Figure S4).
 
@@
 The selectivity at the youngest ages for the two fishery independent indices varied over time but without notable trend.
-The selectivity of fleets and indices in the south increased with age, but were constant over time (Figure \ref{fig:selectivity-south}).
-
-The three sensitivity fits suggest the estimates of stock size and fishing mortality are fairly robust to whether the movement rates are fixed at the values from the companion Stock Synthesis fit or whether no movement is assumed (Figure \ref{fig:sensitivity-plots}).
+The selectivity of fleets and indices in the south increased with age, but were constant over time (Figure S5).
+
+The three sensitivity fits suggest the estimates of stock size and fishing mortality are fairly robust to whether the movement rates are fixed at the values from the companion Stock Synthesis fit or whether no movement is assumed (Figure S6).
 A constant natural mortality rate was estimable, but the estimated rate (0.72) is nearly twice that determined by the Working Group and resulted in generally higher stock sizes and lower fishing mortality rates.
@@
 \ref{eq:newton-obj}).
-Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB$_{40\%}$ for the southern stock component, whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
+Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB$_{40\%}$ for the southern stock component, whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure S7).
 
@@
 However, for the southern component, using the constant expected recruitment is comparable to using median recruitment and the annual variation in the inputs to $\upphi(\widetilde{F})$ indicates no years where SSB was greater than unfished levels throughout the time series.
-The patterns in uncertainty of SSB ratios are the same as those for the SSB reference points (Figure \ref{fig:annual-SSB-status-cvs}).
+The patterns in uncertainty of SSB ratios are the same as those for the SSB reference points (Figure S8).
 
@@
 \ref{eq:newton-obj}, Figure \ref{fig:annual-BRPs}: third row).
-For similar reasons, estimates of uncertainty of $F_{40\%}$ are also similar between the type of recruitment estimates (Figure \ref{fig:annual-F40-cvs}).
+For similar reasons, estimates of uncertainty of $F_{40\%}$ are also similar between the type of recruitment estimates (Figure S9).
 Annual estimates for the northern stock component were lowest in the earliest years of the time series, whereas those for the southern component are relatively stable over time.
@@
 For a given type of recruitment used to calculate $F_{40\%}$, the annual ratios of average $F$ to the corresponding reference point are identical for each stock component and in total because of the selectivity used to define a single F reference point across both regions is exactly the same as that for the annual estimate of fishing mortality at age by fleet (Figure \ref{fig:annual-BRPs}: fourth row).
-As for annual $F_{40\%}$, estimates of uncertainty in the ratios are similar whether expected recruitment or the random effects are used to calculate the reference points (Figure \ref{fig:annual-F-status-cvs}).
+As for annual $F_{40\%}$, estimates of uncertainty in the ratios are similar whether expected recruitment or the random effects are used to calculate the reference points (Figure S10).
 
@@
 Comparing projected recruitment for the northern and southern stock components, the projected expected recruitment has much lower uncertainty than the projected random effects regardless of whether bottom temperature affects recruitment.
-For the northern component, CVs of projected expected recruitments are smaller than those of recruitment random effects, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for future bottom temperature anomalies (Figure \ref{fig:R-F-SSB-CVs}).
+For the northern component, CVs of projected expected recruitments are smaller than those of recruitment random effects, but there are larger differences between CVs of projected expected recruitment assuming AR1 and the other scenarios for future bottom temperature anomalies (Figure S11).
 
@@
 However, the alternative bottom temperature assumptions do affect projections of northern component SSB and the projections do not stabilize when the anomalies are projected to increase linearly.
-When the same AR1 assumption for the bottom temperature anomalies is used in projections, CVs of northern and total SSB projections are larger than when the recent average or linear trend are assumed (Figure \ref{fig:R-F-SSB-CVs}).
+When the same AR1 assumption for the bottom temperature anomalies is used in projections, CVs of northern and total SSB projections are larger than when the recent average or linear trend are assumed (Figure S11).
 
@@
 Estimated reference points in the first few projection years can differ depending on whether the expected recruitment or the random effects are used in their calculation, but longer term projected values are identical because the random effect estimates approach the expected values as the influence of observations diminishes.
-However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures \ref{fig:annual-SSB40-cvs} and \ref{fig:annual-F40-cvs}).
+However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures S7 and S9).
 The reduction in uncertainty using expected recruitment is less for $F_{40\%}$ than SSB$_{40\%}$ and for the northern component which includes bottom temperature effects than the southern component.

### 2025-07-24 paper/paper_accepted_submit.tex
separate supp materials

--- 2025-07-24 paper/paper_accepted_crossref_supp.tex
@@
 \usepackage{upgreek}
-\usepackage{xr}
-\externaldocument{supp_accepted} %allows references from supp_accepted.aux
+% \usepackage{xr}
+% \externaldocument{supp_accepted} %allows references from supp_accepted.aux
 \newcommand{\blandscape}{\begin{landscape}}
@@
 
-We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock component and no random effects on age 1 natural mortality (\(M_1\), Tables \ref{tab:diff-aic-table} and \ref{tab:aic-wts-table}).
+We found the best model of bottom temperature covariate effects included the effect only on recruitment of the northern stock component and no random effects on age 1 natural mortality (\(M_1\), Tables \ref{tab:diff-aic-table} and S2).
 Model \(M_1\) consistently had the lowest AIC across retrospective peels where the terminal years of data are removed from the model.
@@
 The residual variation is reduced because the expected recruitment (Eq.
-\ref{eq:expected-recruitment}) is a function of the covariate (Figure \ref{fig:BT-Ecov-R}).
-
-The posterior estimates of the bottom temperature covariate match the observations well because of the high precision of the observed anomalies (Figure \ref{fig:bottom-temperature}).
+\ref{eq:expected-recruitment}) is a function of the covariate (Figure S2).
+
+The posterior estimates of the bottom temperature covariate match the observations well because of the high precision of the observed anomalies (Figure S3).
 The anomalies for the north and south regions appear highly correlated presumably because of the proximity of the two regions.
 Because the temperature anomalies are treated as latent variables, other data components in the model can affect the estimated anomalies when effects on recruitment are included (e.g., \protect\hyperlink{ref-milleretal18}{Miller et al.
-2018}), but in this case the estimates of anomalies are altered negligibly during the population model years when the effect on recruitment is estimated (Figure \ref{fig:Ecov-M1-rel-M0}).
+2018}), but in this case the estimates of anomalies are altered negligibly during the population model years when the effect on recruitment is estimated (Figure S4).
 
@@
 \ref{eq:newton-obj}).
-Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB\(_{40\%}\) for the southern stock component, whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure \ref{fig:annual-SSB40-cvs}).
+Using the recruitment random effects instead of expected recruitment creates more uncertain estimates of SSB\(_{40\%}\) for the southern stock component, whereas there is little difference for the northern component because there is less difference in the types of recruitment estimates resulting from the covariate effects (Figure S7).
 
@@
 However, for the southern component, using the constant expected recruitment is comparable to using median recruitment and the annual variation in the inputs to \(\upphi(\widetilde{F})\) indicates no years where SSB was greater than unfished levels throughout the time series.
-The patterns in uncertainty of SSB ratios are the same as those for the SSB reference points (Figure \ref{fig:annual-SSB-status-cvs}).
+The patterns in uncertainty of SSB ratios are the same as those for the SSB reference points (Figure S8).
 
@@
 \ref{eq:newton-obj}, Figure \ref{fig:annual-BRPs}: third row).
-For similar reasons, estimates of uncertainty of \(F_{40\%}\) are also similar between the type of recruitment estimates (Figure \ref{fig:annual-F40-cvs}).
+For similar reasons, estimates of uncertainty of \(F_{40\%}\) are also similar between the type of recruitment estimates (Figure S9).
 Annual estimates for the northern stock component were lowest in the earliest years of the time series, whereas those for the southern component are relatively stable over time.
@@
 For a given type of recruitment used to calculate \(F_{40\%}\), the annual ratios of average \(F\) to the corresponding reference point are identical for each stock component and in total because of the selectivity used to define a single F reference point across both regions is exactly the same as that for the annual estimate of fishing mortality at age by fleet (Figure \ref{fig:annual-BRPs}: fourth row).
-As for annual \(F_{40\%}\), estimates of uncertainty in the ratios are similar whether expected recruitment or the random effects are used to calculate the reference points (Figure \ref{fig:annual-F-status-cvs}).
+As for annual \(F_{40\%}\), estimates of uncertainty in the ratios are similar whether expected recruitment or the random effects are used to calculate the reference points (Figure S10).
 
@@
 Estimated reference points in the first few projection years can differ depending on whether the expected recruitment or the random effects are used in their calculation, but longer term projected values are identical because the random effect estimates approach the expected values as the influence of observations diminishes.
-However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures \ref{fig:annual-SSB40-cvs} and \ref{fig:annual-F40-cvs}).
+However, the estimates of uncertainty in the longer term reference points is lower using the expected recruitment rather than the random effects (Figures S7 and S9).
 The reduction in uncertainty using expected recruitment is less for \(F_{40\%}\) than SSB\(_{40\%}\) and for the northern component which includes bottom temperature effects than the southern component.

### 2025-07-24 paper/supp_accepted.Rmd
separate supp materials

--- 2025-07-01 paper/supp_final.Rmd
@@
     includes:
-      in_header: options_final.sty
+      in_header: options_accepted.sty
 csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
 bibliography: bsb-wham.bib
+link-citations: yes
 ---

### 2025-07-24 paper/supp_accepted.tex
separate supp materials

--- 2025-07-01 paper/supp_final.tex
@@
 \usepackage{setspace}
+%\singlespacing
+%\onehalfspacing
 %\doublespacing
@@
 \usepackage{upgreek}
-
+% \usepackage{xr}
+% \externaldocument{supp_final}
 \newcommand{\blandscape}{\begin{landscape}}
@@
 
-The Working Group fit a Stock Synthesis model (Methot and Wetzel 2013) that included tagging data with 2 seasons (6 months each) and 2 regions where a proportion \(\mu^*_1\) of the northern component moves to the south in one season and some proportion \(\mu^*_{2\rightarrow 1}\) move back to the south in the second season (NEFSC 2023).
+The Working Group fit a Stock Synthesis model (\protect\hyperlink{ref-methotwetzel13}{Methot and Wetzel 2013}) that included tagging data with 2 seasons (6 months each) and 2 regions where a proportion \(\mu^*_1\) of the northern component moves to the south in one season and some proportion \(\mu^*_{2\rightarrow 1}\) move back to the south in the second season (\protect\hyperlink{ref-nefsc23}{NEFSC 2023}).
 The seasonal movement matrices for each season are
@@
 The Working Group created bottom temperature observations from a high resolution ocean bottom temperature product by du Pontavice et al.
-(2023).
+(\protect\hyperlink{ref-dupontaviceetal23}{2023}).
 The annual observations for each region are defined by the average over all spatial bottom temperature values for February and March by region and year.
